---
title: "Simulating Functional Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Functional Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

The `fdars` package provides comprehensive tools for simulating functional data. This is useful for:

- **Method validation**: Testing statistical methods on data with known properties
- **Power analysis**: Determining sample sizes and effect sizes
- **Teaching**: Creating examples with controlled characteristics
- **Benchmarking**: Comparing algorithm performance

This vignette covers the Karhunen-Loeve simulation framework and related tools.

## Karhunen-Loeve Simulation

The Karhunen-Loeve (KL) expansion represents a stochastic process as:

$$
X(t) = \mu(t) + \sum_{k=1}^{\infty} \xi_k \phi_k(t)
$$

where:
- $\mu(t)$ is the mean function
- $\phi_k(t)$ are orthonormal eigenfunctions
- $\xi_k \sim N(0, \lambda_k)$ are independent scores with variances $\lambda_k$

For simulation, we truncate to $M$ terms and generate curves via `simFunData()`.

### Basic Example

```{r basic-simulation}
# Define evaluation points
t <- seq(0, 1, length.out = 100)

# Simulate 20 curves with 5 basis functions
fd <- simFunData(n = 20, argvals = t, M = 5, seed = 42)
autoplot(fd) + labs(title = "Simulated Functional Data")
```

### Reproducibility

Set a seed for reproducible results:

```{r reproducibility, fig.height = 3}
fd1 <- simFunData(n = 5, argvals = t, M = 5, seed = 123)
fd2 <- simFunData(n = 5, argvals = t, M = 5, seed = 123)

# Verify they're identical
all.equal(fd1$data, fd2$data)
```

## Eigenfunction Bases

The shape of simulated curves depends on the eigenfunction basis. Use `eFun()` to generate different types.

### Fourier Basis

Best for periodic or smooth oscillating data:

```{r fourier-basis}
phi_fourier <- eFun(t, M = 5, type = "Fourier")
phi_df <- data.frame(t = rep(t, 5),
                     phi = as.vector(phi_fourier),
                     k = factor(rep(1:5, each = length(t))))
ggplot(phi_df, aes(x = t, y = phi, color = k)) +
  geom_line(linewidth = 1) +
  labs(title = "Fourier Eigenfunctions", x = "t", y = expression(phi(t)), color = "k") +
  theme_minimal()
```

```{r fourier-sim}
fd_fourier <- simFunData(n = 20, argvals = t, M = 5,
                         eFun.type = "Fourier", eVal.type = "exponential", seed = 42)
autoplot(fd_fourier) + labs(title = "Fourier Basis Simulation")
```

### Legendre Polynomials

Orthogonal polynomials on [0, 1]:

```{r poly-basis}
phi_poly <- eFun(t, M = 5, type = "Poly")
phi_df <- data.frame(t = rep(t, 5),
                     phi = as.vector(phi_poly),
                     degree = factor(rep(0:4, each = length(t))))
ggplot(phi_df, aes(x = t, y = phi, color = degree)) +
  geom_line(linewidth = 1) +
  labs(title = "Legendre Polynomial Eigenfunctions", x = "t", y = expression(phi(t))) +
  theme_minimal()
```

```{r poly-sim}
fd_poly <- simFunData(n = 20, argvals = t, M = 5,
                      eFun.type = "Poly", eVal.type = "exponential", seed = 42)
autoplot(fd_poly) + labs(title = "Polynomial Basis Simulation")
```

### Wiener Process Eigenfunctions

Eigenfunctions of Brownian motion covariance $K(s,t) = \min(s,t)$:

```{r wiener-basis}
phi_wiener <- eFun(t, M = 5, type = "Wiener")
phi_df <- data.frame(t = rep(t, 5),
                     phi = as.vector(phi_wiener),
                     k = factor(rep(1:5, each = length(t))))
ggplot(phi_df, aes(x = t, y = phi, color = k)) +
  geom_line(linewidth = 1) +
  labs(title = "Wiener Process Eigenfunctions", x = "t", y = expression(phi(t)), color = "k") +
  theme_minimal()
```

```{r wiener-sim}
fd_wiener <- simFunData(n = 20, argvals = t, M = 10,
                        eFun.type = "Wiener", eVal.type = "wiener", seed = 42)
autoplot(fd_wiener) + labs(title = "Wiener Process Simulation")
```

## Eigenvalue Decay

The eigenvalue sequence controls how much each mode contributes to variance.

### Comparing Decay Patterns

```{r eigenvalue-decay}
M <- 20
lambda_lin <- eVal(M, "linear")      # lambda_k = 1/k
lambda_exp <- eVal(M, "exponential") # lambda_k = exp(-k)
lambda_wie <- eVal(M, "wiener")      # lambda_k = 1/((k-0.5)*pi)^2

df_evals <- data.frame(
  k = rep(1:M, 3),
  lambda = c(lambda_lin, lambda_exp, lambda_wie),
  type = rep(c("Linear", "Exponential", "Wiener"), each = M)
)

ggplot(df_evals, aes(x = k, y = lambda, color = type)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_log10() +
  labs(title = "Eigenvalue Decay Patterns",
       x = "Mode k", y = expression(lambda[k]),
       color = "Decay Type") +
  theme(legend.position = "right")
```

### Effect on Curve Smoothness

Faster decay = smoother curves (higher modes contribute less):

```{r decay-comparison, fig.height = 5, fig.width = 10}
# Linear decay - rougher curves
fd_lin <- simFunData(n = 10, argvals = t, M = 10,
                     eFun.type = "Fourier", eVal.type = "linear", seed = 42)
# Exponential decay - smoother curves
fd_exp <- simFunData(n = 10, argvals = t, M = 10,
                     eFun.type = "Fourier", eVal.type = "exponential", seed = 42)

p1 <- autoplot(fd_lin) + labs(title = "Linear Decay (Rougher)")
p2 <- autoplot(fd_exp) + labs(title = "Exponential Decay (Smoother)")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Adding Mean Function

Simulate data with a specified mean:

```{r mean-function}
# Define mean function
mean_fn <- function(t) 2 * sin(2 * pi * t) + t

# Simulate with mean
fd_mean <- simFunData(n = 30, argvals = t, M = 5, mean = mean_fn, seed = 42)

# Visualize with ggplot2
mean_df <- data.frame(t = t, mean = mean_fn(t))
autoplot(fd_mean) +
  geom_line(data = mean_df, aes(x = t, y = mean), color = "red", linewidth = 1.5, inherit.aes = FALSE) +
  labs(title = "Simulated Data with Sinusoidal Mean")
```

## Adding Measurement Error

Use `addError()` to add noise to simulated (or real) data:

```{r add-noise, fig.height = 6, fig.width = 10}
fd_clean <- simFunData(n = 10, argvals = t, M = 5, seed = 42)

# Add different noise levels
fd_low <- addError(fd_clean, sd = 0.1, seed = 123)
fd_med <- addError(fd_clean, sd = 0.3, seed = 123)
fd_high <- addError(fd_clean, sd = 0.5, seed = 123)

p1 <- autoplot(fd_clean) + labs(title = "Clean Data")
p2 <- autoplot(fd_low) + labs(title = "Low Noise (sd = 0.1)")
p3 <- autoplot(fd_med) + labs(title = "Medium Noise (sd = 0.3)")
p4 <- autoplot(fd_high) + labs(title = "High Noise (sd = 0.5)")
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### Curve-Level Noise

Add a constant noise value per curve (useful for modeling measurement bias):

```{r curve-noise}
fd_curve_noise <- addError(fd_clean, sd = 0.5, type = "curve", seed = 123)
autoplot(fd_curve_noise) + labs(title = "Curve-Level Noise")
```

## Creating Sparse/Irregular Data

Use `sparsify()` to simulate irregular sampling:

```{r sparsify}
fd <- simFunData(n = 10, argvals = t, M = 5, seed = 42)

# Create sparse version
ifd <- sparsify(fd, minObs = 10, maxObs = 30, seed = 123)
print(ifd)
```

```{r plot-sparse-sim}
autoplot(ifd) + labs(title = "Sparsified Functional Data")
```

See the "Irregular Sampling" vignette for more details on working with sparse data.

## Multivariate Functional Data

Simulate multiple correlated functional components:

```{r multivariate}
t1 <- seq(0, 1, length.out = 100)
t2 <- seq(0, 0.5, length.out = 50)

mfd <- simMultiFunData(
  n = 15,
  argvals = list(t1, t2),
  M = c(5, 3),
  eFun.type = c("Fourier", "Wiener"),
  eVal.type = c("exponential", "linear"),
  seed = 42
)

print(mfd)
```

```{r multivariate-plot, fig.height = 5, fig.width = 10}
p1 <- autoplot(mfd$components[[1]]) + labs(title = "Component 1 (Fourier)")
p2 <- autoplot(mfd$components[[2]]) + labs(title = "Component 2 (Wiener)")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Comparison with GP Simulation

The package also offers Gaussian Process simulation via `make.gaussian.process()`:

| Aspect | KL (`simFunData`) | GP (`make.gaussian.process`) |
|--------|-------------------|------------------------------|
| Control | Eigenfunctions, eigenvalues | Covariance kernel |
| Interpretation | Modal decomposition | Correlation structure |
| Best for | Known FPCA structure | Known covariance kernel |
| Computation | O(nM) per curve | O(m³) for Cholesky |

```{r gp-comparison, fig.height = 4, fig.width = 10}
# KL simulation
fd_kl <- simFunData(n = 10, argvals = t, M = 10,
                    eFun.type = "Wiener", eVal.type = "wiener", seed = 42)

# GP simulation with Brownian motion kernel
set.seed(42)
fd_gp <- make.gaussian.process(n = 10, t = t,
                                cov = kernel.brownian())

p1 <- autoplot(fd_kl) + labs(title = "KL Simulation (Wiener)")
p2 <- autoplot(fd_gp) + labs(title = "GP Simulation (Brownian Kernel)")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Complete Simulation Recipe

Here's a complete workflow for power analysis:

```{r recipe, fig.height = 4, fig.width = 10}
set.seed(42)

# 1. Simulate two groups with different means
n_per_group <- 30
t <- seq(0, 1, length.out = 100)

mean_group1 <- function(t) sin(2 * pi * t)
mean_group2 <- function(t) sin(2 * pi * t) + 0.5 * cos(4 * pi * t)

group1 <- simFunData(n = n_per_group, argvals = t, M = 5, mean = mean_group1)
group2 <- simFunData(n = n_per_group, argvals = t, M = 5, mean = mean_group2)

# 2. Add measurement noise
group1 <- addError(group1, sd = 0.2)
group2 <- addError(group2, sd = 0.2)

# 3. Visualize with ggplot2
mean_df1 <- data.frame(t = t, mean = mean_group1(t))
mean_df2 <- data.frame(t = t, mean = mean_group2(t))

p1 <- autoplot(group1, alpha = 0.5) +
  geom_line(data = mean_df1, aes(x = t, y = mean), color = "blue", linewidth = 1.5, inherit.aes = FALSE) +
  labs(title = "Group 1")

p2 <- autoplot(group2, alpha = 0.5) +
  geom_line(data = mean_df2, aes(x = t, y = mean), color = "red", linewidth = 1.5, inherit.aes = FALSE) +
  labs(title = "Group 2")

# Side-by-side plots
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Summary

| Function | Purpose |
|----------|---------|
| `simFunData()` | KL simulation with specified eigenfunctions/eigenvalues |
| `simMultiFunData()` | Multivariate functional data simulation |
| `eFun()` | Generate eigenfunction bases (Fourier, Poly, Wiener) |
| `eVal()` | Generate eigenvalue sequences (linear, exponential, wiener) |
| `addError()` | Add Gaussian noise (pointwise or curve-level) |
| `sparsify()` | Create irregular/sparse data from regular |
| `make.gaussian.process()` | GP simulation with covariance kernels |
| `r.brownian()`, `r.ou()`, `r.bridge()` | Specific random process generators |
