---
title: "Advanced Elastic Alignment"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This article covers the advanced alignment methods introduced in fdars-core
0.10.0. These extend the foundational elastic alignment framework (see
[Elastic Curve Alignment](elastic-alignment.html)) with robust estimation,
uncertainty quantification, specialized geometries, and diagnostic tools.

All methods operate on the **elastic manifold** defined by the Square-Root
Slope Function (SRSF) representation $q(t) = \text{sign}(\dot{f}(t))
\sqrt{|\dot{f}(t)|}$, which converts the Fisher-Rao metric to the $L^2$
metric on the Hilbert sphere (Srivastava et al., 2011).

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
set.seed(42)
```

```{r generate-data}
# Simulated curves with amplitude and phase variability
n <- 30
t_grid <- seq(0, 1, length.out = 200)
fd_list <- lapply(seq_len(n), function(i) {
  amp <- rnorm(1, 3, 0.5)
  shift <- rnorm(1, 0.5, 0.08)
  noise <- cumsum(rnorm(200, 0, 0.02))
  amp * dnorm(t_grid, mean = shift, sd = 0.12) + noise
})
fd <- fdata(do.call(rbind, fd_list), argvals = t_grid)
```

---

## 1. Robust Karcher Mean and Median

### Motivation

The standard Karcher mean minimizes the sum of squared elastic distances
and is sensitive to outlier curves. For contaminated data, two alternatives
exist:

- **Karcher median**: minimizes the sum of (unsquared) distances via
  the Weiszfeld algorithm on the elastic manifold -- the functional
  analog of the geometric median.
- **Trimmed Karcher mean**: removes the most distant curves before
  computing the mean.

### Mathematical Background

The **Karcher median** on a Riemannian manifold $M$ is defined as:

$$\tilde{\mu} = \arg\min_{\mu \in M} \sum_{i=1}^{n} d(\mu, f_i)$$

This is solved iteratively using the **Weiszfeld algorithm** adapted to
manifolds (Fletcher et al., 2009): at each step, compute the weighted
Frechet mean where weights are inversely proportional to distances from
the current estimate.

The **trimmed Karcher mean** removes the $k$ most distant curves (in
elastic distance) before computing the standard Karcher mean, following
the approach of Cuesta-Albertos & Fraiman (2007) extended to the elastic
manifold.

```{r robust-karcher}
# Standard Karcher mean (reference)
km <- karcher.mean(fd, max.iter = 15)

# Karcher median (robust to outliers)
kmed <- karcher.median(fd, max.iter = 15)

# Trimmed Karcher mean (remove 10% most distant curves)
km_trimmed <- robust.karcher.mean(fd, trim = 0.1, max.iter = 15)

# Compare the three estimates
km_vals <- as.numeric(km$mean$data[1, ])
kmed_vals <- as.numeric(kmed$mean$data[1, ])
ktrim_vals <- as.numeric(km_trimmed$mean$data[1, ])
# Ensure all same length (use km argvals as reference)
km_t <- as.numeric(km$mean$argvals)
df_means <- data.frame(
  t = rep(km_t, 3),
  value = c(km_vals, kmed_vals[seq_along(km_vals)],
            ktrim_vals[seq_along(km_vals)]),
  method = rep(c("Karcher Mean", "Karcher Median", "Trimmed Mean (10%)"),
               each = length(km_t))
)
ggplot(df_means, aes(x = .data$t, y = .data$value, color = .data$method)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Robust vs Standard Karcher Mean",
       x = "t", y = "f(t)", color = NULL) +
  theme(legend.position = "bottom")
```

**References:**

- Fletcher, P. T., Venkatasubramanian, S., & Joshi, S. (2009).
  The geometric median on Riemannian manifolds with application to
  robust atlas estimation. *NeuroImage*, 45(1), S143--S152.
  DOI: [10.1016/j.neuroimage.2008.10.052](https://doi.org/10.1016/j.neuroimage.2008.10.052)
- Cuesta-Albertos, J. A. & Fraiman, R. (2007).
  Impartial trimmed k-means for functional data.
  *Computational Statistics & Data Analysis*, 51(10), 4864--4877.

---

## 2. Elastic Depth

### Motivation

Functional depth measures how "central" a curve is in a sample. Standard
depth (e.g., MBD) ignores phase variability -- two curves may have similar
depth despite one being time-warped. **Elastic depth** decomposes the
depth into amplitude and phase components using elastic distances.

### Method

For each curve $f_i$, the elastic depth is computed as:

$$D_{\text{elastic}}(f_i) = \frac{1}{1 + \bar{d}_{\text{elastic}}(f_i)}$$

where $\bar{d}_{\text{elastic}}(f_i)$ is the average elastic distance
from $f_i$ to all other curves. The decomposition into amplitude and
phase depths uses the aligned vs unaligned distances:

- **Amplitude depth**: based on $L^2$ distances after alignment
- **Phase depth**: based on the warping function complexity

```{r elastic-depth}
# Elastic depth with amplitude/phase decomposition
ed <- elastic.depth(fd)

cat("Amplitude depth range:", round(range(ed$amplitude_depth), 3), "\n")
cat("Phase depth range:    ", round(range(ed$phase_depth), 3), "\n")
cat("Combined depth range: ", round(range(ed$combined_depth), 3), "\n")
cat("Deepest curve (most central):", which.max(ed$combined_depth), "\n")
```

**References:**

- Lopez-Pintado, S. & Romo, J. (2009). On the concept of depth for
  functional data. *Journal of the American Statistical Association*,
  104(486), 718--734.
  DOI: [10.1198/jasa.2009.0108](https://doi.org/10.1198/jasa.2009.0108)
- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer. Chapter 9: Statistical Summaries.

---

## 3. SRVF Outlier Detection

### Motivation

Outlier detection in the elastic framework accounts for both amplitude
and phase anomalies. A curve can be an outlier because it has an unusual
shape (amplitude outlier), an unusual timing (phase outlier), or both.

### Method

Elastic outlier detection computes the distance from each curve to a
reference (mean or median) and applies Tukey fence thresholding:

$$\text{outlier if } d_{\text{elastic}}(f_i, \mu) > Q_3 + k \cdot \text{IQR}$$

where $k$ is typically 1.5. The amplitude/phase decomposition classifies
outliers into types.

```{r elastic-outlier}
# Outlier detection with amplitude/phase decomposition
outliers <- elastic.outlier.detection(fd)

cat("Outlier indices:", if (length(outliers$outlier_indices) > 0)
  paste(outliers$outlier_indices, collapse = ", ") else "none", "\n")
cat("Threshold:", round(outliers$threshold, 4), "\n")
cat("Distance range:", round(range(outliers$distances), 4), "\n")
```

**References:**

- Xie, W., Kurtek, S., Bharath, K., & Sun, Y. (2017). A geometric
  approach to visualization of variability in functional data.
  *Journal of the American Statistical Association*, 112(519), 979--993.
  DOI: [10.1080/01621459.2016.1256813](https://doi.org/10.1080/01621459.2016.1256813)
- Dai, W. & Genton, M. G. (2018). Functional boxplots. *Journal of
  Computational and Graphical Statistics*, 20(2), 316--334.

---

## 4. Shape Confidence Intervals

### Motivation

After computing the Karcher mean, how certain are we about its shape?
**Shape confidence intervals** provide bootstrap confidence bands for
the elastic mean, quantifying estimation uncertainty.

### Method

1. Resample $B$ bootstrap samples from the original curves
2. For each sample, compute the Karcher mean
3. Align all bootstrap means to the original mean
4. Compute pointwise percentile bands (e.g., 2.5% and 97.5%)

The alignment step is crucial: without it, phase variability between
bootstrap means inflates the bands artificially.

```{r shape-ci, fig.height = 4}
# Bootstrap confidence bands for the Karcher mean
ci <- shape.confidence.interval(fd, n.bootstrap = 50, max.iter = 10)

df_ci <- data.frame(
  t = rep(t_grid, 3),
  value = c(ci$lower_band, as.numeric(ci$mean$data[1, ]), ci$upper_band),
  type = rep(c("Lower 95%", "Mean", "Upper 95%"), each = length(t_grid))
)

ggplot(df_ci, aes(x = .data$t, y = .data$value, linetype = .data$type)) +
  geom_line(color = "#4A90D9", linewidth = 0.8) +
  scale_linetype_manual(values = c("Lower 95%" = "dashed",
                                    "Mean" = "solid",
                                    "Upper 95%" = "dashed")) +
  labs(title = "Shape Confidence Interval for Karcher Mean",
       x = "t", y = "f(t)", linetype = NULL) +
  theme(legend.position = "bottom")
```

**References:**

- Kurtek, S. & Marron, J. S. (2017). Elastic shape analysis of
  boundaries of planar objects. *Annals of Applied Statistics*, 11(3),
  1490--1521. DOI: [10.1214/17-AOAS1059](https://doi.org/10.1214/17-AOAS1059)
- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer. Chapter 8: Statistical Analysis of Shapes.

---

## 5. Bayesian Alignment

### Motivation

Standard dynamic programming alignment yields a point estimate of the
warping function. **Bayesian alignment** provides a posterior distribution
over warping functions, enabling uncertainty quantification: credible
bands on aligned curves, posterior predictive distributions, and model
comparison.

### Method

The method uses the **preconditioned Crank-Nicolson (pCN)** MCMC sampler
on the Hilbert sphere (Cotter et al., 2013), adapted for functional
alignment by Cheng et al. (2016). The prior over warping functions is
a Dirichlet process or Gaussian process on the sphere of SRSFs; the
likelihood measures the $L^2$ fit between the SRSFs after warping.

The pCN proposal

$$\gamma^* = \sqrt{1 - \beta^2} \, \gamma^{(k)} + \beta \, \xi, \quad \xi \sim \Pi_0$$

ensures dimension-independent acceptance rates, making the algorithm
scalable to fine time grids.

```{r bayesian-align}
# Bayesian alignment of two curves
ba <- bayesian.align.pair(fd$data[1, ], fd$data[2, ],
                           argvals = t_grid,
                           n.samples = 500, burn.in = 100)

cat("Acceptance rate:", round(ba$acceptance_rate, 3), "\n")

# Plot original vs Bayesian-aligned
df_ba <- data.frame(
  t = rep(t_grid, 3),
  value = c(fd$data[1, ], fd$data[2, ], ba$f_aligned_mean),
  curve = rep(c("f1 (reference)", "f2 (original)", "f2 (Bayesian aligned)"),
              each = length(t_grid))
)
ggplot(df_ba, aes(x = .data$t, y = .data$value, color = .data$curve)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Bayesian Pairwise Alignment",
       subtitle = sprintf("Acceptance rate: %.1f%%", 100 * ba$acceptance_rate),
       x = "t", y = "f(t)", color = NULL) +
  theme(legend.position = "bottom")
```

**References:**

- Cheng, W., Dryden, I. L., & Huang, X. (2016). Bayesian registration
  of functions and curves. *Bayesian Analysis*, 11(2), 447--481.
  DOI: [10.1214/15-BA957](https://doi.org/10.1214/15-BA957)
- Cotter, S. L., Roberts, G. O., Stuart, A. M., & White, D. (2013).
  MCMC methods for functions: modifying old algorithms to make them
  faster. *Statistical Science*, 28(3), 424--446.
  DOI: [10.1214/13-STS421](https://doi.org/10.1214/13-STS421)
- Lu, Y., Herbei, R., & Kurtek, S. (2017). Bayesian registration of
  functions with a Gaussian process prior. *Journal of Computational and
  Graphical Statistics*, 26(4), 894--904.

---

## 6. Multi-Resolution Alignment

### Motivation

Dynamic programming alignment has $O(T^2)$ complexity, making it slow
for long curves ($T > 1000$). **Multi-resolution alignment** performs
a coarse alignment on a subsampled grid, then refines on the original
resolution -- achieving near-optimal alignment in $O(T \log T)$ time.

### Method

1. **Coarse stage**: Subsample the curve to $T/k$ points and run DP
   alignment to get an approximate warping path
2. **Fine stage**: Use the coarse warp as initialization for gradient
   refinement (e.g., L-BFGS) on the full-resolution grid

This is related to the FastDTW approach (Salvador & Chan, 2007) but
adapted to the elastic (SRSF) framework.

```{r multires}
# Multi-resolution alignment (faster for long curves)
mr <- elastic.align.pair.multires(fd$data[1, ], fd$data[2, ],
                                   argvals = t_grid,
                                   coarsen.factor = 4)

cat("Multi-res distance:", round(mr$distance, 4), "\n")

# Compare with standard DP alignment
std <- elastic.align(fdata(matrix(c(fd$data[1, ], fd$data[2, ]), nrow = 2),
                            argvals = t_grid))
cat("Standard DP distance:", round(std$distance[1], 4), "\n")
```

**References:**

- Salvador, S. & Chan, P. (2007). Toward accurate dynamic time warping
  in linear time and space. *Intelligent Data Analysis*, 11(5), 561--580.
  DOI: [10.3233/IDA-2007-11508](https://doi.org/10.3233/IDA-2007-11508)
- Robinson, D. T. (2012). Functional data analysis and partial shape
  matching in the square root velocity framework. PhD Thesis, Florida
  State University.

---

## 7. Closed Curve Alignment

### Motivation

Closed curves (e.g., cell boundaries, periodic signals) require alignment
that accounts for the **circular starting point ambiguity**: the same
closed curve can be parameterized starting from any point on its boundary.
Standard alignment fixes the starting point and misses this degree of
freedom.

### Method

Closed curve alignment adds an optimization over the starting point
rotation $\theta$ in addition to the warping function $\gamma$:

$$d(f_1, f_2) = \min_{\gamma, \theta} \|q_1 - (q_2 \circ R_\theta \circ \gamma) \sqrt{\dot{\gamma}}\|_{L^2}$$

The optimization uses a **coarse-to-fine** search: evaluate elastic
distances at $K$ uniformly-spaced rotations, then refine the best
candidate with gradient descent.

```{r closed-align}
# Create periodic curves (simulate closed shapes)
closed1 <- sin(2 * pi * t_grid) + 0.3 * sin(4 * pi * t_grid)
closed2 <- sin(2 * pi * (t_grid - 0.1)) + 0.4 * sin(4 * pi * (t_grid - 0.1))

# Align closed curves
ca <- elastic.align.pair.closed(closed1, closed2, t_grid)
cat("Closed alignment distance:", round(ca$distance, 4), "\n")
cat("Optimal rotation index:", ca$optimal_rotation, "\n")

# Karcher mean for closed curves
fd_closed <- fdata(rbind(closed1, closed2,
  sin(2*pi*(t_grid+0.05)) + 0.35*sin(4*pi*(t_grid+0.05))),
  argvals = t_grid)
km_c <- karcher.mean.closed(fd_closed, max.iter = 10)
cat("Closed Karcher mean converged:", km_c$converged, "\n")
```

**References:**

- Srivastava, A., Klassen, E., Joshi, S. H., & Jermyn, I. H. (2011).
  Shape analysis of elastic curves in Euclidean spaces. *IEEE Transactions
  on Pattern Analysis and Machine Intelligence*, 33(7), 1415--1428.
  DOI: [10.1109/TPAMI.2010.184](https://doi.org/10.1109/TPAMI.2010.184)
- Kurtek, S., Srivastava, A., Klassen, E., & Ding, Z. (2012). Statistical
  modeling of curves using shapes and related features. *Journal of the
  American Statistical Association*, 107(499), 1152--1165.

---

## 8. Elastic Partial Matching

### Motivation

When comparing curves of different lengths or where only a portion of
one curve matches the other (e.g., a subsequence search), standard
alignment forces the full curves to match. **Partial matching** finds
the best-aligned subcurve of a longer target.

### Method

The algorithm searches over all subsequences $[a, b] \subset [0, 1]$
of the target curve and computes the elastic distance to the query
curve, returning the optimal segment and alignment:

$$d_{\text{partial}}(f_1, f_2) = \min_{0 \leq a < b \leq 1}
\min_\gamma \|q_1 - (q_2|_{[a,b]} \circ \gamma) \sqrt{\dot{\gamma}}\|$$

A sliding variable-length window search avoids the $O(T^3)$ brute-force
cost.

```{r partial-match}
# Template: short curve (first 50 points of curve 1)
template <- fd$data[1, 1:50]
target <- fd$data[2, ]

pm <- elastic.partial.match(template, target,
                             t_grid[1:50], t_grid,
                             min.span = 0.2)

cat("Match range: index", pm$start_index, "to", pm$end_index, "\n")
cat("Domain fraction:", round(pm$domain_fraction, 3), "\n")
cat("Match distance:", round(pm$distance, 4), "\n")
```

**References:**

- Robinson, D. T. (2012). *Functional data analysis and partial shape
  matching in the square root velocity framework*. PhD Thesis, Florida
  State University.
- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer. Section 4.5: Partial Matching.

---

## 9. Transfer Alignment

### Motivation

When curves come from two different populations (e.g., healthy vs
diseased, two manufacturing sites), direct alignment may fail because
the population-level shape differences confound the within-subject
warping. **Transfer alignment** aligns curves from a target population
to a source population's coordinate system via bridging warps.

### Method

1. Compute Karcher means $\mu_S$ and $\mu_T$ for source and target
2. Find the bridging warp $\gamma_B$ that aligns $\mu_T$ to $\mu_S$
3. For each target curve $f_i^T$, compute its within-target warp
   $\gamma_i^T$ (aligning to $\mu_T$)
4. Compose: $\gamma_i^{S} = \gamma_B \circ \gamma_i^T$
5. Apply $\gamma_i^{S}$ to place $f_i^T$ in the source coordinate system

```{r transfer}
# Split data into two "populations"
fd_source <- fdata(fd$data[1:15, ], argvals = t_grid)
fd_target <- fdata(fd$data[16:30, ], argvals = t_grid)

tr <- transfer.alignment(fd_source, fd_target, max.iter = 10)

cat("Aligned", nrow(tr$aligned$data), "target curves to source frame\n")
cat("Distance range:", round(range(tr$distances), 4), "\n")
```

**References:**

- Matuk, J., Bharath, K., Chkrebtii, O., & Kurtek, S. (2022).
  Bayesian framework for simultaneous registration and estimation of
  noisy, sparse, and fragmented functional data. *Journal of the American
  Statistical Association*, 117(540), 1964--1980.
  DOI: [10.1080/01621459.2021.1893179](https://doi.org/10.1080/01621459.2021.1893179)

---

## 10. Curve Geodesic Interpolation

### Motivation

Given two curves, what do the "intermediate" curves look like? A
**geodesic** on the elastic manifold is the shortest path between two
shapes, providing a natural interpolation that smoothly morphs one curve
into another through both amplitude and phase changes.

### Method

The geodesic path between curves $f_1$ and $f_2$ in the elastic
framework is computed by:

1. **Amplitude interpolation** in SRSF space: linearly interpolate
   between $q_1$ and $q_2 \circ \gamma^*$ in $L^2$
2. **Phase interpolation** on the Hilbert sphere: geodesic path
   between $\text{id}$ and $\gamma^*$ on the diffeomorphism group

At parameter $\alpha \in [0, 1]$:
$$q_\alpha = (1 - \alpha) q_1 + \alpha (q_2 \circ \gamma^* \sqrt{\dot{\gamma}^*})$$
$$\gamma_\alpha = \text{exp}_{\text{id}}(\alpha \cdot \log_{\text{id}}(\gamma^*))$$

```{r geodesic, fig.height = 4}
# Geodesic path between two curves (8 intermediate steps)
geo <- curve.geodesic(fd$data[1, ], fd$data[5, ],
                       argvals = t_grid, n.points = 8)

# Plot the geodesic path
n_steps <- nrow(geo$curves$data)
df_geo <- data.frame(
  t = rep(t_grid, n_steps),
  value = as.vector(t(geo$curves$data)),
  step = rep(seq_len(n_steps), each = length(t_grid)),
  alpha = rep(geo$parameter_values, each = length(t_grid))
)

ggplot(df_geo, aes(x = .data$t, y = .data$value,
                    group = .data$step, color = .data$alpha)) +
  geom_line(linewidth = 0.6) +
  scale_color_gradientn(
    colours = c("#313695", "#ABD9E9", "#FEE090", "#A50026")) +
  labs(title = "Geodesic Path Between Two Curves",
       subtitle = "Smooth interpolation in elastic metric space",
       x = "t", y = "f(t)", color = expression(alpha))
```

**References:**

- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer. Chapter 4: Elastic Shape Analysis of Curves.
- Srivastava, A., Klassen, E., Joshi, S. H., & Jermyn, I. H. (2011).
  Shape analysis of elastic curves in Euclidean spaces. *IEEE TPAMI*,
  33(7), 1415--1428.

---

## 11. Gaussian Generative Model

### Motivation

After separating amplitude and phase variability via FPCA, we can build
a **generative model** that synthesizes new curves by sampling from the
estimated amplitude and phase distributions. This is useful for data
augmentation, simulation studies, and testing.

### Method

The generative model fits Gaussian distributions to the vertical (amplitude)
and horizontal (phase) FPC scores separately:

1. Align curves to the Karcher mean and compute vertical FPCA on
   aligned curves and horizontal FPCA on warping functions
2. Fit $\mathbf{z}_V \sim N(\mathbf{0}, \Lambda_V)$ to amplitude scores
   and $\mathbf{z}_H \sim N(\mathbf{0}, \Lambda_H)$ to phase scores
3. Generate new curves by sampling scores and reconstructing:
   $f_{\text{new}} = \bar{f} + \sum_k z_{V,k} \xi_k$ warped by
   $\gamma_{\text{new}} = \text{exp}(\sum_k z_{H,k} \psi_k)$

The **joint model** preserves amplitude-phase correlations by fitting
a joint Gaussian to the concatenated score vectors.

```{r generative, fig.height = 4}
# Fit generative model from Karcher mean result
gm <- gauss.model(km, n.components = 3, n.samples = 20)

cat("Generated", nrow(gm$samples$data), "synthetic curves\n")

# Plot original vs generated
df_gen <- rbind(
  data.frame(t = rep(t_grid, nrow(fd$data)),
             value = as.vector(t(fd$data)),
             curve = rep(seq_len(nrow(fd$data)), each = length(t_grid)),
             type = "Original"),
  data.frame(t = rep(t_grid, nrow(gm$samples$data)),
             value = as.vector(t(gm$samples$data)),
             curve = rep(seq_len(nrow(gm$samples$data)) + 100, each = length(t_grid)),
             type = "Generated")
)

ggplot(df_gen, aes(x = .data$t, y = .data$value,
                    group = .data$curve, color = .data$type)) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  scale_color_manual(values = c(Original = "#4A90D9", Generated = "#D55E00")) +
  labs(title = "Gaussian Generative Model: Original vs Synthetic Curves",
       x = "t", y = "f(t)", color = NULL) +
  theme(legend.position = "bottom")
```

**References:**

- Tucker, J. D., Wu, W., & Srivastava, A. (2013). Generative models for
  functional data using phase and amplitude separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.
  DOI: [10.1016/j.csda.2012.12.001](https://doi.org/10.1016/j.csda.2012.12.001)

---

## 12. Peak Persistence Diagram

### Motivation

The alignment regularization parameter $\lambda$ controls the trade-off
between perfect alignment and smoothness of warping functions. Choosing
$\lambda$ is typically ad hoc. **Peak persistence** provides a topology-based
method: by sweeping $\lambda$ from 0 to $\infty$ and tracking when peaks
in the Karcher mean appear ("birth") and disappear ("death"), persistent
peaks indicate genuine features while transient peaks indicate noise.

### Method

The algorithm computes Karcher means at a grid of $\lambda$ values and
builds a persistence diagram:

1. For each $\lambda$, compute the Karcher mean $\bar{f}_\lambda$
2. Detect peaks in $\bar{f}_\lambda$ and track them across $\lambda$ values
3. A peak is "born" at the $\lambda$ where it first appears and "dies"
   at the $\lambda$ where it merges with a neighbor
4. The optimal $\lambda$ is chosen where all persistent peaks are present
   but noise peaks have died

```{r persistence}
# Peak persistence diagram for lambda selection
pers <- peak.persistence(fd, lambdas = seq(0, 3, length.out = 15),
                          max.iter = 8)

cat("Optimal lambda:", round(pers$optimal_lambda, 3), "\n")
cat("Peak counts across lambdas:", pers$peak_counts, "\n")
```

**References:**

- Edelsbrunner, H. & Harer, J. (2010). *Computational Topology: An
  Introduction*. American Mathematical Society.
- Chazal, F., de Silva, V., Glisse, M., & Oudot, S. (2016). The
  structure and stability of persistence modules. Springer.
- Tucker, J. D., Wu, W., & Srivastava, A. (2014). Analysis of
  proteomics data: phase-amplitude separation using an extended
  Fisher-Rao metric. *Electronic Journal of Statistics*, 8, 1724--1733.

---

## 13. Horizontal FPNS

### Motivation

Standard horizontal FPCA linearizes the space of warping functions
(diffeomorphisms) and performs PCA in the tangent space. This linear
approximation can be poor when warps are far from the identity.
**Functional Principal Nested Spheres (FPNS)** performs PCA directly
on the Hilbert sphere, extracting nonlinear principal directions.

### Method

FPNS iteratively extracts principal geodesic directions on the sphere
of SRSF-transformed warping functions:

1. Map warping functions to the Hilbert sphere via $\psi = \sqrt{\dot{\gamma}}$
2. Find the great circle (geodesic subspace) that maximizes the
   explained variance of the projected data
3. Project the data onto this subspace and repeat on the residuals
   (on the orthogonal subsphere)

This produces nested spheres $S^{d-1} \supset S^{d-2} \supset \cdots$
that capture decreasing levels of phase variability.

```{r fpns}
# Horizontal FPNS (nonlinear PCA on warping functions)
fpns_result <- horiz.fpns(km, n.components = 3)

cat("Explained variance:", round(fpns_result$explained_variance, 4), "\n")
cat("Score dimensions:", paste(dim(fpns_result$scores), collapse = " x "), "\n")
```

**References:**

- Jung, S., Dryden, I. L., & Marron, J. S. (2012). Analysis of
  principal nested spheres. *Biometrika*, 99(3), 551--568.
  DOI: [10.1093/biomet/ass022](https://doi.org/10.1093/biomet/ass022)
- Tucker, J. D. (2014). *Functional component analysis and regression
  using elastic methods*. PhD Thesis, Florida State University.

---

## 14. Multivariate Curve Alignment

### Motivation

When curves are vector-valued -- e.g., 3D motion trajectories
$(x(t), y(t), z(t))$ or multivariate sensor data -- alignment must
find a single warping function that simultaneously aligns all components.
The scalar elastic framework extends naturally by using the
multivariate SRSF.

### Method

For an $\mathbb{R}^d$-valued curve $\mathbf{f}(t) = (f_1(t), \ldots, f_d(t))$,
the multivariate SRSF is:

$$\mathbf{q}(t) = \frac{\dot{\mathbf{f}}(t)}{\sqrt{\|\dot{\mathbf{f}}(t)\|}}$$

The elastic distance and Karcher mean generalize directly. The alignment
finds a single $\gamma$ that minimizes:

$$\sum_{j=1}^{d} \|q_{1,j} - (q_{2,j} \circ \gamma) \sqrt{\dot{\gamma}}\|^2$$

```{r nd-karcher, eval = FALSE}
# Multivariate Karcher mean (e.g., for 3D motion data)
# Note: N-D alignment requires FdCurveSet inputs -- R wrappers pending
km_nd <- karcher.mean.nd(fd_nd, max.iter = 20)

# Covariance estimation in the tangent space
cov_nd <- karcher.covariance.nd(fd_nd, km_nd)

# Multivariate elastic PCA
pca_nd <- pca.nd(fd_nd, km_nd, n_components = 3)
```

**References:**

- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer. Chapter 5: Shapes of Planar Curves.
- Kurtek, S. (2017). A geometric approach to pairwise Bayesian alignment
  of functional data using importance sampling. *Electronic Journal of
  Statistics*, 11(1), 502--531.

---

## Feature Summary

| Feature | Use Case | Key Function | Section |
|---------|----------|--------------|---------|
| Robust Karcher mean | Outlier-contaminated data | `karcher.median`, `robust.karcher.mean` | 1 |
| Elastic depth | Centrality with phase awareness | `elastic.depth` | 2 |
| SRVF outlier detection | Amplitude/phase outlier classification | `elastic.outlier.detection` | 3 |
| Shape confidence intervals | Uncertainty on Karcher mean | `shape.confidence.interval` | 4 |
| Bayesian alignment | Posterior over warping functions | `bayesian.align.pair` | 5 |
| Multi-resolution alignment | Fast alignment for long curves | `elastic.align.pair.multires` | 6 |
| Closed curve alignment | Periodic/closed shapes | `elastic.align.pair.closed` | 7 |
| Elastic partial matching | Subsequence search | `elastic.partial.match` | 8 |
| Transfer alignment | Cross-population alignment | `transfer.alignment` | 9 |
| Curve geodesic | Shape interpolation | `curve.geodesic` | 10 |
| Gaussian generative model | Synthetic curve generation | `gauss.model` | 11 |
| Peak persistence | Automatic lambda selection | `peak.persistence` | 12 |
| Horizontal FPNS | Nonlinear phase PCA | `horiz.fpns` | 13 |
| Multivariate alignment | Vector-valued curves | `karcher.mean.nd` | 14 |

## See Also

- [Elastic Curve Alignment](elastic-alignment.html) -- foundational
  SRSF framework and standard alignment
- [Comparing Alignment Methods](alignment-comparison.html) -- when to
  use elastic vs landmark vs TSRVF alignment
- [Elastic FPCA](elastic-fpca.html) -- vertical, horizontal, and joint
  FPCA after alignment
- [Shape Analysis](shape-analysis.html) -- quotient space analysis
  under reparameterization, translation, and scale

## References

- Srivastava, A. & Klassen, E. P. (2016). *Functional and Shape Data
  Analysis*. Springer Series in Statistics.
  DOI: [10.1007/978-1-4939-4020-2](https://doi.org/10.1007/978-1-4939-4020-2)
- Srivastava, A., Klassen, E., Joshi, S. H., & Jermyn, I. H. (2011).
  Shape analysis of elastic curves in Euclidean spaces. *IEEE TPAMI*,
  33(7), 1415--1428.
- Tucker, J. D., Wu, W., & Srivastava, A. (2013). Generative models for
  functional data using phase and amplitude separation. *CSDA*, 61, 50--66.
- Marron, J. S., Ramsay, J. O., Sangalli, L. M., & Srivastava, A. (2015).
  Functional data analysis of amplitude and phase variation. *Statistical
  Science*, 30(4), 468--484.
  DOI: [10.1214/15-STS524](https://doi.org/10.1214/15-STS524)
