---
title: "Advanced Statistical Process Monitoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Statistical Process Monitoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

The basic $T^2$/SPE Shewhart chart from `spm.phase1()` is a good starting
point, but it has well-known limitations. It uses a single fixed control limit,
treats each observation independently, and assumes the training data is clean.
In practice, functional process data often involves contaminated Phase I
samples, small persistent drifts that Shewhart charts miss, and non-standard
distributional behavior that invalidates parametric control limits.

```{=html}
<img src="../reference/figures/advanced-spm-overview.svg" alt="Advanced SPM methods: component selection, robust limits, iterative Phase I, run rules, CUSUM, MEWMA, AMEWMA, and ARL analysis" style="width:100%; max-width:720px;">
```

This article covers the advanced SPM toolkit in fdars: automated component
selection, robust control limits, iterative Phase I cleaning, Western Electric
and Nelson run rules, CUSUM and MEWMA charts for detecting small persistent
shifts, adaptive MEWMA, and Average Run Length analysis for comparing methods.

## Simulating Process Data

We simulate semiconductor wafer etch rate profiles: each curve is a spectral
measurement across the wafer surface. Phase I uses 150 in-control runs.
Phase II has 60 new runs with a gradual drift beginning at observation 30.

```{r simulate}
set.seed(42)
n_train <- 150
n_test <- 60
m <- 50
argvals <- seq(0, 1, length.out = m)

# In-control: smooth curves with random variation
basis <- outer(argvals, 1:5, function(t, k) sin(k * pi * t))
X_train <- matrix(rnorm(n_train * 5), n_train, 5) %*% t(basis) +
           matrix(rnorm(n_train * m, sd = 0.3), n_train, m)
fd_train <- fdata(X_train, argvals = argvals)

# Phase II: gradual drift starting at observation 30
X_test <- matrix(rnorm(n_test * 5), n_test, 5) %*% t(basis) +
          matrix(rnorm(n_test * m, sd = 0.3), n_test, m)
drift <- pmax(0, (1:n_test - 30) / 30)
for (i in 1:n_test) {
  X_test[i, ] <- X_test[i, ] + drift[i] * sin(pi * argvals)
}
fd_test <- fdata(X_test, argvals = argvals)
```

```{r plot-data, fig.height = 4}
plot(fd_train, main = "Phase I: In-Control Etch Rate Profiles",
     xlab = "Wafer Position", ylab = "Etch Rate")
```

We also build a standard Phase I chart as a baseline for the advanced methods:

```{r baseline-chart}
chart <- spm.phase1(fd_train, ncomp = 5, alpha = 0.05, seed = 42)
print(chart)
```

## Component Selection

Choosing the number of principal components is critical: too few misses
important variation (inflating SPE), too many inflates $T^2$ noise.
`spm.ncomp.select()` provides four automated methods.

```{r ncomp-select}
eig <- chart$eigenvalues

ncomp_var90  <- spm.ncomp.select(eig, method = "variance90", threshold = 0.9)
ncomp_kaiser <- spm.ncomp.select(eig, method = "kaiser")
ncomp_elbow  <- spm.ncomp.select(eig, method = "elbow")
ncomp_fixed  <- spm.ncomp.select(eig, method = "fixed", threshold = 3)

cat("variance90:", ncomp_var90, "\n")
cat("kaiser:    ", ncomp_kaiser, "\n")
cat("elbow:     ", ncomp_elbow, "\n")
cat("fixed(3):  ", ncomp_fixed, "\n")
```

```{r ncomp-plot, fig.height = 4}
cum_var <- cumsum(eig) / sum(eig)
df_eig <- data.frame(
  pc = seq_along(eig),
  eigenvalue = eig,
  cum_variance = cum_var
)

ggplot(df_eig, aes(x = .data$pc, y = .data$eigenvalue)) +
  geom_col(fill = "#4A90D9", width = 0.6) +
  geom_line(aes(y = max(.data$eigenvalue) * .data$cum_variance),
            color = "#D55E00", linewidth = 0.8) +
  geom_point(aes(y = max(.data$eigenvalue) * .data$cum_variance),
             color = "#D55E00", size = 2) +
  geom_hline(yintercept = mean(eig), linetype = "dashed",
             color = "grey50", linewidth = 0.5) +
  annotate("text", x = length(eig), y = mean(eig),
           label = "Kaiser threshold", hjust = 1, vjust = -0.5,
           size = 3, color = "grey40") +
  geom_vline(xintercept = ncomp_var90 + 0.5, linetype = "dotted",
             color = "#009E73", linewidth = 0.6) +
  annotate("text", x = ncomp_var90 + 0.5, y = max(eig) * 0.95,
           label = paste0("90% var (k=", ncomp_var90, ")"),
           hjust = -0.1, size = 3, color = "#009E73") +
  geom_vline(xintercept = ncomp_elbow + 0.5, linetype = "dotted",
             color = "#CC79A7", linewidth = 0.6) +
  annotate("text", x = ncomp_elbow + 0.5, y = max(eig) * 0.85,
           label = paste0("Elbow (k=", ncomp_elbow, ")"),
           hjust = -0.1, size = 3, color = "#CC79A7") +
  scale_y_continuous(
    name = "Eigenvalue",
    sec.axis = sec_axis(~ . / max(df_eig$eigenvalue),
                        name = "Cumulative Variance",
                        labels = scales::percent)
  ) +
  labs(title = "Component Selection: Scree Plot with Method Cutoffs",
       x = "Principal Component")
```

The variance-based and elbow methods often agree. Kaiser can be conservative
when eigenvalue decay is gradual. For SPM, retaining slightly more components
is generally safer than too few -- a missed mode of variation will produce
SPE alarms that could have been $T^2$ alarms (and therefore more interpretable).

## Robust Control Limits

The parametric control limit for $T^2$ assumes an $F$-distribution, which may
not hold when the training sample is small or the score distribution is
non-Gaussian. `spm.limit.robust()` provides empirical, bootstrap, and KDE
alternatives.

```{r robust-limits}
t2_vals <- chart$t2.phase1

lim_param     <- spm.limit.robust(t2_vals, ncomp = chart$ncomp,
                                  type = "t2", method = "parametric")
lim_empirical <- spm.limit.robust(t2_vals, ncomp = chart$ncomp,
                                  type = "t2", method = "empirical")
lim_bootstrap <- spm.limit.robust(t2_vals, ncomp = chart$ncomp,
                                  type = "t2", method = "bootstrap",
                                  n.bootstrap = 2000, seed = 42)
lim_kde       <- spm.limit.robust(t2_vals, ncomp = chart$ncomp,
                                  type = "t2", method = "kde")

cat("Parametric UCL:", format(lim_param$ucl, digits = 4), "\n")
cat("Empirical UCL: ", format(lim_empirical$ucl, digits = 4), "\n")
cat("Bootstrap UCL: ", format(lim_bootstrap$ucl, digits = 4), "\n")
cat("KDE UCL:       ", format(lim_kde$ucl, digits = 4), "\n")
```

```{r robust-plot, fig.height = 4}
df_t2 <- data.frame(obs = seq_along(t2_vals), t2 = t2_vals)

ucl_lines <- data.frame(
  method = c("Parametric", "Empirical", "Bootstrap", "KDE"),
  ucl = c(lim_param$ucl, lim_empirical$ucl, lim_bootstrap$ucl, lim_kde$ucl)
)

ggplot(df_t2, aes(x = .data$obs, y = .data$t2)) +
  geom_segment(aes(xend = .data$obs, y = 0, yend = .data$t2),
               color = "#4A90D9", linewidth = 0.5) +
  geom_hline(data = ucl_lines,
             aes(yintercept = .data$ucl, color = .data$method, linetype = .data$method),
             linewidth = 0.7) +
  scale_color_manual(values = c("Parametric" = "#D55E00", "Empirical" = "#009E73",
                                "Bootstrap" = "#CC79A7", "KDE" = "#E69F00")) +
  scale_linetype_manual(values = c("Parametric" = "dashed", "Empirical" = "dotted",
                                   "Bootstrap" = "solid", "KDE" = "longdash")) +
  labs(title = expression("Phase I T"^2 * " with Parametric vs Robust UCLs"),
       x = "Calibration Observation", y = expression("T"^2),
       color = "Method", linetype = "Method") +
  theme(legend.position = "bottom")
```

When the limits diverge substantially, the parametric assumption is suspect.
Bootstrap limits are generally the most reliable: they make no distributional
assumptions and provide stable estimates with 1000+ resamples.

## Iterative Phase I

Real training data often contains a few out-of-control observations that
contaminate the FPCA and inflate control limits. `spm.phase1.iterative()`
repeatedly fits the chart, removes flagged observations, and re-fits until
convergence or a maximum removal fraction is reached.

We create a contaminated training set by injecting 10% outliers:

```{r iterative-setup}
set.seed(42)
n_outlier <- 15
outlier_idx <- sample(n_train, n_outlier)

X_contaminated <- X_train
for (i in outlier_idx) {
  X_contaminated[i, ] <- X_contaminated[i, ] + 3 * sin(2 * pi * argvals)
}
fd_contaminated <- fdata(X_contaminated, argvals = argvals)
```

```{r iterative-fit}
chart_naive <- spm.phase1(fd_contaminated, ncomp = 3, alpha = 0.05, seed = 42)

chart_iter <- spm.phase1.iterative(fd_contaminated, ncomp = 3, alpha = 0.05,
                                    seed = 42, max.iterations = 10,
                                    max.removal.fraction = 0.3)

cat("Naive T2 UCL:", format(chart_naive$t2.ucl, digits = 4), "\n")
cat("Iterative T2 UCL:", format(chart_iter$t2.ucl, digits = 4), "\n")
cat("Iterations:", chart_iter$n.iterations, "\n")
cat("Removed:", length(chart_iter$removed.indices), "observations\n")
cat("Remaining:", chart_iter$n.remaining, "observations\n")
```

```{r iterative-plot, fig.height = 6}
n_cal_naive <- length(chart_naive$t2.phase1)
n_cal_iter <- length(chart_iter$t2.phase1)

df_before <- data.frame(
  obs = seq_len(n_cal_naive),
  t2 = chart_naive$t2.phase1,
  panel = "Before: Naive Phase I"
)
df_before$alarm <- df_before$t2 > chart_naive$t2.ucl

df_after <- data.frame(
  obs = seq_len(n_cal_iter),
  t2 = chart_iter$t2.phase1,
  panel = "After: Iterative Phase I"
)
df_after$alarm <- df_after$t2 > chart_iter$t2.ucl

df_comb <- rbind(df_before, df_after)
df_comb$panel <- factor(df_comb$panel,
                        levels = c("Before: Naive Phase I",
                                   "After: Iterative Phase I"))

ucl_df <- data.frame(
  panel = factor(c("Before: Naive Phase I", "After: Iterative Phase I"),
                 levels = c("Before: Naive Phase I", "After: Iterative Phase I")),
  ucl = c(chart_naive$t2.ucl, chart_iter$t2.ucl),
  label = paste("UCL =", format(c(chart_naive$t2.ucl, chart_iter$t2.ucl),
                                digits = 4))
)

ggplot(df_comb, aes(x = .data$obs, y = .data$t2)) +
  geom_segment(aes(xend = .data$obs, y = 0, yend = .data$t2,
                   color = .data$alarm), linewidth = 0.6) +
  geom_hline(data = ucl_df, aes(yintercept = .data$ucl),
             linetype = "dashed", color = "#D55E00", linewidth = 0.7) +
  geom_text(data = ucl_df,
            aes(x = -Inf, y = .data$ucl, label = .data$label),
            hjust = -0.05, vjust = -0.5, size = 3, color = "#D55E00") +
  scale_color_manual(values = c("FALSE" = "#4A90D9", "TRUE" = "#D55E00"),
                     guide = "none") +
  facet_wrap(~ .data$panel, ncol = 1, scales = "free") +
  labs(title = expression("Iterative Phase I: Outlier Removal Effect on T"^2),
       x = "Observation", y = expression("T"^2)) +
  theme(strip.text = element_text(face = "bold"))
```

The iterative procedure tightens the control limits by removing contaminating
observations. The resulting chart is more sensitive to real out-of-control
behavior in Phase II because the baseline is no longer inflated by outliers.

## Western Electric and Nelson Rules

Beyond simply comparing a statistic to a fixed UCL, run rules detect
non-random patterns in the monitoring statistics: runs above/below the
center line, trending sequences, and zone violations. These patterns often
indicate developing process issues before a single point crosses the UCL.

```{r rules-setup}
# Monitor Phase II data with the clean chart
chart_clean <- spm.phase1(fd_train, ncomp = 3, alpha = 0.05, seed = 42)
monitor <- spm.monitor(chart_clean, fd_test)

# Apply Western Electric rules to T2 statistics
t2_center <- mean(chart_clean$t2.phase1)
t2_sigma <- sd(chart_clean$t2.phase1)

we_rules <- spm.rules(monitor$t2, center = t2_center, sigma = t2_sigma,
                       rule.set = "western.electric")
print(we_rules)

# Also try Nelson rules (more comprehensive, 8 rules)
nelson_rules <- spm.rules(monitor$t2, center = t2_center, sigma = t2_sigma,
                           rule.set = "nelson")
print(nelson_rules)
```

```{r rules-plot, fig.height = 4.5}
# Collect all violation indices from Western Electric rules
violation_idx <- unique(unlist(we_rules$indices))

df_rules <- data.frame(
  obs = seq_along(monitor$t2),
  t2 = monitor$t2,
  violation = seq_along(monitor$t2) %in% violation_idx,
  ucl_alarm = monitor$t2.alarm
)

ggplot(df_rules, aes(x = .data$obs, y = .data$t2)) +
  geom_line(color = "#4A90D9", linewidth = 0.5) +
  geom_point(aes(color = interaction(.data$violation, .data$ucl_alarm)),
             size = 1.8) +
  geom_hline(yintercept = chart_clean$t2.ucl, linetype = "dashed",
             color = "#D55E00", linewidth = 0.6) +
  geom_hline(yintercept = t2_center, linetype = "solid",
             color = "grey50", linewidth = 0.4) +
  geom_hline(yintercept = t2_center + c(1, 2) * t2_sigma,
             linetype = "dotted", color = "grey70", linewidth = 0.3) +
  annotate("text", x = n_test, y = t2_center + c(1, 2, 3) * t2_sigma,
           label = c(expression(paste("1", sigma)),
                     expression(paste("2", sigma)),
                     "UCL"),
           hjust = 1.1, vjust = -0.3, size = 2.5, color = "grey50") +
  geom_vline(xintercept = 30.5, linetype = "dashed", color = "#009E73",
             linewidth = 0.4, alpha = 0.7) +
  annotate("text", x = 30.5, y = max(monitor$t2) * 0.95,
           label = "Drift onset", hjust = -0.1, size = 3, color = "#009E73") +
  scale_color_manual(
    values = c("FALSE.FALSE" = "#4A90D9",   # normal
               "TRUE.FALSE"  = "#E69F00",   # rule violation only
               "FALSE.TRUE"  = "#D55E00",   # UCL alarm only
               "TRUE.TRUE"   = "#D55E00"),  # both
    labels = c("In-control", "Rule violation", "UCL alarm", "Both"),
    name = "Status"
  ) +
  labs(title = expression("Western Electric Rules Applied to T"^2 * " Statistics"),
       x = "Monitoring Observation", y = expression("T"^2)) +
  theme(legend.position = "bottom")
```

The Western Electric rules flag patterns that the Shewhart UCL alone would
miss. In the gradual drift scenario, run rules often detect the shift before
any single observation crosses the UCL, because the rules are sensitive to
sustained deviations from the center line.

## CUSUM Charts

The Cumulative Sum (CUSUM) chart accumulates evidence of a mean shift over
time. Unlike the Shewhart chart that only looks at each observation
individually, CUSUM maintains a running total that grows when the process
deviates from its in-control state. This makes it particularly powerful for
detecting small, persistent shifts.

The two key parameters are the allowance $k$ (the shift size the CUSUM is
tuned to detect, in standard deviation units) and the decision interval $h$
(an alarm fires when the CUSUM exceeds $h$).

```{r cusum}
cusum_no_restart <- spm.cusum(chart_clean, fd_test,
                               k = 0.5, h = 5.0, restart = FALSE)
cusum_restart    <- spm.cusum(chart_clean, fd_test,
                               k = 0.5, h = 5.0, restart = TRUE)

cat("Without restart — alarms:", sum(cusum_no_restart$alarm), "\n")
cat("With restart    — alarms:", sum(cusum_restart$alarm), "\n")
```

```{r cusum-plot, fig.height = 5}
df_cusum <- data.frame(
  obs = rep(seq_len(n_test), 2),
  cusum = c(cusum_no_restart$cusum.statistic, cusum_restart$cusum.statistic),
  variant = factor(rep(c("No Restart", "With Restart"), each = n_test))
)

ggplot(df_cusum, aes(x = .data$obs, y = .data$cusum, color = .data$variant)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 5.0, linetype = "dashed", color = "#D55E00",
             linewidth = 0.6) +
  annotate("text", x = 1, y = 5.0, label = "h = 5.0",
           hjust = -0.1, vjust = -0.5, size = 3, color = "#D55E00") +
  geom_vline(xintercept = 30.5, linetype = "dashed", color = "#009E73",
             linewidth = 0.4, alpha = 0.7) +
  annotate("text", x = 30.5, y = max(df_cusum$cusum) * 0.9,
           label = "Drift onset", hjust = -0.1, size = 3, color = "#009E73") +
  scale_color_manual(values = c("No Restart" = "#4A90D9",
                                "With Restart" = "#CC79A7")) +
  labs(title = "CUSUM Monitoring: Restart vs No-Restart",
       x = "Monitoring Observation", y = "CUSUM Statistic",
       color = "Variant") +
  theme(legend.position = "bottom")
```

Without restart, the CUSUM climbs monotonically after the shift and stays
elevated. With restart, it resets to zero after each alarm, which is useful
when you want to detect the *next* shift after corrective action. In both
cases, the CUSUM crosses the threshold well before the gradual drift becomes
large enough for a single Shewhart observation to trigger an alarm.

## MEWMA Monitoring

The Multivariate Exponentially Weighted Moving Average (MEWMA) chart smooths
the FPC score *vectors* over time, then computes a Hotelling-type statistic
on the smoothed scores. This preserves the multivariate correlation structure
while gaining the sensitivity of exponential smoothing to small persistent
shifts.

```{r mewma}
mewma_result <- spm.mewma(chart_clean, fd_test, lambda = 0.2)
print(mewma_result)
```

```{r mewma-plot, fig.height = 5}
df_compare <- data.frame(
  obs = rep(seq_len(n_test), 2),
  value = c(mewma_result$mewma.statistic, monitor$t2),
  method = factor(rep(c("MEWMA (lambda=0.2)", "Raw T\u00b2"), each = n_test))
)

ggplot(df_compare, aes(x = .data$obs, y = .data$value, color = .data$method)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = mewma_result$ucl, linetype = "dashed",
             color = "#4A90D9", linewidth = 0.5) +
  geom_hline(yintercept = chart_clean$t2.ucl, linetype = "dashed",
             color = "#D55E00", linewidth = 0.5) +
  geom_vline(xintercept = 30.5, linetype = "dashed", color = "#009E73",
             linewidth = 0.4, alpha = 0.7) +
  annotate("text", x = 30.5, y = max(df_compare$value) * 0.95,
           label = "Drift onset", hjust = -0.1, size = 3, color = "#009E73") +
  scale_color_manual(values = c("MEWMA (lambda=0.2)" = "#4A90D9",
                                "Raw T\u00b2" = "#D55E00")) +
  labs(title = expression("MEWMA vs Raw T"^2 * ": Detecting Gradual Drift"),
       x = "Monitoring Observation", y = "Statistic",
       color = "Method") +
  theme(legend.position = "bottom")
```

The MEWMA statistic shows a clear upward trend after the drift onset at
observation 30, while the raw $T^2$ values remain noisy. The smoothing
filters out observation-to-observation variability and accumulates evidence
of a sustained shift. Smaller $\lambda$ values give more smoothing and
earlier detection of small shifts, at the cost of slower response to sudden
large shifts.

## AMEWMA

The Adaptive MEWMA (AMEWMA) extends MEWMA by allowing the smoothing parameter
$\lambda$ to change over time. When the process is in control, $\lambda$ stays
small (strong smoothing for sensitivity to small shifts). When a large
deviation is detected, $\lambda$ increases toward 1 (weak smoothing for fast
response to large shifts). This gives good performance across a wide range
of shift magnitudes.

```{r amewma}
amewma_result <- spm.amewma(chart_clean, fd_test,
                             lambda.min = 0.05, lambda.max = 0.95,
                             lambda.init = 0.2, eta = 0.1)
print(amewma_result)
```

```{r amewma-plot, fig.height = 6}
# Panel 1: AMEWMA statistic
df_amewma_t2 <- data.frame(
  obs = seq_len(n_test),
  value = amewma_result$t2.statistic,
  alarm = amewma_result$alarm
)

p1 <- ggplot(df_amewma_t2, aes(x = .data$obs, y = .data$value)) +
  geom_line(color = "#4A90D9", linewidth = 0.6) +
  geom_point(data = df_amewma_t2[df_amewma_t2$alarm, , drop = FALSE],
             aes(x = .data$obs, y = .data$value),
             color = "#D55E00", size = 1.5) +
  geom_hline(yintercept = amewma_result$ucl, linetype = "dashed",
             color = "#D55E00", linewidth = 0.6) +
  annotate("text", x = 1, y = amewma_result$ucl,
           label = paste("UCL =", format(amewma_result$ucl, digits = 4)),
           hjust = -0.05, vjust = -0.5, size = 3, color = "#D55E00") +
  geom_vline(xintercept = 30.5, linetype = "dashed", color = "#009E73",
             linewidth = 0.4, alpha = 0.7) +
  labs(title = "AMEWMA Monitoring",
       x = NULL, y = expression("T"^2 * " Statistic"))

# Panel 2: adaptive lambda
df_lambda <- data.frame(
  obs = seq_len(n_test),
  lambda = amewma_result$lambda.t
)

p2 <- ggplot(df_lambda, aes(x = .data$obs, y = .data$lambda)) +
  geom_line(color = "#4A90D9", linewidth = 0.6) +
  geom_hline(yintercept = 0.2, linetype = "dotted", color = "grey50") +
  annotate("text", x = n_test, y = 0.2,
           label = expression(lambda[init] == 0.2),
           hjust = 1, vjust = -0.5, size = 3, color = "grey40") +
  geom_vline(xintercept = 30.5, linetype = "dashed", color = "#009E73",
             linewidth = 0.4, alpha = 0.7) +
  ylim(0.05, 0.95) +
  labs(x = "Monitoring Observation", y = expression("Adaptive " * lambda))

p1 / p2 + patchwork::plot_layout(heights = c(2, 1))
```

The adaptive $\lambda$ trace reveals the AMEWMA's self-tuning behavior. During
the in-control period (observations 1--30), $\lambda$ remains near its initial
value. As the drift develops, $\lambda$ adjusts to balance sensitivity and
responsiveness. This adaptivity means you do not need to tune $\lambda$ for a
specific shift magnitude -- the chart handles both small and large shifts.

## ARL Analysis

The Average Run Length (ARL) is the expected number of observations before an
alarm. The in-control ARL ($\text{ARL}_0$) measures the false alarm rate:
larger is better. The out-of-control ARL ($\text{ARL}_1$) measures detection
speed: smaller is better.

```{r arl, message = FALSE}
# In-control ARL for Shewhart T2
arl0_t2 <- spm.arl(chart_clean, type = "t2", n.sim = 5000, seed = 42)

# Out-of-control ARL with a shift in the first component
shift_vec <- rep(0, chart_clean$ncomp)
shift_vec[1] <- 1.0
arl1_t2 <- spm.arl(chart_clean, type = "t2", shift.magnitude = shift_vec,
                    n.sim = 5000, seed = 42)

# In-control ARL for EWMA
arl0_ewma <- spm.arl.ewma(chart_clean, lambda = 0.2, n.sim = 5000, seed = 42)

cat("Shewhart T2 — ARL0:", format(arl0_t2$arl, digits = 4),
    " ARL1:", format(arl1_t2$arl, digits = 4), "\n")
cat("EWMA T2     — ARL0:", format(arl0_ewma$arl, digits = 4), "\n")
```

```{r arl-table}
arl_table <- data.frame(
  Method = c("Shewhart T\u00b2", "EWMA T\u00b2 (lambda=0.2)"),
  ARL0 = c(format(arl0_t2$arl, digits = 1),
           format(arl0_ewma$arl, digits = 1)),
  ARL0_SD = c(format(arl0_t2$std.dev, digits = 1),
              format(arl0_ewma$std.dev, digits = 1)),
  Median_RL0 = c(arl0_t2$median.rl, arl0_ewma$median.rl)
)

knitr::kable(arl_table,
             col.names = c("Method", "ARL0", "Std. Dev.", "Median RL"),
             caption = "In-control Average Run Length comparison",
             align = "lrrr")
```

When comparing methods at a fixed $\text{ARL}_0$, the chart with the smaller
$\text{ARL}_1$ is preferred. EWMA charts typically achieve much lower
$\text{ARL}_1$ for small to moderate shifts while maintaining comparable
$\text{ARL}_0$, which is why they are the default choice for detecting
gradual process degradation.

## Method Comparison

| Method | Detects | Strengths | Limitations |
|--------|---------|-----------|-------------|
| **Shewhart $T^2$/SPE** | Large sudden shifts | Simple, no tuning | Misses small persistent shifts |
| **Western Electric / Nelson rules** | Runs, trends, zone violations | No extra parameters | Requires centerline and sigma |
| **CUSUM** | Small persistent shifts | Optimal for known shift size | Must specify $k$ and $h$ |
| **MEWMA** | Small persistent shifts | Multivariate smoothing | Must choose $\lambda$ |
| **AMEWMA** | Shifts of any size | Self-tuning $\lambda$ | More complex, more parameters |
| **Robust limits** | Same as Shewhart | No distributional assumptions | Requires larger Phase I sample |
| **Iterative Phase I** | Cleans training data | Automatic outlier removal | May remove valid extreme runs |

**Rules of thumb:**

- Start with `spm.phase1()` + `spm.monitor()` as a baseline.
- If Phase I data quality is uncertain, use `spm.phase1.iterative()`.
- For gradual degradation, add `spm.mewma()` or `spm.cusum()`.
- When the shift magnitude is unknown, prefer `spm.amewma()`.
- Always validate with `spm.arl()` to compare false alarm rates.

## See Also

- `vignette("articles/statistical-process-monitoring")` -- Basic SPM workflow
- `vignette("articles/fpca")` -- Functional PCA fundamentals
- `vignette("articles/streaming-depth")` -- Online depth-based monitoring

## References

- Montgomery, D.C. (2013). *Statistical Quality Control*. 7th edition. Wiley.
- Lowry, C.A., Woodall, W.H., Champ, C.W., and Rigdon, S.E. (1992).
  A multivariate exponentially weighted moving average control chart.
  *Technometrics*, 34(1), 46--53.
- Mahmoud, M.A. and Maravelakis, P.E. (2011). The performance of multivariate
  CUSUM control charts with estimated parameters. *Journal of Statistical
  Computation and Simulation*, 81(4), 497--514.
- Sparks, R.S. (2000). CUSUM charts for signalling varying location shifts.
  *Journal of Quality Technology*, 32(2), 157--171.
