---
title: "Comparing Alignment Methods"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

fdars provides four approaches to curve alignment, each with different trade-offs
in terms of automation, smoothness, interpretability, and speed. This article
compares them side-by-side on the same data, and provides guidance on when to
use each method.

The four methods are:

1. **Elastic alignment** (`elastic.align`, `karcher.mean`) -- global optimization
   via dynamic programming in SRSF space
2. **Landmark registration** (`landmark.register`) -- feature-based piecewise-linear
   warping
3. **Constrained elastic alignment** (`elastic.align.constrained`) -- elastic
   alignment with landmark pass-through constraints
4. **TSRVF** (`tsrvf.transform`) -- not an alignment method per se, but a
   linearization of elastic alignment for downstream analysis

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## Test Data

We use two datasets that highlight different alignment challenges.

### Dataset 1: Smooth Phase Shifts

Curves with smooth, global timing differences -- the classic alignment scenario:

```{r data-smooth}
set.seed(42)
argvals <- seq(0, 1, length.out = 200)
n <- 15

data_smooth <- matrix(0, n, 200)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  amp <- rnorm(1, 1, 0.15)
  data_smooth[i, ] <- amp * sin(2 * pi * (argvals - shift)) +
                       0.4 * amp * sin(4 * pi * (argvals - shift * 0.7))
}
fd_smooth <- fdata(data_smooth, argvals = argvals)
```

### Dataset 2: Distinct Feature Shifts

Curves with clearly identifiable peaks that shift independently -- where domain
knowledge matters:

```{r data-features}
data_feat <- matrix(0, n, 200)
for (i in 1:n) {
  pk1 <- runif(1, 0.15, 0.35)
  pk2 <- runif(1, 0.55, 0.75)
  w1 <- rnorm(1, 1, 0.2)
  w2 <- rnorm(1, 0.8, 0.15)
  data_feat[i, ] <- w1 * exp(-200 * (argvals - pk1)^2) +
                     w2 * exp(-200 * (argvals - pk2)^2) +
                     0.05 * rnorm(200)
}
fd_feat <- fdata(data_feat, argvals = argvals)
```

```{r data-plot}
p1 <- plot(fd_smooth) + labs(title = "Dataset 1: Smooth Phase Shifts")
p2 <- plot(fd_feat) + labs(title = "Dataset 2: Distinct Features")
p1 + p2
```

## Method 1: Elastic Alignment

Elastic alignment finds the globally optimal warping function by minimizing the
$L^2$ distance between SRSFs via dynamic programming. It requires no feature
detection and produces smooth, diffeomorphic warping functions.

**Mathematical basis**: minimize
$\|\, q_1 - (q_2 \circ \gamma) \sqrt{\dot\gamma}\,\|_{L^2}^2$ over the warping
group $\Gamma$, where $q_i$ is the SRSF of $f_i$.

```{r elastic}
ea_smooth <- elastic.align(fd_smooth)
ea_feat <- elastic.align(fd_feat)
```

```{r elastic-plot}
p1 <- plot(ea_smooth, type = "aligned") + labs(title = "Elastic (smooth data)")
p2 <- plot(ea_feat, type = "aligned") + labs(title = "Elastic (feature data)")
p1 + p2
```

## Method 2: Landmark Registration

Landmark registration detects features (peaks, valleys, etc.) and warps curves
so that corresponding features align to common target positions. The warping is
piecewise-linear between anchor points.

**Mathematical basis**: For landmarks $\tau_{i,j}$ and targets $\tau_j^*$,
construct $\gamma_i$ as a piecewise-linear function satisfying
$\gamma_i(\tau_j^*) = \tau_{i,j}$.

```{r landmark}
lr_smooth <- landmark.register(fd_smooth, kind = "peak", min.prominence = 0.3,
                                expected.count = 1)
lr_feat <- landmark.register(fd_feat, kind = "peak", min.prominence = 0.2,
                              expected.count = 2)
```

```{r landmark-plot}
p1 <- plot(lr_smooth, type = "registered") +
  labs(title = "Landmark (smooth data)")
p2 <- plot(lr_feat, type = "registered") +
  labs(title = "Landmark (feature data)")
p1 + p2
```

## Method 3: Constrained Elastic Alignment

Constrained elastic alignment combines the smooth, globally optimal warping of
elastic alignment with landmark pass-through constraints. The warping is
computed by segmented dynamic programming between landmark anchor points.

**Mathematical basis**: minimize $\|q_1 - (q_2 \circ \gamma) \sqrt{\dot\gamma}\|_{L^2}^2$
subject to $\gamma(\tau_j^*) = \tau_{i,j}$ for each landmark pair.

```{r constrained}
ec_smooth <- elastic.align.constrained(fd_smooth, kind = "peak",
                                        min.prominence = 0.3,
                                        expected.count = 1)
ec_feat <- elastic.align.constrained(fd_feat, kind = "peak",
                                      min.prominence = 0.2,
                                      expected.count = 2)
```

```{r constrained-plot}
p1 <- plot(ec_smooth, type = "aligned") +
  labs(title = "Constrained (smooth data)")
p2 <- plot(ec_feat, type = "aligned") +
  labs(title = "Constrained (feature data)")
p1 + p2
```

## Warping Function Comparison

The warping functions reveal the fundamental difference between methods. Elastic
produces smooth warps, landmark produces piecewise-linear warps, and constrained
is smooth but passes through landmark anchor points:

```{r warps-compare}
p1 <- plot(ea_feat, type = "warps") + labs(title = "Elastic Warps")
p2 <- plot(lr_feat, type = "warps") + labs(title = "Landmark Warps")
p3 <- plot(ec_feat, type = "warps") + labs(title = "Constrained Warps")
p1 + p2 + p3
```

## Quantitative Comparison

### Variance Reduction

Variance reduction (VR) measures how much pointwise variance is removed by
alignment. Higher VR means more phase variability was captured:

$$\text{VR} = 1 - \frac{\text{mean pointwise variance (aligned)}}{\text{mean pointwise variance (original)}}$$

```{r vr-comparison}
vr <- function(original, aligned) {
  var_orig <- mean(apply(original$data, 2, var))
  var_aligned <- mean(apply(aligned$data, 2, var))
  1 - var_aligned / var_orig
}

results <- data.frame(
  Dataset = rep(c("Smooth shifts", "Distinct features"), each = 3),
  Method = rep(c("Elastic", "Landmark", "Constrained"), 2),
  VR = c(
    vr(fd_smooth, ea_smooth$aligned),
    vr(fd_smooth, lr_smooth$registered),
    vr(fd_smooth, ec_smooth$aligned),
    vr(fd_feat, ea_feat$aligned),
    vr(fd_feat, lr_feat$registered),
    vr(fd_feat, ec_feat$aligned)
  )
)

results$VR <- round(results$VR, 3)
print(results)
```

```{r vr-plot}
ggplot(results, aes(x = Method, y = VR, fill = Dataset)) +
  geom_col(position = "dodge") +
  labs(title = "Variance Reduction by Method",
       y = "Variance Reduction (higher = better)") +
  scale_fill_manual(values = c("Smooth shifts" = "steelblue",
                                "Distinct features" = "coral")) +
  ylim(0, 1)
```

### Alignment Quality Diagnostics

For the Karcher mean (elastic), we can compute detailed diagnostics:

```{r quality}
km_smooth <- karcher.mean(fd_smooth, max.iter = 15)
aq_smooth <- alignment.quality(fd_smooth, km_smooth)
print(aq_smooth)
```

### Amplitude-Phase Decomposition

Elastic alignment uniquely provides a principled decomposition of total
variability into amplitude and phase components:

```{r decomposition}
plot(aq_smooth, type = "variance")
```

## Decision Guide

### Summary Table

| Method | Warping | Automation | Speed | Smoothness | Feature control |
|--------|---------|------------|-------|------------|----------------|
| **Elastic** | Smooth diffeomorphism | Fully automatic | $O(nm^2)$ | High | None |
| **Landmark** | Piecewise-linear | Needs feature type | $O(nm + nk)$ | Low (corners at landmarks) | Full |
| **Constrained** | Smooth with anchors | Needs feature type | $O(nm^2/k)$ | High (between landmarks) | Partial |
| **TSRVF** | (uses elastic) | Fully automatic | $O(nm^2)$ + PCA | High | None |

### When to Use Each Method

**Use elastic alignment when:**

- You have no prior knowledge about which features should correspond
- Curves have smooth, continuous timing differences
- You want a principled metric space structure (Fisher-Rao distance)
- You need amplitude-phase decomposition or TSRVF for downstream analysis

**Use landmark registration when:**

- Features are well-defined and domain-meaningful (ECG waves, spectral peaks)
- You need guaranteed feature correspondence
- Speed is critical (linear in grid size per landmark)
- Curves have complex warping that confuses global DP (e.g., features moving in
  opposite directions)

**Use constrained elastic alignment when:**

- You want both smooth warps and feature correspondence
- Some features must align exactly, but the warping between them should be
  optimized
- You are comparing elastic and landmark results and want a middle ground

**Use TSRVF when:**

- You need to perform PCA, regression, or classification on aligned data
- You want a Euclidean representation of curve shapes
- You have already computed a Karcher mean and want to extract tangent vectors

### Pitfalls

| Method | Common pitfall | Mitigation |
|--------|---------------|------------|
| **Elastic** | Over-warping (pinching) on noisy data | Increase `lambda` penalty |
| **Landmark** | Mismatched landmark correspondence | Use `expected.count`, increase `min.prominence` |
| **Constrained** | Too few landmarks = barely constrained | Detect multiple landmark types |
| **TSRVF** | Linearization error for curves far from mean | Check reconstruction quality |

## Workflow Example

A typical analysis workflow combining multiple methods:

```{r workflow}
# 1. Explore data
plot(fd_feat) + labs(title = "Step 1: Explore")
```

```{r workflow-landmarks}
# 2. Detect features to understand structure
lms <- detect.landmarks(fd_feat, kind = "peak", min.prominence = 0.2)
cat("Landmarks per curve:",
    paste(sapply(lms, nrow), collapse = ", "), "\n")
```

```{r workflow-align}
# 3. Align (choose method based on data)
lr <- landmark.register(fd_feat, kind = "peak", min.prominence = 0.2,
                         expected.count = 2)
```

```{r workflow-tsrvf}
# 4. For downstream analysis, compute Karcher mean + TSRVF
km <- karcher.mean(lr$registered, max.iter = 10)
tv <- tsrvf.from.alignment(km)

# 5. Standard PCA on tangent vectors
pca <- prcomp(tv$tangent_vectors$data)
var_exp <- cumsum(pca$sdev^2 / sum(pca$sdev^2))
cat("Variance explained (first 3 PCs):",
    round(var_exp[1:3] * 100, 1), "%\n")
```

## See Also

- `vignette("elastic-alignment", package = "fdars")` -- elastic alignment details
- `vignette("landmark-registration", package = "fdars")` -- landmark registration
  details
- `vignette("tsrvf", package = "fdars")` -- TSRVF for linearized analysis
