---
title: "Andrews Transformation: From Tables to Curves"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Many potential fdars users have multivariate (tabular) data rather than
functional data. The **Andrews transformation** (Andrews, 1972) provides a
mathematically rigorous bridge: it maps each $p$-dimensional observation to a
curve on $[-\pi, \pi]$ using a Fourier expansion where the data values serve as
coefficients.

The key property is that the $L^2$ distance between Andrews curves equals
$\sqrt{\pi}$ times the Euclidean distance between the original observations ---
so all distance-based FDA methods produce results that are directly
interpretable in the original multivariate space.

This article demonstrates the full fdars pipeline --- depth, outlier detection,
clustering, and FPCA --- applied to Andrews-transformed iris data.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

## Mathematical Background

Given a $p$-dimensional observation $\mathbf{x} = (x_1, x_2, \ldots, x_p)$,
the Andrews function is defined as:

$$f_{\mathbf{x}}(t) = \frac{x_1}{\sqrt{2}} + x_2 \sin(t) + x_3 \cos(t) + x_4 \sin(2t) + x_5 \cos(2t) + \cdots$$

for $t \in [-\pi, \pi]$. The pattern is: the first coefficient is scaled by
$1/\sqrt{2}$, then subsequent coefficients multiply alternating $\sin(kt)$ and
$\cos(kt)$ terms.

**Distance preservation theorem.** For any two observations $\mathbf{x}$ and
$\mathbf{y}$:

$$\|f_{\mathbf{x}} - f_{\mathbf{y}}\|_{L^2} = \sqrt{\pi} \cdot \|\mathbf{x} - \mathbf{y}\|_2$$

This means Euclidean distances are preserved (up to a constant factor) when
moving to the functional domain. Every distance-based FDA method --- depth,
clustering, outlier detection --- produces results equivalent to applying the
same method with Euclidean distance in the original space.

**Variable ordering.** Because the first Fourier terms carry the most visual
weight, ordering variables by importance (e.g., by ANOVA F-statistic or
variance) makes the plots more informative. However, the distance preservation
property holds regardless of ordering, so analytical results are invariant.

## The `andrews_transform()` Function

The package provides `andrews_transform()`, which builds the Fourier basis
matrix and returns an `fdata` object. The basis matrix and variable names are
stored as attributes so that `andrews_loadings()` can later project FPCA
eigenfunctions back to the original variable space.

## Iris Dataset and Visualization

The iris dataset has 150 observations with 4 numeric variables. We standardize
before transforming so that no single variable dominates due to scale:

```{r iris-transform}
X_raw <- as.matrix(iris[, 1:4])
X <- scale(X_raw)
species <- iris$Species

fd_iris <- andrews_transform(X)
cat("Andrews curves:", nrow(fd_iris$data), "observations,",
    ncol(fd_iris$data), "grid points\n")
```

```{r iris-plot, fig.height = 6}
# Plot Andrews curves colored by species
n <- nrow(fd_iris$data)
m <- ncol(fd_iris$data)
t_grid <- fd_iris$argvals

df_curves <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_iris$data)),
  curve = rep(1:n, each = m),
  species = rep(species, each = m)
)

ggplot(df_curves, aes(x = t, y = value, group = curve, color = species)) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  scale_color_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                                "virginica" = "#4DAF4A")) +
  labs(title = "Andrews Curves of Iris Data",
       x = expression(t), y = expression(f[x](t)), color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Setosa separates clearly, while versicolor and virginica overlap --- exactly
matching what we know about these species in the original space.

### Group Means

```{r group-means}
# Compute mean curve per species
species_levels <- levels(species)
df_means <- do.call(rbind, lapply(species_levels, function(sp) {
  idx <- which(species == sp)
  mean_curve <- colMeans(fd_iris$data[idx, ])
  data.frame(t = t_grid, value = mean_curve, species = sp)
}))

ggplot(df_means, aes(x = t, y = value, color = species)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                                "virginica" = "#4DAF4A")) +
  labs(title = "Mean Andrews Curves by Species",
       x = expression(t), y = expression(bar(f)(t)), color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Depth and Centrality

Functional depth measures how "central" each curve is relative to the sample.
Deeper curves are more typical; shallow curves are potential outliers.

```{r depth}
# Fraiman-Muniz depth
depth_fm <- depth.FM(fd_iris)

# Modified Band Depth
depth_mbd <- depth.MBD(fd_iris)
```

```{r depth-boxplot}
df_depth <- data.frame(
  FM = depth_fm,
  MBD = depth_mbd,
  species = species
)

# Boxplot of depth by species
df_depth_long <- data.frame(
  depth = c(depth_fm, depth_mbd),
  method = rep(c("Fraiman-Muniz", "Modified Band Depth"), each = length(species)),
  species = rep(species, 2)
)

ggplot(df_depth_long, aes(x = species, y = depth, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ method, scales = "free_y") +
  scale_fill_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                               "virginica" = "#4DAF4A")) +
  labs(title = "Functional Depth by Species",
       x = NULL, y = "Depth") +
  theme(legend.position = "none")
```

### Functional Median and Trimmed Mean

```{r central-curves}
# Functional median (deepest curve)
fd_median <- median(fd_iris)

# Trimmed mean (mean of deepest 50%)
fd_trimmed <- trimmed(fd_iris, trim = 0.5)

df_central <- data.frame(
  t = rep(t_grid, 2),
  value = c(fd_median$data[1, ], fd_trimmed$data[1, ]),
  type = rep(c("Median", "Trimmed Mean (50%)"), each = m)
)

ggplot(df_central, aes(x = t, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Median" = "darkblue", "Trimmed Mean (50%)" = "darkorange")) +
  labs(title = "Functional Median and Trimmed Mean",
       x = expression(t), y = expression(f(t)), color = NULL) +
  theme(legend.position = "bottom")
```

## Outlier Detection

We apply three complementary outlier detection methods.

### Depth-Ponderation

```{r outliers-pond}
out_pond <- outliers.depth.pond(fd_iris, nb = 1000, seed = 123)
cat("Depth-ponderation outliers:", out_pond$outliers, "\n")
cat("Species:", as.character(species[out_pond$outliers]), "\n")
```

### Outliergram

The outliergram plots Modified Epigraph Index against Modified Band Depth,
flagging curves that fall outside the expected parabolic relationship:

```{r outliergram}
og <- outliergram(fd_iris)
plot(og)
```

### Magnitude-Shape Plot

The magnitude-shape plot separates outliers into those that differ in
magnitude (shifted up/down) from those that differ in shape (different pattern):

```{r magnitudeshape}
ms <- magnitudeshape(fd_iris)
plot(ms)
```

### Cross-Referencing with Original Data

```{r outlier-check}
# Collect all flagged observations
all_outliers <- unique(c(out_pond$outliers, og$outliers, ms$outliers))
cat("All flagged observations:", sort(all_outliers), "\n")

if (length(all_outliers) > 0) {
  cat("\nOriginal iris values for flagged observations:\n")
  print(iris[all_outliers, ])
}
```

## Clustering

### K-Means with k = 3

```{r kmeans}
km <- cluster.kmeans(fd_iris, ncl = 3, nstart = 20, seed = 123)

# Confusion matrix
conf_matrix <- table(Predicted = km$cluster, True = species)
print(conf_matrix)

# Accuracy with optimal label matching
max_matches <- 0
for (perm in list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))) {
  matched <- sum(km$cluster == perm[as.numeric(species)])
  max_matches <- max(max_matches, matched)
}
cat("Best matching accuracy:", max_matches / length(species), "\n")
```

```{r kmeans-plot, fig.height = 6}
df_km <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_iris$data)),
  curve = rep(1:n, each = m),
  cluster = factor(rep(km$cluster, each = m))
)

ggplot(df_km, aes(x = t, y = value, group = curve, color = cluster)) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "K-Means Clusters (k = 3)",
       x = expression(t), y = expression(f[x](t)), color = "Cluster") +
  theme(legend.position = "bottom")
```

### Automatic Number of Clusters

```{r cluster-optim}
opt <- cluster.optim(fd_iris, ncl.range = 2:6, criterion = "silhouette", seed = 123)
cat("Optimal number of clusters (silhouette):", opt$best.model$ncl, "\n")
```

### Fuzzy C-Means

Fuzzy clustering captures the overlap between versicolor and virginica:

```{r fcm}
fcm <- cluster.fcm(fd_iris, ncl = 3, seed = 123)

# Find versicolor observations with high membership uncertainty
membership <- fcm$membership
max_membership <- apply(membership, 1, max)

df_fuzzy <- data.frame(
  observation = 1:n,
  max_membership = max_membership,
  species = species
)

ggplot(df_fuzzy, aes(x = species, y = max_membership, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                               "virginica" = "#4DAF4A")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  labs(title = "Fuzzy C-Means: Maximum Membership by Species",
       subtitle = "Lower values indicate more overlap between clusters",
       x = NULL, y = "Maximum Membership Probability") +
  theme(legend.position = "none")
```

Setosa observations have near-perfect membership, while versicolor and
virginica show much more uncertainty --- reflecting the genuine overlap between
these species.

## FPCA

Functional PCA on Andrews curves reveals the dominant modes of variation:

```{r fpca}
fpca <- fdata2pc(fd_iris, ncomp = 4)

# Variance explained
var_explained <- fpca$d^2 / sum(fpca$d^2) * 100
cat("Variance explained by first 4 PCs:\n")
for (i in 1:4) {
  cat(sprintf("  PC%d: %.1f%%\n", i, var_explained[i]))
}
cat(sprintf("  Total: %.1f%%\n", sum(var_explained[1:4])))
```

### Variance Plot

```{r fpca-variance}
plot(fpca, type = "variance")
```

### Score Plot

```{r fpca-scores}
df_scores <- data.frame(
  PC1 = fpca$x[, 1],
  PC2 = fpca$x[, 2],
  species = species
)

ggplot(df_scores, aes(x = PC1, y = PC2, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                                "virginica" = "#4DAF4A")) +
  labs(title = "FPCA Score Plot (Andrews Curves)",
       x = sprintf("PC1 (%.1f%%)", var_explained[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained[2]),
       color = "Species") +
  theme(legend.position = "bottom")
```

### Component Functions

```{r fpca-components, fig.height = 6}
df_components <- data.frame(
  t = rep(t_grid, 3),
  loading = c(fpca$rotation$data[1, ], fpca$rotation$data[2, ],
              fpca$rotation$data[3, ]),
  PC = factor(rep(paste0("PC", 1:3), each = m))
)

ggplot(df_components, aes(x = t, y = loading, color = PC)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Principal Component Functions",
       x = expression(t), y = "Loading", color = NULL) +
  theme(legend.position = "bottom")
```

## Variable Ordering

The Andrews transformation maps the first variable to the $1/\sqrt{2}$ term,
the second to $\sin(t)$, and so on. Because lower-frequency Fourier terms
dominate visually, placing important variables first produces more informative
plots.

We rank the iris variables by ANOVA F-statistic:

```{r variable-ordering}
# Rank variables by ANOVA F-statistic
f_stats <- sapply(1:4, function(j) {
  fit <- aov(X[, j] ~ species)
  summary(fit)[[1]]$`F value`[1]
})
names(f_stats) <- colnames(iris)[1:4]
var_order <- order(f_stats, decreasing = TRUE)

cat("Variables ranked by ANOVA F-statistic:\n")
for (i in var_order) {
  cat(sprintf("  %s: F = %.1f\n", names(f_stats)[i], f_stats[i]))
}
```

```{r ordering-comparison, fig.height = 8}
# Transform with original and reordered variables
fd_original <- andrews_transform(X)
fd_reordered <- andrews_transform(X[, var_order])

# Plot both
df_orig <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_original$data)),
  curve = rep(1:n, each = m),
  species = rep(species, each = m),
  ordering = "Original Order"
)

df_reord <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_reordered$data)),
  curve = rep(1:n, each = m),
  species = rep(species, each = m),
  ordering = "Reordered by F-statistic"
)

p1 <- ggplot(df_orig, aes(x = t, y = value, group = curve, color = species)) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  scale_color_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                                "virginica" = "#4DAF4A")) +
  labs(title = "Original Variable Order",
       x = expression(t), y = expression(f[x](t)), color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(df_reord, aes(x = t, y = value, group = curve, color = species)) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  scale_color_manual(values = c("setosa" = "#E41A1C", "versicolor" = "#377EB8",
                                "virginica" = "#4DAF4A")) +
  labs(title = "Reordered by ANOVA F-statistic",
       x = expression(t), y = expression(f[x](t)), color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")

if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p1 / p2
} else {
  print(p1)
  print(p2)
}
```

The reordered plot shows clearer species separation because the most
discriminating variable (petal length) now drives the dominant low-frequency
component.

### Clustering is Invariant to Ordering

The distance preservation theorem guarantees that analytical results are
identical regardless of variable ordering:

```{r ordering-invariance}
km_original <- cluster.kmeans(fd_original, ncl = 3, nstart = 20, seed = 123)
km_reordered <- cluster.kmeans(fd_reordered, ncl = 3, nstart = 20, seed = 123)

# Match cluster labels and compare
best_match <- function(c1, c2) {
  max_matches <- 0
  for (perm in list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))) {
    relabeled <- perm[c2]
    max_matches <- max(max_matches, sum(c1 == relabeled))
  }
  max_matches
}

agreement <- best_match(km_original$cluster, km_reordered$cluster)
cat("Cluster agreement:", agreement, "/", n, "(", round(100 * agreement / n, 1), "%)\n")
```

## Distance Verification

We verify the theoretical result: $\|f_{\mathbf{x}} - f_{\mathbf{y}}\|_{L^2} = \sqrt{\pi} \cdot \|\mathbf{x} - \mathbf{y}\|_2$.

```{r distance-verification}
# Compute pairwise Andrews (L2) distances
dist_andrews <- metric.lp(fd_iris)

# Compute pairwise Euclidean distances on standardized data
dist_euclid <- as.matrix(dist(X))

# Extract upper triangle
idx_upper <- upper.tri(dist_andrews)
d_a <- dist_andrews[idx_upper]
d_e <- dist_euclid[idx_upper]

# Exclude pairs with zero Euclidean distance (iris has 1 duplicate observation)
nonzero <- d_e > 1e-10
ratio <- d_a[nonzero] / d_e[nonzero]
cat("Distance ratio d_andrews / d_euclidean:\n")
cat(sprintf("  Mean:   %.4f\n", mean(ratio)))
cat(sprintf("  Median: %.4f\n", median(ratio)))
cat(sprintf("  SD:     %.2e\n", sd(ratio)))
cat(sprintf("  sqrt(pi) = %.4f\n", sqrt(pi)))
```

```{r distance-scatter}
df_dist <- data.frame(
  euclidean = d_e[nonzero],
  andrews = d_a[nonzero]
)

ggplot(df_dist, aes(x = euclidean, y = andrews)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_abline(slope = sqrt(pi), intercept = 0, color = "red", linewidth = 1) +
  labs(title = "Andrews L2 Distance vs Euclidean Distance",
       subtitle = sprintf("Red line: slope = sqrt(pi) ≈ %.4f", sqrt(pi)),
       x = "Euclidean Distance (standardized data)",
       y = "Andrews L2 Distance") +
  theme_minimal() +
  coord_equal()
```

The points fall exactly on the $\sqrt{\pi}$ line, confirming the distance
preservation theorem.

## When Does FDA Add Value?

The Andrews transformation is isometric: $L^2$ distances between curves equal
$\sqrt{\pi}$ times Euclidean distances between the original vectors. This means
that for methods whose only input is a distance matrix --- PCA, k-means, basic
outlier detection by Mahalanobis distance --- **classical multivariate statistics
and FDA on Andrews curves give the same numerical answers.** FPCA on Andrews
curves and `prcomp()` on the raw matrix produce scores that correlate at $\pm 1$.
K-means produces identical clusters.

So why bother? The value comes not from the transformation itself, but from what
you can do *afterward* in function space:

1. **Functional tools with no tabular equivalent.** Functional depth
   (Fraiman--Muniz, modified band depth), the functional boxplot, the
   outliergram, and tolerance bands are inherently functional concepts. They
   classify outliers by *type* (magnitude vs shape), provide simultaneous
   monitoring of all variables in a single chart, and define nonparametric
   tolerance regions. There is no direct multivariate equivalent of
   `outliergram()` or `tolerance.band()`.

2. **Smoothing as structured regularization.** When $p$ is large or variables
   are noisy, you can smooth the Andrews curves (e.g., via basis expansion or
   penalized splines) before analysis. This damps high-frequency Fourier terms
   --- corresponding to less important variables --- while preserving
   low-frequency structure. Standard PCA has no natural way to impose this
   frequency-aware regularization.

3. **Unified pipeline.** With Andrews curves, depth, outlier detection,
   clustering, FPCA, and the functional boxplot all operate on the same `fdata`
   object with the same distance semantics. In a classical workflow, each of
   these is a separate analysis step with its own data format and assumptions.

4. **Communication.** Eigenfunctions, mean curves, and cluster centroids are
   visual objects. A continuous curve showing the dominant mode of variation
   communicates structure more intuitively than a table of 13 signed loadings,
   especially in reports and presentations.

> **Rule of thumb:** If your analysis stops at PCA + k-means on a clean,
> moderate-$p$ matrix, `prcomp()` and `kmeans()` are simpler and give you the
> same answer. Use Andrews curves when you want the full functional toolbox ---
> depth-based outlier detection, tolerance bands, simultaneous monitoring ---
> or when you're building a pipeline where multiple FDA methods feed into each
> other.

## Best Practices

1. **Standardize before transforming.** Variables on different scales will
   dominate the Andrews function. Always center and scale (or use robust
   standardization) before applying the transformation.

2. **Order variables by importance for visualization.** Place the most
   informative variables first so they correspond to low-frequency Fourier
   terms that are visually dominant. Use ANOVA F-statistics, variance, or
   domain knowledge to rank.

3. **Use $m \geq 200$ grid points.** Fewer points can introduce numerical
   artifacts in distance computations. The default of 200 is sufficient for
   most applications.

4. **Works best for small-to-moderate $p$.** With many variables, higher-order
   Fourier terms oscillate rapidly and contribute little visual information.
   For high-dimensional data, consider dimension reduction (e.g., PCA) before
   applying the Andrews transformation.

5. **Analytical results are ordering-invariant.** While variable ordering
   affects the appearance of plots, all distance-based methods (clustering,
   depth, outlier detection) give identical results regardless of ordering.

## Summary

| FDA Method | What It Reveals on Andrews Curves |
|---|---|
| `depth.FM()` / `depth.MBD()` | Centrality of multivariate observations |
| `median()` / `trimmed()` | Robust center of the multivariate distribution |
| `outliers.depth.pond()` | Observations far from the multivariate center |
| `outliergram()` | Shape outliers in the multivariate space |
| `magnitudeshape()` | Magnitude vs shape decomposition of outlyingness |
| `cluster.kmeans()` | Euclidean-distance-based partitioning |
| `cluster.fcm()` | Soft cluster memberships (overlap detection) |
| `fdata2pc()` | Dominant directions of multivariate variation |
| `metric.lp()` | $\sqrt{\pi} \times$ Euclidean distance |

## See Also

- `vignette("depth-functions", package = "fdars")` — functional depth measures and ranking
- `vignette("clustering", package = "fdars")` — functional data clustering
- `vignette("outlier-detection", package = "fdars")` — functional outlier detection methods

## References

- Andrews, D.F. (1972). Plots of high-dimensional data. *Biometrics*, 28(1),
  125--136.
- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  Springer. (Chapter 1: Introduction to FDA)
