---
title: "Conformal Prediction for Classification"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Conformal prediction for classification produces **prediction sets** — sets of
plausible labels with a coverage guarantee — instead of point predictions.
For a new observation $x_{\text{new}}$, the conformal prediction set
$C_\alpha(x_{\text{new}})$ satisfies:

$$P(y_{\text{new}} \in C_\alpha) \geq 1 - \alpha$$

for any distribution, with no assumptions on the data-generating process.

Unlike regression conformal (which produces intervals), classification conformal
answers: "which classes are plausible for this observation?" A set of size 1
means high confidence; larger sets indicate ambiguity.

```{=html}
<img src="../reference/figures/conformal-classification-overview.svg" alt="Conformal classification: prediction sets, three methods, and scoring rules" style="width:100%; max-width:700px;">
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

## Simulated Data

We simulate a three-class functional classification problem where each class
has a distinct mean curve:

```{r data}
n_per_class <- 30
n <- 3 * n_per_class
m <- 50
t_grid <- seq(0, 1, length.out = m)

X <- matrix(0, n, m)
for (i in 1:n_per_class) {
  X[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.2)
  X[n_per_class + i, ] <- cos(2 * pi * t_grid) + rnorm(m, sd = 0.2)
  X[2 * n_per_class + i, ] <- 0.5 * t_grid + rnorm(m, sd = 0.2)
}
labels <- rep(1:3, each = n_per_class)

# Train/test split
train_idx <- c(1:25, 31:55, 61:85)
test_idx <- setdiff(1:n, train_idx)

fd_train <- fdata(X[train_idx, ], argvals = t_grid)
fd_test <- fdata(X[test_idx, ], argvals = t_grid)
y_train <- labels[train_idx]
y_test <- labels[test_idx]
```

```{r plot-data, fig.height = 4}
df_curves <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(X)),
  curve = rep(1:n, each = m),
  class = factor(rep(labels, each = m))
)

ggplot(df_curves, aes(x = .data$t, y = .data$value,
                      group = .data$curve, color = .data$class)) +
  geom_line(alpha = 0.3) +
  labs(title = "Three-Class Functional Data",
       x = "t", y = "X(t)", color = "Class")
```

## Scoring Rules

Conformal classification uses a **nonconformity score** to measure how
surprising a label is for a given observation. **fdars** supports two scoring
rules:

| Score | Name | Description | Set size |
|:------|:-----|:------------|:---------|
| **LAC** | Least Ambiguous Criterion | $1 - \hat{p}(y \mid x)$ | Smaller (adaptive) |
| **APS** | Adaptive Prediction Sets | Cumulative probability until true class included | Larger (conservative) |

**LAC** produces smaller prediction sets on average but provides only marginal
coverage. **APS** guarantees coverage even conditionally on the true class,
at the cost of slightly larger sets.

## Split Conformal Classification

The simplest approach splits the data into a proper training set and a
calibration set. The model is fitted on the training set, nonconformity scores
are computed on the calibration set, and the $(1-\alpha)$-quantile of scores
determines the prediction set threshold.

```{r split-conformal}
# Split conformal with LDA classifier and LAC scoring
conf_split <- conformal.classif(
  fd_train, y_train, fd_test,
  ncomp = 5, classifier = "lda",
  score.type = "lac",
  cal.fraction = 0.25,
  alpha = 0.10, seed = 42
)

cat("Predicted classes:", conf_split$predicted_classes, "\n")
cat("Average set size:", round(conf_split$average_set_size, 2), "\n")
cat("Coverage:", round(conf_split$coverage * 100, 1), "%\n")
```

### Examining Prediction Sets

```{r set-sizes}
df_sets <- data.frame(
  Observation = seq_along(conf_split$set_sizes),
  Set_Size = conf_split$set_sizes,
  Correct = factor(ifelse(conf_split$predicted_classes == y_test,
                          "Correct", "Wrong"))
)

ggplot(df_sets, aes(x = .data$Observation, y = .data$Set_Size,
                    fill = .data$Correct)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Correct" = "#2E8B57", "Wrong" = "#D55E00")) +
  labs(title = "Prediction Set Sizes (Split Conformal, LAC)",
       subtitle = "Size 1 = confident prediction, size > 1 = ambiguous",
       x = "Test Observation", y = "Set Size", fill = NULL)
```

## CV+ Conformal Classification

Cross-conformal (CV+) avoids the data-splitting penalty of split conformal.
Each fold serves as the calibration set for the model trained on the remaining
folds. All data contributes to both training and calibration.

```{r cv-conformal}
cv_conf <- cv.conformal.classification(
  fd_train, y_train, fd_test,
  ncomp = 5, classifier = "lda",
  score.type = "lac",
  n.folds = 5,
  alpha = 0.10, seed = 42
)

cat("CV+ average set size:", round(cv_conf$average_set_size, 2), "\n")
cat("CV+ coverage:", round(cv_conf$coverage * 100, 1), "%\n")
```

## Generic Conformal Classification

If you already have a fitted `functional.logistic` model, you can construct
conformal prediction sets without re-fitting. Only binary classification
(2 classes) is supported.

> **Note:** Generic conformal uses in-sample calibration (the model was
> trained on all data including the calibration set). The coverage guarantee
> is broken — use this as a fast heuristic only. For valid coverage, prefer
> `conformal.classif()` or `cv.conformal.classification()`.

```{r generic-conformal}
# Binary classification for generic conformal (requires logistic model)
# functional.logistic expects 0/1 labels
idx_train_bin <- y_train %in% c(1, 2)
idx_test_bin <- y_test %in% c(1, 2)
train_binary <- fd_train[idx_train_bin, ]
test_binary <- fd_test[idx_test_bin, ]
y_train_bin <- as.integer(y_train[idx_train_bin] == 2)  # recode to 0/1
y_test_bin <- as.integer(y_test[idx_test_bin] == 2)

# Fit logistic model
log_model <- functional.logistic(train_binary, y_train_bin, ncomp = 3)

# Generic conformal from the fitted model
gen_conf <- conformal.generic.classification(
  log_model, train_binary, y_train_bin, test_binary,
  score.type = "lac",
  cal.fraction = 0.25, alpha = 0.10, seed = 42
)

cat("Generic conformal coverage:", round(gen_conf$coverage * 100, 1), "%\n")
cat("Average set size:", round(gen_conf$average_set_size, 2), "\n")
```

## Comparing Classifiers

Split conformal works with different base classifiers. Compare LDA, QDA,
and kNN:

```{r compare-classifiers}
classifiers <- c("lda", "qda", "knn")
results <- lapply(classifiers, function(clf) {
  res <- conformal.classif(
    fd_train, y_train, fd_test,
    ncomp = 5, classifier = clf,
    score.type = "lac",
    cal.fraction = 0.25,
    alpha = 0.10, seed = 42
  )
  data.frame(
    classifier = toupper(clf),
    coverage = res$coverage,
    avg_set_size = res$average_set_size
  )
})
df_clf <- do.call(rbind, results)
knitr::kable(df_clf, digits = 3,
             col.names = c("Classifier", "Coverage", "Avg Set Size"))
```

## LAC vs APS Scoring

```{r lac-vs-aps}
conf_lac <- conformal.classif(
  fd_train, y_train, fd_test,
  ncomp = 5, classifier = "lda",
  score.type = "lac",
  cal.fraction = 0.25, alpha = 0.10, seed = 42
)

conf_aps <- conformal.classif(
  fd_train, y_train, fd_test,
  ncomp = 5, classifier = "lda",
  score.type = "aps",
  cal.fraction = 0.25, alpha = 0.10, seed = 42
)

df_score <- data.frame(
  Scoring = rep(c("LAC", "APS"), each = length(y_test)),
  Set_Size = c(conf_lac$set_sizes, conf_aps$set_sizes)
)

ggplot(df_score, aes(x = .data$Scoring, y = .data$Set_Size,
                     fill = .data$Scoring)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("LAC" = "#4A90D9", "APS" = "#D55E00")) +
  labs(title = "Prediction Set Size: LAC vs APS",
       subtitle = "APS produces larger sets but stronger conditional coverage",
       y = "Set Size") +
  theme(legend.position = "none")
```

## Method Comparison

| Method | Function | Base models | Coverage | Sets |
|:-------|:---------|:------------|:---------|:-----|
| Split | `conformal.classif()` | LDA, QDA, kNN | $\geq 1 - \alpha$ | Tightest |
| CV+ | `cv.conformal.classification()` | LDA, QDA, kNN | $\geq 1 - 2\alpha$ | Tighter (no split penalty) |
| Generic | `conformal.generic.classification()` | Logistic (pre-fitted, binary only) | Heuristic only | From existing model |

**Choosing a method:**

- **Split conformal** when data is plentiful and you want the strongest
  coverage guarantee ($1 - \alpha$).
- **CV+** when data is limited and you want tighter sets — all data is used
  for both training and calibration.
- **Generic** when you already have a fitted logistic model and want a
  quick heuristic — but note that coverage is not guaranteed (in-sample
  calibration).

## See Also

- `vignette("articles/uncertainty-quantification")` — conformal prediction for
  regression (intervals, not sets)
- `vignette("articles/functional-classification")` — base classification methods
  (LDA, QDA, kNN)

## References

- Vovk, V., Gammerman, A. and Shafer, G. (2005). *Algorithmic Learning in
  a Random World*. Springer.

- Sadinle, M., Lei, J. and Wasserman, L. (2019). Least ambiguous set-valued
  classifiers with bounded error levels. *Journal of the American Statistical
  Association*, 114(525), 223--234.

- Romano, Y., Sesia, M. and Candes, E.J. (2020). Classification with valid
  and adaptive coverage. *Advances in Neural Information Processing Systems*,
  33.
