---
title: "Conformal Prediction Guide"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

**fdars** provides 14 conformal prediction functions covering regression,
classification, and elastic models. This guide helps you choose the right
function for your problem and summarizes the available options.

All conformal methods share the same guarantee: for coverage level
$1 - \alpha$,

$$P(y_{\text{new}} \in C_\alpha) \geq 1 - \alpha$$

holds for any data distribution, with no parametric assumptions.

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Decision Flowchart

```{=html}
<img src="../reference/figures/cpg-decision-flowchart.svg" alt="Conformal prediction decision flowchart" style="width:100%; max-width:700px;">
```

Choose your conformal method based on three questions:

**1. Regression or classification?**

- **Regression** (predict a number $\hat{y}$) $\rightarrow$ prediction *intervals*
- **Classification** (predict a label $\hat{y} \in \{1, \ldots, K\}$) $\rightarrow$ prediction *sets*

**2. Which base model?**

| Model | Regression | Classification |
|:------|:-----------|:---------------|
| `fregre.lm` | `conformal.fregre.lm()` | — |
| `fregre.np` | `conformal.fregre.np()` | — |
| Elastic regression | `conformal.elastic.regression()` | — |
| Elastic PCR | `conformal.elastic.pcr()` | — |
| LDA / QDA / kNN | — | `conformal.classif()` |
| Logistic | `conformal.logistic()` | `conformal.elastic.logistic()` |

**3. How much data can you afford?**

| Variant | Data use | # fits | Guarantee | Function suffix |
|:--------|:---------|:-------|:----------|:---------------|
| **Split** | Wastes calibration fraction | 1 | $\geq 1 - \alpha$ | model-specific |
| **CV+** | All data used | $K$ folds | $\geq 1 - 2\alpha$ | `cv.conformal.*()` |
| **Jackknife+** | All data used | $n$ LOO fits | $\geq 1 - 2\alpha$ | `jackknife.plus()` |
| **Generic** | Pre-fitted model | 0 | $\geq 1 - \alpha$ | `conformal.generic.*()` |

## Complete Function Reference

### Split Conformal — Regression

```{=html}
<img src="../reference/figures/cpg-split-regression.svg" alt="Split conformal regression pipeline" style="width:100%; max-width:700px;">
```

Each function takes training `fdata`, response `y`, test `fdata`, and returns
prediction intervals.

| Function | Model | Key parameters |
|:---------|:------|:---------------|
| `conformal.fregre.lm()` | FPC linear | `ncomp`, `cal.fraction` |
| `conformal.fregre.np()` | Nonparametric | `ncomp`, `cal.fraction` |
| `conformal.elastic.regression()` | Elastic regression | `ncomp`, `cal.fraction` |
| `conformal.elastic.pcr()` | Elastic PCR | `ncomp`, `cal.fraction` |
| `conformal.logistic()` | Logistic (binary) | `ncomp`, `cal.fraction` |

All return: `predictions`, `lower`, `upper`, `residual.quantile`, `coverage`.

### Split Conformal — Classification

```{=html}
<img src="../reference/figures/cpg-split-classification.svg" alt="Split conformal classification pipeline" style="width:100%; max-width:700px;">
```

| Function | Model | Key parameters |
|:---------|:------|:---------------|
| `conformal.classif()` | LDA, QDA, kNN | `classifier`, `score.type` |
| `conformal.elastic.logistic()` | Elastic logistic | `score.type`, `cal.fraction` |

Returns: `predicted_classes`, `set_sizes`, `average_set_size`, `coverage`,
`score_quantile`.

### Cross-Conformal (CV+)

```{=html}
<img src="../reference/figures/cpg-advanced-variants.svg" alt="Advanced conformal variants: CV+, Jackknife+, Generic" style="width:100%; max-width:700px;">
```

| Function | Task | Key parameters |
|:---------|:-----|:---------------|
| `cv.conformal.regression()` | Regression | `method` ("fregre.lm" or "fregre.np"), `n.folds` |
| `cv.conformal.classification()` | Classification | `classifier`, `score.type`, `n.folds` |

### Jackknife+

| Function | Task | Key parameters |
|:---------|:-----|:---------------|
| `jackknife.plus()` | Regression | `method` ("fregre.lm" or "fregre.np") |

### Generic Conformal (pre-fitted model)

| Function | Task | Input model |
|:---------|:-----|:------------|
| `conformal.generic.regression()` | Regression | Fitted `fregre.lm` object |
| `conformal.generic.classification()` | Classification | Fitted `functional.logistic` object |

## Quick Examples

### Regression: Split vs CV+ vs Jackknife+

```{r regression-example}
set.seed(42)
n <- 80; m <- 50
t_grid <- seq(0, 1, length.out = m)

X <- matrix(0, n, m)
for (i in 1:n) X[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)
y <- rowMeans(X) + rnorm(n, sd = 0.3)

fd_train <- fdata(X[1:60, ], argvals = t_grid)
fd_test <- fdata(X[61:80, ], argvals = t_grid)
y_train <- y[1:60]; y_test <- y[61:80]

# Split conformal
split_res <- conformal.fregre.lm(fd_train, y_train, fd_test,
                                  ncomp = 3, alpha = 0.10, seed = 42)

# CV+ conformal
cv_res <- cv.conformal.regression(fd_train, y_train, fd_test,
                                   method = "fregre.lm", ncomp = 3,
                                   n.folds = 5, alpha = 0.10, seed = 42)

# Jackknife+
jk_res <- jackknife.plus(fd_train, y_train, fd_test,
                          method = "fregre.lm", ncomp = 3, alpha = 0.10)

# Compare widths
cat("Split mean width:     ", round(mean(split_res$upper - split_res$lower), 4), "\n")
cat("CV+ mean width:       ", round(mean(cv_res$upper - cv_res$lower), 4), "\n")
cat("Jackknife+ mean width:", round(mean(jk_res$upper - jk_res$lower), 4), "\n")
```

```{r regression-plot, fig.height = 4}
df_width <- data.frame(
  Method = rep(c("Split", "CV+", "Jackknife+"), each = 20),
  Width = c(split_res$upper - split_res$lower,
            cv_res$upper - cv_res$lower,
            jk_res$upper - jk_res$lower)
)
df_width$Method <- factor(df_width$Method, levels = c("Split", "CV+", "Jackknife+"))

ggplot(df_width, aes(x = .data$Method, y = .data$Width, fill = .data$Method)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Split" = "#2E8B57", "CV+" = "#4A90D9",
                                "Jackknife+" = "#D55E00")) +
  labs(title = "Prediction Interval Width by Method",
       y = "Interval Width") +
  theme(legend.position = "none")
```

### Classification: Split vs CV+

```{r classification-example}
set.seed(42)
n_cl <- 60; m_cl <- 50
t_cl <- seq(0, 1, length.out = m_cl)
X_cl <- matrix(0, n_cl, m_cl)
for (i in 1:30) X_cl[i, ] <- sin(2 * pi * t_cl) + rnorm(m_cl, sd = 0.2)
for (i in 31:60) X_cl[i, ] <- cos(2 * pi * t_cl) + rnorm(m_cl, sd = 0.2)
y_cl <- rep(1:2, each = 30)

fd_cl_train <- fdata(X_cl[c(1:25, 31:55), ], argvals = t_cl)
fd_cl_test <- fdata(X_cl[c(26:30, 56:60), ], argvals = t_cl)
y_cl_train <- y_cl[c(1:25, 31:55)]

# Split conformal classification
split_cl <- conformal.classif(fd_cl_train, y_cl_train, fd_cl_test,
                               ncomp = 3, classifier = "lda",
                               score.type = "lac", alpha = 0.10, seed = 42)

# CV+ conformal classification
cv_cl <- cv.conformal.classification(fd_cl_train, y_cl_train, fd_cl_test,
                                      ncomp = 3, classifier = "lda",
                                      score.type = "lac", n.folds = 5,
                                      alpha = 0.10, seed = 42)

cat("Split avg set size:", round(split_cl$average_set_size, 2), "\n")
cat("CV+ avg set size:  ", round(cv_cl$average_set_size, 2), "\n")
```

## Practical Guidance

### Sample Size Requirements

- **Split conformal**: needs enough calibration data for a reliable quantile.
  With `cal.fraction = 0.25` and $n = 100$, you have 25 calibration points.
  Rule of thumb: $n_{\text{cal}} \geq 1/\alpha$ (e.g., $\geq 10$ for
  $\alpha = 0.10$).
- **CV+**: works well with smaller $n$ since all data contributes. Effective
  with $n \geq 30$.
- **Jackknife+**: most data-efficient but requires $n$ model fits. Practical
  for $n \leq 500$.

### Computational Cost

| Method | Model fits | Relative cost |
|:-------|:-----------|:--------------|
| Split | 1 | Fastest |
| Generic | 0 (pre-fitted) | Fastest |
| CV+ (5-fold) | 5 | Moderate |
| Jackknife+ | $n$ | Slowest |

### Common Pitfalls

1. **Too few calibration points**: split conformal with small $n$ and small
   `cal.fraction` gives noisy intervals. Use CV+ instead.
2. **Too many FPC components**: overfitting the base model produces optimistic
   residuals, widening conformal intervals. Use `model.selection.ncomp()` to
   choose `ncomp`.
3. **Confusing coverage guarantees**: split and generic give $1 - \alpha$;
   CV+ and jackknife+ give $1 - 2\alpha$ in theory but often achieve
   near-$1 - \alpha$ empirically.

## See Also

- `vignette("articles/uncertainty-quantification")` — detailed UQ with
  parametric intervals, LOO-CV, and bootstrap CI
- `vignette("articles/conformal-classification")` — in-depth classification
  conformal with scoring rules and classifier comparison
- `vignette("articles/scalar-on-function")` — base regression models

## References

- Vovk, V., Gammerman, A. and Shafer, G. (2005). *Algorithmic Learning in
  a Random World*. Springer.

- Barber, R.F., Candes, E.J., Ramdas, A. and Tibshirani, R.J. (2021).
  Predictive inference with the jackknife+. *Annals of Statistics*, 49(1),
  486--507.

- Romano, Y., Patterson, E. and Candes, E. (2019). Conformalized quantile
  regression. *Advances in Neural Information Processing Systems*, 32.
