---
title: "Cross-Validation for Functional Data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Why Cross-Validation?

A model's training error is always optimistic: it measures how well the model
fits data it has already seen, not how well it will predict new observations.
Cross-validation (CV) provides an honest estimate of **out-of-sample**
performance by systematically holding out portions of the data for evaluation.

In functional data analysis, cross-validation serves two purposes:

1. **Hyperparameter selection** — choosing the number of principal components,
   the regularization parameter $\lambda$, or the bandwidth $h$
2. **Model evaluation** — estimating how well a fitted model will perform on
   new functional observations

The existing `*.cv` functions in **fdars** (`fregre.pc.cv`, `fregre.basis.cv`,
`fregre.np.cv`) handle purpose (1). The `cv.fdata()` function provides a
unified framework for both, producing **out-of-fold (OOF) predictions** where
every observation is predicted exactly once.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## K-Fold Cross-Validation

In $k$-fold CV, the dataset of $n$ observations is partitioned into $k$
roughly equal-sized **folds**. For each fold $j = 1, \ldots, k$:

1. Hold out fold $j$ as the **test set**
2. Train the model on the remaining $k-1$ folds
3. Predict on the held-out fold

$$\text{CV}(k) = \frac{1}{n} \sum_{j=1}^{k} \sum_{i \in \mathcal{F}_j} L(y_i, \hat{y}_i^{(-j)})$$

where $\mathcal{F}_j$ is the set of indices in fold $j$, $\hat{y}_i^{(-j)}$ is
the prediction for observation $i$ from the model trained without fold $j$,
and $L$ is a loss function (e.g., squared error).

The key property: **every observation is predicted exactly once**, when it is in
the test fold. The resulting predictions are called **out-of-fold (OOF)
predictions**.

### Choosing $k$

| $k$ | Name | Bias | Variance | Cost |
|-----|------|------|----------|------|
| $n$ | Leave-one-out (LOO) | Low | High | $n$ model fits |
| 10 | 10-fold (standard) | Moderate | Moderate | 10 model fits |
| 5 | 5-fold | Higher | Lower | 5 model fits |

In practice, $k = 5$ or $k = 10$ provides a good bias-variance tradeoff. LOO
is used by the hat-matrix-based `*.cv` functions for linear methods where it
can be computed efficiently.

## Simple Example

```{r data}
set.seed(42)
n <- 80
m <- 50
t_grid <- seq(0, 1, length.out = m)

X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) * rnorm(1, 1, 0.3) +
            cos(4 * pi * t_grid) * rnorm(1, 0, 0.2) +
            rnorm(m, sd = 0.1)
}
fd <- fdata(X, argvals = t_grid)

beta_true <- sin(2 * pi * t_grid)
y <- numeric(n)
for (i in 1:n) {
  y[i] <- sum(beta_true * X[i, ]) / m + rnorm(1, sd = 0.5)
}
```

### Running `cv.fdata`

The interface is simple: pass any fitting function and `cv.fdata` handles the
fold loop.

```{r cv-basic}
cv_result <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.pc(fd, y, ncomp = 3),
  kfold = 5, seed = 42)
print(cv_result)
```

The return value contains:

- `$oof.predictions` — one prediction per observation (length $n$)
- `$folds` — which fold each observation belongs to
- `$fold.models` — the fitted model from each fold
- `$metrics` — overall RMSE, MAE, R2
- `$fold.metrics` — per-fold metrics

```{r cv-check}
# Every observation predicted exactly once
cat("Predictions:", length(cv_result$oof.predictions), "\n")
cat("Any missing:", any(is.na(cv_result$oof.predictions)), "\n")
cat("Fold sizes:", table(cv_result$folds), "\n")
```

## Nested Cross-Validation

When a model has hyperparameters that are themselves selected by
cross-validation, a single level of CV is **not enough** for unbiased
evaluation. The hyperparameter selection uses information from the test fold
(indirectly), leading to optimistic error estimates.

**Nested CV** separates these two tasks:

- **Outer loop** ($k_{\text{out}}$ folds): produces OOF predictions for
  evaluation
- **Inner loop** ($k_{\text{in}}$ folds, on training data only): selects
  hyperparameters

```{=html}
<img src="../reference/figures/nested-cv-diagram.svg" alt="Nested Cross-Validation Diagram" style="width:100%; max-width:700px;">
```

With `cv.fdata`, nested CV is achieved by putting the existing `*.cv` function
inside `fit.fn`:

```{r nested}
cv_nested <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) {
    # Inner CV selects optimal ncomp on training data only
    cv_inner <- fregre.pc.cv(fd, y, ncomp.range = 1:10, kfold = 5)
    cv_inner$model
  },
  kfold = 5, seed = 42)
print(cv_nested)
```

The hyperparameter selected can vary across outer folds — this is expected and
desirable, as it reflects genuine uncertainty:

```{r nested-ncomp}
ncomps <- sapply(cv_nested$fold.models, function(m) m$ncomp)
cat("ncomp per fold:", ncomps, "\n")
```

## Stratified Folds

When the response distribution is skewed or has outliers, purely random fold
assignment can create folds with very different $y$ distributions. This adds
noise to the CV estimate.

**Stratified** fold assignment bins the response into quantile groups and
samples within each bin, ensuring every fold has a similar $y$ distribution.

```{=html}
<img src="../reference/figures/stratified-folds-diagram.svg" alt="Stratified Folds Diagram" style="width:100%; max-width:700px;">
```

`cv.fdata` uses stratified folds by default (`stratified = TRUE`).

```{r stratified, fig.height = 3.5}
# Skewed response
set.seed(1)
y_skew <- exp(y)

folds_strat <- fdars::: .create_folds(y_skew, kfold = 5, type = "regression",
                                       stratified = TRUE, seed = 1)
folds_rand <- fdars::: .create_folds(y_skew, kfold = 5, type = "regression",
                                      stratified = FALSE, seed = 1)

df_folds <- data.frame(
  y = rep(y_skew, 2),
  Fold = factor(c(folds_strat, folds_rand)),
  Type = rep(c("Stratified", "Random"), each = n)
)

ggplot(df_folds, aes(x = Fold, y = y)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  facet_wrap(~ Type) +
  labs(title = "Stratified vs Random Fold Assignment (skewed response)",
       y = "y (skewed)")
```

For classification, stratification preserves class proportions in each fold.

## Repeated Cross-Validation

A single k-fold split can produce noisy estimates — different random partitions
can give different OOF predictions and metrics. **Repeated cross-validation**
runs the entire k-fold procedure multiple times with different random fold
assignments, giving each observation multiple predictions.

This lets you assess **prediction variability**: how much a prediction changes
depending on which observations happen to be in its training set.

```{=html}
<img src="../reference/figures/repeated-cv-diagram.svg" alt="Repeated Cross-Validation Diagram" style="width:100%; max-width:700px;">
```

```{r repeated-cv}
cv_rep <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.pc(fd, y, ncomp = 3),
  kfold = 5, nrep = 10, seed = 42)
print(cv_rep)
```

The `$oof.matrix` contains the raw predictions from each repetition (an
$n \times \text{nrep}$ matrix), while `$oof.predictions` contains the
aggregated (mean) predictions. The `$oof.sd` vector gives the standard
deviation of predictions across repetitions for each observation.

```{r repeated-variability}
# Observations with highest prediction variability
top_var <- head(sort(cv_rep$oof.sd, decreasing = TRUE))
cat("Top 6 per-observation SDs:\n")
print(round(top_var, 4))

# Per-repetition metrics
knitr::kable(cv_rep$rep.metrics, digits = 4,
             caption = "Metrics across 10 repetitions")
```

The `plot` method for repeated CV shows error bars ($\pm 1$ SD) on the
observed-vs-predicted scatter, plus a histogram of prediction variability:

```{r repeated-plot, fig.height = 4}
plot(cv_rep)
```

High per-observation variability can flag **influential observations** or
regions of the predictor space where the model is unstable.

## Comparing Methods Fairly

A key advantage of `cv.fdata` is that it allows **fair comparisons** between
methods by fixing the fold assignments with the `seed` parameter. When two
methods use the same seed, they are evaluated on identical train/test splits,
removing a source of variability.

```{r compare}
cv_pc <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.pc(fd, y, ncomp = 3),
  kfold = 5, seed = 42)

cv_basis <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.basis(fd, y, nbasis = 11, lambda = 1),
  kfold = 5, seed = 42)

cv_np <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.np(fd, y, type.S = "S.NW"),
  kfold = 5, seed = 42)

comparison <- data.frame(
  Method = c("PC Regression", "Basis Regression", "Nadaraya-Watson"),
  RMSE = round(c(cv_pc$metrics$RMSE, cv_basis$metrics$RMSE,
                  cv_np$metrics$RMSE), 4),
  MAE = round(c(cv_pc$metrics$MAE, cv_basis$metrics$MAE,
                 cv_np$metrics$MAE), 4),
  R2 = round(c(cv_pc$metrics$R2, cv_basis$metrics$R2,
                cv_np$metrics$R2), 4)
)
knitr::kable(comparison, caption = "5-fold OOF performance (same folds)")
```

## Per-Fold Stability

Large variation in per-fold metrics can indicate:

- **Small sample size** — not enough data for stable estimates
- **Outliers** — one fold contains unusual observations
- **Model sensitivity** — the method is fragile to small data changes

```{r stability}
knitr::kable(cv_nested$fold.metrics, digits = 4,
             caption = "Per-fold metrics (nested CV)")
```

## Visualising Results

The `plot` method for `cv.fdata` objects produces observed-vs-predicted and
per-fold residual boxplots:

```{r plot, fig.height = 4}
plot(cv_nested)
```

For multi-method comparison, build a ggplot directly:

```{r compare-plot, fig.height = 4}
df_oof <- data.frame(
  Observed = rep(y, 3),
  Predicted = c(cv_pc$oof.predictions, cv_basis$oof.predictions,
                cv_np$oof.predictions),
  Method = rep(c("PC Regression", "Basis Regression", "Nadaraya-Watson"),
               each = n)
)

ggplot(df_oof, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.5, colour = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "red") +
  facet_wrap(~ Method) +
  labs(title = "Out-of-Fold Predictions by Method",
       x = "Observed", y = "OOF Predicted")
```

## Custom Predict Functions

By default, `cv.fdata` calls `predict(model, newdata)`. For models that need
a different prediction interface, pass a `predict.fn`:

```{r custom}
cv_custom <- cv.fdata(fd, y,
  fit.fn = function(fd, y, ...) fregre.basis(fd, y, nbasis = 11, lambda = 0.1),
  predict.fn = function(model, newdata) predict(model, newdata),
  kfold = 5, seed = 42)
cat("RMSE:", round(cv_custom$metrics$RMSE, 4), "\n")
```

This is especially useful for classification models (like `fclassif`) that
may not have a standard `predict` method.

## Classification

`cv.fdata` supports classification when `y` is a factor. Fold stratification
preserves class proportions, and metrics include accuracy and a confusion
matrix.

```{r classif}
# Binary classification from the simulated data
y_class <- factor(ifelse(y > median(y), "high", "low"))

cv_classif <- cv.fdata(fd, y_class,
  fit.fn = function(fd, y, ...) {
    fregre.pc(fd, as.numeric(y), ncomp = 3)
  },
  predict.fn = function(model, newdata) {
    preds <- predict(model, newdata)
    ifelse(preds > 1.5, "high", "low")
  },
  kfold = 5, seed = 42)

cat("Accuracy:", round(cv_classif$metrics$accuracy, 3), "\n")
cv_classif$metrics$confusion
```

## Relationship to Existing `*.cv` Functions

| Function | Purpose | Inner tuning | OOF predictions |
|---|---|---|---|
| `fregre.pc.cv` | Select optimal `ncomp` | LOO or k-fold | No |
| `fregre.basis.cv` | Select optimal `lambda` | LOO or k-fold | No |
| `fregre.np.cv` | Select optimal `h` | LOO or k-fold | No |
| `fregre.lm.cv` | Select optimal `k` | k-fold | No |
| `fclassif.cv` | Select optimal `ncomp` | LOO or k-fold | No |
| **`cv.fdata`** | **Evaluate any model** | **Via `fit.fn`** | **Yes** |

Use `*.cv` functions *inside* `cv.fdata`'s `fit.fn` for proper nested CV.

## Summary

- **`cv.fdata`** wraps any fit/predict workflow in a k-fold CV loop
- **OOF predictions** give honest generalisation estimates — each observation
  predicted exactly once
- **Nested CV** = `cv.fdata` (outer) + `*.cv` (inner) → unbiased evaluation
  with automatic tuning
- **Stratified folds** (default) ensure balanced $y$ distribution
- **Same seed** → same folds → fair method comparison
- **`$fold.models`** lets you inspect per-fold hyperparameters

## See Also

- `vignette("articles/regression")` — regression methods overview
- `vignette("articles/example-cross-validation")` — real-data example with Tecator NIR spectra
- `vignette("articles/functional-classification")` — classification methods
