---
title: "Functional Depth Functions"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Functional depth generalizes the notion of "how central is this observation?" from
univariate statistics (where the median has maximum depth) to the space of curves.
Given a sample of curves, each curve receives a depth value in $[0, 1]$: high values
indicate typical, central curves; low values indicate unusual, peripheral curves.

Depth is a fundamental building block in functional data analysis because it provides:

- **Nonparametric ranking** --- order curves from most central to most outlying without assuming a distribution
- **Robust summary statistics** --- the functional median (deepest curve) and trimmed mean are resistant to outliers
- **Outlier detection** --- curves with low depth are candidate outliers, used by `outliergram()` and `magnitudeshape()`
- **Tolerance bands** --- the central region containing the deepest $\alpha\%$ of curves defines simultaneous confidence bands
- **Functional boxplot** --- `boxplot(fd)` uses depth to define the 50% central region and fence
- **Classification** --- depth values serve as features for DD-classifiers (`fclassif(..., method = "DD")`)
- **Quality control** --- streaming depth enables online monitoring of new curves against a reference sample

```{=html}
<img src="../reference/figures/depth-functions-diagram.svg" alt="Functional Depth: Ranking Curves by Centrality" style="width:100%; max-width:720px; display:block; margin:1em auto;" />
```

### Available Methods

| Method | Function | Description | Supports 2D |
|--------|----------|-------------|:-----------:|
| **FM** | `depth.FM()` | Fraiman-Muniz: integrates pointwise univariate depth | Yes |
| **MBD** | `depth.MBD()` | Modified Band Depth: proportion of time inside random bands | No |
| **BD** | `depth.BD()` | Band Depth: fraction of bands fully containing the curve | No |
| **MEI** | `depth.MEI()` | Modified Epigraph Index: proportion above/below other curves | No |
| **mode** | `depth.mode()` | Modal Depth: kernel density in function space | Yes |
| **RP** | `depth.RP()` | Random Projection: halfspace depth of projected scores | Yes |
| **RT** | `depth.RT()` | Random Tukey: Tukey depth via random projections | Yes |
| **RPD** | `depth.RPD()` | RP with Derivatives: includes curve shape information | No |
| **FSD** | `depth.FSD()` | Functional Spatial Depth: based on spatial signs | Yes |
| **KFSD** | `depth.KFSD()` | Kernel FSD: smoothed functional spatial depth | Yes |
| **streaming** | `streaming.depth()` | O(T log N) per query via pre-sorted reference | No |

All methods are accessible through the unified `depth(fd, method = "...")` interface.

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

## Generating Example Data

We simulate a sample of smooth curves with two outliers: one shifted in magnitude
and one with a different shape.

```{r generate-data}
n <- 30
m <- 100
argvals <- seq(0, 1, length.out = m)

# Normal curves: sine + individual variation
data <- matrix(0, n, m)
for (i in 1:(n - 2)) {
  data[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.2) + rnorm(m, 0, 0.05)
}

# Magnitude outlier (shifted up)
data[n - 1, ] <- sin(2 * pi * argvals) + 2.5 + rnorm(m, 0, 0.05)

# Shape outlier (different frequency)
data[n, ] <- cos(4 * pi * argvals) + rnorm(m, 0, 0.05)

fd <- fdata(data, argvals = argvals)

plot(fd, main = "30 Curves with 2 Outliers",
     xlab = "t", ylab = "X(t)")
```

Curve 29 (magnitude outlier) is shifted up; curve 30 (shape outlier) has a
different oscillation pattern.

## Pointwise Depth Methods

### Fraiman-Muniz Depth (FM)

FM depth integrates univariate depth (based on the empirical CDF) at each time
point over the domain:

$$D_{FM}(X) = \int_0^1 D_n(X(t))\, dt, \quad D_n(x) = 1 - |2 F_n(x) - 1|$$

```{r depth-fm}
d_fm <- depth.FM(fd)
cat("FM depth range:", round(range(d_fm), 3), "\n")
cat("Magnitude outlier (curve 29):", round(d_fm[n - 1], 4), "\n")
cat("Shape outlier (curve 30):", round(d_fm[n], 4), "\n")
cat("Deepest curve:", which.max(d_fm), "with depth", round(max(d_fm), 4), "\n")
```

### Modified Band Depth (MBD)

MBD measures the proportion of time a curve lies within the band formed by
random pairs of other curves. It captures both magnitude and shape information.

```{r depth-mbd}
d_mbd <- depth.MBD(fd)
cat("MBD depth - outlier 29:", round(d_mbd[n - 1], 4),
    "  outlier 30:", round(d_mbd[n], 4), "\n")
```

### Band Depth (BD)

BD counts the fraction of random bands (formed by pairs) that fully contain
the curve. It is binary at each time point --- a curve is either inside or
outside the band.

```{r depth-bd}
d_bd <- depth.BD(fd)
cat("BD depth - outlier 29:", round(d_bd[n - 1], 4),
    "  outlier 30:", round(d_bd[n], 4), "\n")
```

### Modified Epigraph Index (MEI)

MEI measures the proportion of time each curve lies below (or above) other
curves in the sample.

```{r depth-mei}
d_mei <- depth.MEI(fd)
cat("MEI depth - outlier 29:", round(d_mei[n - 1], 4),
    "  outlier 30:", round(d_mei[n], 4), "\n")
```

## Projection-Based Methods

### Random Projection Depth (RP)

RP projects functional data to random directions and computes halfspace depth
in the projected space. This method works for both 1D and 2D functional data.

```{r depth-rp}
d_rp <- depth.RP(fd, nproj = 50, seed = 123)
cat("RP depth - outlier 29:", round(d_rp[n - 1], 4),
    "  outlier 30:", round(d_rp[n], 4), "\n")
```

### Random Tukey Depth (RT)

RT uses Tukey's halfspace depth on random projections. It tends to produce
coarser depth rankings but is computationally fast.

```{r depth-rt}
d_rt <- depth.RT(fd, nproj = 50, seed = 123)
cat("RT depth - outlier 29:", round(d_rt[n - 1], 4),
    "  outlier 30:", round(d_rt[n], 4), "\n")
```

### Random Projection with Derivatives (RPD)

RPD extends RP by including derivatives. This makes it sensitive to shape
differences even when curves have the same magnitude.

```{r depth-rpd}
d_rpd <- depth.RPD(fd, nproj = 20, deriv = c(0, 1))
cat("RPD depth - outlier 29:", round(d_rpd[n - 1], 4),
    "  outlier 30:", round(d_rpd[n], 4), "\n")
```

## Kernel and Spatial Methods

### Modal Depth

Modal depth is based on kernel density estimation in function space. Curves
in high-density regions receive high depth.

```{r depth-mode}
d_mode <- depth.mode(fd)
cat("Mode depth - outlier 29:", round(d_mode[n - 1], 4),
    "  outlier 30:", round(d_mode[n], 4), "\n")
```

### Functional Spatial Depth (FSD)

FSD is based on spatial signs (unit vectors pointing from the curve to all
other curves). It is affine-invariant and robust.

```{r depth-fsd}
d_fsd <- depth.FSD(fd)
cat("FSD depth - outlier 29:", round(d_fsd[n - 1], 4),
    "  outlier 30:", round(d_fsd[n], 4), "\n")
```

### Kernel Functional Spatial Depth (KFSD)

KFSD smooths the spatial depth using a kernel, improving performance in
finite samples.

```{r depth-kfsd}
d_kfsd <- depth.KFSD(fd)
cat("KFSD depth - outlier 29:", round(d_kfsd[n - 1], 4),
    "  outlier 30:", round(d_kfsd[n], 4), "\n")
```

## Comparing Methods

```{r compare-all, fig.height = 7}
df_depths <- data.frame(
  curve = 1:n,
  FM = d_fm,
  MBD = d_mbd,
  BD = d_bd,
  MEI = d_mei,
  RP = d_rp,
  RT = d_rt,
  mode = d_mode,
  FSD = d_fsd,
  KFSD = d_kfsd,
  RPD = d_rpd,
  type = c(rep("Normal", n - 2), "Magnitude", "Shape")
)

# Reshape for plotting
df_long <- do.call(rbind, lapply(
  c("FM", "MBD", "BD", "MEI", "RP", "RT", "mode", "FSD", "KFSD", "RPD"),
  function(method) {
    data.frame(
      curve = df_depths$curve,
      depth = df_depths[[method]],
      method = method,
      type = df_depths$type
    )
  }
))

ggplot(df_long, aes(x = curve, y = depth, fill = type)) +
  geom_col(alpha = 0.8, width = 0.8) +
  facet_wrap(~ method, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Normal" = "#4A90D9", "Magnitude" = "#D55E00",
                                "Shape" = "#E6A020")) +
  labs(x = "Curve index", y = "Depth", fill = "Type",
       title = "All 10 Depth Methods: Outliers Get Low Depth") +
  theme(legend.position = "bottom")
```

All methods assign low depth to both outliers, but they differ in sensitivity:

- **FM, MBD, BD, MEI** are pointwise --- they detect magnitude outliers well but
  may miss subtle shape outliers
- **RP, RT, RPD** use random projections --- RPD is particularly good at shape
  outliers because it includes derivatives
- **mode, FSD, KFSD** use density/spatial approaches --- they are robust to the
  choice of metric

## Depth-Based Statistics

### Functional Median and Trimmed Mean

```{r median-trimmed}
# Functional median (deepest curve)
fd_median <- median(fd, method = "FM")

# Trimmed mean (exclude 20% shallowest)
fd_trim <- trimmed(fd, trim = 0.2, method = "FM")

# Ordinary mean (affected by outliers)
fd_mean <- mean(fd)

df_robust <- data.frame(
  t = rep(argvals, 3),
  value = c(fd_mean$data[1, ], fd_median$data[1, ], fd_trim$data[1, ]),
  type = factor(rep(c("Mean", "Median", "Trimmed Mean (20%)"), each = m),
                levels = c("Mean", "Median", "Trimmed Mean (20%)"))
)

ggplot(df_robust, aes(x = t, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Mean" = "#D55E00", "Median" = "#2E8B57",
                                 "Trimmed Mean (20%)" = "#4A90D9")) +
  labs(x = "t", y = "Value", color = NULL,
       title = "Mean vs Robust Alternatives",
       subtitle = "Mean is pulled up by the magnitude outlier") +
  theme(legend.position = "bottom")
```

### Functional Boxplot

```{r boxplot}
boxplot(fd, main = "Functional Boxplot (FM Depth)")
```

## Depth Against a Reference Sample

All depth methods accept a separate reference sample via `fdataori`:

```{r reference-depth}
# Split into reference (first 20) and query (last 10)
fd_ref <- fd[1:20, ]
fd_query <- fd[21:n, ]

# Compute depth of query curves against reference
d_query <- depth.FM(fd_query, fdataori = fd_ref)

cat("Depth of query curves against reference:\n")
for (i in seq_along(d_query)) {
  label <- if (20 + i == n - 1) " (magnitude outlier)"
           else if (20 + i == n) " (shape outlier)"
           else ""
  cat(sprintf("  Curve %d: %.4f%s\n", 20 + i, d_query[i], label))
}
```

## Method Selection Guide

| Use case | Recommended method | Why |
|----------|-------------------|-----|
| General purpose | **FM** or **MBD** | Well-established, interpretable |
| Shape-sensitive | **RPD** | Includes derivatives |
| 2D surfaces | **FM**, **RP**, **FSD** | Support 2D data |
| Functional boxplot | **MBD** | Default in `boxplot.fdata()` |
| Large reference sets | **streaming** | O(T log N) per query |
| Outlier detection | **MBD** + outliergram | Complementary views |
| Robust estimation | **FM** | Used by `median()`, `trimmed()` |

## See Also

- `vignette("articles/outlier-detection")` --- depth-based outlier detection methods
- `vignette("articles/streaming-depth")` --- streaming depth for online monitoring
- `vignette("articles/tolerance-bands")` --- depth-based tolerance bands and central regions

## References

- Fraiman, R. and Muniz, G. (2001). Trimmed means for functional data. *Test*, 10(2), 419--440.
- Lopez-Pintado, S. and Romo, J. (2009). On the concept of depth for functional data. *JASA*, 104(486), 718--734.
- Cuevas, A., Febrero, M., and Fraiman, R. (2007). Robust estimation and classification for functional data via projection-based depth notions. *Computational Statistics*, 22, 481--496.
- Chakraborty, A. and Chaudhuri, P. (2014). The spatial distribution in infinite dimensional spaces and related quantiles and depths. *Annals of Statistics*, 42(3), 1203--1231.
