---
title: "Distance Metrics and Semimetrics"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Distance measures are fundamental to many FDA methods including clustering,
k-nearest neighbors regression, and outlier detection. **fdars** provides
both true metrics (satisfying the triangle inequality) and semimetrics
(which may not).

Choosing the right distance measure depends on what kind of "similarity" matters
for your application. Some distances compare curves point-by-point (L2), others
allow for timing shifts before comparing (DTW, Soft-DTW, elastic), and still
others focus on specific aspects like curve shape (derivative-based) or frequency
content (Fourier-based).

A **metric** $d$ on a space $\mathcal{X}$ satisfies:

1. $d(f, g) \geq 0$ (non-negativity)
2. $d(f, g) = 0 \Leftrightarrow f = g$ (identity of indiscernibles)
3. $d(f, g) = d(g, f)$ (symmetry)
4. $d(f, h) \leq d(f, g) + d(g, h)$ (triangle inequality)

A **semimetric** satisfies only conditions 1-3.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())

# Create example data with different curve types
set.seed(42)
m <- 100
t_grid <- seq(0, 1, length.out = m)

# Four types of curves
curve1 <- sin(2 * pi * t_grid)                    # Sine
curve2 <- sin(2 * pi * t_grid + 0.5)              # Phase-shifted sine
curve3 <- cos(2 * pi * t_grid)                    # Cosine
curve4 <- sin(4 * pi * t_grid)                    # Higher frequency

X <- rbind(curve1, curve2, curve3, curve4)
fd <- fdata(X, argvals = t_grid,
            names = list(main = "Four Curve Types"))
plot(fd)
```

## True Metrics

### Lp Distance (metric.lp)

The $L^p$ distance is the most common choice for functional data. It computes
the integrated $L^p$ norm of the difference between two functions:

$$d_p(f, g) = \left( \int_{\mathcal{T}} |f(t) - g(t)|^p \, dt \right)^{1/p}$$

For discrete observations, this is approximated using numerical integration
(Simpson's rule):

$$d_p(f, g) \approx \left( \sum_{j=1}^{m} w_j |f(t_j) - g(t_j)|^p \right)^{1/p}$$

where $w_j$ are quadrature weights.

Special cases:

- **$p = 2$ (L2/Euclidean)**: Most common, corresponds to the standard
  functional norm. Sensitive to vertical differences.
- **$p = 1$ (L1/Manhattan)**: More robust to outliers than L2.
- **$p = \infty$ (L-infinity/Chebyshev)**: Maximum absolute difference,
  $d_\infty(f, g) = \max_t |f(t) - g(t)|$.

```{r metric-lp}
# L2 (Euclidean) distance - default
dist_l2 <- metric.lp(fd)
print(round(as.matrix(dist_l2), 3))

# L1 (Manhattan) distance
dist_l1 <- metric.lp(fd, lp = 1)

# L-infinity (maximum) distance
dist_linf <- metric.lp(fd, lp = Inf)
```

### Weighted Lp Distance

Apply different weights to different parts of the domain:

$$d_{p,w}(f, g) = \left( \int_{\mathcal{T}} w(t) |f(t) - g(t)|^p \, dt \right)^{1/p}$$

This is useful when certain parts of the domain are more important than others.

```{r weighted-lp}
# Weight emphasizing the middle of the domain
w <- dnorm(t_grid, mean = 0.5, sd = 0.2)
w <- w / sum(w) * length(w)  # Normalize

dist_weighted <- metric.lp(fd, lp = 2, w = w)
```

### Hausdorff Distance (metric.hausdorff)

The Hausdorff distance treats curves as sets of points $(t, f(t))$ in 2D space
and computes the maximum of minimum distances:

$$d_H(f, g) = \max\left\{ \sup_{t \in \mathcal{T}} \inf_{s \in \mathcal{T}} \|P_f(t) - P_g(s)\|, \sup_{s \in \mathcal{T}} \inf_{t \in \mathcal{T}} \|P_f(t) - P_g(s)\| \right\}$$

where $P_f(t) = (t, f(t))$ is the point on the graph of $f$ at time $t$.

The Hausdorff distance is useful when:

- Curves may cross each other
- The timing of features is less important than their shape
- Comparing curves with different supports

```{r hausdorff}
dist_haus <- metric.hausdorff(fd)
print(round(as.matrix(dist_haus), 3))
```

### Dynamic Time Warping (metric.DTW)

DTW allows for non-linear alignment of curves before computing the distance.
It finds the optimal warping path $\pi$ that minimizes:

$$d_{DTW}(f, g) = \min_{\pi} \sqrt{\sum_{(i,j) \in \pi} |f(t_i) - g(t_j)|^2}$$

subject to boundary conditions, monotonicity, and continuity constraints.

The **Sakoe-Chiba band** constraint limits the warping to $|i - j| \leq w$,
preventing excessive distortion:

$$d_{DTW}^{SC}(f, g) = \min_{\pi: |i-j| \leq w} \sqrt{\sum_{(i,j) \in \pi} |f(t_i) - g(t_j)|^2}$$

DTW is particularly effective for:

- Phase-shifted signals
- Time series with varying speed
- Comparing signals with local time distortions

```{r dtw}
dist_dtw <- metric.DTW(fd)
print(round(as.matrix(dist_dtw), 3))
```

Notice that DTW gives a smaller distance between the original sine and
phase-shifted sine compared to L2, because it can align them:

```{r dtw-comparison}
# Compare L2 vs DTW for phase-shifted curves
cat("L2 distance (sine vs phase-shifted):", round(as.matrix(dist_l2)[1, 2], 3), "\n")
cat("DTW distance (sine vs phase-shifted):", round(as.matrix(dist_dtw)[1, 2], 3), "\n")
```

DTW with band constraint:

```{r dtw-band}
# Restrict warping to 10% of the series length
dist_dtw_band <- metric.DTW(fd, w = round(m * 0.1))
```

### Soft-DTW (metric.softDTW)

**Soft-DTW** (Cuturi & Blondel, 2017) is a differentiable relaxation of DTW
that replaces the hard minimum with a soft minimum controlled by a smoothing
parameter $\gamma > 0$:

$$\text{sdtw}_\gamma(f, g) = \min_\gamma^{\text{soft}} \sum_{(i,j) \in \pi} |f(t_i) - g(t_j)|^2$$

where $\min^\text{soft}$ is the log-sum-exp soft minimum:
$\min^\text{soft}(a_1, \ldots, a_n) = -\gamma \log \sum_k \exp(-a_k / \gamma)$.

As $\gamma \to 0$, Soft-DTW converges to standard DTW. As $\gamma \to \infty$,
all warping paths receive equal weight. **Soft-DTW is not a metric** --
it can assign negative values to self-distances. The **Soft-DTW divergence**
corrects this:

$$D_\gamma(f, g) = \text{sdtw}_\gamma(f, g) - \frac{1}{2}\left[ \text{sdtw}_\gamma(f, f) + \text{sdtw}_\gamma(g, g) \right]$$

The divergence is non-negative, zero iff $f = g$, and symmetric.

```{r softdtw}
# Soft-DTW distance matrix (raw)
dist_sdtw <- metric.softDTW(fd, gamma = 1.0)
cat("Soft-DTW distance matrix:\n")
print(round(as.matrix(dist_sdtw), 3))

# Soft-DTW divergence (non-negative, symmetric)
dist_sdtw_div <- metric.softDTW(fd, gamma = 1.0, divergence = TRUE)
cat("\nSoft-DTW divergence matrix:\n")
print(round(as.matrix(dist_sdtw_div), 3))
```

#### Effect of the Smoothing Parameter $\gamma$

Small $\gamma$ approximates hard DTW (sharper, less smooth gradients); large
$\gamma$ averages over more warping paths (smoother, less discriminative):

```{r softdtw-gamma, fig.height = 4}
gammas <- c(0.01, 0.1, 1.0, 10.0)
sdtw_12 <- sapply(gammas, function(g) {
  as.matrix(metric.softDTW(fd, gamma = g, divergence = TRUE))[1, 2]
})

df_gamma <- data.frame(gamma = gammas, divergence = sdtw_12)
ggplot(df_gamma, aes(x = gamma, y = divergence)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(title = "Soft-DTW Divergence vs Smoothing Parameter",
       subtitle = "Distance between sine and phase-shifted sine",
       x = expression(gamma ~ "(log scale)"), y = "Divergence")
```

#### Soft-DTW Barycenter

Soft-DTW provides a differentiable loss, making it suitable for computing
**barycenters** (weighted averages under DTW alignment) via gradient descent:

```{r softdtw-barycenter, fig.height = 4}
# Create shifted bumps for barycenter computation
set.seed(42)
n_bary <- 10
X_bary <- matrix(0, n_bary, m)
for (i in 1:n_bary) {
  shift <- runif(1, -0.1, 0.1)
  X_bary[i, ] <- exp(-80 * (t_grid - 0.5 - shift)^2)
}
fd_bary <- fdata(X_bary, argvals = t_grid)

bc <- softdtw.barycenter(fd_bary, gamma = 1.0)
cross_mean <- colMeans(fd_bary$data)

df_bc <- data.frame(
  t = rep(t_grid, 2),
  value = c(as.numeric(bc$barycenter$data), cross_mean),
  type = rep(c("Soft-DTW Barycenter", "Cross-sectional Mean"), each = m)
)

ggplot() +
  geom_line(data = data.frame(
    t = rep(t_grid, n_bary),
    value = as.vector(t(fd_bary$data)),
    curve = rep(1:n_bary, each = m)
  ), aes(x = t, y = value, group = curve), alpha = 0.25) +
  geom_line(data = df_bc, aes(x = t, y = value, color = type), linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional Mean" = "steelblue",
                                 "Soft-DTW Barycenter" = "firebrick")) +
  labs(title = "Soft-DTW Barycenter vs Cross-Sectional Mean",
       x = "t", y = "f(t)", color = NULL)
```

The cross-sectional mean is attenuated by the phase shifts; the Soft-DTW
barycenter recovers the peak shape by accounting for alignment.

### Kullback-Leibler Divergence (metric.kl)

The symmetric Kullback-Leibler divergence treats normalized curves as
probability density functions:

$$d_{KL}(f, g) = \frac{1}{2}\left[ \int f(t) \log\frac{f(t)}{g(t)} \, dt + \int g(t) \log\frac{g(t)}{f(t)} \, dt \right]$$

Before computing, curves are shifted to be non-negative and normalized to
integrate to 1:

$$\tilde{f}(t) = \frac{f(t) - \min_s f(s) + \epsilon}{\int [f(s) - \min_s f(s) + \epsilon] \, ds}$$

This metric is useful for:

- Comparing density-like functions
- Distribution comparison
- Information-theoretic analysis

```{r metric-kl}
# Shift curves to be positive for KL
X_pos <- X - min(X) + 0.1
fd_pos <- fdata(X_pos, argvals = t_grid)

dist_kl <- metric.kl(fd_pos)
print(round(as.matrix(dist_kl), 3))
```

## Semimetrics

Semimetrics may not satisfy the triangle inequality but can be more appropriate
for certain applications or provide computational advantages.

### PCA-Based Semimetric (semimetric.pca)

Projects curves onto the first $q$ principal components and computes the
Euclidean distance in the reduced space:

$$d_{PCA}(f, g) = \sqrt{\sum_{k=1}^{q} (\xi_k^f - \xi_k^g)^2}$$

where $\xi_k^f = \langle f - \bar{f}, \phi_k \rangle$ is the $k$-th principal
component score of $f$, and $\phi_k$ are the eigenfunctions from functional PCA.

This semimetric is useful for:

- Dimension reduction before distance computation
- Noise reduction (low-rank approximation)
- Computational efficiency with many evaluation points

```{r semimetric-pca}
# Distance using first 3 PCs
dist_pca <- semimetric.pca(fd, ncomp = 3)
print(round(as.matrix(dist_pca), 3))
```

### Derivative-Based Semimetric (semimetric.deriv)

Computes the $L^p$ distance based on the $r$-th derivative of the curves:

$$d_{deriv}^{(r)}(f, g) = \left( \int_{\mathcal{T}} |f^{(r)}(t) - g^{(r)}(t)|^p \, dt \right)^{1/p}$$

where $f^{(r)}$ denotes the $r$-th derivative of $f$.

This semimetric focuses on the shape and dynamics of curves rather than their
absolute values. It is particularly useful when:

- Comparing rate of change (first derivative)
- Analyzing curvature (second derivative)
- Functions are measured with different baselines

```{r semimetric-deriv}
# First derivative (velocity)
dist_deriv1 <- semimetric.deriv(fd, nderiv = 1)

# Second derivative (acceleration/curvature)
dist_deriv2 <- semimetric.deriv(fd, nderiv = 2)

cat("First derivative distances:\n")
print(round(as.matrix(dist_deriv1), 3))
```

### Basis Semimetric (semimetric.basis)

Projects curves onto a basis (B-spline or Fourier) and computes the Euclidean
distance between the coefficient vectors:

$$d_{basis}(f, g) = \|c^f - c^g\|_2 = \sqrt{\sum_{k=1}^{K} (c_k^f - c_k^g)^2}$$

where $c^f = (c_1^f, \ldots, c_K^f)$ are the basis coefficients from
$f(t) \approx \sum_{k=1}^{K} c_k^f B_k(t)$.

For **B-splines**: Local support provides good approximation of local features.

For **Fourier basis**: Global support captures periodic patterns efficiently.

```{r semimetric-basis}
# B-spline basis (local features)
dist_bspline <- semimetric.basis(fd, nbasis = 15, basis = "bspline")

# Fourier basis (periodic patterns)
dist_fourier <- semimetric.basis(fd, nbasis = 11, basis = "fourier")

cat("B-spline basis distances:\n")
print(round(as.matrix(dist_bspline), 3))
```

### Fourier Semimetric (semimetric.fourier)

Uses the Fast Fourier Transform (FFT) to compute Fourier coefficients and
measures distance based on the first $K$ frequency components:

$$d_{FFT}(f, g) = \sqrt{\sum_{k=0}^{K} |\hat{f}_k - \hat{g}_k|^2}$$

where $\hat{f}_k$ is the $k$-th Fourier coefficient computed via FFT.

This is computationally efficient for large datasets and particularly useful
for periodic or frequency-domain analysis.

```{r semimetric-fourier}
dist_fft <- semimetric.fourier(fd, nfreq = 5)
cat("Fourier (FFT) distances:\n")
print(round(as.matrix(dist_fft), 3))
```

### Horizontal Shift Semimetric (semimetric.hshift)

Finds the optimal horizontal shift before computing the $L^2$ distance:

$$d_{hshift}(f, g) = \min_{|h| \leq h_{max}} \left( \int_{\mathcal{T}} |f(t) - g(t+h)|^2 \, dt \right)^{1/2}$$

where $h$ is the shift in discrete time units and $h_{max}$ is the maximum
allowed shift.

This semimetric is simpler than DTW (only horizontal shifts, no warping) but
can be very effective for:

- Phase-shifted periodic signals
- ECG or other physiological signals with timing variations
- Comparing curves where horizontal alignment is meaningful

```{r semimetric-hshift}
dist_hshift <- semimetric.hshift(fd)
cat("Horizontal shift distances:\n")
print(round(as.matrix(dist_hshift), 3))
```

## Elastic Distances

Elastic alignment provides a family of distances based on the Fisher-Rao metric
and the SRSF transform. These distances decompose curve differences into
amplitude (shape) and phase (timing) components. See
`vignette("elastic-alignment", package = "fdars")` for the full framework.

### Elastic Distance (metric.elastic / elastic.distance)

The elastic distance computes the $L^2$ distance between optimally aligned
SRSFs:

$$d_e(f, g) = \min_{\gamma \in \Gamma} \| q_f - (q_g \circ \gamma) \sqrt{\dot\gamma} \|_{L^2}$$

where $q_f, q_g$ are the SRSFs of $f, g$ and $\Gamma$ is the warping group.
This is a proper metric that is invariant to reparameterization.

```{r elastic-dist}
dist_elastic <- elastic.distance(fd)
cat("Elastic distance matrix:\n")
print(round(dist_elastic, 3))
```

### Amplitude Distance (amplitude.distance)

The **amplitude distance** measures the shape difference between curves *after*
optimal alignment -- it captures how much the curves differ in height, depth,
and feature magnitude:

$$d_a(f_1, f_2) = \| q_1 - (q_2 \circ \gamma^*) \sqrt{\dot\gamma^*} \|_{L^2} - \lambda \int_0^1 |\dot\gamma^*(t) - 1| \, dt$$

where $\lambda$ controls the penalty on warping complexity.

```{r amplitude-dist}
dist_amp <- amplitude.distance(fd, lambda = 0)
cat("Amplitude distance matrix:\n")
print(round(as.matrix(dist_amp), 3))
```

### Phase Distance (phase.distance)

The **phase distance** measures only the timing difference between curves --
how much one curve's time axis must be warped to match another:

$$d_p(f_1, f_2) = \arccos\left( \int_0^1 \sqrt{\dot\gamma^*(t)} \, dt \right)$$

This is the geodesic distance of the warping function from the identity on the
warping group $\Gamma$, which forms a Riemannian manifold.

```{r phase-dist}
dist_phase <- phase.distance(fd, lambda = 0)
cat("Phase distance matrix:\n")
print(round(as.matrix(dist_phase), 3))
```

### Comparing Amplitude and Phase Distances

The amplitude and phase distances capture orthogonal aspects of curve
variability. Curves that differ mainly in timing (phase-shifted copies of the
same shape) will have small amplitude distance but large phase distance:

```{r elastic-compare, fig.height = 4}
amp_vals <- as.vector(as.matrix(dist_amp))[upper.tri(as.matrix(dist_amp))]
phase_vals <- as.vector(as.matrix(dist_phase))[upper.tri(as.matrix(dist_phase))]

plot(amp_vals, phase_vals, pch = 19, col = "steelblue",
     xlab = "Amplitude Distance", ylab = "Phase Distance",
     main = "Amplitude vs Phase Pairwise Distances")
abline(h = mean(phase_vals), lty = 2, col = "grey50")
abline(v = mean(amp_vals), lty = 2, col = "grey50")
```

## Unified Interface

Use `metric()` for a unified interface to all distance functions:

```{r metric-unified}
# Different methods via single function
d1 <- metric(fd, method = "lp", lp = 2)
d2 <- metric(fd, method = "dtw")
d3 <- metric(fd, method = "hausdorff")
d4 <- metric(fd, method = "pca", ncomp = 2)
d5 <- metric(fd, method = "softdtw", gamma = 1.0)
d6 <- metric(fd, method = "elastic")
d7 <- metric(fd, method = "amplitude")
d8 <- metric(fd, method = "phase")
```

## Comparing Distance Measures

Different distance measures capture different aspects of curve similarity:

```{r compare}
# Create comparison data
dists <- list(
  L2 = as.vector(as.matrix(dist_l2)),
  L1 = as.vector(as.matrix(dist_l1)),
  DTW = as.vector(as.matrix(dist_dtw)),
  Hausdorff = as.vector(as.matrix(dist_haus))
)

# Correlation between distance measures
dist_mat <- do.call(cbind, dists)
cat("Correlation between distance measures:\n")
print(round(cor(dist_mat), 2))
```

## Distance Matrices for Larger Samples

```{r large-sample}
# Generate larger sample
set.seed(123)
n <- 50
X_large <- matrix(0, n, m)
for (i in 1:n) {
  phase <- runif(1, 0, 2*pi)
  freq <- sample(1:3, 1)
  X_large[i, ] <- sin(freq * pi * t_grid + phase) + rnorm(m, sd = 0.1)
}
fd_large <- fdata(X_large, argvals = t_grid)

# Compute distance matrix
dist_matrix <- metric.lp(fd_large)

# Visualize as heatmap using ggplot2
dist_df <- expand.grid(Curve1 = 1:n, Curve2 = 1:n)
dist_df$Distance <- as.vector(as.matrix(dist_matrix))

ggplot(dist_df, aes(x = Curve1, y = Curve2, fill = Distance)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma") +
  coord_equal() +
  labs(x = "Curve", y = "Curve", title = "L2 Distance Matrix") +
  theme_minimal()
```

## Metric Properties Summary

| Distance | Type | Triangle Ineq. | Phase Invariant | Computational Cost |
|----------|------|----------------|-----------------|-------------------|
| Lp | Metric | Yes | No | O(nm) |
| Hausdorff | Metric | Yes | Partial | O(nm²) |
| DTW | Metric* | Yes* | Yes | O(nm²) |
| Soft-DTW | Semimetric | No | Yes | O(nm²) |
| Soft-DTW Div. | Semimetric | No*** | Yes | O(nm²) |
| Elastic | Metric | Yes | Yes | O(nm²) |
| Amplitude | Semimetric | No | Yes | O(nm²) |
| Phase | Metric | Yes | Yes | O(nm²) |
| KL | Semimetric | No | No | O(nm) |
| PCA | Semimetric | Yes** | No | O(nm + nq²) |
| Derivative | Semimetric | Yes** | No | O(nm) |
| Basis | Semimetric | Yes** | No | O(nmK) |
| Fourier | Semimetric | Yes** | No | O(nm log m) |
| H-shift | Semimetric | No | Yes | O(nm·h_max) |

\* DTW satisfies triangle inequality when using the same warping path
\** These satisfy triangle inequality in the reduced/transformed space
\*** Soft-DTW divergence is non-negative and symmetric but does not satisfy triangle inequality

## Choosing a Distance Measure

| Scenario | Recommended Distance |
|----------|---------------------|
| General purpose | L2 (`metric.lp`) |
| Heavy-tailed noise | L1 (`metric.lp`, p=1) |
| Worst-case analysis | L-infinity (`metric.lp`, p=Inf) |
| Phase-shifted signals | DTW, H-shift, or Elastic |
| Differentiable DTW loss | Soft-DTW (`metric.softDTW`) |
| Reparameterization-invariant | Elastic (`elastic.distance`) |
| Separate shape vs timing | Amplitude + Phase distances |
| Curves that cross | Hausdorff |
| High-dimensional data | PCA-based |
| Shape comparison | Derivative-based |
| Periodic data | Fourier-based |
| Density-like curves | Kullback-Leibler |

## Performance

Distance computations are parallelized in Rust:

```{r performance, eval=FALSE}
# Benchmark for 500 curves
X_bench <- matrix(rnorm(500 * 200), 500, 200)
fd_bench <- fdata(X_bench)

system.time(metric.lp(fd_bench))
#>    user  system elapsed
#>   0.089   0.000   0.045

system.time(metric.DTW(fd_bench))
#>    user  system elapsed
#>   1.234   0.000   0.312
```

## Using Distances in Other Methods

Distance functions can be passed to clustering and regression:

```{r use-in-methods}
# K-means with L1 distance
km_l1 <- cluster.kmeans(fd_large, ncl = 3, metric = "L1", seed = 42)

# K-means with custom metric function
km_dtw <- cluster.kmeans(fd_large, ncl = 3, metric = metric.DTW, seed = 42)

# Nonparametric regression uses distances internally
y <- rnorm(n)
fit_np <- fregre.np(fd_large, y, metric = metric.lp)
```

## See Also

- `vignette("elastic-alignment", package = "fdars")` -- elastic alignment
  framework, Karcher mean, amplitude-phase decomposition
- `vignette("alignment-comparison", package = "fdars")` -- comparing alignment
  methods side-by-side

## References

- Berndt, D.J. and Clifford, J. (1994). Using Dynamic Time Warping to Find
  Patterns in Time Series. *KDD Workshop*, 359-370.
- Cuturi, M. and Blondel, M. (2017). Soft-DTW: a Differentiable Loss Function
  for Time-Series. *Proceedings of the 34th International Conference on Machine
  Learning*, 894-903.
- Ferraty, F. and Vieu, P. (2006). *Nonparametric Functional Data Analysis*.
  Springer.
- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  Springer, 2nd edition.
- Sakoe, H. and Chiba, S. (1978). Dynamic programming algorithm optimization
  for spoken word recognition. *IEEE Transactions on Acoustics, Speech, and
  Signal Processing*, 26(1), 43-49.
- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data Analysis*.
  Springer.
