---
title: "Elastic Curve Alignment"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

When comparing functional data, observed curves often exhibit both **amplitude
variability** (differences in height/shape) and **phase variability** (differences
in timing/speed). Standard methods like the cross-sectional mean or PCA treat all
variation as amplitude, which can produce blurred averages and inflated variance.

**Elastic alignment** separates these two sources of variability by finding optimal
**warping functions** $\gamma$ that retime each curve before comparison. The
mathematical framework uses the **Square-Root Slope Function (SRSF)** representation,
which makes the Fisher-Rao metric equivalent to the $L^2$ metric -- enabling efficient
dynamic programming alignment.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## Quick Start

```{r quick-start}
set.seed(42)
argvals <- seq(0, 1, length.out = 100)

# Simulate curves with horizontal shifts
n <- 15
shifts <- runif(n, -0.1, 0.1)
data <- matrix(0, n, 100)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * (argvals - shifts[i]))
}
fd <- fdata(data, argvals = argvals)

# Align
res <- elastic.align(fd)
print(res)
```

```{r quick-start-plot}
plot(res, type = "both")
```

## SRSF Transform

The SRSF of a function $f$ is defined as:

$$q(t) = \text{sign}(f'(t)) \sqrt{|f'(t)|}$$

This representation has the key property that the Fisher-Rao distance between two
curves equals the $L^2$ distance between their SRSFs after optimal alignment.

```{r srsf-demo}
q <- srsf.transform(fd)

p1 <- plot(fd) + labs(title = "Original Curves")
p2 <- plot(q) + labs(title = "SRSF Representation")
```

The SRSF transform is invertible: given $q$ and an initial value $f_0$, we can
reconstruct:

$$f(t) = f_0 + \int_0^t q(s)|q(s)|\,ds$$

```{r srsf-inverse}
f_original <- fd$data[1, ]
q_values <- q$data[1, ]
f_reconstructed <- srsf.inverse(q_values, argvals, f_original[1])

max_error <- max(abs(f_original - f_reconstructed))
cat("Round-trip max error:", format(max_error, digits = 4), "\n")
```

## Alignment in Detail

### How It Works

`elastic.align()` aligns all curves to a common target using dynamic programming
to find the optimal warping function $\gamma$ for each curve. The warping functions
satisfy $\gamma(0) = 0$ and $\gamma(1) = 1$ (boundary-preserving) and are monotone
increasing (time never runs backward).

By default, the cross-sectional mean is used as the target:

```{r align-detail}
res <- elastic.align(fd)
```

You can also supply a custom target -- as a numeric vector or an `fdata` object:

```{r align-target}
# Use the deepest (most central) curve as target
depths <- depth(fd, method = "FM")
deepest <- fd[which.max(depths)]
res_deep <- elastic.align(fd, target = deepest)
```

### Inspecting Warping Functions

The warping functions show how each curve's time axis is stretched or compressed
to match the target:

```{r warps}
plot(res, type = "warps")
```

A warping function above the diagonal means that part of the curve has been
sped up (compressed in time); below the diagonal means slowed down (stretched).

## Karcher Mean

The **Karcher mean** (Frechet mean in the elastic metric) computes the mean shape
by iteratively aligning curves to a running estimate of the mean. This produces a
sharper average than the cross-sectional mean because it removes phase variability:

```{r karcher}
km <- karcher.mean(fd, max.iter = 20, tol = 1e-4)
print(km)
```

```{r karcher-compare, fig.height = 4}
cross_mean <- colMeans(fd$data)
karcher <- as.numeric(km$mean$data)

df_means <- data.frame(
  t = rep(argvals, 2),
  value = c(cross_mean, karcher),
  type = rep(c("Cross-sectional mean", "Karcher mean"), each = 100)
)

ggplot(df_means, aes(x = t, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Cross-Sectional vs Karcher Mean",
       x = "t", y = "f(t)", color = NULL)
```

The cross-sectional mean is attenuated by the phase shifts, while the Karcher mean
recovers the true amplitude.

## Elastic Distance Matrix

The elastic distance provides a reparameterization-invariant metric for
clustering, MDS, and other distance-based methods:

```{r distance}
D <- elastic.distance(fd)

# Metric properties
cat("Symmetric:             ", max(abs(D - t(D))) < 1e-10, "\n")
cat("Non-negative:          ", all(D >= -1e-10), "\n")
cat("Zero diagonal:         ", all(abs(diag(D)) < 1e-10), "\n")
```

```{r mds, fig.height = 4}
mds <- cmdscale(D, k = 2)
plot(mds, pch = 19, xlab = "MDS 1", ylab = "MDS 2",
     main = "MDS of Elastic Distances", col = "steelblue", cex = 1.3)
```

The elastic distance is also available through the `metric()` dispatcher:

```{r metric-elastic}
D2 <- metric(fd, method = "elastic")
all.equal(D, D2)
```

## When to Use Elastic Alignment

Elastic alignment works best when curves share the **same shape** but differ
in **timing**. The table below summarizes where it excels and where it does not.

### Works Well

| Scenario | Typical VR | Notes |
|:---------|:----------:|:------|
| Shifted peaks / bumps | 98-100% | Best case: localized features with horizontal shifts |
| ECG-like sharp peaks | 98-100% | Localized timing differences in multi-peak curves |
| Spectral data (shifted peaks) | 95-98% | Multiple peaks with coherent timing shifts |
| Phase-shifted smooth curves | 90-93% | Sinusoidal, polynomial, sigmoid curves with shifts |
| Non-periodic curves | 95-100% | Works well when boundary values are similar |

### Where It Degrades

| Scenario | Typical VR | Explanation |
|:---------|:----------:|:------------|
| Mixed amplitude + phase | 40-65% | Phase correction is partial when amplitude dominates |
| Noisy data (raw) | 0-30% | Noise in derivatives overwhelms SRSF; smooth first |
| Large boundary mismatch | 70-85% | Fixed boundary ($\gamma(0)=0$, $\gamma(1)=1$) can't correct endpoint values |
| Coarse grids (m < 50) | 70-75% | DP resolution limits alignment accuracy |

### Where It Should NOT Be Used

| Scenario | What happens | Recommendation |
|:---------|:-------------|:---------------|
| Pure amplitude differences | Identity warp (no harm, no help) | Use standard methods |
| Rate-varying monotone curves | Spurious warping, VR < 0 | These are amplitude/shape differences, not phase |
| Genuinely different shapes | Forced warping, VR < 0 | Use depth or distance methods instead |
| Very noisy data (unsmoothed) | Noise-driven warping, VR < 0 | Pre-smooth with `fdata2basis()` + `basis2fdata()` |

## Visual Examples by Scenario

The following examples illustrate alignment behavior across different data
scenarios. Each shows the **original** curves (left) and **aligned** curves
(right) with variance reduction (VR) reported.

### Works Well: ECG-like Multi-Peak Curves

Sharp, localized features with timing differences are the ideal use case:

```{r gallery-ecg, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.05, 0.05)
  t_shifted <- argvals - shift
  # P wave, QRS complex, T wave
  dm[i, ] <- 0.2 * exp(-200 * (t_shifted - 0.2)^2) -
             0.6 * exp(-500 * (t_shifted - 0.4)^2) +
             1.0 * exp(-800 * (t_shifted - 0.43)^2) -
             0.2 * exp(-500 * (t_shifted - 0.46)^2) +
             0.3 * exp(-150 * (t_shifted - 0.65)^2)
}
fd_ecg <- fdata(dm, argvals = argvals)
res_ecg <- elastic.align(fd_ecg)
vr <- 1 - mean(apply(res_ecg$aligned$data, 2, var)) /
          mean(apply(fd_ecg$data, 2, var))
```

```{r gallery-ecg-plot, fig.height = 4}
plot(res_ecg, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Works Well: Spectral Peaks

Multiple peaks at varying positions, common in spectroscopy and chromatography:

```{r gallery-spectral, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.04, 0.04)
  dm[i, ] <- 0.8 * exp(-300 * (argvals - 0.25 - shift)^2) +
             1.0 * exp(-500 * (argvals - 0.50 - shift * 1.2)^2) +
             0.5 * exp(-300 * (argvals - 0.75 - shift * 0.8)^2)
}
fd_spec <- fdata(dm, argvals = argvals)
res_spec <- elastic.align(fd_spec)
vr <- 1 - mean(apply(res_spec$aligned$data, 2, var)) /
          mean(apply(fd_spec$data, 2, var))
```

```{r gallery-spectral-plot, fig.height = 4}
plot(res_spec, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Works Well: Growth-like Sigmoid Curves

Monotone S-shaped curves with timing shifts (e.g., growth curves with
differing onset times):

```{r gallery-sigmoid, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- 1 / (1 + exp(-20 * (argvals - 0.5 - shift)))
}
fd_sig <- fdata(dm, argvals = argvals)
res_sig <- elastic.align(fd_sig)
vr <- 1 - mean(apply(res_sig$aligned$data, 2, var)) /
          mean(apply(fd_sig$data, 2, var))
```

```{r gallery-sigmoid-plot, fig.height = 4}
plot(res_sig, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Medium: Mixed Amplitude and Phase Variability

When curves differ in both height *and* timing, alignment corrects the phase
component but leaves amplitude variability untouched:

```{r gallery-mixed, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 1.5)
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- amp * sin(2 * pi * (argvals - shift))
}
fd_mix <- fdata(dm, argvals = argvals)
res_mix <- elastic.align(fd_mix)
vr <- 1 - mean(apply(res_mix$aligned$data, 2, var)) /
          mean(apply(fd_mix$data, 2, var))
```

```{r gallery-mixed-plot, fig.height = 4}
plot(res_mix, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

After alignment, curves still differ in amplitude (vertical spread) but peaks
and valleys are now synchronized. The residual variance is real amplitude
variability, not a failure of alignment.

### Medium: Noisy Data (Raw vs Smoothed)

Noise in derivatives overwhelms the SRSF. Pre-smoothing recovers alignment:

```{r gallery-noise, fig.height = 7}
set.seed(42)
n <- 20
dm_clean <- matrix(0, n, 100)
shifts <- runif(n, -0.1, 0.1)
for (i in 1:n) dm_clean[i, ] <- sin(2 * pi * (argvals - shifts[i]))
dm_noisy <- dm_clean + matrix(rnorm(n * 100, sd = 0.3), n, 100)
fd_noisy <- fdata(dm_noisy, argvals = argvals)

# Align raw noisy data
res_noisy <- elastic.align(fd_noisy)
vr_raw <- 1 - mean(apply(res_noisy$aligned$data, 2, var)) /
              mean(apply(fd_noisy$data, 2, var))

# Smooth first, then align
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_sm <- basis2fdata(coefs, argvals)
res_sm <- elastic.align(fd_sm)
vr_sm <- 1 - mean(apply(res_sm$aligned$data, 2, var)) /
             mean(apply(fd_sm$data, 2, var))
```

```{r gallery-noise-plot, fig.height = 7}
p_raw <- plot(res_noisy, type = "both") +
  plot_annotation(subtitle = paste0("Raw data — VR = ", round(vr_raw * 100, 1), "%"))
p_sm <- plot(res_sm, type = "both") +
  plot_annotation(subtitle = paste0("After smoothing — VR = ", round(vr_sm * 100, 1), "%"))
p_raw / p_sm
```

### Medium: Boundary Mismatch

When shifted curves have different values at the domain boundaries, alignment
is excellent in the interior but constrained at the edges:

```{r gallery-boundary, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
shifts <- runif(n, -0.15, 0.15)
for (i in 1:n) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd_bnd <- fdata(dm, argvals = argvals)
res_bnd <- elastic.align(fd_bnd)
vr <- 1 - mean(apply(res_bnd$aligned$data, 2, var)) /
          mean(apply(fd_bnd$data, 2, var))
```

```{r gallery-boundary-plot, fig.height = 4}
plot(res_bnd, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — note residual spread at boundaries"))
```

### Bad: Pure Amplitude Differences

When curves differ only in height (no phase variability), alignment correctly
produces identity warps — but also provides no benefit:

```{r gallery-amplitude, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 2.0)
  dm[i, ] <- amp * sin(2 * pi * argvals)
}
fd_amp <- fdata(dm, argvals = argvals)
res_amp <- elastic.align(fd_amp)
vr <- 1 - mean(apply(res_amp$aligned$data, 2, var)) /
          mean(apply(fd_amp$data, 2, var))
```

```{r gallery-amplitude-plot, fig.height = 4}
plot(res_amp, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — no phase to correct, identity warps applied"))
```

This is correct behavior: no spurious warping is introduced.

### Bad: Genuinely Different Shapes

Alignment should **not** be used when curves have fundamentally different
shapes. Forced warping produces meaningless results:

```{r gallery-different, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  if (i %% 5 == 1) dm[i, ] <- sin(2 * pi * argvals)
  else if (i %% 5 == 2) dm[i, ] <- cos(6 * pi * argvals)
  else if (i %% 5 == 3) dm[i, ] <- argvals^3
  else if (i %% 5 == 4) dm[i, ] <- sin(10 * pi * argvals) * exp(-3 * argvals)
  else dm[i, ] <- dnorm(argvals, 0.5, 0.05) / dnorm(0.5, 0.5, 0.05)
}
fd_diff <- fdata(dm, argvals = argvals)
res_diff <- elastic.align(fd_diff)
vr <- 1 - mean(apply(res_diff$aligned$data, 2, var)) /
          mean(apply(fd_diff$data, 2, var))
```

```{r gallery-different-plot, fig.height = 4}
plot(res_diff, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — forced warping of unrelated shapes"))
```

When VR is very low or negative, it indicates the data is not suitable for
elastic alignment. Use depth or distance methods instead.

## Periodic Functional Data

### The Problem

Elastic alignment uses fixed boundary constraints $\gamma(0) = 0$, $\gamma(1) = 1$.
This prevents the warping function from performing circular shifts, which means
periodic data (e.g., curves on $[0, 2\pi]$ where $f(0) = f(2\pi)$) with large
phase offsets cannot be properly aligned.

For example, two copies of $\sin(t)$ shifted by half a period have identical
shape but the boundary constraints force alignment to fail:

```{r periodic-problem, fig.height = 4}
set.seed(42)
m <- 100
argvals_p <- seq(0, 2 * pi, length.out = m)
n <- 15
data_p <- matrix(0, n, m)
for (i in 1:n) {
  shift <- sample(0:(m - 1), 1)
  idx <- ((seq_len(m) - 1L + shift) %% m) + 1L
  data_p[i, ] <- sin(argvals_p[idx])
}
fd_p <- fdata(data_p, argvals = argvals_p)

# Standard alignment struggles with circular shifts
res_std <- elastic.align(fd_p)
var_orig <- mean(apply(fd_p$data, 2, var))
vr_std <- 1 - mean(apply(res_std$aligned$data, 2, var)) / var_orig
```

```{r periodic-problem-plot, fig.height = 4}
plot(res_std, type = "both") +
  plot_annotation(subtitle = paste0(
    "Standard alignment on periodic data — VR = ", round(vr_std * 100, 1), "%"))
```

### The Solution

Setting `periodic = TRUE` applies a two-stage approach:

1. **Circular rotation**: Each curve is circularly shifted so that its global
   maximum is at a canonical grid position
2. **Elastic alignment**: Standard alignment on the rotated data

```{r periodic-solution, fig.height = 4}
res_per <- elastic.align(fd_p, periodic = TRUE)
vr_per <- 1 - mean(apply(res_per$aligned$data, 2, var)) / var_orig
```

```{r periodic-solution-plot, fig.height = 4}
plot(res_per, type = "both") +
  plot_annotation(subtitle = paste0(
    "Periodic alignment — VR = ", round(vr_per * 100, 1), "%"))
```

The rotation step removes large circular shifts that the boundary-constrained
warping cannot handle, and the elastic step corrects any remaining timing
differences.

You can also apply the rotation step manually using `periodic.rotate()`:

```{r periodic-rotate}
rot <- periodic.rotate(fd_p)
cat("Shifts applied:", head(rot$shifts), "...\n")
```

The `karcher.mean()` function also accepts `periodic = TRUE`:

```{r periodic-karcher}
km_per <- karcher.mean(fd_p, max.iter = 15, periodic = TRUE)
print(km_per)
```

## Practical Guide

### Pre-smooth Noisy Data

The SRSF transform involves derivatives, so noise gets amplified. Always smooth
noisy data before alignment:

```{r presmooth}
set.seed(42)
noisy <- matrix(0, 15, 100)
shifts <- runif(15, -0.1, 0.1)
for (i in 1:15) {
  noisy[i, ] <- sin(2 * pi * (argvals - shifts[i])) + rnorm(100, sd = 0.15)
}
fd_noisy <- fdata(noisy, argvals = argvals)

# Smooth with B-spline basis
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_smooth <- basis2fdata(coefs, argvals)

var_before <- mean(apply(fd_noisy$data, 2, var))

# Align raw vs smoothed
res_raw <- elastic.align(fd_noisy)
res_smooth <- elastic.align(fd_smooth)

vr_raw <- 1 - mean(apply(res_raw$aligned$data, 2, var)) / var_before
vr_smooth <- 1 - mean(apply(res_smooth$aligned$data, 2, var)) / var_before
cat("VR (raw):     ", round(vr_raw * 100, 1), "%\n")
cat("VR (smoothed):", round(vr_smooth * 100, 1), "%\n")
```

### Use Sufficient Grid Resolution

The dynamic programming alignment operates on the evaluation grid. Finer grids
yield more accurate warping:

```{r grid-resolution}
set.seed(42)
for (m in c(20, 50, 100, 200)) {
  av <- seq(0, 1, length.out = m)
  shifts <- runif(15, -0.1, 0.1)
  dm <- matrix(0, 15, m)
  for (i in 1:15) dm[i, ] <- sin(2 * pi * (av - shifts[i]))
  fd_m <- fdata(dm, argvals = av)
  ea <- elastic.align(fd_m)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                  mean(apply(fd_m$data, 2, var))
  cat(sprintf("  m = %3d: VR = %5.1f%%\n", m, var_red * 100))
}
```

A minimum of 50-100 grid points is recommended.

### Choose an Appropriate Target

The target curve affects alignment quality. Three options:

| Target | When to use |
|:-------|:------------|
| Cross-sectional mean (default) | General-purpose; robust with many curves |
| Deepest curve | When outliers may distort the mean |
| Known reference | When aligning to a standard template |

```{r target-choice}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)
var_before <- mean(apply(fd$data, 2, var))

# Default (cross-sectional mean)
vr1 <- 1 - mean(apply(elastic.align(fd)$aligned$data, 2, var)) / var_before

# Deepest curve
depths <- depth(fd, method = "FM")
vr2 <- 1 - mean(apply(elastic.align(fd, target = fd[which.max(depths)])$aligned$data, 2, var)) / var_before

cat("Cross-sectional mean: VR =", round(vr1 * 100, 1), "%\n")
cat("Deepest curve:        VR =", round(vr2 * 100, 1), "%\n")
```

### Alignment + FPCA

Alignment before FPCA concentrates variance into the first principal component
by removing phase variability that would otherwise spread across components:

```{r alignment-fpca}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)

# FPCA before and after alignment
pc_before <- fdata2pc(fd, ncomp = 3)
ea <- elastic.align(fd)
pc_after <- fdata2pc(ea$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2], "%\n")
```

## Example: Localized Peak Alignment

Elastic alignment excels at aligning curves with localized features at varying
positions -- the pattern seen in spectroscopy, ECG, and chromatography data:

```{r peaks-example}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  # Two peaks with varying positions
  p1 <- 0.3 + runif(1, -0.05, 0.05)
  p2 <- 0.7 + runif(1, -0.05, 0.05)
  dm[i, ] <- exp(-100 * (argvals - p1)^2) + 0.6 * exp(-100 * (argvals - p2)^2)
}
fd_peaks <- fdata(dm, argvals = argvals)

res_peaks <- elastic.align(fd_peaks)

var_red <- 1 - mean(apply(res_peaks$aligned$data, 2, var)) /
               mean(apply(fd_peaks$data, 2, var))
cat("Variance reduction:", round(var_red * 100, 1), "%\n")
```

```{r peaks-plot}
plot(res_peaks, type = "both")
```

## Example: Shifted Gaussian Bumps

```{r bumps-example}
set.seed(42)
dm <- matrix(0, 20, 100)
shifts <- runif(20, -0.15, 0.15)
for (i in 1:20) {
  dm[i, ] <- exp(-25 * (argvals - 0.5 - shifts[i])^2)
}
fd_bumps <- fdata(dm, argvals = argvals)
res_bumps <- elastic.align(fd_bumps)

var_red <- 1 - mean(apply(res_bumps$aligned$data, 2, var)) /
               mean(apply(fd_bumps$data, 2, var))
cat("Variance reduction:", round(var_red * 100, 1), "%\n")
```

```{r bumps-plot}
plot(res_bumps, type = "both")
```

## Karcher Mean with Custom Settings

For complex data, increase `max.iter` and check the aligned curves:

```{r karcher-custom}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- exp(-25 * (argvals - 0.5 - shift)^2)
}
fd <- fdata(dm, argvals = argvals)

km <- karcher.mean(fd, max.iter = 30, tol = 1e-4)
cat("Converged:", km$converged, "\n")
cat("Iterations:", km$n.iter, "\n")
```

```{r karcher-custom-plot}
plot(km, type = "mean")
```

## Limitations

### Fixed Boundary Conditions

Warping functions are constrained to $\gamma(0) = 0$ and $\gamma(1) = 1$.
This means the alignment cannot correct mismatches at the domain boundaries.
For shifted curves, the residual at the boundary equals the value difference:

```{r boundary}
f1 <- sin(2 * pi * argvals)
f2 <- sin(2 * pi * (argvals - 0.1))

fd2 <- fdata(rbind(f1, f2), argvals = argvals)
ea <- elastic.align(fd2)

# Boundary residual = |sin(0) - sin(-0.2*pi)| = 0.588
cat("Boundary gap:          ", round(abs(f2[1] - f1[1]), 3), "\n")
cat("Max aligned difference:", round(max(abs(ea$aligned$data[1, ] - ea$aligned$data[2, ])), 3), "\n")
cat("Interior max diff:     ", round(max(abs(ea$aligned$data[1, 10:90] - ea$aligned$data[2, 10:90])), 3), "\n")
```

The interior alignment is excellent even when boundaries cannot be corrected.

### Amplitude-Only Differences

When curves differ only in amplitude (e.g., $f$ vs $2f$), alignment correctly
produces identity warping -- no spurious time distortion:

```{r amplitude-only}
f1 <- sin(2 * pi * argvals)
f2 <- 2 * sin(2 * pi * argvals)
fd_amp <- fdata(rbind(f1, f2), argvals = argvals)
res_amp <- elastic.align(fd_amp)

max_warp_dev <- max(abs(res_amp$gammas$data -
                        matrix(argvals, 2, 100, byrow = TRUE)))
cat("Max warp deviation from identity:", max_warp_dev, "\n")
```

This is the correct behavior: amplitude differences should not produce warping.
However, when both amplitude and phase variability are present, the phase correction
may be partial -- alignment focuses on phase and leaves amplitude untouched.

### Domain Independence

The alignment works correctly on any domain, not just $[0, 1]$:

```{r domains}
for (domain in list(c(0, 1), c(0, 2*pi), c(0, 10), c(-1, 1))) {
  av <- seq(domain[1], domain[2], length.out = 100)
  range <- domain[2] - domain[1]
  dm <- matrix(0, 10, 100)
  set.seed(42)
  for (i in 1:10) {
    shift <- runif(1, -0.1 * range, 0.1 * range)
    dm[i, ] <- sin(2 * pi * (av - shift) / range)
  }
  fd <- fdata(dm, argvals = av)
  ea <- elastic.align(fd)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                 mean(apply(fd$data, 2, var))
  cat(sprintf("  [%g, %g]: VR = %.1f%%\n", domain[1], domain[2], var_red * 100))
}
```

## References

- Srivastava, A., Klassen, E., Joshi, S.H., and Jermyn, I.H. (2011). Shape analysis of elastic curves in Euclidean spaces. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 33(7):1415--1428.
- Tucker, J.D., Wu, W., and Srivastava, A. (2013). Generative models for functional data using phase and amplitude separation. *Computational Statistics & Data Analysis*, 61:50--66.
- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data Analysis*. Springer.
