---
title: "Elastic FPCA"
description: >
  Separate amplitude and phase variability using vertical, horizontal,
  and joint FPCA after elastic alignment.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elastic FPCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Motivation

Functional principal component analysis (FPCA) is a powerful technique for
understanding variation in functional data. However, standard FPCA conflates
two distinct sources of variation:

- **Amplitude variability**: differences in the *values* or *shape* of functions
- **Phase variability**: differences in the *timing* or *alignment* of features

**Elastic FPCA** addresses this by first aligning curves using elastic
alignment (via the SRVF framework), then performing FPCA on the aligned
curves and warping functions separately.

```{=html}
<img src="../reference/figures/elastic-fpca-diagram.svg" alt="Elastic FPCA: Separating Amplitude and Phase Variation" style="width:100%; max-width:700px;">
```

## Mathematical Framework

### The SRVF Representation

Given a function $f : [0,1] \to \mathbb{R}$, its **square-root velocity
function** (SRVF) is

$$q(t) = \text{sign}\bigl(\dot{f}(t)\bigr)\,\sqrt{|\dot{f}(t)|}.$$

The key property of the SRVF is that the Fisher-Rao metric between two
functions reduces to the $\mathbb{L}^2$ distance between their SRVFs after
optimal alignment:

$$d_{FR}(f_1, f_2) = \inf_{\gamma \in \Gamma} \|q_1 - (q_2 \circ \gamma)\sqrt{\dot{\gamma}}\|_2,$$

where $\Gamma$ is the group of orientation-preserving diffeomorphisms
$\gamma : [0,1] \to [0,1]$.

### Karcher Mean

The **Karcher mean** $\bar{f}$ of a sample $\{f_1,\ldots,f_n\}$ is the
Fr\u00e9chet mean under the Fisher-Rao metric:

$$\bar{f} = \arg\min_{f} \sum_{i=1}^{n} d_{FR}(f, f_i)^2.$$

The algorithm alternates between (1) aligning each curve to the current mean
and (2) computing a new mean from the aligned curves.

### Three Modes of FPCA

After computing the Karcher mean, we obtain aligned curves
$\{\tilde{f}_i\}$ and warping functions $\{\gamma_i\}$. These are analyzed
with three variants of FPCA:

**Vertical (amplitude) FPCA**: PCA on the aligned SRVFs $\{\tilde{q}_i\}$,
capturing pure shape variation. The covariance operator is

$$C_V(s,t) = \frac{1}{n}\sum_{i=1}^{n} \bigl(\tilde{q}_i(s) - \bar{q}(s)\bigr)\bigl(\tilde{q}_i(t) - \bar{q}(t)\bigr).$$

**Horizontal (phase) FPCA**: PCA on the shooting vectors
$v_i = \exp_{\text{id}}^{-1}(\gamma_i)$ in the tangent space at the
identity warp, capturing pure timing variation.

**Joint FPCA**: Concatenates amplitude and phase representations into a
combined vector and performs PCA, revealing mixed amplitude-phase modes of
variation.

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

## Simulated Data with Amplitude + Phase Variation

```{r simulate-data}
n <- 30
m <- 101
t <- seq(0, 1, length.out = m)

# Generate curves with amplitude + phase variation
X <- matrix(0, n, m)
for (i in 1:n) {
  amp <- rnorm(1, 1, 0.3)           # amplitude variation
  shift <- rnorm(1, 0, 0.05)        # phase variation (timing shift)
  X[i, ] <- amp * sin(2 * pi * (t - shift)) +
            0.3 * amp * cos(4 * pi * (t - shift))
}
fd <- fdata(X, argvals = t)
```

```{r plot-data}
plot(fd, main = "Simulated Curves: Mixed Amplitude + Phase",
     xlab = "t", ylab = "X(t)")
```

## Elastic Alignment

Elastic FPCA begins by aligning curves with `karcher.mean()`:

```{r alignment}
karcher <- karcher.mean(fd, max.iter = 10)
```

```{r plot-aligned}
plot(karcher$aligned, main = "After Elastic Alignment",
     xlab = "t", ylab = "X(t)")
```

```{r plot-warps}
plot(karcher$gammas, main = "Warping Functions (Phase Information)",
     xlab = "t", ylab = expression(gamma(t)))
```

## Vertical FPCA: Amplitude Variability

Vertical FPCA performs PCA on the **aligned** curves, capturing pure shape
variation:

```{r vertical-fpca}
vert_result <- vert.fpca(karcher, ncomp = 3)

cat("Vertical FPCA - cumulative variance:\n")
print(round(vert_result$cumulative.variance, 4))
```

```{r plot-vert-scores}
plot(vert_result, type = "scores")
```

```{r plot-vert-eigen}
plot(vert_result, type = "eigenfunctions")
```

```{r plot-vert-variance}
plot(vert_result, type = "variance")
```

## Horizontal FPCA: Phase Variability

Horizontal FPCA analyzes the **warping functions**, capturing pure timing
variation:

```{r horizontal-fpca}
horiz_result <- horiz.fpca(karcher, ncomp = 3)

cat("Horizontal FPCA - cumulative variance:\n")
print(round(horiz_result$cumulative.variance, 4))
```

```{r plot-horiz-scores}
plot(horiz_result, type = "scores")
```

```{r plot-horiz-warps}
plot(horiz_result, type = "warps")
```

## Joint FPCA: Combined Analysis

Joint FPCA simultaneously analyzes both amplitude and phase:

```{r joint-fpca}
joint_result <- joint.fpca(karcher, ncomp = 3)

cat("Joint FPCA - cumulative variance:\n")
print(round(joint_result$cumulative.variance, 4))
```

```{r plot-joint-scores}
plot(joint_result, type = "scores")
```

```{r plot-joint-balance}
plot(joint_result, type = "balance")
```

## Practical Guidance

**Use vertical (amplitude) FPCA when:**

- You want to understand pure shape variation after removing timing differences

**Use horizontal (phase) FPCA when:**

- You want to understand timing or alignment variation

**Use joint FPCA when:**

- You want to simultaneously analyze amplitude and phase variation
- You need to understand their relative contributions

**Example applications:**

- **Growth curves**: Vertical for shape differences, horizontal for
  developmental timing
- **Gait analysis**: Vertical for movement patterns, horizontal for stride
  timing
- **Weather**: Vertical for seasonal temperature patterns, horizontal for
  seasonal shift

## References

- Srivastava, A., Wu, W., Kurtek, S., Klassen, E. and Marron, J.S. (2011).
  Registration of functional data using the Fisher-Rao metric. *arXiv
  preprint arXiv:1103.3817*.

- Tucker, J.D., Wu, W. and Srivastava, A. (2013). Generative models for
  functional data using phase and amplitude separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.

- Srivastava, A. and Klassen, E.P. (2016). *Functional and Shape Data
  Analysis*. Springer.

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  2nd ed. Springer.

## See Also

- `vignette("elastic-alignment")` --- alignment without FPCA
- `vignette("fpca")` --- standard (non-elastic) FPCA
- `vignette("elastic-regression")` --- regression using elastic FPCA scores
