---
title: "Elastic Curve Alignment"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

When comparing functional data, observed curves often exhibit both **amplitude variability** (differences in height/shape) and **phase variability** (differences in timing/speed). Standard methods like the cross-sectional mean or PCA treat all variation as amplitude, which can produce blurred averages and inflated variance.

**Elastic alignment** separates these two sources of variability by finding optimal **warping functions** $\gamma$ that retime each curve before comparison. The mathematical framework uses the **Square-Root Slope Function (SRSF)** representation, which makes the Fisher-Rao metric equivalent to the $L^2$ metric -- enabling efficient dynamic programming alignment.

## SRSF Transform

The SRSF of a function $f$ is defined as:

$$q(t) = \text{sign}(f'(t)) \sqrt{|f'(t)|}$$

This representation has the key property that the Fisher-Rao distance between two curves equals the $L^2$ distance between their SRSFs after optimal alignment.

```{r srsf-demo}
library(fdars)

set.seed(42)
argvals <- seq(0, 1, length.out = 100)

# Create sample curves with different phases
data <- matrix(0, 10, 100)
for (i in 1:10) {
  phase <- runif(1, -0.3, 0.3)
  amp <- rnorm(1, 1, 0.2)
  data[i, ] <- amp * sin(2 * pi * (argvals + phase))
}
fd <- fdata(data, argvals = argvals)

# Compute SRSF transform
q <- srsf.transform(fd)

# Visualize original curves and their SRSFs
p1 <- plot(fd) + ggplot2::labs(title = "Original Curves")
p2 <- plot(q) + ggplot2::labs(title = "SRSF Representation")
```

The SRSF transform is invertible: given $q$ and an initial value $f_0$, we can reconstruct:

$$f(t) = f_0 + \int_0^t q(s)|q(s)|\,ds$$

```{r srsf-inverse}
# Round-trip: SRSF -> inverse -> original
f_original <- fd$data[1, ]
q_values <- q$data[1, ]
f_reconstructed <- srsf.inverse(q_values, argvals, f_original[1])

plot(argvals, f_original, type = "l", col = "blue", lwd = 2,
     xlab = "t", ylab = "f(t)", main = "SRSF Round-Trip")
lines(argvals, f_reconstructed, col = "red", lty = 2, lwd = 2)
legend("topright", c("Original", "Reconstructed"), col = c("blue", "red"), lty = 1:2)
```

## Pairwise Alignment

The `elastic.align()` function aligns all curves to a common target. When no target is specified, the cross-sectional mean is used:

```{r align}
res <- elastic.align(fd)
print(res)
```

The alignment separates amplitude from phase:

```{r align-plot}
plot(res, type = "both")
```

We can also inspect the warping functions:

```{r warps}
plot(res, type = "warps")
```

## Karcher Mean

The **Karcher mean** (or Fréchet mean) computes the mean shape in the elastic metric by iteratively aligning curves to a running estimate of the mean:

```{r karcher}
km <- karcher.mean(fd, max.iter = 20, tol = 1e-4)
print(km)
```

```{r karcher-plot}
plot(km, type = "mean")
```

The Karcher mean provides a sharper, more representative average than the cross-sectional mean because it eliminates phase variability:

```{r compare-means}
cross_mean <- colMeans(fd$data)
karcher <- as.numeric(km$mean$data)

plot(argvals, cross_mean, type = "l", col = "blue", lwd = 2,
     xlab = "t", ylab = "f(t)", main = "Cross-Sectional vs Karcher Mean")
lines(argvals, karcher, col = "red", lwd = 2)
legend("topright", c("Cross-sectional", "Karcher"), col = c("blue", "red"), lty = 1)
```

## Elastic Distance Matrix

The Fisher-Rao distance provides a proper metric on function space that is invariant to reparameterization:

```{r distance}
D <- elastic.distance(fd)
round(D[1:5, 1:5], 3)
```

This can be used in any distance-based method (clustering, MDS, etc.):

```{r mds}
mds <- cmdscale(D, k = 2)
plot(mds, pch = 19, xlab = "MDS 1", ylab = "MDS 2",
     main = "MDS of Elastic Distances")
```

The elastic distance is also available through the `metric()` dispatcher:

```{r metric-elastic}
D2 <- metric(fd, method = "elastic")
all.equal(D, D2)
```
