---
title: "Elastic Changepoint Detection"
description: >
  Detect structural breaks in functional time series using amplitude,
  phase, and FPCA-based changepoint tests.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elastic Changepoint Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

**Changepoint detection** identifies the index at which a chronologically
ordered sequence of functional observations undergoes a structural change.
The `elastic.changepoint()` function provides three approaches, each
sensitive to a different type of change:

- **Amplitude changepoints**: Detect shifts in functional shape
- **Phase changepoints**: Detect shifts in feature timing
- **FPCA changepoints**: Detect changes in principal component structure
  (sensitive to both amplitude and phase)

```{=html}
<img src="../reference/figures/elastic-changepoint-diagram.svg" alt="Elastic Changepoint Detection" style="width:100%; max-width:700px;">
```

## Two Time Indices

Functional changepoint detection involves **two distinct time axes**:

- **Intra-curve time** $t \in \mathcal{T}$: the domain over which each
  individual curve is observed (the `argvals` of the fdata object). For
  example, $t$ might represent seconds within a single production cycle or
  hours within a single day.
- **Inter-curve index** $i = 1,\ldots,n$: the chronological ordering of
  curves (the row index in the fdata object). For example, $i$ might index
  successive production runs, calendar days, or patient visits.

The changepoint $\tau$ is detected on the **inter-curve axis**: "at which
curve index does the distribution of curves change?" The intra-curve domain
$t$ is used internally to compute distances, align SRVFs, and extract FPCA
features.

**Important**: The rows of the fdata object must be in chronological order.
Shuffling the rows would destroy the sequential structure that the CUSUM
test relies on.

## Mathematical Framework

### CUSUM Test for Functional Data

Given a functional time series $\{f_1,\ldots,f_n\}$ ordered chronologically
by the inter-curve index $i$, we test $H_0$: no changepoint (the
distribution is stationary across $i$) against $H_1$: there exists
$\tau \in \{1,\ldots,n-1\}$ such that the distribution changes at $\tau$.

The **CUSUM (cumulative sum)** statistic for a scalar summary $S_i$
extracted from each curve is

$$T_n(k) = \frac{1}{\hat{\sigma}\sqrt{n}} \left| \sum_{i=1}^{k} S_i - \frac{k}{n}\sum_{i=1}^{n} S_i \right|, \qquad k = 1,\ldots,n-1,$$

and the test statistic is $T_n = \max_{k} T_n(k)$, with the estimated
changepoint $\hat{\tau} = \arg\max_{k} T_n(k)$.

### How the Summary Statistic Differs

- **Amplitude**: $S_i^{(\text{amp})} = \|\tilde{q}_i - \bar{q}\|_2$
  (distance between aligned SRVF and Karcher mean SRVF). Sensitive to shape
  changes only.
- **Phase**: $S_i^{(\text{ph})} = d_{\Gamma}(\gamma_i, \text{id})$ (elastic
  distance between warp function and the identity). Sensitive to timing
  changes only.
- **FPCA**: Multivariate CUSUM on the FPC scores $\boldsymbol{\xi}_i$.
  Sensitive to changes in either amplitude, phase, or both.

### Permutation Inference

Under $H_0$, the ordering of curves is exchangeable. The p-value is
estimated by randomly permuting the sequence $B$ times:

$$\hat{p} = \frac{1}{B}\sum_{b=1}^{B} \mathbf{1}\{T_n^{(b)} \geq T_n\}.$$

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)

n <- 60
m <- 101
tau <- 30
t <- seq(0, 1, length.out = m)
```

## Example 1: Pure Amplitude Changepoint

In this scenario, the **shape** of the curves changes at $\tau = 30$ but
the timing stays the same. Before the break: simple sine waves. After: the
amplitude doubles and a second harmonic appears.

```{=html}
<img src="../reference/figures/changepoint-amplitude-diagram.svg" alt="Amplitude Changepoint: shape changes, timing stays the same" style="width:100%; max-width:700px;">
```

```{r sim-amplitude}
curves_amp <- matrix(0, n, m)
for (i in 1:tau)
  curves_amp[i, ] <- sin(2 * pi * t) + rnorm(m, sd = 0.12)
for (i in (tau + 1):n)
  curves_amp[i, ] <- 2 * sin(2 * pi * t) + 0.8 * sin(4 * pi * t) +
                     rnorm(m, sd = 0.12)
fd_amp <- fdata(curves_amp, argvals = t)
```

```{r plot-amp-data}
plot(fd_amp, main = "Example 1: Pure Amplitude Change at obs 30",
     xlab = "t", ylab = "X(t)")
```

We run both the amplitude and phase tests. Since only shape changed, the
amplitude test should produce a much larger CUSUM statistic:

```{r test-amplitude}
res_amp_a <- elastic.changepoint(fd_amp, type = "amplitude", n.mc = 199, seed = 42)
res_amp_p <- elastic.changepoint(fd_amp, type = "phase", n.mc = 199, seed = 42)

cat("Amplitude test: detected at", res_amp_a$changepoint,
    "| CUSUM =", round(res_amp_a$test.statistic, 5),
    "| p =", format(res_amp_a$p.value, digits = 3), "\n")
cat("Phase test:     detected at", res_amp_p$changepoint,
    "| CUSUM =", round(res_amp_p$test.statistic, 5),
    "| p =", format(res_amp_p$p.value, digits = 3), "\n")
cat("Amplitude/Phase CUSUM ratio:", round(res_amp_a$test.statistic /
                                     res_amp_p$test.statistic, 1), "x\n")
```

The amplitude test is highly significant (small p-value) with a much larger
CUSUM, confirming the change is in shape, not timing. The phase test may
also reach significance because shape changes can induce spurious phase
variation, but the CUSUM ratio clearly indicates a shape-driven change.

```{r plot-amp-cusum}
plot(res_amp_a, type = "statistic")
```

```{r plot-amp-segments}
plot(res_amp_a, type = "data")
```

## Example 2: Pure Phase Changepoint

Here the **shape** is identical before and after, but the **timing** shifts:
after $\tau = 30$, the same template is observed through a nonlinear
time warp (as if the process runs faster at the start and slower at the
end).

```{=html}
<img src="../reference/figures/changepoint-phase-diagram.svg" alt="Phase Changepoint: timing changes, shape stays the same" style="width:100%; max-width:700px;">
```

```{r sim-phase}
template_ph <- sin(2 * pi * t) + 0.4 * cos(4 * pi * t)

curves_ph <- matrix(0, n, m)
for (i in 1:tau)
  curves_ph[i, ] <- template_ph + rnorm(m, sd = 0.12)
for (i in (tau + 1):n) {
  gamma <- pbeta(t, 1.6, 0.7)  # nonlinear warp
  curves_ph[i, ] <- approx(t, template_ph, xout = gamma, rule = 2)$y +
                    rnorm(m, sd = 0.12)
}
fd_ph <- fdata(curves_ph, argvals = t)
```

```{r plot-ph-data}
plot(fd_ph, main = "Example 2: Pure Phase Change at obs 30",
     xlab = "t", ylab = "X(t)")
```

Now the phase test should dominate:

```{r test-phase}
res_ph_a <- elastic.changepoint(fd_ph, type = "amplitude", n.mc = 199, seed = 42)
res_ph_p <- elastic.changepoint(fd_ph, type = "phase", n.mc = 199, seed = 42)

cat("Amplitude test: detected at", res_ph_a$changepoint,
    "| CUSUM =", round(res_ph_a$test.statistic, 5),
    "| p =", format(res_ph_a$p.value, digits = 3), "\n")
cat("Phase test:     detected at", res_ph_p$changepoint,
    "| CUSUM =", round(res_ph_p$test.statistic, 5),
    "| p =", format(res_ph_p$p.value, digits = 3), "\n")
cat("Phase/Amplitude CUSUM ratio:", round(res_ph_p$test.statistic /
                                     res_ph_a$test.statistic, 1), "x\n")
```

The phase CUSUM is larger and significant, confirming the change is in
timing, not shape.

```{r plot-ph-cusum}
plot(res_ph_p, type = "statistic")
```

```{r plot-ph-segments}
plot(res_ph_p, type = "data")
```

## Example 3: Mixed Changepoint (Amplitude + Phase)

In practice, structural breaks often involve **both** shape and timing
changes simultaneously. After $\tau = 30$, the curves are scaled, gain a
harmonic, and are nonlinearly warped.

```{=html}
<img src="../reference/figures/changepoint-mixed-diagram.svg" alt="Mixed Changepoint: both shape and timing change" style="width:100%; max-width:700px;">
```

```{r sim-mixed}
curves_mix <- matrix(0, n, m)
for (i in 1:tau)
  curves_mix[i, ] <- sin(2 * pi * t) + rnorm(m, sd = 0.12)
for (i in (tau + 1):n) {
  gamma <- pbeta(t, 1.4, 0.7)
  template_mix <- 1.5 * sin(2 * pi * t) + 0.5 * cos(4 * pi * t)
  curves_mix[i, ] <- approx(t, template_mix, xout = gamma, rule = 2)$y +
                     rnorm(m, sd = 0.12)
}
fd_mix <- fdata(curves_mix, argvals = t)
```

```{r plot-mix-data}
plot(fd_mix, main = "Example 3: Mixed Change at obs 30",
     xlab = "t", ylab = "X(t)")
```

All three tests detect the changepoint, but the FPCA-based test — which
captures both amplitude and phase variation — produces the strongest signal:

```{r test-mixed}
res_mix_a <- elastic.changepoint(fd_mix, type = "amplitude", n.mc = 199, seed = 42)
res_mix_p <- elastic.changepoint(fd_mix, type = "phase", n.mc = 199, seed = 42)
res_mix_f <- elastic.changepoint(fd_mix, type = "fpca", pca.method = "vertical",
                                 ncomp = 3, n.mc = 199, seed = 42)

cat("Amplitude test: detected at", res_mix_a$changepoint,
    "| CUSUM =", round(res_mix_a$test.statistic, 5),
    "| p =", format(res_mix_a$p.value, digits = 3), "\n")
cat("Phase test:     detected at", res_mix_p$changepoint,
    "| CUSUM =", round(res_mix_p$test.statistic, 5),
    "| p =", format(res_mix_p$p.value, digits = 3), "\n")
cat("FPCA test:      detected at", res_mix_f$changepoint,
    "| CUSUM =", round(res_mix_f$test.statistic, 5),
    "| p =", format(res_mix_f$p.value, digits = 3), "\n")
```

```{r plot-mix-cusum}
plot(res_mix_f, type = "statistic")
```

```{r plot-mix-segments}
plot(res_mix_f, type = "data")
```

## Summary: Which Test for Which Change?

```{r summary-table}
summary_df <- data.frame(
  Example = c("1. Pure Amplitude", "1. Pure Amplitude",
              "2. Pure Phase", "2. Pure Phase",
              "3. Mixed", "3. Mixed", "3. Mixed"),
  Test = c("amplitude", "phase",
           "amplitude", "phase",
           "amplitude", "phase", "fpca"),
  Detected = c(res_amp_a$changepoint, res_amp_p$changepoint,
               res_ph_a$changepoint, res_ph_p$changepoint,
               res_mix_a$changepoint, res_mix_p$changepoint,
               res_mix_f$changepoint),
  CUSUM = round(c(res_amp_a$test.statistic, res_amp_p$test.statistic,
                   res_ph_a$test.statistic, res_ph_p$test.statistic,
                   res_mix_a$test.statistic, res_mix_p$test.statistic,
                   res_mix_f$test.statistic), 5),
  p.value = c(res_amp_a$p.value, res_amp_p$p.value,
              res_ph_a$p.value, res_ph_p$p.value,
              res_mix_a$p.value, res_mix_p$p.value,
              res_mix_f$p.value)
)
knitr::kable(summary_df, digits = c(0, 0, 0, 5, 4))
```

| Change type | Best test | Rationale |
|:---|:---|:---|
| Shape only | `type = "amplitude"` | SRVF distance isolates shape after removing phase |
| Timing only | `type = "phase"` | Warp distance isolates timing after alignment |
| Both | `type = "fpca"` | FPCA scores capture combined amplitude + phase variation |

## Quantifying Change Strength

After detecting a changepoint at $\hat{\tau}$, we can quantify *how large*
the change is by comparing the two segments. Two natural effect size measures
are:

- **L2 effect size**: the root-mean-square distance between the mean curves
  of the before and after segments,
  $$\Delta_{\text{L2}} = \left(\int_{\mathcal{T}} \bigl(\bar{f}_{\text{before}}(t) - \bar{f}_{\text{after}}(t)\bigr)^2 \, dt\right)^{1/2}.$$
- **Relative change**: $\Delta_{\text{L2}} / \|\bar{f}_{\text{before}}\|_2
  \times 100\%$, expressing the shift as a percentage of the baseline signal
  magnitude.

```{r effect-size}
effect_size <- function(fdataobj, changepoint) {
  X <- fdataobj$data
  n <- nrow(X)
  before <- colMeans(X[1:changepoint, , drop = FALSE])
  after  <- colMeans(X[(changepoint + 1):n, , drop = FALSE])
  l2 <- sqrt(mean((after - before)^2))
  baseline <- sqrt(mean(before^2))
  list(l2 = l2, relative_pct = l2 / baseline * 100,
       max_pointwise = max(abs(after - before)))
}

# Amplitude example
es_amp <- effect_size(fd_amp, res_amp_a$changepoint)
cat("Example 1 (Amplitude):\n")
cat("  L2 effect size:", round(es_amp$l2, 3), "\n")
cat("  Relative change:", round(es_amp$relative_pct, 1), "%\n")
cat("  Max pointwise diff:", round(es_amp$max_pointwise, 3), "\n\n")

# Phase example
es_ph <- effect_size(fd_ph, res_ph_p$changepoint)
cat("Example 2 (Phase):\n")
cat("  L2 effect size:", round(es_ph$l2, 3), "\n")
cat("  Relative change:", round(es_ph$relative_pct, 1), "%\n")
cat("  Max pointwise diff:", round(es_ph$max_pointwise, 3), "\n\n")

# Mixed example
es_mix <- effect_size(fd_mix, res_mix_f$changepoint)
cat("Example 3 (Mixed):\n")
cat("  L2 effect size:", round(es_mix$l2, 3), "\n")
cat("  Relative change:", round(es_mix$relative_pct, 1), "%\n")
cat("  Max pointwise diff:", round(es_mix$max_pointwise, 3), "\n")
```

The **CUSUM statistic itself** is also a measure of change strength: it scales
monotonically with the signal-to-noise ratio. Comparing CUSUM values across
the amplitude and phase tests reveals the *nature* of the change
(shape vs timing), while the magnitude indicates its severity.

## Detection Sensitivity and the Role of Noise

A key practical question is: **how strong must a break be for the test to
detect it?** The answer depends not just on the signal strength but on the
**noise level**. The same 50% amplitude increase is easy to detect at
$\sigma = 0.05$ but may be invisible at $\sigma = 0.4$. We therefore sweep
both dimensions to produce **detection surfaces**.

### Amplitude × Noise Surface

We vary the amplitude scaling factor and the observation noise simultaneously.

```{r sens-amp-surface, fig.height=5}
amp_factors <- c(1.0, 1.1, 1.25, 1.5, 2.0, 3.0)
noise_sds <- c(0.05, 0.12, 0.25, 0.4)

# Helper: compute rect boundaries for non-uniform grids (no white gaps)
add_rect_bounds <- function(df, x_col, y_col) {
  midbreaks <- function(v) {
    u <- sort(unique(v))
    mids <- (u[-1] + u[-length(u)]) / 2
    c(u[1] - (mids[1] - u[1]), mids,
      u[length(u)] + (u[length(u)] - mids[length(mids)]))
  }
  xb <- midbreaks(df[[x_col]])
  yb <- midbreaks(df[[y_col]])
  xs <- sort(unique(df[[x_col]]))
  ys <- sort(unique(df[[y_col]]))
  xi <- match(df[[x_col]], xs)
  yi <- match(df[[y_col]], ys)
  df$xmin <- xb[xi];  df$xmax <- xb[xi + 1]
  df$ymin <- yb[yi];  df$ymax <- yb[yi + 1]
  df
}

amp_grid <- expand.grid(factor = amp_factors, noise_sd = noise_sds)
amp_grid$cusum <- NA_real_
amp_grid$p_value <- NA_real_
amp_grid$detected <- NA_integer_

for (r in seq_len(nrow(amp_grid))) {
  af <- amp_grid$factor[r]
  sd_val <- amp_grid$noise_sd[r]
  set.seed(42)
  curves_s <- matrix(0, n, m)
  for (i in 1:tau)
    curves_s[i, ] <- sin(2 * pi * t) + rnorm(m, sd = sd_val)
  for (i in (tau + 1):n)
    curves_s[i, ] <- af * sin(2 * pi * t) + (af - 1) * 0.4 * sin(4 * pi * t) +
                     rnorm(m, sd = sd_val)
  fd_s <- fdata(curves_s, argvals = t)
  res_s <- elastic.changepoint(fd_s, type = "amplitude", n.mc = 199, seed = 42)
  amp_grid$cusum[r] <- res_s$test.statistic
  amp_grid$p_value[r] <- res_s$p.value
  amp_grid$detected[r] <- res_s$changepoint
}
```

The heatmap shows the full detection surface, with the $p = 0.05$ contour
marking the detection frontier:

```{r amp-heatmap, fig.height=4.5}
amp_grid <- add_rect_bounds(amp_grid, "factor", "noise_sd")

ggplot(amp_grid, aes(fill = p_value)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_contour(aes(x = factor, y = noise_sd, z = p_value), breaks = 0.05,
               color = "white", linewidth = 1) +
  scale_fill_viridis_c(option = "D", direction = -1, limits = c(0, 1),
                       name = "p-value") +
  labs(title = "Amplitude Changepoint Detection Surface",
       subtitle = "White contour = p = 0.05 boundary",
       x = "Amplitude Scaling Factor", y = "Noise SD") +
  theme(legend.position = "right")
```

For comparison, here is the 1D slice at the default noise level
($\sigma = 0.12$):

```{r amp-slice, fig.height=4}
amp_slice <- amp_grid[amp_grid$noise_sd == 0.12, ]

ggplot(amp_slice, aes(x = factor, y = p_value)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "#D55E00") +
  annotate("text", x = 2.5, y = 0.08, label = "alpha == 0.05",
           parse = TRUE, color = "#D55E00") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Amplitude Test at SD = 0.12 (Slice of Surface)",
       x = "Amplitude Scaling Factor",
       y = "Permutation p-value")
```

### Phase × Noise Surface

We vary the warp strength and noise level. The warp is a Beta CDF
$\gamma(t) = F_{\text{Beta}}(t; a, 1/a)$ and the "max deviation" is
$\max_t |\gamma(t) - t|$.

```{r sens-phase-surface, fig.height=5}
warp_strengths <- c(1.0, 1.05, 1.1, 1.2, 1.4, 2.0)

ph_grid <- expand.grid(warp = warp_strengths, noise_sd = noise_sds)
ph_grid$max_dev <- NA_real_
ph_grid$cusum <- NA_real_
ph_grid$p_value <- NA_real_

for (r in seq_len(nrow(ph_grid))) {
  ws <- ph_grid$warp[r]
  sd_val <- ph_grid$noise_sd[r]
  set.seed(42)
  curves_s <- matrix(0, n, m)
  for (i in 1:tau)
    curves_s[i, ] <- template_ph + rnorm(m, sd = sd_val)
  gamma_s <- pbeta(t, ws, 1 / ws)
  ph_grid$max_dev[r] <- max(abs(gamma_s - t))
  for (i in (tau + 1):n)
    curves_s[i, ] <- approx(t, template_ph, xout = gamma_s, rule = 2)$y +
                     rnorm(m, sd = sd_val)
  fd_s <- fdata(curves_s, argvals = t)
  res_s <- elastic.changepoint(fd_s, type = "phase", n.mc = 199, seed = 42)
  ph_grid$cusum[r] <- res_s$test.statistic
  ph_grid$p_value[r] <- res_s$p.value
}
```

```{r phase-heatmap, fig.height=4.5}
ph_grid <- add_rect_bounds(ph_grid, "warp", "noise_sd")

ggplot(ph_grid, aes(fill = p_value)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_contour(aes(x = warp, y = noise_sd, z = p_value), breaks = 0.05,
               color = "white", linewidth = 1) +
  scale_fill_viridis_c(option = "D", direction = -1, limits = c(0, 1),
                       name = "p-value") +
  labs(title = "Phase Changepoint Detection Surface",
       subtitle = "White contour = p = 0.05 boundary",
       x = "Warp Strength", y = "Noise SD") +
  theme(legend.position = "right")
```

### CUSUM Degradation with Noise

Fixing the amplitude factor at 1.5, we see how the CUSUM statistic degrades
as noise increases. The test statistic shrinks because the signal is
increasingly buried in noise.

```{r cusum-vs-noise, fig.height=4}
noise_sweep <- seq(0.05, 0.5, by = 0.05)
cusum_noise <- data.frame(noise_sd = numeric(), cusum = numeric(),
                          p_value = numeric())

for (sd_val in noise_sweep) {
  set.seed(42)
  curves_s <- matrix(0, n, m)
  for (i in 1:tau)
    curves_s[i, ] <- sin(2 * pi * t) + rnorm(m, sd = sd_val)
  for (i in (tau + 1):n)
    curves_s[i, ] <- 1.5 * sin(2 * pi * t) + 0.5 * 0.4 * sin(4 * pi * t) +
                     rnorm(m, sd = sd_val)
  fd_s <- fdata(curves_s, argvals = t)
  res_s <- elastic.changepoint(fd_s, type = "amplitude", n.mc = 199, seed = 42)
  cusum_noise <- rbind(cusum_noise, data.frame(
    noise_sd = sd_val, cusum = res_s$test.statistic, p_value = res_s$p.value
  ))
}

ggplot(cusum_noise, aes(x = noise_sd, y = cusum)) +
  geom_line(linewidth = 1) +
  geom_point(aes(color = p_value < 0.05), size = 3) +
  scale_color_manual(values = c("TRUE" = "#2166AC", "FALSE" = "#B2182B"),
                     labels = c("TRUE" = "Significant", "FALSE" = "Not significant"),
                     name = "") +
  labs(title = "CUSUM Degradation: Factor = 1.5 Across Noise Levels",
       subtitle = "Test statistic shrinks as noise buries the signal",
       x = "Noise SD", y = "CUSUM Statistic")
```

### Minimum Detectable Effect

For each noise level, the smallest amplitude factor and warp strength where
the test is significant:

```{r mde-changepoint, results='asis'}
# Amplitude: minimum factor per noise level
min_amp_cp <- do.call(rbind, lapply(noise_sds, function(sd_val) {
  sub <- amp_grid[amp_grid$noise_sd == sd_val & amp_grid$factor > 1, ]
  det <- sub$factor[sub$p_value < 0.05]
  data.frame(noise_sd = sd_val,
             min_factor = if (length(det) > 0) min(det) else NA)
}))

# Phase: minimum warp (as max deviation %) per noise level
min_ph_cp <- do.call(rbind, lapply(noise_sds, function(sd_val) {
  sub <- ph_grid[ph_grid$noise_sd == sd_val & ph_grid$warp > 1, ]
  det <- sub[sub$p_value < 0.05, ]
  data.frame(noise_sd = sd_val,
             min_warp_dev = if (nrow(det) > 0)
               paste0(round(min(det$max_dev) * 100, 1), "%") else NA)
}))

mde_cp <- data.frame(
  `Noise SD` = noise_sds,
  `Min Amp Factor` = min_amp_cp$min_factor,
  `Min Warp (max dev %)` = min_ph_cp$min_warp_dev,
  check.names = FALSE
)
knitr::kable(mde_cp, caption = "Minimum detectable effect at alpha = 0.05")
```

### Practical Guidance

- **Low noise** ($\sigma \leq 0.1$): even 10% amplitude changes and 5%
  temporal displacements are detectable.
- **Moderate noise** ($\sigma \approx 0.2$): need ~50% amplitude change
  or ~10% temporal displacement.
- **High noise** ($\sigma \geq 0.4$): only dramatic changes (2x+ amplitude)
  are reliably detected.
- **Phase detection is inherently more robust**: elastic alignment removes
  amplitude variation before testing, so the phase test is less affected by
  amplitude noise. This makes it the preferred first test when the noise
  level is uncertain.

## Process Monitoring: Statistical Decision Boundary

For process monitoring applications, a binary **OK / Not OK** decision
requires a statistically valid threshold. The permutation p-value provides
exactly this: at significance level $\alpha$, reject $H_0$ (no change)
when $p \leq \alpha$.

### Setting the Decision Rule

The decision framework works as follows:

1. **Collect a calibration set** of $n$ functional observations in
   chronological order during a period of known stable operation.
2. **Run `elastic.changepoint()`** with `n.mc >= 1000` for a stable p-value.
3. **Decision**:
   - If $p \leq \alpha$ (e.g., $\alpha = 0.05$): **reject** $H_0$.
     A structural change is detected at the estimated $\hat{\tau}$.
   - If $p > \alpha$: **do not reject**. No significant change detected.

### Interpreting the Result

When a changepoint is detected:

| What to inspect | How |
|:---|:---|
| **Where** did it happen? | `result$changepoint` (curve index) |
| **How strong** is the change? | Segment L2 effect size (see above) |
| **What kind** of change? | Compare amplitude vs phase CUSUM |
| **How confident** are we? | `result$p.value` (smaller = stronger evidence) |

### Important Considerations

- **Multiple testing**: If you test the same data with amplitude, phase, and
  FPCA tests, apply a Bonferroni correction ($\alpha / 3$) or use a single
  FPCA test as a screening test.
- **Sample size**: The permutation test has a minimum achievable p-value of
  $1 / (B + 1)$, where $B$ = `n.mc`. Use `n.mc >= 999` for decisions at
  $\alpha = 0.05$, and `n.mc >= 9999` for $\alpha = 0.01$.
- **Prospective monitoring**: For ongoing monitoring of a production process,
  apply the test to a sliding window of the most recent $n$ curves. This
  provides near-real-time detection of process shifts.

## Applications

Elastic changepoint detection is useful whenever a sequence of functional
observations may undergo a structural break. The separation of amplitude
and phase is the key advantage over classical changepoint methods.

### Manufacturing and Quality Control

Sensor profiles from a production line (e.g., temperature curves from a
furnace, spectral measurements of output material) can shift when a tool
wears out, a raw material batch changes, or a machine is recalibrated.

- **Amplitude changepoint**: detects when the *shape* of production curves
  changes (defective product, drift in process output).
- **Phase changepoint**: detects when the *timing* of a process step shifts
  (e.g., a heating ramp slows down) even if the final product still meets
  specs.

### Biomedicine and Clinical Trials

Longitudinal patient curves (EEG signals, growth trajectories, gait cycles)
may change during treatment or disease progression.

- Detect when a patient's gait pattern changes after surgery (amplitude) or
  when stride timing shifts (phase).
- In clinical trials, identify the time point at which a drug begins to take
  effect by monitoring functional biomarker profiles.

### Environmental and Climate Monitoring

Seasonal temperature profiles, river discharge curves, or air quality
time series may undergo regime changes.

- **Amplitude changepoint**: identifies when seasonal patterns intensify or
  weaken (e.g., hotter summers after a climate shift).
- **Phase changepoint**: identifies when seasons shift in timing (e.g.,
  earlier spring onset due to warming trends).

### Speech and Audio Processing

Speaker intonation profiles or phoneme spectrograms may change mid-recording
due to emotion, fatigue, or speaker switches.

- Phase changepoints detect changes in speaking rate or rhythm.
- Amplitude changepoints detect changes in pitch contour shape.

### Finance

Yield curves, volatility surfaces, or term structure functions observed
daily may undergo structural breaks during policy changes or market crises.

## Practical Guidance

**Use amplitude changepoints** for changes in functional shape (quality
control, growth curves).

**Use phase changepoints** for changes in timing (speech patterns, seasonal
shifts).

**Use FPCA changepoints** for changes in principal component structure.

**Tips:**

- Use `n.mc` >= 1000 for stable p-values (200 used here for speed)
- Set `seed` for reproducibility
- Compare CUSUM magnitudes across test types to diagnose the nature of a
  detected change
- Combine amplitude and phase tests for a complete picture: a process
  may change in shape, timing, or both

## References

- Aue, A. and Horv\u00e1th, L. (2013). Structural breaks in time series.
  *Journal of Time Series Analysis*, 34(1), 1--16.

- Berkes, I., Gabrys, R., Horv\u00e1th, L. and Kokoszka, P. (2009).
  Detecting changes in the mean of functional observations. *Journal of the
  Royal Statistical Society: Series B*, 71(5), 927--946.

- Srivastava, A., Wu, W., Kurtek, S., Klassen, E. and Marron, J.S. (2011).
  Registration of functional data using the Fisher-Rao metric. *arXiv
  preprint arXiv:1103.3817*.

## See Also

- `vignette("elastic-alignment")` --- elastic curve alignment
- `vignette("elastic-fpca")` --- elastic FPCA for amplitude/phase separation
- `vignette("streaming-depth")` --- online monitoring of functional data
