---
title: "Elastic Regression"
description: >
  Joint alignment and regression for functional data using elastic methods.
  Includes scalar-on-function regression, logistic classification, and PCR.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elastic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Functional regression often assumes curves are meaningfully aligned before
analysis. However, when functional predictors contain **phase variability**
(differences in timing or alignment), standard regression methods blend
amplitude and phase variation, reducing predictive power and interpretability
of the coefficient function.

**Elastic regression** addresses this by jointly estimating the alignment
warping functions and regression coefficients, separating phase from
amplitude variation.

```{=html}
<img src="../reference/figures/elastic-regression-diagram.svg" alt="Elastic Regression: Joint Alignment + Prediction" style="width:100%; max-width:700px;">
```

## Mathematical Framework

### Standard Functional Linear Model

The scalar-on-function model is

$$y_i = \alpha + \int_0^1 X_i(t)\,\beta(t)\,\mathrm{d}t + \varepsilon_i.$$

When curves $X_i$ are misaligned, the integral mixes amplitude and phase
effects, and the estimated $\hat{\beta}(t)$ becomes blurred and
uninterpretable.

### Elastic Regression Model

Elastic regression introduces warping functions $\gamma_i \in \Gamma$:

$$y_i = \alpha + \int_0^1 \tilde{X}_i(t)\,\beta(t)\,\mathrm{d}t + \varepsilon_i, \qquad \tilde{X}_i = X_i \circ \gamma_i.$$

The algorithm alternates between:

1. **Alignment step**: Fix $\beta(t)$, update warping functions $\gamma_i$
   to minimize the residual sum of squares via the Fisher-Rao elastic metric.
2. **Regression step**: Fix warping functions, re-estimate $\alpha$ and
   $\beta(t)$ from the aligned curves.

Convergence is measured by the relative change in SSE. The SRVF
representation $q_i = \text{SRVF}(\tilde{X}_i)$ ensures the alignment
respects the elastic geometry (Srivastava et al., 2011).

### Elastic Logistic Model

For binary classification $y_i \in \{0,1\}$, the model becomes

$$\log\frac{P(y_i = 1)}{P(y_i = 0)} = \alpha + \int_0^1 \tilde{X}_i(t)\,\beta(t)\,\mathrm{d}t,$$

with the same alternating alignment-regression optimization.

### Elastic PCR

Elastic Principal Component Regression first computes elastic FPCA scores
(vertical, horizontal, or joint), then regresses $y$ on the scores:

$$y_i = \alpha + \sum_{k=1}^{K} b_k\,\xi_{ik} + \varepsilon_i,$$

where $\xi_{ik}$ are scores from the chosen FPCA mode.

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

## Simulated Data: Warped Template with Amplitude Signal

We simulate a scenario that strongly favours elastic regression: all curves
share a common shape (template), but each is **nonlinearly warped** by
a random diffeomorphism. The response depends on a subtle *amplitude*
perturbation that is invisible to standard FPCA because the dominant
variation is phase:

```{r create-data}
n <- 60
m <- 80
t <- seq(0, 1, length.out = m)

template <- function(s) sin(2 * pi * s) + 0.3 * sin(6 * pi * s)

X <- matrix(0, n, m)
y <- numeric(n)
for (i in 1:n) {
  # Random nonlinear warp via Beta CDF
  a <- runif(1, 0.5, 2.0)
  b <- runif(1, 0.5, 2.0)
  gamma <- pbeta(t, a, b)

  # Amplitude perturbation (the signal)
  delta <- rnorm(1, 0, 0.4)

  # Warped and scaled template
  X[i, ] <- (1 + 0.2 * delta) *
            approx(t, template(t), xout = gamma, rule = 2)$y +
            rnorm(m, sd = 0.1)

  # Response depends on amplitude perturbation
  y[i] <- 2 * delta + rnorm(1, sd = 0.3)
}
fd <- fdata(X, argvals = t)
```

```{r plot-data}
plot(fd, main = "Functional Predictors: Nonlinearly Warped Template",
     xlab = "t", ylab = "X(t)")
```

Standard FPCA spends nearly all its components capturing the dominant warp
variation, leaving little capacity for the amplitude signal that drives
the response.

## Standard Regression (Baseline)

```{r standard-regression}
fit_std <- fregre.lm(fd, y, ncomp = 5)
cat("Standard fregre.lm R\u00b2:", round(fit_std$r.squared, 4), "\n")
```

```{r plot-standard}
df_std <- data.frame(Observed = y, Fitted = fit_std$fitted.values)
ggplot(df_std, aes(x = Observed, y = Fitted)) +
  geom_point(color = "#4A90D9", size = 2.5, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Standard Regression: Observed vs Fitted",
       subtitle = paste("R\u00b2 =", round(fit_std$r.squared, 3)),
       x = "Observed", y = "Fitted")
```

Standard regression fails because the FPCs capture warp variation rather
than the amplitude signal the response depends on.

## Elastic Scalar-on-Function Regression

`elastic.regression()` jointly estimates alignment and regression, removing
the phase variation before fitting:

```{r elastic-regression}
fit_elastic <- elastic.regression(fd, y, ncomp.beta = 5, lambda = 0.01,
                                  max.iter = 20, tol = 1e-3)
cat("Elastic regression R\u00b2:", round(fit_elastic$r.squared, 4), "\n")
cat("Iterations:", fit_elastic$n.iter, "\n")
```

```{r plot-elastic-fit}
plot(fit_elastic, type = "fit")
```

```{r plot-elastic-beta}
plot(fit_elastic, type = "beta")
```

```{r plot-elastic-residuals}
plot(fit_elastic, type = "residuals")
```

### Comparison

```{r comparison}
comp <- data.frame(
  Method = c("Standard (fregre.lm)", "Elastic Regression"),
  R2 = c(fit_std$r.squared, fit_elastic$r.squared),
  RMSE = c(sqrt(mean(fit_std$residuals^2)),
           sqrt(mean(fit_elastic$residuals^2)))
)
print(comp)
```

## Elastic Logistic Classification

For binary classification:

```{r elastic-logistic}
y_bin <- as.numeric(y > median(y))

fit_logistic <- elastic.logistic(fd, y_bin, ncomp.beta = 5, lambda = 0.01,
                                 max.iter = 15, tol = 1e-3)
cat("Accuracy:", round(fit_logistic$accuracy * 100, 1), "%\n")
```

```{r plot-logistic}
plot(fit_logistic, type = "probabilities")
```

## Elastic PCR

Compare vertical, horizontal, and joint PCA methods:

```{r elastic-pcr}
fit_vert <- elastic.pcr(fd, y, ncomp = 3, pca.method = "vertical")
fit_horiz <- elastic.pcr(fd, y, ncomp = 3, pca.method = "horizontal")
fit_joint <- elastic.pcr(fd, y, ncomp = 3, pca.method = "joint")

comp_pcr <- data.frame(
  Method = c("Vertical", "Horizontal", "Joint"),
  R2 = c(fit_vert$r.squared, fit_horiz$r.squared, fit_joint$r.squared)
)
print(comp_pcr)
```

```{r plot-pcr-fit}
plot(fit_joint, type = "fit")
```

```{r plot-pcr-coef}
plot(fit_joint, type = "coefficients")
```

## Amplitude vs Phase Attribution

`elastic.attribution()` quantifies how much of the prediction accuracy is
attributable to amplitude variation (shape after alignment) versus phase
variation (warping), using elastic PCR scores. This helps decide whether
alignment is worthwhile for a given dataset.

```{r attribution}
attr <- elastic.attribution(fd, y, ncomp = 3, pca.method = "vertical",
                            n.perm = 50, seed = 42)

cat("Amplitude importance:", round(attr$amplitude_importance, 4), "\n")
cat("Phase importance:", round(attr$phase_importance, 4), "\n")
```

```{r plot-attribution}
df_attr <- data.frame(
  Source = c("Amplitude", "Phase"),
  Importance = c(attr$amplitude_importance, attr$phase_importance)
)

ggplot(df_attr, aes(x = Source, y = Importance, fill = Source)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(values = c("Amplitude" = "#4A90D9", "Phase" = "#D55E00")) +
  labs(title = "Amplitude vs Phase Attribution",
       subtitle = "Relative contribution to prediction",
       y = "Importance") +
  theme(legend.position = "none")
```

In our simulated data, the response depends on amplitude perturbations, so
`elastic.attribution()` correctly identifies amplitude as the dominant
predictive source.

## When to Use Elastic Regression

- **Phase variability dominates**: when nonlinear warping is the primary
  source of variation, standard FPCA wastes components on phase
- **Interpretability important**: want a clear coefficient function after
  removing phase
- **Classification with misalignment**: use `elastic.logistic()` for groups
  with different peak timings
- **Attribution needed**: `elastic.attribution()` can quantify amplitude
  vs phase contributions before committing to a full elastic model

Avoid elastic regression if curves are already well-aligned or if the
response depends on phase variation itself.

## References

- Srivastava, A., Wu, W., Kurtek, S., Klassen, E. and Marron, J.S. (2011).
  Registration of functional data using the Fisher-Rao metric. *arXiv
  preprint arXiv:1103.3817*.

- Tucker, J.D., Wu, W. and Srivastava, A. (2013). Generative models for
  functional data using phase and amplitude separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  2nd ed. Springer.

## See Also

- `vignette("elastic-alignment")` --- alignment without regression
- `vignette("articles/scalar-on-function")` --- standard scalar-on-function regression
- `vignette("elastic-fpca")` --- elastic FPCA for amplitude/phase separation
