---
title: "Elastic Clustering"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Standard clustering methods like k-means use Euclidean (L2) distances to
group curves. This works well when curves are aligned --- but functional data
often exhibits **phase variation**, where the same underlying pattern appears
at different speeds or timings across observations. A peak that occurs early
in one curve and late in another will inflate the Euclidean distance, even
when the two curves have the same shape.

**Elastic clustering** solves this by replacing the Euclidean distance with
the **elastic (Fisher-Rao) distance**, which factors out reparameterization
before measuring dissimilarity. Cluster centers are computed as Karcher means
in the elastic metric, so they represent the average *shape* of each group
rather than a blurred pointwise average.

```{=html}
<img src="../reference/figures/elastic-clustering-overview.svg" alt="Elastic Clustering Overview" style="width:100%; max-width:720px; display:block; margin:1em auto;" />
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Simulated Gait Data

We simulate 60 knee-angle-like curves from three distinct movement patterns,
each with random phase variation. Group 1 produces sinusoidal flexion--extension
cycles, Group 2 produces cosine-shaped curves (shifted phase pattern), and
Group 3 produces an asymmetric push-off profile. Within each group, curves
share the same shape but are randomly retimed.

```{r sim-data}
set.seed(42)
n_per_group <- 20
m <- 50
argvals <- seq(0, 1, length.out = m)

# 3 groups with different base patterns + phase variation
X <- matrix(0, 3 * n_per_group, m)
for (i in 1:n_per_group) {
  phase_shift <- runif(1, -0.1, 0.1)
  X[i, ] <- sin(2 * pi * (argvals + phase_shift)) + rnorm(m, sd = 0.1)
}
for (i in 1:n_per_group) {
  phase_shift <- runif(1, -0.1, 0.1)
  X[n_per_group + i, ] <- cos(2 * pi * (argvals + phase_shift)) +
    rnorm(m, sd = 0.1)
}
for (i in 1:n_per_group) {
  phase_shift <- runif(1, -0.1, 0.1)
  X[2 * n_per_group + i, ] <- argvals^2 * (1 - argvals) * 4 +
    rnorm(m, sd = 0.1) + phase_shift
}
fd <- fdata(X, argvals = argvals)
true_labels <- rep(1:3, each = n_per_group)
```

```{r plot-data}
plot(fd, main = "Simulated Gait Curves (3 groups with phase variation)",
     xlab = "Normalized Stride Cycle", ylab = "Knee Angle (a.u.)")
```

The three groups are visually distinct in shape, but the random timing shifts
make pointwise comparison unreliable. This is exactly the scenario where
elastic clustering outperforms standard methods.

## Elastic K-Means

`elastic.kmeans()` partitions curves into k clusters using the elastic
distance. At each iteration, cluster centers are updated as Karcher means ---
the curve that minimizes total squared elastic distance to all members.

```{r elastic-kmeans}
ekm <- elastic.kmeans(fd, k = 3, max.iter = 50, seed = 42)
print(ekm)
```

The built-in plot method shows each cluster's mean curve (red) with member
curves in the background:

```{r plot-ekm}
plot(ekm)
```

### Comparing to True Labels

```{r ekm-confusion}
table(Elastic = ekm$labels, True = true_labels)
```

Elastic k-means recovers the three groups accurately despite the phase
variation, because the elastic distance aligns curves before measuring
dissimilarity.

## Elastic Hierarchical Clustering

`elastic.hclust()` builds an agglomerative dendrogram using pairwise elastic
distances. It supports single, complete, and average linkage. Unlike k-means,
hierarchical clustering does not require choosing k in advance --- you can
cut the dendrogram at any level.

```{r elastic-hclust}
ehc <- elastic.hclust(fd, method = "complete")
print(ehc)
```

```{r plot-dendro}
plot(ehc)
```

The dendrogram reveals the three-group structure as three well-separated
subtrees. Cut the tree with `elastic.cutree()`:

```{r cutree}
hc_labels <- elastic.cutree(ehc, k = 3)
table(Hierarchical = hc_labels, True = true_labels)
```

### Choosing a Linkage Method

| Method | Behavior | Best for |
|--------|----------|----------|
| `complete` | Merges clusters whose farthest members are closest | Compact, similar-size clusters |
| `single` | Merges clusters whose nearest members are closest | Elongated or chain-like clusters |
| `average` | Merges by average pairwise distance | General purpose |

## Choosing k

When the true number of groups is unknown, compare within-cluster distances
across several values of k. A sharp drop followed by a plateau suggests the
right number of clusters.

```{r choose-k}
k_range <- 2:6
within_dist <- numeric(length(k_range))

for (idx in seq_along(k_range)) {
  fit <- elastic.kmeans(fd, k = k_range[idx], max.iter = 50, seed = 42)
  within_dist[idx] <- fit$total.within.distance
}

df_k <- data.frame(k = k_range, total_within = within_dist)
ggplot(df_k, aes(x = .data$k, y = .data$total_within)) +
  geom_line(linewidth = 1, color = "#4A90D9") +
  geom_point(size = 3, color = "#4A90D9") +
  geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
  labs(title = "Elastic K-Means: Choosing k",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Distance") +
  scale_x_continuous(breaks = k_range)
```

The elbow at k = 3 confirms three clusters. Beyond k = 3, splitting
further does not substantially reduce within-cluster distance.

## Comparison: Standard vs Elastic Clustering

To demonstrate the advantage of elastic clustering, we compare it to
standard (Euclidean) k-means on the same data.

```{r standard-kmeans}
# Standard k-means (Euclidean)
km_std <- cluster.kmeans(fd, ncl = 3, metric = "L2", nstart = 10, seed = 42)
table(Standard = km_std$cluster, True = true_labels)
```

```{r side-by-side}
# Build comparison data frame
n <- nrow(X)

df_compare <- data.frame(
  curve_id = rep(rep(seq_len(n), each = m), 2),
  argval = rep(rep(argvals, n), 2),
  value = rep(as.vector(t(X)), 2),
  cluster = factor(c(rep(ekm$labels, each = m),
                     rep(km_std$cluster, each = m))),
  method = factor(rep(c("Elastic K-Means", "Standard K-Means"),
                      each = n * m))
)

ggplot(df_compare, aes(x = .data$argval, y = .data$value,
                       group = .data$curve_id, color = .data$cluster)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~ .data$method) +
  labs(title = "Cluster Assignments: Elastic vs Standard",
       x = "Normalized Stride Cycle", y = "Knee Angle (a.u.)",
       color = "Cluster") +
  theme(legend.position = "bottom")
```

Elastic k-means assigns curves based on their underlying shape, grouping
together curves that have the same pattern regardless of timing. Standard
k-means is confused by the phase shifts and may mix groups or split a single
shape pattern into multiple clusters.

### Accuracy Summary

```{r accuracy}
match_accuracy <- function(predicted, true_labels) {
  # Try all label permutations for k = 3
  perms <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))
  best <- 0
  for (p in perms) {
    matched <- sum(predicted == p[true_labels])
    best <- max(best, matched)
  }
  best / length(true_labels)
}

cat("Elastic k-means accuracy:", match_accuracy(ekm$labels, true_labels), "\n")
cat("Standard k-means accuracy:", match_accuracy(km_std$cluster, true_labels), "\n")
```

## Best Practices

1. **Use elastic clustering when phase variation is present.** If curves are
   already well-aligned, standard clustering is faster and equally effective.
2. **Start with hierarchical clustering** to explore the group structure
   without committing to a specific k.
3. **Compare linkage methods.** Complete linkage tends to produce compact
   clusters; single linkage is better for elongated structures.
4. **Check convergence.** If `elastic.kmeans()` reports `converged = FALSE`,
   increase `max.iter` or `karcher.max.iter`.
5. **Regularization.** Set `lambda > 0` to penalize extreme warpings when
   data is noisy.

## See Also

- `vignette("articles/clustering")` --- standard functional clustering
  (k-means, fuzzy c-means, optimal k selection)
- `vignette("articles/elastic-alignment")` --- elastic curve alignment
  and Karcher means
- `vignette("articles/shape-analysis")` --- shape space analysis
  (orbit representatives, shape distances, shape means)
- `vignette("articles/distance-metrics")` --- distance and semimetric
  measures for functional data

## References

- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data
  Analysis*. Springer.

- Tucker, J.D., Wu, W. and Srivastava, A. (2013). Generative Models for
  Functional Data Using Phase and Amplitude Separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.

- Kurtek, S., Srivastava, A., Klassen, E. and Ding, Z. (2012). Statistical
  Modeling of Curves Using Shapes and Related Features. *Journal of the
  American Statistical Association*, 107(499), 1152--1165.
