---
title: "Elastic Curve Alignment"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

When comparing functional data, observed curves often exhibit both **amplitude
variability** (differences in height/shape) and **phase variability** (differences
in timing/speed). Standard methods like the cross-sectional mean or PCA treat all
variation as amplitude, which can produce blurred averages and inflated variance.

**Elastic alignment** separates these two sources of variability by finding optimal
**warping functions** $\gamma$ that retime each curve before comparison. The
mathematical framework uses the **Square-Root Slope Function (SRSF)** representation,
which makes the Fisher-Rao metric equivalent to the $L^2$ metric -- enabling efficient
dynamic programming alignment.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## Quick Start

```{r quick-start}
set.seed(42)
argvals <- seq(0, 1, length.out = 100)

# Simulate curves with horizontal shifts
n <- 15
shifts <- runif(n, -0.1, 0.1)
data <- matrix(0, n, 100)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * (argvals - shifts[i]))
}
fd <- fdata(data, argvals = argvals)

# Align
res <- elastic.align(fd)
print(res)
```

```{r quick-start-plot}
plot(res, type = "both")
```

## SRSF Transform

The SRSF of a function $f$ is defined as:

$$q(t) = \text{sign}(f'(t)) \sqrt{|f'(t)|}$$

This representation has the key property that the Fisher-Rao distance between two
curves equals the $L^2$ distance between their SRSFs after optimal alignment.

```{r srsf-demo}
q <- srsf.transform(fd)

p1 <- plot(fd) + labs(title = "Original Curves")
p2 <- plot(q) + labs(title = "SRSF Representation")
```

The SRSF transform is invertible: given $q$ and an initial value $f_0$, we can
reconstruct:

$$f(t) = f_0 + \int_0^t q(s)|q(s)|\,ds$$

```{r srsf-inverse}
f_original <- fd$data[1, ]
q_values <- q$data[1, ]
f_reconstructed <- srsf.inverse(q_values, argvals, f_original[1])

max_error <- max(abs(f_original - f_reconstructed))
cat("Round-trip max error:", format(max_error, digits = 4), "\n")
```

## Alignment in Detail

### How It Works

`elastic.align()` aligns all curves to a common target using dynamic programming
to find the optimal warping function $\gamma$ for each curve. The warping functions
satisfy $\gamma(0) = 0$ and $\gamma(1) = 1$ (boundary-preserving) and are monotone
increasing (time never runs backward).

By default, the cross-sectional mean is used as the target:

```{r align-detail}
res <- elastic.align(fd)
```

You can also supply a custom target -- as a numeric vector or an `fdata` object:

```{r align-target}
# Use the deepest (most central) curve as target
depths <- depth(fd, method = "FM")
deepest <- fd[which.max(depths)]
res_deep <- elastic.align(fd, target = deepest)
```

### Inspecting Warping Functions

The warping functions show how each curve's time axis is stretched or compressed
to match the target:

```{r warps}
plot(res, type = "warps")
```

A warping function above the diagonal means that part of the curve has been
sped up (compressed in time); below the diagonal means slowed down (stretched).

## Karcher Mean

The **Karcher mean** (Frechet mean in the elastic metric) computes the mean shape
by iteratively aligning curves to a running estimate of the mean. This produces a
sharper average than the cross-sectional mean because it removes phase variability:

```{r karcher}
km <- karcher.mean(fd, max.iter = 20, tol = 1e-4)
print(km)
```

```{r karcher-compare, fig.height = 4}
cross_mean <- colMeans(fd$data)
karcher <- as.numeric(km$mean$data)

df_means <- data.frame(
  t = rep(argvals, 2),
  value = c(cross_mean, karcher),
  type = rep(c("Cross-sectional mean", "Karcher mean"), each = 100)
)

ggplot(df_means, aes(x = t, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Cross-Sectional vs Karcher Mean",
       x = "t", y = "f(t)", color = NULL)
```

The cross-sectional mean is attenuated by the phase shifts, while the Karcher mean
recovers the true amplitude.

## Elastic Distance Matrix

The elastic distance provides a reparameterization-invariant metric for
clustering, MDS, and other distance-based methods:

```{r distance}
D <- elastic.distance(fd)

# Metric properties
cat("Symmetric:             ", max(abs(D - t(D))) < 1e-10, "\n")
cat("Non-negative:          ", all(D >= -1e-10), "\n")
cat("Zero diagonal:         ", all(abs(diag(D)) < 1e-10), "\n")
```

```{r mds, fig.height = 4}
mds <- cmdscale(D, k = 2)
plot(mds, pch = 19, xlab = "MDS 1", ylab = "MDS 2",
     main = "MDS of Elastic Distances", col = "steelblue", cex = 1.3)
```

The elastic distance is also available through the `metric()` dispatcher:

```{r metric-elastic}
D2 <- metric(fd, method = "elastic")
all.equal(D, D2)
```

## When to Use Elastic Alignment

Elastic alignment works best when curves share the **same shape** but differ
in **timing**. The table below summarizes where it excels and where it does not.

### Works Well

| Scenario | Typical VR | Notes |
|:---------|:----------:|:------|
| Shifted peaks / bumps | 98-100% | Best case: localized features with horizontal shifts |
| ECG-like sharp peaks | 98-100% | Localized timing differences in multi-peak curves |
| Spectral data (shifted peaks) | 95-98% | Multiple peaks with coherent timing shifts |
| Phase-shifted smooth curves | 90-93% | Sinusoidal, polynomial, sigmoid curves with shifts |
| Non-periodic curves | 95-100% | Works well when boundary values are similar |

### Where It Degrades

| Scenario | Typical VR | Explanation |
|:---------|:----------:|:------------|
| Mixed amplitude + phase | 40-65% | Phase correction is partial when amplitude dominates |
| Noisy data (raw) | 0-30% | Noise in derivatives overwhelms SRSF; smooth first |
| Large boundary mismatch | 70-85% | Fixed boundary ($\gamma(0)=0$, $\gamma(1)=1$) can't correct endpoint values |
| Coarse grids (m < 50) | 70-75% | DP resolution limits alignment accuracy |

### Where It Should NOT Be Used

| Scenario | What happens | Recommendation |
|:---------|:-------------|:---------------|
| Pure amplitude differences | Identity warp (no harm, no help) | Use standard methods |
| Rate-varying monotone curves | Spurious warping, VR < 0 | These are amplitude/shape differences, not phase |
| Genuinely different shapes | Forced warping, VR < 0 | Use depth or distance methods instead |
| Very noisy data (unsmoothed) | Noise-driven warping, VR < 0 | Pre-smooth with `fdata2basis()` + `basis2fdata()` |

## Comparison with DTW and Shift Registration

Three alignment strategies are available in **fdars**:

| | Elastic (SRSF) | DTW | Shift Registration |
|:--|:--|:--|:--|
| **Function** | `elastic.align()` | `metric.DTW()` | `register.fd()` |
| **Warping** | Smooth, monotone diffeomorphism | Piecewise constant, allows repeated indices | Global translation only |
| **Metric** | Proper metric (triangle inequality holds) | Not a metric | Not applicable (no warping) |
| **Amplitude/Phase** | Cleanly separates the two | Mixes them — can absorb amplitude differences into warping | Pure phase (shift), amplitude untouched |
| **Degeneracy** | No "pinching" — SRSF penalizes compression | Can collapse entire regions to a single point | No degeneracy, but very limited |
| **Mean** | Karcher mean is well-defined | No principled mean | Shift-corrected cross-sectional mean |
| **Complexity** | $O(m^2)$ per pair (DP) | $O(m^2)$ per pair (DP) | $O(m)$ per curve (cross-correlation) |

### Head-to-Head: Shifted Bumps

The clearest way to see the differences is on data with localized features:

```{r comparison-setup}
set.seed(42)
m_cmp <- 200
argvals_cmp <- seq(0, 1, length.out = m_cmp)
n_cmp <- 15

# Two bumps with varying positions
data_cmp <- matrix(0, n_cmp, m_cmp)
for (i in 1:n_cmp) {
  p1 <- 0.3 + runif(1, -0.06, 0.06)
  p2 <- 0.7 + runif(1, -0.06, 0.06)
  data_cmp[i, ] <- exp(-300 * (argvals_cmp - p1)^2) +
                   exp(-300 * (argvals_cmp - p2)^2)
}
fd_cmp <- fdata(data_cmp, argvals = argvals_cmp)
var_orig_cmp <- mean(apply(fd_cmp$data, 2, var))
```

**Elastic alignment** finds smooth, local warping for each bump independently:

```{r comparison-elastic}
res_elastic <- elastic.align(fd_cmp)
vr_elastic <- 1 - mean(apply(res_elastic$aligned$data, 2, var)) / var_orig_cmp
cat("Elastic VR:", round(vr_elastic * 100, 1), "%\n")
```

**Shift registration** applies a single global shift — it can center one bump
but not both when they move independently:

```{r comparison-shift}
res_shift <- register.fd(fd_cmp)
vr_shift <- 1 - mean(apply(res_shift$registered$data, 2, var)) / var_orig_cmp
cat("Shift registration VR:", round(vr_shift * 100, 1), "%\n")
```

```{r comparison-plot, fig.height = 8}
plot_curves <- function(data, argvals, title) {
  n <- nrow(data)
  m <- length(argvals)
  df <- data.frame(
    curve_id = rep(seq_len(n), each = m),
    argval = rep(argvals, n),
    value = as.vector(t(data))
  )
  ggplot(df, aes(x = argval, y = value, group = curve_id)) +
    geom_line(alpha = 0.5) +
    labs(title = title, x = "t", y = "value") +
    theme_minimal()
}

p1 <- plot_curves(fd_cmp$data, argvals_cmp, "Original")
p2 <- plot_curves(res_elastic$aligned$data, argvals_cmp,
        paste0("Elastic (VR = ", round(vr_elastic * 100, 1), "%)"))
p3 <- plot_curves(res_shift$registered$data, argvals_cmp,
        paste0("Shift Registration (VR = ", round(vr_shift * 100, 1), "%)"))

p1 / p2 / p3
```

Elastic alignment resolves the independent shifts of both bumps. Shift
registration can only correct a single global offset, leaving residual
phase variability when features move independently.

### DTW Distance vs Elastic Distance

DTW and elastic distances rank curves differently because DTW can absorb
amplitude differences into warping while elastic alignment separates them:

```{r comparison-distances}
# Compute both distance matrices
D_elastic <- elastic.distance(fd_cmp)
D_dtw <- metric.DTW(fd_cmp)

# Compare: correlation of off-diagonal entries
idx <- upper.tri(D_elastic)
cat("Correlation between elastic and DTW distances:",
    round(cor(D_elastic[idx], D_dtw[idx]), 3), "\n")
```

```{r comparison-distances-plot, fig.height = 4}
plot(D_elastic[idx], D_dtw[idx], pch = 19, cex = 0.6, col = "steelblue",
     xlab = "Elastic distance", ylab = "DTW distance",
     main = "Elastic vs DTW Pairwise Distances")
abline(lm(D_dtw[idx] ~ D_elastic[idx]), col = "red", lwd = 1.5)
```

The distances are correlated but not identical. The key difference emerges
when amplitude variability is present:

```{r comparison-amplitude, fig.height = 4}
set.seed(42)
# Curves with BOTH amplitude and phase variability
data_amp <- matrix(0, n_cmp, m_cmp)
for (i in 1:n_cmp) {
  amp <- runif(1, 0.5, 1.5)
  shift <- runif(1, -0.06, 0.06)
  data_amp[i, ] <- amp * exp(-200 * (argvals_cmp - 0.5 - shift)^2)
}
fd_amp <- fdata(data_amp, argvals = argvals_cmp)

D_elastic_amp <- elastic.distance(fd_amp)
D_dtw_amp <- metric.DTW(fd_amp)
idx_amp <- upper.tri(D_elastic_amp)

cat("Correlation (amplitude+phase):",
    round(cor(D_elastic_amp[idx_amp], D_dtw_amp[idx_amp]), 3), "\n")
```

With mixed amplitude and phase, DTW distances are less correlated with elastic
distances because DTW warping partially absorbs amplitude differences. Elastic
distances preserve the amplitude/phase separation, making them more suitable
for downstream analysis (clustering, MDS) when both sources of variability
carry distinct meaning.

### When to Use Each Method

| Scenario | Recommended |
|:---------|:------------|
| Localized features with independent timing shifts | Elastic alignment |
| Simple global time shift or delay | Shift registration (`register.fd`) — fast and sufficient |
| Distance computation for clustering/MDS | Elastic distance (proper metric) or DTW (if metric property not needed) |
| Noisy signals, speech-like data | DTW (more robust to noise without pre-smoothing) |
| Periodic/circular data | `elastic.align(periodic = TRUE)` |
| Need principled mean shape | `karcher.mean()` (no DTW equivalent) |

## Visual Examples by Scenario

The following examples illustrate alignment behavior across different data
scenarios. Each shows the **original** curves (left) and **aligned** curves
(right) with variance reduction (VR) reported.

### Works Well: ECG-like Multi-Peak Curves

Sharp, localized features with timing differences are the ideal use case:

```{r gallery-ecg, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.05, 0.05)
  t_shifted <- argvals - shift
  # P wave, QRS complex, T wave
  dm[i, ] <- 0.2 * exp(-200 * (t_shifted - 0.2)^2) -
             0.6 * exp(-500 * (t_shifted - 0.4)^2) +
             1.0 * exp(-800 * (t_shifted - 0.43)^2) -
             0.2 * exp(-500 * (t_shifted - 0.46)^2) +
             0.3 * exp(-150 * (t_shifted - 0.65)^2)
}
fd_ecg <- fdata(dm, argvals = argvals)
res_ecg <- elastic.align(fd_ecg)
vr <- 1 - mean(apply(res_ecg$aligned$data, 2, var)) /
          mean(apply(fd_ecg$data, 2, var))
```

```{r gallery-ecg-plot, fig.height = 4}
plot(res_ecg, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Works Well: Spectral Peaks

Multiple peaks at varying positions, common in spectroscopy and chromatography:

```{r gallery-spectral, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.04, 0.04)
  dm[i, ] <- 0.8 * exp(-300 * (argvals - 0.25 - shift)^2) +
             1.0 * exp(-500 * (argvals - 0.50 - shift * 1.2)^2) +
             0.5 * exp(-300 * (argvals - 0.75 - shift * 0.8)^2)
}
fd_spec <- fdata(dm, argvals = argvals)
res_spec <- elastic.align(fd_spec)
vr <- 1 - mean(apply(res_spec$aligned$data, 2, var)) /
          mean(apply(fd_spec$data, 2, var))
```

```{r gallery-spectral-plot, fig.height = 4}
plot(res_spec, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Works Well: Growth-like Sigmoid Curves

Monotone S-shaped curves with timing shifts (e.g., growth curves with
differing onset times):

```{r gallery-sigmoid, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- 1 / (1 + exp(-20 * (argvals - 0.5 - shift)))
}
fd_sig <- fdata(dm, argvals = argvals)
res_sig <- elastic.align(fd_sig)
vr <- 1 - mean(apply(res_sig$aligned$data, 2, var)) /
          mean(apply(fd_sig$data, 2, var))
```

```{r gallery-sigmoid-plot, fig.height = 4}
plot(res_sig, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

### Medium: Mixed Amplitude and Phase Variability

When curves differ in both height *and* timing, alignment corrects the phase
component but leaves amplitude variability untouched:

```{r gallery-mixed, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 1.5)
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- amp * sin(2 * pi * (argvals - shift))
}
fd_mix <- fdata(dm, argvals = argvals)
res_mix <- elastic.align(fd_mix)
vr <- 1 - mean(apply(res_mix$aligned$data, 2, var)) /
          mean(apply(fd_mix$data, 2, var))
```

```{r gallery-mixed-plot, fig.height = 4}
plot(res_mix, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1), "%"))
```

After alignment, curves still differ in amplitude (vertical spread) but peaks
and valleys are now synchronized. The residual variance is real amplitude
variability, not a failure of alignment.

### Medium: Noisy Data (Raw vs Smoothed)

Noise in derivatives overwhelms the SRSF. Pre-smoothing recovers alignment:

```{r gallery-noise, fig.height = 7}
set.seed(42)
n <- 20
dm_clean <- matrix(0, n, 100)
shifts <- runif(n, -0.1, 0.1)
for (i in 1:n) dm_clean[i, ] <- sin(2 * pi * (argvals - shifts[i]))
dm_noisy <- dm_clean + matrix(rnorm(n * 100, sd = 0.3), n, 100)
fd_noisy <- fdata(dm_noisy, argvals = argvals)

# Align raw noisy data
res_noisy <- elastic.align(fd_noisy)
vr_raw <- 1 - mean(apply(res_noisy$aligned$data, 2, var)) /
              mean(apply(fd_noisy$data, 2, var))

# Smooth first, then align
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_sm <- basis2fdata(coefs, argvals)
res_sm <- elastic.align(fd_sm)
vr_sm <- 1 - mean(apply(res_sm$aligned$data, 2, var)) /
             mean(apply(fd_sm$data, 2, var))
```

```{r gallery-noise-plot, fig.height = 7}
p_raw <- plot(res_noisy, type = "both") +
  plot_annotation(subtitle = paste0("Raw data — VR = ", round(vr_raw * 100, 1), "%"))
p_sm <- plot(res_sm, type = "both") +
  plot_annotation(subtitle = paste0("After smoothing — VR = ", round(vr_sm * 100, 1), "%"))
p_raw / p_sm
```

### Medium: Boundary Mismatch

When shifted curves have different values at the domain boundaries, alignment
is excellent in the interior but constrained at the edges:

```{r gallery-boundary, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
shifts <- runif(n, -0.15, 0.15)
for (i in 1:n) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd_bnd <- fdata(dm, argvals = argvals)
res_bnd <- elastic.align(fd_bnd)
vr <- 1 - mean(apply(res_bnd$aligned$data, 2, var)) /
          mean(apply(fd_bnd$data, 2, var))
```

```{r gallery-boundary-plot, fig.height = 4}
plot(res_bnd, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — note residual spread at boundaries"))
```

### Bad: Pure Amplitude Differences

When curves differ only in height (no phase variability), alignment correctly
produces identity warps — but also provides no benefit:

```{r gallery-amplitude, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 2.0)
  dm[i, ] <- amp * sin(2 * pi * argvals)
}
fd_amp <- fdata(dm, argvals = argvals)
res_amp <- elastic.align(fd_amp)
vr <- 1 - mean(apply(res_amp$aligned$data, 2, var)) /
          mean(apply(fd_amp$data, 2, var))
```

```{r gallery-amplitude-plot, fig.height = 4}
plot(res_amp, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — no phase to correct, identity warps applied"))
```

This is correct behavior: no spurious warping is introduced.

### Bad: Genuinely Different Shapes

Alignment should **not** be used when curves have fundamentally different
shapes. Forced warping produces meaningless results:

```{r gallery-different, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  if (i %% 5 == 1) dm[i, ] <- sin(2 * pi * argvals)
  else if (i %% 5 == 2) dm[i, ] <- cos(6 * pi * argvals)
  else if (i %% 5 == 3) dm[i, ] <- argvals^3
  else if (i %% 5 == 4) dm[i, ] <- sin(10 * pi * argvals) * exp(-3 * argvals)
  else dm[i, ] <- dnorm(argvals, 0.5, 0.05) / dnorm(0.5, 0.5, 0.05)
}
fd_diff <- fdata(dm, argvals = argvals)
res_diff <- elastic.align(fd_diff)
vr <- 1 - mean(apply(res_diff$aligned$data, 2, var)) /
          mean(apply(fd_diff$data, 2, var))
```

```{r gallery-different-plot, fig.height = 4}
plot(res_diff, type = "both") +
  plot_annotation(subtitle = paste0("VR = ", round(vr * 100, 1),
    "% — forced warping of unrelated shapes"))
```

When VR is very low or negative, it indicates the data is not suitable for
elastic alignment. Use depth or distance methods instead.

## Periodic Functional Data

### The Problem

Elastic alignment uses fixed boundary constraints $\gamma(0) = 0$, $\gamma(1) = 1$.
This prevents the warping function from performing circular shifts, which means
periodic data (e.g., curves on $[0, 2\pi]$ where $f(0) = f(2\pi)$) with large
phase offsets cannot be properly aligned.

For example, two copies of $\sin(t)$ shifted by half a period have identical
shape but the boundary constraints force alignment to fail:

```{r periodic-problem, fig.height = 4}
set.seed(42)
m <- 100
argvals_p <- seq(0, 2 * pi, length.out = m)
n <- 15
data_p <- matrix(0, n, m)
for (i in 1:n) {
  shift <- sample(0:(m - 1), 1)
  idx <- ((seq_len(m) - 1L + shift) %% m) + 1L
  data_p[i, ] <- sin(argvals_p[idx])
}
fd_p <- fdata(data_p, argvals = argvals_p)

# Standard alignment struggles with circular shifts
res_std <- elastic.align(fd_p)
var_orig <- mean(apply(fd_p$data, 2, var))
vr_std <- 1 - mean(apply(res_std$aligned$data, 2, var)) / var_orig
```

```{r periodic-problem-plot, fig.height = 4}
plot(res_std, type = "both") +
  plot_annotation(subtitle = paste0(
    "Standard alignment on periodic data — VR = ", round(vr_std * 100, 1), "%"))
```

### The Solution

Setting `periodic = TRUE` applies a two-stage approach:

1. **Circular rotation**: Each curve is circularly shifted so that its global
   maximum is at a canonical grid position
2. **Elastic alignment**: Standard alignment on the rotated data

```{r periodic-solution, fig.height = 4}
res_per <- elastic.align(fd_p, periodic = TRUE)
vr_per <- 1 - mean(apply(res_per$aligned$data, 2, var)) / var_orig
```

```{r periodic-solution-plot, fig.height = 4}
plot(res_per, type = "both") +
  plot_annotation(subtitle = paste0(
    "Periodic alignment — VR = ", round(vr_per * 100, 1), "%"))
```

The rotation step removes large circular shifts that the boundary-constrained
warping cannot handle, and the elastic step corrects any remaining timing
differences.

You can also apply the rotation step manually using `periodic.rotate()`:

```{r periodic-rotate}
rot <- periodic.rotate(fd_p)
cat("Shifts applied:", head(rot$shifts), "...\n")
```

The `karcher.mean()` function also accepts `periodic = TRUE`:

```{r periodic-karcher}
km_per <- karcher.mean(fd_p, max.iter = 15, periodic = TRUE)
print(km_per)
```

## Practical Guide

### Pre-smooth Noisy Data

The SRSF transform involves derivatives, so noise gets amplified. Always smooth
noisy data before alignment:

```{r presmooth}
set.seed(42)
noisy <- matrix(0, 15, 100)
shifts <- runif(15, -0.1, 0.1)
for (i in 1:15) {
  noisy[i, ] <- sin(2 * pi * (argvals - shifts[i])) + rnorm(100, sd = 0.15)
}
fd_noisy <- fdata(noisy, argvals = argvals)

# Smooth with B-spline basis
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_smooth <- basis2fdata(coefs, argvals)

var_before <- mean(apply(fd_noisy$data, 2, var))

# Align raw vs smoothed
res_raw <- elastic.align(fd_noisy)
res_smooth <- elastic.align(fd_smooth)

vr_raw <- 1 - mean(apply(res_raw$aligned$data, 2, var)) / var_before
vr_smooth <- 1 - mean(apply(res_smooth$aligned$data, 2, var)) / var_before
cat("VR (raw):     ", round(vr_raw * 100, 1), "%\n")
cat("VR (smoothed):", round(vr_smooth * 100, 1), "%\n")
```

### Use Sufficient Grid Resolution

The dynamic programming alignment operates on the evaluation grid. Finer grids
yield more accurate warping:

```{r grid-resolution}
set.seed(42)
for (m in c(20, 50, 100, 200)) {
  av <- seq(0, 1, length.out = m)
  shifts <- runif(15, -0.1, 0.1)
  dm <- matrix(0, 15, m)
  for (i in 1:15) dm[i, ] <- sin(2 * pi * (av - shifts[i]))
  fd_m <- fdata(dm, argvals = av)
  ea <- elastic.align(fd_m)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                  mean(apply(fd_m$data, 2, var))
  cat(sprintf("  m = %3d: VR = %5.1f%%\n", m, var_red * 100))
}
```

A minimum of 50-100 grid points is recommended.

### Choose an Appropriate Target

The target curve affects alignment quality. Three options:

| Target | When to use |
|:-------|:------------|
| Cross-sectional mean (default) | General-purpose; robust with many curves |
| Deepest curve | When outliers may distort the mean |
| Known reference | When aligning to a standard template |

```{r target-choice}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)
var_before <- mean(apply(fd$data, 2, var))

# Default (cross-sectional mean)
vr1 <- 1 - mean(apply(elastic.align(fd)$aligned$data, 2, var)) / var_before

# Deepest curve
depths <- depth(fd, method = "FM")
vr2 <- 1 - mean(apply(elastic.align(fd, target = fd[which.max(depths)])$aligned$data, 2, var)) / var_before

cat("Cross-sectional mean: VR =", round(vr1 * 100, 1), "%\n")
cat("Deepest curve:        VR =", round(vr2 * 100, 1), "%\n")
```

### Alignment + FPCA

Alignment before FPCA concentrates variance into the first principal component
by removing phase variability that would otherwise spread across components:

```{r alignment-fpca}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)

# FPCA before and after alignment
pc_before <- fdata2pc(fd, ncomp = 3)
ea <- elastic.align(fd)
pc_after <- fdata2pc(ea$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2], "%\n")
```

## Example: Localized Peak Alignment

Elastic alignment excels at aligning curves with localized features at varying
positions -- the pattern seen in spectroscopy, ECG, and chromatography data:

```{r peaks-example}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  # Two peaks with varying positions
  p1 <- 0.3 + runif(1, -0.05, 0.05)
  p2 <- 0.7 + runif(1, -0.05, 0.05)
  dm[i, ] <- exp(-100 * (argvals - p1)^2) + 0.6 * exp(-100 * (argvals - p2)^2)
}
fd_peaks <- fdata(dm, argvals = argvals)

res_peaks <- elastic.align(fd_peaks)

var_red <- 1 - mean(apply(res_peaks$aligned$data, 2, var)) /
               mean(apply(fd_peaks$data, 2, var))
cat("Variance reduction:", round(var_red * 100, 1), "%\n")
```

```{r peaks-plot}
plot(res_peaks, type = "both")
```

## Example: Shifted Gaussian Bumps

```{r bumps-example}
set.seed(42)
dm <- matrix(0, 20, 100)
shifts <- runif(20, -0.15, 0.15)
for (i in 1:20) {
  dm[i, ] <- exp(-25 * (argvals - 0.5 - shifts[i])^2)
}
fd_bumps <- fdata(dm, argvals = argvals)
res_bumps <- elastic.align(fd_bumps)

var_red <- 1 - mean(apply(res_bumps$aligned$data, 2, var)) /
               mean(apply(fd_bumps$data, 2, var))
cat("Variance reduction:", round(var_red * 100, 1), "%\n")
```

```{r bumps-plot}
plot(res_bumps, type = "both")
```

## Example: Multi-Bump Alignment

Real-world functional data often contains multiple peaks or features that vary
independently. This section explores how elastic alignment handles curves with
**varying inter-bump distance** (peaks shift independently) and **varying bump
width** (peaks broaden or narrow) as the number of bumps increases.

### Two Bumps

With two bumps, the warping function has enough freedom to independently
adjust each peak's position and width:

```{r multi-bump-2, fig.height = 8}
set.seed(42)
m <- 200
argvals_mb <- seq(0, 1, length.out = m)
n <- 20

# Varying inter-bump distance
data_pos <- matrix(0, n, m)
for (i in 1:n) {
  p1 <- 0.3
  p2 <- 0.7 + runif(1, -0.12, 0.12)
  data_pos[i, ] <- exp(-300 * (argvals_mb - p1)^2) +
                   exp(-300 * (argvals_mb - p2)^2)
}
fd_pos <- fdata(data_pos, argvals = argvals_mb)
res_pos <- elastic.align(fd_pos)
vr_pos <- 1 - mean(apply(res_pos$aligned$data, 2, var)) /
              mean(apply(fd_pos$data, 2, var))

# Varying bump width
data_w <- matrix(0, n, m)
for (i in 1:n) {
  w1 <- runif(1, 200, 500)
  w2 <- runif(1, 200, 500)
  data_w[i, ] <- exp(-w1 * (argvals_mb - 0.3)^2) +
                 exp(-w2 * (argvals_mb - 0.7)^2)
}
fd_w <- fdata(data_w, argvals = argvals_mb)
res_w <- elastic.align(fd_w)
vr_w <- 1 - mean(apply(res_w$aligned$data, 2, var)) /
             mean(apply(fd_w$data, 2, var))

cat("2 bumps — varying position: VR =", round(vr_pos * 100, 1), "%\n")
cat("2 bumps — varying width:    VR =", round(vr_w * 100, 1), "%\n")
```

```{r multi-bump-2-plot, fig.height = 8}
p1 <- plot(res_pos, type = "both") +
  plot_annotation(subtitle = paste0("Varying inter-bump distance — VR = ",
                                    round(vr_pos * 100, 1), "%"))
p2 <- plot(res_w, type = "both") +
  plot_annotation(subtitle = paste0("Varying bump width — VR = ",
                                    round(vr_w * 100, 1), "%"))
p1 / p2
```

Both cases achieve near-perfect alignment. The warping function locally
stretches/compresses the region around each bump to bring peaks into
correspondence.

### Scaling to More Bumps

As the number of bumps increases, alignment becomes progressively harder.
A monotone warping function that stretches one region must compress another,
so with many tightly packed bumps the degrees of freedom become insufficient
to independently correct each peak.

```{r multi-bump-scaling}
set.seed(42)
results <- data.frame(
  bumps = integer(), scenario = character(), VR = numeric(),
  stringsAsFactors = FALSE
)

for (nb in c(2, 4, 5, 7)) {
  positions_base <- seq(0.1, 0.9, length.out = nb)

  # Varying positions
  data_pos <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      p <- positions_base[j] + runif(1, -0.04, 0.04)
      data_pos[i, ] <- data_pos[i, ] + exp(-300 * (argvals_mb - p)^2)
    }
  }
  fd_pos <- fdata(data_pos, argvals = argvals_mb)
  res_pos <- elastic.align(fd_pos)
  vr_pos <- 1 - mean(apply(res_pos$aligned$data, 2, var)) /
                mean(apply(fd_pos$data, 2, var))

  # Varying widths
  data_w <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      w <- runif(1, 200, 600)
      data_w[i, ] <- data_w[i, ] + exp(-w * (argvals_mb - positions_base[j])^2)
    }
  }
  fd_w <- fdata(data_w, argvals = argvals_mb)
  res_w <- elastic.align(fd_w)
  vr_w <- 1 - mean(apply(res_w$aligned$data, 2, var)) /
               mean(apply(fd_w$data, 2, var))

  # Both varying
  data_both <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      p <- positions_base[j] + runif(1, -0.04, 0.04)
      w <- runif(1, 200, 600)
      data_both[i, ] <- data_both[i, ] + exp(-w * (argvals_mb - p)^2)
    }
  }
  fd_both <- fdata(data_both, argvals = argvals_mb)
  res_both <- elastic.align(fd_both)
  vr_both <- 1 - mean(apply(res_both$aligned$data, 2, var)) /
                  mean(apply(fd_both$data, 2, var))

  results <- rbind(results, data.frame(
    bumps = rep(nb, 3),
    scenario = c("Position", "Width", "Both"),
    VR = c(vr_pos, vr_w, vr_both) * 100
  ))
}
```

```{r multi-bump-table}
# Display as a table
df_wide <- reshape(results, idvar = "bumps", timevar = "scenario",
                   direction = "wide")
names(df_wide) <- gsub("VR\\.", "", names(df_wide))
for (col in c("Position", "Width", "Both")) {
  df_wide[[col]] <- paste0(round(df_wide[[col]], 1), "%")
}
knitr::kable(df_wide, col.names = c("Bumps", "Varying Position",
             "Varying Width", "Both Varying"),
             align = "cccc",
             caption = "Variance reduction by number of bumps and variability type")
```

### Visual Comparison: 4 vs 7 Bumps

```{r multi-bump-4v7, fig.height = 8}
set.seed(42)

# 4 bumps — both varying
pos4 <- seq(0.1, 0.9, length.out = 4)
data4 <- matrix(0, n, m)
for (i in 1:n) {
  for (j in 1:4) {
    p <- pos4[j] + runif(1, -0.04, 0.04)
    w <- runif(1, 200, 600)
    data4[i, ] <- data4[i, ] + exp(-w * (argvals_mb - p)^2)
  }
}
fd4 <- fdata(data4, argvals = argvals_mb)
res4 <- elastic.align(fd4)
vr4 <- 1 - mean(apply(res4$aligned$data, 2, var)) /
            mean(apply(fd4$data, 2, var))

# 7 bumps — both varying
pos7 <- seq(0.1, 0.9, length.out = 7)
data7 <- matrix(0, n, m)
for (i in 1:n) {
  for (j in 1:7) {
    p <- pos7[j] + runif(1, -0.04, 0.04)
    w <- runif(1, 200, 600)
    data7[i, ] <- data7[i, ] + exp(-w * (argvals_mb - p)^2)
  }
}
fd7 <- fdata(data7, argvals = argvals_mb)
res7 <- elastic.align(fd7)
vr7 <- 1 - mean(apply(res7$aligned$data, 2, var)) /
            mean(apply(fd7$data, 2, var))
```

```{r multi-bump-4v7-plot, fig.height = 8}
p4 <- plot(res4, type = "both") +
  plot_annotation(subtitle = paste0("4 bumps (both varying) — VR = ",
                                    round(vr4 * 100, 1), "%"))
p7 <- plot(res7, type = "both") +
  plot_annotation(subtitle = paste0("7 bumps (both varying) — VR = ",
                                    round(vr7 * 100, 1), "%"))
p4 / p7
```

With 4 bumps, alignment still captures most of the variability. With 7
tightly packed bumps, the monotonicity constraint on the warping function
limits independent correction of each peak — stretching one inter-peak
region forces compression elsewhere. This is a fundamental limitation of
elastic alignment, not a numerical issue: a single monotone time-warping
cannot independently retime many closely spaced features.

## Real Data: Berkeley Growth Velocity

The **Berkeley Growth Study** recorded the heights of 93 children (39 boys,
54 girls) from ages 1 to 18. The growth *velocity* curves (cm/year) exhibit
clear phase variability: each child's pubertal growth spurt occurs at a
different age, creating horizontal shifts in the peak.

This dataset is a standard benchmark in elastic FDA literature (Tucker et al.,
2013; Srivastava and Klassen, 2016). It is available in the `fdasrvf` package.

```{r growth-data}
data(growth_vel, package = "fdasrvf")

# growth_vel$f is 69 (grid) x 93 (curves) — transpose for fdata
fd_growth <- fdata(t(growth_vel$f), argvals = growth_vel$time)
cat("Curves:", nrow(fd_growth$data), "\n")
cat("Grid points:", ncol(fd_growth$data), "\n")
cat("Age range:", range(fd_growth$argvals), "\n")
```

### Full Age Range

```{r growth-full, fig.height = 5}
plot(fd_growth) + labs(title = "Berkeley Growth Velocity (all 93 children)",
                       x = "Age (years)", y = "Growth velocity (cm/year)")
```

The dominant feature at age 1 (infant growth) and the pubertal spurt around
ages 11--14 are both visible, but the pubertal peak is blurred by phase
variability.

```{r growth-full-align}
res_growth <- elastic.align(fd_growth)
var_orig_growth <- mean(apply(fd_growth$data, 2, var))
vr_growth <- 1 - mean(apply(res_growth$aligned$data, 2, var)) / var_orig_growth
cat("Full range VR:", round(vr_growth * 100, 1), "%\n")
```

```{r growth-full-plot, fig.height = 4}
plot(res_growth, type = "both")
```

### Pubertal Growth Spurt (Ages 8--18)

The pubertal region is where phase variability is strongest. Restricting to
ages 8--18 isolates the growth spurt:

```{r growth-pubertal}
idx_pub <- which(fd_growth$argvals >= 8)
fd_pub <- fdata(fd_growth$data[, idx_pub],
                argvals = fd_growth$argvals[idx_pub])

res_pub <- elastic.align(fd_pub)
var_pub <- mean(apply(fd_pub$data, 2, var))
vr_pub <- 1 - mean(apply(res_pub$aligned$data, 2, var)) / var_pub
cat("Pubertal region VR:", round(vr_pub * 100, 1), "%\n")
```

```{r growth-pubertal-plot, fig.height = 4}
plot(res_pub, type = "both") +
  plot_annotation(subtitle = paste0("Ages 8–18 — VR = ",
                                    round(vr_pub * 100, 1), "%"))
```

### Cross-Sectional Mean vs Karcher Mean

The cross-sectional mean is attenuated because children's growth spurts are
averaged at mismatched ages. The Karcher mean recovers the true peak amplitude:

```{r growth-karcher}
km_pub <- karcher.mean(fd_pub, max.iter = 20)
cross_mean_pub <- colMeans(fd_pub$data)
karcher_pub <- as.numeric(km_pub$mean$data)

cat("Cross-sectional mean peak:", round(max(cross_mean_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(cross_mean_pub)], 1), "\n")
cat("Karcher mean peak:        ", round(max(karcher_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(karcher_pub)], 1), "\n")
```

```{r growth-karcher-plot, fig.height = 5}
df_means <- data.frame(
  age = rep(fd_pub$argvals, 2),
  velocity = c(cross_mean_pub, karcher_pub),
  type = rep(c("Cross-sectional mean", "Karcher mean"),
             each = length(fd_pub$argvals))
)

ggplot(df_means, aes(x = age, y = velocity, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Pubertal Growth Spurt: Cross-Sectional vs Karcher Mean",
       x = "Age (years)", y = "Growth velocity (cm/year)", color = NULL)
```

The Karcher mean is sharper and taller because it aligns the growth spurts
before averaging, rather than smearing them across different ages.

### FPCA Before and After Alignment

Alignment concentrates variance into fewer principal components:

```{r growth-fpca}
pc_before <- fdata2pc(fd_pub, ncomp = 3)
pc_after <- fdata2pc(km_pub$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2],
    "%, PC3 =", pve_before[3], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2],
    "%, PC3 =", pve_after[3], "%\n")
```

Before alignment, phase variability spreads across multiple PCs. After
alignment, the remaining (amplitude) variability is more efficiently captured
by the first PC.

### Warping Functions Reveal Growth Timing

The estimated warping functions show how each child's biological clock
differs from the average:

```{r growth-warps, fig.height = 4}
plot(res_pub, type = "warps") +
  labs(title = "Warping Functions (Ages 8–18)",
       x = "Chronological age", y = "Biological age")
```

Curves above the diagonal correspond to children whose growth spurt occurred
earlier than average (biological clock runs ahead); curves below the diagonal
indicate later-maturing children.

## Karcher Mean with Custom Settings

For complex data, increase `max.iter` and check the aligned curves:

```{r karcher-custom}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- exp(-25 * (argvals - 0.5 - shift)^2)
}
fd <- fdata(dm, argvals = argvals)

km <- karcher.mean(fd, max.iter = 30, tol = 1e-4)
cat("Converged:", km$converged, "\n")
cat("Iterations:", km$n.iter, "\n")
```

```{r karcher-custom-plot}
plot(km, type = "mean")
```

## Limitations

### Fixed Boundary Conditions

Warping functions are constrained to $\gamma(0) = 0$ and $\gamma(1) = 1$.
This means the alignment cannot correct mismatches at the domain boundaries.
For shifted curves, the residual at the boundary equals the value difference:

```{r boundary}
f1 <- sin(2 * pi * argvals)
f2 <- sin(2 * pi * (argvals - 0.1))

fd2 <- fdata(rbind(f1, f2), argvals = argvals)
ea <- elastic.align(fd2)

# Boundary residual = |sin(0) - sin(-0.2*pi)| = 0.588
cat("Boundary gap:          ", round(abs(f2[1] - f1[1]), 3), "\n")
cat("Max aligned difference:", round(max(abs(ea$aligned$data[1, ] - ea$aligned$data[2, ])), 3), "\n")
cat("Interior max diff:     ", round(max(abs(ea$aligned$data[1, 10:90] - ea$aligned$data[2, 10:90])), 3), "\n")
```

The interior alignment is excellent even when boundaries cannot be corrected.

### Amplitude-Only Differences

When curves differ only in amplitude (e.g., $f$ vs $2f$), alignment correctly
produces identity warping -- no spurious time distortion:

```{r amplitude-only}
f1 <- sin(2 * pi * argvals)
f2 <- 2 * sin(2 * pi * argvals)
fd_amp <- fdata(rbind(f1, f2), argvals = argvals)
res_amp <- elastic.align(fd_amp)

max_warp_dev <- max(abs(res_amp$gammas$data -
                        matrix(argvals, 2, 100, byrow = TRUE)))
cat("Max warp deviation from identity:", max_warp_dev, "\n")
```

This is the correct behavior: amplitude differences should not produce warping.
However, when both amplitude and phase variability are present, the phase correction
may be partial -- alignment focuses on phase and leaves amplitude untouched.

### Domain Independence

The alignment works correctly on any domain, not just $[0, 1]$:

```{r domains}
for (domain in list(c(0, 1), c(0, 2*pi), c(0, 10), c(-1, 1))) {
  av <- seq(domain[1], domain[2], length.out = 100)
  range <- domain[2] - domain[1]
  dm <- matrix(0, 10, 100)
  set.seed(42)
  for (i in 1:10) {
    shift <- runif(1, -0.1 * range, 0.1 * range)
    dm[i, ] <- sin(2 * pi * (av - shift) / range)
  }
  fd <- fdata(dm, argvals = av)
  ea <- elastic.align(fd)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                 mean(apply(fd$data, 2, var))
  cat(sprintf("  [%g, %g]: VR = %.1f%%\n", domain[1], domain[2], var_red * 100))
}
```

## References

- Srivastava, A., Klassen, E., Joshi, S.H., and Jermyn, I.H. (2011). Shape analysis of elastic curves in Euclidean spaces. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 33(7):1415--1428.
- Tucker, J.D., Wu, W., and Srivastava, A. (2013). Generative models for functional data using phase and amplitude separation. *Computational Statistics & Data Analysis*, 61:50--66.
- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data Analysis*. Springer.
