---
title: "Elastic Curve Alignment"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

When comparing functional data, observed curves often exhibit both **amplitude
variability** (differences in height/shape) and **phase variability** (differences
in timing/speed). Standard methods like the cross-sectional mean or PCA treat all
variation as amplitude, which can produce blurred averages and inflated variance.

**Elastic alignment** separates these two sources of variability by finding optimal
**warping functions** $\gamma$ that retime each curve before comparison. The
mathematical framework uses the **Square-Root Slope Function (SRSF)** representation,
which makes the Fisher-Rao metric equivalent to the $L^2$ metric -- enabling efficient
dynamic programming alignment.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## Quick Start

```{r quick-start}
set.seed(42)
argvals <- seq(0, 1, length.out = 100)

# Simulate curves with horizontal shifts
n <- 15
shifts <- runif(n, -0.1, 0.1)
data <- matrix(0, n, 100)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * (argvals - shifts[i]))
}
fd <- fdata(data, argvals = argvals)

# Align to cross-sectional mean (default)
res <- elastic.align(fd)
print(res)
```

```{r quick-start-plot}
plot(res, type = "both")
```

You can also supply a custom target -- as a numeric vector or an `fdata` object:

```{r align-target}
# Use the deepest (most central) curve as target
depths <- depth(fd, method = "FM")
res_deep <- elastic.align(fd, target = fd[which.max(depths)])
```

### Inspecting Warping Functions

The warping functions $\gamma_i$ are the key output of elastic alignment. Each
$\gamma_i(t)$ is a monotonically increasing function from $[0, 1]$ to $[0, 1]$
that maps the **reference time axis** to the **original curve's time axis**.
The aligned curve is then $f_i(\gamma_i(t))$ -- time is reparameterised so that
features line up with the target.

```{r warps}
plot(res, type = "warps")
```

#### How to Read a Warping Function

The **diagonal** ($\gamma(t) = t$) represents the identity warp -- no time
change needed. Deviations from the diagonal tell you how the curve's timing
differs from the reference:

- $\gamma_i(t) < t$ (below diagonal): at reference time $t$, the
  corresponding feature in curve $i$ occurs at an **earlier** time. The
  alignment must look backward to find the matching feature.
- $\gamma_i(t) > t$ (above diagonal): the feature occurs **later** in curve
  $i$. The alignment must look forward.

The **slope** of $\gamma_i$ encodes local speed:

- $\dot\gamma_i(t) > 1$ (steeper than diagonal): this region of the original
  curve is **stretched** -- the alignment slows it down to spread features over
  more reference time.
- $\dot\gamma_i(t) < 1$ (shallower): this region is **compressed** -- the
  alignment speeds it up.

Let's visualise this for a single curve to make the interpretation concrete:

```{r warp-anatomy, fig.height = 12}
# Pick the curve with the largest shift
i_max <- which.max(abs(shifts))
gamma_i <- as.numeric(res$gammas$data[i_max, ])

# Panel 1: the warping function with annotations
df_warp <- data.frame(t = argvals, gamma = gamma_i)
df_diag <- data.frame(t = c(0, 1), gamma = c(0, 1))

p_warp <- ggplot(df_warp, aes(x = t, y = gamma)) +
  geom_line(data = df_diag, linetype = "dashed", color = "grey60") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_ribbon(aes(ymin = pmin(gamma, t), ymax = pmax(gamma, t)),
              fill = "steelblue", alpha = 0.15) +
  labs(title = paste0("Warping function for curve ", i_max,
                      " (shift = ", round(shifts[i_max], 3), ")"),
       x = "Reference time t", y = expression(gamma(t))) +
  coord_equal()

# Panel 2: original curve, aligned curve, and target
f_orig <- as.numeric(fd$data[i_max, ])
f_aligned <- as.numeric(res$aligned$data[i_max, ])
f_target <- as.numeric(res$target)

df_curves <- data.frame(
  t = rep(argvals, 3),
  value = c(f_orig, f_aligned, f_target),
  curve = rep(c("Original", "Aligned", "Target"), each = length(argvals))
)

p_curves <- ggplot(df_curves, aes(x = t, y = value, color = curve,
                                  linetype = curve)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c(Original = "coral", Aligned = "steelblue",
                                Target = "grey30")) +
  scale_linetype_manual(values = c(Original = "solid", Aligned = "solid",
                                   Target = "dashed")) +
  labs(title = "Effect of warping on the curve",
       x = "t", y = "value", color = NULL, linetype = NULL) +
  theme(legend.position = "top")

# Panel 3: local speed via deriv()
warp_speed <- deriv(res$gammas)
speed_i <- as.numeric(warp_speed$data[i_max, ])
df_speed <- data.frame(t = warp_speed$argvals, speed = speed_i)

p_speed <- ggplot(df_speed, aes(x = t, y = speed)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60") +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_ribbon(aes(ymin = 1, ymax = speed), fill = "steelblue", alpha = 0.15) +
  labs(title = "Local warping speed: deriv(res$gammas)",
       x = "t",
       y = expression(dot(gamma)(t))) +
  annotate("text", x = 0.05, y = max(speed_i) * 0.95, label = "stretched",
           hjust = 0, size = 3, color = "grey40") +
  annotate("text", x = 0.05, y = min(speed_i) * 1.05, label = "compressed",
           hjust = 0, size = 3, color = "grey40")

p_warp / p_curves / p_speed
```

The three panels show the full picture for a single curve:

1. **Top**: the warping function $\gamma_i(t)$ -- the shaded area shows
   deviation from the identity. This curve had a positive shift, so
   $\gamma_i(t) > t$ over most of the domain (the alignment needs to look
   forward to find matching features).
2. **Middle**: the original curve (coral), the target (dashed), and the
   aligned result $f_i(\gamma_i(t))$ (blue). After warping, the aligned curve
   matches the target closely.
3. **Bottom**: the local warping speed $\dot\gamma_i(t)$, computed via
   `deriv(res$gammas)`. Since the warping functions are `fdata` objects,
   `deriv()` returns an `fdata` of derivatives. Where $\dot\gamma > 1$, time
   is being stretched; where $\dot\gamma < 1$, time is compressed. The regions
   of fastest change correspond to where the original curve's features were
   most offset from the target.

## How It Works (Intuition)

Imagine you record the heartbeat signal of several patients. Each heartbeat has
the same features -- a P wave, a QRS complex, a T wave -- but the timing differs
from patient to patient. If you simply average the raw signals, the peaks blur
out because they don't line up.

Elastic alignment fixes this by **stretching and compressing the time axis** of
each curve until corresponding features line up. The key insight is to work with
the **Square-Root Slope Function (SRSF)** -- a transformed version of the curve
that makes the math tractable. In SRSF space, finding the best alignment
reduces to a standard dynamic programming problem that can be solved efficiently.

After alignment, variability splits cleanly into two parts:

- **Amplitude variability**: genuine differences in curve shape (how tall the
  peaks are, how deep the valleys)
- **Phase variability**: differences in timing (when the peaks occur)

The **Karcher mean** is the average curve shape, computed after alignment. Unlike
the ordinary mean, it preserves sharp features because the peaks are aligned
before averaging.

## Mathematical Framework

### The Space of Warping Functions

Let $\mathcal{F}$ denote the space of absolutely continuous functions
$f : [0, 1] \to \mathbb{R}$. The **warping group** is the set of
orientation-preserving diffeomorphisms of $[0, 1]$:

$$\Gamma = \{\gamma : [0, 1] \to [0, 1] \mid \gamma(0) = 0,\; \gamma(1) = 1,\; \dot\gamma > 0\}$$

$\Gamma$ acts on $\mathcal{F}$ by composition: for $f \in \mathcal{F}$ and
$\gamma \in \Gamma$, the warped function is $(f \circ \gamma)(t) = f(\gamma(t))$.
This group action reparameterizes the curve without changing its shape -- only the
speed at which the curve is traversed changes.

### The Fisher-Rao Metric

The **Fisher-Rao metric** on $\mathcal{F}$ is the unique Riemannian metric (up to
scaling) that is invariant under the action of $\Gamma$. For two tangent vectors
$\delta f_1, \delta f_2 \in T_f \mathcal{F}$, it is defined as:

$$\langle \delta f_1, \delta f_2 \rangle_{FR} = \int_0^1 \frac{\delta f_1'(t) \, \delta f_2'(t)}{|f'(t)|} \, dt$$

This metric is difficult to compute directly because it depends on the derivative
of $f$ in the denominator. The SRSF transform resolves this by isometrically
mapping the Fisher-Rao geometry to a flat $L^2$ space.

### SRSF Transform

The **Square-Root Slope Function** of $f$ is defined as:

$$q(t) = \text{sign}(f'(t)) \sqrt{|f'(t)|}$$

This transform has three key properties:

1. **Isometry**: The Fisher-Rao distance between $f_1$ and $f_2$ equals the
   $L^2$ distance between their SRSFs $q_1$ and $q_2$ (after optimal alignment)
2. **Equivariance**: Under the warping action $f \mapsto f \circ \gamma$, the
   SRSF transforms as $q \mapsto (q \circ \gamma) \sqrt{\dot\gamma}$. This
   is simply the standard action of $\Gamma$ on $L^2$
3. **Invertibility**: Given $q$ and an initial value $f_0 = f(0)$, the original
   function is recovered by:

$$f(t) = f_0 + \int_0^t q(s)|q(s)|\,ds$$

```{r srsf-demo}
q <- srsf.transform(fd)
```

```{r srsf-plot}
p1 <- plot(fd) + labs(title = "Original Curves")
p2 <- plot(q) + labs(title = "SRSF Representation")
p1 + p2
```

```{r srsf-inverse}
f_original <- fd$data[1, ]
q_values <- q$data[1, ]
f_reconstructed <- srsf.inverse(q_values, argvals, f_original[1])

max_error <- max(abs(f_original - f_reconstructed))
cat("Round-trip max error:", format(max_error, digits = 4), "\n")
```

### Alignment as Optimization

Given two functions $f_1, f_2 \in \mathcal{F}$ with SRSFs $q_1, q_2$, the
**optimal alignment** finds:

$$\gamma^* = \arg\min_{\gamma \in \Gamma} \| q_1 - (q_2 \circ \gamma) \sqrt{\dot\gamma} \|_{L^2}$$

This is solved via dynamic programming in $O(m^2)$ time, where $m$ is the
number of grid points. The aligned function is $f_2 \circ \gamma^*$, and the
**elastic distance** between $f_1$ and $f_2$ is:

$$d_e(f_1, f_2) = \| q_1 - (q_2 \circ \gamma^*) \sqrt{\dot\gamma^*} \|_{L^2}$$

This distance satisfies all metric axioms (non-negativity, symmetry, triangle
inequality) and is invariant to reparameterization: $d_e(f_1, f_2) = d_e(f_1 \circ \gamma, f_2 \circ \gamma)$
for any $\gamma \in \Gamma$.

### Amplitude-Phase Decomposition

Elastic alignment provides a principled decomposition of total variability.
For a sample $f_1, \ldots, f_n$ with aligned versions $\tilde{f}_i = f_i \circ \gamma_i^*$:

- **Amplitude variability**: The residual variance in $\tilde{f}_1, \ldots, \tilde{f}_n$ after alignment. This captures genuine differences in curve shape (height, depth of features)
- **Phase variability**: The variability in the warping functions $\gamma_1^*, \ldots, \gamma_n^*$ themselves. This captures differences in timing (when features occur)

The **variance reduction** (VR) quantifies the proportion of total variance
attributable to phase:

$$\text{VR} = 1 - \frac{\text{Var}(\tilde{f}_1, \ldots, \tilde{f}_n)}{\text{Var}(f_1, \ldots, f_n)}$$

where $\text{Var}$ denotes the mean pointwise variance
$\frac{1}{T} \sum_{t} \text{Var}_i(f_i(t))$. A VR close to 1 indicates that
most variability was phase (and has been removed); a VR near 0 indicates the
variability is primarily amplitude.

### Karcher Mean

The **Karcher mean** (Frechet mean in the elastic metric) is the function
$\mu^*$ that minimizes the sum of squared elastic distances:

$$\mu^* = \arg\min_{\mu \in \mathcal{F}} \sum_{i=1}^n d_e(\mu, f_i)^2$$

Since $d_e$ accounts for reparameterization, $\mu^*$ is a **shape-preserving
average**: it recovers the common shape without blurring from phase variability.
The algorithm iterates:

1. Initialize $\hat\mu = \bar{f}$ (cross-sectional mean)
2. Align all curves to $\hat\mu$: $\tilde{f}_i = f_i \circ \gamma_i^*$
3. Update $\hat\mu \leftarrow \frac{1}{n} \sum_{i=1}^n \tilde{f}_i$
4. Repeat until convergence ($\|\hat\mu^{(k+1)} - \hat\mu^{(k)}\|_{L^2} < \varepsilon$)

In practice, convergence is guaranteed to a local minimum and typically occurs in
5--20 iterations.

## Core Functions

### Karcher Mean

```{r karcher}
km <- karcher.mean(fd, max.iter = 20, tol = 1e-4)
print(km)
```

```{r karcher-compare, fig.height = 4}
cross_mean <- as.numeric(mean(fd)$data)
karcher <- as.numeric(km$mean$data)

df_means <- data.frame(
  t = rep(argvals, 2),
  value = c(cross_mean, karcher),
  type = rep(c("Cross-sectional mean", "Karcher mean"), each = 100)
)

ggplot(df_means, aes(x = t, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Cross-Sectional vs Karcher Mean",
       x = "t", y = "f(t)", color = NULL)
```

The cross-sectional mean is attenuated by the phase shifts, while the Karcher mean
recovers the true amplitude.

For complex data, increase `max.iter` and check convergence:

```{r karcher-custom}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- exp(-25 * (argvals - 0.5 - shift)^2)
}
fd_bumps <- fdata(dm, argvals = argvals)

km <- karcher.mean(fd_bumps, max.iter = 30, tol = 1e-4)
cat("Converged:", km$converged, "\n")
cat("Iterations:", km$n.iter, "\n")
```

```{r karcher-custom-plot}
plot(km, type = "mean")
```

### Elastic Distance Matrix

The elastic distance $d_e$ provides a reparameterization-invariant metric for
clustering, MDS, and other distance-based methods:

```{r distance}
D <- elastic.distance(fd)

# Metric properties
cat("Symmetric:             ", max(abs(D - t(D))) < 1e-10, "\n")
cat("Non-negative:          ", all(D >= -1e-10), "\n")
cat("Zero diagonal:         ", all(abs(diag(D)) < 1e-10), "\n")
```

```{r mds, fig.height = 4}
mds <- cmdscale(D, k = 2)
plot(mds, pch = 19, xlab = "MDS 1", ylab = "MDS 2",
     main = "MDS of Elastic Distances", col = "steelblue", cex = 1.3)
```

The elastic distance is also available through the `metric()` dispatcher:

```{r metric-elastic}
D2 <- metric(fd, method = "elastic")
all.equal(D, D2)
```

### Amplitude and Phase Distances

The elastic framework uniquely decomposes the total distance between two curves
into **amplitude** (shape) and **phase** (timing) components:

```{r amp-phase-dist}
D_amp <- amplitude.distance(fd, lambda = 0)
D_phase <- phase.distance(fd, lambda = 0)

cat("Amplitude distance (curves 1-2):", round(as.matrix(D_amp)[1, 2], 3), "\n")
cat("Phase distance (curves 1-2):    ", round(as.matrix(D_phase)[1, 2], 3), "\n")
cat("Elastic distance (curves 1-2):  ", round(D[1, 2], 3), "\n")
```

For these phase-shifted sine curves, most of the distance comes from phase
(timing differences), with very little amplitude distance (the shapes are
similar after alignment).

## Alignment Quality Diagnostics

After alignment, we can quantify exactly what changed. First, compute the
Karcher mean and alignment quality:

```{r quality}
km_diag <- karcher.mean(fd, max.iter = 20, tol = 1e-4)
aq <- alignment.quality(fd, km_diag)
print(aq)
```

### Before vs After: What Changed?

To see the effect of alignment concretely, we compare key metrics computed on
the original and aligned curves:

```{r before-after-table}
# Pairwise L2 distances
D_before <- as.matrix(metric.lp(fd))
D_after <- as.matrix(metric.lp(km_diag$aligned))
mean_d_before <- mean(D_before[upper.tri(D_before)])
mean_d_after <- mean(D_after[upper.tri(D_after)])

# Pointwise variance
var_before <- apply(fd$data, 2, var)
var_after <- apply(km_diag$aligned$data, 2, var)
mean_var_before <- mean(var_before)
mean_var_after <- mean(var_after)

# Format as table
pct <- function(b, a) paste0(round((b - a) / b * 100, 1), "%")
tab <- data.frame(
  Metric = c("Mean pairwise L\u00b2 distance",
             "Mean pointwise variance",
             "Mean cross-sectional SD"),
  Before = round(c(mean_d_before, mean_var_before, mean(sqrt(var_before))), 4),
  After = round(c(mean_d_after, mean_var_after, mean(sqrt(var_after))), 4),
  Reduction = c(pct(mean_d_before, mean_d_after),
                pct(mean_var_before, mean_var_after),
                pct(mean(sqrt(var_before)), mean(sqrt(var_after))))
)
knitr::kable(tab, align = "lrrr")
```

### Pointwise Variance: Where Did Alignment Help?

The pointwise variance at each time point shows *where* alignment had the
greatest effect. The shaded region highlights the variance removed:

```{r pointwise-var-plot, fig.height = 4}
df_var <- data.frame(
  t = rep(argvals, 2),
  variance = c(var_before, var_after),
  stage = rep(c("Before", "After"), each = length(argvals))
)

ggplot() +
  geom_ribbon(
    data = data.frame(t = argvals, ymin = var_after, ymax = var_before),
    aes(x = t, ymin = ymin, ymax = ymax),
    fill = "steelblue", alpha = 0.2
  ) +
  geom_line(data = df_var,
            aes(x = t, y = variance, color = stage, linetype = stage),
            linewidth = 0.8) +
  scale_color_manual(values = c(Before = "grey40", After = "steelblue")) +
  scale_linetype_manual(values = c(Before = "dashed", After = "solid")) +
  labs(title = "Pointwise Variance: Before vs After Alignment",
       x = "t", y = "Variance", color = NULL, linetype = NULL) +
  theme(legend.position = "top")
```

### Mean Curve: How Does the Average Sharpen?

The cross-sectional mean of the original curves is blurred by timing
differences. The Karcher mean, computed after alignment, recovers a sharper
representative shape:

```{r mean-comparison-plot, fig.height = 4}
n <- nrow(fd$data)
m <- ncol(fd$data)

# Original curves + cross-sectional mean
df_orig <- data.frame(
  t = rep(argvals, n), value = as.vector(t(fd$data)),
  curve = rep(seq_len(n), each = m)
)
cross_mean <- as.numeric(mean(fd)$data)
df_cmean <- data.frame(t = argvals, value = cross_mean)

p1 <- ggplot() +
  geom_line(data = df_orig, aes(x = t, y = value, group = curve),
            alpha = 0.15, color = "grey50") +
  geom_line(data = df_cmean, aes(x = t, y = value),
            color = "coral", linewidth = 1.2) +
  labs(title = "Before: Cross-Sectional Mean", x = "t", y = "value")

# Aligned curves + Karcher mean
df_aln <- data.frame(
  t = rep(argvals, n), value = as.vector(t(km_diag$aligned$data)),
  curve = rep(seq_len(n), each = m)
)
df_kmean <- data.frame(t = argvals, value = as.numeric(km_diag$mean$data[1, ]))

p2 <- ggplot() +
  geom_line(data = df_aln, aes(x = t, y = value, group = curve),
            alpha = 0.15, color = "grey50") +
  geom_line(data = df_kmean, aes(x = t, y = value),
            color = "steelblue", linewidth = 1.2) +
  labs(title = "After: Karcher Mean", x = "t", y = "value")

p1 + p2
```

### Variance Decomposition

The `alignment.quality()` result decomposes total variance into amplitude
(shape) and phase (timing) components:

- **Variance reduction**: fraction of total pointwise variance removed by alignment
- **Phase-amplitude ratio**: how much of the total variability was due to phase
- **Warp complexity**: average deviation of warping functions from the identity (higher = more warping needed)
- **Warp smoothness**: measures the regularity of the warping functions

```{r quality-plot, fig.height = 4}
plot(aq, type = "variance")
```

### Pairwise Consistency

The `alignment.pairwise.consistency()` function checks whether pairwise
alignments are consistent -- i.e., whether aligning A to C directly gives the
same result as aligning A to B then B to C. A value near 1 indicates high
consistency:

```{r consistency}
pc <- alignment.pairwise.consistency(fd, lambda = 0, max.triplets = 100)
cat("Pairwise consistency:", round(pc, 3), "\n")
```

Low consistency can indicate that the data contains subgroups with different
shapes, or that the alignment is sensitive to the choice of target.

## Amplitude-Phase Decomposition

For individual curve pairs, `elastic.decomposition()` provides the full
decomposition of the elastic distance into amplitude and phase components:

```{r decomposition}
f1 <- fd[1]
f2 <- fd[5]

decomp <- elastic.decomposition(f1, f2, lambda = 0)
cat("Amplitude distance:", round(decomp$d_amplitude, 4), "\n")
cat("Phase distance:    ", round(decomp$d_phase, 4), "\n")
```

```{r decomp-plot, fig.height = 4}
# Visualize the aligned pair
f1_data <- as.numeric(f1$data)
f2_aligned <- decomp$f_aligned
df_pair <- data.frame(
  t = rep(argvals, 2),
  value = c(f1_data, f2_aligned),
  curve = rep(c("f1 (target)", "f2 (aligned)"), each = length(argvals))
)
ggplot(df_pair, aes(x = t, y = value, color = curve)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("f1 (target)" = "steelblue", "f2 (aligned)" = "firebrick")) +
  labs(title = sprintf("Amplitude dist = %.3f, Phase dist = %.3f",
                        decomp$d_amplitude, decomp$d_phase),
       x = "t", y = "f(t)", color = NULL)
```

This decomposition is useful for understanding *why* two curves differ:
is it because they have different shapes (amplitude) or different timing (phase)?

## Constrained Alignment

When curves have identifiable features (peaks, valleys), you can constrain the
elastic alignment to pass through specific landmark positions. This combines the
smoothness of elastic warping with guaranteed feature correspondence:

```{r constrained-example}
# Detect peaks and constrain alignment to pass through them
ec <- elastic.align.constrained(fd, kind = "peak", min.prominence = 0.3,
                                 expected.count = 1)
```

```{r constrained-plot, fig.height = 5}
p1 <- plot(res, type = "warps") + labs(title = "Elastic Warps")
p2 <- plot(ec, type = "warps") + labs(title = "Constrained Warps")
p1 + p2
```

The constrained warps are smooth between landmarks (like elastic) but anchored
at the peak positions (like landmark registration). For a detailed comparison
of all alignment methods, see `vignette("alignment-comparison", package = "fdars")`.

## When to Use Elastic Alignment

Elastic alignment works best when curves share the **same shape** but differ
in **timing**. The table below summarizes where it excels and where it does not.

### Works Well

| Scenario | Typical VR | Notes |
|:---------|:----------:|:------|
| Shifted peaks / bumps | 98-100% | Best case: localized features with horizontal shifts |
| ECG-like sharp peaks | 98-100% | Localized timing differences in multi-peak curves |
| Spectral data (shifted peaks) | 95-98% | Multiple peaks with coherent timing shifts |
| Phase-shifted smooth curves | 90-93% | Sinusoidal, polynomial, sigmoid curves with shifts |
| Non-periodic curves | 95-100% | Works well when boundary values are similar |

### Where It Degrades

| Scenario | Typical VR | Explanation |
|:---------|:----------:|:------------|
| Mixed amplitude + phase | 40-65% | Phase correction is partial when amplitude dominates |
| Noisy data (raw) | 0-30% | Noise in derivatives overwhelms SRSF; smooth first |
| Large boundary mismatch | 70-85% | Fixed boundary ($\gamma(0)=0$, $\gamma(1)=1$) can't correct endpoint values |
| Coarse grids (m < 50) | 70-75% | DP resolution limits alignment accuracy |

### Where It Should NOT Be Used

| Scenario | What happens | Recommendation |
|:---------|:-------------|:---------------|
| Pure amplitude differences | Identity warp (no harm, no help) | Use standard methods |
| Rate-varying monotone curves | Spurious warping, VR < 0 | These are amplitude/shape differences, not phase |
| Genuinely different shapes | Forced warping, VR < 0 | Use depth or distance methods instead |
| Very noisy data (unsmoothed) | Noise-driven warping, VR < 0 | Pre-smooth with `fdata2basis()` + `basis2fdata()` |

## Scenario Gallery

The following examples illustrate alignment behavior across different data
scenarios. Each shows the **original** curves (left) and **aligned** curves
(right) with variance reduction (VR) reported.

### ECG-like Multi-Peak Curves

Sharp, localized features with timing differences are the ideal use case:

```{r gallery-ecg, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.05, 0.05)
  t_shifted <- argvals - shift
  # P wave, QRS complex, T wave
  dm[i, ] <- 0.2 * exp(-200 * (t_shifted - 0.2)^2) -
             0.6 * exp(-500 * (t_shifted - 0.4)^2) +
             1.0 * exp(-800 * (t_shifted - 0.43)^2) -
             0.2 * exp(-500 * (t_shifted - 0.46)^2) +
             0.3 * exp(-150 * (t_shifted - 0.65)^2)
}
fd_ecg <- fdata(dm, argvals = argvals)
res_ecg <- elastic.align(fd_ecg)
vr <- 1 - mean(apply(res_ecg$aligned$data, 2, var)) /
          mean(apply(fd_ecg$data, 2, var))
cat("ECG alignment VR:", round(vr * 100, 1), "%\n")
```

```{r gallery-ecg-plot, fig.height = 4}
plot(res_ecg, type = "both")
```

### Spectral Peaks

Multiple peaks at varying positions, common in spectroscopy and chromatography:

```{r gallery-spectral, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.04, 0.04)
  dm[i, ] <- 0.8 * exp(-300 * (argvals - 0.25 - shift)^2) +
             1.0 * exp(-500 * (argvals - 0.50 - shift * 1.2)^2) +
             0.5 * exp(-300 * (argvals - 0.75 - shift * 0.8)^2)
}
fd_spec <- fdata(dm, argvals = argvals)
res_spec <- elastic.align(fd_spec)
vr <- 1 - mean(apply(res_spec$aligned$data, 2, var)) /
          mean(apply(fd_spec$data, 2, var))
cat("Spectral alignment VR:", round(vr * 100, 1), "%\n")
```

```{r gallery-spectral-plot, fig.height = 4}
plot(res_spec, type = "both")
```

### Growth-like Sigmoid Curves

Monotone S-shaped curves with timing shifts (e.g., growth curves with
differing onset times):

```{r gallery-sigmoid, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- 1 / (1 + exp(-20 * (argvals - 0.5 - shift)))
}
fd_sig <- fdata(dm, argvals = argvals)
res_sig <- elastic.align(fd_sig)
vr <- 1 - mean(apply(res_sig$aligned$data, 2, var)) /
          mean(apply(fd_sig$data, 2, var))
cat("Sigmoid alignment VR:", round(vr * 100, 1), "%\n")
```

```{r gallery-sigmoid-plot, fig.height = 4}
plot(res_sig, type = "both")
```

### Mixed Amplitude and Phase Variability

When curves differ in both height *and* timing, alignment corrects the phase
component but leaves amplitude variability untouched:

```{r gallery-mixed, fig.height = 4}
set.seed(42)
n <- 20
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 1.5)
  shift <- runif(1, -0.1, 0.1)
  dm[i, ] <- amp * sin(2 * pi * (argvals - shift))
}
fd_mix <- fdata(dm, argvals = argvals)
res_mix <- elastic.align(fd_mix)
vr <- 1 - mean(apply(res_mix$aligned$data, 2, var)) /
          mean(apply(fd_mix$data, 2, var))
cat("Mixed amplitude+phase VR:", round(vr * 100, 1), "%\n")
```

```{r gallery-mixed-plot, fig.height = 4}
plot(res_mix, type = "both")
```

After alignment, curves still differ in amplitude (vertical spread) but peaks
and valleys are now synchronized. The residual variance is real amplitude
variability, not a failure of alignment.

### Noisy Data: Raw vs Smoothed

Noise in derivatives overwhelms the SRSF. Pre-smoothing recovers alignment:

```{r gallery-noise, fig.height = 4}
set.seed(42)
n <- 20
dm_clean <- matrix(0, n, 100)
shifts <- runif(n, -0.1, 0.1)
for (i in 1:n) dm_clean[i, ] <- sin(2 * pi * (argvals - shifts[i]))
dm_noisy <- dm_clean + matrix(rnorm(n * 100, sd = 0.3), n, 100)
fd_noisy <- fdata(dm_noisy, argvals = argvals)

# Align raw noisy data
res_noisy <- elastic.align(fd_noisy)
vr_raw <- 1 - mean(apply(res_noisy$aligned$data, 2, var)) /
              mean(apply(fd_noisy$data, 2, var))

# Smooth first, then align
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_sm <- basis2fdata(coefs, argvals)
res_sm <- elastic.align(fd_sm)
vr_sm <- 1 - mean(apply(res_sm$aligned$data, 2, var)) /
             mean(apply(fd_sm$data, 2, var))

cat("Raw VR:     ", round(vr_raw * 100, 1), "%\n")
cat("Smoothed VR:", round(vr_sm * 100, 1), "%\n")
```

```{r gallery-noise-raw-plot, fig.height = 4}
plot(res_noisy, type = "both")
```

```{r gallery-noise-smooth-plot, fig.height = 4}
plot(res_sm, type = "both")
```

### Boundary Mismatch

When shifted curves have different values at the domain boundaries, alignment
is excellent in the interior but constrained at the edges:

```{r gallery-boundary, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
shifts <- runif(n, -0.15, 0.15)
for (i in 1:n) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd_bnd <- fdata(dm, argvals = argvals)
res_bnd <- elastic.align(fd_bnd)
vr <- 1 - mean(apply(res_bnd$aligned$data, 2, var)) /
          mean(apply(fd_bnd$data, 2, var))
cat("Boundary mismatch VR:", round(vr * 100, 1), "% — note residual spread at boundaries\n")
```

```{r gallery-boundary-plot, fig.height = 4}
plot(res_bnd, type = "both")
```

### Pure Amplitude Differences

When curves differ only in height (no phase variability), alignment correctly
produces identity warps -- no benefit, but no harm:

```{r gallery-amplitude, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- runif(1, 0.5, 2.0)
  dm[i, ] <- amp * sin(2 * pi * argvals)
}
fd_amp <- fdata(dm, argvals = argvals)
res_amp <- elastic.align(fd_amp)
vr <- 1 - mean(apply(res_amp$aligned$data, 2, var)) /
          mean(apply(fd_amp$data, 2, var))
cat("Pure amplitude VR:", round(vr * 100, 1), "% — identity warps applied\n")
```

```{r gallery-amplitude-plot, fig.height = 4}
plot(res_amp, type = "both")
```

### Genuinely Different Shapes

Alignment should **not** be used when curves have fundamentally different
shapes. Forced warping produces meaningless results:

```{r gallery-different, fig.height = 4}
set.seed(42)
n <- 15
dm <- matrix(0, n, 100)
for (i in 1:n) {
  if (i %% 5 == 1) dm[i, ] <- sin(2 * pi * argvals)
  else if (i %% 5 == 2) dm[i, ] <- cos(6 * pi * argvals)
  else if (i %% 5 == 3) dm[i, ] <- argvals^3
  else if (i %% 5 == 4) dm[i, ] <- sin(10 * pi * argvals) * exp(-3 * argvals)
  else dm[i, ] <- dnorm(argvals, 0.5, 0.05) / dnorm(0.5, 0.5, 0.05)
}
fd_diff <- fdata(dm, argvals = argvals)
res_diff <- elastic.align(fd_diff)
vr <- 1 - mean(apply(res_diff$aligned$data, 2, var)) /
          mean(apply(fd_diff$data, 2, var))
cat("Different shapes VR:", round(vr * 100, 1), "% — forced warping of unrelated shapes\n")
```

```{r gallery-different-plot, fig.height = 4}
plot(res_diff, type = "both")
```

When VR is very low or negative, it indicates the data is not suitable for
elastic alignment. Use depth or distance methods instead.

## Multi-Bump Scaling

Real-world functional data often contains multiple peaks or features that vary
independently. This section explores how elastic alignment handles curves with
**varying inter-bump distance** (peaks shift independently) and **varying bump
width** (peaks broaden or narrow) as the number of bumps increases.

### Two Bumps

With two bumps, the warping function has enough freedom to independently
adjust each peak's position and width:

```{r multi-bump-2}
set.seed(42)
m <- 200
argvals_mb <- seq(0, 1, length.out = m)
n <- 20

# Varying inter-bump distance
data_pos <- matrix(0, n, m)
for (i in 1:n) {
  p1 <- 0.3
  p2 <- 0.7 + runif(1, -0.12, 0.12)
  data_pos[i, ] <- exp(-300 * (argvals_mb - p1)^2) +
                   exp(-300 * (argvals_mb - p2)^2)
}
fd_pos <- fdata(data_pos, argvals = argvals_mb)
res_pos <- elastic.align(fd_pos)
vr_pos <- 1 - mean(apply(res_pos$aligned$data, 2, var)) /
              mean(apply(fd_pos$data, 2, var))

# Varying bump width
data_w <- matrix(0, n, m)
for (i in 1:n) {
  w1 <- runif(1, 200, 500)
  w2 <- runif(1, 200, 500)
  data_w[i, ] <- exp(-w1 * (argvals_mb - 0.3)^2) +
                 exp(-w2 * (argvals_mb - 0.7)^2)
}
fd_w <- fdata(data_w, argvals = argvals_mb)
res_w <- elastic.align(fd_w)
vr_w <- 1 - mean(apply(res_w$aligned$data, 2, var)) /
             mean(apply(fd_w$data, 2, var))

cat("2 bumps — varying position: VR =", round(vr_pos * 100, 1), "%\n")
cat("2 bumps — varying width:    VR =", round(vr_w * 100, 1), "%\n")
```

```{r multi-bump-2-pos-plot, fig.height = 4}
plot(res_pos, type = "both")
```

```{r multi-bump-2-width-plot, fig.height = 4}
plot(res_w, type = "both")
```

Both cases achieve near-perfect alignment. The warping function locally
stretches/compresses the region around each bump to bring peaks into
correspondence.

### Scaling to More Bumps

As the number of bumps increases, alignment becomes progressively harder.
A monotone warping function that stretches one region must compress another,
so with many tightly packed bumps the degrees of freedom become insufficient
to independently correct each peak.

```{r multi-bump-scaling}
set.seed(42)
results <- data.frame(
  bumps = integer(), scenario = character(), VR = numeric(),
  stringsAsFactors = FALSE
)

for (nb in c(2, 4, 5, 7)) {
  positions_base <- seq(0.1, 0.9, length.out = nb)

  # Varying positions
  data_pos <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      p <- positions_base[j] + runif(1, -0.04, 0.04)
      data_pos[i, ] <- data_pos[i, ] + exp(-300 * (argvals_mb - p)^2)
    }
  }
  fd_pos <- fdata(data_pos, argvals = argvals_mb)
  res_pos <- elastic.align(fd_pos)
  vr_pos <- 1 - mean(apply(res_pos$aligned$data, 2, var)) /
                mean(apply(fd_pos$data, 2, var))

  # Varying widths
  data_w <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      w <- runif(1, 200, 600)
      data_w[i, ] <- data_w[i, ] + exp(-w * (argvals_mb - positions_base[j])^2)
    }
  }
  fd_w <- fdata(data_w, argvals = argvals_mb)
  res_w <- elastic.align(fd_w)
  vr_w <- 1 - mean(apply(res_w$aligned$data, 2, var)) /
               mean(apply(fd_w$data, 2, var))

  # Both varying
  data_both <- matrix(0, n, m)
  for (i in 1:n) {
    for (j in 1:nb) {
      p <- positions_base[j] + runif(1, -0.04, 0.04)
      w <- runif(1, 200, 600)
      data_both[i, ] <- data_both[i, ] + exp(-w * (argvals_mb - p)^2)
    }
  }
  fd_both <- fdata(data_both, argvals = argvals_mb)
  res_both <- elastic.align(fd_both)
  vr_both <- 1 - mean(apply(res_both$aligned$data, 2, var)) /
                  mean(apply(fd_both$data, 2, var))

  results <- rbind(results, data.frame(
    bumps = rep(nb, 3),
    scenario = c("Position", "Width", "Both"),
    VR = c(vr_pos, vr_w, vr_both) * 100
  ))
}
```

```{r multi-bump-table}
# Display as a table
df_wide <- reshape(results, idvar = "bumps", timevar = "scenario",
                   direction = "wide")
names(df_wide) <- gsub("VR\\.", "", names(df_wide))
for (col in c("Position", "Width", "Both")) {
  df_wide[[col]] <- paste0(round(df_wide[[col]], 1), "%")
}
knitr::kable(df_wide, col.names = c("Bumps", "Varying Position",
             "Varying Width", "Both Varying"),
             align = "cccc",
             caption = "Variance reduction by number of bumps and variability type")
```

### Visual Comparison: 4 vs 7 Bumps

```{r multi-bump-4v7}
set.seed(42)

# 4 bumps — both varying
pos4 <- seq(0.1, 0.9, length.out = 4)
data4 <- matrix(0, n, m)
for (i in 1:n) {
  for (j in 1:4) {
    p <- pos4[j] + runif(1, -0.04, 0.04)
    w <- runif(1, 200, 600)
    data4[i, ] <- data4[i, ] + exp(-w * (argvals_mb - p)^2)
  }
}
fd4 <- fdata(data4, argvals = argvals_mb)
res4 <- elastic.align(fd4)
vr4 <- 1 - mean(apply(res4$aligned$data, 2, var)) /
            mean(apply(fd4$data, 2, var))

# 7 bumps — both varying
pos7 <- seq(0.1, 0.9, length.out = 7)
data7 <- matrix(0, n, m)
for (i in 1:n) {
  for (j in 1:7) {
    p <- pos7[j] + runif(1, -0.04, 0.04)
    w <- runif(1, 200, 600)
    data7[i, ] <- data7[i, ] + exp(-w * (argvals_mb - p)^2)
  }
}
fd7 <- fdata(data7, argvals = argvals_mb)
res7 <- elastic.align(fd7)
vr7 <- 1 - mean(apply(res7$aligned$data, 2, var)) /
            mean(apply(fd7$data, 2, var))

cat("4 bumps (both varying) — VR =", round(vr4 * 100, 1), "%\n")
cat("7 bumps (both varying) — VR =", round(vr7 * 100, 1), "%\n")
```

```{r multi-bump-4-plot, fig.height = 4}
plot(res4, type = "both")
```

```{r multi-bump-7-plot, fig.height = 4}
plot(res7, type = "both")
```

With 4 bumps, alignment still captures most of the variability. With 7
tightly packed bumps, the monotonicity constraint on the warping function
limits independent correction of each peak -- stretching one inter-peak
region forces compression elsewhere. This is a fundamental limitation of
elastic alignment, not a numerical issue: a single monotone time-warping
cannot independently retime many closely spaced features.

## Comparison with DTW and Shift Registration

Three alignment strategies are available in **fdars**:

| | Elastic (SRSF) | DTW | Shift Registration |
|:--|:--|:--|:--|
| **Function** | `elastic.align()` | `metric.DTW()` | `register.fd()` |
| **Warping** | Smooth, monotone diffeomorphism | Piecewise constant, allows repeated indices | Global translation only |
| **Metric** | Proper metric (triangle inequality holds) | Not a metric | Not applicable (no warping) |
| **Amplitude/Phase** | Cleanly separates the two | Mixes them — can absorb amplitude differences into warping | Pure phase (shift), amplitude untouched |
| **Degeneracy** | No "pinching" — SRSF penalizes compression | Can collapse entire regions to a single point | No degeneracy, but very limited |
| **Mean** | Karcher mean is well-defined | No principled mean | Shift-corrected cross-sectional mean |
| **Complexity** | $O(m^2)$ per pair (DP) | $O(m^2)$ per pair (DP) | $O(m)$ per curve (cross-correlation) |

### Why Elastic Alignment Avoids Degeneracy

A key advantage of the SRSF framework is that the elastic distance penalizes
"pinching" — collapsing a region of the curve to a single point. The SRSF
of a warped function is $(q \circ \gamma)\sqrt{\dot\gamma}$, so when
$\dot\gamma \to 0$ (extreme compression), the SRSF contribution vanishes
rather than concentrating. This acts as a natural regularizer. In contrast,
DTW allows repeated indices that effectively collapse entire regions to a
single point, which can produce degenerate alignments.

### Head-to-Head: Shifted Bumps

The clearest way to see the differences is on data with localized features:

```{r comparison-setup}
set.seed(42)
m_cmp <- 200
argvals_cmp <- seq(0, 1, length.out = m_cmp)
n_cmp <- 15

# Two bumps with varying positions
data_cmp <- matrix(0, n_cmp, m_cmp)
for (i in 1:n_cmp) {
  p1 <- 0.3 + runif(1, -0.06, 0.06)
  p2 <- 0.7 + runif(1, -0.06, 0.06)
  data_cmp[i, ] <- exp(-300 * (argvals_cmp - p1)^2) +
                   exp(-300 * (argvals_cmp - p2)^2)
}
fd_cmp <- fdata(data_cmp, argvals = argvals_cmp)
var_orig_cmp <- mean(apply(fd_cmp$data, 2, var))
```

**Elastic alignment** finds smooth, local warping for each bump independently:

```{r comparison-elastic}
res_elastic <- elastic.align(fd_cmp)
vr_elastic <- 1 - mean(apply(res_elastic$aligned$data, 2, var)) / var_orig_cmp
cat("Elastic VR:", round(vr_elastic * 100, 1), "%\n")
```

**Shift registration** applies a single global shift -- it can center one bump
but not both when they move independently:

```{r comparison-shift}
res_shift <- register.fd(fd_cmp)
vr_shift <- 1 - mean(apply(res_shift$registered$data, 2, var)) / var_orig_cmp
cat("Shift registration VR:", round(vr_shift * 100, 1), "%\n")
```

```{r comparison-plot, fig.height = 6}
plot_curves <- function(data, argvals, title) {
  n <- nrow(data)
  m <- length(argvals)
  df <- data.frame(
    curve_id = rep(seq_len(n), each = m),
    argval = rep(argvals, n),
    value = as.vector(t(data))
  )
  ggplot(df, aes(x = argval, y = value, group = curve_id)) +
    geom_line(alpha = 0.5) +
    labs(title = title, x = "t", y = "value") +
    theme_minimal()
}

p1 <- plot_curves(fd_cmp$data, argvals_cmp, "Original")
p2 <- plot_curves(res_elastic$aligned$data, argvals_cmp,
        paste0("Elastic (VR = ", round(vr_elastic * 100, 1), "%)"))
p3 <- plot_curves(res_shift$registered$data, argvals_cmp,
        paste0("Shift Registration (VR = ", round(vr_shift * 100, 1), "%)"))

p1 / p2 / p3
```

Elastic alignment resolves the independent shifts of both bumps. Shift
registration can only correct a single global offset, leaving residual
phase variability when features move independently.

### DTW Distance vs Elastic Distance

DTW and elastic distances rank curves differently because DTW can absorb
amplitude differences into warping while elastic alignment separates them:

```{r comparison-distances}
# Compute both distance matrices
D_elastic <- elastic.distance(fd_cmp)
D_dtw <- metric.DTW(fd_cmp)

# Compare: correlation of off-diagonal entries
idx <- upper.tri(D_elastic)
cat("Correlation between elastic and DTW distances:",
    round(cor(D_elastic[idx], D_dtw[idx]), 3), "\n")
```

```{r comparison-distances-plot, fig.height = 4}
df_ed <- data.frame(elastic = D_elastic[idx], dtw = D_dtw[idx])
ggplot(df_ed, aes(x = elastic, y = dtw)) +
  geom_point(color = "steelblue", size = 1, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  labs(title = "Elastic vs DTW Pairwise Distances",
       x = "Elastic distance", y = "DTW distance") +
  theme_minimal()
```

The distances are correlated but not identical. The key difference emerges
when amplitude variability is present:

```{r comparison-amplitude, fig.height = 4}
set.seed(42)
# Curves with BOTH amplitude and phase variability
data_amp <- matrix(0, n_cmp, m_cmp)
for (i in 1:n_cmp) {
  amp <- runif(1, 0.5, 1.5)
  shift <- runif(1, -0.06, 0.06)
  data_amp[i, ] <- amp * exp(-200 * (argvals_cmp - 0.5 - shift)^2)
}
fd_amp <- fdata(data_amp, argvals = argvals_cmp)

D_elastic_amp <- elastic.distance(fd_amp)
D_dtw_amp <- metric.DTW(fd_amp)
idx_amp <- upper.tri(D_elastic_amp)

cat("Correlation (amplitude+phase):",
    round(cor(D_elastic_amp[idx_amp], D_dtw_amp[idx_amp]), 3), "\n")
```

With mixed amplitude and phase, DTW distances are less correlated with elastic
distances because DTW warping partially absorbs amplitude differences. Elastic
distances preserve the amplitude/phase separation, making them more suitable
for downstream analysis (clustering, MDS) when both sources of variability
carry distinct meaning.

### When to Use Each Method

| Scenario | Recommended |
|:---------|:------------|
| Localized features with independent timing shifts | Elastic alignment |
| Simple global time shift or delay | Shift registration (`register.fd`) — fast and sufficient |
| Distance computation for clustering/MDS | Elastic distance (proper metric) or DTW (if metric property not needed) |
| Noisy signals, speech-like data | DTW (more robust to noise without pre-smoothing) |
| Periodic/circular data | `elastic.align(periodic = TRUE)` |
| Need principled mean shape | `karcher.mean()` (no DTW equivalent) |

## Periodic Functional Data

### The Problem

Elastic alignment uses fixed boundary constraints $\gamma(0) = 0$, $\gamma(1) = 1$.
This prevents the warping function from performing circular shifts, which means
periodic data (e.g., curves on $[0, 2\pi]$ where $f(0) = f(2\pi)$) with large
phase offsets cannot be properly aligned.

For example, two copies of $\sin(t)$ shifted by half a period have identical
shape but the boundary constraints force alignment to fail:

```{r periodic-problem, fig.height = 4}
set.seed(42)
m <- 100
argvals_p <- seq(0, 2 * pi, length.out = m)
n <- 15
data_p <- matrix(0, n, m)
for (i in 1:n) {
  shift <- sample(0:(m - 1), 1)
  idx <- ((seq_len(m) - 1L + shift) %% m) + 1L
  data_p[i, ] <- sin(argvals_p[idx])
}
fd_p <- fdata(data_p, argvals = argvals_p)

# Standard alignment struggles with circular shifts
res_std <- elastic.align(fd_p)
var_orig <- mean(apply(fd_p$data, 2, var))
vr_std <- 1 - mean(apply(res_std$aligned$data, 2, var)) / var_orig
cat("Standard alignment on periodic data — VR:", round(vr_std * 100, 1), "%\n")
```

```{r periodic-problem-plot, fig.height = 4}
plot(res_std, type = "both")
```

### The Solution

Setting `periodic = TRUE` applies a two-stage approach:

1. **Circular rotation**: For each curve, find the position of the global
   maximum. Circularly shift the curve so that this maximum lands at a fixed
   canonical grid position (by default, one quarter of the domain). The shift
   for curve $i$ is simply `which.max(f_i) - target_pos`. This is a
   deterministic, fast heuristic -- no optimisation is involved.
2. **Elastic alignment**: Standard boundary-constrained alignment on the
   rotated data, correcting any remaining timing differences.

**Caveat**: The rotation heuristic relies on a single dominant peak. If curves
have multiple peaks of similar height, `which.max` may select different peaks
for different curves, producing inconsistent rotations. In such cases, you may
need to supply a custom `target.pos` to `periodic.rotate()`, or pre-process
the data to ensure a consistent landmark is the global maximum.

```{r periodic-solution, fig.height = 4}
res_per <- elastic.align(fd_p, periodic = TRUE)
vr_per <- 1 - mean(apply(res_per$aligned$data, 2, var)) / var_orig
cat("Periodic alignment — VR:", round(vr_per * 100, 1), "%\n")
```

```{r periodic-solution-plot, fig.height = 4}
plot(res_per, type = "both")
```

The rotation step removes large circular shifts that the boundary-constrained
warping cannot handle, and the elastic step corrects any remaining timing
differences.

### What Happens to the Warping Functions?

Comparing the warping functions from standard vs periodic alignment reveals
why the standard approach fails. Let's look at the warping functions side by
side, and inspect the warping speed:

```{r periodic-warps-compare, fig.height = 4}
p_warp_std <- plot(res_std, type = "warps") +
  labs(title = "Standard: warping functions")
p_warp_per <- plot(res_per, type = "warps") +
  labs(title = "Periodic: warping functions")
p_warp_std + p_warp_per
```

The standard warps are forced through $(0, 0)$ and $(2\pi, 2\pi)$. When the
true offset is a large circular shift, the warp must bend drastically to
compensate, producing extreme stretching in some regions and compression in
others. The periodic warps are much closer to the identity because the circular
rotation has already removed most of the phase offset.

We can quantify this by looking at the warping speed $\dot\gamma(t)$ via
`deriv()`:

```{r periodic-speed-compare, fig.height = 4}
speed_std <- deriv(res_std$gammas)
speed_per <- deriv(res_per$gammas)

cat("Standard alignment — warping speed range:",
    round(min(speed_std$data), 2), "to", round(max(speed_std$data), 2), "\n")
cat("Periodic alignment — warping speed range:",
    round(min(speed_per$data), 2), "to", round(max(speed_per$data), 2), "\n")
```

```{r periodic-speed-plot, fig.height = 4}
n_std <- nrow(speed_std$data)
m_std <- ncol(speed_std$data)
df_speed_std <- data.frame(
  t = rep(speed_std$argvals, n_std),
  speed = as.vector(t(speed_std$data)),
  curve = rep(seq_len(n_std), each = m_std)
)
n_per <- nrow(speed_per$data)
m_per <- ncol(speed_per$data)
df_speed_per <- data.frame(
  t = rep(speed_per$argvals, n_per),
  speed = as.vector(t(speed_per$data)),
  curve = rep(seq_len(n_per), each = m_per)
)

p_sp_std <- ggplot(df_speed_std, aes(x = t, y = speed, group = curve)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60") +
  geom_line(alpha = 0.3, color = "coral") +
  labs(title = "Standard: warping speed", x = "t",
       y = expression(dot(gamma)(t)))

p_sp_per <- ggplot(df_speed_per, aes(x = t, y = speed, group = curve)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60") +
  geom_line(alpha = 0.3, color = "steelblue") +
  coord_cartesian(ylim = range(df_speed_std$speed)) +
  labs(title = "Periodic: warping speed", x = "t",
       y = expression(dot(gamma)(t)))

p_sp_std + p_sp_per
```

The standard warping speeds swing wildly (extreme stretching and compression),
while the periodic warping speeds stay close to 1 (the identity). This
illustrates the benefit of the two-stage periodic approach: the rotation
handles the gross circular shift, leaving only minor residual timing
differences for the elastic alignment to correct.

Let's examine one curve in detail to see the full decomposition -- the
circular rotation followed by the residual elastic warp:

```{r periodic-anatomy, fig.height = 12}
# Pick the curve with the largest rotation shift
rot <- periodic.rotate(fd_p)
i_max <- which.max(abs(rot$shifts))
shift_amount <- rot$shifts[i_max]

# Three stages: original, after rotation, after alignment
f_orig <- as.numeric(fd_p$data[i_max, ])
f_rotated <- as.numeric(rot$fdataobj$data[i_max, ])
f_aligned <- as.numeric(res_per$aligned$data[i_max, ])
f_target <- as.numeric(res_per$target)

# Panel 1: the three curve stages
df_stages <- data.frame(
  t = rep(argvals_p, 4),
  value = c(f_orig, f_rotated, f_aligned, f_target),
  stage = rep(c("1. Original", "2. After rotation", "3. After alignment",
                "Target"), each = m)
)

p1 <- ggplot(df_stages, aes(x = t, y = value, color = stage,
                             linetype = stage)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("1. Original" = "coral",
                                "2. After rotation" = "goldenrod",
                                "3. After alignment" = "steelblue",
                                "Target" = "grey30")) +
  scale_linetype_manual(values = c("1. Original" = "solid",
                                   "2. After rotation" = "solid",
                                   "3. After alignment" = "solid",
                                   "Target" = "dashed")) +
  labs(title = paste0("Curve ", i_max, ": two-stage periodic alignment",
                      " (rotation = ", shift_amount, " grid points)"),
       x = "t", y = "value", color = NULL, linetype = NULL) +
  theme(legend.position = "top")

# Panel 2: warping function (standard vs periodic)
gamma_std <- as.numeric(res_std$gammas$data[i_max, ])
gamma_per <- as.numeric(res_per$gammas$data[i_max, ])
df_warps <- data.frame(
  t = rep(argvals_p, 2),
  gamma = c(gamma_std, gamma_per),
  method = rep(c("Standard", "Periodic"), each = m)
)

p2 <- ggplot(df_warps, aes(x = t, y = gamma, color = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(Standard = "coral", Periodic = "steelblue")) +
  labs(title = "Warping function: standard vs periodic",
       x = "t", y = expression(gamma(t)), color = NULL) +
  theme(legend.position = "top")

# Panel 3: warping speed comparison
speed_std_i <- as.numeric(deriv(res_std$gammas)$data[i_max, ])
speed_per_i <- as.numeric(deriv(res_per$gammas)$data[i_max, ])
df_sp <- data.frame(
  t = rep(deriv(res_per$gammas)$argvals, 2),
  speed = c(speed_std_i, speed_per_i),
  method = rep(c("Standard", "Periodic"), each = length(speed_per_i))
)

p3 <- ggplot(df_sp, aes(x = t, y = speed, color = method)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(Standard = "coral", Periodic = "steelblue")) +
  labs(title = "Local warping speed: deriv(gammas)",
       x = "t", y = expression(dot(gamma)(t)), color = NULL) +
  theme(legend.position = "top")

p1 / p2 / p3
```

The three panels show the full story for this curve:

1. **Top**: The original curve (coral) is circularly shifted far from the
   target. Rotation (gold) brings it close; the residual elastic warp (blue)
   finishes the job.
2. **Middle**: The standard warping function (coral) deviates wildly from the
   diagonal because it must encode a large circular shift as a boundary-
   constrained warp. The periodic warping function (blue) stays near the
   identity.
3. **Bottom**: The standard warp speed swings between extreme compression and
   stretching. The periodic warp speed is nearly constant at 1.

### Where Did the Shift Go?

Notice that the periodic warping function is nearly the identity -- it does
**not** show the large circular shift visible in the original data. This is
because the shift was already removed by `periodic.rotate()` *before* the
elastic alignment runs. The total phase correction in the periodic approach
is decomposed into two pieces:

1. **Circular rotation** (stored in `res_per$rotations`) -- the gross shift,
   measured in grid points
2. **Elastic warp** (stored in `res_per$gammas`) -- the residual fine-tuning

You can inspect both:

```{r periodic-decompose}
cat("Curve", i_max, "phase decomposition:\n")
cat("  Circular rotation:", res_per$rotations[i_max], "grid points (",
    round(res_per$rotations[i_max] / m * 2 * pi, 2), "radians)\n")
warp_deviation <- as.numeric(res_per$gammas$data[i_max, ]) - argvals_p
cat("  Residual warp: mean |gamma(t) - t| =",
    round(mean(abs(warp_deviation)), 4), "\n")
```

This decomposition is the key difference from the non-periodic case. In the
Quick Start example, the warping function had to absorb the *entire* shift,
producing a visible deviation from the diagonal. Here, the rotation absorbs
the gross shift, and the warping function only encodes what's left.

### Alternative Rotation Methods

The default `"peak"` rotation uses `which.max` to locate each curve's global
maximum and shift it to a canonical position. This is fast but fragile when
curves have **multiple peaks of similar height** -- `which.max` may pick
different peaks for different curves, producing inconsistent rotations.

Three additional methods are available via the `method` argument to
`periodic.rotate()` (and via `rotate.method` in `elastic.align()` and
`karcher.mean()`):

#### Cross-correlation (`method = "xcorr"`)

Cross-correlation finds the circular shift that best aligns each curve to a
reference, using all features simultaneously rather than a single landmark.
This is robust to multi-peak shapes:

```{r rotate-xcorr, fig.height = 4}
# Multi-peak curves where peak method fails
set.seed(42)
base_multi <- sin(argvals_p) + 0.95 * sin(2 * argvals_p + 0.3)
n_multi <- 12
data_multi <- matrix(0, n_multi, m)
for (i in 1:n_multi) {
  shift <- sample(0:(m - 1), 1)
  idx <- ((seq_len(m) - 1L + shift) %% m) + 1L
  data_multi[i, ] <- base_multi[idx]
}
fd_multi <- fdata(data_multi, argvals = argvals_p)

rot_peak <- periodic.rotate(fd_multi, method = "peak")
rot_xcorr <- periodic.rotate(fd_multi, method = "xcorr")

# Compare variance reduction
var_orig <- mean(apply(fd_multi$data, 2, var))
vr_peak <- 1 - mean(apply(rot_peak$fdataobj$data, 2, var)) / var_orig
vr_xcorr <- 1 - mean(apply(rot_xcorr$fdataobj$data, 2, var)) / var_orig
cat("Multi-peak rotation VR — peak:", round(vr_peak * 100, 1), "%,",
    "xcorr:", round(vr_xcorr * 100, 1), "%\n")
```

```{r rotate-xcorr-plot, fig.height = 4}
p_peak <- plot(rot_peak$fdataobj) + labs(title = "Peak rotation")
p_xcorr <- plot(rot_xcorr$fdataobj) + labs(title = "Cross-correlation rotation")
p_peak + p_xcorr
```

The cross-correlation method aligns curves coherently because it matches the
**entire shape** rather than a single point. You can supply a custom reference
via the `reference` argument; by default it uses the cross-sectional mean.

#### Landmark-based (`method = "landmark"`)

The landmark method generalises the peak approach: instead of `which.max`, you
supply any function that locates a feature of interest. For example, a
"steepest rise" function finds the point of maximum first derivative:

```{r rotate-landmark}
steepest_rise <- function(curve) {
  diffs <- diff(curve)
  which.max(diffs)
}

rot_lm <- periodic.rotate(fd_p, method = "landmark",
                           landmark.func = steepest_rise)
cat("Landmark shifts:", rot_lm$shifts, "\n")
```

This is useful when the relevant feature isn't the peak but some other
characteristic point -- a zero crossing, the start of a rise, etc.

#### Iterative (`method = "iterative"`)

The iterative method alternates between cross-correlation rotation and elastic
alignment. This handles cases where the optimal rotation depends on the
alignment and vice versa:

```{r rotate-iterative}
rot_iter <- periodic.rotate(fd_p, method = "iterative", max.iter = 5)
cat("Iterative shifts:", rot_iter$shifts, "\n")
```

The iterative approach converges when no further shifts are needed (all lags
become zero). It is the most expensive method but can improve results when
curves have complex shapes.

#### When to Use Each Method

| Method | Best for | Cost |
|--------|----------|------|
| `"peak"` | Single dominant peak, fast pre-screening | O(m) per curve |
| `"xcorr"` | Multi-peak curves, noisy data | O(m log m) per curve |
| `"landmark"` | Domain-specific features (zero crossing, onset) | O(m) per curve |
| `"iterative"` | Complex shapes where rotation and alignment interact | O(m log m) × iters |

You can pass the rotation method through to `elastic.align()` and
`karcher.mean()` via the `rotate.method` parameter:

```{r rotate-method-passthrough}
res_xcorr <- elastic.align(fd_multi, periodic = TRUE, rotate.method = "xcorr")
print(res_xcorr)
```

You can also apply the rotation step manually using `periodic.rotate()`:

```{r periodic-rotate}
rot <- periodic.rotate(fd_p)
cat("Shifts applied:", rot$shifts, "\n")
```

The `karcher.mean()` function also accepts `periodic = TRUE`:

```{r periodic-karcher}
km_per <- karcher.mean(fd_p, max.iter = 15, periodic = TRUE)
print(km_per)
```

## Real Data: Berkeley Growth Velocity

The **Berkeley Growth Study** recorded the heights of 93 children (39 boys,
54 girls) from ages 1 to 18. The growth *velocity* curves (cm/year) exhibit
clear phase variability: each child's pubertal growth spurt occurs at a
different age, creating horizontal shifts in the peak.

The primary output of elastic alignment here is the **warping functions** --
each $\gamma_i$ encodes *when* child $i$'s growth spurt occurred relative to
the population, i.e., their biological maturation timing. These warpings are
directly useful for downstream analysis (correlating puberty timing with other
variables, clustering by maturation pattern, etc.). The aligned curves
themselves are a byproduct: they represent pure amplitude variability (how
*tall* the growth spurt is) after timing differences have been removed.

This dataset is a standard benchmark in elastic FDA literature (Tucker et al.,
2013; Srivastava and Klassen, 2016). It is available in the `fdasrvf` package.

```{r growth-data}
data(growth_vel, package = "fdasrvf")

# growth_vel$f is 69 (grid) x 93 (curves) — transpose for fdata
# Standard ordering: first 39 = boys, last 54 = girls
fd_growth_raw <- fdata(t(growth_vel$f), argvals = growth_vel$time)
gender <- factor(c(rep("Boy", 39), rep("Girl", 54)))
cat("Curves:", nrow(fd_growth_raw$data),
    "(", sum(gender == "Boy"), "boys,", sum(gender == "Girl"), "girls)\n")
cat("Grid points:", ncol(fd_growth_raw$data), "\n")
cat("Age range:", range(fd_growth_raw$argvals), "\n")
```

The raw data has only 69 grid points, which is coarse for elastic alignment
(the DP warping resolution is limited). We smooth with P-splines and evaluate
on a finer grid:

```{r growth-smooth}
# Smooth with P-splines and evaluate on a finer grid
fine_argvals <- seq(min(fd_growth_raw$argvals), max(fd_growth_raw$argvals),
                    length.out = 200)
coefs_growth <- fdata2basis(fd_growth_raw, nbasis = 25, type = "bspline")
fd_growth <- basis2fdata(coefs_growth, fine_argvals)
cat("Smoothed grid points:", ncol(fd_growth$data), "\n")
```

### Full Age Range

```{r growth-full, fig.height = 4}
plot(fd_growth) + labs(title = "Berkeley Growth Velocity (all 93 children)",
                       x = "Age (years)", y = "Growth velocity (cm/year)")
```

The dominant feature at age 1 (infant growth) and the pubertal spurt around
ages 11--14 are both visible, but the pubertal peak is blurred by phase
variability.

```{r growth-full-align}
res_growth <- elastic.align(fd_growth)
var_orig_growth <- mean(apply(fd_growth$data, 2, var))
vr_growth <- 1 - mean(apply(res_growth$aligned$data, 2, var)) / var_orig_growth
cat("Full range VR:", round(vr_growth * 100, 1), "%\n")
```

```{r growth-full-plot, fig.height = 4}
plot(res_growth, type = "both")
```

### Pubertal Growth Spurt (Ages 8--18)

The pubertal region is where phase variability is strongest. Restricting to
ages 8--18 isolates the growth spurt:

```{r growth-pubertal}
idx_pub <- which(fd_growth$argvals >= 8)
fd_pub <- fdata(fd_growth$data[, idx_pub],
                argvals = fd_growth$argvals[idx_pub])

res_pub <- elastic.align(fd_pub)
var_pub <- mean(apply(fd_pub$data, 2, var))
vr_pub <- 1 - mean(apply(res_pub$aligned$data, 2, var)) / var_pub
cat("Pubertal region VR:", round(vr_pub * 100, 1), "%\n")
```

```{r growth-pubertal-plot, fig.height = 4}
plot(res_pub, type = "both")
```

### Cross-Sectional Mean vs Karcher Mean

The Karcher mean serves as the **reference frame** that gives the warping
functions their meaning -- each $\gamma_i$ is defined as "warp curve $i$ to
match *this* mean." Without a consistent reference, the warpings would not be
comparable across children.

As a bonus, the Karcher mean also recovers a de-blurred population average.
The cross-sectional mean is attenuated because children's growth spurts are
averaged at mismatched ages; the Karcher mean recovers the true peak amplitude:

```{r growth-karcher}
km_pub <- karcher.mean(fd_pub, max.iter = 20)
cross_mean_pub <- as.numeric(mean(fd_pub)$data)
karcher_pub <- as.numeric(km_pub$mean$data)

cat("Cross-sectional mean peak:", round(max(cross_mean_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(cross_mean_pub)], 1), "\n")
cat("Karcher mean peak:        ", round(max(karcher_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(karcher_pub)], 1), "\n")
```

```{r growth-karcher-plot, fig.height = 4}
df_means <- data.frame(
  age = rep(fd_pub$argvals, 2),
  velocity = c(cross_mean_pub, karcher_pub),
  type = rep(c("Cross-sectional mean", "Karcher mean"),
             each = length(fd_pub$argvals))
)

ggplot(df_means, aes(x = age, y = velocity, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Pubertal Growth Spurt: Cross-Sectional vs Karcher Mean",
       x = "Age (years)", y = "Growth velocity (cm/year)", color = NULL)
```

The Karcher mean is sharper and taller because it aligns the growth spurts
before averaging, rather than smearing them across different ages. This sharper
shape is a better estimate of the typical growth velocity profile -- but its
primary role is as the anchor that makes the per-child warping functions
interpretable and mutually comparable.

### FPCA Before and After Alignment

Alignment concentrates variance into fewer principal components:

```{r growth-fpca}
pc_before <- fdata2pc(fd_pub, ncomp = 3)
pc_after <- fdata2pc(km_pub$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2],
    "%, PC3 =", pve_before[3], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2],
    "%, PC3 =", pve_after[3], "%\n")
```

Before alignment, phase variability spreads across multiple PCs. After
alignment, the remaining (amplitude) variability is more efficiently captured
by the first PC.

### Warping Functions Reveal Growth Timing

The warping functions are the main extracted quantity from this analysis.
Each $\gamma_i$ maps population-average time to child $i$'s biological time,
revealing how their developmental clock differs from the Karcher mean reference:

```{r growth-warps, fig.height = 4}
n_pub <- nrow(res_pub$gammas$data)
m_pub <- ncol(res_pub$gammas$data)
df_warps <- data.frame(
  t = rep(res_pub$gammas$argvals, n_pub),
  gamma = as.vector(t(res_pub$gammas$data)),
  curve = rep(seq_len(n_pub), each = m_pub),
  sex = rep(gender, each = m_pub)
)

ggplot(df_warps, aes(x = t, y = gamma, group = curve, color = sex)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(alpha = 0.4) +
  scale_color_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "Warping Functions by Sex",
       x = "Reference age (years)",
       y = expression(gamma(t)), color = NULL) +
  theme(legend.position = "top")
```

Each $\gamma_i(t)$ maps from the reference time axis (the Karcher mean) to the
corresponding age in child $i$'s original curve. Curves **below** the diagonal
($\gamma_i(t) < t$) indicate that the growth spurt occurred at a younger age
than the population average -- i.e., earlier maturation. Curves **above** the
diagonal indicate later maturation.

The colour separation confirms the well-known pattern: girls (coral) cluster
below the diagonal, reflecting earlier pubertal timing, while boys (blue)
cluster above.

### Testing the Gender Difference in Timing

The visual pattern suggests boys and girls differ in pubertal timing. We can
test this formally with `fequiv.test()`, which checks whether the mean warping
functions of two groups are equivalent within a margin $\delta$. If equivalence
is *not* declared, we have evidence the timing patterns genuinely differ.

We use $\delta = 0.5$ years -- differences smaller than half a year would be
considered practically negligible:

```{r growth-gender-test}
warps_boys <- km_pub$gammas[gender == "Boy"]
warps_girls <- km_pub$gammas[gender == "Girl"]

set.seed(42)
equiv <- fequiv.test(warps_boys, warps_girls, delta = 0.5,
                     n.boot = 1000, seed = 42)
print(equiv)
```

```{r growth-gender-test-plot, fig.height = 4}
plot(equiv) +
  labs(x = "Age (years)",
       y = expression(hat(gamma)[boys](t) - hat(gamma)[girls](t)))
```

The mean difference curve (solid line) and its simultaneous confidence band
(shaded) both far exceed the $\pm\delta = 0.5$ year margins (dashed lines),
so equivalence is decisively rejected.

The mean warping functions for each group show the separation directly:

```{r growth-mean-warps, fig.height = 4}
t_pub <- warps_boys$argvals
mean_warp_boys <- as.numeric(mean(warps_boys)$data)
mean_warp_girls <- as.numeric(mean(warps_girls)$data)

# Pointwise 95% CI via bootstrap
set.seed(42)
n_boot <- 1000
boot_boys <- matrix(0, n_boot, length(t_pub))
boot_girls <- matrix(0, n_boot, length(t_pub))
for (b in seq_len(n_boot)) {
  boot_boys[b, ] <- colMeans(warps_boys$data[sample(nrow(warps_boys$data),
                              replace = TRUE), ])
  boot_girls[b, ] <- colMeans(warps_girls$data[sample(nrow(warps_girls$data),
                               replace = TRUE), ])
}
ci_boys_lo <- apply(boot_boys, 2, quantile, 0.025)
ci_boys_hi <- apply(boot_boys, 2, quantile, 0.975)
ci_girls_lo <- apply(boot_girls, 2, quantile, 0.025)
ci_girls_hi <- apply(boot_girls, 2, quantile, 0.975)

df_ci <- rbind(
  data.frame(t = t_pub, lo = ci_boys_lo, hi = ci_boys_hi, group = "Boys"),
  data.frame(t = t_pub, lo = ci_girls_lo, hi = ci_girls_hi, group = "Girls")
)
df_mean_warps <- data.frame(
  t = rep(t_pub, 3),
  gamma = c(mean_warp_boys, mean_warp_girls, t_pub),
  group = rep(c("Boys", "Girls", "Identity"), each = length(t_pub))
)

ggplot() +
  geom_ribbon(data = df_ci, aes(x = t, ymin = lo, ymax = hi, fill = group),
              alpha = 0.2) +
  geom_line(data = df_mean_warps,
            aes(x = t, y = gamma, color = group, linetype = group),
            linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(Boys = "steelblue", Girls = "coral",
                                Identity = "grey60")) +
  scale_fill_manual(values = c(Boys = "steelblue", Girls = "coral")) +
  scale_linetype_manual(values = c(Boys = "solid", Girls = "solid",
                                   Identity = "dashed")) +
  labs(title = "Mean Warping Functions by Sex (95% CI)",
       x = "Reference age (years)", y = expression(bar(gamma)(t)),
       color = NULL, linetype = NULL, fill = NULL) +
  guides(fill = "none") +
  theme(legend.position = "top")
```

Girls' mean warping sits below the identity (the growth spurt features occur
at younger ages), while boys' sits above. The vertical gap between the two
curves represents the timing difference at each reference age.

We can quantify the timing gap directly from the original curves:

```{r growth-peak-ages}
peak_ages <- fd_pub$argvals[apply(fd_pub$data, 1, which.max)]
cat("Boys: mean peak age =", round(mean(peak_ages[gender == "Boy"]), 1),
    "years (sd =", round(sd(peak_ages[gender == "Boy"]), 1), ")\n")
cat("Girls: mean peak age =", round(mean(peak_ages[gender == "Girl"]), 1),
    "years (sd =", round(sd(peak_ages[gender == "Girl"]), 1), ")\n")
cat("Difference:", round(mean(peak_ages[gender == "Boy"]) -
    mean(peak_ages[gender == "Girl"]), 1), "years\n")
```

The equivalence test on the warping functions confirms what the peak ages
show: boys and girls have meaningfully different growth timing.

### Working with Warping Functions

The warping functions $\gamma_i$ are `fdata` objects, so they can be used with
any fdars function. Here are concrete ways to extract information from them:

**Scalar summaries per child.** The deviation from the identity
$\gamma_i(t) - t$ summarises how much each child's timing differs from the
reference. Averaging this gives a single "timing shift" score per child:

```{r warp-scores}
# Timing shift: mean deviation from identity (positive = later, negative = earlier)
t_pub <- km_pub$gammas$argvals
identity <- matrix(t_pub, nrow = nrow(km_pub$gammas$data),
                   ncol = length(t_pub), byrow = TRUE)
timing_shift <- rowMeans(km_pub$gammas$data - identity)

cat("Earliest maturer: child", which.min(timing_shift),
    "(shift =", round(min(timing_shift), 2), "years)\n")
cat("Latest maturer:   child", which.max(timing_shift),
    "(shift =", round(max(timing_shift), 2), "years)\n")
```

These scalar scores can be used as covariates in standard regression, or for
correlation with other variables (e.g., adult height, BMI):

```{r warp-boxplot, fig.height = 4}
df_shift <- data.frame(shift = timing_shift, sex = gender)
ggplot(df_shift, aes(x = sex, y = shift, fill = sex)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "Timing Shift Scores by Sex",
       x = NULL, y = "Mean timing shift (years)") +
  guides(fill = "none")
```

**FPCA on warping functions.** The warpings themselves can be analysed as
functional data. Functional PCA on the warping functions identifies the main
modes of timing variation:

```{r warp-fpca}
pc_warps <- fdata2pc(km_pub$gammas, ncomp = 2)
pve_warps <- round(pc_warps$d^2 / sum(pc_warps$d^2) * 100, 1)
cat("Warp PC1:", pve_warps[1], "% — PC2:", pve_warps[2], "%\n")
```

```{r warp-fpca-plot, fig.height = 4}
df_scores <- data.frame(
  PC1 = pc_warps$x[, 1],
  PC2 = pc_warps$x[, 2],
  sex = gender
)
ggplot(df_scores, aes(x = PC1, y = PC2, color = sex)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "PCA of Warping Functions",
       x = paste0("PC1 (", pve_warps[1], "%)"),
       y = paste0("PC2 (", pve_warps[2], "%)"),
       color = NULL)
```

The PC1 score is almost perfectly correlated with the mean timing shift
($r \approx 0.99$) and clearly separates boys from girls. PC2 captures a
secondary mode of timing variation orthogonal to the early/late shift --
its interpretation would require inspecting the eigenfunctions.

## Practical Tips

### Pre-smooth Noisy Data

The SRSF transform involves derivatives, so noise gets amplified. Always smooth
noisy data before alignment:

```{r presmooth}
set.seed(42)
noisy <- matrix(0, 15, 100)
shifts <- runif(15, -0.1, 0.1)
for (i in 1:15) {
  noisy[i, ] <- sin(2 * pi * (argvals - shifts[i])) + rnorm(100, sd = 0.15)
}
fd_noisy <- fdata(noisy, argvals = argvals)

# Smooth with B-spline basis
coefs <- fdata2basis(fd_noisy, nbasis = 15, type = "bspline")
fd_smooth <- basis2fdata(coefs, argvals)

var_before <- mean(apply(fd_noisy$data, 2, var))

# Align raw vs smoothed
res_raw <- elastic.align(fd_noisy)
res_smooth <- elastic.align(fd_smooth)

vr_raw <- 1 - mean(apply(res_raw$aligned$data, 2, var)) / var_before
vr_smooth <- 1 - mean(apply(res_smooth$aligned$data, 2, var)) / var_before
cat("VR (raw):     ", round(vr_raw * 100, 1), "%\n")
cat("VR (smoothed):", round(vr_smooth * 100, 1), "%\n")
```

### Use Sufficient Grid Resolution

The dynamic programming alignment operates on the evaluation grid. Finer grids
yield more accurate warping:

```{r grid-resolution}
set.seed(42)
for (m in c(20, 50, 100, 200)) {
  av <- seq(0, 1, length.out = m)
  shifts <- runif(15, -0.1, 0.1)
  dm <- matrix(0, 15, m)
  for (i in 1:15) dm[i, ] <- sin(2 * pi * (av - shifts[i]))
  fd_m <- fdata(dm, argvals = av)
  ea <- elastic.align(fd_m)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                  mean(apply(fd_m$data, 2, var))
  cat(sprintf("  m = %3d: VR = %5.1f%%\n", m, var_red * 100))
}
```

A minimum of 50-100 grid points is recommended.

### Choose an Appropriate Target

The target curve affects alignment quality. Three options:

| Target | When to use |
|:-------|:------------|
| Cross-sectional mean (default) | General-purpose; robust with many curves |
| Deepest curve | When outliers may distort the mean |
| Known reference | When aligning to a standard template |

```{r target-choice}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)
var_before <- mean(apply(fd$data, 2, var))

# Default (cross-sectional mean)
vr1 <- 1 - mean(apply(elastic.align(fd)$aligned$data, 2, var)) / var_before

# Deepest curve
depths <- depth(fd, method = "FM")
vr2 <- 1 - mean(apply(elastic.align(fd, target = fd[which.max(depths)])$aligned$data, 2, var)) / var_before

cat("Cross-sectional mean: VR =", round(vr1 * 100, 1), "%\n")
cat("Deepest curve:        VR =", round(vr2 * 100, 1), "%\n")
```

### Alignment + FPCA

Alignment before FPCA concentrates variance into the first principal component
by removing phase variability that would otherwise spread across components:

```{r alignment-fpca}
set.seed(42)
shifts <- runif(15, -0.1, 0.1)
dm <- matrix(0, 15, 100)
for (i in 1:15) dm[i, ] <- sin(2 * pi * (argvals - shifts[i]))
fd <- fdata(dm, argvals = argvals)

# FPCA before and after alignment
pc_before <- fdata2pc(fd, ncomp = 3)
ea <- elastic.align(fd)
pc_after <- fdata2pc(ea$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2], "%\n")
```

## Limitations

### Fixed Boundary Conditions

Warping functions are constrained to $\gamma(0) = 0$ and $\gamma(1) = 1$.
This means the alignment cannot correct mismatches at the domain boundaries.
For shifted curves, the residual at the boundary equals the value difference:

```{r boundary}
f1 <- sin(2 * pi * argvals)
f2 <- sin(2 * pi * (argvals - 0.1))

fd2 <- fdata(rbind(f1, f2), argvals = argvals)
ea <- elastic.align(fd2)

# Boundary residual = |sin(0) - sin(-0.2*pi)| = 0.588
cat("Boundary gap:          ", round(abs(f2[1] - f1[1]), 3), "\n")
cat("Max aligned difference:", round(max(abs(ea$aligned$data[1, ] - ea$aligned$data[2, ])), 3), "\n")
cat("Interior max diff:     ", round(max(abs(ea$aligned$data[1, 10:90] - ea$aligned$data[2, 10:90])), 3), "\n")
```

The interior alignment is excellent even when boundaries cannot be corrected.

### Amplitude-Only Differences

When curves differ only in amplitude (e.g., $f$ vs $2f$), alignment correctly
produces identity warping -- no spurious time distortion:

```{r amplitude-only}
f1 <- sin(2 * pi * argvals)
f2 <- 2 * sin(2 * pi * argvals)
fd_amp <- fdata(rbind(f1, f2), argvals = argvals)
res_amp <- elastic.align(fd_amp)

max_warp_dev <- max(abs(res_amp$gammas$data -
                        matrix(argvals, 2, 100, byrow = TRUE)))
cat("Max warp deviation from identity:", max_warp_dev, "\n")
```

This is the correct behavior: amplitude differences should not produce warping.
However, when both amplitude and phase variability are present, the phase correction
may be partial -- alignment focuses on phase and leaves amplitude untouched.

### Domain Independence

The alignment works correctly on any domain, not just $[0, 1]$:

```{r domains}
for (domain in list(c(0, 1), c(0, 2*pi), c(0, 10), c(-1, 1))) {
  av <- seq(domain[1], domain[2], length.out = 100)
  range <- domain[2] - domain[1]
  dm <- matrix(0, 10, 100)
  set.seed(42)
  for (i in 1:10) {
    shift <- runif(1, -0.1 * range, 0.1 * range)
    dm[i, ] <- sin(2 * pi * (av - shift) / range)
  }
  fd <- fdata(dm, argvals = av)
  ea <- elastic.align(fd)
  var_red <- 1 - mean(apply(ea$aligned$data, 2, var)) /
                 mean(apply(fd$data, 2, var))
  cat(sprintf("  [%g, %g]: VR = %.1f%%\n", domain[1], domain[2], var_red * 100))
}
```

## References

- Srivastava, A., Klassen, E., Joshi, S.H., and Jermyn, I.H. (2011). Shape analysis of elastic curves in Euclidean spaces. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 33(7):1415--1428.
- Tucker, J.D., Wu, W., and Srivastava, A. (2013). Generative models for functional data using phase and amplitude separation. *Computational Statistics & Data Analysis*, 61:50--66.
- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data Analysis*. Springer.
