---
title: "Functional Equivalence Testing"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Classical hypothesis tests (like `group.test`) ask whether two groups of curves
are *different*. But in many applications the relevant question is the opposite:
**are two groups of curves equivalent?**

Examples include:

- **Bioequivalence**: Do a generic and brand-name drug produce equivalent
  concentration-time profiles?
- **Process validation**: Does a new manufacturing process produce output curves
  within tolerance of the old one?
- **Reproducibility**: Do two labs produce equivalent functional measurements?

`fequiv.test()` implements a functional **TOST** (Two One-Sided Tests) procedure
based on the supremum norm. It constructs a simultaneous confidence band (SCB)
for the mean difference and checks whether the entire band lies within an
equivalence margin $[-\delta, \delta]$.

**Hypotheses:**

- $H_0$: $\sup_t |\mu_1(t) - \mu_2(t)| \geq \delta$ (NOT equivalent)
- $H_1$: $\sup_t |\mu_1(t) - \mu_2(t)| < \delta$ (equivalent)

```{=html}
<img src="../reference/figures/equivalence-testing-diagram.svg" alt="Functional Equivalence Testing Diagram" style="width:100%; max-width:700px;">
```

## Setup

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())

set.seed(42)
m <- 100
t_grid <- seq(0, 1, length.out = m)
```

## Example 1: Equivalent curves

Two samples drawn from the same distribution should be declared equivalent
with a reasonable $\delta$.

```{r equivalent}
# Two groups from the same process
n1 <- 30
n2 <- 25
X1 <- matrix(0, n1, m)
X2 <- matrix(0, n2, m)
for (i in 1:n1) X1[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.3)
for (i in 1:n2) X2[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.3)

fd1 <- fdata(X1, argvals = t_grid)
fd2 <- fdata(X2, argvals = t_grid)

result <- fequiv.test(fd1, fd2, delta = 0.5, n.boot = 1000, seed = 42)
print(result)
```

The plot shows the mean difference curve (black), the simultaneous confidence
band (green = equivalence declared, red = not declared), and the equivalence
margins (dashed lines at $\pm\delta$).

```{r plot-equivalent}
plot(result)
```

## Example 2: Non-equivalent curves

When the two groups have genuinely different mean functions, the test should
*not* declare equivalence.

```{r non-equivalent}
# Group 2 has a shifted mean
X2_shifted <- matrix(0, n2, m)
for (i in 1:n2) X2_shifted[i, ] <- sin(2 * pi * t_grid) + 0.8 + rnorm(m, sd = 0.3)

fd2_shifted <- fdata(X2_shifted, argvals = t_grid)

result2 <- fequiv.test(fd1, fd2_shifted, delta = 0.5, n.boot = 1000, seed = 42)
print(result2)
```

```{r plot-non-equivalent}
plot(result2)
```

The red band extends well beyond the dashed equivalence margins, so
equivalence is not declared.

## Choosing delta

The equivalence margin $\delta$ is the maximum sup-norm difference you consider
scientifically negligible. It must be chosen *before* looking at the data,
based on domain knowledge:

- **Too large**: trivially declares equivalence (low power against real
  differences)
- **Too small**: almost never declares equivalence (conservative)

A useful diagnostic is to sweep over $\delta$ values and see where the
decision flips:

```{r delta-sweep}
deltas <- seq(0.1, 1.0, by = 0.05)
decisions <- sapply(deltas, function(d) {
  fequiv.test(fd1, fd2, delta = d, n.boot = 500, seed = 42)$reject
})

df_sweep <- data.frame(delta = deltas, equivalent = decisions)
ggplot(df_sweep, aes(x = delta, y = as.numeric(equivalent))) +
  geom_step(linewidth = 0.8) +
  geom_point(aes(color = equivalent), size = 2) +
  scale_color_manual(values = c("FALSE" = "#d62728", "TRUE" = "#2ca02c")) +
  labs(x = expression(delta), y = "Equivalence declared",
       title = "Decision as a function of equivalence margin") +
  scale_y_continuous(breaks = c(0, 1), labels = c("No", "Yes"))
```

## One-sample test

You can also test whether a single sample's mean is equivalent to a known
reference function $\mu_0$.

```{r one-sample}
# Test if the sample mean is equivalent to the true generating function
mu0 <- sin(2 * pi * t_grid)
result_one <- fequiv.test(fd1, delta = 0.5, mu0 = mu0, n.boot = 1000, seed = 42)
print(result_one)
plot(result_one)
```

## Bootstrap methods

`fequiv.test` supports two bootstrap methods:

- **`"multiplier"`** (default): Gaussian multiplier bootstrap. Fast and
  asymptotically valid. Recommended for most applications.
- **`"percentile"`**: Resampling-based bootstrap. More robust with small
  samples or heavy-tailed data.

```{r percentile}
result_pct <- fequiv.test(fd1, fd2, delta = 0.5, n.boot = 1000,
                          method = "percentile", seed = 42)
print(result_pct)
```

## See Also

- `vignette("tolerance-bands", package = "fdars")` — tolerance and confidence bands for functional data

## References

- Dette, H. and Kokot, K. (2021). Detecting relevant differences in the covariance operators of functional time series. *Biometrika*, 108(4):895--913.
