---
title: "Phoneme Recognition: Shape-Based Sound Classification"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

Five phoneme classes --- "aa", "ao", "dcl", "iy", "sh" --- produce distinct
log-periodogram spectra over 150 frequency points. This example applies
**shape-based analysis** to classify phonemes by their spectral geometry
rather than raw amplitude profiles.

| Step | What It Does | Outcome |
|---|---|---|
| Phase variation check | Measure amplitude/phase variance ratio | Confirms that phase variation is present in spectral data |
| Shape distance matrix | Pairwise elastic distances modulo reparameterization | Block-diagonal structure reveals phoneme grouping |
| Shape means | Karcher mean per phoneme class | Canonical spectral signature for each sound |
| MDS visualization | Embed shape distances in 2D | Shape distances yield tighter class clusters than Euclidean |
| Elastic clustering | K-means and hierarchical clustering in shape space | Recovers phoneme classes without label supervision |

**Key finding:** Shape distances produce cleaner class separation than
Euclidean distances on these spectral curves. Elastic k-means achieves
substantially higher cluster purity than standard k-means, and the
shape-based MDS embedding reveals five distinct phoneme clusters.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## 1. The Data

The `phoneme` dataset from the `fda.usc` package contains log-periodogram
spectra recorded from speakers producing five phoneme classes. We use the
learning set of 250 curves (50 per class) evaluated at 150 frequency points.

```{r load-data}
data(phoneme, package = "fda.usc")

# Create fdars fdata from the learning set
fd <- fdata(phoneme$learn$data, argvals = as.numeric(phoneme$learn$argvals))

# Map numeric class labels to standard phoneme names
classes <- phoneme$classlearn
levels(classes) <- c("aa", "ao", "dcl", "iy", "sh")

cat("Curves:", nrow(fd$data), "| Frequency points:", ncol(fd$data), "\n")
cat("Classes:", paste(levels(classes), collapse = ", "),
    "(50 each)\n")
```

```{r plot-all-spectra, fig.height = 4}
# Build a data frame for faceted plotting
n <- nrow(fd$data)
m <- ncol(fd$data)
argvals <- as.numeric(fd$argvals)

df_all <- data.frame(
  curve_id = rep(seq_len(n), each = m),
  freq = rep(argvals, n),
  intensity = as.vector(t(fd$data)),
  class = rep(classes, each = m)
)

ggplot(df_all, aes(x = .data$freq, y = .data$intensity,
                   group = .data$curve_id, color = .data$class)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  facet_wrap(~ .data$class, ncol = 5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Log-Periodogram Spectra by Phoneme Class",
       x = "Frequency", y = "Log-Periodogram (dB)") +
  guides(color = "none") +
  theme(strip.text = element_text(face = "bold"))
```

Each phoneme class produces a characteristic spectral shape. The vowels
"aa" and "ao" show broad low-frequency energy, "sh" (fricative) peaks at
high frequencies, and "dcl" and "iy" have intermediate patterns. Despite
clear group structure, there is substantial within-class variability.

## 2. Phase Variation Check

We compute the Karcher mean and alignment quality for one phoneme class to
check whether meaningful phase variation is present.

```{r phase-check}
set.seed(42)

# Use the "aa" class (first 50 curves) for the phase check
fd_aa <- fd[classes == "aa", ]

# Elastic alignment via Karcher mean
km_aa <- karcher.mean(fd_aa, max.iter = 15, tol = 1e-4)
aq_aa <- alignment.quality(fd_aa, km_aa)

print(aq_aa)
```

```{r phase-plot, fig.height = 4}
plot(aq_aa, type = "variance") +
  labs(title = "Amplitude vs Phase Variance ('aa' Phoneme)")
```

A non-trivial fraction of total variability is due to phase differences ---
speakers produce the same phoneme with slightly shifted frequency peaks.
Shape analysis, which factors out reparameterization, is appropriate here.

## 3. Shape Distance Matrix

We take a balanced subset of 50 curves (10 per class) for computational
tractability.

```{r subset-data}
set.seed(42)

# Select 10 curves per class
idx <- unlist(lapply(levels(classes), function(k) {
  which(classes == k)[1:10]
}))

fd_sub <- fd[idx, ]
cls_sub <- classes[idx]
n_sub <- length(idx)

cat("Subset size:", n_sub, "curves\n")
cat("Per class:", table(cls_sub), "\n")
```

```{r compute-distances}
# Shape distance matrix (elastic, modulo reparameterization)
D_shape <- shape.distance.matrix(fd_sub, quotient = "reparameterization")

# Euclidean (L2) distance matrix for comparison
D_euclid <- metric.lp(fd_sub, p = 2)
```

```{r heatmap-comparison, fig.height = 5}
# Helper to convert distance matrix to long format
dist_to_long <- function(D, name) {
  nn <- nrow(D)
  df <- expand.grid(i = seq_len(nn), j = seq_len(nn))
  df$dist <- as.vector(D)
  df$type <- name
  df
}

df_euclid <- dist_to_long(D_euclid, "Euclidean Distance")
df_shape <- dist_to_long(D_shape, "Shape Distance")
df_heat <- rbind(df_euclid, df_shape)
df_heat$type <- factor(df_heat$type,
                       levels = c("Euclidean Distance", "Shape Distance"))

# Add class annotation breaks
class_breaks <- cumsum(table(cls_sub)) + 0.5

p_heat <- ggplot(df_heat, aes(x = .data$i, y = .data$j, fill = .data$dist)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "Distance") +
  facet_wrap(~ .data$type) +
  geom_hline(yintercept = class_breaks, color = "white",
             linewidth = 0.3, linetype = "dashed") +
  geom_vline(xintercept = class_breaks, color = "white",
             linewidth = 0.3, linetype = "dashed") +
  coord_equal() +
  labs(title = "Euclidean vs Shape Distance Matrices",
       subtitle = "Dashed lines mark class boundaries (10 curves each)",
       x = "Curve Index", y = "Curve Index") +
  theme(axis.text = element_text(size = 6))

p_heat
```

Both matrices show block-diagonal structure, but the shape distance matrix
has sharper within/between-class contrast because it collapses phase-shifted
variants of the same phoneme to near-zero distance.

## 4. Shape Means

The shape mean (Karcher mean in the quotient space) provides a canonical
spectral signature per class, not blurred by phase variation.

```{r compute-shape-means}
shape_means <- list()
for (k in seq_along(levels(classes))) {
  fd_k <- fd[classes == levels(classes)[k], ]
  shape_means[[k]] <- shape.mean(fd_k, quotient = "reparameterization",
                                  max.iter = 15, tol = 1e-4)
}
names(shape_means) <- levels(classes)
```

```{r plot-shape-means, fig.height = 6}
# Build data frame: individual curves (light) + shape means (bold)
df_means_all <- do.call(rbind, lapply(levels(classes), function(lev) {
  fd_k <- fd[classes == lev, ]
  n_k <- nrow(fd_k$data)
  rbind(
    data.frame(curve_id = rep(seq_len(n_k), each = m), freq = rep(argvals, n_k),
               intensity = as.vector(t(fd_k$data)), class = lev, is_mean = FALSE),
    data.frame(curve_id = 0L, freq = argvals,
               intensity = as.numeric(shape_means[[lev]]$mean),
               class = lev, is_mean = TRUE)
  )
}))

ggplot(df_means_all, aes(x = .data$freq, y = .data$intensity)) +
  geom_line(data = df_means_all[!df_means_all$is_mean, ],
            aes(group = interaction(.data$class, .data$curve_id)),
            alpha = 0.15, color = "grey60", linewidth = 0.2) +
  geom_line(data = df_means_all[df_means_all$is_mean, ],
            color = "#D55E00", linewidth = 1.2) +
  facet_wrap(~ .data$class, ncol = 5) +
  labs(title = "Shape Means by Phoneme Class",
       subtitle = "Red = Karcher mean; grey = individual spectra",
       x = "Frequency", y = "Log-Periodogram (dB)") +
  theme(strip.text = element_text(face = "bold"))
```

Each shape mean captures the distinctive spectral profile for its phoneme,
sharper than a pointwise average because phase variation has been removed.

```{r plot-means-overlay, fig.height = 4}
# Overlay all five shape means
df_overlay <- do.call(rbind, lapply(levels(classes), function(lev) {
  data.frame(freq = argvals, intensity = as.numeric(shape_means[[lev]]$mean),
             class = lev)
}))

ggplot(df_overlay, aes(x = .data$freq, y = .data$intensity,
                       color = .data$class)) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Shape Mean Comparison Across Phonemes",
       x = "Frequency", y = "Log-Periodogram (dB)",
       color = "Phoneme") +
  theme(legend.position = "bottom")
```

The overlay shows how the five phonemes partition the frequency spectrum,
with vowels in different formant regions and consonants showing distinct
spectral profiles.

## 5. MDS Visualization

We embed the pairwise distance matrices in 2D via classical MDS.

```{r mds-embed}
mds_shape  <- cmdscale(D_shape, k = 2)
mds_euclid <- cmdscale(D_euclid, k = 2)
```

```{r mds-plot, fig.height = 4.5}
df_mds_euclid <- data.frame(MDS1 = mds_euclid[, 1], MDS2 = mds_euclid[, 2],
                              class = cls_sub)
df_mds_shape  <- data.frame(MDS1 = mds_shape[, 1], MDS2 = mds_shape[, 2],
                              class = cls_sub)

mds_panel <- function(df, title) {
  ggplot(df, aes(x = .data$MDS1, y = .data$MDS2, color = .data$class)) +
    geom_point(size = 2.5, alpha = 0.8) +
    scale_color_brewer(palette = "Set1") +
    labs(title = title, x = "MDS Dim 1", y = "MDS Dim 2",
         color = "Phoneme") +
    coord_equal()
}

mds_panel(df_mds_euclid, "MDS: Euclidean Distance") +
  mds_panel(df_mds_shape, "MDS: Shape Distance") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The shape-based MDS embedding shows tighter, more separated clusters because
the shape metric collapses within-class phase variation. The Euclidean
embedding spreads classes more diffusely.

## 6. Shape-Based Clustering

We compare elastic k-means (Fisher-Rao distance, Karcher mean updates)
with standard Euclidean k-means.

```{r elastic-kmeans}
set.seed(42)

# Elastic k-means on the subset
ekm <- elastic.kmeans(fd_sub, k = 5, max.iter = 50, seed = 42)
print(ekm)
```

```{r plot-elastic-kmeans, fig.height = 5}
plot(ekm) +
  labs(title = "Elastic K-Means Clusters (k = 5)",
       x = "Frequency", y = "Log-Periodogram (dB)")
```

The elastic k-means plot shows each cluster's Karcher mean (red) with
member curves in grey. Each cluster captures a distinct spectral shape
corresponding to a phoneme class.

```{r standard-kmeans}
set.seed(42)

# Standard Euclidean k-means for comparison
km_std <- cluster.kmeans(fd_sub, ncl = 5, metric = "L2",
                          nstart = 10, seed = 42)
```

```{r cluster-comparison-plot, fig.height = 5}
# Side-by-side cluster assignments
m_sub <- ncol(fd_sub$data)
argvals_sub <- as.numeric(fd_sub$argvals)

df_ekm <- data.frame(
  curve_id = rep(seq_len(n_sub), each = m_sub),
  freq = rep(argvals_sub, n_sub),
  intensity = as.vector(t(fd_sub$data)),
  cluster = factor(rep(ekm$labels, each = m_sub)),
  method = "Elastic K-Means"
)

df_std <- data.frame(
  curve_id = rep(seq_len(n_sub), each = m_sub),
  freq = rep(argvals_sub, n_sub),
  intensity = as.vector(t(fd_sub$data)),
  cluster = factor(rep(km_std$cluster, each = m_sub)),
  method = "Standard K-Means"
)

df_compare <- rbind(df_ekm, df_std)

ggplot(df_compare, aes(x = .data$freq, y = .data$intensity,
                       group = .data$curve_id, color = .data$cluster)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  facet_wrap(~ .data$method) +
  labs(title = "Cluster Assignments: Elastic vs Standard K-Means",
       x = "Frequency", y = "Log-Periodogram (dB)",
       color = "Cluster") +
  theme(legend.position = "bottom")
```

Elastic k-means clusters are more homogeneous in shape. Standard k-means
may mix classes that share amplitude levels but differ in spectral shape.

## 7. Clustering Accuracy

Purity measures the fraction of each cluster belonging to the majority class.

```{r purity-metric}
purity <- function(clusters, true_classes) {
  tab <- table(clusters, true_classes)
  sum(apply(tab, 1, max)) / sum(tab)
}

purity_elastic <- purity(ekm$labels, cls_sub)
purity_standard <- purity(km_std$cluster, cls_sub)

cat("Elastic k-means purity: ", round(purity_elastic, 3), "\n")
cat("Standard k-means purity:", round(purity_standard, 3), "\n")
```

```{r confusion-tables}
cat("\n--- Elastic K-Means ---\n")
print(table(Elastic = ekm$labels, True = cls_sub))

cat("\n--- Standard K-Means ---\n")
print(table(Standard = km_std$cluster, True = cls_sub))
```

```{r purity-barplot, fig.height = 3.5}
df_purity <- data.frame(
  method = c("Standard K-Means", "Elastic K-Means"),
  purity = c(purity_standard, purity_elastic)
)

ggplot(df_purity, aes(x = reorder(.data$method, .data$purity),
                      y = .data$purity, fill = .data$method)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = sprintf("%.1f%%", .data$purity * 100)),
            hjust = -0.2, size = 4) +
  scale_fill_manual(values = c("Standard K-Means" = "#4A90D9",
                                "Elastic K-Means" = "#D55E00")) +
  coord_flip(ylim = c(0, 1.1)) +
  labs(title = "Cluster Purity: Elastic vs Standard",
       x = "", y = "Purity") +
  guides(fill = "none")
```

Elastic k-means achieves higher purity because it groups curves by spectral
shape rather than pointwise amplitude.

## 8. Hierarchical View

A dendrogram reveals relationships between phoneme classes at multiple
granularities.

```{r elastic-hclust}
ehc <- elastic.hclust(fd_sub, method = "complete")
print(ehc)
```

```{r plot-dendrogram, fig.height = 5}
plot(ehc) +
  labs(title = "Elastic Hierarchical Clustering Dendrogram",
       subtitle = "Complete linkage on Fisher-Rao distances")
```

```{r cutree-compare}
hc_labels <- elastic.cutree(ehc, k = 5)

purity_hc <- purity(hc_labels, cls_sub)
cat("Hierarchical clustering purity:", round(purity_hc, 3), "\n")

cat("\n--- Hierarchical Clustering (k = 5) ---\n")
print(table(HClust = hc_labels, True = cls_sub))
```

Vowels tend to merge at lower heights than phoneme pairs that differ in
manner of articulation. Cutting at k = 5 recovers classes with purity
comparable to elastic k-means.

```{r summary-table, fig.height = 3}
df_summary <- data.frame(
  Method = c("Standard K-Means (L2)", "Elastic K-Means",
             "Elastic Hierarchical"),
  Purity = c(purity_standard, purity_elastic, purity_hc)
)

knitr::kable(df_summary, digits = 3,
             caption = "Clustering Purity Comparison (50 curves, 5 classes)")
```

## 9. Conclusions

The elastic (Fisher-Rao) distance captures phonetically meaningful
differences that Euclidean metrics miss. Shape distances yield cleaner
block-diagonal structure, better MDS separation, and higher clustering
purity. For production-scale recognition, the shape means or distance
matrix computed here can serve as features in a supervised classifier.

## See Also

- `vignette("articles/shape-analysis")` --- shape space theory,
  orbit representatives, and shape distance computation
- `vignette("articles/elastic-clustering")` --- elastic k-means and
  hierarchical clustering on simulated data
- `vignette("articles/distance-metrics")` --- comparison of functional
  distance measures

## References

- Ferraty, F. and Vieu, P. (2006). *Nonparametric Functional Data Analysis*.
  Springer. Chapter on the phoneme dataset.

- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data
  Analysis*. Springer.

- Kurtek, S., Srivastava, A., Klassen, E. and Ding, Z. (2012). Statistical
  Modeling of Curves Using Shapes and Related Features. *Journal of the
  American Statistical Association*, 107(499), 1152--1165.
