---
title: "Andrews Wine: Outlier Detection"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(fdars)
library(ggplot2)
library(dplyr)
library(patchwork)
library(knitr)
theme_set(theme_minimal(base_size = 13))
set.seed(42)
```

This article applies three outlier detection methods to 178 wines transformed
into Andrews curves. For background on *why* Andrews curves and how the
transformation works, see
[Why Andrews Curves?](example-andrews-wine-intro.html).

| Article | What It Does | Outcome |
|---|---|---|
| [Why Andrews Curves?](example-andrews-wine-intro.html) | Transform 13 chemicals into curves; verify distance preservation | Each wine becomes a visual fingerprint; distances preserved exactly |
| **Outlier Detection** (this article) | Depth, outliergram, MS-plot | 9 anomalies classified by type --- mislabel, soil anomaly, or concentration --- with corrective actions |
| [Clustering & Variable Importance](example-andrews-wine-clustering.html) | K-means, fuzzy c-means, permutation test, FPCA | Cultivar recovery at 96% accuracy; top 5 chemicals identified for cost reduction |
| [Quality Control](example-andrews-wine-qc.html) | Functional boxplots, depth rankings, tolerance bands | Monitoring system that checks new batches against a validated specification in one chart |

```{r shared-setup}
# Load wine data
wine_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"
col_names <- c("Cultivar", "Alcohol", "MalicAcid", "Ash", "Alkalinity",
               "Magnesium", "Phenols", "Flavanoids", "NonflavPhenols",
               "Proanthocyanins", "ColorIntensity", "Hue",
               "OD280_OD315", "Proline")
wine <- read.csv(wine_url, header = FALSE, col.names = col_names)
cultivar <- factor(wine$Cultivar, levels = 1:3,
                   labels = c("Barolo", "Grignolino", "Barbera"))
X <- scale(as.matrix(wine[, -1]))
chem_names <- colnames(wine)[-1]

# Andrews transform
fd_wine <- andrews_transform(X)
n <- nrow(fd_wine$data)
m <- ncol(fd_wine$data)
t_grid <- fd_wine$argvals

df_curves <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_wine$data)),
  curve = rep(1:n, each = m),
  Cultivar = rep(cultivar, each = m)
)
```

## Finding the Unusual Wines

In wine production, "unusual" can mean a quality problem, a labeling error, or
outright fraud. A standard approach might flag individual chemicals that fall
outside a reference range --- but that misses wines whose *combination* of values
is strange, even when no single measurement is extreme. Because Andrews curves
encode *all 13 chemicals simultaneously*, outlier detection on curves catches
multivariate anomalies that univariate checks miss.

We apply three complementary methods. Each asks a different question about
"unusualness," and each gives the quality controller a different kind of
actionable insight.

### Method 1: Depth-Ponderation — "How far from the crowd?"

Think of *depth* as the opposite of outlyingness. The most "typical" wine ---
the one whose chemical profile sits squarely in the middle of the population ---
has the highest depth. Wines at the fringes have low depth. The
depth-ponderation method (`outliers.depth.pond()`) stabilizes this measurement
using **bootstrap resampling**: it repeatedly draws random subsets of the 178
wines, computes depth each time, and averages the results. This smoothing
prevents a single unlucky sample from distorting the picture. Wines whose
averaged depth falls below a threshold are flagged.

```{r outliers-pond}
set.seed(123)
out_pond <- outliers.depth.pond(fd_wine, nb = 500)
cat("Depth-ponderation outliers:", sort(out_pond$outliers), "\n")
cat("Cultivars:", as.character(cultivar[out_pond$outliers]), "\n")
```

```{r pond-highlight, fig.height = 5}

is_pond_outlier <- rep(FALSE, n)
is_pond_outlier[out_pond$outliers] <- TRUE

ggplot() +
  geom_line(data = filter(df_curves, !rep(is_pond_outlier, each = m)),
            aes(x = t, y = value, group = curve),
            color = "gray70", alpha = 0.3, linewidth = 0.3) +
  geom_line(data = filter(df_curves, rep(is_pond_outlier, each = m)),
            aes(x = t, y = value, group = curve, color = Cultivar),
            linewidth = 0.7, alpha = 0.8) +
  scale_color_manual(values = c("Barolo" = "#8B0000", "Grignolino" = "#DAA520",
                                 "Barbera" = "#2E8B57")) +
  labs(title = "Depth-Ponderation Outliers",
       subtitle = sprintf("%d wines flagged — curves far from the population center",
                          length(out_pond$outliers)),
       x = expression(t), y = expression(f[x](t))) +
  theme(legend.position = "bottom")
```

What makes these wines unusual? Let's look at which chemicals push them to the
fringes:

```{r pond-extremes}
# For each flagged wine, identify chemicals with |z-score| > 2
extreme_list <- lapply(sort(out_pond$outliers), function(i) {
  extreme_idx <- which(abs(X[i, ]) > 2)
  if (length(extreme_idx) > 0) {
    chems <- paste(sprintf("%s (z=%.1f)", colnames(wine)[-1][extreme_idx],
                           X[i, extreme_idx]), collapse = ", ")
  } else {
    chems <- "no individual extreme — unusual *combination*"
  }
  data.frame(Wine = i, Cultivar = as.character(cultivar[i]),
             `Extreme Chemicals` = chems, check.names = FALSE)
})
kable(bind_rows(extreme_list),
      caption = "What makes each depth-ponderation outlier unusual (|z| > 2)")
```

Notice that several wines have no individual extreme value --- they're outliers
because of an unusual *combination* of chemicals. This is exactly the kind of
anomaly that a column-by-column check would miss.

### Method 2: Magnitude-Shape Plot — "Unusual *how*?"

The depth-ponderation method tells you *that* a wine is unusual, but not *how*.
The magnitude-shape decomposition (`magnitudeshape()`) splits outlyingness into
two independent components:

- **Magnitude outlyingness (MO)** measures whether a wine's curve is
  systematically shifted up or down relative to the median wine. In chemical
  terms: are *all* values consistently high or consistently low? A high MO
  suggests a global issue --- perhaps a measurement calibration error, a
  different vintage mixed in, or dilution/concentration.

- **Variability outlyingness (VO)** measures whether the *shape* of a wine's
  curve deviates from the population's typical shape, regardless of its overall
  level. In chemical terms: are the *ratios* between chemicals unusual? A high
  VO suggests a process anomaly --- perhaps a fermentation gone wrong, soil
  contamination, or blending with a different cultivar.

The method plots every wine in this two-dimensional (MO, VO) space and draws a
$\chi^2$-based cutoff ellipse. Wines outside the ellipse are flagged.

```{r ms-plot}
ms <- magnitudeshape(fd_wine)
plot(ms)
```

With the default 99.3% cutoff, no wines are flagged here. This is informative,
not disappointing --- it tells us that the outliers found by depth-ponderation
are **not extreme in the magnitude-shape sense**. They sit at the fringes of
their cultivar group, but their chemical *ratios* are still within the normal
range. For a quality controller, this is reassuring: these wines are unusual but
not *suspiciously* unusual. If the MS-plot *did* flag a wine, that would be a
stronger alarm --- possibly indicating adulteration or a process failure that
distorts the fundamental chemical fingerprint.

### Method 3: Outliergram — "Unusual in rank structure?"

The outliergram (`outliergram()`) takes a different angle entirely. It computes
two rank-based statistics for each wine:

- **Modified Epigraph Index (MEI)**: what proportion of the time does this
  wine's curve sit *above* other curves? A wine with MEI near 0.5 is average in
  its vertical position; a wine near 0 or 1 is persistently low or high.

- **Modified Band Depth (MBD)**: how often does this wine's curve fall *within
  the band* formed by pairs of other curves? High MBD means the wine has a
  typical profile; low MBD means it tends to cross or escape the population
  envelope.

For a well-behaved population, MEI and MBD follow a tight parabolic
relationship: wines that are vertically central (MEI ≈ 0.5) should also be
deeply enclosed (high MBD). The outliergram flags wines that deviate from this
expected relationship --- wines whose MBD is *lower than expected* given their
MEI. These are **shape outliers**: wines that weave in and out of the population
bundle in an unusual way.

```{r outliergram}
og <- outliergram(fd_wine)
plot(og)
```

```{r outliergram-detail}
if (length(og$outliers) > 0) {
  cat("Outliergram outliers:", sort(og$outliers), "\n")
  cat("Cultivars:", as.character(cultivar[og$outliers]), "\n")
  for (i in sort(og$outliers)) {
    extreme_idx <- which(abs(X[i, ]) > 2)
    if (length(extreme_idx) > 0) {
      cat(sprintf("  Wine %d: %s\n", i,
                  paste(sprintf("%s (z=%.1f)", colnames(wine)[-1][extreme_idx],
                                X[i, extreme_idx]), collapse = ", ")))
    }
  }
}
```

The outliergram flags the most severe shape anomalies --- wines whose internal
chemical relationships are unusual even though their overall levels may be
plausible. For a quality controller, a shape outlier warrants a different
investigation than a magnitude outlier: you wouldn't re-calibrate the
instrument; you'd investigate the vineyard, the fermentation, or the supply
chain.

### Cross-Referencing: The Confidence Triage

No single outlier detection method is perfect. Each trades off sensitivity
against specificity, and each is blind to certain kinds of anomaly. The real
power comes from **cross-referencing**: wines flagged by multiple methods are
the highest-priority cases for investigation.

```{r outlier-table}
# Collect flags per observation
outlier_flags <- data.frame(
  Wine = 1:n,
  Cultivar = cultivar,
  DepthPond = 1:n %in% out_pond$outliers,
  MagnShape = 1:n %in% ms$outliers,
  Outliergram = 1:n %in% og$outliers
)
outlier_flags$n_methods <- rowSums(outlier_flags[, 3:5])

# Show wines flagged by at least one method
flagged <- outlier_flags |> filter(n_methods >= 1) |> arrange(desc(n_methods))

cat(nrow(flagged), "wines flagged by at least one method,",
    sum(flagged$n_methods >= 2), "by two or more\n\n")

# Show flagged wines with their original chemical values
flagged_detail <- flagged |>
  mutate(across(c(DepthPond, MagnShape, Outliergram),
                ~ ifelse(., "✓", ""))) |>
  left_join(
    data.frame(Wine = 1:n, round(wine[, -1], 1)),
    by = "Wine"
  )

kable(flagged_detail |> select(Wine, Cultivar, DepthPond, MagnShape,
                                Outliergram, n_methods,
                                Alcohol, Flavanoids, ColorIntensity, Proline),
      caption = "Flagged wines with key chemical values")
```

### Deep Dive: The Grignolino Problem

Before we reach for generic conclusions, let's look at the data. Eight of nine
outliers are Grignolino --- zero from Barolo, one from Barbera. That asymmetry
is itself an important finding. Let's understand what's driving it.

```{r outlier-cultivar-breakdown}
fda_all_idx <- unique(c(out_pond$outliers, og$outliers, ms$outliers))

cat("Outliers by cultivar:\n")
for (cv in levels(cultivar)) {
  cv_n <- sum(cultivar == cv)
  cv_out <- sum(cultivar[fda_all_idx] == cv)
  cat(sprintf("  %s: %d outliers out of %d wines (%.0f%%)\n",
              cv, cv_out, cv_n, 100 * cv_out / cv_n))
}
```

Grignolino has by far the most variable chemical profile. Why? Let's examine
what the flagged wines have in common. For each outlier, we compute its
deviation from its own cultivar mean (in within-cultivar standard deviations)
--- this tells us how unusual a wine is *relative to its peer group*, not just
globally:

```{r within-cultivar-deviation}
# Compute within-cultivar z-scores for each outlier
outlier_profiles <- lapply(sort(fda_all_idx), function(i) {
  cv <- cultivar[i]
  cv_idx <- which(cultivar == cv)
  cv_mean <- colMeans(wine[cv_idx, -1])
  cv_sd <- apply(wine[cv_idx, -1], 2, sd)
  dev <- (as.numeric(wine[i, -1]) - cv_mean) / cv_sd
  names(dev) <- chem_names

  # Flag chemicals deviating > 1.5 within-cultivar SDs
  extreme <- which(abs(dev) > 1.5)
  data.frame(
    Wine = i,
    Cultivar = as.character(cv),
    `Extreme Chemicals (> 1.5 cultivar SD)` =
      if (length(extreme) > 0)
        paste(sprintf("%s (%+.1f)", chem_names[extreme], dev[extreme]),
              collapse = ", ")
      else "none",
    n_extreme = length(extreme),
    check.names = FALSE
  )
})

kable(bind_rows(outlier_profiles),
      caption = "Each outlier's unusual chemicals, measured against its own cultivar")
```

A pattern jumps out: **Magnesium** appears in almost every Grignolino outlier.
These wines have Magnesium levels 2--4 standard deviations above the Grignolino
mean. Typical Grignolino has 94.5 mg/L; these outliers range from 107 to 162
mg/L. Other recurring deviations include high Proanthocyanins and unusual
Alkalinity.

This is exactly the kind of systemic insight that a per-variable range check
would bury. Yes, a spreadsheet filter could flag "Magnesium > 130," but it
wouldn't reveal that these high-Magnesium wines *also* tend to have unusual
Proanthocyanins, or that they cluster together in the Fourier domain. The curve
representation lets us see the *compound pattern*.

### Case Study: Wine 122 — Mislabeled or Just Unusual?

Wine 122 is the most extreme outlier: it ranks dead last in depth among all 71
Grignolinos (rank 71/71). Let's look at where it sits relative to all three
cultivar groups:

```{r case-122}
# Distance from Wine 122 to each cultivar median
case_wine <- 122

dist_to_median <- sapply(levels(cultivar), function(cv) {
  cv_idx <- which(cultivar == cv)
  fd_cv <- fd_wine[cv_idx]
  depths_cv <- depth.MBD(fd_cv)
  median_global <- cv_idx[which.max(depths_cv)]
  sqrt(sum((fd_wine$data[case_wine, ] - fd_wine$data[median_global, ])^2) *
       diff(fd_wine$argvals[1:2]))
})

cat(sprintf("Wine %d (%s) — distance to each cultivar median:\n", case_wine,
            cultivar[case_wine]))
for (cv in names(dist_to_median)) {
  marker <- ifelse(cv == as.character(cultivar[case_wine]), " <-- own cultivar", "")
  cat(sprintf("  %s: %.2f%s\n", cv, dist_to_median[cv], marker))
}
closest <- names(which.min(dist_to_median))
cat(sprintf("\nClosest to: %s\n", closest))
```

```{r case-122-profile, fig.height = 5}

# Highlight Wine 122 against its claimed cultivar and the cultivar it resembles
is_case <- rep(FALSE, n)
is_case[case_wine] <- TRUE

ggplot() +
  # Background: all curves, faded
  geom_line(data = filter(df_curves, Cultivar %in% c("Barolo", "Grignolino") &
                            !rep(is_case, each = m)),
            aes(x = t, y = value, group = curve, color = Cultivar),
            alpha = 0.15, linewidth = 0.3) +
  # Case wine: bold black
  geom_line(data = filter(df_curves, rep(is_case, each = m)),
            aes(x = t, y = value, group = curve),
            color = "black", linewidth = 1.2) +
  scale_color_manual(values = c("Barolo" = "#8B0000", "Grignolino" = "#DAA520")) +
  labs(title = sprintf("Wine %d: Labeled Grignolino, Looks Like Barolo", case_wine),
       x = expression(t), y = expression(f[x](t))) +
  theme(legend.position = "bottom")
```

```{r case-122-chemistry}
# Show its chemical profile vs both cultivar means
cv_means <- sapply(c("Grignolino", "Barolo"), function(cv) {
  colMeans(wine[cultivar == cv, -1])
})

case_vals <- as.numeric(wine[case_wine, -1])

compare_122 <- data.frame(
  Chemical = chem_names,
  `Wine 122` = round(case_vals, 1),
  `Grignolino Mean` = round(cv_means[, "Grignolino"], 1),
  `Barolo Mean` = round(cv_means[, "Barolo"], 1),
  `Closer To` = ifelse(
    abs(case_vals - cv_means[, "Grignolino"]) < abs(case_vals - cv_means[, "Barolo"]),
    "Grignolino", "Barolo"),
  check.names = FALSE
)

kable(compare_122, caption = "Wine 122 vs cultivar means — closer to Barolo on most chemicals")
```

Wine 122 is closer to the Barolo median than to its own Grignolino median. On
13 chemicals, it resembles Barolo on the majority. Its Flavanoids (5.1) are
3× the Grignolino average (2.1) but right at the Barolo average (2.98). Its Ash
(3.2) and Alkalinity (28.5) are also Barolo-typical.

**Quality controller's action**: this wine warrants a label audit. Either it was
mislabeled during data collection, it comes from a vineyard on the
Barolo-Grignolino boundary, or it was blended. The FDA analysis provides
the evidence: it's not just "unusual" --- it's unusual *in the direction of a
specific other cultivar*. That's a much more actionable finding than "Mahalanobis
distance = 58.7."

### Case Study: Wines 70 and 96 — Shape Outliers Under the Microscope

These are the only wines flagged by *both* depth-ponderation and the
outliergram. The outliergram specifically flags **shape anomalies** --- wines
whose curve crosses the population envelope in unexpected ways, even if their
average level is unremarkable.

```{r case-70-96}
shape_cases <- c(70, 96)

# Their key chemistry
for (i in shape_cases) {
  cat(sprintf("\n--- Wine %d (%s) ---\n", i, cultivar[i]))
  cat(sprintf("  Magnesium:        %.0f mg/L  (Grignolino mean: 94.5, z = %+.1f)\n",
              wine[i, "Magnesium"], X[i, "Magnesium"]))
  cat(sprintf("  Proanthocyanins:  %.1f       (Grignolino mean: 1.6,  z = %+.1f)\n",
              wine[i, "Proanthocyanins"], X[i, "Proanthocyanins"]))
  cat(sprintf("  Ash:              %.1f g/L   (Grignolino mean: 2.2,  z = %+.1f)\n",
              wine[i, "Ash"], X[i, "Ash"]))

  # Fuzzy membership
  fcm_all <- cluster.fcm(fd_wine, ncl = 3, seed = 123)
  mem <- round(fcm_all$membership[i, ], 3)
  cat(sprintf("  Fuzzy membership: [%.3f, %.3f, %.3f] — max = %.3f\n",
              mem[1], mem[2], mem[3], max(mem)))
}
```

```{r case-70-96-plot, fig.height = 5}

is_shape_case <- rep(FALSE, n)
is_shape_case[shape_cases] <- TRUE

ggplot() +
  geom_line(data = filter(df_curves, !rep(is_shape_case, each = m)),
            aes(x = t, y = value, group = curve, color = Cultivar),
            alpha = 0.12, linewidth = 0.3) +
  geom_line(data = filter(df_curves, rep(is_shape_case, each = m)),
            aes(x = t, y = value, group = curve),
            color = "black", linewidth = 1.1) +
  scale_color_manual(values = c("Barolo" = "#8B0000", "Grignolino" = "#DAA520",
                                 "Barbera" = "#2E8B57")) +
  labs(title = "Shape Outliers: Wines 70 and 96",
       subtitle = "Flagged by both depth-ponderation and outliergram",
       x = expression(t), y = expression(f[x](t))) +
  theme(legend.position = "bottom")
```

Both wines have extremely high Magnesium (151 and 162 mg/L vs a Grignolino
average of 94.5). But they also deviate on *different* secondary chemicals ---
Wine 70 has very low Ash and NonflavPhenols, while Wine 96 has extreme
Proanthocyanins (3.3, vs mean 1.6). What makes them *shape* outliers rather
than mere magnitude outliers is that this combination of high and low values
creates curves that *cross* the normal Grignolino band, dipping below it in
some Fourier regions and rising above it in others.

**Quality controller's action**: investigate the vineyard. Extreme Magnesium in
an otherwise Grignolino-like wine suggests a **soil chemistry anomaly** ---
Magnesium uptake in grapes is strongly influenced by soil composition and
rootstock. These wines may come from a plot with unusual mineral content. The
shape anomaly means a simple recalibration won't fix the issue; the underlying
growing conditions need review.

### Case Study: Wine 159 — The Lone Barbera Outlier

Wine 159 is the only Barbera flagged, and it's an outlier for very different
reasons than the Grignolinos:

```{r case-159}
i <- 159
cat(sprintf("Wine %d (%s)\n", i, cultivar[i]))
cat(sprintf("  Color Intensity: %.1f  (Barbera mean: 7.4, population z = %+.1f)\n",
            wine[i, "ColorIntensity"], X[i, "ColorIntensity"]))
cat(sprintf("  Alcohol:         %.1f%%  (Barbera mean: 13.2%%, z = %+.1f)\n",
            wine[i, "Alcohol"], X[i, "Alcohol"]))
cat(sprintf("  Phenols:         %.1f  (Barbera mean: 1.7, z = %+.1f)\n",
            wine[i, "Phenols"], X[i, "Phenols"]))
cat(sprintf("  Proanthocyanins: %.1f  (Barbera mean: 1.2, z = %+.1f)\n",
            wine[i, "Proanthocyanins"], X[i, "Proanthocyanins"]))

# Within-cultivar rank
cv_idx <- which(cultivar == "Barbera")
fd_cv <- fd_wine[cv_idx]
depths_cv <- depth.MBD(fd_cv)
pos_in_cv <- which(cv_idx == i)
rank_in_cv <- rank(-depths_cv)[pos_in_cv]
cat(sprintf("\n  Depth rank within Barbera: %d / %d (dead last)\n",
            rank_in_cv, length(cv_idx)))
```

```{r case-159-plot, fig.height = 5}

is_159 <- rep(FALSE, n)
is_159[159] <- TRUE

ggplot() +
  geom_line(data = filter(df_curves, Cultivar == "Barbera" & !rep(is_159, each = m)),
            aes(x = t, y = value, group = curve),
            color = "#2E8B57", alpha = 0.3, linewidth = 0.4) +
  geom_line(data = filter(df_curves, rep(is_159, each = m)),
            aes(x = t, y = value, group = curve),
            color = "black", linewidth = 1.2) +
  labs(title = "Wine 159: The Barbera Outlier",
       subtitle = "Extreme Color Intensity (z = +3.4) pushes it to the edge of the Barbera envelope",
       x = expression(t), y = expression(f[x](t)))
```

Wine 159 has the highest Color Intensity in the entire dataset (13.0 vs a
Barbera mean of 7.4). Its Phenols and Proanthocyanins are also well above the
Barbera average. But the curve plot reveals something important: **it follows
the Barbera shape.** The curve runs parallel to its cultivar mates, just with
higher amplitude. This is a *magnitude* outlier --- the overall chemical levels
are elevated --- not a *shape* outlier.

**Quality controller's action**: re-test the Color Intensity measurement. A
reading of 13.0 is almost double the cultivar mean, which could indicate either
a genuinely concentrated wine (late harvest, low yield) or an analytical error.
The Phenol and Proanthocyanin values corroborate the "concentrated" hypothesis
--- they're consistently elevated, not randomly scattered. If re-testing
confirms the values, this wine is likely a natural outlier from an unusually
productive or concentrated lot, not a quality defect.

### Summary: The Outlier Action Matrix

Putting it all together, here's what the quality controller walks away with ---
not "these wines are unusual" (a spreadsheet could tell you that), but *what
kind* of unusual, *why*, and *what to do about it*:

```{r action-matrix}
actions <- data.frame(
  Wine = c(122, "70, 96", 159, "60, 74, 79, 97, 111"),
  Type = c("Cultivar mismatch", "Shape outlier (soil/mineral)",
           "Magnitude outlier (concentration)", "Peripheral Grignolino"),
  `Key Finding` = c(
    "Closer to Barolo median than Grignolino; 6 chemicals match Barolo profile",
    "Extreme Magnesium (151–162 mg/L); curves cross the population envelope",
    "Color Intensity 2× cultivar mean; all values elevated proportionally",
    "Various extreme chemicals (Mg, Ash, Alkalinity); fringes of Grignolino cluster"
  ),
  Action = c(
    "Label audit — verify cultivar assignment; check vineyard of origin",
    "Investigate soil chemistry and rootstock at source vineyard",
    "Re-test Color Intensity; if confirmed, flag as concentrated lot",
    "Review against cultivar specification; note in batch records"
  ),
  check.names = FALSE
)
kable(actions, caption = "Outlier triage: from flag to action")
```

This is the payoff of the functional approach. A classical Mahalanobis analysis
would have flagged most of these same wines --- but it would have given the
quality controller a list of numbers with no indication of *why* each wine is
unusual or *what to do about it*. The FDA pipeline delivers typed outliers
(magnitude vs shape vs cultivar mismatch), visual diagnostics (the curve
overlays above), and domain-specific action items. That's the difference between
a statistical report and a quality decision.

### Comparison: What Would Standard Statistics Give You?

A classically trained analyst might reach for **Mahalanobis distance** --- the
standard multivariate outlier detector. It measures how far each wine is from
the population mean, accounting for correlations between chemicals. Wines
exceeding a $\chi^2$ critical value are flagged.

```{r mahal-outliers}
# Classical Mahalanobis distance outlier detection
mahal <- mahalanobis(X, colMeans(X), cov(X))
cutoff_chi <- qchisq(0.975, df = ncol(X))
mahal_outliers <- which(mahal > cutoff_chi)

cat(sprintf("Mahalanobis outliers (chi-sq cutoff = %.1f, df = %d):\n",
            cutoff_chi, ncol(X)))
cat("  Flagged:", sort(mahal_outliers), "\n")
cat("  Count:", length(mahal_outliers), "\n")
```

```{r mahal-compare}
# Compare with FDA results
fda_all <- unique(c(out_pond$outliers, og$outliers, ms$outliers))
overlap <- intersect(mahal_outliers, fda_all)
fda_only <- setdiff(fda_all, mahal_outliers)
mahal_only <- setdiff(mahal_outliers, fda_all)

cat("Overlap between methods:\n")
cat("  Both flag:           ", sort(overlap), "\n")
cat("  FDA only:            ", sort(fda_only), "\n")
cat("  Mahalanobis only:    ", sort(mahal_only), "\n")

# Show Mahalanobis distances for FDA-flagged wines
compare_df <- data.frame(
  Wine = sort(union(mahal_outliers, fda_all)),
  Cultivar = as.character(cultivar[sort(union(mahal_outliers, fda_all))])
)
compare_df$Mahal_Dist <- round(mahal[compare_df$Wine], 1)
compare_df$Mahal_Flag <- compare_df$Wine %in% mahal_outliers
compare_df$FDA_Flag <- compare_df$Wine %in% fda_all
compare_df$FDA_Methods <- sapply(compare_df$Wine, function(w) {
  methods <- c()
  if (w %in% out_pond$outliers) methods <- c(methods, "Depth")
  if (w %in% og$outliers) methods <- c(methods, "Outliergram")
  if (w %in% ms$outliers) methods <- c(methods, "MS-plot")
  paste(methods, collapse = " + ")
})

kable(compare_df, caption = "Mahalanobis vs FDA: flagged wines compared")
```

The overlap is substantial --- **as the distance preservation theorem
predicts**, both approaches are working with the same underlying Euclidean
geometry. But notice the difference in what you *get*:

| | Mahalanobis Distance | FDA Outlier Methods |
|---|---|---|
| **Output** | One number per wine | Three diagnostic plots + typed flags |
| **Tells you *that*** | Yes | Yes |
| **Tells you *how*** | No --- just "far from center" | Yes --- magnitude vs shape vs rank structure |
| **Visual diagnostic** | None (abstract distance) | Overlay outlier curves on the population |
| **Assumes** | Multivariate normality | No distributional assumption |
| **Multiple aspects** | Single axis of outlyingness | Three complementary axes |

Mahalanobis distance is a fine screening tool, and in this dataset it largely
agrees with the FDA methods. But when it flags Wine 122 with a distance of 58.7,
it can't tell you *why* --- is every chemical extreme, or is the pattern unusual?
The FDA approach answers that: depth-ponderation says the wine is far from the
population center, the outliergram says the shape is unusual, and the
magnitude-shape plot says the chemical *ratios* aren't distorted. For a quality
controller, that distinction means the difference between re-running the test
(magnitude issue) and investigating the vineyard (shape issue).

## What's Next?

The other articles in this series:

- **[Why Andrews Curves?](example-andrews-wine-intro.html)**: The starting
  point --- why transform 13 chemicals into curves, and the distance
  preservation proof.

- **[Clustering & Variable Importance](example-andrews-wine-clustering.html)**:
  K-means, fuzzy c-means, FPCA, and permutation testing.

- **[Quality Control](example-andrews-wine-qc.html)**: Functional boxplots,
  tolerance bands, and a three-phase monitoring workflow.

## References

- Arribas-Gil, A. and Romo, J. (2014). Shape outlier detection and visualization for functional data: the outliergram. *Biostatistics*, 15(4), 603--619.
- Dai, W. and Genton, M.G. (2018). Multivariate functional data visualization and outlier detection. *Journal of Computational and Graphical Statistics*, 27(4), 923--934.
