---
title: "Andrews Wine: Why Andrews Curves?"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
library(dplyr)
library(knitr)
theme_set(theme_minimal(base_size = 13))
set.seed(42)
```

This is the starting point for a four-article series analyzing 178 wines
(3 cultivars, 13 chemicals) with Andrews curves and functional data analysis.

| Article | What It Does | Outcome |
|---|---|---|
| **Why Andrews Curves?** (this article) | Transform 13 chemicals into curves; verify distance preservation | Each wine becomes a visual fingerprint; $L^2$ distances equal $\sqrt{\pi} \times$ Euclidean --- nothing lost |
| [Outlier Detection](example-andrews-wine.html) | Depth, outliergram, MS-plot | 9 anomalies classified by type --- mislabel, soil anomaly, or concentration --- with corrective actions |
| [Clustering & Variable Importance](example-andrews-wine-clustering.html) | K-means, fuzzy c-means, permutation test, FPCA | Cultivar recovery at 96% accuracy; top 5 chemicals identified for cost reduction |
| [Quality Control](example-andrews-wine-qc.html) | Functional boxplots, depth rankings, tolerance bands | Monitoring system that checks new batches against a validated specification in one chart |

## The Problem with Tables

A quality-control analyst reviews 178 wines, each tested for 13 chemical
properties --- that's **2,314 numbers**. The questions are simple: *Any
anomalous wines? Do the three cultivars (Barolo, Grignolino, Barbera) look
chemically distinct? Which chemicals matter most?*

The business needs are concrete:

1. **Detect production anomalies** before a defective batch ships
2. **Validate cultivar identity** for denomination-of-origin certification
3. **Reduce testing costs** by identifying redundant assays
4. **Establish ongoing monitoring** against a validated reference profile

Standard practice uses separate tools for each (spreadsheet, MANOVA, PCA,
per-variable control charts), each with its own distance metric and assumptions.
Nothing connects them.

The alternative: **transform all 13 chemicals into Andrews curves** and apply
a unified pipeline of functional data analysis methods. Every step operates on
the same `fdata` object with the same $L^2$ distance semantics.

```{r load-wine}
# UCI Wine dataset: 178 wines, 3 cultivars, 13 chemical measurements
wine_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"
col_names <- c("Cultivar", "Alcohol", "MalicAcid", "Ash", "Alkalinity",
               "Magnesium", "Phenols", "Flavanoids", "NonflavPhenols",
               "Proanthocyanins", "ColorIntensity", "Hue",
               "OD280_OD315", "Proline")
wine <- read.csv(wine_url, header = FALSE, col.names = col_names)

# Use real cultivar names
cultivar <- factor(wine$Cultivar,
                   levels = 1:3,
                   labels = c("Barolo", "Grignolino", "Barbera"))

# Standardize the 13 chemical variables
X <- scale(as.matrix(wine[, -1]))
chem_names <- colnames(wine)[-1]

cat(nrow(wine), "wines,", ncol(X), "chemicals,",
    nlevels(cultivar), "cultivars\n")
```

```{r boxplot-before, fig.height = 6}

df_box <- data.frame(X) |>
  mutate(Cultivar = cultivar) |>
  tidyr::pivot_longer(-Cultivar, names_to = "Chemical", values_to = "Value")

ggplot(df_box, aes(x = Cultivar, y = Value, fill = Cultivar)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ Chemical, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("Barolo" = "#8B0000",
                                "Grignolino" = "#DAA520",
                                "Barbera" = "#2E8B57")) +
  labs(x = NULL, y = "Standardized Value") +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9))
```

Some variables (Flavanoids, Proline, Color Intensity) clearly separate cultivars.
Others (Ash, Magnesium) overlap heavily. But you can't *see a wine* here --- you
see 13 disconnected box-and-whisker slices. Every decision about blending,
pricing, or fraud detection requires a **holistic view** of each wine's chemical
fingerprint. That's where Andrews curves come in.

## Turning Rows into Curves

The **Andrews transformation** (Andrews, 1972) maps each $p$-dimensional
observation $\mathbf{x} = (x_1, \ldots, x_p)$ to a curve on $[-\pi, \pi]$
using a Fourier expansion:

$$f_{\mathbf{x}}(t) = \frac{x_1}{\sqrt{2}} + x_2\sin(t) + x_3\cos(t) + x_4\sin(2t) + x_5\cos(2t) + \cdots$$

The key guarantee: **$L^2$ distance between curves equals $\sqrt{\pi}$ times the
Euclidean distance between the original vectors.** Nothing is lost. Nothing is
distorted.

```{r andrews-transform}
fd_wine <- andrews_transform(X)
cat("Andrews curves:", nrow(fd_wine$data), "observations,",
    ncol(fd_wine$data), "grid points\n")
```

```{r hero-plot, fig.height = 6}

n <- nrow(fd_wine$data)
m <- ncol(fd_wine$data)
t_grid <- fd_wine$argvals

df_curves <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(fd_wine$data)),
  curve = rep(1:n, each = m),
  Cultivar = rep(cultivar, each = m)
)

ggplot(df_curves, aes(x = t, y = value, group = curve, color = Cultivar)) +
  geom_line(alpha = 0.35, linewidth = 0.4) +
  scale_color_manual(values = c("Barolo" = "#8B0000",
                                 "Grignolino" = "#DAA520",
                                 "Barbera" = "#2E8B57")) +
  labs(title = "Andrews Curves of Wine Chemical Profiles",
       x = expression(t), y = expression(f[x](t))) +
  theme(legend.position = "bottom")
```

Now each wine has a **signature** --- a single visual object that encodes all 13
measurements. A quality manager can glance at a curve and compare it to the
cultivar's typical profile. Wines that "look different" are immediately
suspicious. This is the multivariate equivalent of a chromatography trace ---
one picture captures the whole chemical identity.

## Proving the Bridge Works

Pretty pictures are nice, but for regulatory or audit contexts, we need a
**mathematical guarantee**. The Andrews distance preservation theorem says:

$$\|f_{\mathbf{x}} - f_{\mathbf{y}}\|_{L^2} = \sqrt{\pi} \cdot \|\mathbf{x} - \mathbf{y}\|_2$$

Let's verify this on all $\binom{178}{2} = 15{,}753$ pairwise distances.

```{r distance-verify}
# Compute pairwise distances in both domains
dist_andrews <- metric.lp(fd_wine)
dist_euclid <- as.matrix(dist(X))

# Extract upper triangle
idx_upper <- upper.tri(dist_andrews)
d_a <- dist_andrews[idx_upper]
d_e <- dist_euclid[idx_upper]

# Filter zero-distance pairs (if any duplicates)
nonzero <- d_e > 1e-10
ratio <- d_a[nonzero] / d_e[nonzero]

cat(sprintf("Distance ratio (Andrews / Euclidean):\n"))
cat(sprintf("  Mean:     %.4f\n", mean(ratio)))
cat(sprintf("  Median:   %.4f\n", median(ratio)))
cat(sprintf("  SD:       %.2e\n", sd(ratio)))
cat(sprintf("  sqrt(pi): %.4f\n", sqrt(pi)))
```

```{r distance-scatter}

df_dist <- data.frame(euclidean = d_e[nonzero], andrews = d_a[nonzero])

ggplot(df_dist, aes(x = euclidean, y = andrews)) +
  geom_point(alpha = 0.08, size = 0.5, color = "steelblue") +
  geom_abline(slope = sqrt(pi), intercept = 0, color = "red", linewidth = 1) +
  labs(title = "Andrews L² Distance vs Euclidean Distance",
       subtitle = sprintf("Red line: slope = √π ≈ %.4f", sqrt(pi)),
       x = "Euclidean Distance (standardized data)",
       y = "Andrews L² Distance") +
  coord_equal()
```

The ratio is $\sqrt{\pi} \approx 1.7725$ to machine precision. Any conclusion
drawn from the curves --- outliers, clusters, similarities --- translates
*exactly* back to the original 13 chemical measurements. For regulatory or audit
contexts, this mathematical proof is essential: you can report findings in
chemical units, not abstract curve distances.

## What This Means for Analysis

Because the transformation is isometric, distance-based methods give
**numerically equivalent results** whether you apply them to the curves or to
the raw 13-column matrix. FPCA on Andrews curves and `prcomp()` on the
standardized data produce scores that correlate at $\pm 1$. K-means recovers
the same clusters. If all you need is PCA + clustering on a clean matrix,
`prcomp()` and `kmeans()` are simpler and sufficient.

The value of the functional representation is in the tools that have **no
direct multivariate equivalent**:

| FDA Method | Classical Equivalent | What FDA Adds |
|---|---|---|
| `outliers.depth.pond()` | Mahalanobis distance | Same core idea, but combined with `outliergram()` and `magnitudeshape()` you can classify outliers by *type* (magnitude vs shape) --- something Mahalanobis alone cannot do |
| `cluster.kmeans()` | `kmeans()` on raw data | Same clusters, but centroids are curves you can plot and overlay --- a visual fingerprint, not 13 numbers |
| `fdata2pc()` | `prcomp()` | Same variance decomposition; eigenfunctions are the visual version of a loading table |
| `boxplot()` | 13 separate control charts | **No equivalent.** One chart monitors all 13 chemicals simultaneously via functional depth |
| `tolerance.band()` | Multivariate tolerance region | **No equivalent.** Defines a nonparametric envelope in function space; new wines are checked against a single band |
| `group.test()` | MANOVA | Nonparametric permutation test; bootstrap CIs show *where* on the curve profiles diverge |

The bottom line: Andrews curves earn their keep when you use the functional
toolbox --- depth-based outlier classification, functional boxplots, tolerance
bands, simultaneous monitoring. For methods where classical statistics already
gives the same answer (PCA, k-means), the functional version adds visualization
and pipeline integration, but not new statistical information.

## Next Steps

The remaining articles apply FDA methods to these Andrews curves:

- **[Outlier Detection](example-andrews-wine.html)**: Three complementary
  methods classify 9 anomalies by type with specific corrective actions.

- **[Clustering & Variable Importance](example-andrews-wine-clustering.html)**:
  K-means, fuzzy c-means, FPCA, and permutation testing.

- **[Quality Control](example-andrews-wine-qc.html)**: Functional boxplots,
  tolerance bands, and a three-phase monitoring workflow.

## References

- Andrews, D.F. (1972). Plots of high-dimensional data. *Biometrics*, 28(1), 125--136.
