---
title: "Berkeley Growth Study: Elastic Alignment of Growth Velocity"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

93 children's growth velocity curves (ages 1--18) are elastically aligned to
separate **when** each child's growth spurt occurs from **how tall** it is.

| Step | What It Does | Outcome |
|---|---|---|
| Elastic alignment | Remove timing differences across children | Variance reduction of ~60% in pubertal region; de-blurred Karcher mean recovers true peak shape |
| Warping functions | Extract per-child maturation timing | Scalar "timing shift" score per child, usable as a covariate in downstream regression |
| Equivalence test | Test whether boys and girls differ in timing | Equivalence rejected at $\delta = 0.5$ years --- ~2 year gap confirmed with bootstrap CIs |
| FPCA on warpings | Identify main modes of timing variation | PC1 (83%) separates boys from girls; PC2 captures secondary timing structure |

**Key result:** Girls' pubertal growth spurt peaks ~2 years earlier than boys'
(mean age 11.5 vs 13.5). The warping functions quantify this precisely for each
child and can be correlated with other variables (adult height, BMI, etc.).

This dataset is a standard benchmark in elastic FDA literature (Tucker et al.,
2013; Srivastava and Klassen, 2016). It is available in the `fdasrvf` package.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Data Preparation

```{r growth-data}
data(growth_vel, package = "fdasrvf")

# growth_vel$f is 69 (grid) x 93 (curves) — transpose for fdata
# Standard ordering: first 39 = boys, last 54 = girls
fd_growth_raw <- fdata(t(growth_vel$f), argvals = growth_vel$time)
gender <- factor(c(rep("Boy", 39), rep("Girl", 54)))
cat("Curves:", nrow(fd_growth_raw$data),
    "(", sum(gender == "Boy"), "boys,", sum(gender == "Girl"), "girls)\n")
cat("Grid points:", ncol(fd_growth_raw$data), "\n")
cat("Age range:", range(fd_growth_raw$argvals), "\n")
```

The raw data has only 69 grid points, which is coarse for elastic alignment
(the DP warping resolution is limited). We smooth with P-splines and evaluate
on a finer grid:

```{r growth-smooth}
# Smooth with P-splines and evaluate on a finer grid
fine_argvals <- seq(min(fd_growth_raw$argvals), max(fd_growth_raw$argvals),
                    length.out = 200)
coefs_growth <- fdata2basis(fd_growth_raw, nbasis = 25, type = "bspline")
fd_growth <- basis2fdata(coefs_growth, fine_argvals)
cat("Smoothed grid points:", ncol(fd_growth$data), "\n")
```

## Full Age Range

```{r growth-full, fig.height = 4}
plot(fd_growth) + labs(title = "Berkeley Growth Velocity (all 93 children)",
                       x = "Age (years)", y = "Growth velocity (cm/year)")
```

The dominant feature at age 1 (infant growth) and the pubertal spurt around
ages 11--14 are both visible, but the pubertal peak is blurred by phase
variability.

```{r growth-full-align}
res_growth <- elastic.align(fd_growth)
var_orig_growth <- mean(apply(fd_growth$data, 2, var))
vr_growth <- 1 - mean(apply(res_growth$aligned$data, 2, var)) / var_orig_growth
cat("Full range VR:", round(vr_growth * 100, 1), "%\n")
```

```{r growth-full-plot, fig.height = 4}
plot(res_growth, type = "both")
```

## Pubertal Growth Spurt (Ages 8--18)

The pubertal region is where phase variability is strongest. Restricting to
ages 8--18 isolates the growth spurt:

```{r growth-pubertal}
idx_pub <- which(fd_growth$argvals >= 8)
fd_pub <- fdata(fd_growth$data[, idx_pub],
                argvals = fd_growth$argvals[idx_pub])

res_pub <- elastic.align(fd_pub)
var_pub <- mean(apply(fd_pub$data, 2, var))
vr_pub <- 1 - mean(apply(res_pub$aligned$data, 2, var)) / var_pub
cat("Pubertal region VR:", round(vr_pub * 100, 1), "%\n")
```

```{r growth-pubertal-plot, fig.height = 4}
plot(res_pub, type = "both")
```

## Cross-Sectional Mean vs Karcher Mean

The Karcher mean serves as the **reference frame** that gives the warping
functions their meaning -- each $\gamma_i$ is defined as "warp curve $i$ to
match *this* mean." Without a consistent reference, the warpings would not be
comparable across children.

As a bonus, the Karcher mean also recovers a de-blurred population average.
The cross-sectional mean is attenuated because children's growth spurts are
averaged at mismatched ages; the Karcher mean recovers the true peak amplitude:

```{r growth-karcher}
km_pub <- karcher.mean(fd_pub, max.iter = 20)
cross_mean_pub <- as.numeric(mean(fd_pub)$data)
karcher_pub <- as.numeric(km_pub$mean$data)

cat("Cross-sectional mean peak:", round(max(cross_mean_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(cross_mean_pub)], 1), "\n")
cat("Karcher mean peak:        ", round(max(karcher_pub), 2), "cm/yr at age",
    round(fd_pub$argvals[which.max(karcher_pub)], 1), "\n")
```

```{r growth-karcher-plot, fig.height = 4}
df_means <- data.frame(
  age = rep(fd_pub$argvals, 2),
  velocity = c(cross_mean_pub, karcher_pub),
  type = rep(c("Cross-sectional mean", "Karcher mean"),
             each = length(fd_pub$argvals))
)

ggplot(df_means, aes(x = age, y = velocity, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Cross-sectional mean" = "steelblue",
                                "Karcher mean" = "firebrick")) +
  labs(title = "Pubertal Growth Spurt: Cross-Sectional vs Karcher Mean",
       x = "Age (years)", y = "Growth velocity (cm/year)", color = NULL)
```

The Karcher mean is sharper and taller because it aligns the growth spurts
before averaging, rather than smearing them across different ages. This sharper
shape is a better estimate of the typical growth velocity profile -- but its
primary role is as the anchor that makes the per-child warping functions
interpretable and mutually comparable.

## FPCA Before and After Alignment

Alignment concentrates variance into fewer principal components:

```{r growth-fpca}
pc_before <- fdata2pc(fd_pub, ncomp = 3)
pc_after <- fdata2pc(km_pub$aligned, ncomp = 3)

pve_before <- round(pc_before$d^2 / sum(pc_before$d^2) * 100, 1)
pve_after <- round(pc_after$d^2 / sum(pc_after$d^2) * 100, 1)

cat("Before alignment: PC1 =", pve_before[1], "%, PC2 =", pve_before[2],
    "%, PC3 =", pve_before[3], "%\n")
cat("After alignment:  PC1 =", pve_after[1], "%, PC2 =", pve_after[2],
    "%, PC3 =", pve_after[3], "%\n")
```

Before alignment, phase variability spreads across multiple PCs. After
alignment, the remaining (amplitude) variability is more efficiently captured
by the first PC.

## Warping Functions Reveal Growth Timing

The warping functions are the main extracted quantity from this analysis.
Each $\gamma_i$ maps population-average time to child $i$'s biological time,
revealing how their developmental clock differs from the Karcher mean reference:

```{r growth-warps, fig.height = 4}
n_pub <- nrow(res_pub$gammas$data)
m_pub <- ncol(res_pub$gammas$data)
df_warps <- data.frame(
  t = rep(res_pub$gammas$argvals, n_pub),
  gamma = as.vector(t(res_pub$gammas$data)),
  curve = rep(seq_len(n_pub), each = m_pub),
  sex = rep(gender, each = m_pub)
)

ggplot(df_warps, aes(x = t, y = gamma, group = curve, color = sex)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(alpha = 0.4) +
  scale_color_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "Warping Functions by Sex",
       x = "Reference age (years)",
       y = expression(gamma(t)), color = NULL) +
  theme(legend.position = "top")
```

Each $\gamma_i(t)$ maps from the reference time axis (the Karcher mean) to the
corresponding age in child $i$'s original curve. Curves **below** the diagonal
($\gamma_i(t) < t$) indicate that the growth spurt occurred at a younger age
than the population average -- i.e., earlier maturation. Curves **above** the
diagonal indicate later maturation.

The colour separation confirms the well-known pattern: girls (coral) cluster
below the diagonal, reflecting earlier pubertal timing, while boys (blue)
cluster above.

## Testing the Gender Difference in Timing

The visual pattern suggests boys and girls differ in pubertal timing. We can
test this formally with `fequiv.test()`, which checks whether the mean warping
functions of two groups are equivalent within a margin $\delta$. If equivalence
is *not* declared, we have evidence the timing patterns genuinely differ.

We use $\delta = 0.5$ years -- differences smaller than half a year would be
considered practically negligible:

```{r growth-gender-test}
warps_boys <- km_pub$gammas[gender == "Boy"]
warps_girls <- km_pub$gammas[gender == "Girl"]

set.seed(42)
equiv <- fequiv.test(warps_boys, warps_girls, delta = 0.5,
                     n.boot = 1000, seed = 42)
print(equiv)
```

```{r growth-gender-test-plot, fig.height = 4}
plot(equiv) +
  labs(x = "Age (years)",
       y = expression(hat(gamma)[boys](t) - hat(gamma)[girls](t)))
```

The mean difference curve (solid line) and its simultaneous confidence band
(shaded) both far exceed the $\pm\delta = 0.5$ year margins (dashed lines),
so equivalence is decisively rejected.

The mean warping functions for each group show the separation directly:

```{r growth-mean-warps, fig.height = 4}
t_pub <- warps_boys$argvals
mean_warp_boys <- as.numeric(mean(warps_boys)$data)
mean_warp_girls <- as.numeric(mean(warps_girls)$data)

# Pointwise 95% CI via bootstrap
set.seed(42)
n_boot <- 1000
boot_boys <- matrix(0, n_boot, length(t_pub))
boot_girls <- matrix(0, n_boot, length(t_pub))
for (b in seq_len(n_boot)) {
  boot_boys[b, ] <- colMeans(warps_boys$data[sample(nrow(warps_boys$data),
                              replace = TRUE), ])
  boot_girls[b, ] <- colMeans(warps_girls$data[sample(nrow(warps_girls$data),
                               replace = TRUE), ])
}
ci_boys_lo <- apply(boot_boys, 2, quantile, 0.025)
ci_boys_hi <- apply(boot_boys, 2, quantile, 0.975)
ci_girls_lo <- apply(boot_girls, 2, quantile, 0.025)
ci_girls_hi <- apply(boot_girls, 2, quantile, 0.975)

df_ci <- rbind(
  data.frame(t = t_pub, lo = ci_boys_lo, hi = ci_boys_hi, group = "Boys"),
  data.frame(t = t_pub, lo = ci_girls_lo, hi = ci_girls_hi, group = "Girls")
)
df_mean_warps <- data.frame(
  t = rep(t_pub, 3),
  gamma = c(mean_warp_boys, mean_warp_girls, t_pub),
  group = rep(c("Boys", "Girls", "Identity"), each = length(t_pub))
)

ggplot() +
  geom_ribbon(data = df_ci, aes(x = t, ymin = lo, ymax = hi, fill = group),
              alpha = 0.2) +
  geom_line(data = df_mean_warps,
            aes(x = t, y = gamma, color = group, linetype = group),
            linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(Boys = "steelblue", Girls = "coral",
                                Identity = "grey60")) +
  scale_fill_manual(values = c(Boys = "steelblue", Girls = "coral")) +
  scale_linetype_manual(values = c(Boys = "solid", Girls = "solid",
                                   Identity = "dashed")) +
  labs(title = "Mean Warping Functions by Sex (95% CI)",
       x = "Reference age (years)", y = expression(bar(gamma)(t)),
       color = NULL, linetype = NULL, fill = NULL) +
  guides(fill = "none") +
  theme(legend.position = "top")
```

Girls' mean warping sits below the identity (the growth spurt features occur
at younger ages), while boys' sits above. The vertical gap between the two
curves represents the timing difference at each reference age.

We can quantify the timing gap directly from the original curves:

```{r growth-peak-ages}
peak_ages <- fd_pub$argvals[apply(fd_pub$data, 1, which.max)]
cat("Boys: mean peak age =", round(mean(peak_ages[gender == "Boy"]), 1),
    "years (sd =", round(sd(peak_ages[gender == "Boy"]), 1), ")\n")
cat("Girls: mean peak age =", round(mean(peak_ages[gender == "Girl"]), 1),
    "years (sd =", round(sd(peak_ages[gender == "Girl"]), 1), ")\n")
cat("Difference:", round(mean(peak_ages[gender == "Boy"]) -
    mean(peak_ages[gender == "Girl"]), 1), "years\n")
```

The equivalence test on the warping functions confirms what the peak ages
show: boys and girls have meaningfully different growth timing.

## Working with Warping Functions

The warping functions $\gamma_i$ are `fdata` objects, so they can be used with
any fdars function. Here are concrete ways to extract information from them:

**Scalar summaries per child.** The deviation from the identity
$\gamma_i(t) - t$ summarises how much each child's timing differs from the
reference. Averaging this gives a single "timing shift" score per child:

```{r warp-scores}
# Timing shift: mean deviation from identity (positive = later, negative = earlier)
t_pub <- km_pub$gammas$argvals
identity <- matrix(t_pub, nrow = nrow(km_pub$gammas$data),
                   ncol = length(t_pub), byrow = TRUE)
timing_shift <- rowMeans(km_pub$gammas$data - identity)

cat("Earliest maturer: child", which.min(timing_shift),
    "(shift =", round(min(timing_shift), 2), "years)\n")
cat("Latest maturer:   child", which.max(timing_shift),
    "(shift =", round(max(timing_shift), 2), "years)\n")
```

These scalar scores can be used as covariates in standard regression, or for
correlation with other variables (e.g., adult height, BMI):

```{r warp-boxplot, fig.height = 4}
df_shift <- data.frame(shift = timing_shift, sex = gender)
ggplot(df_shift, aes(x = sex, y = shift, fill = sex)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "Timing Shift Scores by Sex",
       x = NULL, y = "Mean timing shift (years)") +
  guides(fill = "none")
```

**FPCA on warping functions.** The warpings themselves can be analysed as
functional data. Functional PCA on the warping functions identifies the main
modes of timing variation:

```{r warp-fpca}
pc_warps <- fdata2pc(km_pub$gammas, ncomp = 2)
pve_warps <- round(pc_warps$d^2 / sum(pc_warps$d^2) * 100, 1)
cat("Warp PC1:", pve_warps[1], "% — PC2:", pve_warps[2], "%\n")
```

```{r warp-fpca-plot, fig.height = 4}
df_scores <- data.frame(
  PC1 = pc_warps$x[, 1],
  PC2 = pc_warps$x[, 2],
  sex = gender
)
ggplot(df_scores, aes(x = PC1, y = PC2, color = sex)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c(Boy = "steelblue", Girl = "coral")) +
  labs(title = "PCA of Warping Functions",
       x = paste0("PC1 (", pve_warps[1], "%)"),
       y = paste0("PC2 (", pve_warps[2], "%)"),
       color = NULL)
```

The PC1 score is almost perfectly correlated with the mean timing shift
($r \approx 0.99$) and clearly separates boys from girls. PC2 captures a
secondary mode of timing variation orthogonal to the early/late shift --
its interpretation would require inspecting the eigenfunctions.

## References

- Tucker, J.D., Wu, W., and Srivastava, A. (2013). Generative models for functional data using phase and amplitude separation. *Computational Statistics & Data Analysis*, 61:50--66.
- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data Analysis*. Springer.
- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd ed. Springer.
