---
title: "Canadian Precipitation: Geographic Effects on Rainfall Profiles"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

Precipitation across Canada varies dramatically with geography. Pacific
stations receive heavy winter rain from moisture-laden air masses off the
ocean, Continental stations see summer-dominated convective rainfall, and
Arctic stations remain dry year-round. This article uses function-on-scalar
regression and functional ANOVA to quantify how latitude and longitude shape
daily precipitation profiles.

| Step | What It Does | Outcome |
|---|---|---|
| Data preparation | Load 35 stations, log-transform daily precipitation | `fdata` object with 365-day curves |
| Regional patterns | Faceted plots by region with mean overlays | Pacific wet-winter, Continental summer-peak, Arctic low |
| Smoothing | B-spline basis expansion (40 basis functions) | Noise-reduced curves for regression |
| FOSR | Regress precipitation on latitude + longitude | Coefficient functions showing when/where geography matters |
| FPC-based FOSR | Re-fit using FPC representation | Smoother coefficient estimates |
| FANOVA | Permutation test for regional differences | Significant regional separation |
| Prediction | Predict profile for a hypothetical station | Curve with geographic context |
| Model comparison | Latitude-only vs longitude-only vs both | Relative explanatory power |

**Key result:** Longitude is the dominant geographic predictor of precipitation
seasonality in Canada. Its effect peaks in winter, reflecting the Pacific
moisture corridor that dumps rain on coastal stations while leaving inland
stations dry. Latitude has a smaller but consistent effect, with higher
latitudes receiving less precipitation overall.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## 1. The Data

The `CanadianWeather` dataset from the **fda** package contains daily
temperature and precipitation averages (1960--1994) at 35 weather stations
grouped into four geographic regions. The precipitation values (mm/day)
include many near-zero days, so a `log1p` transform stabilizes the variance
and reduces right-skew while naturally handling zeros.

```{r data-load}
data(CanadianWeather, package = "fda")

# Extract precipitation: 365 days x 35 stations
precip_raw <- CanadianWeather$dailyAv[, , "Precipitation.mm"]

# log1p handles zeros gracefully
precip_log <- log1p(precip_raw)

# fdata wants rows = observations, columns = time points
fd <- fdata(t(precip_log), argvals = seq(1, 365, by = 1))

# Geographic metadata
coords <- CanadianWeather$coordinates
region <- factor(CanadianWeather$region)
stations <- CanadianWeather$place

cat("Stations:", nrow(fd$data), "\n")
cat("Grid points:", ncol(fd$data), "(daily)\n")
cat("Regions:", paste(levels(region), table(region), sep = ": ", collapse = ", "), "\n")
```

```{r plot-all-curves, fig.height = 5}
region_colors <- c("Arctic" = "#56B4E9", "Atlantic" = "#E69F00",
                    "Continental" = "#009E73", "Pacific" = "#CC79A7")

# Build plotting data frame
curve_df <- data.frame(
  Day = rep(1:365, each = nrow(fd$data)),
  Precip = as.vector(t(fd$data)),
  Station = rep(stations, 365),
  Region = rep(as.character(region), 365)
)

ggplot(curve_df, aes(x = .data$Day, y = .data$Precip,
                     group = .data$Station, color = .data$Region)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  scale_color_manual(values = region_colors) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "Daily Precipitation Profiles (35 Canadian Stations)",
       x = "Day of year", y = "log1p(precipitation, mm)")
```

Several patterns are immediately visible: Pacific stations (pink) show high
winter precipitation that drops sharply in summer, Continental stations
(green) peak in mid-summer, and Arctic stations (blue) remain near the
bottom throughout. The Atlantic stations are intermediate.

## 2. Regional Patterns

Faceting by region makes the within-group variation and between-group
contrast easier to see.

```{r regional-facets, fig.height = 6}
# Compute regional means
mean_df <- do.call(rbind, lapply(levels(region), function(r) {
  idx <- which(region == r)
  mean_curve <- colMeans(fd$data[idx, , drop = FALSE])
  data.frame(Day = 1:365, Precip = mean_curve, Region = r)
}))

ggplot(curve_df, aes(x = .data$Day, y = .data$Precip)) +
  geom_line(aes(group = .data$Station), alpha = 0.3, linewidth = 0.3,
            color = "grey50") +
  geom_line(data = mean_df, aes(x = .data$Day, y = .data$Precip),
            color = "firebrick", linewidth = 1) +
  facet_wrap(~ .data$Region) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "Precipitation by Region (red = regional mean)",
       x = "Day of year", y = "log1p(precipitation, mm)")
```

- **Pacific**: Pronounced wet-winter / dry-summer cycle, driven by Pacific
  frontal systems. Vancouver receives over 10 times more rain in December
  than in July.
- **Continental**: Summer-dominated convective precipitation. The annual range
  is moderate.
- **Atlantic**: Relatively uniform year-round precipitation with a slight
  autumn peak from Atlantic storms.
- **Arctic**: Uniformly low precipitation. Cold air holds little moisture.

## 3. Smoothing

Daily precipitation is noisy even after averaging over 30+ years. Before
fitting regression models, a B-spline basis expansion with 40 basis functions
removes high-frequency fluctuations while preserving the seasonal signal.

```{r smoothing}
set.seed(42)
coefs <- fdata2basis(fd, nbasis = 40, type = "bspline")
fd_s <- basis2fdata(coefs, argvals = seq(1, 365, by = 1), type = "bspline")
```

```{r smooth-comparison, fig.height = 4}
# Compare raw vs smoothed for Vancouver (a Pacific station)
van_idx <- which(stations == "Vancouver")

smooth_df <- data.frame(
  Day = rep(1:365, 2),
  Precip = c(fd$data[van_idx, ], fd_s$data[van_idx, ]),
  Type = rep(c("Raw", "Smoothed"), each = 365)
)

ggplot(smooth_df, aes(x = .data$Day, y = .data$Precip,
                      color = .data$Type, linewidth = .data$Type)) +
  geom_line() +
  scale_color_manual(values = c("Raw" = "grey60", "Smoothed" = "steelblue")) +
  scale_linewidth_manual(values = c("Raw" = 0.3, "Smoothed" = 0.9)) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "Vancouver: Raw vs B-spline Smoothed Precipitation",
       x = "Day of year", y = "log1p(precipitation, mm)")
```

The smoothed curve captures the strong wet-winter / dry-summer cycle without
the day-to-day noise. All subsequent analyses use the smoothed curves
(`fd_s`).

## 4. Geographic Predictors

Latitude and longitude are the two scalar predictors for function-on-scalar
regression. The station map shows how the four regions cluster geographically.

```{r station-map, fig.height = 5}
geo <- data.frame(
  latitude = coords[, "N.latitude"],
  longitude = -abs(coords[, "W.longitude"]),
  station = stations,
  region = as.character(region)
)

ggplot(geo, aes(x = .data$longitude, y = .data$latitude,
                color = .data$region)) +
  geom_point(size = 3) +
  scale_color_manual(values = region_colors) +
  labs(title = "Canadian Weather Stations by Region",
       x = "Longitude (°E)", y = "Latitude (°N)", color = "Region") +
  coord_fixed(ratio = 1.3)
```

Pacific stations cluster along the west coast (longitude near -125),
Atlantic stations hug the east coast (longitude near -60), and Continental
and Arctic stations span the interior and north.

## 5. FOSR: Latitude and Longitude Effects

Function-on-scalar regression models each station's precipitation curve as a
linear function of its latitude and longitude:

$$Y_i(t) = \mu(t) + \beta_{\text{lat}}(t)\,\text{latitude}_i + \beta_{\text{lon}}(t)\,\text{longitude}_i + \varepsilon_i(t)$$

The coefficient functions $\beta_{\text{lat}}(t)$ and $\beta_{\text{lon}}(t)$
reveal when during the year each geographic variable matters most.

```{r fosr-fit}
predictors <- cbind(geo$latitude, geo$longitude)
fit <- fosr(fd_s, predictors, lambda = 1)
print(fit)
```

```{r fosr-plot, fig.height = 5}
# Extract coefficient functions
beta_data <- fit$beta$data
argvals <- fit$fdataobj$argvals

beta_df <- data.frame(
  Day = rep(argvals, 2),
  Beta = c(beta_data[1, ], beta_data[2, ]),
  Predictor = rep(c("Latitude", "Longitude"), each = length(argvals))
)

ggplot(beta_df, aes(x = .data$Day, y = .data$Beta, color = .data$Predictor)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  scale_color_manual(values = c("Latitude" = "#D55E00", "Longitude" = "#0072B2")) +
  labs(title = "FOSR Coefficient Functions: Precipitation ~ Latitude + Longitude",
       x = "Day of year", y = "beta(t)", color = "Predictor")
```

### Interpretation

- **Latitude effect** $\hat{\beta}_{\text{lat}}(t)$: Generally negative,
  meaning higher-latitude stations receive less precipitation. The effect is
  relatively stable across the year, with a modest winter amplification.
- **Longitude effect** $\hat{\beta}_{\text{lon}}(t)$: This is the more
  dynamic coefficient. In winter months (November--March), the longitude
  effect is strongly positive — more westerly stations (larger negative
  longitude values, mapped to more negative inputs) receive substantially
  more rain from Pacific moisture. In summer, the longitude effect diminishes
  or reverses, as continental convection dominates regardless of east--west
  position.

### Pointwise R-squared

```{r fosr-rsq, fig.height = 4}
rsq_df <- data.frame(
  Day = argvals,
  R2 = fit$r.squared.t
)

ggplot(rsq_df, aes(x = .data$Day, y = .data$R2)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  ylim(0, 1) +
  labs(title = "Pointwise R-squared: Precipitation ~ Latitude + Longitude",
       x = "Day of year", y = expression(R^2 * "(t)"))
```

The explanatory power of latitude and longitude varies across the year.
Winter months typically show higher R-squared because the Pacific moisture
gradient creates a strong east--west contrast that is well captured by
longitude.

## 6. FOSR with FPC Basis

The FPC-based variant projects the response onto its leading principal
components before fitting the regression, which can produce smoother
coefficient estimates and reduce overfitting when the sample size is small.

```{r fosr-fpc}
fit_fpc <- fosr.fpc(fd_s, predictors, ncomp = 5)
cat("Penalized FOSR R-squared:", round(fit$r.squared, 4), "\n")
cat("FPC-based FOSR R-squared:", round(fit_fpc$r.squared, 4), "\n")
```

```{r fosr-comparison, fig.height = 5}
# Compare beta(t) from both methods
beta_fpc_data <- fit_fpc$beta$data

comp_df <- data.frame(
  Day = rep(argvals, 4),
  Beta = c(beta_data[1, ], beta_fpc_data[1, ],
           beta_data[2, ], beta_fpc_data[2, ]),
  Predictor = rep(c("Latitude", "Latitude", "Longitude", "Longitude"),
                  each = length(argvals)),
  Method = rep(c("Penalized", "FPC-based", "Penalized", "FPC-based"),
               each = length(argvals))
)

ggplot(comp_df, aes(x = .data$Day, y = .data$Beta,
                    color = .data$Method, linetype = .data$Method)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~ .data$Predictor, scales = "free_y") +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  scale_color_manual(values = c("Penalized" = "#D55E00", "FPC-based" = "#0072B2")) +
  labs(title = "Coefficient Functions: Penalized vs FPC-based FOSR",
       x = "Day of year", y = "beta(t)")
```

The FPC-based approach generally produces smoother coefficient functions
because the response variation is compressed into a small number of principal
components. The broad seasonal patterns agree between methods, but the
FPC-based estimates avoid some of the high-frequency wiggles.

## 7. Regional FANOVA

Functional ANOVA tests whether the four regions have significantly different
mean precipitation profiles, using a permutation-based global F-test.

```{r fanova}
set.seed(123)
fan <- fanova(fd_s, region, n.perm = 1000)
print(fan)
```

```{r fanova-plot, fig.height = 5}
# Extract group means and build plot data
gm_data <- fan$group.means$data
ng <- fan$n.groups

gm_df <- data.frame(
  Day = rep(argvals, ng),
  Precip = as.vector(t(gm_data)),
  Region = factor(rep(levels(region), each = length(argvals)),
                  levels = levels(region))
)

ggplot(gm_df, aes(x = .data$Day, y = .data$Precip, color = .data$Region)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = region_colors) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = paste0("FANOVA: Regional Mean Precipitation (p = ",
                      format.pval(fan$p.value), ")"),
       x = "Day of year", y = "log1p(precipitation, mm)",
       color = "Region")
```

The permutation test confirms that the four regions have significantly
different precipitation curves. The group means highlight the distinct
seasonal patterns: Pacific's winter peak, Continental's summer peak,
Atlantic's relative uniformity, and Arctic's consistently low values.

## 8. Prediction

With the fitted FOSR model, we can predict the precipitation profile for a
hypothetical station at any latitude/longitude combination.

```{r predict, fig.height = 5}
# Hypothetical station: 55 N, 100 W (northern Saskatchewan)
new_station <- matrix(c(55, -100), nrow = 1)
pred <- predict(fit, new.predictors = new_station)

# Find actual stations near latitude 55 for context
near_idx <- which(abs(geo$latitude - 55) < 5)
near_names <- stations[near_idx]

pred_df <- data.frame(
  Day = argvals,
  Precip = as.numeric(pred$data[1, ]),
  Type = "Predicted (55N, 100W)"
)

# Overlay actual curves from nearby-latitude stations
actual_df <- do.call(rbind, lapply(near_idx, function(i) {
  data.frame(
    Day = argvals,
    Precip = fd_s$data[i, ],
    Type = paste0("Actual: ", stations[i])
  )
}))

overlay_df <- rbind(pred_df, actual_df)

ggplot(overlay_df, aes(x = .data$Day, y = .data$Precip, color = .data$Type)) +
  geom_line(aes(linewidth = ifelse(.data$Type == "Predicted (55N, 100W)",
                                   "pred", "actual"))) +
  scale_linewidth_manual(values = c("pred" = 1.2, "actual" = 0.5), guide = "none") +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "Predicted vs Actual Precipitation Profiles Near 55°N",
       x = "Day of year", y = "log1p(precipitation, mm)",
       color = NULL) +
  theme(legend.position = "bottom")
```

The predicted curve for a northern Saskatchewan station (55N, 100W) falls
within the range of actual stations at similar latitudes. The model captures
the Continental summer-peak pattern driven by the station's inland location.

## 9. Model Comparison

How much explanatory power do latitude and longitude provide individually
versus together?

```{r model-comparison}
# Latitude only
fit_lat <- fosr(fd_s, as.matrix(geo$latitude), lambda = 1)

# Longitude only
fit_lon <- fosr(fd_s, as.matrix(geo$longitude), lambda = 1)

# Both (already fitted)
comparison <- data.frame(
  Model = c("Latitude only", "Longitude only", "Latitude + Longitude"),
  R_squared = round(c(fit_lat$r.squared, fit_lon$r.squared, fit$r.squared), 4),
  GCV = round(c(fit_lat$gcv, fit_lon$gcv, fit$gcv), 6)
)

knitr::kable(comparison, col.names = c("Model", "R-squared", "GCV"),
             caption = "FOSR model comparison")
```

```{r model-comparison-plot, fig.height = 5}
rsq_comp_df <- data.frame(
  Day = rep(argvals, 3),
  R2 = c(fit_lat$r.squared.t, fit_lon$r.squared.t, fit$r.squared.t),
  Model = rep(c("Latitude only", "Longitude only", "Both"),
              each = length(argvals))
)

ggplot(rsq_comp_df, aes(x = .data$Day, y = .data$R2, color = .data$Model)) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  ylim(0, 1) +
  scale_color_manual(values = c("Latitude only" = "#D55E00",
                                "Longitude only" = "#0072B2",
                                "Both" = "#009E73")) +
  labs(title = "Pointwise R-squared by Model",
       x = "Day of year", y = expression(R^2 * "(t)"),
       color = "Model")
```

At first glance, the joint model appears to perform **worse** than either
single-predictor model. However, this is a **regularization artifact**:
the same $\lambda = 1$ is applied to all three models, but the joint model
has two coefficient functions to estimate, making it more sensitive to the
penalty strength.

### Lambda Sensitivity

We can verify this by refitting the models across a range of $\lambda$
values:

```{r lambda-sensitivity}
lambdas <- c(0.01, 0.1, 1, 10)
sens_df <- do.call(rbind, lapply(lambdas, function(lam) {
  fit_l <- fosr(fd_s, as.matrix(geo$latitude), lambda = lam)
  fit_o <- fosr(fd_s, as.matrix(geo$longitude), lambda = lam)
  fit_b <- fosr(fd_s, cbind(geo$latitude, geo$longitude), lambda = lam)
  data.frame(
    Lambda = lam,
    Model = c("Latitude only", "Longitude only", "Latitude + Longitude"),
    R2 = c(fit_l$r.squared, fit_o$r.squared, fit_b$r.squared)
  )
}))

knitr::kable(sens_df, col.names = c("Lambda", "Model", "R-squared"),
             digits = 4, caption = "R-squared across lambda values")
```

```{r lambda-sensitivity-plot, fig.height = 4}
ggplot(sens_df, aes(x = factor(.data$Lambda), y = .data$R2,
                    fill = .data$Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Latitude only" = "#D55E00",
                                "Longitude only" = "#0072B2",
                                "Latitude + Longitude" = "#009E73")) +
  labs(title = "FOSR R-squared is Sensitive to Lambda",
       subtitle = "At low lambda, the joint model wins; at high lambda, it loses",
       x = expression(lambda), y = expression(R^2),
       fill = "Model") +
  theme(legend.position = "bottom")
```

With $\lambda = 0.01$, the joint model outperforms both single-predictor
models (as expected when predictors carry complementary information).
With $\lambda = 1$, the heavier penalty shrinks the two coefficient
functions more aggressively, degrading the fit. The single-predictor
models are less affected because they have only one coefficient function
to penalize.

### Station Geography

The moderate correlation between latitude and longitude (r = -0.53)
contributes to the sensitivity. Canadian stations cluster along a
northwest-to-southeast pattern:

```{r station-scatter, fig.height = 5}
r_lat_lon <- cor(geo$latitude, geo$longitude)

station_df <- data.frame(
  Latitude = geo$latitude,
  Longitude = geo$longitude,
  Region = region
)

ggplot(station_df,
       aes(x = .data$Longitude, y = .data$Latitude,
           color = .data$Region)) +
  geom_point(size = 3) +
  scale_color_manual(values = region_colors) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40",
              linetype = "dashed", linewidth = 0.5,
              inherit.aes = FALSE,
              aes(x = .data$Longitude, y = .data$Latitude)) +
  labs(title = sprintf("Station Locations (r = %.2f)", r_lat_lon),
       subtitle = "NW-SE clustering increases penalty sensitivity in joint model",
       x = "Longitude", y = "Latitude",
       color = "Region")
```

The practical takeaway: when comparing FOSR models with different numbers
of predictors, either (a) select $\lambda$ separately for each model via
cross-validation, or (b) compare models at the same, well-chosen $\lambda$.

## 10. Conclusion

This analysis demonstrates how function-on-scalar regression reveals the
geographic determinants of precipitation seasonality:

- **Latitude dominates** for precipitation in this dataset, with
  higher-latitude stations tending to be drier overall.
- **Longitude plays a secondary role**, reflecting the west-to-east
  moisture gradient from the Pacific.
- **Lambda sensitivity**: the joint model's apparent underperformance is a
  regularization artifact. At low $\lambda$, latitude + longitude together
  outperform either alone. Always select $\lambda$ per model specification.
- **Functional ANOVA** confirms highly significant regional differences in
  precipitation profiles, with Pacific stations having the most distinctive
  pattern.
- **Smoothing** with B-splines removes daily noise while preserving seasonal
  structure, and the FPC-based FOSR variant provides an alternative approach
  that naturally regularizes the coefficient estimates.

With only 35 stations and 2 predictors, the models are necessarily limited,
but they capture the dominant geographic signal. Adding elevation or distance
from the coast could further improve the fit.

## See Also

- [Function-on-Scalar Regression](function-on-scalar.html) — FOSR and
  FANOVA methodology
- [Scalar-on-Function Regression](scalar-on-function.html) — the reverse
  problem: predicting a scalar from a curve
- [Canadian Weather: Regional Climate Patterns](example-canadian-weather.html)
  — temperature-focused analysis of the same dataset

## References

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd
  ed. Springer. Chapter 13 (Canadian Weather case study).
- Ramsay, J.O. and Silverman, B.W. (2002). *Applied Functional Data
  Analysis*. Springer.
