---
title: "Canadian Temperature: Annual Cycle Detection and Decomposition"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

A climatologist studying long-term temperature trends at Edmonton, Canada wants
to automatically detect, decompose, and characterize the annual temperature
cycle from a multi-year record. Starting from the observed 1960–1994 average
annual cycle, we simulate 8 years of daily temperatures with a gradual warming
trend, increasing seasonal amplitude, and realistic year-to-year variability.
This provides an ideal test bed for the full suite of seasonal analysis tools.

| Step | What It Does | Outcome |
|---|---|---|
| Data preparation | Build 8-year series from Edmonton's annual cycle + trend + noise | `fdata` object with 2,920 daily values |
| Period detection | `detect.period()`, `sazed()`, `autoperiod()`, `cfd.autoperiod()` | All methods find period ≈ 365 days |
| Spectral analysis | `lomb.scargle()`, `detect.periods()` | Multiple periodicities: annual cycle + harmonics with periods, confidence, and strength |
| Matrix Profile | `matrix.profile()` | Shape-based motif confirms annual pattern |
| Decomposition | `stl.fd()`, `ssa.fd()`, `decompose()` | Trend + seasonal + remainder compared |
| Seasonal strength | `seasonal.strength()` (3 methods), `classify.seasonality()` | StableSeasonal classification |
| Peak timing | `detect.peaks()` across all 35 stations | Summer peak day vs latitude and region |
| Change detection | `detect.seasonality.changes.auto()`, `detect_amplitude_modulation()` | Increasing amplitude detected |

**Key result:** All four period detection methods converge on the 365-day
annual cycle. Decomposition recovers the warming trend. Amplitude modulation
is detected as the annual temperature swing gradually increases over the
8-year period, simulating the effect of increasing climate variability.

The Canadian Weather dataset is a classic benchmark in functional data analysis
(Ramsay and Silverman, 2005).

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Data Preparation

The CanadianWeather dataset contains daily temperature averaged over 1960–1994
at 35 stations. We use Edmonton (53.6°N) — a continental station with a strong
annual cycle — as the base pattern, then simulate 8 years of daily data with
a warming trend and gradually increasing seasonal amplitude.

```{r data-load}
data(CanadianWeather, package = "fda")

# 35 stations × 365 daily temperatures
temp_mat <- t(CanadianWeather$dailyAv[, , "Temperature.C"])
coords <- CanadianWeather$coordinates

# Edmonton's average annual temperature cycle (53.3°N, continental)
edmonton <- as.numeric(temp_mat["Edmonton", ])
cat("Edmonton annual cycle: min", round(min(edmonton), 1),
    "°C, max", round(max(edmonton), 1), "°C\n")

# Simulate 8 years: warming trend + increasing amplitude + weather noise
set.seed(42)
n_years <- 8
temp_long <- numeric(0)

for (yr in seq_len(n_years)) {
  trend <- 0.3 * (yr - 1)               # +0.3 °C per year warming
  amp_factor <- 1 + 0.03 * (yr - 1)     # amplitude grows ~3%/year
  noise <- rnorm(365, sd = 1.5)          # year-to-year weather variability
  yearly <- edmonton * amp_factor + trend + noise
  temp_long <- c(temp_long, yearly)
}

days <- as.numeric(seq_len(length(temp_long)))
fd_temp <- fdata(matrix(temp_long, nrow = 1), argvals = days)

cat("Series length:", length(temp_long), "days (", n_years, "years)\n")
cat("Temperature range:", round(range(temp_long), 1), "°C\n")
```

```{r raw-plot, fig.height = 4}
df_raw <- data.frame(Day = days, Temp = temp_long)
boundaries <- seq(0, n_years * 365, by = 365)

ggplot(df_raw, aes(x = Day, y = Temp)) +
  geom_vline(xintercept = boundaries[-c(1, length(boundaries))],
             linetype = "dotted", alpha = 0.3) +
  geom_line(color = "steelblue", linewidth = 0.3) +
  labs(title = "Edmonton: 8-Year Simulated Daily Temperature",
       x = "Day", y = "Temperature (°C)") +
  scale_x_continuous(
    breaks = boundaries[-length(boundaries)] + 182,
    labels = paste0("Yr ", 1:n_years)
  )
```

Each year shows the characteristic continental annual cycle: deep winter
troughs (~−20°C) and summer peaks (~+20°C). The warming trend gradually shifts
the curve upward, and the increasing amplitude factor makes winters slightly
colder and summers slightly warmer in later years.

```{r zoom-plot, fig.height = 4}
# Zoom: first 2 years
ggplot(df_raw[1:730, ], aes(x = Day, y = Temp)) +
  geom_vline(xintercept = 365, linetype = "dashed", alpha = 0.3) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  labs(title = "First Two Years (zoom)",
       x = "Day", y = "Temperature (°C)") +
  annotate("text", x = 182, y = max(temp_long[1:365]) + 3,
           label = "Year 1", size = 3, color = "grey40") +
  annotate("text", x = 547, y = max(temp_long[366:730]) + 3,
           label = "Year 2", size = 3, color = "grey40")
```

The annual cycle is unmistakable: winter troughs near the start/end of each
year and summer peaks around day 180–200 (July). Year-to-year weather noise
adds realistic variability without obscuring the dominant pattern.

## Period Detection Showdown

Four algorithms with fundamentally different approaches all attempt to
identify the dominant period. SAZED combines 5 sub-methods by voting,
Autoperiod validates FFT candidates with ACF, CFD-Autoperiod detrends
and clusters spectral peaks, and the FFT method uses raw periodogram peaks.

```{r period-detection}
# Method 1: SAZED (ensemble of 5 sub-methods)
res_sazed <- sazed(fd_temp)

# Method 2: Autoperiod (FFT + ACF validation)
res_auto <- autoperiod(fd_temp)

# Method 3: CFD-Autoperiod (clustering-based)
res_cfd <- cfd.autoperiod(fd_temp)

# Method 4: detect.period wrapper (FFT)
res_fft <- detect.period(fd_temp, method = "fft")

period_results <- data.frame(
  Method = c("SAZED", "Autoperiod", "CFD-Autoperiod", "detect.period (FFT)"),
  Period = round(c(res_sazed$period, res_auto$period,
                    res_cfd$period, res_fft$period), 1),
  Confidence = round(c(res_sazed$confidence, res_auto$confidence,
                        res_cfd$confidence, res_fft$confidence), 3)
)
knitr::kable(period_results,
             caption = "Period detection: convergence on the 365-day annual cycle")
```

All four methods converge on the 365-day annual cycle. CFD-Autoperiod achieves
the highest confidence because its detrending step removes the +0.3°C/year
warming trend before spectral analysis, letting the annual peak dominate
the periodogram. Cross-method agreement provides high confidence in the result.

## Spectral Analysis

The Lomb-Scargle periodogram reveals the full frequency content: the
fundamental annual cycle and its harmonics.

```{r lomb-scargle, fig.height = 4}
ls_result <- lomb.scargle(fd_temp, oversampling = 4)
plot(ls_result)
```

```{r detect-periods}
# Detect multiple concurrent periodicities
multi <- detect.periods(fd_temp, max_periods = 5, min_confidence = 0.2)

cat(sprintf("Number of periods detected: %d\n\n", multi$n_periods))

if (multi$n_periods > 0) {
  period_table <- data.frame(
    Period = round(multi$periods, 1),
    Confidence = round(multi$confidence, 3),
    Strength = round(multi$strength, 3),
    Amplitude = round(multi$amplitude, 2),
    Harmonic = ifelse(
      abs(multi$periods - 365) < 30, "fundamental",
      ifelse(abs(multi$periods - 365/2) < 20, "2nd harmonic",
        ifelse(abs(multi$periods - 365/3) < 15, "3rd harmonic",
          ifelse(abs(multi$periods - 365/4) < 10, "4th harmonic", ""))))
  )
  knitr::kable(period_table,
               caption = "Detected periodicities (fundamental + harmonics)")
}
```

`detect.periods()` searches for up to `max_periods` concurrent periodicities.
The dominant period at ≈ 365 days is the fundamental annual cycle. Additional
harmonics (half-year, quarter-year) arise from the **asymmetric shape** of the
temperature waveform: the winter trough is broader and flatter than the sharp
summer peak. Any non-sinusoidal periodic signal will produce harmonics — their
presence confirms that the annual cycle is real and not a sinusoid.

## Matrix Profile

The matrix profile finds repeating motifs using shape similarity rather than
spectral decomposition — robust to non-sinusoidal waveforms.

```{r matrix-profile, fig.height = 5}
mp <- matrix.profile(fd_temp, subsequence_length = 365)
plot(mp, type = "both")
```

```{r mp-summary}
cat("Primary period:", mp$primary_period, "days\n")
cat("Top detected periods:", mp$detected_periods, "days\n")
cat("Confidence:", round(mp$confidence, 3), "\n")
```

**How period extraction works:** For every position *i* in the series, the
matrix profile finds its nearest neighbor *j* (the most similar 365-day
subsequence). The *arc count* histogram (bottom) records the index distance
|*i* − *j*| across all such pairs. If the signal has a true period *P*,
most nearest neighbors will be exactly *P* steps apart, producing a dominant
spike. The primary period is the index distance with the highest arc count —
here, 365 days. The additional spikes at 730, 1095, … are integer multiples
(year 1 matching year 3, etc.).

The *matrix profile* itself (top) shows the Euclidean distance to the nearest
neighbor at each position. Low values mark positions where the annual pattern
repeats most closely; high values indicate unique segments. The downward trend
in the profile reflects that later years are closer to each other due to the
amplitude growth making them more similar.

The moderate confidence (`r round(mp$confidence, 2)`) is expected:
year-to-year noise and the amplitude trend mean no two annual cycles are
identical, so the profile distance never reaches zero.

## Decomposition: Three Methods

Extract trend, seasonal component, and remainder using three complementary
approaches.

```{r detrend}
# First, examine the overall trend
fd_detrend <- detrend(fd_temp, method = "loess", return_trend = TRUE)
cat("Detrending method:", fd_detrend$method, "\n")
```

### STL (Seasonal-Trend Decomposition using Loess)

```{r stl-decomp, fig.height = 7}
stl_result <- stl.fd(fd_temp, period = 365, s.window = 731, robust = FALSE)
plot(stl_result)
```

### Singular Spectrum Analysis (SSA)

SSA decomposes a series into additive components by embedding it in a
trajectory matrix and performing SVD. The window length should be at least
one full period.

```{r ssa-decomp, fig.height = 7}
ssa_result <- ssa.fd(fd_temp, window.length = 730)
plot(ssa_result)
```

```{r ssa-spectrum, fig.height = 4}
plot(ssa_result, type = "spectrum")
```

**Interpreting the SSA output:** The first two singular values are nearly
equal (`r round(ssa_result$contributions[1], 3)` and
`r round(ssa_result$contributions[2], 3)`) and together explain
`r round(100 * sum(ssa_result$contributions[1:2]), 1)`% of the variance.
This paired structure is the signature of a periodic component — the annual
cycle. SSA correctly assigns this oscillation to the "seasonal" slot, while
the slowly varying warming trend goes to the "trend" slot.

```{r ssa-detail}
cat("Top 5 component contributions:",
    round(ssa_result$contributions[1:5], 4), "\n")
cat("Components 1-2 (annual cycle):",
    round(100 * sum(ssa_result$contributions[1:2]), 1), "% variance\n")
```

### Classical Decomposition

```{r classical-decomp, fig.height = 6}
dec_result <- decompose(fd_temp, period = 365, method = "additive")

dec_df <- data.frame(
  Day = rep(days, 3),
  Value = c(dec_result$trend$data[1, ],
            dec_result$seasonal$data[1, ],
            dec_result$remainder$data[1, ]),
  Component = factor(rep(c("Trend", "Seasonal", "Remainder"), each = length(days)),
                     levels = c("Trend", "Seasonal", "Remainder"))
)

ggplot(dec_df, aes(x = Day, y = Value)) +
  geom_line(color = "steelblue", linewidth = 0.3) +
  facet_wrap(~Component, ncol = 1, scales = "free_y") +
  labs(title = "Classical Additive Decomposition (period = 365 days)",
       x = "Day", y = "Temperature (°C)")
```

### Comparing Seasonal Components

```{r compare-seasonal, fig.height = 4}
# Overlay the first 365 days of each method's seasonal component
n_show <- 365
seasonal_compare <- data.frame(
  Day = rep(1:n_show, 3),
  Value = c(stl_result$seasonal$data[1, 1:n_show],
            ssa_result$seasonal$data[1, 1:n_show],
            dec_result$seasonal$data[1, 1:n_show]),
  Method = rep(c("STL", "SSA", "Classical"), each = n_show)
)

ggplot(seasonal_compare, aes(x = Day, y = Value, color = Method)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Seasonal Component Comparison (first annual cycle)",
       x = "Day of year", y = "Temperature (°C)", color = "") +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct"))
```

All three methods extract a consistent annual seasonal pattern. The trend
component captures the gradual warming over the 8-year period — the simulated
+0.3°C/year trend is recovered by the decomposition.

**Why does STL's seasonal component show noise?** Unlike classical decomposition
(which forces the seasonal to be perfectly periodic), STL allows the seasonal
component to evolve slowly over time — that is its key strength. With only 8
annual cycles, the LOESS smoothing in each cycle subseries (e.g., all Jan 1st
values across 8 years) has limited noise-averaging power: the standard error
at each position is sd/sqrt(8) ≈ 0.5°C. This residual noise, combined with the
genuine 3%/year amplitude change, appears as small cycle-to-cycle differences
in the STL seasonal. Setting `robust = FALSE` (as above) avoids robustness
iterations that can amplify this effect when there are no true outliers.

## Seasonal Strength & Classification

How strong is the annual seasonality in this 8-year record?

```{r seasonal-strength}
# Three methods for quantifying seasonal strength
ss_var <- seasonal.strength(fd_temp, period = 365, method = "variance")
ss_spec <- seasonal.strength(fd_temp, period = 365, method = "spectral")
ss_wav <- seasonal.strength(fd_temp, period = 365, method = "wavelet")

strength_df <- data.frame(
  Method = c("Variance", "Spectral", "Wavelet"),
  Strength = round(c(ss_var, ss_spec, ss_wav), 3)
)
knitr::kable(strength_df, caption = "Seasonal strength by method (0 = none, 1 = perfect)")
```

```{r strength-curve, fig.height = 4}
# Time-varying seasonal strength
sc <- seasonal.strength.curve(fd_temp, period = 365)
sc_df <- data.frame(
  Day = as.numeric(sc$argvals),
  Strength = as.numeric(sc$data[1, ])
)

ggplot(sc_df, aes(x = Day, y = Strength)) +
  geom_line(color = "#D55E00", linewidth = 0.5) +
  geom_vline(xintercept = seq(365, n_years * 365, by = 365),
             linetype = "dotted", alpha = 0.3) +
  labs(title = "Time-Varying Seasonal Strength",
       subtitle = "Dotted lines = year boundaries",
       x = "Day", y = "Strength")
```

```{r classify}
# Classify the seasonality type
class_result <- classify.seasonality(fd_temp, period = 365)
print(class_result)
```

The annual temperature cycle is classified as **StableSeasonal**: the period
is constant (365 days) and the seasonal pattern dominates the signal. The
time-varying strength curve shows consistent seasonality throughout the
record, with the increasing amplitude reflected in slightly higher strength
in later years.

## Peak Timing Analysis Across Stations

When does the summer temperature peak occur at each station, and does it
shift with latitude or region? We apply `detect.peaks()` to all 35 individual
station curves and extract the summer maximum.

```{r peak-example, fig.height = 5}
# detect.peaks() smooths by default (smooth_first = TRUE)
peaks_full <- detect.peaks(fd_temp, min_distance = 300)
plot(peaks_full)
```

Fourier smoothing (enabled by default) removes daily weather noise, letting
`detect.peaks()` reliably find one summer maximum per year. Peak temperatures
increase across years due to the warming trend and growing amplitude.

```{r peak-detect}
# Detect peaks for each station's 365-day annual curve
region <- factor(CanadianWeather$region)
n_stations <- nrow(temp_mat)

peak_info <- data.frame(
  Station = character(n_stations),
  Latitude = numeric(n_stations),
  Region = character(n_stations),
  PeakDay = integer(n_stations),
  PeakTemp = numeric(n_stations),
  stringsAsFactors = FALSE
)

for (i in seq_len(n_stations)) {
  fd_i <- fdata(matrix(as.numeric(temp_mat[i, ]), nrow = 1),
                argvals = as.numeric(1:365))
  # These are already smooth (1960-1994 averages), no need for Fourier smoothing
  pk <- detect.peaks(fd_i, min_distance = 200, smooth_first = FALSE)
  p <- pk$peaks[[1]]
  # Keep the warmest peak (summer maximum)
  best <- which.max(p$value)
  peak_info$Station[i] <- rownames(temp_mat)[i]
  peak_info$Latitude[i] <- coords[i, "N.latitude"]
  peak_info$Region[i] <- as.character(region[i])
  peak_info$PeakDay[i] <- p$time[best]
  peak_info$PeakTemp[i] <- round(p$value[best], 1)
}

cat("Peak day range:", range(peak_info$PeakDay), "(day of year)\n")
cat("Peak temp range:", range(peak_info$PeakTemp), "°C\n")
```

```{r peak-lat-plot, fig.height = 6, fig.width = 8}
ggplot(peak_info, aes(x = Latitude, y = PeakDay, color = Region)) +
  geom_point(size = 2.5, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Station), size = 2.3, max.overlaps = 20,
                            segment.alpha = 0.3, show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40",
              linetype = "dashed", linewidth = 0.5) +
  labs(title = "Summer Peak Timing vs Latitude (35 Canadian Stations)",
       subtitle = "Each dot = one station (1960\u20131994 average annual cycle)",
       x = "Latitude (\u00b0N)", y = "Peak day of year")
```

```{r peak-temp-plot, fig.height = 6, fig.width = 8}
ggplot(peak_info, aes(x = Latitude, y = PeakTemp, color = Region)) +
  geom_point(size = 2.5, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Station), size = 2.3, max.overlaps = 20,
                            segment.alpha = 0.3, show.legend = FALSE) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 0.5,
              color = "grey40", linetype = "dashed") +
  labs(title = "Summer Peak Temperature vs Latitude",
       subtitle = "Each dot = one station (1960\u20131994 average annual cycle)",
       x = "Latitude (\u00b0N)", y = "Peak temperature (\u00b0C)")
```

```{r peak-table}
# Summarize by region
peak_summary <- aggregate(
  cbind(PeakDay, PeakTemp) ~ Region,
  data = peak_info,
  FUN = function(x) c(mean = round(mean(x), 1), sd = round(sd(x), 1))
)
cat("Peak timing by region:\n")
for (r in levels(region)) {
  sub <- peak_info[peak_info$Region == r, ]
  cat(sprintf("  %-12s: day %5.1f ± %4.1f  |  %.1f ± %.1f °C  (n=%d)\n",
              r, mean(sub$PeakDay), sd(sub$PeakDay),
              mean(sub$PeakTemp), sd(sub$PeakTemp), nrow(sub)))
}
```

Summer peaks cluster around day 200–210 (late July) across all stations.
Pacific stations tend to peak slightly later (oceanic thermal lag), while
Continental and Arctic stations peak earlier. Peak temperature decreases
strongly with latitude — about 0.5°C per degree northward.

## Peak Timing Variability: Is It Shifting?

The cross-station analysis above shows *spatial* variation in peak timing.
But does peak timing change *over time* within a station? With warming
temperatures, does summer arrive earlier — or later?

`analyze.peak.timing()` detects one peak per annual cycle and fits a linear
trend to the peak day-of-year. A positive trend means peaks are shifting later;
negative means earlier.

### Edmonton (8-year simulated series)

```{r peak-timing-edmonton}
pt_edm <- analyze.peak.timing(fd_temp, period = 365)
print(pt_edm)
```

```{r peak-timing-edmonton-plot, fig.height = 4}
plot(pt_edm, period = 365) +
  labs(x = "Year", y = "Peak day of year")
```

```{r peak-timing-edmonton-detail}
cat("Peak days:", round(pt_edm$peak_times %% 365, 1), "\n")
cat("Peak temps:", round(pt_edm$peak_values, 1), "°C\n")
cat("Std dev of timing:", round(pt_edm$std_timing * 365, 1), "days\n")
```

### Multi-Station Comparison

To see whether peak timing variability differs by region, we simulate 8 years
for one representative station per region and run `analyze.peak.timing()` on
each.

```{r peak-timing-multi}
# Representative stations: one per region
representatives <- c("Halifax", "Ottawa", "Edmonton", "Vancouver", "Resolute")
rep_regions <- c("Atlantic", "Continental", "Continental", "Pacific", "Arctic")

set.seed(42)
timing_comparison <- data.frame(
  Station = character(), Region = character(),
  Year = integer(), PeakDay = numeric(), PeakTemp = numeric(),
  stringsAsFactors = FALSE
)

for (j in seq_along(representatives)) {
  base_curve <- as.numeric(temp_mat[representatives[j], ])
  temp_sim <- numeric(0)
  for (yr in seq_len(n_years)) {
    trend <- 0.3 * (yr - 1)
    amp_factor <- 1 + 0.03 * (yr - 1)
    noise <- rnorm(365, sd = 1.5)
    temp_sim <- c(temp_sim, base_curve * amp_factor + trend + noise)
  }
  fd_sim <- fdata(matrix(temp_sim, nrow = 1),
                  argvals = as.numeric(seq_len(n_years * 365)))
  pt <- analyze.peak.timing(fd_sim, period = 365)

  for (k in seq_along(pt$peak_times)) {
    timing_comparison <- rbind(timing_comparison, data.frame(
      Station = representatives[j],
      Region = rep_regions[j],
      Year = pt$cycle_indices[k],
      PeakDay = pt$peak_times[k] %% 365,
      PeakTemp = pt$peak_values[k],
      stringsAsFactors = FALSE
    ))
  }
}
```

```{r peak-timing-multi-plot, fig.height = 5}
ggplot(timing_comparison, aes(x = Year, y = PeakDay,
                               color = Station, shape = Region)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, linetype = "dashed") +
  labs(title = "Peak Timing Trends Across Stations (8-year simulations)",
       subtitle = "Same warming trend (+0.3\u00b0C/yr) and noise applied to each station",
       x = "Year", y = "Peak day of year")
```

```{r peak-timing-summary}
# Summarize timing trend per station
cat("Peak timing trends (days/year):\n")
for (s in representatives) {
  sub <- timing_comparison[timing_comparison$Station == s, ]
  fit <- lm(PeakDay ~ Year, data = sub)
  cat(sprintf("  %-12s: %+.2f days/year (sd = %.1f days)\n",
              s, coef(fit)[2], sd(sub$PeakDay)))
}
```

Since all stations share the same warming scenario (+0.3°C/year, +3%/year
amplitude), the peak timing trends are primarily driven by weather noise rather
than a systematic shift. In a real climate analysis with station-specific
trends, regional differences in peak timing drift would reveal differential
warming patterns (e.g., Arctic amplification causing earlier peaks at high
latitudes).

## Change Detection & Amplitude Modulation

The simulated warming trend and increasing amplitude should be detectable.
`detect_amplitude_modulation()` characterizes the growing seasonal swing,
while `detect.seasonality.changes.auto()` finds transitions in local
seasonal strength.

```{r amplitude-modulation}
am <- detect_amplitude_modulation(fd_temp, period = 365)
print(am)
```

```{r am-plot, fig.height = 4}
if (!is.null(am$amplitude_curve)) {
  plot(am)
}
```

**The amplitude envelope** is computed via the Hilbert transform: the analytic
signal *z(t) = x(t) + i H[x(t)]* wraps the detrended temperature series into
a complex signal whose modulus |*z(t)*| traces the instantaneous amplitude.
The resulting curve shows how the peak-to-trough temperature swing evolves
across the 8 years. An upward-sloping envelope confirms the +3%/year amplitude
growth. When the slope exceeds the `modulation_threshold` (default 15%), the
function classifies the pattern as `"emerging"` — seasonality is getting stronger.

```{r am-detail}
cat("Modulation type:", am$modulation_type, "\n")
cat("Amplitude trend:", round(am$amplitude_trend, 4), "\n")
cat("Modulation score:", round(am$modulation_score, 4), "\n")
cat("Amplitude range:", round(am$min_strength, 2), "to", round(am$max_strength, 2), "\n")
```

```{r change-detection}
changes <- detect.seasonality.changes.auto(fd_temp, period = 365)
cat("Auto threshold:", round(changes$computed_threshold, 3), "\n")
cat("Change points found:", nrow(changes$change_points), "\n")

if (nrow(changes$change_points) > 0) {
  knitr::kable(changes$change_points, digits = 2,
               caption = "Detected seasonality changes")
}
```

```{r change-plot, fig.height = 4}
sc_changes <- data.frame(
  Day = seq_along(changes$strength_curve),
  Strength = changes$strength_curve
)

ggplot(sc_changes, aes(x = Day, y = Strength)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_hline(yintercept = changes$computed_threshold,
             linetype = "dashed", color = "red") +
  geom_vline(xintercept = seq(365, n_years * 365, by = 365),
             linetype = "dotted", alpha = 0.3) +
  labs(title = "Local Seasonal Strength with Auto Threshold (Otsu)",
       subtitle = "Dashed red = threshold; dotted = year boundaries",
       x = "Day", y = "Strength")
```

Amplitude modulation is detected because the annual temperature swing grows
~3% per year — later years have both warmer summers and colder winters. The
change detection may flag transitions where the accumulated amplitude
increase crosses a threshold.

## Conclusions

- **Period detection:** All four methods (SAZED, Autoperiod, CFD-Autoperiod,
  FFT) converge on the 365-day annual cycle. Cross-method agreement provides
  high confidence in the result.
- **Spectral analysis:** The Lomb-Scargle periodogram shows a strong 365-day
  peak. `detect.periods()` identifies the fundamental annual cycle and its
  harmonics (half-year, quarter-year), each with period, confidence, and
  strength — confirming the non-sinusoidal shape of the seasonal pattern.
- **Matrix Profile** confirms the annual motif using non-parametric shape
  matching — robust even with year-to-year noise.
- **Decomposition:** STL, SSA, and classical decomposition all extract a
  consistent seasonal component and recover the +0.3°C/year warming trend.
- **Seasonal strength** is high, classified as **StableSeasonal** — the
  period is reliable and the seasonal pattern dominates the signal.
- **SSA** correctly assigns the dominant annual oscillation (89% variance)
  to the seasonal slot, with the warming trend extracted separately.
- **Amplitude modulation** is detected: the annual temperature swing
  increases over the 8-year period, simulating the effect of growing
  climate variability.
- **Peak timing** across all 35 stations shows the summer maximum occurs
  around day 200–210 (late July), with Pacific stations peaking slightly
  later. Peak temperature decreases strongly with latitude.

## See Also

- `vignette("articles/seasonal-analysis")` — seasonal analysis tutorial
  (methodology and synthetic examples)
- `vignette("articles/example-canadian-weather")` — the same dataset used
  for FANOVA, FOSR, and classification
- `vignette("articles/simulation-toolbox")` — simulating functional data

## References

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd ed.
  Springer. Chapter 13 (Canadian Weather example).
- Cleveland, R.B., Cleveland, W.S., McRae, J.E. and Terpenning, I. (1990).
  STL: A seasonal-trend decomposition procedure based on loess. *Journal
  of Official Statistics*, 6, 3-73.
