---
title: "Biopharmaceutical Batch Monitoring: Penicillin Fermentation"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
set.seed(2024)
```

## Overview

A biopharmaceutical manufacturer monitors 100,000 L penicillin fed-batch
fermentations. Each batch produces time-series trajectories for temperature,
dissolved oxygen, and penicillin concentration. Batches are run under three
control strategies (recipe-driven, operator-controlled, advanced process
control), and some batches experience process faults. The goals are:

1. Characterize batch-to-batch variability via FPCA
2. Build a monitoring chart from normal batches and detect faults
3. Compare control strategy tightness
4. Predict final yield from early-stage process trajectories

This article uses synthetic data that mimics the structure of the
[IndPenSim](https://www.kaggle.com/datasets/stephengoldie/big-databiopharmaceutical-manufacturing)
dataset (Goldrick et al., 2015). The simulation reproduces the key features:
logistic penicillin production curves, controlled fermentation temperature,
dissolved oxygen consumption during growth, and realistic batch-to-batch
variability under different control strategies.

| Step | What It Does | Outcome |
|------|-------------|---------|
| 1. Simulate batches | Generate 100 batch trajectories under 4 control regimes | Penicillin, temperature, DO2 as `fdata` |
| 2. Exploratory analysis | Spaghetti plots and group means by control strategy | Visual comparison of regimes |
| 3. FPCA | Extract dominant modes of batch variation | Eigenfunction interpretation |
| 4. Process monitoring | Phase I from recipe batches, Phase II fault detection | T-squared chart + detection rates |
| 5. Precision & F1 | Evaluate monitoring quality | Precision-recall trade-off |
| 6. Yield prediction | Functional regression: early trajectory to final yield | Early warning model |

**Dataset reference:** Goldrick S. et al. (2015), *The development of an
industrial-scale fed-batch fermentation simulation*, Journal of Biotechnology,
193:70-82. The full 100-batch dataset (2.5 GB, 39 process variables + 2200
Raman channels) is available on
[Kaggle](https://www.kaggle.com/datasets/stephengoldie/big-databiopharmaceutical-manufacturing).

## 1. Simulating Batch Fermentation Data

We simulate 100 batches on a common 200-hour time grid. The penicillin
production curve follows logistic kinetics; temperature is controlled near
a setpoint with group-specific tightness; dissolved oxygen dips during the
exponential growth phase and recovers as growth slows.

```{r batch-parameters}
# Simulation parameters
n_batches <- 100
m <- 201                                  # time points (every 1 hour)
t_grid <- seq(0, 200, length.out = m)

# Control strategy groups
groups <- factor(c(
  rep("Recipe",   30),                    # batches 1-30
  rep("Operator", 30),                    # batches 31-60
  rep("APC",      30),                    # batches 61-90 (Advanced Process Control)
  rep("Fault",    10)                     # batches 91-100
), levels = c("Recipe", "Operator", "APC", "Fault"))

# Smooth basis for correlated noise
M_basis <- 6
phi <- eFun(t_grid, M = M_basis, type = "Fourier")
```

```{r generate-batches}
# Group-specific parameters:
#   Pmax: max penicillin concentration (g/L)
#   tmid: inflection point of logistic growth (hours)
#   noise: trajectory noise level
group_params <- list(
  Recipe   = list(Pmax_mu = 50, Pmax_sd = 3, tmid_mu = 80, tmid_sd = 3,
                  T_sd = 0.3, DO2_sd = 0.3, pen_noise = 1.2),
  Operator = list(Pmax_mu = 47, Pmax_sd = 5, tmid_mu = 85, tmid_sd = 5,
                  T_sd = 0.8, DO2_sd = 0.6, pen_noise = 2.0),
  APC      = list(Pmax_mu = 52, Pmax_sd = 2, tmid_mu = 78, tmid_sd = 2,
                  T_sd = 0.15, DO2_sd = 0.2, pen_noise = 0.6),
  Fault    = list(Pmax_mu = 45, Pmax_sd = 6, tmid_mu = 88, tmid_sd = 6,
                  T_sd = 0.5, DO2_sd = 0.5, pen_noise = 2.5)
)

pen_data  <- matrix(0, n_batches, m)
temp_data <- matrix(0, n_batches, m)
do2_data  <- matrix(0, n_batches, m)
yields    <- numeric(n_batches)

for (i in seq_len(n_batches)) {
  gp <- group_params[[as.character(groups[i])]]

  # Per-batch kinetic parameters
  Pmax_i <- rnorm(1, gp$Pmax_mu, gp$Pmax_sd)
  tmid_i <- rnorm(1, gp$tmid_mu, gp$tmid_sd)
  k_i    <- rnorm(1, 0.06, 0.005)

  # Penicillin: logistic growth + smooth noise
  pen_mean <- Pmax_i / (1 + exp(-k_i * (t_grid - tmid_i)))
  pen_noise <- phi %*% (rnorm(M_basis) * gp$pen_noise *
                         exp(-(seq_len(M_basis)) / 2))
  pen_data[i, ] <- pmax(0, pen_mean + as.numeric(pen_noise))

  # Temperature: controlled at 298 K + noise
  temp_noise <- phi %*% (rnorm(M_basis) * gp$T_sd *
                          exp(-(seq_len(M_basis)) / 2))
  temp_data[i, ] <- 298 + as.numeric(temp_noise)

  # Dissolved oxygen: dip during growth, recover
  do2_mean <- 8 - 4.5 * exp(-((t_grid - tmid_i + 10)^2) / (2 * 25^2))
  do2_noise <- phi %*% (rnorm(M_basis) * gp$DO2_sd *
                         exp(-(seq_len(M_basis)) / 2))
  do2_data[i, ] <- pmax(0, do2_mean + as.numeric(do2_noise))

  yields[i] <- pen_data[i, m]
}

# --- Inject specific faults into batches 91-100 ---

# Batches 91-93: temperature drift (+4 K over last 80 hours)
for (i in 91:93) {
  drift <- pmax(0, (t_grid - 120) / 80) * 4
  temp_data[i, ] <- temp_data[i, ] + drift
}

# Batches 94-96: dissolved oxygen crash (sudden drop at ~100 h)
for (i in 94:96) {
  t_fault <- sample(90:110, 1)
  drop <- -3 * pnorm(t_grid, mean = t_fault, sd = 5)
  do2_data[i, ] <- pmax(0, do2_data[i, ] + drop)
}

# Batches 97-99: substrate feed fault (penicillin production stalls at ~120 h)
for (i in 97:99) {
  t_stall <- sample(110:130, 1)
  pen_data[i, t_grid > t_stall] <- pen_data[i, which.min(abs(t_grid - t_stall))]
  yields[i] <- pen_data[i, m]
}

# Batch 100: combined fault (temperature drift + DO2 crash)
temp_data[100, ] <- temp_data[100, ] + pmax(0, (t_grid - 100) / 100) * 5
do2_data[100, ]  <- pmax(0, do2_data[100, ] -
                          3 * pnorm(t_grid, mean = 105, sd = 5))
```

```{r create-fdata}
# Convert to fdata objects
fd_pen  <- fdata(pen_data,  argvals = t_grid)
fd_temp <- fdata(temp_data, argvals = t_grid)
fd_do2  <- fdata(do2_data,  argvals = t_grid)

cat("Batches:", n_batches, "\n")
cat("Time points per batch:", m, "(0-200 hours)\n")
cat("Control strategies:", paste(levels(groups), collapse = ", "), "\n")
cat("Fault batches: 91-100\n")
```

## 2. Batch Trajectory Exploration

Each batch produces a penicillin production curve over 200 hours. We plot
all trajectories colored by control strategy.

```{r trajectory-plot, fig.height = 7}
# Build long data frame for plotting
build_df <- function(fd, varname) {
  n <- nrow(fd$data)
  data.frame(
    time = rep(t_grid, each = n),
    value = as.vector(t(fd$data)),
    batch = rep(seq_len(n), times = m),
    group = rep(groups, times = m),
    variable = varname
  )
}

df_all <- rbind(
  build_df(fd_pen, "Penicillin (g/L)"),
  build_df(fd_temp, "Temperature (K)"),
  build_df(fd_do2, "Dissolved O2 (mg/L)")
)
df_all$variable <- factor(df_all$variable,
                           levels = c("Penicillin (g/L)", "Temperature (K)",
                                      "Dissolved O2 (mg/L)"))

ggplot(df_all, aes(x = .data$time, y = .data$value,
                    group = .data$batch, color = .data$group)) +
  geom_line(alpha = 0.25, linewidth = 0.3) +
  facet_wrap(~ .data$variable, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(Recipe = "#4A90D9", Operator = "#E69F00",
                                APC = "#009E73", Fault = "#D55E00")) +
  labs(title = "100 Batch Trajectories by Control Strategy",
       x = "Time (hours)", y = NULL, color = "Strategy") +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

Key observations:

- **Penicillin** follows a logistic growth curve; APC batches (green) are
  tightest, operator batches (yellow) show the most scatter.
- **Temperature** is held near 298 K; fault batches (red) show clear drift.
- **Dissolved oxygen** dips during the growth phase; fault batches show
  abnormal profiles (early crash, sustained depression).

```{r group-means, fig.height = 4}
# Mean penicillin curves per group
group_means <- do.call(rbind, lapply(levels(groups), function(g) {
  idx <- which(groups == g)
  data.frame(
    time = t_grid,
    mean_pen = colMeans(pen_data[idx, , drop = FALSE]),
    group = g
  )
}))

ggplot(group_means, aes(x = .data$time, y = .data$mean_pen,
                         color = .data$group)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(Recipe = "#4A90D9", Operator = "#E69F00",
                                APC = "#009E73", Fault = "#D55E00")) +
  labs(title = "Mean Penicillin Production by Control Strategy",
       x = "Time (hours)", y = "Penicillin (g/L)", color = "Strategy") +
  theme(legend.position = "bottom")
```

APC batches reach the highest mean yield with the fastest onset, reflecting
tighter glucose feed control. Fault batches plateau earlier due to injected
process deviations.

## 3. FPCA: Modes of Batch Variation

We apply FPCA to the 90 non-fault penicillin trajectories (batches 1-90) to
extract the dominant modes of variation.

```{r fpca-chart}
# FPCA via spm.phase1 on non-fault batches
fd_normal <- fdata(pen_data[1:90, ], argvals = t_grid)
chart_all <- spm.phase1(fd_normal, ncomp = 8, alpha = 0.01, seed = 42)

# Cumulative variance
eig <- chart_all$eigenvalues
cum_var <- cumsum(eig) / sum(eig)

ncomp_sel <- spm.ncomp.select(eig, method = "variance90", threshold = 0.9)
cat("Selected components (90% variance):", ncomp_sel, "\n")
```

```{r scree, fig.height = 4}
df_scree <- data.frame(
  pc = seq_along(eig),
  eigenvalue = eig,
  cum_var = cum_var
)

p1 <- ggplot(df_scree, aes(x = .data$pc, y = .data$eigenvalue)) +
  geom_col(fill = "#4A90D9", width = 0.6) +
  geom_vline(xintercept = ncomp_sel + 0.5, linetype = "dashed",
             color = "#D55E00") +
  labs(title = "Scree Plot", x = "Component", y = "Eigenvalue")

p2 <- ggplot(df_scree, aes(x = .data$pc, y = .data$cum_var)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "#D55E00") +
  geom_vline(xintercept = ncomp_sel + 0.5, linetype = "dashed",
             color = "#D55E00", alpha = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Cumulative Variance", x = "Component",
       y = "Variance Explained")

p1 + p2
```

The first few components capture almost all batch-to-batch variation. The
dominant mode typically represents variation in maximum yield (vertical shift
of the logistic plateau), while the second mode captures timing differences
(horizontal shift of the growth onset).

```{r fpca-scores, fig.height = 5}
# Standard PCA on centered batch trajectories
pca_res <- prcomp(pen_data[1:90, ], center = TRUE, scale. = FALSE)

df_scores <- data.frame(
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  group = groups[1:90]
)

ggplot(df_scores, aes(x = .data$PC1, y = .data$PC2, color = .data$group)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.9, linetype = "dashed", linewidth = 0.6) +
  scale_color_manual(values = c(Recipe = "#4A90D9", Operator = "#E69F00",
                                APC = "#009E73", Fault = "#D55E00")) +
  labs(title = "FPC Score Plot: Batch Variability by Control Strategy",
       subtitle = "90% confidence ellipses",
       x = "FPC 1 (yield level)", y = "FPC 2 (growth timing)",
       color = "Strategy") +
  theme(legend.position = "bottom")
```

The score plot reveals clear separation between control strategies:

- **APC** batches (green) cluster tightly near the center -- consistent
  process performance with low variability.
- **Operator** batches (yellow) show the widest scatter -- human
  intervention introduces batch-to-batch inconsistency.
- **Recipe** batches (blue) fall between the two -- predictable but
  less optimized than APC.

## 4. Process Monitoring: Phase I Calibration

We build a T-squared control chart from the recipe-driven batches (1-30),
which represent the baseline "in-control" process.

```{r phase1}
fd_recipe <- fdata(pen_data[1:30, ], argvals = t_grid)
chart <- spm.phase1(fd_recipe, ncomp = ncomp_sel, alpha = 0.01, seed = 42)
print(chart)
```

```{r phase1-plot}
plot(chart)
```

## 5. Fault Detection: Monitoring Batches 91-100

We now monitor the 10 fault batches against the recipe-derived control
limits using Shewhart, CUSUM, and MEWMA charts.

```{r monitor-faults}
fd_fault <- fdata(pen_data[91:100, ], argvals = t_grid)
n_fault <- 10

# Shewhart
mon_fault <- spm.monitor(chart, fd_fault)
n_det_shew <- sum(mon_fault$t2.alarm)

# CUSUM
cusum_fault <- spm.cusum(chart, fd_fault, k = 0.5, h = 5.0)
n_det_cusum <- sum(cusum_fault$alarm)

# MEWMA
mewma_fault <- spm.mewma(chart, fd_fault, lambda = 0.2)
n_det_mewma <- sum(mewma_fault$alarm)

cat("Fault detection (out of", n_fault, "fault batches):\n")
cat("  Shewhart:", n_det_shew, sprintf("(%.0f%%)\n", 100 * n_det_shew / n_fault))
cat("  CUSUM:   ", n_det_cusum, sprintf("(%.0f%%)\n", 100 * n_det_cusum / n_fault))
cat("  MEWMA:   ", n_det_mewma, sprintf("(%.0f%%)\n", 100 * n_det_mewma / n_fault))
```

```{r fault-panel, fig.height = 5}
df_fault <- data.frame(
  batch = rep(91:100, 3),
  value = c(mon_fault$t2, cusum_fault$cusum.statistic,
            mewma_fault$mewma.statistic),
  ucl = rep(c(chart$t2.ucl, cusum_fault$ucl, mewma_fault$ucl), each = n_fault),
  alarm = c(mon_fault$t2.alarm, cusum_fault$alarm, mewma_fault$alarm),
  method = factor(rep(c("Shewhart T2", "CUSUM", "MEWMA"), each = n_fault),
                  levels = c("Shewhart T2", "CUSUM", "MEWMA"))
)

ggplot(df_fault, aes(x = .data$batch, y = .data$value)) +
  geom_col(aes(fill = .data$alarm), width = 0.7) +
  geom_hline(aes(yintercept = .data$ucl), linetype = "dashed",
             color = "#D55E00", linewidth = 0.6) +
  scale_fill_manual(values = c("FALSE" = "#4A90D9", "TRUE" = "#D55E00"),
                    labels = c("In-control", "Alarm")) +
  facet_wrap(~ .data$method, scales = "free_y", ncol = 1) +
  labs(title = "Monitoring Fault Batches (91-100)",
       x = "Batch", y = "Statistic", fill = "") +
  scale_x_continuous(breaks = 91:100) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

### False Positive Check

We also monitor 30 fresh operator batches (31-60) as a non-fault comparison.
While they differ from the recipe baseline, they should produce moderate
alarm rates -- not the high detection seen for true faults.

```{r fpr-check}
fd_operator <- fdata(pen_data[31:60, ], argvals = t_grid)
mon_op <- spm.monitor(chart, fd_operator)
cusum_op <- spm.cusum(chart, fd_operator, k = 0.5, h = 5.0)
mewma_op <- spm.mewma(chart, fd_operator, lambda = 0.2)

fpr_shew  <- mean(mon_op$t2.alarm)
fpr_cusum <- mean(cusum_op$alarm)
fpr_mewma <- mean(mewma_op$alarm)
```

### Precision, Recall, and F1

We compute precision, recall, and F1 treating fault batches as positives
and operator batches as the non-fault reference.

```{r pr-compute}
fp_counts <- c(
  Shewhart = sum(mon_op$t2.alarm),
  CUSUM = sum(cusum_op$alarm),
  MEWMA = sum(mewma_op$alarm)
)

pr_df <- do.call(rbind, lapply(c("Shewhart", "CUSUM", "MEWMA"), function(meth) {
  if (meth == "Shewhart") { tp <- n_det_shew }
  else if (meth == "CUSUM") { tp <- n_det_cusum }
  else { tp <- n_det_mewma }

  fp <- fp_counts[meth]
  fn <- n_fault - tp
  prec <- if ((tp + fp) > 0) tp / (tp + fp) else 0
  rec  <- tp / n_fault
  f1   <- if ((prec + rec) > 0) 2 * prec * rec / (prec + rec) else 0

  data.frame(Method = meth, TP = tp, FP = fp, FN = fn,
             Precision = prec, Recall = rec, F1 = f1)
}))

pr_table <- pr_df
pr_table$Precision <- sprintf("%.1f%%", 100 * pr_df$Precision)
pr_table$Recall <- sprintf("%.1f%%", 100 * pr_df$Recall)
pr_table$F1 <- sprintf("%.3f", pr_df$F1)

knitr::kable(pr_table, align = c("l", "r", "r", "r", "r", "r", "r"),
             caption = "Fault detection: precision, recall, and F1")
```

```{r pr-plot, fig.height = 4}
pr_long <- rbind(
  data.frame(Method = pr_df$Method, Metric = "Precision",
             Value = pr_df$Precision),
  data.frame(Method = pr_df$Method, Metric = "Recall",
             Value = pr_df$Recall),
  data.frame(Method = pr_df$Method, Metric = "F1",
             Value = pr_df$F1)
)
pr_long$Method <- factor(pr_long$Method,
                          levels = c("Shewhart", "CUSUM", "MEWMA"))
pr_long$Metric <- factor(pr_long$Metric,
                          levels = c("Precision", "Recall", "F1"))

ggplot(pr_long, aes(x = .data$Method, y = .data$Value, fill = .data$Metric)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c(Precision = "#4A90D9", Recall = "#D55E00",
                                F1 = "#009E73")) +
  labs(title = "Precision, Recall, and F1 for Fault Detection",
       subtitle = "Fault batches vs operator batches as reference",
       x = NULL, y = "Score", fill = "Metric") +
  theme(legend.position = "bottom")
```

## 6. Control Strategy Comparison

How does each control strategy compare on process consistency? We compute the
pointwise standard deviation of the penicillin trajectories within each
non-fault group.

```{r strategy-variability, fig.height = 5}
var_df <- do.call(rbind, lapply(c("Recipe", "Operator", "APC"), function(g) {
  idx <- which(groups == g)
  sd_curve <- apply(pen_data[idx, ], 2, sd)
  data.frame(time = t_grid, sd = sd_curve, group = g)
}))

ggplot(var_df, aes(x = .data$time, y = .data$sd, color = .data$group)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(Recipe = "#4A90D9", Operator = "#E69F00",
                                APC = "#009E73")) +
  labs(title = "Batch-to-Batch Variability: Pointwise SD of Penicillin",
       x = "Time (hours)", y = "SD (g/L)", color = "Strategy") +
  theme(legend.position = "bottom")
```

```{r strategy-table}
strat_summary <- do.call(rbind, lapply(c("Recipe", "Operator", "APC"), function(g) {
  idx <- which(groups == g)
  data.frame(
    Strategy = g,
    N = length(idx),
    Mean_Yield = round(mean(yields[idx]), 1),
    SD_Yield = round(sd(yields[idx]), 2),
    CV_Yield = sprintf("%.1f%%", 100 * sd(yields[idx]) / mean(yields[idx]))
  )
}))

knitr::kable(strat_summary,
             caption = "Yield statistics by control strategy")
```

APC achieves the highest mean yield with the lowest coefficient of variation
(CV). Operator control shows the widest spread -- human decision-making adds
variability that automated control eliminates.

## 7. Yield Prediction from Early Process Data

Can we predict final penicillin yield from the first 100 hours of a batch?
Early prediction enables intervention before a low-yield batch completes.

```{r yield-regression}
# Functional predictor: first 100 hours of penicillin trajectory
early_idx <- t_grid <= 100
fd_early <- fdata(pen_data[1:90, early_idx], argvals = t_grid[early_idx])

# Scalar response: final yield
y_normal <- yields[1:90]

# Functional linear regression
fit <- fregre.lm(fd_early, y_normal, ncomp = ncomp_sel)

cat("R-squared:", round(fit$r.squared, 3), "\n")
cat("GCV:", round(fit$gcv, 3), "\n")
```

```{r yield-fit-plot, fig.height = 4}
df_fit <- data.frame(
  observed = y_normal,
  predicted = fit$fitted.values,
  group = groups[1:90]
)

ggplot(df_fit, aes(x = .data$observed, y = .data$predicted,
                    color = .data$group)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c(Recipe = "#4A90D9", Operator = "#E69F00",
                                APC = "#009E73")) +
  labs(title = "Yield Prediction: Observed vs Fitted",
       subtitle = "Using first 100 hours of penicillin trajectory",
       x = "Observed yield (g/L)", y = "Predicted yield (g/L)",
       color = "Strategy") +
  theme(legend.position = "bottom")
```

```{r beta-plot, fig.height = 4}
if (!is.null(fit$beta_t)) {
  df_beta <- data.frame(
    time = t_grid[early_idx],
    beta = as.numeric(fit$beta_t)
  )

  ggplot(df_beta, aes(x = .data$time, y = .data$beta)) +
    geom_line(color = "#4A90D9", linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
    labs(title = expression("Coefficient Function " * hat(beta)(t)),
         subtitle = "Weight each time point contributes to yield prediction",
         x = "Time (hours)", y = expression(hat(beta)(t)))
}
```

The coefficient function reveals **when** the process trajectory is most
predictive of final yield. Regions with large absolute beta values indicate
critical phases where process deviations have the greatest impact on the
outcome. For penicillin fermentation, the transition from lag to exponential
growth (around the inflection point) is typically the most informative.

## 8. Conclusions

- **Control strategy matters.** APC batches show the tightest clustering in
  FPCA score space and the lowest yield variability. Operator-controlled
  batches have the highest scatter, confirming that automated control
  reduces batch-to-batch inconsistency.

- **Fault detection works.** Building a Phase I chart from normal (recipe)
  batches and monitoring fault batches achieves high recall. CUSUM and MEWMA
  may catch faults that the Shewhart chart misses when the deviation is
  gradual (temperature drift, slow DO2 decline).

- **Early yield prediction is feasible.** Functional regression using only
  the first 100 hours (50% of the batch) predicts final yield with
  reasonable accuracy, enabling early intervention for underperforming
  batches.

- **FDA provides interpretable summaries.** FPCA eigenfunctions describe
  meaningful modes of variation (yield level, growth timing). The
  coefficient function from regression shows exactly which time windows
  drive yield prediction.

This workflow -- FPCA for characterization, SPM for monitoring, and
functional regression for prediction -- applies to any batch manufacturing
process where time-series trajectories are recorded.

## Using Real IndPenSim Data

To use the actual Kaggle dataset instead of synthetic data, download the
CSV from [Kaggle](https://www.kaggle.com/datasets/stephengoldie/big-databiopharmaceutical-manufacturing)
and replace the simulation section with:

```{r load-real-data, eval = FALSE}
# Read the full dataset (2.5 GB, ~113k rows x 2239 columns)
raw <- read.csv("100_Batches_IndPenSim_V3.csv")

# Split by batch and extract key process variables
batch_ids <- sort(unique(raw$Batch_ref))
t_common <- seq(0, 200, length.out = 201)

pen_data <- matrix(NA, 100, length(t_common))
for (i in seq_along(batch_ids)) {
  batch <- raw[raw$Batch_ref == batch_ids[i], ]
  pen_data[i, ] <- approx(batch$Time..h., batch$Penicillin.concentration.P.g.L.,
                           xout = t_common)$y
}

fd_pen <- fdata(pen_data, argvals = t_common)
# Then proceed with FPCA, monitoring, and regression as above
```

## See Also

- [Inline Monitoring: Detection Power](example-inline-monitoring.html) --
  FPR and power analysis on synthetic data
- [Tecator: Inline Quality Monitoring](example-tecator-monitoring.html) --
  SPM with real NIR spectra
- [Statistical Process Monitoring](statistical-process-monitoring.html) --
  SPM fundamentals
