---
title: "Sonar: Mine vs Rock Classification via Elastic TSRVF"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

Sonar signals bounced off metal cylinders (mines) and rough rocks produce
different spectral return patterns. Each observation in the UCI Sonar dataset
consists of 60 frequency-band energy measurements — a spectral profile that
we treat as a functional curve. The challenge: classify Mine vs Rock from
these profiles.

This example applies the full elastic TSRVF pipeline and compares it against
raw, smoothed, and derivative-based features. The ablation study reveals
*when* elastic alignment helps — and when simpler approaches suffice.

| Step | What It Does | Outcome |
|---|---|---|
| Signal conditioning | Smooth + resample to [0,1] | Ensure smooth derivatives for SRSF |
| Derivatives | 1st and 2nd derivative of smoothed curves | Enhance spectral features, remove baseline |
| Karcher mean | Elastic alignment via SRSF + iterative mean | Aligned curves + warping functions $\gamma_i(t)$ |
| TSRVF projection | Log map to tangent space at mean | Euclidean tangent vectors $v_i$ |
| Feature extraction | PCA on tangent vectors (amplitude) + PCA on warpings (phase) | Combined feature vector per sample |
| Ablation study | CV accuracy across 7 feature representations | Compare raw, derivatives, aligned, and elastic |

**Key finding:** On this spectral dataset, smoothed raw features with
sufficient FPCs outperform the elastic pipeline. Derivatives and elastic
alignment do not improve accuracy — the 60 frequency bands are fixed physical
measurements without genuine phase variability. This is an instructive
negative result: TSRVF shines on temporal data with timing variation (e.g.,
growth curves), but spectral data needs different tools.

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## 1. Data Preparation

The Sonar dataset contains 208 observations: 111 mines (M) and 97 rocks (R).
Each has 60 frequency-band energy measurements.

```{r data-load}
data(Sonar, package = "mlbench")

X <- as.matrix(Sonar[, 1:60])
y <- as.integer(Sonar$Class)  # 1 = M (mine), 2 = R (rock)
class_labels <- ifelse(y == 1, "Mine", "Rock")

cat("Observations:", nrow(X), "\n")
cat("  Mines:", sum(y == 1), "| Rocks:", sum(y == 2), "\n")
cat("Frequency bands:", ncol(X), "\n")

fd_raw <- fdata(X, argvals = seq(0, 1, length.out = 60))
```

```{r raw-plot, fig.height = 4}
plot(fd_raw, color = factor(class_labels),
     palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
     alpha = 0.3) +
  labs(title = "Raw Sonar Returns by Class",
       x = "Frequency Band (normalized)", y = "Energy",
       color = "Class")
```

The raw profiles overlap substantially — mines and rocks share similar spectral
shapes, making this a challenging classification task.

## 2. Signal Conditioning & Derivatives

B-spline smoothing provides well-defined derivatives for the SRSF transform
and for derivative-based features. We compute both first and second
derivatives — commonly used in spectroscopy to remove baselines and sharpen
features.

```{r conditioning}
# Smooth with B-splines on [0,1]
t_fine <- seq(0, 1, length.out = 100)
coefs <- fdata2basis(fd_raw, nbasis = 25, type = "bspline")
fd_smooth <- basis2fdata(coefs, t_fine)

# First and second derivatives
fd_d1 <- deriv(fd_smooth)
fd_d2 <- deriv(fd_d1)

cat("Smoothed grid:", ncol(fd_smooth$data), "points\n")
```

```{r deriv-plot, fig.height = 10}
p1 <- plot(fd_smooth, color = factor(class_labels),
           palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
           alpha = 0.3) +
  labs(title = "Smoothed Spectra", x = "t", y = "f(t)", color = "Class")

p2 <- plot(fd_d1, color = factor(class_labels),
           palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
           alpha = 0.3) +
  labs(title = "First Derivative", x = "t", y = "f'(t)", color = "Class")

p3 <- plot(fd_d2, color = factor(class_labels),
           palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
           alpha = 0.3) +
  labs(title = "Second Derivative", x = "t", y = "f''(t)", color = "Class")

p1 / p2 / p3
```

The first derivative highlights slope differences between classes. The second
derivative enhances curvature features — peaks and troughs in the original
signal become more pronounced. Whether these enhance or degrade classification
depends on the signal-to-noise ratio.

## 3. Elastic Alignment & Karcher Mean

The Karcher mean aligns all curves elastically via the SRSF framework. This
separates **amplitude** variation (shape differences after alignment) from
**phase** variation (timing differences captured by warping functions).

```{r alignment}
km <- karcher.mean(fd_smooth, max.iter = 20, tol = 1e-4)
aq <- alignment.quality(fd_smooth, km)
cat("Alignment quality:\n")
print(aq)
```

```{r alignment-plot, fig.height = 8}
p_orig <- plot(fd_smooth, color = factor(class_labels),
               palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
               alpha = 0.3) +
  labs(title = "Before Alignment", x = "t", y = "f(t)", color = "Class")

p_aligned <- plot(km$aligned, color = factor(class_labels),
                  palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
                  alpha = 0.3) +
  labs(title = "After Elastic Alignment", x = "t", y = "f(t)", color = "Class")

p_orig / p_aligned
```

```{r warping-plot, fig.height = 4}
plot(km$gammas, color = factor(class_labels),
     palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
     alpha = 0.3) +
  labs(title = "Warping Functions by Class",
       x = "t", y = expression(gamma(t)),
       color = "Class")
```

The warping functions cluster tightly near the identity — indicating that
the curves had little genuine phase variability to begin with. This is
expected: frequency bands are fixed physical measurements, not time points
that shift between observations.

```{r variance-decomp, fig.height = 4}
plot(aq, type = "variance") +
  labs(title = "Amplitude vs Phase Variance Decomposition")
```

## 4. TSRVF Projection

The TSRVF maps aligned curves to a tangent space at the Karcher mean, where
each curve is represented by a tangent vector capturing its shape deviation.

```{r tsrvf}
tv <- tsrvf.from.alignment(km)

# Verify reconstruction quality
recon <- tsrvf.inverse(tv)
recon_err <- mean((km$aligned$data - recon$data)^2)
cat("Mean reconstruction error:", format(recon_err, digits = 4), "\n")
```

```{r tangent-plot, fig.height = 4}
plot(tv$tangent_vectors, color = factor(class_labels),
     palette = c("Mine" = "#0072B2", "Rock" = "#D55E00"),
     alpha = 0.3) +
  labs(title = "TSRVF Tangent Vectors by Class",
       x = "t", y = expression(v[i](t)),
       color = "Class")
```

## 5. Feature Extraction

We extract features from both the amplitude (tangent vectors) and phase
(warping functions) spaces via PCA.

```{r pca-features}
# Amplitude PCA (tangent vectors)
pca_amp <- prcomp(tv$tangent_vectors$data, center = TRUE)
var_amp <- pca_amp$sdev^2 / sum(pca_amp$sdev^2)

# Phase PCA (warping functions)
pca_phase <- prcomp(km$gammas$data, center = TRUE)
var_phase <- pca_phase$sdev^2 / sum(pca_phase$sdev^2)

# Choose components for ~90% variance
k_amp <- which(cumsum(var_amp) >= 0.90)[1]
k_phase <- which(cumsum(var_phase) >= 0.90)[1]
cat("Amplitude PCs for 90%:", k_amp, "\n")
cat("Phase PCs for 90%:", k_phase, "\n")
```

```{r pca-variance, fig.height = 4}
df_var <- rbind(
  data.frame(PC = 1:min(15, length(var_amp)),
             CumVar = cumsum(var_amp[1:min(15, length(var_amp))]) * 100,
             Space = "Amplitude (tangent vectors)"),
  data.frame(PC = 1:min(15, length(var_phase)),
             CumVar = cumsum(var_phase[1:min(15, length(var_phase))]) * 100,
             Space = "Phase (warping functions)")
)

ggplot(df_var, aes(x = PC, y = CumVar, color = Space)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 90, linetype = "dashed", color = "grey50") +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative % Variance",
       color = "") +
  scale_x_continuous(breaks = seq(2, 15, by = 2))
```

```{r pc-scatter, fig.height = 4}
df_amp <- data.frame(
  PC1 = pca_amp$x[, 1], PC2 = pca_amp$x[, 2],
  Class = class_labels
)
df_ph <- data.frame(
  PC1 = pca_phase$x[, 1], PC2 = pca_phase$x[, 2],
  Class = class_labels
)

p_amp <- ggplot(df_amp, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Mine" = "#0072B2", "Rock" = "#D55E00")) +
  labs(title = "Amplitude PC Scores", x = "PC1", y = "PC2") +
  coord_equal()

p_ph <- ggplot(df_ph, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Mine" = "#0072B2", "Rock" = "#D55E00")) +
  labs(title = "Phase PC Scores", x = "PC1", y = "PC2") +
  coord_equal()

p_amp + p_ph + plot_layout(guides = "collect")
```

Neither amplitude nor phase PC scores show strong class separation in the
first two components — the discriminative information is spread across many
dimensions.

```{r combined-features}
amp_scores <- pca_amp$x[, 1:k_amp]
phase_scores <- pca_phase$x[, 1:k_phase]
combined <- cbind(amp_scores, phase_scores)
cat("Combined feature matrix:", nrow(combined), "x", ncol(combined), "\n")
```

## 6. Classification — Ablation Study

Seven feature representations, each evaluated with LDA and kNN via 10-fold
cross-validation. We use `ncomp = 10` for curve-based features (where FPCs
are computed internally) and adapt for the pre-computed combined features.

1. **Raw**: original spectra on [0,1] (no smoothing, no alignment)
2. **Smoothed**: B-spline smoothed spectra
3. **1st derivative**: first derivative of smoothed spectra
4. **2nd derivative**: second derivative of smoothed spectra
5. **Aligned**: elastically aligned curves (no tangent projection)
6. **Amplitude only**: TSRVF tangent vectors
7. **Full elastic**: combined amplitude + phase PCA scores

```{r ablation}
# Prepare feature sets
fd_combined <- fdata(combined, argvals = seq_len(ncol(combined)))

feature_sets <- list(
  "Raw"            = fd_raw,
  "Smoothed"       = fd_smooth,
  "1st derivative" = fd_d1,
  "2nd derivative" = fd_d2,
  "Aligned"        = km$aligned,
  "Amplitude only" = tv$tangent_vectors,
  "Full elastic"   = fd_combined
)

methods <- c("lda", "knn")

# Run ablation
set.seed(42)
ablation <- data.frame(
  Features = character(),
  Method = character(),
  CV_Accuracy = numeric(),
  stringsAsFactors = FALSE
)

for (feat_name in names(feature_sets)) {
  for (meth in methods) {
    fd_i <- feature_sets[[feat_name]]
    nc <- min(10, ncol(fd_i$data) - 1)
    cv_i <- fclassif.cv(fd_i, y, method = meth, ncomp = nc,
                         nfold = 10, seed = 42)
    ablation <- rbind(ablation, data.frame(
      Features = feat_name,
      Method = toupper(meth),
      CV_Accuracy = round(1 - cv_i$error.rate, 3),
      stringsAsFactors = FALSE
    ))
  }
}

# Order features for display
ablation$Features <- factor(ablation$Features,
                             levels = names(feature_sets))

knitr::kable(ablation, caption = "10-Fold CV Accuracy: Ablation Study",
             row.names = FALSE)
```

```{r ablation-plot, fig.height = 5}
ggplot(ablation, aes(x = Features, y = CV_Accuracy, fill = Method)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("LDA" = "#0072B2", "KNN" = "#D55E00")) +
  labs(title = "Classification Accuracy by Feature Representation",
       x = "", y = "10-Fold CV Accuracy") +
  ylim(0, 1) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 7. Best Model Deep Dive

Train the best configuration on the full data and examine the confusion matrix.

```{r best-model, fig.height = 4}
# Use the best method from ablation
best_idx <- which.max(ablation$CV_Accuracy)
best_method <- tolower(as.character(ablation$Method[best_idx]))
best_feat <- as.character(ablation$Features[best_idx])
best_fd <- feature_sets[[best_feat]]
best_nc <- min(10, ncol(best_fd$data) - 1)

cat("Best configuration:", best_feat, "+", toupper(best_method), "\n")

best_fit <- fclassif(best_fd, y, method = best_method, ncomp = best_nc)
cat("Training accuracy:", round(best_fit$accuracy, 3), "\n")

# Confusion matrix
plot(best_fit)
```

```{r per-class}
# Per-class accuracy
cm <- table(Predicted = best_fit$predicted, Actual = y)
class_acc <- diag(cm) / colSums(cm)
cat("Mine accuracy:", round(class_acc[1], 3), "\n")
cat("Rock accuracy:", round(class_acc[2], 3), "\n")
```

## 8. Conclusions

- **Smoothed raw features win** on this dataset. kNN with 10 FPCs on the
  smoothed (or raw) spectra achieves the highest CV accuracy.

- **Derivatives do not help.** First and second derivatives lose information
  from the spectral *levels* without adding discriminative power. Unlike
  chemometric data (where derivatives remove baseline shifts), sonar spectra
  have meaningful absolute energy levels.

- **Elastic alignment does not improve classification.** The warping functions
  cluster tightly near the identity, confirming that these frequency-band
  measurements lack genuine phase variability. The alignment step fits noise
  rather than removing nuisance variation.

- **When TSRVF shines vs. when it doesn't.** Elastic alignment and TSRVF are
  designed for temporal curves where observations share common features that
  occur at different times (e.g., growth spurts, ECG peaks). For spectral
  data — where the x-axis represents fixed frequency bands rather than
  stretchable time — there is no phase variability to separate, and the
  simpler Euclidean analysis is both faster and more accurate.

- **Lesson for practitioners:** Always include the raw baseline in an ablation
  study. Sophisticated machinery can degrade performance when its assumptions
  (here: meaningful phase variability) do not hold.

## See Also

- `vignette("articles/tsrvf")` — TSRVF theory and linearized elastic analysis
- `vignette("articles/elastic-alignment")` — elastic alignment and Karcher mean
- `vignette("articles/functional-classification")` — supervised classification
  methods (LDA, QDA, kNN, kernel, DD)
- `vignette("articles/example-tecator-regression")` — Tecator NIR spectra where
  derivatives *do* improve prediction

## References

- Gorman, R.P. and Sejnowski, T.J. (1988). Analysis of hidden units in a
  layered network trained to classify sonar targets. *Neural Networks*,
  1(1), 75--89.
- Srivastava, A., Wu, W., Kurtek, S., Klassen, E., and Marron, J.S. (2011).
  Registration of functional data using the Fisher-Rao metric.
  *arXiv:1103.3817*.
- Tucker, J.D. (2014). Generative models for functional data using phase and
  amplitude separation. *Computational Statistics & Data Analysis*, 61, 50--66.
