---
title: "Canadian Weather: Regional Climate Patterns"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

A climate analyst wants to characterize how geography shapes seasonal
temperature and precipitation patterns across Canada. Functional ANOVA tests
whether climate curves differ by region, while function-on-scalar regression
reveals which geographic variables (latitude, longitude) drive seasonal
variation.

| Step | What It Does | Outcome |
|---|---|---|
| Data preparation | Load 35 stations × 365 daily temperature/precipitation curves | `fdata` objects colored by 4 regions |
| FANOVA (temperature) | Permutation test for regional mean differences | Strong rejection: regions differ significantly |
| Pairwise FANOVA | Test all region pairs | Pacific/Arctic most distinct; Atlantic/Continental overlap in summer |
| FOSR: temperature ~ geography | Regress temperature curves on latitude + longitude | Latitude dominates in winter; longitude effect peaks in spring/fall |
| Precipitation analysis | FANOVA + FOSR on precipitation curves | Regions also differ; Pacific has distinct wet-season pattern |
| Classification | Classify region from temperature curves | Cross-validated accuracy by region |

**Key result:** Latitude is the strongest predictor of temperature seasonality
— its effect is largest in winter when Arctic/Continental stations drop far
below Atlantic/Pacific ones. The permutation FANOVA confirms highly significant
regional differences (p < 0.002).

The Canadian Weather dataset is a classic benchmark in functional data analysis
(Ramsay and Silverman, 2005).

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Data Preparation

The dataset contains daily temperature and precipitation averaged over 1960–1994
at 35 Canadian weather stations, grouped into 4 regions.

```{r data-load}
data(CanadianWeather, package = "fda")

# Temperature: 365 x 35 matrix (days x stations) — transpose for fdata
temp_mat <- t(CanadianWeather$dailyAv[, , "Temperature.C"])
fd_temp <- fdata(temp_mat, argvals = 1:365)

# Precipitation (log mm)
precip_mat <- t(CanadianWeather$dailyAv[, , "log10precip"])
fd_precip <- fdata(precip_mat, argvals = 1:365)

# Region labels
region <- factor(CanadianWeather$region)

# Geographic coordinates
coords <- CanadianWeather$coordinates
latitude <- coords[, "N.latitude"]
longitude <- -abs(coords[, "W.longitude"])  # negative for West

cat("Stations:", nrow(fd_temp$data), "\n")
cat("Grid points:", ncol(fd_temp$data), "(daily)\n")
cat("Regions:", paste(levels(region), table(region), sep = ": ", collapse = ", "), "\n")
```

## Temperature Curves by Region

```{r temp-by-region, fig.height = 5}
region_colors <- c("Arctic" = "#56B4E9", "Atlantic" = "#E69F00",
                    "Continental" = "#009E73", "Pacific" = "#CC79A7")

# Create data frame for ggplot
temp_df <- data.frame(
  Day = rep(1:365, each = nrow(fd_temp$data)),
  Temp = as.vector(t(fd_temp$data)),
  Station = rep(rownames(fd_temp$data), 365),
  Region = rep(as.character(region), 365)
)

ggplot(temp_df, aes(x = Day, y = Temp, group = Station, color = Region)) +
  geom_line(alpha = 0.6) +
  scale_color_manual(values = region_colors) +
  labs(title = "Daily Temperature by Region (35 Canadian Stations)",
       x = "Day of year", y = "Temperature (°C)") +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct"))
```

Arctic and Continental stations show the widest annual swing (cold winters, warm
summers), while Pacific stations have the mildest climate. The key question:
are these regional differences statistically significant across the whole year?

## Functional ANOVA: Do Regions Differ?

```{r fanova-temp}
anova_temp <- fanova(fd_temp, region, n.perm = 500)
print(anova_temp)
```

The small p-value confirms that the four regions have significantly different
mean temperature curves — the visual separation is real, not a sampling
artifact.

```{r fanova-temp-plot, fig.height = 4}
plot(anova_temp) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "FANOVA: Regional Mean Temperature Curves",
       x = "Day of year", y = "Temperature (°C)")
```

The group means show the expected pattern: Arctic coldest year-round,
Continental with the largest amplitude, Pacific mildest, Atlantic intermediate.
The separation is greatest in winter (days 1–90 and 275–365).

## Pairwise FANOVA

Which specific region pairs differ?

```{r pairwise-fanova}
# Test each pair of regions
pairs <- combn(levels(region), 2)
pairwise_results <- data.frame(
  Region1 = character(), Region2 = character(),
  F_stat = numeric(), p_value = numeric(),
  stringsAsFactors = FALSE
)

for (j in 1:ncol(pairs)) {
  idx <- region %in% pairs[, j]
  result <- fanova(fd_temp[idx, ], region[idx, drop = TRUE], n.perm = 500)
  pairwise_results <- rbind(pairwise_results, data.frame(
    Region1 = pairs[1, j], Region2 = pairs[2, j],
    F_stat = round(result$global.statistic, 2),
    p_value = result$p.value
  ))
}

knitr::kable(pairwise_results, caption = "Pairwise FANOVA p-values")
```

Arctic vs Pacific shows the largest F-statistic (greatest separation).
Continental vs Atlantic may overlap more, particularly in summer when both
regions reach similar peak temperatures.

## FOSR: Temperature ~ Latitude + Longitude

Function-on-scalar regression reveals *how* geographic variables shape the
temperature curve at each day of the year.

```{r fosr-temp}
predictors <- cbind(latitude, longitude)
fosr_temp <- fosr(fd_temp, predictors, lambda = 1)
print(fosr_temp)
```

```{r fosr-temp-plot, fig.height = 5}
plot(fosr_temp) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct"))
```

### Interpreting the Coefficient Functions

- **Latitude $\hat{\beta}_1(t)$**: Negative throughout the year (higher latitude
  → colder). The effect is *strongest in winter* — a 1° increase in latitude
  has a bigger cooling effect in January than in July.
- **Longitude $\hat{\beta}_2(t)$**: More subtle. Western stations (more negative
  longitude) tend to be warmer in winter (Pacific influence) but the effect
  reverses or diminishes in summer.

### FPC-Based FOSR

```{r fosr-fpc}
fosr_fpc <- fosr.fpc(fd_temp, predictors, ncomp = 5)
cat("FPC-based R-squared:", round(fosr_fpc$r.squared, 4), "\n")
```

### Prediction: What Would a Station at Given Coordinates Look Like?

```{r fosr-predict, fig.height = 4}
# Predict temperature curve for hypothetical stations
new_locs <- matrix(c(
  45, -75,   # Montreal-like (Continental)
  55, -120,  # Northern BC (Continental/Pacific)
  65, -135   # Yukon (Arctic)
), nrow = 3, byrow = TRUE)

pred_curves <- predict(fosr_temp, new_locs)

pred_df <- data.frame(
  Day = rep(1:365, 3),
  Temp = as.vector(t(pred_curves$data)),
  Location = rep(c("45°N, 75°W (Montreal-like)",
                    "55°N, 120°W (Northern BC)",
                    "65°N, 135°W (Yukon)"), each = 365)
)

ggplot(pred_df, aes(x = Day, y = Temp, color = Location)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "Predicted Temperature Curves by Location",
       x = "Day of year", y = "Temperature (°C)")
```

## Precipitation Analysis

### FANOVA on Precipitation

```{r fanova-precip}
anova_precip <- fanova(fd_precip, region, n.perm = 500)
print(anova_precip)
```

```{r fanova-precip-plot, fig.height = 4}
plot(anova_precip) +
  scale_x_continuous(breaks = c(1, 91, 182, 274),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  labs(title = "FANOVA: Regional Mean Precipitation Curves",
       x = "Day of year", y = "log10(precipitation, mm)")
```

Pacific stations show a distinct wet-winter/dry-summer Mediterranean-like
pattern, while Continental stations have summer-dominated precipitation.

### FOSR on Precipitation

```{r fosr-precip}
fosr_precip <- fosr(fd_precip, predictors, lambda = 1)
cat("Precipitation FOSR R-squared:", round(fosr_precip$r.squared, 4), "\n")
```

## Classification: Region from Temperature

Can we classify a station's region from its temperature curve alone?

```{r classify}
# Functional classification with cross-validation
classif_result <- fclassif.cv(fd_temp, region, nfold = 5)
print(classif_result)
```

```{r classify-detail}
# Also fit on full data to see confusion matrix
classif_full <- fclassif(fd_temp, region)

cat("Overall accuracy:", round(classif_full$accuracy * 100, 1), "%\n")
print(classif_full$confusion)
```

Pacific and Arctic stations are typically classified correctly due to their
distinctive temperature profiles. Misclassification is most common between
Atlantic and Continental stations, whose summer temperatures overlap.

## Conclusions

- **FANOVA** confirms highly significant regional differences in both
  temperature and precipitation curves (p < 0.002 in both cases).
- **Latitude** is the dominant geographic predictor of temperature seasonality.
  Its effect is strongest in winter, when the Arctic/Continental vs
  Atlantic/Pacific divide is most pronounced.
- **Longitude** has a more subtle effect, primarily reflecting the moderating
  influence of the Pacific Ocean on western stations.
- **FOSR** with 2 scalar predictors explains a substantial fraction of the
  curve-to-curve temperature variation — geography alone goes a long way.
- **Classification** from temperature curves correctly identifies most regions,
  with Pacific and Arctic being the most distinctive.

## See Also

- `vignette("articles/scalar-on-function")` — scalar-on-function regression methods
- `vignette("articles/function-on-scalar")` — FOSR and FANOVA methodology
- `vignette("articles/example-tecator-regression")` — real-data regression:
  NIR spectra → fat content
- `vignette("articles/clustering")` — alternative approach: unsupervised
  grouping of curves

## References

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd ed.
  Springer. Chapter 13 (Canadian Weather example).
- Cuevas, A., Febrero, M., and Fraiman, R. (2004). An anova test for functional
  data. *Computational Statistics & Data Analysis*, 47(1), 111-122.
