---
title: "Model Explainability"
description: >
  Comprehensive toolkit for interpreting functional regression models:
  22 methods across global, domain-level, and local explanations.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Explainability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

Understanding how functional regression models make predictions is crucial for
building trust and ensuring reliable deployments. The fdars explainability
toolkit provides **22 methods** organized into four categories.

```{=html}
<img src="../reference/figures/explainability-diagram.svg" alt="Explainability Toolkit Overview: 22 methods across 4 categories" style="width:100%; max-width:700px;">
```

### Supported Models

Two model types are supported. Most methods work with **both**; a few are
specific to one:

| Model | Class | Response | Fitted by |
|:------|:------|:---------|:----------|
| **Linear** | `fregre.lm` | Continuous $y \in \mathbb{R}$ | `fregre.lm()` |
| **Logistic** | `fregre.logistic` | Binary $y \in \{0,1\}$ | `functional.logistic()` |

Both models project curves onto FPC scores and regress on them. The same R
functions (e.g. `fregre.pdp`) automatically dispatch to the correct Rust
backend based on the model class.

### Method Reference

| Method | Function | Models | Category |
|:-------|:---------|:------:|:---------|
| Partial Dependence (PDP) | `fregre.pdp` | lm, logistic | Global (Score) |
| Accumulated Local Effects | `fregre.ale` | lm, logistic | Global (Score) |
| Sobol Sensitivity | `fregre.sobol` | lm, logistic | Global (Score) |
| Friedman H-statistic | `fregre.friedman` | lm, logistic | Global (Score) |
| Beta Decomposition | `fregre.beta.decomp` | lm, logistic | Global (Score) |
| Pointwise Importance | `fregre.pointwise` | lm, logistic | Global (Domain) |
| Saliency Maps | `fregre.saliency` | lm, logistic | Global (Domain) |
| Domain Importance | `fregre.domain` | lm, logistic | Global (Domain) |
| Significant Regions | `fregre.significant.regions` | lm, logistic | Global (Domain) |
| SHAP Values | `fregre.shap` | lm, logistic | Local |
| LIME | `fregre.lime` | lm, logistic | Local |
| Counterfactual | `fregre.counterfactual` | lm, logistic | Local |
| Anchor | `fregre.anchor` | lm, logistic | Local |
| Prototypes | `fregre.prototype` | lm, logistic | Local |
| Permutation Importance | `fregre.importance` | lm, logistic | Diagnostics |
| VIF | `fregre.vif` | lm, logistic | Diagnostics |
| Regression Depth | `fregre.depth` | lm, logistic | Diagnostics |
| Conditional Importance | `fregre.conditional.importance` | lm, logistic | Diagnostics |
| Influence (Leverage, Cook's D) | `fregre.influence` | **lm only** | Diagnostics |
| DFBETAS / DFFITS | `fregre.dfbetas` | **lm only** | Diagnostics |
| Calibration / ECE | `fregre.calibration`, `fregre.ece` | **logistic only** | Diagnostics |
| Explanation Stability | `fregre.stability` | *model-free* | Diagnostics |

## Setup

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

### Simulated Data

We create a dataset where the response depends on specific regions of the
functional domain. In applied terms, imagine spectroscopy curves where only
certain wavelengths predict a chemical concentration — the goal of
explainability is to recover which wavelengths matter.

```{r create-data}
n <- 80
m <- 100
t_grid <- seq(0, 1, length.out = m)

# True coefficient function: peaked around t=0.3 and t=0.7
beta_true <- 2 * dnorm(t_grid, 0.3, 0.08) - 1.5 * dnorm(t_grid, 0.7, 0.08)
beta_true <- beta_true / max(abs(beta_true))

# Generate functional predictors
X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) + 0.5 * rnorm(1) * cos(4 * pi * t_grid) +
            rnorm(m, sd = 0.1)
}

# Generate response: y_i = integral of beta(t) * X_i(t) dt + noise
dt <- t_grid[2] - t_grid[1]
y <- numeric(n)
for (i in 1:n) {
  y[i] <- sum(beta_true * X[i, ]) * dt + rnorm(1, sd = 0.3)
}

fdataobj <- fdata(X, argvals = t_grid)
```

```{r plot-true-beta}
df_beta <- data.frame(t = t_grid, beta = beta_true)

ggplot(df_beta, aes(x = t, y = beta)) +
  geom_line(color = "#2E8B57", linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_ribbon(aes(ymin = 0, ymax = beta), fill = "#2E8B57", alpha = 0.15) +
  annotate("text", x = 0.3, y = max(beta_true) * 1.1, label = "Important\nregion",
           color = "#2E8B57", size = 3) +
  annotate("text", x = 0.7, y = min(beta_true) * 1.1, label = "Important\nregion",
           color = "#D55E00", size = 3) +
  labs(title = expression("True Coefficient Function " * beta(t)),
       subtitle = "Positive peak at t=0.3, negative peak at t=0.7",
       x = "t", y = expression(beta(t)))
```

### Fitting a Linear Model

The functional linear model represents the scalar response as

$$y_i = \alpha + \int_0^1 \beta(t)\, X_i(t)\, dt + \varepsilon_i.$$

In practice, `fregre.lm()` projects each curve onto $K$ functional principal
components (FPCs) and fits OLS on the resulting scores:

$$y_i \approx \alpha + \sum_{k=1}^{K} b_k \xi_{ik}, \qquad \xi_{ik} = \int X_i(t)\,\phi_k(t)\,dt.$$

This is the model that all explainability methods operate on.

```{r fit-model}
model <- fregre.lm(fdataobj, y, ncomp = 5)
cat("R\u00b2:", round(model$r.squared, 4), "\n")
cat("RMSE:", round(sqrt(mean(model$residuals^2)), 4), "\n")
```

```{r plot-fit}
df_fit <- data.frame(Observed = y, Fitted = model$fitted.values)

ggplot(df_fit, aes(x = Observed, y = Fitted)) +
  geom_point(color = "#4A90D9", size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Observed vs Fitted Values",
       subtitle = paste("R\u00b2 =", round(model$r.squared, 3)),
       x = "Observed y", y = "Fitted y")
```

---

## Global Explanations: Score-Level

These methods answer: **which FPC scores drive predictions?** They operate on
the multivariate regression $y_i = \alpha + \sum_k b_k \xi_{ik}$ and reveal
how changing each score affects the predicted response.

*All methods in this section support both `fregre.lm` and
`functional.logistic`.*

```{=html}
<img src="../reference/figures/explainability-global-score.svg" alt="Score-level global explanations: PDP, ALE, Sobol, Beta Decomposition, Importance" style="width:100%; max-width:700px;">
```

### Partial Dependence Plots (PDP)

**In plain English:** A PDP shows how the model's prediction changes as one FPC
score varies, averaging over all other scores. Think of it as: "if we could
separately control just PC1, what would happen to the prediction?"

**Mathematically:** The partial dependence function for the $k$-th score is

$$\hat{f}_k(\xi_k) = \frac{1}{n}\sum_{i=1}^{n} \hat{f}(\xi_{i1},\ldots,\xi_{i,k-1},\,\xi_k,\,\xi_{i,k+1},\ldots,\xi_{iK}).$$

For each value of $\xi_k$, we plug it into every observation's score vector
(keeping all other scores at their actual values) and average the predictions.
A flat PDP means the score has no effect; a steep slope means it matters.

**Limitation:** When FPC scores are correlated, PDP evaluates the model at
unrealistic score combinations (e.g. high PC1 with low PC2 even if they always
co-occur). See ALE below for a correction.

```{r pdp}
pdp1 <- fregre.pdp(model, data = fdataobj, component = 1, n.grid = 30)
pdp2 <- fregre.pdp(model, data = fdataobj, component = 2, n.grid = 30)

df_pdp <- rbind(
  data.frame(x = pdp1$grid_values, y = pdp1$pdp_curve, Component = "PC1"),
  data.frame(x = pdp2$grid_values, y = pdp2$pdp_curve, Component = "PC2")
)

ggplot(df_pdp, aes(x = x, y = y)) +
  geom_line(color = "#4A90D9", linewidth = 1) +
  facet_wrap(~ Component, scales = "free_x") +
  labs(title = "Partial Dependence Plots",
       subtitle = "Marginal effect of each PC on prediction",
       x = "PC Score", y = "Partial Dependence")
```

### Accumulated Local Effects (ALE)

**In plain English:** ALE fixes PDP's correlation problem. Instead of
substituting scores globally, it looks at how predictions change **locally**
within narrow bins of the score distribution. This avoids extrapolating into
unrealistic score combinations.

**Mathematically:** The ALE for score $k$ accumulates local differences:

$$\widehat{\text{ALE}}_k(\xi_k) = \sum_{j=1}^{J(\xi_k)} \frac{1}{|N_j|}\sum_{i \in N_j}\left[\hat{f}(\xi_{ij,k}^{(U)}) - \hat{f}(\xi_{ij,k}^{(L)})\right],$$

where $N_j$ is the set of observations in bin $j$, and $\xi^{(U)}$,
$\xi^{(L)}$ are the bin's upper and lower bounds. The result is centred so
$E[\widehat{\text{ALE}}_k] = 0$.

**When to prefer ALE over PDP:** Always, unless FPC scores are (nearly)
uncorrelated — which they are by construction in FPCA. So for `fregre.lm`,
PDP and ALE will typically agree closely. They may diverge for logistic models
or when scalar covariates are correlated with FPC scores.

```{r ale}
ale1 <- fregre.ale(model, data = fdataobj, component = 1, n.bins = 15)

df_ale <- data.frame(x = ale1$bin_midpoints, y = ale1$ale_values)

ggplot(df_ale, aes(x = x, y = y)) +
  geom_step(color = "#4A90D9", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  labs(title = "ALE Plot for PC1",
       subtitle = "Accumulated Local Effects (robust to score correlations)",
       x = "PC1 Score", y = "ALE")
```

### Sobol Sensitivity Indices

**In plain English:** Sobol indices answer "what fraction of the variation in
predictions is due to each FPC score?" First-order indices measure the direct
effect of each score alone; total-order indices add all interactions involving
that score. If first-order and total-order are nearly equal, the score acts
independently.

**Mathematically:** The first-order index for score $k$ is

$$S_k = \frac{\text{Var}_{\xi_k}\!\left[E_{\boldsymbol{\xi}_{-k}}[\hat{y} \mid \xi_k]\right]}{\text{Var}[\hat{y}]},$$

and the total-order index is

$$S_k^T = 1 - \frac{\text{Var}_{\boldsymbol{\xi}_{-k}}\!\left[E_{\xi_k}[\hat{y} \mid \boldsymbol{\xi}_{-k}]\right]}{\text{Var}[\hat{y}]}.$$

Both are estimated via Monte Carlo sampling. First-order indices sum to 1 when
there are no interactions; total-order indices can sum to more than 1.

```{r sobol}
sob <- fregre.sobol(model, data = fdataobj, y = y)

df_sobol <- data.frame(
  Component = paste0("PC", seq_along(sob$first_order)),
  First = sob$first_order,
  Total = sob$total_order
)

df_sobol_long <- rbind(
  data.frame(Component = df_sobol$Component, Value = df_sobol$First, Index = "First-order"),
  data.frame(Component = df_sobol$Component, Value = df_sobol$Total, Index = "Total-order")
)

ggplot(df_sobol_long, aes(x = Component, y = Value, fill = Index)) +
  geom_col(position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("First-order" = "#4A90D9", "Total-order" = "#D55E00")) +
  labs(title = "Sobol Sensitivity Indices",
       subtitle = "First-order vs total-order variance decomposition",
       x = "Component", y = "Index") +
  theme(legend.position = "bottom")
```

### Friedman H-Statistic

**In plain English:** The H-statistic tells you whether two FPC scores interact
— that is, whether the effect of one score depends on the value of the other.
A value near 0 means the scores affect the prediction independently; a large
value means their combined effect is more (or less) than the sum of parts.

**Mathematically:** For scores $j$ and $k$, the H-statistic is

$$H_{jk}^2 = \frac{\sum_i \left[\hat{f}_{jk}(\xi_{ij}, \xi_{ik}) - \hat{f}_j(\xi_{ij}) - \hat{f}_k(\xi_{ik})\right]^2}{\sum_i \hat{f}_{jk}(\xi_{ij}, \xi_{ik})^2},$$

where $\hat{f}_{jk}$ is the two-dimensional PDP and $\hat{f}_j$, $\hat{f}_k$
are the marginal PDPs. For a purely linear model (no interactions),
$H_{jk}^2 = 0$ exactly.

```{r friedman}
fried <- fregre.friedman(model, data = fdataobj, component.j = 1,
                         component.k = 2, n.grid = 15)
cat("H-statistic (PC1 x PC2):", round(fried$h_squared, 4), "\n")
```

### Beta Decomposition

**In plain English:** Since the model predicts using $y = \alpha + b_1 \xi_1 + \ldots + b_K \xi_K$, we can ask "how much of the variation in predictions comes
from each component?" Beta decomposition answers this by weighting each
coefficient by the variance of its score.

**Mathematically:** The variance proportion for component $k$ is

$$\text{VP}_k = \frac{b_k^2\,\text{Var}(\xi_k)}{\sum_{j=1}^K b_j^2\,\text{Var}(\xi_j)}.$$

This is the simplest global importance measure — it uses only the model
coefficients and score variances, with no Monte Carlo sampling.

```{r beta-decomp}
decomp <- fregre.beta.decomp(model)

df_decomp <- data.frame(
  Component = paste0("PC", 1:length(decomp$variance_proportion)),
  Weight = decomp$variance_proportion
)
df_decomp$Component <- factor(df_decomp$Component,
                               levels = df_decomp$Component[order(-df_decomp$Weight)])

ggplot(df_decomp, aes(x = Component, y = Weight)) +
  geom_col(fill = "#4A90D9", alpha = 0.7) +
  labs(title = "Beta Decomposition: Variance Proportion",
       subtitle = "Fraction of prediction variance attributable to each PC",
       x = "Component", y = "Variance Proportion")
```

---

## Global Explanations: Domain-Level

These methods answer: **where on the functional domain** $t$ **does the model
focus?** While score-level methods operate on FPC scores (multivariate), these
methods translate importance back to the original curve domain — telling you
which time points, wavelengths, or spatial locations matter.

*All methods in this section support both `fregre.lm` and
`functional.logistic`.*

```{=html}
<img src="../reference/figures/explainability-global-domain.svg" alt="Domain-level global explanations: Pointwise importance, Saliency, Domain importance, Significant regions" style="width:100%; max-width:700px;">
```

### Pointwise Importance

**In plain English:** For each point $t$ on the domain, this method combines
how much each eigenfunction $\phi_k(t)$ contributes at that point with how
important the corresponding score is to the prediction. The result is a curve
showing which parts of the domain the model "pays attention to".

**Mathematically:** The pointwise importance at $t$ is

$$I(t) = \sum_{k=1}^{K} |b_k| \cdot |\phi_k(t)| \cdot \text{Var}(\xi_k)^{1/2}.$$

High values indicate regions where (a) the eigenfunctions are large and
(b) the regression coefficients are large. Compare the result to the true
$\beta(t)$ — the peaks should align.

```{r pointwise}
pw <- fregre.pointwise(model)

df_pw <- data.frame(t = t_grid, importance = pw$importance)

ggplot(df_pw, aes(x = t, y = importance)) +
  geom_line(color = "#7B2D8E", linewidth = 1) +
  geom_ribbon(aes(ymin = 0, ymax = importance), fill = "#7B2D8E", alpha = 0.15) +
  labs(title = "Pointwise Importance",
       subtitle = "Which domain points drive predictions most?",
       x = "t", y = "Importance")
```

### Saliency Maps

**In plain English:** Saliency maps show how sensitive the prediction is to
small perturbations of the curve at each point $t$. If wiggling the curve at
$t = 0.3$ changes the prediction a lot, that point has high saliency. This is
the functional analogue of gradient-based saliency in image classification.

**Mathematically:** For each observation $i$, the saliency at $t$ is

$$s_i(t) = \frac{\partial \hat{y}_i}{\partial X_i(t)}.$$

In the linear model this equals $\hat{\beta}(t)$ for every observation.
`fregre.saliency()` returns the mean absolute saliency across all observations,
which highlights where the model is most sensitive.

```{r saliency}
sal <- fregre.saliency(model, data = fdataobj)

df_sal <- data.frame(t = t_grid, saliency = sal$mean_absolute_saliency)

ggplot(df_sal, aes(x = t, y = saliency)) +
  geom_line(color = "#C0392B", linewidth = 1) +
  geom_ribbon(aes(ymin = 0, ymax = saliency), fill = "#C0392B", alpha = 0.15) +
  labs(title = "Mean Absolute Saliency Map",
       subtitle = "Gradient-based sensitivity across the domain",
       x = "t", y = "|Saliency|")
```

### Domain Importance

**In plain English:** Instead of looking at individual points, domain importance
uses a **sliding window** along $t$ to assess which regions matter. This is
useful when the effect is spread over a neighbourhood rather than concentrated
at a single point.

**Mathematically:** The domain importance at grid point $t_j$ aggregates the
absolute coefficient function within a window of width $w$:

$$D(t_j) = \sum_{|t_l - t_j| \leq w/2} |\hat{\beta}(t_l)|.$$

Larger `window.width` values produce smoother importance curves.

```{r domain}
dom <- fregre.domain(model, window.width = 5)

df_dom <- data.frame(t = t_grid, importance = dom$pointwise_importance)

ggplot(df_dom, aes(x = t, y = importance)) +
  geom_line(color = "#2E8B57", linewidth = 1) +
  geom_ribbon(aes(ymin = 0, ymax = importance), fill = "#2E8B57", alpha = 0.15) +
  labs(title = "Domain Importance (Sliding Window)",
       subtitle = paste("Window width:", dom$window_width),
       x = "t", y = "Importance")
```

### Significant Regions

**In plain English:** This method identifies contiguous intervals of $t$ where
the coefficient $\beta(t)$ is statistically different from zero. Unlike
pointwise importance (which gives a continuous importance score), this gives a
binary answer: significant or not. Use this when you need formal inference
about which domain regions matter.

**Mathematically:** At each $t$, a pointwise test is performed using the
standard error of $\hat{\beta}(t)$. Contiguous intervals exceeding the
significance threshold at level $\alpha$ are reported.

```{r sig-regions}
sig <- fregre.significant.regions(model, alpha = 0.05)

if (is.null(sig) || length(sig$start_idx) == 0) {
  cat("No significant regions detected at alpha = 0.05.\n")
  cat("This can happen when the signal is spread across many FPCs or\n")
  cat("when pointwise tests lack power. Try bootstrap CI instead.\n")
} else {
  cat("Significant regions found:", length(sig$start_idx), "\n")
  for (i in seq_along(sig$start_idx)) {
    cat(sprintf("  Region %d: t = [%.3f, %.3f] (%s)\n",
                i, t_grid[sig$start_idx[i]], t_grid[sig$end_idx[i]],
                sig$direction[i]))
  }
}
```

---

## Local Explanations

These methods answer: **why did the model make this specific prediction?** Each
method explains a single observation's prediction, which is essential when you
need to justify individual decisions (e.g. medical diagnostics, process
control).

*All methods in this section support both `fregre.lm` and
`functional.logistic`.*

```{=html}
<img src="../reference/figures/explainability-local.svg" alt="Local explanations: SHAP, LIME, Counterfactual, Anchor, Prototype, Calibration" style="width:100%; max-width:700px;">
```

### SHAP Values

**In plain English:** SHAP assigns each FPC score a "contribution" to a
specific prediction. If observation $i$ has a high predicted value, SHAP
tells you which scores pushed the prediction up and which pulled it down. The
contributions always add up exactly to the difference between the prediction
and the average prediction — nothing is missing.

**Mathematically:** SHAP values are the unique Shapley values from cooperative
game theory. For observation $i$ and score $k$:

$$\hat{y}_i - E[\hat{y}] = \sum_{k=1}^{K} \phi_k^{(i)}.$$

Each $\phi_k^{(i)}$ averages the marginal contribution of score $k$ across
all possible orderings of scores (hence: fair, additive, exact).

**The beeswarm plot** below shows every observation as a dot, coloured by its
feature value. This reveals whether high values of a score push predictions up
(dots on the right) or down (dots on the left).

```{r shap}
shap <- fregre.shap(model, data = fdataobj)

obs_idx <- 1
ncomp <- ncol(shap$values)
df_shap <- data.frame(
  Component = paste0("PC", 1:ncomp),
  SHAP = shap$values[obs_idx, ]
)
df_shap$Direction <- ifelse(df_shap$SHAP > 0, "Positive", "Negative")
df_shap$Component <- factor(df_shap$Component,
                             levels = df_shap$Component[order(abs(df_shap$SHAP))])

ggplot(df_shap, aes(x = Component, y = SHAP, fill = Direction)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(values = c("Positive" = "#4A90D9", "Negative" = "#D55E00")) +
  coord_flip() +
  labs(title = paste("SHAP Values for Observation", obs_idx),
       subtitle = paste("Predicted:", round(model$fitted.values[obs_idx], 3),
                        "| Observed:", round(y[obs_idx], 3)),
       x = NULL, y = "SHAP Value") +
  theme(legend.position = "bottom")
```

```{r shap-beeswarm}
df_shap_all <- data.frame(
  Component = rep(paste0("PC", 1:ncomp), each = n),
  SHAP = as.vector(shap$values),
  Feature_Value = as.vector(model$.fpca_scores[, 1:ncomp])
)

ggplot(df_shap_all, aes(x = SHAP, y = Component, color = Feature_Value)) +
  geom_jitter(width = 0, height = 0.2, size = 1.5, alpha = 0.6) +
  scale_color_gradient2(low = "#4A90D9", mid = "gray90", high = "#D55E00",
                        midpoint = 0) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  labs(title = "SHAP Summary Plot",
       subtitle = "Each point = one observation's attribution for that component",
       x = "SHAP Value", y = NULL, color = "Feature\nValue")
```

### LIME

**In plain English:** LIME explains one prediction by fitting a simple linear
model in the **neighbourhood** of that observation. It perturbs the FPC scores
slightly, observes how the prediction changes, and fits a weighted linear model.
The coefficients of this local model tell you which scores matter for
*this particular observation*.

**Mathematically:** LIME minimises

$$\sum_{i=1}^{M} w_i \left[\hat{f}(\tilde{\boldsymbol{\xi}}_i) - g(\tilde{\boldsymbol{\xi}}_i)\right]^2,$$

where $g$ is a simple (e.g. linear) surrogate, $\tilde{\boldsymbol{\xi}}_i$ are
perturbed score vectors, and $w_i = \exp(-d_i^2 / \sigma^2)$ weights by
distance from the target observation.

The **local $R^2$** indicates how well the linear surrogate approximates the
model in this neighbourhood. Low $R^2$ means the model has local nonlinearities.

```{r lime}
lime_res <- fregre.lime(model, data = fdataobj, observation = 1,
                        n.samples = 300, kernel.width = 1, seed = 42)

ncomp_lime <- length(lime_res$attributions)
df_lime <- data.frame(
  Component = paste0("PC", seq_len(ncomp_lime)),
  Attribution = lime_res$attributions
)
df_lime$Direction <- ifelse(df_lime$Attribution > 0, "Positive", "Negative")

ggplot(df_lime, aes(x = reorder(Component, abs(Attribution)), y = Attribution,
                     fill = Direction)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(values = c("Positive" = "#4A90D9", "Negative" = "#D55E00")) +
  coord_flip() +
  labs(title = paste("LIME Explanation (Obs 1, local R\u00b2 =",
                     round(lime_res$local_r_squared, 3), ")"),
       x = NULL, y = "Local Attribution") +
  theme(legend.position = "none")
```

### Counterfactual Explanations

**In plain English:** A counterfactual answers "what is the smallest change
to this observation's scores that would make the model predict a target value?"
This is useful for actionable feedback — e.g. "if PC1 were 0.5 units higher,
the predicted quality score would cross the acceptance threshold."

**Mathematically:** The counterfactual finds

$$\boldsymbol{\xi}^* = \arg\min_{\boldsymbol{\xi}} \|\boldsymbol{\xi} - \boldsymbol{\xi}_i\|_2 \quad \text{s.t.} \quad \hat{f}(\boldsymbol{\xi}^*) = y_{\text{target}}.$$

The distance $\|\boldsymbol{\xi}^* - \boldsymbol{\xi}_i\|$ indicates how far
the observation is from the target in score space.

```{r counterfactual}
target_val <- mean(y) + sd(y)
cf <- fregre.counterfactual(model, data = fdataobj, observation = 1,
                            target.value = target_val)

cat("Original prediction:", round(cf$original_prediction, 3), "\n")
cat("Counterfactual prediction:", round(cf$counterfactual_prediction, 3), "\n")
cat("Score distance:", round(cf$distance, 3), "\n")
```

### Anchor Explanations

**In plain English:** An anchor is a set of **sufficient conditions** on the
FPC scores such that any observation meeting those conditions receives a similar
prediction. Think of it as an if-then rule: "IF PC1 is between 2.1 and 3.4
AND PC3 is between -1.0 and 0.5, THEN the prediction is approximately the
same, with 95% precision."

**Mathematically:** An anchor $A$ satisfies

$$P\left[\hat{f}(\boldsymbol{\xi}) \approx \hat{f}(\boldsymbol{\xi}_i) \mid A(\boldsymbol{\xi})\right] \geq 1 - \delta,$$

where $\delta$ is the precision threshold and $A(\boldsymbol{\xi})$ denotes
that $\boldsymbol{\xi}$ satisfies the anchor conditions.

**Coverage** is the fraction of observations satisfying the anchor. Narrow
anchors have high precision but low coverage; broader anchors are more general
but less precise.

```{r anchor}
anch <- fregre.anchor(model, data = fdataobj, observation = 1,
                      precision = 0.90, n.bins = 5)

cat("Anchor precision:", round(anch$precision, 3), "\n")
cat("Coverage:", round(anch$coverage, 3), "\n")
cat("Matching observations:", anch$n_matching, "\n")
if (length(anch$condition_components) > 0) {
  cat("Conditions:\n")
  for (i in seq_along(anch$condition_components)) {
    cat(sprintf("  PC%d in [%.3f, %.3f]\n",
                anch$condition_components[i],
                anch$condition_lower[i], anch$condition_upper[i]))
  }
}
```

### Prototypes and Criticisms

**In plain English:** Prototypes are the most representative observations —
the "typical" curves in each part of score space. Criticisms are the opposite:
observations that are poorly represented by the prototypes. Together, they give
you a small set of example curves that summarise the entire dataset.

**Mathematically:** Prototypes and criticisms are selected using the Maximum
Mean Discrepancy (MMD) witness function. Prototypes minimise the MMD between
the selected subset and the full dataset; criticisms are points where the
witness function is maximally negative (underrepresented).

```{r prototype}
proto <- fregre.prototype(model, ncomp = model$ncomp,
                          n.prototypes = 3, n.criticisms = 3)

cat("Prototype indices:", paste(proto$prototype_indices, collapse = ", "), "\n")
cat("Criticism indices:", paste(proto$criticism_indices, collapse = ", "), "\n")
```

```{r plot-proto}
proto_curves <- fdataobj$data[proto$prototype_indices, , drop = FALSE]
crit_curves <- fdataobj$data[proto$criticism_indices, , drop = FALSE]

df_proto <- do.call(rbind, lapply(seq_len(nrow(proto_curves)), function(i) {
  data.frame(t = t_grid, y = proto_curves[i, ],
             type = "Prototype", id = paste("P", proto$prototype_indices[i]))
}))
df_crit <- do.call(rbind, lapply(seq_len(nrow(crit_curves)), function(i) {
  data.frame(t = t_grid, y = crit_curves[i, ],
             type = "Criticism", id = paste("C", proto$criticism_indices[i]))
}))
df_pc <- rbind(df_proto, df_crit)

ggplot(df_pc, aes(x = t, y = y, color = type, group = id)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("Prototype" = "#4A90D9", "Criticism" = "#D55E00")) +
  labs(title = "Prototypes and Criticisms",
       subtitle = "Representative vs atypical curves",
       x = "t", y = "X(t)", color = NULL) +
  theme(legend.position = "bottom")
```

---

## Classification Explainability

All the methods above also work with `functional.logistic()` models via
automatic dispatch — you call the same R function and it detects the model
class. In addition, two methods are **logistic-only**: calibration and ECE.

| Method | lm | logistic | Notes |
|:-------|:--:|:--------:|:------|
| PDP, ALE, Sobol, ... | yes | yes | On probability scale for logistic |
| Calibration | | yes | Hosmer-Lemeshow test |
| ECE / MCE | | yes | Binned calibration error |

### Logistic Setup

```{r logistic-setup}
y_binary <- as.numeric(y > median(y))
logit_model <- functional.logistic(fdataobj, y_binary, ncomp = 3)
cat("Accuracy:", round(logit_model$accuracy, 4), "\n")
```

### PDP and ALE for Classification

For logistic models, PDP and ALE show the effect of FPC scores on the
**predicted probability** of the positive class (rather than a continuous
response).

```{r logistic-pdp}
pdp_logit <- fregre.pdp(logit_model, data = fdataobj, component = 1, n.grid = 30)

df_pdp_logit <- data.frame(x = pdp_logit$grid_values, y = pdp_logit$pdp_curve)

ggplot(df_pdp_logit, aes(x = x, y = y)) +
  geom_line(color = "#D55E00", linewidth = 1) +
  labs(title = "PDP for Logistic Model (PC1)",
       subtitle = "Effect of PC1 on predicted probability",
       x = "PC1 Score", y = "Predicted Probability")
```

```{r logistic-ale}
ale_logit <- fregre.ale(logit_model, data = fdataobj, component = 1, n.bins = 15)

df_ale_logit <- data.frame(x = ale_logit$bin_midpoints, y = ale_logit$ale_values)

ggplot(df_ale_logit, aes(x = x, y = y)) +
  geom_step(color = "#D55E00", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  labs(title = "ALE for Logistic Model (PC1)",
       x = "PC1 Score", y = "ALE")
```

### Beta Decomposition (Logistic)

```{r logistic-beta-decomp}
decomp_logit <- fregre.beta.decomp(logit_model)

df_decomp_logit <- data.frame(
  Component = paste0("PC", 1:length(decomp_logit$variance_proportion)),
  Weight = decomp_logit$variance_proportion
)
df_decomp_logit$Component <- factor(df_decomp_logit$Component,
  levels = df_decomp_logit$Component[order(-df_decomp_logit$Weight)])

ggplot(df_decomp_logit, aes(x = Component, y = Weight)) +
  geom_col(fill = "#D55E00", alpha = 0.7) +
  labs(title = "Beta Decomposition (Logistic)",
       subtitle = "Importance of each FPC for the classification boundary",
       x = "Component", y = "Variance Proportion")
```

### Calibration Diagnostics

**In plain English:** Calibration checks whether predicted probabilities match
observed frequencies. If the model says "70% chance of class 1" for a group of
observations, about 70% should actually be class 1. Poor calibration means the
model is over- or under-confident.

**logistic only.** Continuous (`fregre.lm`) models do not produce
probabilities.

- **Brier score**: Mean squared error of probability forecasts (lower = better).
- **Hosmer-Lemeshow $\chi^2$**: Goodness-of-fit test for calibration.
- **ECE** (Expected Calibration Error): Average absolute gap between predicted
  probability and observed frequency.
- **MCE** (Maximum Calibration Error): Worst-case gap across bins.

```{r logistic-calibration}
cal <- fregre.calibration(logit_model, y = y_binary, n.groups = 5)
cat("Brier score:", round(cal$brier_score, 4), "\n")
cat("Hosmer-Lemeshow chi2:", round(cal$hosmer_lemeshow_chi2, 4), "\n")
```

```{r logistic-ece}
ece <- fregre.ece(logit_model, y = y_binary, n.bins = 5)
cat("ECE:", round(ece$ece, 4), "\n")
cat("MCE:", round(ece$mce, 4), "\n")
```

---

## Perturbation Spaces: Current Approach and Alternatives

All explainability methods in fdars work by **perturbing** the model inputs
and observing how predictions change. The key design choice is: *in which
space do we define perturbations?* For functional data this is nontrivial,
because a curve $X(t)$ is infinite-dimensional and different representations
capture different aspects of its variation.

```{=html}
<img src="../reference/figures/explainability-perturbation.svg" alt="Four perturbation spaces: FPC scores, basis coefficients, pointwise, SRVF+warping" style="width:100%; max-width:700px;">
```

### Current implementation: FPC scores

The fdars explainability toolkit operates in **FPC score space**. All
perturbations — PDP grid sweeps, SHAP coalitions, LIME neighbourhood
sampling, counterfactual optimisation — modify the $K$-dimensional score
vector $\boldsymbol{\xi} = (\xi_1,\ldots,\xi_K)$ rather than the curve
itself.

**Why this works well:** For `fregre.lm` and `functional.logistic`, the model
is *defined* in FPC score space — the prediction function is
$\hat{y} = f(\boldsymbol{\xi})$. Perturbing scores is exact: we change the
actual model inputs. The dimensionality is low ($K \approx 3$--$15$), so
methods like SHAP (which enumerate subsets) and Sobol (which require Monte
Carlo sampling) are computationally tractable.

### Alternative perturbation spaces

For other model types or data structures, FPC scores may not be the most
faithful perturbation space:

| Space | Perturb what? | Best for | Pros | Cons |
|:------|:-------------|:---------|:-----|:-----|
| **FPC scores** | $\xi_k$ | `fregre.lm`, `functional.logistic` | Low-dim, exact for FPC-based models | Assumes FPCA captures relevant variation |
| **Basis coefficients** | B-spline / Fourier $c_j$ | Basis-represented models | Low-dim, preserves smoothness | Tied to a specific basis choice |
| **Pointwise / segments** | $X(t_j)$ or region averages | Nonparametric (`fregre.np`), any model | Model-agnostic | High-dim unless segmented; may produce unrealistic curves |
| **SRVF + warping** | Amplitude and phase separately | Elastic alignment models | Respects shape geometry | Complex; requires elastic framework |
| **Wavelet coefficients** | Multi-resolution $w_{j,k}$ | Localized features | Captures local + global structure | Basis-dependent; less common in FDA |

### When FPC scores are the wrong space

FPC scores are a **linear, global** representation. They can be misleading
when:

1. **Phase variation dominates.** Elastic models exist precisely because FPCA
   cannot separate timing from amplitude. Perturbing FPC scores of unaligned
   curves produces unrealistic shapes that mix the two. The natural space for
   elastic models is SRVF amplitude scores + warping function scores.

2. **Localized features matter.** FPCA eigenfunctions are global (nonzero
   everywhere). If a small bump at $t = 0.3$ is the key predictor, many FPCs
   are needed to represent it. A B-spline or wavelet perturbation space would
   be more faithful and lower-dimensional for that feature.

3. **Nonlinear manifold structure.** If curves live on a curved manifold
   (e.g. positive functions, shapes, compositional data), linear FPC
   perturbations can produce off-manifold curves — negative where they should
   be positive, wrong topology, etc.

### The pragmatic path: FPC projection as a universal adapter

For models that do not internally use FPC scores (e.g. `fregre.np`), a
practical approach is to use FPCA as a **perturbation coordinate system**
without modifying the model:

1. Perform FPCA on the training data.
2. Project each observation into score space.
3. Perturb scores as usual (PDP, SHAP, etc.).
4. Reconstruct the perturbed curve from the modified scores.
5. Call the model's `predict()` on the reconstructed curve.

This treats FPCA as the *perturbation language*, not part of the model itself.
It works well as a first-order approximation — "which modes of variation
matter most?" — but the explanations are in terms of FPC modes, which may not
align with the features the nonparametric model actually uses.

**The principled alternative** is model-specific perturbation spaces: elastic
models perturb in (aligned amplitude, warping) space; basis models perturb
coefficients; NP models use functional distance-based neighbourhoods. This
is a potential direction for future fdars versions.

### Practical recipe: Elastic alignment + linear regression

You don't need native elastic explainability to get explanations for data with
phase variation. The key insight: **elastic alignment separates amplitude from
phase**. Once curves are aligned, FPCA captures amplitude variation only, and
a standard `fregre.lm()` model operates in a clean amplitude score space.

The pipeline is:

1. **Align curves** using `alignment.karcher.mean()` to obtain aligned
   curves $\tilde{X}_i(t)$ and warping functions $\gamma_i(t)$.
2. **Fit `fregre.lm()`** on the aligned curves $\tilde{X}_i(t)$.
3. **Apply any of the 22 explainability methods** — they all work on the
   resulting `fregre.lm` model.

```{r align-then-explain, eval=FALSE}
# Step 1: Elastic alignment (removes phase variation)
aligned <- alignment.karcher.mean(fdataobj)

# Step 2: Fit linear model on aligned curves
model_aligned <- fregre.lm(aligned$aligned, y, ncomp = 5)

# Step 3: All explainability methods work!
pdp_aligned <- fregre.pdp(model_aligned, data = aligned$aligned, component = 1)
shap_aligned <- fregre.shap(model_aligned, data = aligned$aligned)
lime_aligned <- fregre.lime(model_aligned, data = aligned$aligned, observation = 1)
```

**What this gives you:**

| Aspect | Before alignment | After alignment |
|:-------|:-----------------|:----------------|
| FPC scores capture | Amplitude + phase (mixed) | Amplitude only (clean) |
| Explainability of | Mixed variation | Pure shape effects |
| Phase information | Confounded with amplitude | Available separately via $\gamma_i(t)$ |
| Model support | Need `elastic.regression` | Standard `fregre.lm` (all 22 methods) |

**What about phase effects?** The warping functions $\gamma_i(t)$ from
alignment capture timing variation. You can analyse them separately — e.g.
via `horiz.fpca()` to extract phase modes — or include summary statistics of
$\gamma_i$ (such as warp complexity or mean shift) as scalar covariates in
the regression model.

**When this is preferable to native elastic regression:**

- You want the full explainability toolkit (22 methods) rather than the
  limited `elastic.attribution()` (amplitude vs phase decomposition only).
- You want model diagnostics (influence, VIF, LOO-CV) for quality control.
- The amplitude component is the primary predictor and phase variation is
  nuisance — alignment removes it cleanly.

**When this is not enough:**

- Phase variation *is* the predictor (e.g. timing of a peak predicts the
  response). Then elastic regression models the relationship directly in
  SRVF space, and the align-then-regress pipeline discards the signal.
- Alignment is unreliable (e.g. curves have very different shapes, not just
  different timing). Forcing alignment can distort the amplitude component.

### Summary

The current FPC score approach covers ~90% of use cases — it is the natural
coordinate system for FDA, low-dimensional, and exact for the two supported
model types. The align-then-regress pipeline extends coverage to elastic data
by using alignment as a preprocessing step. For the remaining cases, FPC
projection provides a reasonable approximation with documented caveats, while
model-specific perturbation spaces would provide more faithful explanations at
greater implementation cost.

---

## References

- Apley, D.W. and Zhu, J. (2020). Visualizing the effects of predictor
  variables in black box supervised learning models. *JRSS-B*, 82(4), 1059--1086.

- Friedman, J.H. and Popescu, B.E. (2008). Predictive learning via rule
  ensembles. *Annals of Applied Statistics*, 2(3), 916--954.

- Goldstein, A., Kapelner, A., Bleich, J. and Pitkin, E. (2015). Peeking
  inside the black box. *JCGS*, 24(1), 44--65.

- Kim, B., Khanna, R. and Koyejo, O.O. (2016). Examples are not enough,
  learn to criticize! *NeurIPS*, 29.

- Lundberg, S.M. and Lee, S.-I. (2017). A unified approach to interpreting
  model predictions. *NeurIPS*, 30.

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  2nd ed. Springer.

- Ribeiro, M.T., Singh, S. and Guestrin, C. (2016). "Why should I trust
  you?": Explaining the predictions of any classifier. *KDD*.

- Ribeiro, M.T., Singh, S. and Guestrin, C. (2018). Anchors: High-precision
  model-agnostic explanations. *AAAI*.

- Sobol, I.M. (1993). Sensitivity estimates for nonlinear mathematical
  models. *Mathematical Modelling and Computational Experiments*, 1, 407--414.

## See Also

- `vignette("regression-diagnostics")` --- DFBETAS, permutation importance, stability
- `vignette("uncertainty-quantification")` --- prediction intervals, bootstrap CI
- `vignette("articles/scalar-on-function")` --- scalar-on-function regression methods
