---
title: "Functional Regression: Scalar and Function Responses"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Functional regression extends classical regression to settings where predictors
or responses are functions. **fdars** provides two complementary directions:

- **Scalar-on-function regression**: scalar response, functional predictors
  (`fregre.lm()`, `functional.logistic()`)
- **Function-on-scalar regression**: functional response, scalar predictors
  (`fosr()`, `fosr.fpc()`, `fanova()`)

This article demonstrates both, including cross-validation and hypothesis testing.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Scalar-on-Function Regression

### FPC-Based Functional Linear Model

The functional linear model relates a scalar response $y_i$ to a functional
predictor $X_i(t)$ via:

$$y_i = \alpha + \int X_i(t) \beta(t) \, dt + \varepsilon_i$$

`fregre.lm()` estimates $\beta(t)$ by projecting onto functional principal
components.

```{r fregre-lm}
set.seed(42)
n <- 80
m <- 100
t_grid <- seq(0, 1, length.out = m)

# Simulate functional predictor
X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) * runif(1, 0.5, 2) +
    cos(4 * pi * t_grid) * rnorm(1, sd = 0.3) +
    rnorm(m, sd = 0.1)
}

# Scalar response depends on the first FPC direction
beta_true <- sin(2 * pi * t_grid)
y <- X %*% beta_true / m + rnorm(n, sd = 0.3)

fd <- fdata(X, argvals = t_grid)
fit <- fregre.lm(fd, y, ncomp = 3)
print(fit)
```

### Cross-Validation for Component Selection

`fregre.lm.cv()` selects the optimal number of FPC components via k-fold
cross-validation.

```{r fregre-lm-cv}
cv_result <- fregre.lm.cv(fd, y, k.range = 1:8, nfold = 10)
cat("Optimal ncomp:", cv_result$optimal.k, "\n")
cat("CV errors:", round(cv_result$cv.errors, 4), "\n")
```

### Comparison with fregre.pc

`fregre.lm()` uses the Rust backend while `fregre.pc()` uses the same
FPC-based approach. Both produce similar fits:

```{r compare-pc}
fit_pc <- fregre.pc(fd, y, ncomp = 3)
cat("fregre.lm R-squared:", round(fit$r.squared, 4), "\n")
cat("fregre.pc residual MSE:", round(mean(fit_pc$residuals^2), 4), "\n")
cat("fregre.lm residual MSE:", round(mean(fit$residuals^2), 4), "\n")
```

### Functional Logistic Regression

For binary classification, `functional.logistic()` fits a logistic model
with FPC features.

```{r logistic}
# Create binary response (0/1)
y_bin <- as.numeric(y > median(y))

logit_fit <- functional.logistic(fd, y_bin, ncomp = 3)
print(logit_fit)

cat("Probabilities range:", range(logit_fit$probabilities), "\n")
```

## Function-on-Scalar Regression

### Penalized FOSR

When the response is a function and predictors are scalar, we have:

$$Y_i(t) = \mu(t) + \sum_j x_{ij} \beta_j(t) + \varepsilon_i(t)$$

`fosr()` estimates the coefficient functions $\beta_j(t)$ with optional
ridge penalty.

```{r fosr}
set.seed(42)
n <- 60
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Two scalar predictors: treatment and age
treatment <- rep(c(0, 1), each = n/2)
age <- rnorm(n)
predictors <- cbind(treatment, age)

# Functional response depends on predictors
Y <- matrix(0, n, m)
for (i in 1:n) {
  Y[i, ] <- sin(2 * pi * t_grid) +
    treatment[i] * 0.5 * cos(2 * pi * t_grid) +
    age[i] * 0.2 * t_grid +
    rnorm(m, sd = 0.15)
}

fd_y <- fdata(Y, argvals = t_grid)
fosr_fit <- fosr(fd_y, predictors, lambda = 0.1)
print(fosr_fit)
```

### Visualizing Coefficient Functions

```{r fosr-plot, fig.height = 4}
plot(fosr_fit)
```

### FPC-Based FOSR

`fosr.fpc()` uses functional principal components instead of penalization:

```{r fosr-fpc}
fosr_fpc_fit <- fosr.fpc(fd_y, predictors, ncomp = 5)
cat("FPC-based R-squared:", round(fosr_fpc_fit$r.squared, 4), "\n")
```

### Prediction

```{r fosr-predict}
new_pred <- matrix(c(1, 0.5,   # treated, average age
                     0, -1.0), # control, young
                   nrow = 2, byrow = TRUE)
pred <- predict(fosr_fit, new_pred)
```

## Functional ANOVA

`fanova()` tests whether mean functions differ across groups using a
permutation-based F-test.

```{r fanova}
set.seed(42)
n_per_group <- 25
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Three treatment groups with different mean functions
Y_anova <- matrix(0, 3 * n_per_group, m)
for (i in 1:n_per_group) {
  Y_anova[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.15)
  Y_anova[n_per_group + i, ] <- sin(2 * pi * t_grid) + 0.5 * cos(2 * pi * t_grid) +
    rnorm(m, sd = 0.15)
  Y_anova[2 * n_per_group + i, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.15)
}
groups <- rep(1:3, each = n_per_group)

fd_anova <- fdata(Y_anova, argvals = t_grid)
anova_result <- fanova(fd_anova, groups, n.perm = 500)
print(anova_result)
```

### Visualizing Group Means

```{r fanova-plot, fig.height = 4}
plot(anova_result)
```

## See Also

- `vignette("articles/regression")` for kernel and basis regression
- `vignette("articles/functional-classification")` for classification with
  `fclassif()`
- `vignette("articles/functional-mixed-models")` for repeated measures with
  `fmm()`
