---
title: "Supervised Classification of Functional Data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Supervised classification assigns functional observations to discrete groups.
**fdars** provides five methods through a unified `fclassif()` interface:

| Method | Description | Strengths |
|--------|-------------|-----------|
| **LDA** | Linear discriminant analysis on FPC scores | Fast, interpretable |
| **QDA** | Quadratic discriminant analysis on FPC scores | Handles class-specific covariance |
| **kNN** | k-nearest neighbors in FPC space | Nonparametric, flexible |
| **Kernel** | Kernel classifier with functional bandwidth | Fully nonparametric |
| **DD** | Depth-vs-depth classification | Robust to outliers |

All methods support optional scalar covariates alongside functional predictors.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Iris as Functional Data

Instead of a toy simulation we use the classic **iris** dataset, transformed
into functional data via the Andrews transformation. This maps each
4-dimensional observation to a curve on $[-\pi, \pi]$ using a Fourier
expansion --- preserving Euclidean distances up to a constant factor.

The three species (*setosa*, *versicolor*, *virginica*) are the classes.
Crucially, *versicolor* and *virginica* overlap substantially, making this a
non-trivial classification task.

```{r andrews-transform}
andrews_transform <- function(X, m = 200) {
  if (is.data.frame(X)) X <- as.matrix(X)
  n <- nrow(X)
  p <- ncol(X)
  t_grid <- seq(-pi, pi, length.out = m)

  basis <- matrix(0, m, p)
  basis[, 1] <- 1 / sqrt(2)
  for (j in 2:p) {
    k <- ceiling((j - 1) / 2)
    if (j %% 2 == 0) {
      basis[, j] <- sin(k * t_grid)
    } else {
      basis[, j] <- cos(k * t_grid)
    }
  }

  fdata(X %*% t(basis), argvals = t_grid)
}
```

```{r data}
X_raw <- as.matrix(iris[, 1:4])
X <- scale(X_raw)
species <- iris$Species
y <- as.integer(species)  # 1 = setosa, 2 = versicolor, 3 = virginica

fd <- andrews_transform(X)
```

```{r iris-plot, fig.height = 5}
plot(fd, color = species, alpha = 0.35, show.mean = TRUE,
     palette = c("setosa" = "#E69F00", "versicolor" = "#56B4E9",
                 "virginica" = "#009E73")) +
  labs(x = expression(t), y = expression(f[x](t)),
       title = "Iris Andrews Curves by Species")
```

Setosa is well separated, but versicolor and virginica overlap ---
exactly the kind of challenge where classifier choice matters.

## Comparing All Five Methods

```{r compare-methods}
methods <- c("lda", "qda", "knn", "kernel", "dd")
results <- list()

for (meth in methods) {
  if (meth == "knn") {
    results[[meth]] <- fclassif(fd, y, method = meth, ncomp = 3, k = 7)
  } else {
    results[[meth]] <- fclassif(fd, y, method = meth, ncomp = 3)
  }
}

acc_df <- data.frame(
  Method = toupper(methods),
  Training_Accuracy = sapply(results, function(r) round(r$accuracy, 3))
)
print(acc_df)
```

Training accuracy alone can be misleading --- QDA and kNN may overfit.
Cross-validation gives a fairer picture (see below).

## Confusion Matrix

The LDA confusion matrix reveals where errors concentrate:

```{r confusion, fig.height = 4}
plot(results[["lda"]])
```

Most errors are between versicolor and virginica, as expected from the
overlapping Andrews curves.

## Cross-Validation

`fclassif.cv()` evaluates out-of-sample error via k-fold cross-validation:

```{r cv}
set.seed(42)
cv_results <- list()
for (meth in methods) {
  cv_results[[meth]] <- fclassif.cv(fd, y, method = meth,
                                     ncomp = 3, nfold = 10, seed = 42)
}

cv_df <- data.frame(
  Method = toupper(methods),
  CV_Error = sapply(cv_results, function(r) round(r$error.rate, 3)),
  CV_Accuracy = sapply(cv_results, function(r) round(1 - r$error.rate, 3))
)
print(cv_df)
```

## Choosing the Number of Components

Cross-validation can also help select the optimal number of FPC components.
Too few loses discriminative information; too many introduces noise.

```{r ncomp-selection}
ncomp_range <- 1:6
cv_by_ncomp <- sapply(ncomp_range, function(k) {
  fclassif.cv(fd, y, method = "lda", ncomp = k, nfold = 10, seed = 42)$error.rate
})

df_ncomp <- data.frame(ncomp = ncomp_range, error = cv_by_ncomp)
ggplot(df_ncomp, aes(x = ncomp, y = error)) +
  geom_line() +
  geom_point(size = 2) +
  labs(x = "Number of FPC Components", y = "10-Fold CV Error Rate",
       title = "LDA on Iris: Component Selection") +
  scale_x_continuous(breaks = ncomp_range)
```

## Method Selection Guidelines

- **LDA**: Good default when classes have similar covariance. Fast and
  interpretable. Works well on iris.
- **QDA**: Better when each class has its own covariance pattern. Needs
  enough observations per class to estimate class-specific covariances.
- **kNN**: Flexible nonparametric option. Tune $k$ via CV. Robust to
  non-linear boundaries.
- **Kernel**: Fully nonparametric --- works directly in function space
  without dimension reduction. Can be slow on large datasets.
- **DD**: Robust to outliers. Uses depth-vs-depth plots for classification.
  Less sensitive to distributional assumptions.

## See Also

- `vignette("articles/andrews-transformation")` for more on the Andrews
  transform and its distance-preservation property
- `vignette("articles/functional-regression")` for scalar-on-function
  regression including `functional.logistic()`
- `vignette("articles/clustering")` for unsupervised functional clustering
- `vignette("articles/fpca")` for functional PCA, which underlies LDA/QDA/kNN
