---
title: "Functional Mixed Models for Repeated Measures"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

In many studies, each subject is observed multiple times, producing repeated
functional measurements. A **functional mixed model** (FMM) decomposes the
variation into:

$$Y_{ij}(t) = \mu(t) + \sum_k x_{ij,k} \beta_k(t) + b_i(t) + \varepsilon_{ij}(t)$$

where:

- $\mu(t)$ is the population mean function
- $\beta_k(t)$ are fixed-effect coefficient functions (e.g., treatment effects)
- $b_i(t)$ are subject-level random effects
- $\varepsilon_{ij}(t)$ is measurement error

**fdars** provides `fmm()` for fitting, `fmm.predict()` for prediction, and
`fmm.test.fixed()` for testing fixed effects via permutation.

```{=html}
<img src="../reference/figures/functional-mixed-models-diagram.svg" alt="Functional Mixed Model Decomposition" style="width:100%; max-width:700px;">
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Why Mixed Models for Functional Data?

When subjects contribute multiple curves, the observations are no longer
independent. Ignoring this structure has serious consequences:

1. **Pseudoreplication.** Treating all $n_{\text{total}}$ curves as independent
   inflates the effective sample size. Standard errors are too small, p-values
   are too optimistic, and Type I error rates exceed their nominal level.

2. **Confounding sources of variation.** Without subject-level random effects,
   between-subject variability is absorbed into the residual, blurring the
   distinction between "subjects differ" and "measurements are noisy."

3. **Inefficient estimation.** The within-subject correlation carries
   information about fixed effects. Mixed models exploit this correlation
   structure, producing more efficient (lower-variance) estimates than methods
   that ignore it.

The functional mixed model resolves these issues by decomposing each curve into
four additive components:

| Component | Notation | Interpretation |
|-----------|----------|----------------|
| Population mean | $\mu(t)$ | Average curve across all subjects and conditions |
| Fixed effects | $\sum_k x_{ij,k}\beta_k(t)$ | Systematic effects of covariates (e.g., treatment) |
| Random effects | $b_i(t)$ | Subject-specific deviations from the population mean |
| Residual | $\varepsilon_{ij}(t)$ | Within-subject measurement noise |

This decomposition is the functional analog of the classical scalar linear mixed
model $y_{ij} = \mu + x_{ij}^T\beta + b_i + \varepsilon_{ij}$, with each
scalar parameter replaced by a function of $t$.

## Mathematical Framework

### Model Specification

The full model for the $j$-th observation from subject $i$ is:

$$Y_{ij}(t) = \mu(t) + \sum_{p=1}^P \beta_p(t)\,x_{ip} + b_i(t) + \varepsilon_{ij}(t)$$

with distributional assumptions:

- $b_i(t) \sim \text{GP}(0, \sigma^2_u)$: subject-specific random effects,
  independent across subjects
- $\varepsilon_{ij}(t) \sim \text{GP}(0, \sigma^2_\varepsilon)$: residual noise,
  independent across observations

Here GP denotes a Gaussian process. The key parameters are the two variance
components $\sigma^2_u$ (between-subject) and $\sigma^2_\varepsilon$
(within-subject).

### FPC-Based Estimation

Direct estimation of the functional parameters $\mu(t), \beta_p(t), b_i(t)$
is infinite-dimensional. The FPC approach reduces this to a set of scalar
mixed models.

**Step 1.** Compute the FPC decomposition of the pooled data:

$$Y_{ij}(t) - \bar{Y}(t) \approx \sum_{k=1}^K \xi_{ij,k}\,\phi_k(t)$$

where $\phi_k(t)$ are the eigenfunctions and $\xi_{ij,k}$ are the scores.

**Step 2.** For each component $k = 1, \ldots, K$, fit a scalar mixed model:

$$\xi_{ij,k} = \gamma_{0k} + \sum_{p=1}^P \gamma_{pk}\,x_{ip} + u_{ik} + e_{ijk}$$

where $u_{ik} \sim \mathcal{N}(0, \sigma^2_{u,k})$ are subject-level random
effects and $e_{ijk} \sim \mathcal{N}(0, \sigma^2_{\varepsilon,k})$ are
residuals.

**Step 3.** Reconstruct the functional parameters:

$$\hat\mu(t) = \bar{Y}(t) + \sum_k \hat\gamma_{0k}\,\phi_k(t), \qquad \hat\beta_p(t) = \sum_k \hat\gamma_{pk}\,\phi_k(t)$$

$$\hat{b}_i(t) = \sum_k \hat{u}_{ik}\,\phi_k(t)$$

This approach is elegant: the infinite-dimensional functional mixed model
decomposes into $K$ independent scalar mixed models, each of which can be
fitted using standard REML estimation.

### REML Estimation

Within each FPC component, the variance components are estimated via
**Restricted Maximum Likelihood (REML)**, which accounts for the loss of
degrees of freedom from estimating fixed effects.

The key quantities are:

**Shrinkage weights:**

$$w_i = \frac{\sigma^2_u}{\sigma^2_u + \sigma^2_\varepsilon / n_i}$$

where $n_i$ is the number of observations for subject $i$. When $n_i$ is large
or $\sigma^2_u \gg \sigma^2_\varepsilon$, $w_i \approx 1$ and the random
effect estimate is close to the raw subject mean. When $n_i$ is small or
$\sigma^2_u \ll \sigma^2_\varepsilon$, $w_i \approx 0$ and the estimate is
shrunk toward zero (the population mean).

**BLUP (Best Linear Unbiased Predictor) for random effects:**

$$\hat{u}_i = w_i\bigl(\bar{\xi}_i - \bar{x}_i^T\hat\gamma\bigr)$$

The BLUP "borrows strength" across subjects: subjects with few observations
are pulled toward the population mean, while subjects with many observations
retain their individual character. This is the functional analog of the
well-known shrinkage property of scalar mixed models.

**REML divisor:** The variance estimate uses $n - p$ in the denominator (where
$p$ is the number of fixed-effect parameters) rather than $n$, correcting for
the bias that would otherwise arise from estimating fixed effects and variance
components simultaneously.

### Hypothesis Testing for Fixed Effects

Testing whether a covariate has a significant functional effect:

$$H_0: \beta_p(t) = 0 \text{ for all } t$$

is analogous to functional ANOVA. The test uses a permutation approach that
respects the within-subject correlation:

1. Compute the observed F-statistic $F_{\text{obs}}$ from the fitted model.
2. For $b = 1, \ldots, B$: permute subject-level labels (not individual
   observations) and refit the model to obtain $F_b^*$.
3. The p-value is:

$$p = \frac{1 + \#\{F_b^* \geq F_{\text{obs}}\}}{B + 1}$$

Permuting at the subject level (rather than the observation level) preserves the
within-subject correlation structure under the null hypothesis, ensuring valid
inference even with complex dependency patterns.

## Simulated Repeated Measures Data

We simulate a clinical trial with 15 subjects, each measured at 3 time points,
with a binary treatment covariate.

```{r data}
set.seed(42)
n_subjects <- 15
n_visits <- 3
n_total <- n_subjects * n_visits
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Subject-specific random effects
subject_effects <- matrix(0, n_subjects, m)
for (i in 1:n_subjects) {
  subject_effects[i, ] <- 0.3 * rnorm(1) * sin(2 * pi * t_grid) +
    0.2 * rnorm(1) * cos(4 * pi * t_grid)
}

# Treatment assignment (first 8 subjects treated, rest control)
treatment <- c(rep(1, 8), rep(0, 7))

# Generate observations
Y <- matrix(0, n_total, m)
subject_ids <- rep(1:n_subjects, each = n_visits)
covariates <- matrix(0, n_total, 1)

for (i in 1:n_total) {
  subj <- subject_ids[i]
  trt <- treatment[subj]
  covariates[i, 1] <- trt

  Y[i, ] <- sin(2 * pi * t_grid) +                         # Mean function
    trt * 0.5 * cos(2 * pi * t_grid) +                      # Treatment effect
    subject_effects[subj, ] +                                 # Random effect
    rnorm(m, sd = 0.1)                                        # Noise
}

fd <- fdata(Y, argvals = t_grid)
```

The simulation has three variance sources: the treatment effect
($0.5\cos(2\pi t)$ for treated subjects), subject-level random effects (smooth
random curves with $\sigma \approx 0.3$), and measurement noise ($\sigma = 0.1$).
The model should recover these components.

## Fitting the Functional Mixed Model

```{r fmm-fit}
fit <- fmm(fd, subject_ids, covariates = covariates, ncomp = 3)
print(fit)
```

The output shows the estimated variance components and the number of FPC
components used. The ratio $\sigma^2_u / \sigma^2_\varepsilon$ (the
intraclass correlation in functional form) indicates how much of the residual
variation is between-subject vs. within-subject.

## Visualizing Fixed and Random Effects

The plot shows the population mean (red) overlaid with subject-specific
random effects (blue):

```{r fmm-plot}
plot(fit)
```

Subject-specific curves that deviate substantially from the population mean
indicate large random effects for those subjects. The spread of the blue curves
reflects $\sigma^2_u$: wider spread means more between-subject variability.

## Examining Model Components

```{r components}
# Residual variance
cat("Residual variance:", round(fit$sigma2.eps, 4), "\n")

# Number of subjects
cat("Number of subjects:", fit$n.subjects, "\n")

# FPC components used for random effects
cat("FPC components:", fit$ncomp, "\n")
```

The residual variance $\hat\sigma^2_\varepsilon$ should be close to
$0.1^2 = 0.01$ (the simulation noise level). A much larger value would
suggest the model is not capturing all systematic variation.

## Population-Level Prediction

`fmm.predict()` generates predictions using fixed effects only (no random
effects), useful for predicting the population-level response for new
covariate values.

```{r predict}
# Predict for treated and control groups
new_cov <- matrix(c(1, 0), nrow = 2)
pred <- fmm.predict(fit, new.covariates = new_cov)

# Plot predictions
df_pred <- data.frame(
  t = rep(t_grid, 2),
  value = as.vector(t(pred$data)),
  group = factor(rep(c("Treated", "Control"), each = m))
)

ggplot(df_pred, aes(x = t, y = value, color = group)) +
  geom_line(linewidth = 1) +
  labs(x = "t", y = "Y(t)",
       title = "Predicted Population-Level Responses",
       color = "Group")
```

The gap between the treated and control curves is the estimated treatment
effect $\hat\beta_1(t)$. Where the curves diverge most is where the treatment
has its strongest functional effect.

## Testing Fixed Effects

`fmm.test.fixed()` tests whether the treatment has a significant effect
using a permutation F-test that respects within-subject correlation:

```{r test}
test_result <- fmm.test.fixed(fd, subject_ids, covariates,
                               ncomp = 3, n.perm = 500, seed = 123)
print(test_result)
```

A small p-value indicates that the treatment effect $\beta_1(t)$ is
significantly different from zero. Because the permutation is done at the
subject level, this test correctly accounts for the repeated measures structure
--- unlike a naive test that would treat all 45 curves as independent.

## Comparison: Ignoring Subject Structure

What happens if we ignore the repeated measures structure and treat
all observations as independent?

```{r naive-comparison}
# Naive approach: function-on-scalar regression ignoring subjects
fosr_naive <- fosr(fd, covariates)
cat("Naive FOSR R-squared:", round(fosr_naive$r.squared, 4), "\n")

# FMM accounts for subject-level variation
cat("FMM residual variance:", round(fit$sigma2.eps, 4), "\n")
```

The FMM separates subject-level variation from residual noise, giving a
cleaner estimate of the treatment effect. The naive FOSR lumps both sources
together, producing an inflated residual and potentially misleading inference
(standard errors that are too small because the effective sample size is
overstated).

## Interpretation and Diagnostics

### Variance Ratio

The ratio $\hat\sigma^2_u / (\hat\sigma^2_u + \hat\sigma^2_\varepsilon)$ is
the **intraclass correlation coefficient (ICC)**. It describes the proportion
of non-fixed-effect variability attributable to subject differences:

- **ICC near 1**: subjects are very different from each other; within-subject
  replicates are nearly identical. Mixed models are essential.
- **ICC near 0**: subjects are similar; most variation is within-subject noise.
  A simpler model (ignoring subjects) may suffice.

### Shrinkage Toward the Population Mean

Random effects are shrunk toward zero by the factor $w_i$. Subjects with fewer
observations are shrunk more heavily. This is desirable --- it prevents
overfitting to individual subjects with sparse data --- but can be misleading if
interpreted as "this subject truly has a small effect." The amount of shrinkage
is itself informative: heavy shrinkage indicates that the data for that subject
are too noisy to estimate a reliable individual effect.

### When to Suspect Model Misspecification

- **Systematic residual patterns**: if residuals show temporal structure (e.g.,
  autocorrelation), the residual process $\varepsilon_{ij}(t)$ may not be white
  noise. Consider adding more FPC components or modeling serial correlation.
- **Non-constant variance**: if residual variance changes across the domain,
  a heteroscedastic model may be needed.
- **Random effects not Gaussian**: the REML estimator is still consistent under
  non-Gaussianity, but prediction intervals for random effects may be poorly
  calibrated.

## References

- Morris, J.S. (2015). Functional regression. *Annual Review of Statistics and Its Application*, 2, 321--359.
- Di, C.Z., Crainiceanu, C.M., Caffo, B.S., and Punjabi, N.M. (2009). Multilevel functional principal component analysis. *Annals of Applied Statistics*, 3(1), 458--488.
- Greven, S., Crainiceanu, C., Caffo, B., and Reich, D. (2010). Longitudinal functional principal component analysis. *Electronic Journal of Statistics*, 4, 1022--1054.
- Crainiceanu, C.M., Staicu, A.M., and Di, C.Z. (2009). Generalized multilevel functional regression. *Journal of the American Statistical Association*, 104(488), 1550--1561.

## See Also

- `vignette("articles/scalar-on-function")` for scalar-on-function regression
- `vignette("articles/function-on-scalar")` for function-on-scalar regression
  (FOSR, FANOVA)
- `vignette("articles/fpca")` for the FPC decomposition underlying the
  random effects
