---
title: "Functional Regression: Scalar and Function Responses"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Functional regression extends classical regression to settings where predictors
or responses are functions. **fdars** provides two complementary directions:

- **Scalar-on-function regression**: scalar response, functional predictors
  (`fregre.lm()`, `functional.logistic()`)
- **Function-on-scalar regression**: functional response, scalar predictors
  (`fosr()`, `fosr.fpc()`, `fanova()`)

This article demonstrates both, including cross-validation and hypothesis testing.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Why Functional Regression?

Classical regression assumes that predictors live in $\mathbb{R}^p$ with $p$
finite and fixed. When the predictor is a curve $X_i(t)$ observed on a
continuum, we face a fundamentally different problem: the regression coefficient
$\beta(t)$ is itself a function, and the "design matrix" is infinite-dimensional.

This creates two immediate challenges:

1. **Ill-posedness.** The functional linear model
   $y_i = \alpha + \int X_i(t)\beta(t)\,dt + \varepsilon_i$ has infinitely many
   solutions because the integral operator is compact. Without regularization,
   the least-squares estimate of $\beta(t)$ does not exist or is wildly unstable.

2. **Curse of dimensionality.** Naively discretizing $X_i(t)$ at $m$ grid
   points and running OLS gives $m$ parameters from $n$ observations. When
   $m \gg n$ (the typical FDA setting), the model is massively overparameterized.

Two strategies resolve these issues:

- **Dimension reduction**: project $X_i(t)$ onto its leading functional
  principal components (FPCs) and regress on the scores. This is the approach
  taken by `fregre.lm()` and `fregre.pc()`.
- **Penalization**: estimate $\beta(t)$ directly but penalize its roughness via
  a second-derivative penalty. This is the approach taken by `fosr()` for the
  function-on-scalar direction.

Both approaches replace the ill-posed infinite-dimensional problem with a
well-conditioned finite-dimensional one.

## Scalar-on-Function Regression

### Mathematical Framework

The **functional linear model** relates a scalar response $y_i$ to a functional
predictor $X_i(t)$ via:

$$y_i = \alpha + \int_0^1 X_i(t)\,\beta(t)\,dt + \varepsilon_i, \quad \varepsilon_i \overset{\text{iid}}{\sim} (0, \sigma^2)$$

The coefficient function $\beta(t)$ describes *how* the predictor curve at time
$t$ influences the response. A positive $\beta(t)$ in some interval means that
curves with higher values there tend to have higher $y_i$.

#### FPC-Based Estimation

Let $\phi_1(t), \phi_2(t), \ldots$ be the eigenfunctions of the covariance
operator of $X$, and let $\xi_{ik} = \int (X_i(t) - \bar{X}(t))\phi_k(t)\,dt$
be the FPC scores. Expanding $\beta(t) = \sum_{k=1}^\infty \gamma_k \phi_k(t)$
and substituting gives:

$$y_i \approx \alpha + \sum_{k=1}^K \gamma_k \xi_{ik} + \varepsilon_i$$

This reduces the functional regression to an ordinary multiple regression on
$K$ FPC scores. The coefficient function is recovered as:

$$\hat\beta(t) = \sum_{k=1}^K \hat\gamma_k \phi_k(t)$$

The number of components $K$ controls the bias--variance trade-off: too few
components oversimplify $\beta(t)$; too many introduce noise. Cross-validation
selects $K$ in a data-driven way.

#### Standard Errors and GCV

With the FPC scores as predictors, standard OLS inference applies. The variance
of $\hat\gamma$ is $\sigma^2 (\Xi^T\Xi)^{-1}$ where $\Xi$ is the $n \times K$
score matrix. Generalized cross-validation provides a computationally efficient
alternative to leave-one-out CV:

$$\text{GCV} = \frac{1}{n}\sum_{i=1}^n \left(\frac{y_i - \hat{y}_i}{1 - h_{ii}}\right)^2$$

where $h_{ii}$ are the diagonal elements of the hat matrix
$H = \Xi(\Xi^T\Xi)^{-1}\Xi^T$.

### FPC-Based Functional Linear Model

`fregre.lm()` estimates $\beta(t)$ by projecting onto functional principal
components.

```{r fregre-lm}
set.seed(42)
n <- 80
m <- 100
t_grid <- seq(0, 1, length.out = m)

# Simulate functional predictor
X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) * runif(1, 0.5, 2) +
    cos(4 * pi * t_grid) * rnorm(1, sd = 0.3) +
    rnorm(m, sd = 0.1)
}

# Scalar response depends on the first FPC direction
beta_true <- sin(2 * pi * t_grid)
y <- X %*% beta_true / m + rnorm(n, sd = 0.3)

fd <- fdata(X, argvals = t_grid)
fit <- fregre.lm(fd, y, ncomp = 3)
print(fit)
```

The R-squared tells us how much scalar variance is explained by the leading
FPC directions of the predictor curves. The estimated $\hat\beta(t)$ should
approximate the true $\beta(t) = \sin(2\pi t)$ used in the simulation.

### Cross-Validation for Component Selection

`fregre.lm.cv()` selects the optimal number of FPC components via k-fold
cross-validation. This is the data-driven way to choose $K$, balancing
approximation accuracy against overfitting.

```{r fregre-lm-cv}
cv_result <- fregre.lm.cv(fd, y, k.range = 1:8, nfold = 10)
cat("Optimal ncomp:", cv_result$optimal.k, "\n")
cat("CV errors:", round(cv_result$cv.errors, 4), "\n")
```

The CV error typically decreases as components are added (capturing more signal),
then plateaus or increases (adding noise). The optimal $K$ sits at the elbow.

### Comparison with fregre.pc

`fregre.lm()` uses the Rust backend while `fregre.pc()` uses the same
FPC-based approach. Both produce similar fits:

```{r compare-pc}
fit_pc <- fregre.pc(fd, y, ncomp = 3)
cat("fregre.lm R-squared:", round(fit$r.squared, 4), "\n")
cat("fregre.pc residual MSE:", round(mean(fit_pc$residuals^2), 4), "\n")
cat("fregre.lm residual MSE:", round(mean(fit$residuals^2), 4), "\n")
```

### Functional Logistic Regression

For binary classification, `functional.logistic()` fits a logistic model
with FPC features.

#### The IRLS Algorithm

The functional logistic model uses the logit link on FPC scores:

$$\log\frac{P(Y_i = 1)}{1 - P(Y_i = 1)} = \alpha + \sum_{k=1}^K \gamma_k \xi_{ik}$$

This is estimated via **Iteratively Reweighted Least Squares (IRLS)**, which
alternates between:

1. **Working response**: $z_i = \hat\eta_i + (y_i - \hat{p}_i) / [\hat{p}_i(1 - \hat{p}_i)]$
2. **Working weights**: $w_i = \hat{p}_i(1 - \hat{p}_i)$
3. **Weighted least squares**: update $\hat\gamma$ by regressing $z$ on $\Xi$ with weights $w$

Convergence is monitored via the log-likelihood:

$$\ell(\gamma) = \sum_{i=1}^n \bigl[y_i \log p_i + (1 - y_i)\log(1 - p_i)\bigr]$$

where $p_i = \text{logit}^{-1}(\alpha + \xi_i^T\gamma)$.

```{r logistic}
# Create binary response (0/1)
y_bin <- as.numeric(y > median(y))

logit_fit <- functional.logistic(fd, y_bin, ncomp = 3)
print(logit_fit)

cat("Probabilities range:", range(logit_fit$probabilities), "\n")
```

The estimated probabilities near 0 or 1 indicate confident classification;
values near 0.5 indicate curves in the overlap region where the predictor's
FPC scores do not clearly separate the two classes.

## Function-on-Scalar Regression

### Mathematical Framework

When the *response* is a function and the predictors are scalar, the model is:

$$Y_i(t) = \mu(t) + \sum_{j=1}^p x_{ij}\,\beta_j(t) + \varepsilon_i(t)$$

Each coefficient $\beta_j(t)$ is a function describing how predictor $j$
modifies the response curve at each time point $t$. The intercept $\mu(t)$
is the mean response when all predictors are zero.

#### Penalized Estimation

Direct pointwise OLS at each $t$ would produce coefficient estimates that are
noisy and discontinuous across the time domain. `fosr()` imposes smoothness
via a second-difference roughness penalty. The penalized objective is:

$$\sum_{i=1}^n \|Y_i - \mu - X_i\beta\|^2 + \lambda \sum_{j=1}^p \int [\beta_j''(t)]^2\,dt$$

where $\lambda > 0$ controls the trade-off between fidelity to the data and
smoothness. In matrix form, the penalty uses the second-difference matrix $D$:

$$\hat\beta = \arg\min_\beta \|Y - X\beta\|_F^2 + \lambda\,\text{tr}(\beta^T D^T D\,\beta)$$

The solution at each grid point is a ridge-type estimator:

$$\hat\beta(\cdot) = (X^TX + \lambda D^TD)^{-1} X^T Y(\cdot)$$

#### Pointwise R-squared

The goodness of fit is measured by the **pointwise R-squared**:

$$R^2(t) = 1 - \frac{\sum_i [Y_i(t) - \hat{Y}_i(t)]^2}{\sum_i [Y_i(t) - \bar{Y}(t)]^2}$$

The overall (mean) R-squared averages this across the domain. Values near 1
everywhere indicate the scalar predictors explain the functional response well
at all time points; regions with low $R^2(t)$ suggest additional predictors or
nonlinear effects may be needed.

### Penalized FOSR

`fosr()` estimates the coefficient functions $\beta_j(t)$ with optional
ridge penalty.

```{r fosr}
set.seed(42)
n <- 60
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Two scalar predictors: treatment and age
treatment <- rep(c(0, 1), each = n/2)
age <- rnorm(n)
predictors <- cbind(treatment, age)

# Functional response depends on predictors
Y <- matrix(0, n, m)
for (i in 1:n) {
  Y[i, ] <- sin(2 * pi * t_grid) +
    treatment[i] * 0.5 * cos(2 * pi * t_grid) +
    age[i] * 0.2 * t_grid +
    rnorm(m, sd = 0.15)
}

fd_y <- fdata(Y, argvals = t_grid)
fosr_fit <- fosr(fd_y, predictors, lambda = 0.1)
print(fosr_fit)
```

The treatment coefficient $\hat\beta_1(t)$ should recover the simulated
$0.5\cos(2\pi t)$ pattern, while the age coefficient $\hat\beta_2(t)$ should
approximate the linear $0.2t$ trend.

### Visualizing Coefficient Functions

```{r fosr-plot, fig.height = 4}
plot(fosr_fit)
```

Each panel shows one coefficient function $\hat\beta_j(t)$. The shape reveals
*when* during the functional domain each predictor has its strongest effect.

### FPC-Based FOSR

`fosr.fpc()` uses functional principal components instead of penalization:

```{r fosr-fpc}
fosr_fpc_fit <- fosr.fpc(fd_y, predictors, ncomp = 5)
cat("FPC-based R-squared:", round(fosr_fpc_fit$r.squared, 4), "\n")
```

The FPC-based approach avoids choosing $\lambda$ but requires selecting the
number of components. When the response curves have strong low-rank structure
(most variance in a few FPCs), this approach is efficient and stable.

### Prediction

```{r fosr-predict}
new_pred <- matrix(c(1, 0.5,   # treated, average age
                     0, -1.0), # control, young
                   nrow = 2, byrow = TRUE)
pred <- predict(fosr_fit, new_pred)
```

## Functional ANOVA

### Mathematical Framework

Functional ANOVA tests whether the mean functions of $G$ groups are equal:

$$H_0: \mu_1(t) = \mu_2(t) = \cdots = \mu_G(t) \quad \text{for all } t$$

At each grid point $t$, a pointwise F-statistic is computed:

$$F(t) = \frac{\sum_{g=1}^G n_g [\bar{Y}_g(t) - \bar{Y}(t)]^2 / (G - 1)}{\sum_{g=1}^G \sum_{i \in g} [Y_i(t) - \bar{Y}_g(t)]^2 / (N - G)}$$

These are summarized into a **global test statistic** by integration:

$$F_{\text{global}} = \int F(t)\,dt$$

#### Permutation Testing

Because the null distribution of $F_{\text{global}}$ is intractable, `fanova()`
uses a permutation approach:

1. Compute the observed $F_{\text{obs}}$ from the original data.
2. For $b = 1, \ldots, B$: randomly permute group labels and compute $F_b^*$.
3. The p-value is:

$$p = \frac{1 + \#\{F_b^* \geq F_{\text{obs}}\}}{B + 1}$$

The $+1$ in numerator and denominator ensures the p-value is never exactly zero
and accounts for the observed statistic being one of the $B + 1$ values under
consideration. This makes the test exact (valid at any sample size) and
assumption-free regarding the distribution of $\varepsilon_i(t)$.

### Permutation F-Test

`fanova()` tests whether mean functions differ across groups using a
permutation-based F-test.

```{r fanova}
set.seed(42)
n_per_group <- 25
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Three treatment groups with different mean functions
Y_anova <- matrix(0, 3 * n_per_group, m)
for (i in 1:n_per_group) {
  Y_anova[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.15)
  Y_anova[n_per_group + i, ] <- sin(2 * pi * t_grid) + 0.5 * cos(2 * pi * t_grid) +
    rnorm(m, sd = 0.15)
  Y_anova[2 * n_per_group + i, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.15)
}
groups <- rep(1:3, each = n_per_group)

fd_anova <- fdata(Y_anova, argvals = t_grid)
anova_result <- fanova(fd_anova, groups, n.perm = 500)
print(anova_result)
```

A small p-value rejects the null hypothesis that all groups share the same mean
function. In this simulation, the three groups have clearly different shapes
(sine, sine + cosine, linear), so we expect strong rejection.

### Visualizing Group Means

```{r fanova-plot, fig.height = 4}
plot(anova_result)
```

The group mean curves show the estimated $\hat\mu_g(t)$ for each group. Where
the curves separate most, the pointwise F-statistic is largest.

## Assumptions and Diagnostics

### When FPC-Based Regression Works Well

- **Low intrinsic dimensionality**: most predictor variability is captured by
  a few leading FPCs. This is common when curves are smooth.
- **Linear relationship**: the link between FPC scores and response is
  approximately linear. For nonlinear relationships, consider kernel regression
  (see `vignette("articles/regression")`).
- **Adequate sample size**: $n$ should be substantially larger than $K$ to
  avoid overfitting.

### When to Be Cautious

- **Local effects**: if $\beta(t)$ is nonzero only on a small interval, the
  FPC basis (which captures *global* variance modes) may need many components
  to represent it, defeating the purpose of dimension reduction.
- **Non-Gaussian errors**: the permutation tests in `fanova()` and
  `fmm.test.fixed()` are distribution-free, but the confidence intervals from
  `fregre.lm()` assume normality.
- **Phase variability**: if curves are misaligned, FPCs mix amplitude and phase
  variation. Consider elastic alignment first
  (see `vignette("articles/elastic-alignment")`).

## References

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd ed. Springer.
- Reiss, P.T. and Ogden, R.T. (2007). Functional Principal Component Regression and Functional Partial Least Squares. *Journal of the American Statistical Association*, 102(479), 984--996.
- Ferraty, F. and Vieu, P. (2006). *Nonparametric Functional Data Analysis: Theory and Practice*. Springer.
- Cuevas, A., Febrero, M., and Fraiman, R. (2004). An anova test for functional data. *Computational Statistics & Data Analysis*, 47(1), 111--122.

## See Also

- `vignette("articles/regression")` for kernel and basis regression
- `vignette("articles/functional-classification")` for classification with
  `fclassif()`
- `vignette("articles/functional-mixed-models")` for repeated measures with
  `fmm()`
