---
title: "Supervised Classification of Functional Data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Supervised classification assigns functional observations to discrete groups.
**fdars** provides five methods through a unified `fclassif()` interface:

| Method | Description | Strengths |
|--------|-------------|-----------|
| **LDA** | Linear discriminant analysis on FPC scores | Fast, interpretable |
| **QDA** | Quadratic discriminant analysis on FPC scores | Handles class-specific covariance |
| **kNN** | k-nearest neighbors in FPC space | Nonparametric, flexible |
| **Kernel** | Kernel classifier with functional bandwidth | Fully nonparametric |
| **DD** | Depth-vs-depth classification | Robust to outliers |

All methods support optional scalar covariates alongside functional predictors.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Why Functional Classification?

The fundamental problem is: given a curve $X_i(t)$, predict a class label
$g_i \in \{1, \ldots, G\}$. One might be tempted to discretize each curve at
$m$ grid points and apply a standard multivariate classifier. However, this
approach faces two obstacles:

1. **Curse of dimensionality.** With $m$ grid points (often 100--1000),
   multivariate classifiers struggle unless $n \gg m$. Covariance matrices
   become singular, nearest-neighbor distances concentrate, and decision
   boundaries overfit.

2. **Ignoring smoothness.** Discretization treats $X_i(t_1)$ and $X_i(t_2)$
   as unrelated features, discarding the continuity and smoothness of the
   underlying curves.

Functional classification resolves this by first projecting curves onto a
low-dimensional basis --- typically the leading FPC directions --- and then
classifying in that reduced space. This simultaneously regularizes the problem
(since $K \ll m$) and respects the functional structure (since FPCs capture the
dominant modes of variation).

For methods that work directly in function space (kernel classifier, DD
classifier), distances are computed using proper functional metrics ($L^2$ norm
or depth), avoiding discretization artifacts entirely.

## Mathematical Framework

All FPC-based methods (LDA, QDA, kNN) start by computing the scores
$\boldsymbol\xi_i = (\xi_{i1}, \ldots, \xi_{iK})^T$ where
$\xi_{ik} = \int [X_i(t) - \bar{X}(t)]\phi_k(t)\,dt$ and $\phi_k$ are the
eigenfunctions of the pooled covariance operator. Classification then operates
in $\mathbb{R}^K$.

### Linear Discriminant Analysis (LDA)

LDA assumes that the FPC scores within each class are Gaussian with a common
covariance:

$$\boldsymbol\xi \mid g = k \;\sim\; \mathcal{N}(\boldsymbol\mu_k,\, \boldsymbol\Sigma_{\text{pool}})$$

The **pooled covariance** is:

$$\hat{\boldsymbol\Sigma}_{\text{pool}} = \frac{1}{n - G}\sum_{g=1}^G \sum_{i \in g}(\boldsymbol\xi_i - \bar{\boldsymbol\xi}_g)(\boldsymbol\xi_i - \bar{\boldsymbol\xi}_g)^T$$

Classification uses the **log-posterior discriminant**:

$$\delta_g(\boldsymbol\xi) = \log\pi_g - \tfrac{1}{2}(\boldsymbol\xi - \boldsymbol\mu_g)^T \boldsymbol\Sigma_{\text{pool}}^{-1}(\boldsymbol\xi - \boldsymbol\mu_g)$$

where $\pi_g$ is the prior probability of class $g$ (estimated by class
proportions). The observation is assigned to $\hat{g} = \arg\max_g \delta_g(\boldsymbol\xi)$.
Because $\boldsymbol\Sigma_{\text{pool}}$ is shared, the decision boundaries are linear
(hyperplanes) in the FPC score space.

### Quadratic Discriminant Analysis (QDA)

QDA relaxes the equal-covariance assumption by estimating a **class-specific
covariance** $\boldsymbol\Sigma_g$ for each group. The discriminant becomes:

$$\delta_g(\boldsymbol\xi) = \log\pi_g - \tfrac{1}{2}\log|\boldsymbol\Sigma_g| - \tfrac{1}{2}(\boldsymbol\xi - \boldsymbol\mu_g)^T\boldsymbol\Sigma_g^{-1}(\boldsymbol\xi - \boldsymbol\mu_g)$$

The additional $\log|\boldsymbol\Sigma_g|$ term penalizes classes with large
spread. Decision boundaries are now quadratic (ellipsoidal). QDA is more
flexible than LDA but requires enough observations per class to reliably
estimate $\boldsymbol\Sigma_g$ --- roughly $n_g > K(K+1)/2$ as a rule of thumb.

### k-Nearest Neighbors (kNN)

kNN classifies by majority vote among the $k$ closest training curves in FPC
space, using Euclidean distance:

$$d(\boldsymbol\xi_i, \boldsymbol\xi_j) = \|\boldsymbol\xi_i - \boldsymbol\xi_j\|_2 = \sqrt{\sum_{l=1}^K (\xi_{il} - \xi_{jl})^2}$$

The predicted class is whichever label appears most often among the $k$
neighbors. kNN makes no distributional assumptions, making it robust to
non-Gaussian score distributions and nonlinear class boundaries. The choice
of $k$ is typically made via cross-validation: small $k$ gives flexible but
noisy boundaries; large $k$ gives smoother but potentially biased boundaries.

### Kernel Classifier

The kernel classifier works directly in function space without dimension
reduction. It uses the $L^2$ functional distance with Simpson's rule
integration weights:

$$d(X_i, X_j) = \sqrt{\int_0^1 [X_i(t) - X_j(t)]^2\,dt}$$

Classification uses a Gaussian kernel to weight contributions from training
curves:

$$K(d, h) = \exp\!\left(-\frac{d^2}{2h^2}\right)$$

The predicted class is:

$$\hat{g}(X) = \arg\max_g \sum_{j:\,y_j = g} K\bigl(d(X, X_j),\, h\bigr)$$

That is, for each class, the kernel weights of all training curves in that class
are summed, and the class with the largest total weight wins. The **bandwidth**
$h$ controls the locality of the classifier. `fclassif()` selects $h$ via
leave-one-out cross-validation over a grid of distance percentiles.

### DD-Classifier (Depth-vs-Depth)

The depth-based classifier avoids both dimension reduction and distance
computation. Instead, it computes the **statistical depth** of each curve
with respect to each class distribution:

$$D_g(X_i) = \text{depth of } X_i \text{ w.r.t. class } g$$

using the Fraiman--Muniz integrated depth:

$$D_{\text{FM}}(X) = \int_0^1 D_1\bigl(X(t);\, F_t\bigr)\,dt$$

where $D_1$ is the univariate depth (one minus twice the distance from the
median CDF value) and $F_t$ is the marginal distribution at time $t$.

For $G$ classes, each curve is mapped to a point
$(D_1(X_i), \ldots, D_G(X_i)) \in \mathbb{R}^G$ and classification proceeds
in this "depth-depth" space. The simplest rule assigns to
$\hat{g} = \arg\max_g D_g(X_i)$: the class in which the curve is most
"central". This approach is robust to outliers because depth is a rank-based
measure.

## Iris as Functional Data

Instead of a toy simulation we use the classic **iris** dataset, transformed
into functional data via the Andrews transformation. This maps each
4-dimensional observation to a curve on $[-\pi, \pi]$ using a Fourier
expansion --- preserving Euclidean distances up to a constant factor.

The three species (*setosa*, *versicolor*, *virginica*) are the classes.
Crucially, *versicolor* and *virginica* overlap substantially, making this a
non-trivial classification task.

```{r andrews-transform}
andrews_transform <- function(X, m = 200) {
  if (is.data.frame(X)) X <- as.matrix(X)
  n <- nrow(X)
  p <- ncol(X)
  t_grid <- seq(-pi, pi, length.out = m)

  basis <- matrix(0, m, p)
  basis[, 1] <- 1 / sqrt(2)
  for (j in 2:p) {
    k <- ceiling((j - 1) / 2)
    if (j %% 2 == 0) {
      basis[, j] <- sin(k * t_grid)
    } else {
      basis[, j] <- cos(k * t_grid)
    }
  }

  fdata(X %*% t(basis), argvals = t_grid)
}
```

```{r data}
X_raw <- as.matrix(iris[, 1:4])
X <- scale(X_raw)
species <- iris$Species
y <- as.integer(species)  # 1 = setosa, 2 = versicolor, 3 = virginica

fd <- andrews_transform(X)
```

```{r iris-plot, fig.height = 5}
plot(fd, color = species, alpha = 0.35, show.mean = TRUE,
     palette = c("setosa" = "#E69F00", "versicolor" = "#56B4E9",
                 "virginica" = "#009E73")) +
  labs(x = expression(t), y = expression(f[x](t)),
       title = "Iris Andrews Curves by Species")
```

Setosa is well separated, but versicolor and virginica overlap ---
exactly the kind of challenge where classifier choice matters.

## Comparing All Five Methods

```{r compare-methods}
methods <- c("lda", "qda", "knn", "kernel", "dd")
results <- list()

for (meth in methods) {
  if (meth == "knn") {
    results[[meth]] <- fclassif(fd, y, method = meth, ncomp = 3, k = 7)
  } else {
    results[[meth]] <- fclassif(fd, y, method = meth, ncomp = 3)
  }
}

acc_df <- data.frame(
  Method = toupper(methods),
  Training_Accuracy = sapply(results, function(r) round(r$accuracy, 3))
)
print(acc_df)
```

Training accuracy alone can be misleading --- QDA and kNN may overfit.
Cross-validation gives a fairer picture (see below).

## Confusion Matrix

The LDA confusion matrix reveals where errors concentrate:

```{r confusion, fig.height = 4}
plot(results[["lda"]])
```

Most errors are between versicolor and virginica, as expected from the
overlapping Andrews curves. Setosa is perfectly classified because its curves
are well separated in the FPC score space.

## Cross-Validation

`fclassif.cv()` evaluates out-of-sample error via k-fold cross-validation.
This gives the most honest estimate of how each classifier would perform on
new data, guarding against the optimism of training accuracy.

```{r cv}
set.seed(42)
cv_results <- list()
for (meth in methods) {
  cv_results[[meth]] <- fclassif.cv(fd, y, method = meth,
                                     ncomp = 3, nfold = 10, seed = 42)
}

cv_df <- data.frame(
  Method = toupper(methods),
  CV_Error = sapply(cv_results, function(r) round(r$error.rate, 3)),
  CV_Accuracy = sapply(cv_results, function(r) round(1 - r$error.rate, 3))
)
print(cv_df)
```

The CV error rates account for overfitting. Methods that had perfect training
accuracy (like QDA) may show higher CV error than simpler methods (like LDA),
especially when sample sizes per class are small relative to the number of
parameters.

## Choosing the Number of Components

Cross-validation can also help select the optimal number of FPC components.
Too few loses discriminative information; too many introduces noise.

```{r ncomp-selection}
ncomp_range <- 1:6
cv_by_ncomp <- sapply(ncomp_range, function(k) {
  fclassif.cv(fd, y, method = "lda", ncomp = k, nfold = 10, seed = 42)$error.rate
})

df_ncomp <- data.frame(ncomp = ncomp_range, error = cv_by_ncomp)
ggplot(df_ncomp, aes(x = ncomp, y = error)) +
  geom_line() +
  geom_point(size = 2) +
  labs(x = "Number of FPC Components", y = "10-Fold CV Error Rate",
       title = "LDA on Iris: Component Selection") +
  scale_x_continuous(breaks = ncomp_range)
```

The optimal number of components balances two effects: adding early components
captures the dominant modes of between-class variation, improving discrimination.
Adding later components brings in directions that are mainly within-class noise,
diluting the signal.

## Choosing a Classifier

The table below summarizes practical considerations:

| Criterion | LDA | QDA | kNN | Kernel | DD |
|-----------|-----|-----|-----|--------|-----|
| **Assumptions** | Common covariance | Class-specific covariance | None | None | None |
| **Min. $n_g$** | $> K$ | $> K(K+1)/2$ | $> k$ | $> 20$ | $> 20$ |
| **Boundaries** | Linear | Quadratic | Flexible | Flexible | Depth-based |
| **Tuning** | $K$ only | $K$ only | $K$, $k$ | $h$ | $K$ |
| **Speed** | Fast | Fast | Moderate | Slow | Moderate |
| **Robustness** | Low | Low | Moderate | Moderate | High |

**Rules of thumb:**

- Start with **LDA** as a baseline --- it is fast, interpretable, and often
  competitive.
- Try **QDA** when you suspect classes have different shapes in FPC space
  (different covariance patterns).
- Use **kNN** when decision boundaries may be nonlinear and you have enough
  data per class.
- Use the **kernel** classifier for small-to-moderate datasets where you want
  to avoid the FPC projection step entirely.
- Use **DD** when robustness to outliers is paramount or distributional
  assumptions are suspect.

## References

- Delaigle, A. and Hall, P. (2012). Achieving near perfect classification for functional data. *Journal of the Royal Statistical Society: Series B*, 74(2), 267--286.
- Cuevas, A., Febrero, M., and Fraiman, R. (2007). Robust estimation and classification for functional data via projection pursuit. *Computational Statistics*, 22(3), 481--496.
- Li, J. and Yu, T. (2008). DD-classifier: nonparametric classification procedure based on DD-plot. *Journal of the American Statistical Association*, 103(483), 737--747.
- Lopez-Pintado, S. and Romo, J. (2009). On the concept of depth for functional data. *Journal of the American Statistical Association*, 104(486), 718--734.

## See Also

- `vignette("articles/andrews-transformation")` for more on the Andrews
  transform and its distance-preservation property
- `vignette("articles/functional-regression")` for scalar-on-function
  regression including `functional.logistic()`
- `vignette("articles/clustering")` for unsupervised functional clustering
- `vignette("articles/fpca")` for functional PCA, which underlies LDA/QDA/kNN
