---
title: "Model-Based Clustering with Gaussian Mixtures"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Gaussian Mixture Models (GMMs) provide model-based clustering with several
advantages over k-means:

- **Soft assignment**: Each curve gets a probability of belonging to each
  cluster, not just a hard label.
- **Automatic model selection**: BIC or ICL criteria select the number of
  clusters.
- **Flexible covariance**: Full, diagonal, or spherical covariance structures
  adapt to cluster shapes.

**fdars** implements GMM clustering through `cluster.gmm()`, which projects
functional data onto a B-spline basis before fitting the mixture model.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Simulated Data

```{r data}
set.seed(42)
n_per_cluster <- 25
m <- 100
t_grid <- seq(0, 1, length.out = m)

# Cluster 1: Sine curves
X1 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X1[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.15)
}

# Cluster 2: Cosine curves
X2 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X2[i, ] <- cos(2 * pi * t_grid) + rnorm(m, sd = 0.15)
}

# Cluster 3: Linear curves
X3 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X3[i, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.15)
}

X <- rbind(X1, X2, X3)
true_clusters <- rep(1:3, each = n_per_cluster)

fd <- fdata(X, argvals = t_grid)
plot(fd)
```

## Fitting a GMM

`cluster.gmm()` searches over a range of k values and selects the best model
using BIC (default) or ICL.

```{r gmm-fit}
gmm_fit <- cluster.gmm(fd, k.range = 2:6, seed = 123)
print(gmm_fit)
```

## Model Selection: BIC Curve

The BIC values across candidate k help visualize model selection:

```{r bic-plot, fig.height = 4}
plot(gmm_fit)
```

## Soft vs Hard Assignment

Unlike k-means, GMM provides membership probabilities:

```{r membership}
# Show membership for first 6 curves
head(round(gmm_fit$membership, 3))

# Maximum membership gives hard assignment
cat("Hard clusters:", head(gmm_fit$cluster), "\n")
```

## Comparison with K-Means

```{r compare}
kmeans_fit <- cluster.kmeans(fd, ncl = 3, seed = 123)

cat("K-means WCSS:", round(kmeans_fit$tot.withinss, 2), "\n")
cat("GMM BIC:", round(gmm_fit$bic, 2), "\n")
cat("GMM converged:", gmm_fit$converged, "\n")
```

## Covariance Types

`cluster.gmm()` supports different covariance structures:

- `"full"` (default): Each cluster has its own full covariance matrix.
- `"diagonal"`: Diagonal covariance (features are independent within clusters).

```{r cov-types}
gmm_diag <- cluster.gmm(fd, k.range = 3, cov.type = "diagonal", seed = 123)
cat("Full GMM BIC:", round(gmm_fit$bic, 2), "\n")
cat("Diagonal GMM BIC:", round(gmm_diag$bic, 2), "\n")
```

## Predicting New Observations

Assign new curves to clusters using the fitted model:

```{r predict}
# Generate new curves
set.seed(99)
X_new <- matrix(0, 3, m)
X_new[1, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)   # Should be cluster 1
X_new[2, ] <- cos(2 * pi * t_grid) + rnorm(m, sd = 0.1)   # Should be cluster 2
X_new[3, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.1)          # Should be cluster 3

fd_new <- fdata(X_new, argvals = t_grid)
pred <- predict(gmm_fit, fd_new)

cat("Predicted clusters:", pred$cluster, "\n")
cat("Membership probabilities:\n")
print(round(pred$membership, 3))
```

## BIC vs ICL

ICL (Integrated Completed Likelihood) penalizes cluster overlap more strongly
than BIC:

```{r icl}
gmm_icl <- cluster.gmm(fd, k.range = 2:6, criterion = "icl", seed = 123)
cat("BIC selected k:", gmm_fit$k, "\n")
cat("ICL selected k:", gmm_icl$k, "\n")
```

## See Also

- `vignette("articles/clustering")` for k-means and fuzzy c-means
- `vignette("articles/functional-classification")` for supervised
  classification with `fclassif()`
