---
title: "Model-Based Clustering with Gaussian Mixtures"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Gaussian Mixture Models (GMMs) provide model-based clustering with several
advantages over k-means:

- **Soft assignment**: Each curve gets a probability of belonging to each
  cluster, not just a hard label.
- **Automatic model selection**: BIC or ICL criteria select the number of
  clusters.
- **Flexible covariance**: Full, diagonal, or spherical covariance structures
  adapt to cluster shapes.

**fdars** implements GMM clustering through `cluster.gmm()`, which projects
functional data onto a B-spline basis before fitting the mixture model.

```{r setup}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Why Model-Based Clustering?

K-means is a natural starting point for clustering, but it makes strong
implicit assumptions:

1. **Spherical clusters.** K-means minimizes within-cluster sum of squares,
   which treats all directions equally. It cannot adapt to elongated or
   correlated cluster shapes.
2. **Equal-size clusters.** K-means assigns each point to its nearest centroid,
   effectively giving all clusters equal "pull" regardless of their size.
3. **Hard assignment.** Each observation belongs to exactly one cluster, with
   no notion of uncertainty.

Gaussian mixture models relax all three assumptions by positing that the data
arise from a mixture of $K$ Gaussian distributions, each with its own mean,
covariance, and mixing weight. The EM algorithm provides:

- **Soft (probabilistic) assignment**: each observation gets a responsibility
  $\tau_{ik} \in [0, 1]$ for each cluster, quantifying classification
  uncertainty.
- **Adaptive cluster shapes**: full covariance matrices capture correlations
  between features within each cluster.
- **Principled model selection**: information criteria (BIC, ICL) balance fit
  against complexity, selecting both the number of clusters and the covariance
  structure.

## Mathematical Framework

### Gaussian Mixture Model

A $K$-component GMM models the density of the data $\mathbf{z} \in \mathbb{R}^d$
as:

$$f(\mathbf{z}) = \sum_{k=1}^K \pi_k \,\mathcal{N}(\mathbf{z};\, \boldsymbol\mu_k,\, \boldsymbol\Sigma_k)$$

where:

- $\pi_k > 0$ are the **mixing weights** with $\sum_{k=1}^K \pi_k = 1$
- $\boldsymbol\mu_k \in \mathbb{R}^d$ is the mean of component $k$
- $\boldsymbol\Sigma_k \in \mathbb{R}^{d \times d}$ is the covariance of
  component $k$
- $\mathcal{N}(\mathbf{z};\, \boldsymbol\mu,\, \boldsymbol\Sigma)$ is the
  multivariate Gaussian density

The **log-likelihood** for observations $\mathbf{z}_1, \ldots, \mathbf{z}_n$ is:

$$\ell(\theta) = \sum_{i=1}^n \log\left[\sum_{k=1}^K \pi_k \,\mathcal{N}(\mathbf{z}_i;\, \boldsymbol\mu_k,\, \boldsymbol\Sigma_k)\right]$$

Direct maximization is difficult because of the log-of-sum structure. The EM
algorithm resolves this.

### EM Algorithm

The Expectation-Maximization algorithm alternates between two steps:

**E-step.** Compute the **responsibilities** (posterior cluster probabilities):

$$\tau_{ik} = \frac{\pi_k \,\mathcal{N}(\mathbf{z}_i;\, \boldsymbol\mu_k,\, \boldsymbol\Sigma_k)}{\sum_{c=1}^K \pi_c \,\mathcal{N}(\mathbf{z}_i;\, \boldsymbol\mu_c,\, \boldsymbol\Sigma_c)}$$

Each $\tau_{ik}$ is the probability that observation $i$ was generated by
component $k$, given the current parameter estimates.

**M-step.** Update parameters using the responsibilities as soft weights:

$$n_k = \sum_{i=1}^n \tau_{ik}, \qquad \pi_k = \frac{n_k}{n}$$

$$\boldsymbol\mu_k = \frac{1}{n_k}\sum_{i=1}^n \tau_{ik}\,\mathbf{z}_i$$

$$\boldsymbol\Sigma_k = \frac{1}{n_k}\sum_{i=1}^n \tau_{ik}\,(\mathbf{z}_i - \boldsymbol\mu_k)(\mathbf{z}_i - \boldsymbol\mu_k)^T$$

The algorithm iterates until convergence:
$|\ell^{(t+1)} - \ell^{(t)}| < \varepsilon$. Each iteration is guaranteed to
increase (or maintain) the log-likelihood, ensuring convergence to a local
maximum.

**Numerical stability.** Computing $\mathcal{N}(\mathbf{z}_i;\, \boldsymbol\mu_k,\, \boldsymbol\Sigma_k)$
involves exponentiating large negative values, which can underflow. The
implementation uses the **log-sum-exp trick**: work in log space and subtract
the maximum before exponentiating.

**Initialization.** EM is sensitive to starting values. `cluster.gmm()` uses
**K-means++ initialization** to obtain well-separated initial centroids, which
substantially reduces the chance of converging to a poor local maximum.

### Model Selection

#### BIC (Bayesian Information Criterion)

$$\text{BIC} = -2\ell(\hat\theta) + p\log n$$

where $p$ is the number of free parameters. For a full-covariance GMM with $K$
components in $d$ dimensions:

$$p = K \cdot \left[\frac{d(d+1)}{2} + d + 1\right] - 1$$

accounting for $d(d+1)/2$ covariance parameters, $d$ mean parameters, and 1
mixing weight per component, minus 1 for the sum-to-one constraint. Lower BIC
is better.

#### ICL (Integrated Completed Likelihood)

$$\text{ICL} = \text{BIC} - 2\sum_{i=1}^n\sum_{k=1}^K \tau_{ik}\log\tau_{ik}$$

ICL adds an **entropy penalty** that penalizes overlap between clusters. When
responsibilities are crisp ($\tau_{ik} \approx 0$ or $1$), the entropy term is
small and ICL $\approx$ BIC. When clusters overlap heavily, the entropy is
large and ICL penalizes more strongly, favoring fewer, better-separated clusters.

**When BIC and ICL disagree:** BIC tends to select more components (it is a
better density estimator), while ICL tends to select fewer (it prioritizes
well-separated clusters). If your goal is clustering, ICL is often more
appropriate; if your goal is density estimation, prefer BIC.

### Covariance Structures

- **Full**: $d(d+1)/2$ free parameters per component. Captures all pairwise
  correlations. Flexible but data-hungry.
- **Diagonal**: $d$ parameters per component. Assumes features are independent
  within each cluster. Parsimonious, useful when the basis coefficients are
  approximately uncorrelated.

## From Curves to Features

`cluster.gmm()` does not operate on raw discretized curves. Instead, it first
projects each curve onto a **B-spline basis**, producing a vector of basis
coefficients $\mathbf{z}_i \in \mathbb{R}^d$ where $d$ equals the number of
basis functions (`nbasis`). This step:

1. **Reduces dimensionality** from $m$ grid points to $d$ basis coefficients
   (typically $d = 10\text{--}30$).
2. **Imposes smoothness** because B-splines are smooth by construction.
3. **Decorrelates** (partially), since B-spline coefficients are less redundant
   than adjacent grid-point values.

The choice of `nbasis` controls how much curve detail is retained. Too few
basis functions smooth away cluster-distinguishing features; too many inflate
the dimensionality and slow convergence.

## Simulated Data

```{r data}
set.seed(42)
n_per_cluster <- 25
m <- 100
t_grid <- seq(0, 1, length.out = m)

# Cluster 1: Sine curves
X1 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X1[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.15)
}

# Cluster 2: Cosine curves
X2 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X2[i, ] <- cos(2 * pi * t_grid) + rnorm(m, sd = 0.15)
}

# Cluster 3: Linear curves
X3 <- matrix(0, n_per_cluster, m)
for (i in 1:n_per_cluster) {
  X3[i, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.15)
}

X <- rbind(X1, X2, X3)
true_clusters <- rep(1:3, each = n_per_cluster)

fd <- fdata(X, argvals = t_grid)
plot(fd)
```

The three clusters have visually distinct shapes: sinusoidal, cosinusoidal,
and linear. The added noise ($\sigma = 0.15$) creates some overlap but the
clusters remain well separated --- an ideal scenario for GMM.

## Fitting a GMM

`cluster.gmm()` searches over a range of k values and selects the best model
using BIC (default) or ICL.

```{r gmm-fit}
gmm_fit <- cluster.gmm(fd, k.range = 2:6, seed = 123)
print(gmm_fit)
```

The selected $k$ should match the true number of clusters (3). The BIC values
for $k < 3$ will be worse because they cannot capture the three distinct shapes,
and $k > 3$ will be penalized for unnecessary parameters.

## Model Selection: BIC Curve

The BIC values across candidate k help visualize model selection:

```{r bic-plot, fig.height = 4}
plot(gmm_fit)
```

A clear minimum (or elbow) at $k = 3$ confirms the true structure. When the
BIC curve is flat, the data may not have well-defined clusters, or multiple
values of $k$ provide comparably good fits.

## Soft vs Hard Assignment

Unlike k-means, GMM provides membership probabilities:

```{r membership}
# Show membership for first 6 curves
head(round(gmm_fit$membership, 3))

# Maximum membership gives hard assignment
cat("Hard clusters:", head(gmm_fit$cluster), "\n")
```

Membership probabilities near 1 indicate confident assignment; values like
$(0.6, 0.3, 0.1)$ indicate the curve sits between clusters. High-entropy
observations are natural candidates for further investigation --- they may be
transitional curves, outliers, or evidence that the model needs refinement.

## Comparison with K-Means

```{r compare}
kmeans_fit <- cluster.kmeans(fd, ncl = 3, seed = 123)

cat("K-means WCSS:", round(kmeans_fit$tot.withinss, 2), "\n")
cat("GMM BIC:", round(gmm_fit$bic, 2), "\n")
cat("GMM converged:", gmm_fit$converged, "\n")
```

K-means and GMM often agree on well-separated clusters. The key differences
emerge when clusters have unequal sizes, non-spherical shapes, or overlapping
boundaries --- situations where GMM's flexibility pays off.

## Covariance Types

`cluster.gmm()` supports different covariance structures:

- `"full"` (default): Each cluster has its own full covariance matrix.
- `"diagonal"`: Diagonal covariance (features are independent within clusters).

```{r cov-types}
gmm_diag <- cluster.gmm(fd, k.range = 3, cov.type = "diagonal", seed = 123)
cat("Full GMM BIC:", round(gmm_fit$bic, 2), "\n")
cat("Diagonal GMM BIC:", round(gmm_diag$bic, 2), "\n")
```

If the diagonal model achieves a comparable BIC, it suggests that the B-spline
coefficients are approximately uncorrelated within clusters --- a simpler model
is adequate and will generalize better.

## Predicting New Observations

Assign new curves to clusters using the fitted model:

```{r predict}
# Generate new curves
set.seed(99)
X_new <- matrix(0, 3, m)
X_new[1, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)   # Should be cluster 1
X_new[2, ] <- cos(2 * pi * t_grid) + rnorm(m, sd = 0.1)   # Should be cluster 2
X_new[3, ] <- 2 * t_grid - 1 + rnorm(m, sd = 0.1)          # Should be cluster 3

fd_new <- fdata(X_new, argvals = t_grid)
pred <- predict(gmm_fit, fd_new)

cat("Predicted clusters:", pred$cluster, "\n")
cat("Membership probabilities:\n")
print(round(pred$membership, 3))
```

The membership probabilities for new curves should be highly concentrated on the
correct cluster, since the test curves were generated with lower noise
($\sigma = 0.1$) than the training data.

## BIC vs ICL

ICL (Integrated Completed Likelihood) penalizes cluster overlap more strongly
than BIC:

```{r icl}
gmm_icl <- cluster.gmm(fd, k.range = 2:6, criterion = "icl", seed = 123)
cat("BIC selected k:", gmm_fit$k, "\n")
cat("ICL selected k:", gmm_icl$k, "\n")
```

When BIC and ICL agree, the clustering structure is unambiguous. When they
disagree (ICL selects fewer clusters), it signals that some BIC-selected
components overlap substantially and might be better interpreted as a single
heterogeneous cluster.

## References

- McLachlan, G.J. and Peel, D. (2000). *Finite Mixture Models*. Wiley.
- Fraley, C. and Raftery, A.E. (2002). Model-based clustering, discriminant analysis, and density estimation. *Journal of the American Statistical Association*, 97(458), 611--631.
- Jacques, J. and Preda, C. (2014). Functional data clustering: a survey. *Advances in Data Analysis and Classification*, 8(3), 231--255.
- Biernacki, C., Celeux, G., and Govaert, G. (2000). Assessing a mixture model for clustering with the integrated completed likelihood. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 22(7), 719--725.

## See Also

- `vignette("articles/clustering")` for k-means and fuzzy c-means
- `vignette("articles/functional-classification")` for supervised
  classification with `fclassif()`
