---
title: "Landmark Registration"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

**Landmark registration** aligns functional data by identifying salient features
-- peaks, valleys, zero-crossings, or inflection points -- and warping curves so
that corresponding features occur at the same time. Unlike elastic alignment,
which searches for the globally optimal warping function via dynamic programming,
landmark registration anchors the warping at specific feature locations and
interpolates between them.

This approach is particularly useful when:

- Curves have **identifiable, biologically or physically meaningful features**
  (e.g., the P, QRS, and T waves in an ECG, peak force in a gait cycle)
- Features are **sparse** and can be reliably detected
- You want alignment to respect **domain knowledge** about which features should
  coincide

```{r setup}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

## Mathematical Framework

### The Registration Problem

Given a sample of curves $f_1, \ldots, f_n$ observed on a common domain
$[a, b]$, registration seeks warping functions
$\gamma_i : [a, b] \to [a, b]$ such that the registered curves
$\tilde{f}_i(t) = f_i(\gamma_i(t))$ have their salient features aligned.

In the landmark approach, we identify $k$ corresponding feature times
$\tau_{i,1}, \ldots, \tau_{i,k}$ in each curve $f_i$, and common target times
$\tau_1^*, \ldots, \tau_k^*$. The warping function $\gamma_i$ must satisfy:

$$\gamma_i(\tau_j^*) = \tau_{i,j}, \quad j = 1, \ldots, k$$

with boundary conditions $\gamma_i(a) = a$ and $\gamma_i(b) = b$.

### Piecewise-Linear Warping

The simplest solution is piecewise-linear interpolation between the $k+2$
anchor points (boundaries plus landmarks). Let $\tau_0^* = a$,
$\tau_{k+1}^* = b$, and similarly $\tau_{i,0} = a$, $\tau_{i,k+1} = b$.
For $t \in [\tau_j^*, \tau_{j+1}^*]$:

$$\gamma_i(t) = \tau_{i,j} + \frac{\tau_{i,j+1} - \tau_{i,j}}{\tau_{j+1}^* - \tau_j^*} (t - \tau_j^*)$$

This warping is monotone (orientation-preserving) as long as the landmarks
are in the same order in every curve. The derivative $\dot\gamma_i$ is
piecewise constant, with value $(\tau_{i,j+1} - \tau_{i,j}) / (\tau_{j+1}^* - \tau_j^*)$
on each segment.

### Target Landmark Selection

The common target times $\tau_j^*$ are chosen as the average of the detected
landmark positions across curves:

$$\tau_j^* = \frac{1}{n} \sum_{i=1}^n \tau_{i,j}$$

This minimizes the total squared warping (in the piecewise-linear sense) and
ensures the target positions are representative of the sample.

### Prominence and Feature Detection

A **peak** at time $t_0$ in a curve $f$ is a local maximum: $f(t_0) \geq f(t)$
for all $t$ in a neighborhood of $t_0$. Its **prominence** is defined as:

$$\text{prom}(t_0) = f(t_0) - \max\left( \min_{t \in [t_L, t_0]} f(t),\; \min_{t \in [t_0, t_R]} f(t) \right)$$

where $t_L$ and $t_R$ are the nearest higher peaks to the left and right
(or the domain boundaries). Prominence measures how much a peak stands out
from its surroundings -- a peak nestled on the flank of a larger peak has
low prominence, while an isolated peak has high prominence.

Valleys are detected analogously (as peaks of $-f$). Zero-crossings are
points where $f$ changes sign, and inflection points are where $f''$ changes
sign.

## Quick Start

```{r quick-start}
set.seed(42)
argvals <- seq(0, 1, length.out = 200)

# Simulate curves with a shifted peak
n <- 12
data <- matrix(0, n, 200)
peak_locs <- runif(n, 0.3, 0.7)
for (i in 1:n) {
  data[i, ] <- exp(-100 * (argvals - peak_locs[i])^2) +
               0.3 * sin(4 * pi * argvals)
}
fd <- fdata(data, argvals = argvals)

# Register using detected peaks
lr <- landmark.register(fd, kind = "peak", min.prominence = 0.5,
                        expected.count = 1)
print(lr)
```

```{r quick-start-plot}
plot(lr, type = "both")
```

The registered curves now have their dominant peak aligned to a common position.

## Detecting Landmarks

Before registration, you can inspect which landmarks are detected in each curve.
The `detect.landmarks()` function finds features of a given type and returns
their position, value, and prominence.

```{r detect-peaks}
lms <- detect.landmarks(fd, kind = "peak", min.prominence = 0.3)

# Show landmarks for first 3 curves
for (i in 1:3) {
  cat("Curve", i, ":\n")
  print(lms[[i]])
  cat("\n")
}
```

### Landmark Types

Four types of landmarks can be detected:

| Kind | Description |
|------|-------------|
| `"peak"` | Local maxima |
| `"valley"` | Local minima |
| `"zero"` | Zero-crossings (sign changes) |
| `"inflection"` | Inflection points (curvature sign changes) |

```{r detect-types}
# Detect valleys and zero-crossings
valleys <- detect.landmarks(fd, kind = "valley", min.prominence = 0.1)
zeros <- detect.landmarks(fd, kind = "zero")

cat("Curve 1 peaks:\n")
print(lms[[1]])
cat("\nCurve 1 valleys:\n")
print(valleys[[1]])
cat("\nCurve 1 zero-crossings:\n")
print(zeros[[1]])
```

### Prominence Filtering

The `min.prominence` parameter filters out minor features. A peak's **prominence**
measures how much it stands out from its surroundings -- the minimum vertical
drop required to reach a higher peak. Setting a higher threshold keeps only the
most salient features:

```{r prominence}
# All peaks (no filtering)
lms_all <- detect.landmarks(fd, kind = "peak", min.prominence = 0)
cat("Curve 1 - all peaks:", nrow(lms_all[[1]]), "\n")

# Only prominent peaks
lms_prominent <- detect.landmarks(fd, kind = "peak", min.prominence = 0.5)
cat("Curve 1 - prominent peaks:", nrow(lms_prominent[[1]]), "\n")
```

## Registration

The `landmark.register()` function detects landmarks in every curve, computes
common target positions (by averaging detected landmark positions), and warps
each curve so that its landmarks map to the targets.

```{r register}
lr <- landmark.register(fd, kind = "peak", min.prominence = 0.5,
                        expected.count = 1)
```

The result is an S3 object of class `"landmark.register"` containing:

- `registered` -- an `fdata` of warped curves
- `gammas` -- an `fdata` of warping functions
- `landmarks` -- detected landmarks for each curve
- `target_landmarks` -- the common target positions

```{r register-warps}
plot(lr, type = "warps")
```

The warping functions are piecewise-linear, with knots at the landmark positions.
Between landmarks, the warping is a simple linear interpolation.

### Expected Count

When `expected.count > 0`, the function selects the most prominent landmarks
up to that count. This is useful when different curves have different numbers
of detected features but you know the true number of corresponding features:

```{r expected-count}
# Data with two peaks per curve
data2 <- matrix(0, n, 200)
for (i in 1:n) {
  pk1 <- runif(1, 0.2, 0.4)
  pk2 <- runif(1, 0.6, 0.8)
  data2[i, ] <- exp(-100 * (argvals - pk1)^2) +
                exp(-100 * (argvals - pk2)^2)
}
fd2 <- fdata(data2, argvals = argvals)

lr2 <- landmark.register(fd2, kind = "peak", min.prominence = 0.3,
                         expected.count = 2)
cat("Target landmarks:", round(lr2$target_landmarks, 3), "\n")
```

```{r expected-count-plot}
p1 <- plot(fd2) + labs(title = "Original (two peaks)")
p2 <- plot(lr2$registered) + labs(title = "Landmark Registered")
p1 + p2
```

## Comparison: Elastic vs Landmark Registration

Elastic alignment and landmark registration solve the same problem -- removing
phase variability -- but with different strategies:

| | Elastic | Landmark |
|---|---------|----------|
| **Approach** | Global DP optimization | Feature-based anchoring |
| **Warping** | Smooth, globally optimal | Piecewise-linear between landmarks |
| **Requires** | Nothing (fully automatic) | Detectable, meaningful features |
| **Best for** | General-purpose alignment | Domain-driven feature alignment |
| **Speed** | $O(nm^2)$ | $O(nm + nk)$ where $k$ = landmarks |

```{r comparison}
# Same data, two approaches
ea <- elastic.align(fd)

p1 <- plot(fd) + labs(title = "Original")
p2 <- plot(ea, type = "aligned") + labs(title = "Elastic Alignment")
p3 <- plot(lr, type = "registered") + labs(title = "Landmark Registration")
p1 + p2 + p3
```

### Constrained Elastic Alignment

You can combine the best of both approaches using `elastic.align.constrained()`,
which runs elastic alignment with landmark constraints. The warping must pass
through the landmark positions, but is otherwise optimized:

```{r constrained}
ec <- elastic.align.constrained(fd, kind = "peak", min.prominence = 0.5,
                                 expected.count = 1)
p_ea <- plot(ea, type = "aligned") + labs(title = "Elastic")
p_ec <- plot(ec, type = "aligned") + labs(title = "Constrained Elastic")
p_lr <- plot(lr, type = "registered") + labs(title = "Landmark")
p_ea + p_ec + p_lr
```

The constrained approach produces smooth warps (like elastic) while respecting
feature correspondence (like landmark).

## See Also

- `vignette("elastic-alignment", package = "fdars")` -- elastic alignment
  framework
- `elastic.align.constrained()` -- constrained elastic alignment
- `alignment.quality()` -- diagnostics for alignment results
