---
title: "Outlier Detection"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Outlier detection in functional data identifies curves that are atypical
or anomalous compared to the rest of the sample. **fdars** provides several
methods based on functional depth and likelihood ratio tests.

```{=html}
<img src="../reference/figures/outlier-types-diagram.svg" alt="Functional Outlier Types Diagram" style="width:100%; max-width:700px;">
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())

# Generate normal data with low noise for clear signal
set.seed(42)
n <- 30
m <- 100
t_grid <- seq(0, 1, length.out = m)

X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)
}

# Add three distinct types of outliers
X[1, ] <- sin(2 * pi * t_grid) + 3           # MAGNITUDE outlier (shifted up)
X[2, ] <- -sin(2 * pi * t_grid)              # SHAPE outlier (inverted pattern)
X[3, ] <- 3 * sin(2 * pi * t_grid)           # AMPLITUDE outlier (larger scale)

fd <- fdata(X, argvals = t_grid)

# Visualize with outliers highlighted
df_curves <- data.frame(
  t = rep(t_grid, n),
  value = as.vector(t(X)),
  curve = rep(1:n, each = m),
  type = rep(c("Magnitude", "Shape", "Amplitude", rep("Normal", n - 3)), each = m)
)
df_curves$type <- factor(df_curves$type, levels = c("Normal", "Magnitude", "Shape", "Amplitude"))

ggplot(df_curves, aes(x = t, y = value, group = curve, color = type)) +
  geom_line(alpha = 0.7) +
  scale_color_manual(values = c("Normal" = "gray50", "Magnitude" = "red",
                                 "Shape" = "blue", "Amplitude" = "green3")) +
  labs(title = "Functional Data with Three Types of Outliers",
       x = "t", y = "X(t)", color = "Type") +
  theme(legend.position = "bottom")
```

## Depth-Based Methods

Depth-based outlier detection identifies curves with unusually low depth
(far from the center of the data).

### Weighted Depth Method (outliers.depth.pond)

Uses bootstrap resampling to estimate the distribution of depths and
identifies curves with depth below a data-driven cutoff. The function supports
three different methods for computing the threshold.

#### Threshold Methods

| Method | Formula | When to Use |
|--------|---------|-------------|
| `"quantile"` | `quantile(depths, quan)` | When you expect a specific proportion of outliers |
| `"mad"` | `median - k × MAD` | More robust when outliers may already exist |
| `"iqr"` | `Q1 - k × IQR` | Boxplot-style detection |

**Default: 95th percentile threshold** (`quan = 0.05`), which flags curves in the
bottom 5% of depths as outliers.

```{r depth-pond}
# Default: quantile method with quan = 0.05 (95th percentile, flags bottom 5%)
out_pond <- outliers.depth.pond(fd, nb = 1000)
print(out_pond)
```

#### Comparing Threshold Methods

```{r threshold-methods}
# Quantile method: flags curves with depth in the bottom 5% (default)
out_quantile <- outliers.depth.pond(fd, nb = 1000,
                                     threshold_method = "quantile", quan = 0.05)
cat("Quantile (5%): ", out_quantile$outliers, "\n")

# More permissive: bottom 10%
out_quantile10 <- outliers.depth.pond(fd, nb = 1000,
                                       threshold_method = "quantile", quan = 0.1)
cat("Quantile (10%):", out_quantile10$outliers, "\n")

# MAD method: more robust, uses median - 2.5*MAD
out_mad <- outliers.depth.pond(fd, nb = 1000,
                                threshold_method = "mad", k = 2.5)
cat("MAD (k=2.5):   ", out_mad$outliers, "\n")

# IQR method: boxplot-like, uses Q1 - 1.5*IQR
out_iqr <- outliers.depth.pond(fd, nb = 1000,
                                threshold_method = "iqr", k = 1.5)
cat("IQR (k=1.5):   ", out_iqr$outliers, "\n")
```

**Choosing the right method:**

- **Quantile** (default): Uses a fixed proportion cutoff. The default `quan = 0.05`
  (95th percentile) flags the bottom 5% of curves. Increase to `quan = 0.1` for
  more permissive detection.

- **MAD**: More robust to existing outliers in the data. The default `k = 2.5`
  corresponds roughly to a 1-2% false positive rate. Increase `k` for stricter
  detection (fewer outliers).

- **IQR**: Similar to boxplot fences. The default `k = 1.5` is the standard
  boxplot rule. Use `k = 3.0` for "far outliers" only.

### Visualizing Outliers

```{r plot-outliers}
plot(out_pond)
```

### Examining Results

```{r examine-pond}
# Which curves are outliers?
out_pond$outliers

# Depth values for all curves
head(out_pond$depths)

# Cutoff used
cat("Cutoff:", out_pond$cutoff, "\n")
cat("Threshold method:", out_pond$threshold_method, "\n")
```

### Understanding depth.pond Results

The `outliers.depth.pond` method uses bootstrap resampling to estimate what depth
values are "normal" for your dataset.

**Key behaviors:**

1. **Edge curves**: Curves near the boundary of the data cloud naturally have lower
   depth, even if they're not true outliers
2. **Bootstrap variability**: Small samples give unstable cutoffs - use at least
   `nb = 200` for stable results
3. **Threshold choice matters**: The `quantile` method with a fixed proportion will
   always flag that proportion as outliers. Use `mad` or `iqr` for data-driven
   thresholds that adapt to the actual depth distribution.

**Recommendation**: Start with `threshold_method = "mad"` for a robust, data-driven
approach. Adjust `k` based on how conservative you want the detection to be.

Compare with `outliers.depth.trim` which uses a fixed trim proportion - more
predictable but requires you to choose the proportion.

### Trimming-Based Method (outliers.depth.trim)

Iteratively removes curves with lowest depth:

```{r depth-trim}
out_trim <- outliers.depth.trim(fd, trim = 0.1, seed = 123)
print(out_trim)
plot(out_trim)
```

### Using Different Depth Functions

Both methods accept a `depth` parameter to specify the depth function:

```{r different-depths}
# Using Random Projection depth
out_rp <- outliers.depth.pond(fd, nb = 1000, seed = 123)

# Using modal depth (default is FM)
out_mode <- outliers.depth.trim(fd, trim = 0.1, seed = 123)
```

## Likelihood Ratio Test (LRT) Method

The LRT method uses a likelihood ratio test to sequentially identify outliers.
It's particularly effective for detecting magnitude outliers.

### Automatic Threshold Computation

The LRT method automatically computes a bootstrap threshold based on a percentile
of the maximum distance distribution under the null hypothesis (no outliers).
By default, the 99th percentile is used, meaning approximately 1% of observations
would be flagged as outliers when there are no true outliers.

```{r lrt-detect}
# The outliers.lrt function automatically computes the threshold
out_lrt <- outliers.lrt(fd, nb = 1000, seed = 123)
print(out_lrt)
plot(out_lrt)
```

### Configuring the Percentile

The `percentile` parameter controls the sensitivity of the LRT method:

- **Higher percentile** (e.g., 0.99): More conservative, fewer false positives
- **Lower percentile** (e.g., 0.95): More sensitive, may catch more subtle outliers

```{r lrt-percentile}
# Default: 99th percentile (conservative)
out_lrt_99 <- outliers.lrt(fd, nb = 1000, seed = 123, percentile = 0.99)
cat("99th percentile outliers:", out_lrt_99$outliers, "\n")

# More sensitive: 95th percentile
out_lrt_95 <- outliers.lrt(fd, nb = 1000, seed = 123, percentile = 0.95)
cat("95th percentile outliers:", out_lrt_95$outliers, "\n")
```

### Manual Threshold Computation

You can also compute the threshold separately if you want to examine it or apply
a custom threshold:

```{r lrt-threshold}
# Compute threshold separately (99th percentile by default)
threshold_99 <- outliers.thres.lrt(fd, nb = 1000, seed = 123)
cat("LRT threshold (99th percentile):", threshold_99, "\n")

# Or with a different percentile
threshold_95 <- outliers.thres.lrt(fd, nb = 1000, seed = 123, percentile = 0.95)
cat("LRT threshold (95th percentile):", threshold_95, "\n")
```

### Full Bootstrap Distribution (`outliers.lrt.dist`)

For finer-grained inference, `outliers.lrt.dist()` returns both the threshold
and the full sorted bootstrap null distribution. This enables per-curve p-value
computation and visualization of the null distribution:

```{r lrt-dist}
lrt_dist <- outliers.lrt.dist(fd, nb = 1000, seed = 123, percentile = 0.99)

cat("Threshold:", round(lrt_dist$threshold, 4), "\n")
cat("Bootstrap distribution: n =", length(lrt_dist$boot.distribution),
    ", range = [", round(min(lrt_dist$boot.distribution), 3), ",",
    round(max(lrt_dist$boot.distribution), 3), "]\n")
```

Visualize the null distribution with the threshold:

```{r lrt-dist-plot, fig.height = 4}
df_boot <- data.frame(max_distance = lrt_dist$boot.distribution)

ggplot(df_boot, aes(x = .data$max_distance)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = lrt_dist$threshold, color = "red",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = lrt_dist$threshold, y = Inf,
           label = paste0("99th percentile = ", round(lrt_dist$threshold, 3)),
           vjust = 2, hjust = -0.1, color = "red", size = 3.5) +
  labs(title = "Bootstrap Null Distribution of Maximum Distances",
       subtitle = "Red line = threshold for outlier detection",
       x = "Maximum Distance", y = "Count")
```

Compute per-curve p-values from the bootstrap distribution:

```{r lrt-pvalues}
# Distance of each curve from the centre
curve_depths <- depth.FM(fd)
curve_distances <- max(curve_depths) - curve_depths  # convert depth to distance

# Per-curve p-value: fraction of bootstrap maxima >= curve's distance
p_values <- sapply(curve_distances, function(d) {
  mean(lrt_dist$boot.distribution >= d)
})

cat("P-values for curves 1-3 (true outliers):", round(p_values[1:3], 4), "\n")
cat("P-values for curves 4-6 (normal):", round(p_values[4:6], 4), "\n")
```

### LRT Results

```{r lrt-results}
# Outlier indices
out_lrt$outliers

# Distance from center for each curve
head(out_lrt$distances)

# Threshold used
out_lrt$threshold

# Bootstrap-calibrated p-values (smaller = more outlying)
cat("P-values for detected outliers:",
    round(out_lrt$p.value[out_lrt$outliers], 4), "\n")

# Bootstrap null distribution (sorted max distances from B bootstrap samples)
cat("Bootstrap null distribution: n =", length(out_lrt$boot_dist),
    ", range = [", round(min(out_lrt$boot_dist), 3), ",",
    round(max(out_lrt$boot_dist), 3), "]\n")
```

### When LRT Works Best

The LRT method is specifically optimized for **magnitude outliers** - curves that are shifted up or down relative to the main data cloud. It computes how far each curve is from the center (mean) of the data.

**What LRT detects well:**
- Curves shifted up or down (magnitude outliers)
- Curves with unusual overall level

**What LRT may miss:**
- Shape outliers (different pattern but similar overall level)
- Amplitude outliers (scaled versions centered at the same level)

**Using the threshold (`outliers.thres.lrt()`):**

The threshold represents the critical value of the LRT statistic. Use it to:

- Apply a custom significance level
- Compare test statistics across different datasets
- Combine with domain knowledge for decision-making

If LRT detects no outliers when you expect some:
1. The outliers may be shape-based rather than magnitude-based
2. Try depth-based methods (`outliers.depth.pond` or `outliers.depth.trim`) instead
3. Use the outliergram or MS-plot for visual detection

## Comparing Methods

Different methods may detect different types of outliers:

```{r compare-methods}
# Run all methods
out1 <- outliers.depth.pond(fd, nb = 1000, seed = 123)
out2 <- outliers.depth.trim(fd, trim = 0.1, seed = 123)
out3 <- outliers.lrt(fd, nb = 1000, seed = 123)

# Compare detected outliers
cat("Depth-pond outliers:", out1$outliers, "\n")
cat("Depth-trim outliers:", out2$outliers, "\n")
cat("LRT outliers:", out3$outliers, "\n")

# True outliers are curves 1, 2, 3
cat("True outliers: 1, 2, 3\n")
```

## Quantifying Outlierness

Binary outlier/not labels don't tell the full story. Every method in **fdars**
now returns a continuous **outlierness** score and a **p-value** for each curve.

```{r outlierness-scores}
# Run three methods on the same data
out_pond <- outliers.depth.pond(fd, nb = 1000, seed = 123)
ms <- magnitudeshape(fd)
og <- outliergram(fd)

# Each method now has p.value and outlierness fields
cat("depth.pond p-values (curves 1-3):", round(out_pond$p.value[1:3], 4), "\n")
cat("magnitudeshape p-values (curves 1-3):", round(ms$p.value[1:3], 4), "\n")
cat("outliergram p-values (curves 1-3):", round(og$p.value[1:3], 4), "\n")
```

The `outlier_summary()` function provides a unified ranked view of any
outlier detection result:

```{r outlier-summary}
# Ranked summary from depth.pond
summary_pond <- outlier_summary(out_pond)
head(summary_pond, 10)
```

```{r compare-outlierness, fig.height=5}
# Compare how different methods rank curves
rankings <- data.frame(
  curve = 1:n,
  pond = out_pond$outlierness,
  ms = ms$outlierness,
  outliergram = og$outlierness,
  type = ifelse(1:n %in% 1:3, "Outlier", "Normal")
)

# Reshape for plotting
library(ggplot2)
df_rank <- data.frame(
  curve = rep(1:n, 3),
  outlierness = c(rankings$pond, rankings$ms, rankings$outliergram),
  method = rep(c("depth.pond", "magnitudeshape", "outliergram"), each = n),
  type = rep(rankings$type, 3)
)

ggplot(df_rank, aes(x = factor(curve), y = outlierness, fill = type)) +
  geom_col(alpha = 0.7) +
  facet_wrap(~method, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("Normal" = "gray60", "Outlier" = "red")) +
  labs(title = "Outlierness Scores by Method",
       x = "Curve Index", y = "Outlierness", fill = "") +
  theme(axis.text.x = element_text(size = 6))
```

## Statistical Significance

Each method now provides per-curve p-values. These allow a formal hypothesis
test: "Is this curve an outlier at significance level $\alpha$?"

### Chi-squared P-values (magnitudeshape)

The magnitude-shape method uses a chi-squared test with 2 degrees of freedom.
The squared distance `dist_sq` follows a $\chi^2_2$ distribution under the null:

```{r chisq-pvalues}
ms <- magnitudeshape(fd)

# Show p-values for all curves
df_pval <- data.frame(
  curve = 1:n,
  dist_sq = ms$dist_sq,
  p.value = ms$p.value,
  is_outlier = ifelse(1:n %in% ms$outliers, "Outlier", "Normal")
)

# Curves 1-3 are true outliers
df_pval[c(1:3, 4:6), ]
```

### Bootstrap-calibrated P-values (LRT)

The LRT method provides **bootstrap-calibrated p-values**: each curve's
distance is compared against the null distribution of maximum distances
from B bootstrap samples. This gives a proper calibration — the p-value
answers: "what fraction of bootstrap null samples produced a maximum
distance at least as extreme as this curve's distance?"

```{r bootstrap-pvalues}
out_lrt <- outliers.lrt(fd, nb = 1000, seed = 123)

cat("LRT p-values for true outliers (1-3):",
    round(out_lrt$p.value[1:3], 4), "\n")
cat("LRT p-values for normal curves (4-6):",
    round(out_lrt$p.value[4:6], 4), "\n")
cat("Bootstrap null distribution size:", length(out_lrt$boot_dist), "\n")
```

### Rank-based P-values (depth methods)

For depth-based methods (`outliers.depth.pond`, `outliers.depth.trim`),
p-values are rank-based: a curve with the lowest depth gets the smallest
p-value.

### Multiple Testing Correction

When testing many curves simultaneously, apply FDR correction:

```{r fdr-correction}
# Adjust p-values for multiple comparisons
p_adjusted <- p.adjust(ms$p.value, method = "BH")

# Compare raw vs adjusted
cat("Raw p-values (curves 1-3):", round(ms$p.value[1:3], 4), "\n")
cat("BH-adjusted (curves 1-3):", round(p_adjusted[1:3], 4), "\n")

# Decision at alpha = 0.05
sig_raw <- which(ms$p.value < 0.05)
sig_adj <- which(p_adjusted < 0.05)
cat("Significant (raw):", sig_raw, "\n")
cat("Significant (BH-adjusted):", sig_adj, "\n")
```

## Detection Sensitivity and the Role of Noise

How strong must an outlier be before a method reliably detects it? The
answer depends not just on the **signal strength** (shift, frequency change,
or scale factor) but also on the **noise level**. A shift of 2 units is
obvious at $\sigma = 0.2$ but invisible at $\sigma = 2.0$. What matters is
the **signal-to-noise ratio (SNR)**.

To quantify this properly, we use a **power analysis**: for each
combination of signal strength and noise level, we run 20 independent
replications and record the **detection rate** — the fraction of
replications where the injected outlier (curve 1) is flagged. A detection
rate of 80% or higher means the method reliably catches that anomaly;
below 50% the method is essentially guessing.

```{r sensitivity-setup}
n_test <- 30
m_test <- 100
t_test <- seq(0, 1, length.out = m_test)
n_rep <- 20  # replications per grid point
noise_sds <- c(0.2, 0.5, 0.75, 1.0, 1.5)  # shared across all grids

# Helper: compute detection rate over n_rep replications
detection_rate <- function(shifts, noise_sds, inject_fn, detect_fn) {
  grid <- expand.grid(signal = shifts, noise_sd = noise_sds)
  grid$rate <- NA_real_
  for (r in seq_len(nrow(grid))) {
    sig <- grid$signal[r]
    sd_val <- grid$noise_sd[r]
    n_det <- 0L
    for (rep_i in seq_len(n_rep)) {
      set.seed(rep_i * 1000L)
      X <- matrix(0, n_test, m_test)
      for (i in seq_len(n_test))
        X[i, ] <- sin(2 * pi * t_test) + rnorm(m_test, sd = sd_val)
      X[1, ] <- inject_fn(sig, sd_val)
      fd <- fdata(X, argvals = t_test)
      if (detect_fn(fd, rep_i)) n_det <- n_det + 1L
    }
    grid$rate[r] <- n_det / n_rep
  }
  grid
}

# Helper: compute rect boundaries for non-uniform grids (no white gaps)
add_rect_bounds <- function(df, x_col, y_col) {
  midbreaks <- function(v) {
    u <- sort(unique(v))
    mids <- (u[-1] + u[-length(u)]) / 2
    c(u[1] - (mids[1] - u[1]), mids,
      u[length(u)] + (u[length(u)] - mids[length(mids)]))
  }
  xb <- midbreaks(df[[x_col]])
  yb <- midbreaks(df[[y_col]])
  xs <- sort(unique(df[[x_col]]))
  ys <- sort(unique(df[[y_col]]))
  xi <- match(df[[x_col]], xs)
  yi <- match(df[[y_col]], ys)
  df$xmin <- xb[xi];  df$xmax <- xb[xi + 1]
  df$ymin <- yb[yi];  df$ymax <- yb[yi + 1]
  df
}
```

### Magnitude Detection Power

We inject a magnitude outlier (vertical shift) and test with `outliers.lrt`.
The LRT uses a likelihood ratio test with bootstrap calibration, giving
well-calibrated detection across a wide range of signal-to-noise ratios.

```{r magnitude-power, fig.height=5}
mag_shifts <- c(0, 0.25, 0.5, 0.75, 1.0, 1.5, 2.0, 3.0, 4.0)

mag_grid <- detection_rate(
  mag_shifts, noise_sds,
  inject_fn = function(sig, sd_val) sin(2 * pi * t_test) + sig,
  detect_fn = function(fd, rep_i) {
    res <- outliers.lrt(fd, nb = 200, seed = rep_i)
    1L %in% res$outliers
  }
)

mag_grid <- add_rect_bounds(mag_grid, "signal", "noise_sd")

ggplot(mag_grid, aes(fill = rate)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.5,
               color = "white", linewidth = 1, linetype = "dashed") +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.8,
               color = "white", linewidth = 1.2) +
  scale_fill_viridis_c(option = "D", limits = c(0, 1),
                       labels = scales::percent,
                       name = "Detection\nRate") +
  labs(title = "Magnitude Outlier: Detection Power (LRT)",
       subtitle = "Solid contour = 80% power, dashed = 50%",
       x = "Shift Magnitude", y = "Noise SD") +
  theme(legend.position = "right")
```

The **detection frontier** (80% contour) follows a diagonal, confirming that
the shift-to-noise ratio — not shift or noise alone — determines
detectability.

### Shape Detection Power

We inject a shape outlier (altered frequency) and test with `outliergram`.
The outliergram detects curves with anomalous relationships between
modified band depth and modified epigraph index — a rank-based approach.

```{r shape-power, fig.height=5}
shape_freqs <- c(1.0, 1.5, 2.0, 2.5, 3.0, 4.0)

shape_grid <- detection_rate(
  shape_freqs, noise_sds,
  inject_fn = function(sig, sd_val) sin(sig * 2 * pi * t_test),
  detect_fn = function(fd, rep_i) {
    res <- outliergram(fd)
    1L %in% res$outliers
  }
)

shape_grid <- add_rect_bounds(shape_grid, "signal", "noise_sd")

ggplot(shape_grid, aes(fill = rate)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.5,
               color = "white", linewidth = 1, linetype = "dashed") +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.8,
               color = "white", linewidth = 1.2) +
  scale_fill_viridis_c(option = "D", limits = c(0, 1),
                       labels = scales::percent,
                       name = "Detection\nRate") +
  labs(title = "Shape Outlier: Detection Power (Outliergram)",
       subtitle = "Solid contour = 80% power, dashed = 50%",
       x = "Frequency Multiplier", y = "Noise SD") +
  theme(legend.position = "right")
```

Unlike magnitude detection, the shape detection boundary is nearly
**horizontal**: when noise exceeds the signal amplitude ($\sigma \gtrsim 0.7$
for a unit-amplitude sine), rank ordering is destroyed and detection power
collapses. Below that threshold, any frequency multiplier $\geq 1.5$
is caught.

### Amplitude Detection Power

We inject an amplitude outlier (scaled curve) and test with `outliers.lrt`.

```{r amplitude-power, fig.height=5}
amp_scales <- c(1.0, 1.25, 1.5, 2.0, 2.5, 3.0, 4.0)

amp_grid <- detection_rate(
  amp_scales, noise_sds,
  inject_fn = function(sig, sd_val) sig * sin(2 * pi * t_test),
  detect_fn = function(fd, rep_i) {
    res <- outliers.lrt(fd, nb = 200, seed = rep_i)
    1L %in% res$outliers
  }
)

amp_grid <- add_rect_bounds(amp_grid, "signal", "noise_sd")

ggplot(amp_grid, aes(fill = rate)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.5,
               color = "white", linewidth = 1, linetype = "dashed") +
  geom_contour(aes(x = signal, y = noise_sd, z = rate), breaks = 0.8,
               color = "white", linewidth = 1.2) +
  scale_fill_viridis_c(option = "D", limits = c(0, 1),
                       labels = scales::percent,
                       name = "Detection\nRate") +
  labs(title = "Amplitude Outlier: Detection Power (LRT)",
       subtitle = "Solid contour = 80% power, dashed = 50%",
       x = "Scale Factor", y = "Noise SD") +
  theme(legend.position = "right")
```

### Minimum Detectable Effect

For each noise level and outlier type, the smallest signal achieving
$\geq$ 80% detection rate:

```{r min-detectable, results='asis'}
mde_at <- function(grid, threshold = 0.8) {
  noise_levels <- sort(unique(grid$noise_sd))
  max_sig <- max(grid$signal)
  do.call(rbind, lapply(noise_levels, function(sd_val) {
    sub <- grid[grid$noise_sd == sd_val & grid$signal > min(grid$signal), ]
    det <- sub$signal[sub$rate >= threshold]
    val <- if (length(det) > 0) as.character(min(det))
           else paste0(">", max_sig)
    data.frame(noise_sd = sd_val, min_signal = val,
               stringsAsFactors = FALSE)
  }))
}

mde_mag   <- mde_at(mag_grid)
mde_shape <- mde_at(shape_grid)
mde_amp   <- mde_at(amp_grid)

mde_table <- data.frame(
  `Noise SD` = noise_sds,
  `Min Shift` = mde_mag$min_signal,
  `Min Freq Mult` = mde_shape$min_signal,
  `Min Scale` = mde_amp$min_signal,
  check.names = FALSE
)
knitr::kable(mde_table,
             caption = "Minimum detectable effect (80% power, n = 30 curves)")
```

### Practical Rules of Thumb

- **Magnitude outliers** (LRT): the detection frontier follows a diagonal —
  roughly shift $\geq 1.0 \times \sigma$ is needed. At $\sigma = 0.5$,
  you need a shift of ~1; at $\sigma = 2$, you need ~3–4.
- **Shape outliers** (outliergram): detection is noise-robust up to a
  threshold. When noise SD approaches the signal amplitude
  ($\sigma \gtrsim 0.8$ for a unit sine), rank ordering collapses and no
  frequency change is detectable. Below that, even 50% frequency changes
  are caught.
- **Amplitude outliers** (LRT): similar diagonal pattern to magnitude —
  scale factor $\geq 1.5$ is detectable at $\sigma = 0.5$, but at
  $\sigma = 1.0$ you need 3x scaling.
- **In practice**: estimate your noise level from the data (e.g., pointwise
  SD of residuals after removing the mean curve), then consult the table
  above to determine whether your expected anomaly size is detectable.

## Process Monitoring Decision Framework

### Method Selection

| Question | Recommended Method | Decision Rule |
|----------|-------------------|---------------|
| Is the curve shifted? | `outliers.lrt` | `p.value < 0.05` |
| Is the shape unusual? | `outliergram` | `p.value < 0.05` |
| Both types? | `magnitudeshape` | `p.value < 0.05` (chi-sq) |
| How extreme is it? | Any method | `outlierness` score |
| Is detection sensitive enough? | Sensitivity analysis | Min detectable severity |

### Practical Guidance

**Multiple testing**: When screening many curves, apply Benjamini-Hochberg
correction to control the false discovery rate:

```r
p_adjusted <- p.adjust(result$p.value, method = "BH")
outliers_fdr <- which(p_adjusted < 0.05)
```

**Bootstrap precision**: For the LRT method, use `nb >= 999` for stable
p-values at $\alpha = 0.05$. The rule of thumb is $nb \geq 1/\alpha - 1$.

**Combining methods**: When using multiple methods on the same data, consider
Bonferroni correction (`alpha / n_methods`) or use the magnitudeshape method
which jointly tests magnitude and shape.

**Streaming monitoring**: For real-time applications, use `streaming_depth_one()`
with depth-based p-values to flag anomalies as they arrive. See the
[streaming depth article](streaming-depth.html) for details.

## Types of Outliers

### Magnitude Outliers

Curves shifted up or down from the main group:

```{r magnitude-outlier, fig.height=4}
# Create clean data with just a magnitude outlier
X_mag <- matrix(0, n, m)
for (i in 1:n) {
  X_mag[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)
}
X_mag[1, ] <- sin(2 * pi * t_grid) + 3  # Large vertical shift

fd_mag <- fdata(X_mag, argvals = t_grid)

# Visualize the magnitude outlier
plot(fd_mag) +
  labs(title = "Magnitude Outlier: Curve 1 Shifted Up",
       subtitle = "Same shape as others, but at a different level")
```

```{r magnitude-detect}
out_mag <- outliers.depth.pond(fd_mag, nb = 1000, seed = 123)
cat("Detected magnitude outlier:", out_mag$outliers, "\n")
```

### Shape Outliers

Curves with different patterns but similar overall level:

```{r shape-outlier, fig.height=4}
# Create clean data with just a shape outlier
X_shape <- matrix(0, n, m)
for (i in 1:n) {
  X_shape[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)
}
X_shape[1, ] <- -sin(2 * pi * t_grid)  # Inverted (phase-shifted by pi)

fd_shape <- fdata(X_shape, argvals = t_grid)

# Visualize the shape outlier
plot(fd_shape) +
  labs(title = "Shape Outlier: Curve 1 Has Inverted Pattern",
       subtitle = "Same amplitude and level, but opposite phase")
```

```{r shape-detect}
out_shape <- outliers.depth.pond(fd_shape, nb = 1000, seed = 123)
cat("Detected shape outlier:", out_shape$outliers, "\n")
```

### Amplitude Outliers

Curves with unusual amplitude (larger or smaller scale):

```{r amplitude-outlier, fig.height=4}
# Create clean data with just an amplitude outlier
X_amp <- matrix(0, n, m)
for (i in 1:n) {
  X_amp[i, ] <- sin(2 * pi * t_grid) + rnorm(m, sd = 0.1)
}
X_amp[1, ] <- 3 * sin(2 * pi * t_grid)  # 3x larger amplitude

fd_amp <- fdata(X_amp, argvals = t_grid)

# Visualize the amplitude outlier
plot(fd_amp) +
  labs(title = "Amplitude Outlier: Curve 1 Has 3x Larger Scale",
       subtitle = "Same shape and center, but much larger oscillations")
```

```{r amplitude-detect}
out_amp <- outliers.depth.pond(fd_amp, nb = 1000, seed = 123)
cat("Detected amplitude outlier:", out_amp$outliers, "\n")
```

## Tuning Parameters

### Number of Bootstrap Samples

More bootstrap samples give more stable results but take longer:

```{r tuning-nb}
# Compare with different nb values
out_nb50 <- outliers.depth.pond(fd, nb = 50, seed = 123)
out_nb200 <- outliers.depth.pond(fd, nb = 200, seed = 123)

cat("nb=50 outliers:", out_nb50$outliers, "\n")
cat("nb=200 outliers:", out_nb200$outliers, "\n")
```

### Trim Proportion

For `outliers.depth.trim`, the trim proportion controls sensitivity:

```{r tuning-trim}
# More aggressive trimming
out_trim05 <- outliers.depth.trim(fd, trim = 0.05, seed = 123)
out_trim20 <- outliers.depth.trim(fd, trim = 0.2, seed = 123)

cat("trim=0.05 outliers:", out_trim05$outliers, "\n")
cat("trim=0.20 outliers:", out_trim20$outliers, "\n")
```

## Handling High Contamination

When outlier contamination is high, use robust methods:

```{r high-contamination}
# Create data with 20% outliers
X_contam <- X
n_outliers <- 6
for (i in 1:n_outliers) {
  X_contam[i, ] <- sin(2 * pi * t_grid) + rnorm(1, 0, 2)
}

fd_contam <- fdata(X_contam, argvals = t_grid)

# Depth-trim with higher trim proportion
out_contam <- outliers.depth.trim(fd_contam, trim = 0.2, seed = 123)
cat("Detected outliers:", out_contam$outliers, "\n")
cat("True outliers: 1-6\n")
```

## Visualizing Depth Distribution

```{r depth-distribution}
# Compute depths
depths <- depth.FM(fd)

# Create histogram
library(ggplot2)
df_depths <- data.frame(
  curve = 1:n,
  depth = depths,
  type = ifelse(1:n %in% c(1, 2, 3), "Outlier", "Normal")
)

ggplot(df_depths, aes(x = depth, fill = type)) +
  geom_histogram(bins = 10, alpha = 0.7, position = "identity") +
  labs(title = "Distribution of Functional Depths",
       x = "Depth", y = "Count", fill = "") +
  theme_minimal()
```

## Performance

The LRT method uses a parallel Rust backend for speed:

```{r performance, eval=FALSE}
# Benchmark with larger dataset
X_large <- matrix(rnorm(200 * 100), 200, 100)
fd_large <- fdata(X_large)

system.time(outliers.lrt(fd_large, nb = 200, seed = 123))
#>    user  system elapsed
#>   0.456   0.000   0.123
```

## Outliergram and MS-Plot

For visual outlier detection, **fdars** provides two powerful diagnostic plots.

### The Outliergram

The outliergram plots the **Modified Epigraph Index (MEI)** against **Modified Band Depth (MBD)**:

```{r outliergram}
og <- outliergram(fd)
plot(og)
```

**How to read the outliergram:**

| Position | MEI (X-axis) | MBD (Y-axis) | Interpretation |
|----------|--------------|--------------|----------------|
| Bottom-left | Low | Low | **Extreme outlier** (unusual shape AND position) |
| Bottom-right | High | Low | **Magnitude outlier** (shifted up/down) |
| Top-left | Low | High | **Shape outlier** (unusual pattern, typical level) |
| Top-right | High | High | **Normal curve** (typical shape and position) |

The parabolic boundary marks the theoretical limit for non-outlying curves. Points below this boundary are flagged as outliers.

### The Magnitude-Shape Plot (MS-Plot)

The MS-plot separates magnitude outlyingness from shape outlyingness:

```{r msplot}
ms <- magnitudeshape(fd)
plot(ms)
```

**How to read the MS-plot:**

| Quadrant | Magnitude Outlyingness | Shape Outlyingness | Type |
|----------|------------------------|-------------------|------|
| Bottom-left | Low | Low | **Normal curve** |
| Bottom-right | High | Low | **Magnitude outlier** only |
| Top-left | Low | High | **Shape outlier** only |
| Top-right | High | High | **Combined outlier** (both types) |

The MS-plot is particularly useful when you want to understand *why* a curve is an outlier - is it because of its level (magnitude) or its pattern (shape)?

### Labeling Outliers by ID or Metadata

When fdata has IDs or metadata, you can label outliers in plots:
```{r outlier-labeling}
# Create fdata with IDs and metadata
meta <- data.frame(
  subject = paste0("S", sprintf("%02d", 1:n)),
  group = rep(c("A", "B"), length.out = n)
)
fd_labeled <- fdata(X, argvals = t_grid,
                    id = paste0("patient_", 1:n),
                    metadata = meta)

# Outliergram with patient IDs
og_labeled <- outliergram(fd_labeled)
plot(og_labeled, label = "id")

# Or with metadata column
plot(og_labeled, label = "subject")

# Label ALL points, not just outliers
plot(og_labeled, label = "id", label_all = TRUE)
```

```{r ms-labeling}
# magnitudeshape with custom labels
magnitudeshape(fd_labeled, label = "subject")
```

## Method Selection Guide

| Method | Best For | Sensitivity |
|--------|----------|-------------|
| depth.pond | General purpose | Moderate |
| depth.trim | Known contamination rate | Controllable |
| LRT | Magnitude outliers | High |
| outliergram | Shape outliers | Visual |
| magnitudeshape | Both magnitude & shape | Visual |

## Best Practices

1. **Start with visualization**: Plot the data to understand outlier types
2. **Try multiple methods**: Different methods catch different outliers
3. **Use sufficient bootstrap samples**: At least 100 for stable results
4. **Consider domain knowledge**: Some "outliers" may be valid observations
5. **Validate findings**: Check detected outliers make sense contextually

## See Also

- `vignette("depth-functions", package = "fdars")` — functional depth measures and ranking
- `vignette("streaming-depth", package = "fdars")` — real-time depth computation for streaming data
- `vignette("clustering", package = "fdars")` — functional data clustering

## References

- Febrero, M., Galeano, P., and González-Manteiga, W. (2008). Outlier
  detection in functional data by depth measures, with application to
  identify abnormal NOx levels. *Environmetrics*, 19(4), 331-345.
- Hyndman, R.J. and Shang, H.L. (2010). Rainbow plots, bagplots, and
  boxplots for functional data. *Journal of Computational and Graphical
  Statistics*, 19(1), 29-45.
