---
title: "Profile and Partial-Domain Monitoring"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

Standard functional SPM (see `vignette("articles/statistical-process-monitoring")`)
monitors curves assuming they are fully observed and independent of external
covariates. Two common situations break these assumptions:

1. **Profile monitoring** — The functional response depends on known scalar
   covariates (e.g., catalyst concentration affects a reactor temperature
   profile). Ignoring covariates inflates false alarm rates because
   covariate-driven variation is mistaken for process shifts.

2. **Partial-domain monitoring** — Curves arrive incrementally: at time $t$
   during a batch, only the portion $[0, t]$ has been observed. Waiting for
   the full curve delays detection; monitoring the partial curve requires
   completing the unobserved tail.

This article demonstrates both capabilities, plus elastic SPM for processes
where phase (timing) variation is present.

## Simulating Process Data

We simulate a chemical reactor scenario. Each batch produces a temperature
profile over a normalised time domain $[0, 1]$. A scalar covariate — catalyst
concentration — modulates the profile shape. Phase II data includes 10
out-of-control batches at the end.

```{r simulate}
set.seed(42)
n_train <- 100
n_test <- 40
m <- 50
argvals <- seq(0, 1, length.out = m)

# Covariate: catalyst concentration
catalyst_train <- runif(n_train, 0.5, 1.5)
catalyst_test <- runif(n_test, 0.5, 1.5)

# Response curves with covariate effect
Y_train <- matrix(0, n_train, m)
for (i in 1:n_train) {
  Y_train[i, ] <- catalyst_train[i] * sin(2 * pi * argvals) +
                  rnorm(m, sd = 0.2)
}
fd_train <- fdata(Y_train, argvals = argvals)
predictors_train <- matrix(catalyst_train, ncol = 1)

# Phase II: last 10 curves have a shifted covariate-response relationship
Y_test <- matrix(0, n_test, m)
for (i in 1:n_test) {
  shift <- if (i > 30) 1.2 * cos(4 * pi * argvals) else 0
  Y_test[i, ] <- catalyst_test[i] * sin(2 * pi * argvals) +
                 shift + rnorm(m, sd = 0.2)
}
fd_test <- fdata(Y_test, argvals = argvals)
predictors_test <- matrix(catalyst_test, ncol = 1)
```

```{r plot-data}
plot(fd_train, main = "Phase I: Reactor Temperature Profiles",
     xlab = "Normalised Time", ylab = "Temperature")
```

## Profile Monitoring

Profile monitoring uses function-on-scalar regression (FOSR) to model how the
covariate influences the curve. Sliding-window FOSR estimates capture
time-varying regression coefficients. FPCA on these coefficient trajectories
gives a reduced representation; $T^2$ statistics on the FPC scores detect
shifts in the covariate-response relationship.

### Phase I

```{r profile-phase1}
profile_chart <- spm.profile.phase1(
  fd_train, predictors_train,
  ncomp = 3, window.size = 20, alpha = 0.05
)
print(profile_chart)
```

### Phase II

```{r profile-monitor}
profile_mon <- spm.profile.monitor(profile_chart, fd_test, predictors_test)

alarm_idx <- which(profile_mon$t2.alarm)
cat("Alarms at windows:", alarm_idx, "\n")
```

```{r profile-plot}
# Residual T-squared control chart
n_mon <- length(profile_mon$t2)
df_profile <- data.frame(
  window = seq_len(n_mon),
  t2 = profile_mon$t2,
  alarm = profile_mon$t2.alarm
)

ggplot(df_profile, aes(x = window, y = t2)) +
  geom_segment(aes(xend = window, y = 0, yend = t2, color = alarm),
               linewidth = 0.8) +
  geom_hline(yintercept = profile_mon$t2.ucl,
             linetype = "dashed", color = "#D55E00", linewidth = 0.7) +
  annotate("text", x = 1, y = profile_mon$t2.ucl,
           label = paste("UCL =", round(profile_mon$t2.ucl, 2)),
           hjust = -0.05, vjust = -0.5, size = 3, color = "#D55E00") +
  scale_color_manual(values = c("FALSE" = "#4A90D9", "TRUE" = "#D55E00"),
                     guide = "none") +
  labs(title = expression("Profile Monitoring: Residual T"^2 * " Chart"),
       x = "Monitoring Window", y = expression("T"^2)) +
  theme(plot.title = element_text(face = "bold"))
```

## Partial-Domain Monitoring

During a running batch, only a prefix of the temperature curve has been
observed. `spm.monitor.partial()` completes the unobserved tail before
computing monitoring statistics.

Three completion methods are available:

| Method | Description |
|--------|-------------|
| `"conditional"` | Conditional expectation (BLUP) given observed portion |
| `"projection"` | Project partial curve onto the FPCA basis directly |
| `"zero_pad"` | Pad unobserved portion with zeros (fast, less accurate) |

We first build a standard Phase I chart (no covariates), then monitor a
single curve at several observation fractions.

```{r partial-setup}
# Phase I chart (standard, no covariates)
chart_ic <- spm.phase1(fd_train, ncomp = 3, alpha = 0.05, seed = 42)

# A test curve observed at increasing fractions
test_curve <- Y_test[35, ]  # an out-of-control curve
fractions <- c(0.4, 0.6, 0.8, 1.0)
n_obs_vals <- as.integer(fractions * m)
```

```{r partial-monitor}
methods <- c("conditional", "projection", "zero_pad")
partial_results <- list()

for (method in methods) {
  for (k in seq_along(fractions)) {
    result <- spm.monitor.partial(
      chart_ic, test_curve,
      n.observed = n_obs_vals[k],
      completion = method
    )
    partial_results[[length(partial_results) + 1]] <- data.frame(
      method = method,
      fraction = fractions[k],
      t2 = result$t2,
      alarm = result$t2.alarm
    )
  }
}
df_partial <- do.call(rbind, partial_results)
```

```{r partial-plot, fig.height = 4}
ggplot(df_partial, aes(x = fraction, y = t2, color = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(shape = alarm), size = 3) +
  geom_hline(yintercept = chart_ic$t2.ucl,
             linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.4, y = chart_ic$t2.ucl,
           label = "UCL", hjust = -0.1, vjust = -0.5,
           size = 3, color = "grey40") +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("In control", "Alarm")) +
  labs(title = expression("Partial-Domain T"^2 * " by Completion Method"),
       x = "Domain Fraction Observed",
       y = expression("T"^2),
       color = "Completion", shape = "Status") +
  theme(legend.position = "bottom")
```

The conditional method generally produces the most reliable statistics at low
observation fractions because it leverages the full covariance structure
learned in Phase I.

## Batch Processing

When multiple partially-observed curves are available (e.g., several batches
at different stages of completion), `spm.monitor.partial.batch()` processes
them in a single call.

```{r batch-setup}
# Simulate 8 batches at varying completion stages
set.seed(123)
n_batch <- 8
obs_fractions <- seq(0.3, 1.0, length.out = n_batch)
n_obs_batch <- as.integer(obs_fractions * m)

batch_curves <- vector("list", n_batch)
for (i in 1:n_batch) {
  # Last 3 batches are out-of-control
  shift <- if (i > 5) 1.5 * cos(4 * pi * argvals) else 0
  batch_curves[[i]] <- sin(2 * pi * argvals) + shift + rnorm(m, sd = 0.2)
}
```

```{r batch-monitor}
batch_results <- spm.monitor.partial.batch(
  chart_ic, batch_curves, n_obs_batch,
  completion = "conditional"
)

df_batch <- data.frame(
  batch = seq_len(n_batch),
  fraction = obs_fractions,
  t2 = sapply(batch_results, function(r) r$t2),
  alarm = sapply(batch_results, function(r) r$t2.alarm)
)
print(df_batch)
```

```{r batch-plot, fig.height = 4}
ggplot(df_batch, aes(x = batch, y = t2)) +
  geom_segment(aes(xend = batch, y = 0, yend = t2, color = alarm),
               linewidth = 1) +
  geom_point(aes(color = alarm), size = 3) +
  geom_hline(yintercept = chart_ic$t2.ucl,
             linetype = "dashed", color = "#D55E00") +
  geom_text(aes(label = paste0(round(fraction * 100), "%")),
            vjust = -1, size = 2.8) +
  scale_color_manual(values = c("FALSE" = "#4A90D9", "TRUE" = "#D55E00"),
                     labels = c("In control", "Alarm"), name = "Status") +
  scale_x_continuous(breaks = seq_len(n_batch)) +
  labs(title = expression("Batch Monitoring: T"^2 * " with Partial Observations"),
       subtitle = "Labels show percentage of domain observed",
       x = "Batch", y = expression("T"^2)) +
  theme(legend.position = "bottom")
```

## Elastic SPM

When process timing varies between batches (e.g., a reaction reaches peak
temperature earlier or later), standard SPM conflates amplitude changes
(how high the peak is) with phase changes (when it occurs). Elastic SPM
separates these sources of variation using elastic alignment to a Karcher
mean.

### Phase I

```{r elastic-phase1}
# Simulate curves with phase variability
set.seed(42)
n_el <- 60
m_el <- 50
argvals_el <- seq(0, 1, length.out = m_el)

X_elastic <- matrix(0, n_el, m_el)
for (i in 1:n_el) {
  phase_shift <- runif(1, -0.08, 0.08)
  amp <- rnorm(1, mean = 1, sd = 0.15)
  X_elastic[i, ] <- amp * sin(2 * pi * (argvals_el + phase_shift)) +
                    rnorm(m_el, sd = 0.1)
}
fd_elastic <- fdata(X_elastic, argvals = argvals_el)

elastic_chart <- spm.elastic.phase1(
  fd_elastic, ncomp = 3, warp.ncomp = 2,
  alpha = 0.05, seed = 42
)
print(elastic_chart)
```

### Phase II

```{r elastic-monitor}
# New data: 15 in-control, then 5 with amplitude shift + phase distortion
set.seed(99)
n_new_el <- 20
X_new_el <- matrix(0, n_new_el, m_el)
for (i in 1:n_new_el) {
  phase_shift <- runif(1, -0.08, 0.08)
  amp <- rnorm(1, mean = 1, sd = 0.15)
  if (i > 15) {
    amp <- amp + 1.5           # amplitude fault
    phase_shift <- phase_shift + 0.15  # phase fault
  }
  X_new_el[i, ] <- amp * sin(2 * pi * (argvals_el + phase_shift)) +
                   rnorm(m_el, sd = 0.1)
}
fd_new_el <- fdata(X_new_el, argvals = argvals_el)

elastic_mon <- spm.elastic.monitor(elastic_chart, fd_new_el)

cat("Amplitude T2 alarms:", sum(elastic_mon$amplitude$t2.alarm), "\n")
cat("Phase T2 alarms:", sum(elastic_mon$phase$t2.alarm), "\n")
```

```{r elastic-plot, fig.height = 6}
n_pts <- length(elastic_mon$amplitude$t2)

df_elastic <- data.frame(
  obs = rep(seq_len(n_pts), 4),
  value = c(
    elastic_mon$amplitude$t2, elastic_mon$amplitude$spe,
    elastic_mon$phase$t2, elastic_mon$phase$spe
  ),
  panel = factor(rep(
    c("Amplitude T\u00b2", "Amplitude SPE",
      "Phase T\u00b2", "Phase SPE"),
    each = n_pts
  ), levels = c("Amplitude T\u00b2", "Amplitude SPE",
                "Phase T\u00b2", "Phase SPE")),
  alarm = c(
    elastic_mon$amplitude$t2.alarm, elastic_mon$amplitude$spe.alarm,
    elastic_mon$phase$t2.alarm, elastic_mon$phase$spe.alarm
  )
)

ucl_df <- data.frame(
  panel = factor(
    c("Amplitude T\u00b2", "Amplitude SPE",
      "Phase T\u00b2", "Phase SPE"),
    levels = c("Amplitude T\u00b2", "Amplitude SPE",
               "Phase T\u00b2", "Phase SPE")
  ),
  ucl = c(
    elastic_chart$amplitude$t2.ucl, elastic_chart$amplitude$spe.ucl,
    elastic_chart$phase$t2.ucl, elastic_chart$phase$spe.ucl
  )
)

ggplot(df_elastic, aes(x = obs, y = value)) +
  geom_segment(aes(xend = obs, y = 0, yend = value, color = alarm),
               linewidth = 0.7) +
  geom_hline(data = ucl_df, aes(yintercept = ucl),
             linetype = "dashed", color = "#D55E00", linewidth = 0.6) +
  geom_vline(xintercept = 15.5, linetype = "dotted", color = "grey50") +
  annotate("text", x = 15.5, y = Inf, label = "Fault onset",
           hjust = 1.1, vjust = 1.5, size = 3, color = "grey50") +
  scale_color_manual(values = c("FALSE" = "#4A90D9", "TRUE" = "#D55E00"),
                     guide = "none") +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(title = "Elastic SPM: Amplitude and Phase Monitoring",
       x = "Observation", y = "Statistic") +
  theme(strip.text = element_text(face = "bold"))
```

The amplitude charts detect the magnitude shift, while the phase charts
detect the timing distortion. Separating these makes root-cause diagnosis
straightforward: an amplitude alarm points to process intensity changes,
a phase alarm points to timing or kinetics changes.

## See Also

- `vignette("articles/statistical-process-monitoring")` — Standard functional SPM
- `vignette("articles/fpca")` — Functional PCA fundamentals
- `vignette("articles/elastic-alignment")` — Elastic alignment background

## References

- Colosimo, B.M. and Pacella, M. (2010). A comparison study of control charts
  for statistical monitoring of functional data. *International Journal of
  Production Research*, 48(6), 1575--1601.
- Capezza, C., Lepore, A., Menafoglio, A., Palumbo, B., and Vantini, S.
  (2020). Control charts for monitoring ship operating conditions and CO2
  emissions based on scalar-on-function regression. *Applied Stochastic
  Models in Business and Industry*, 36(3), 477--500.
- Srivastava, A. and Klassen, E.P. (2016). *Functional and Shape Data
  Analysis*. Springer.
