---
title: "Regression Diagnostics"
description: >
  Diagnostic tools for functional regression models: influence analysis,
  DFBETAS, variance inflation, permutation importance, regression depth,
  and explanation stability.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

After fitting a functional regression model, it is essential to check whether
the assumptions hold, whether any observations unduly influence the fit, and
whether the results are stable. The fdars package provides seven diagnostic
tools adapted for FPC-based functional regression.

```{=html}
<img src="../reference/figures/regression-diagnostics-diagram.svg" alt="Regression Diagnostics: 6 panels showing influence, DFBETAS, importance, depth, residuals, VIF" style="width:100%; max-width:700px;">
```

### Supported Models

| Method | Function | `fregre.lm` | `functional.logistic` |
|:-------|:---------|:-----------:|:---------------------:|
| Influence (leverage, Cook's D) | `fregre.influence` | yes | |
| DFBETAS / DFFITS | `fregre.dfbetas` | yes | |
| VIF | `fregre.vif` | yes | yes |
| Permutation importance | `fregre.importance` | yes | yes |
| Conditional importance | `fregre.conditional.importance` | yes | yes |
| Regression depth | `fregre.depth` | yes | yes |
| Explanation stability | `fregre.stability` | *model-free* | *model-free* |

**Why only `fregre.lm` and `functional.logistic`?** All diagnostic functions
operate on the FPC scores and regression coefficients produced by
`fregre.lm()` or `functional.logistic()`. Other model types (`fregre.pc`,
`fregre.np`, `fregre.basis`) use different internal structures and are not
currently supported by the Rust diagnostics backend.

**Why not nonparametric models?** Nonparametric models like `fregre.np()` (kernel
regression) and KNN do not produce coefficients or an explicit hat matrix, so
leverage, Cook's distance, DFBETAS, VIF, and beta decomposition are undefined.
Permutation importance could in principle work for any model, but the current
implementation requires the FPC score representation from `fregre.lm`.

## Setup

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

### Simulated Data

We simulate functional data with a known coefficient function. In applied terms,
think of a manufacturing process where sensor curves predict a quality measure —
diagnostics help identify faulty sensors (influential observations) and unstable
model regions.

```{r create-data}
n <- 80
m <- 100
t_grid <- seq(0, 1, length.out = m)

beta_true <- 2 * dnorm(t_grid, 0.3, 0.08) - 1.5 * dnorm(t_grid, 0.7, 0.08)
beta_true <- beta_true / max(abs(beta_true))

X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) + 0.5 * rnorm(1) * cos(4 * pi * t_grid) +
            rnorm(m, sd = 0.1)
}

dt <- t_grid[2] - t_grid[1]
y <- numeric(n)
for (i in 1:n) {
  y[i] <- sum(beta_true * X[i, ]) * dt + rnorm(1, sd = 0.3)
}

fdataobj <- fdata(X, argvals = t_grid)
model <- fregre.lm(fdataobj, y, ncomp = 5)
```

---

## Influence and Case Deletion

These methods identify observations that disproportionately affect the
fitted model. Removing a single influential observation can substantially
change the coefficients, fitted values, and predictions.

**Supported models: `fregre.lm` only.** These methods require the hat matrix,
which is only defined for linear models.

```{=html}
<img src="../reference/figures/diag-influence.svg" alt="Influence diagnostics: leverage, Cook's D, DFBETAS, VIF" style="width:100%; max-width:700px;">
```

### Leverage and Cook's Distance

**In plain English:** Leverage measures how far an observation's FPC scores are
from the centre of the score space — think of it as "how unusual is this
curve's shape?" Cook's distance combines leverage with residual size to measure
the overall influence on the model. A point with high leverage *and* a large
residual is influential: removing it would noticeably change the regression
coefficients.

**Mathematically:** In the FPC regression $\hat{\boldsymbol{y}} = \boldsymbol{H}\boldsymbol{y}$, the hat matrix is

$$\boldsymbol{H} = \boldsymbol{\Xi}(\boldsymbol{\Xi}^\top\boldsymbol{\Xi})^{-1}\boldsymbol{\Xi}^\top,$$

where $\boldsymbol{\Xi}$ is the $n \times (K+1)$ augmented score matrix. The
diagonal $h_{ii}$ is the leverage of observation $i$. Observations with
$h_{ii} > 2(K+1)/n$ are flagged as high-leverage.

Cook's distance combines leverage and residual:

$$D_i = \frac{e_i^2}{(K+1)\,\hat{\sigma}^2} \cdot \frac{h_{ii}}{(1-h_{ii})^2}.$$

The classic threshold is $D_i > 4/n$. A high Cook's $D$ means the observation
pulls the fitted surface toward itself.

```{r influence}
influence <- fregre.influence(model, data = fdataobj)

df_inf <- data.frame(
  Observation = 1:n,
  Leverage = influence$leverage,
  CooksD = influence$cooks_distance
)
```

```{r plot-influence}
ggplot(df_inf, aes(x = .data$Leverage, y = .data$CooksD)) +
  geom_point(color = "#7B2D8E", size = 2, alpha = 0.7) +
  geom_text(data = subset(df_inf, CooksD > quantile(CooksD, 0.95)),
            aes(label = .data$Observation), hjust = -0.3, size = 3,
            color = "#D55E00") +
  geom_hline(yintercept = 4 / n, linetype = "dashed",
             color = "#D55E00", alpha = 0.5) +
  annotate("text", x = max(df_inf$Leverage) * 0.9, y = 4 / n + 0.001,
           label = "4/n threshold", color = "#D55E00", size = 3) +
  labs(title = "Influence Diagnostics",
       subtitle = "High leverage + high Cook's D = influential observations",
       x = "Leverage (Hat value)", y = "Cook's Distance")
```

#### Leverage Distribution

Observations above the dashed threshold have FPC scores far from the centroid.
In an industrial setting, these might be sensors operating under unusual
conditions.

```{r plot-leverage}
ggplot(df_inf, aes(x = .data$Observation, y = .data$Leverage)) +
  geom_segment(aes(xend = .data$Observation, yend = 0), color = "gray70") +
  geom_point(color = "#4A90D9", size = 2) +
  geom_hline(yintercept = 2 * (model$ncomp + 1) / n, linetype = "dashed",
             color = "#D55E00") +
  labs(title = "Leverage per Observation",
       subtitle = paste("Threshold: 2(K+1)/n =",
                        round(2 * (model$ncomp + 1) / n, 3)),
       x = "Observation", y = "Leverage")
```

### DFBETAS and DFFITS

**In plain English:** DFBETAS asks, for each observation and each regression
coefficient: "how much would this coefficient change if I removed this
observation?" DFFITS is similar but for the fitted value itself: "how much
would observation $i$'s prediction change if I refitted without it?"

Large DFBETAS values for a specific component indicate that the observation
is driving that component's regression coefficient. This helps pinpoint *which
aspect* of the model an influential point affects, not just that it matters.

**Mathematically:**

$$\text{DFBETAS}_{ij} = \frac{\hat{b}_j - \hat{b}_j^{(-i)}}{\hat{\sigma}^{(-i)}\sqrt{(\boldsymbol{\Xi}^\top\boldsymbol{\Xi})^{-1}_{jj}}},$$

$$\text{DFFITS}_i = \frac{\hat{y}_i - \hat{y}_i^{(-i)}}{\hat{\sigma}^{(-i)}\sqrt{h_{ii}}}.$$

The standard cutoff for DFBETAS is $2/\sqrt{n}$; for DFFITS it is
$2\sqrt{(K+1)/n}$.

```{r dfbetas}
dfb <- fregre.dfbetas(model, data = fdataobj)

cat("DFBETAS cutoff (2/sqrt(n)):", round(dfb$dfbetas_cutoff, 3), "\n")
cat("DFFITS cutoff:", round(dfb$dffits_cutoff, 3), "\n")
```

```{r plot-dffits}
df_dffits <- data.frame(
  Observation = seq_len(n),
  DFFITS = dfb$dffits
)

ggplot(df_dffits, aes(x = .data$Observation, y = abs(.data$DFFITS))) +
  geom_segment(aes(xend = .data$Observation, yend = 0), color = "gray70") +
  geom_point(color = "#7B2D8E", size = 2) +
  geom_hline(yintercept = dfb$dffits_cutoff, linetype = "dashed",
             color = "#D55E00") +
  labs(title = "DFFITS: Case Influence on Fitted Values",
       subtitle = paste("Cutoff:", round(dfb$dffits_cutoff, 3)),
       x = "Observation", y = "|DFFITS|")
```

```{r plot-dfbetas}
ncomp_show <- min(3, ncol(dfb$dfbetas))
df_dfb <- do.call(rbind, lapply(seq_len(ncomp_show), function(k) {
  data.frame(Observation = seq_len(n),
             DFBETAS = dfb$dfbetas[, k],
             Component = paste0("PC", k))
}))

ggplot(df_dfb, aes(x = .data$Observation, y = .data$DFBETAS)) +
  geom_point(color = "#4A90D9", size = 1.5, alpha = 0.6) +
  geom_hline(yintercept = c(-dfb$dfbetas_cutoff, dfb$dfbetas_cutoff),
             linetype = "dashed", color = "#D55E00") +
  facet_wrap(~ Component) +
  labs(title = "DFBETAS per FPC Component",
       subtitle = "Per-coefficient case influence — points beyond cutoff shift that coefficient",
       x = "Observation", y = "DFBETAS")
```

### Variance Inflation Factor

**In plain English:** VIF checks whether the FPC scores are collinear with each
other (or with scalar covariates). If two predictors carry the same information,
the regression coefficients become unstable. Since FPC scores are orthogonal by
construction, VIF should be near 1 for a pure FPCA model. VIF > 5 indicates a
problem — typically when scalar covariates correlate with FPC scores or when
the score matrix is numerically ill-conditioned.

**Mathematically:** The VIF for score $k$ is

$$\text{VIF}_k = \frac{1}{1 - R_k^2},$$

where $R_k^2$ is the $R^2$ from regressing $\xi_k$ on all other predictors.
VIF = 1 means no collinearity; VIF = 10 means 90% of the variance of $\xi_k$
is explained by other predictors.

**Supported models: `fregre.lm` + `functional.logistic`.**

```{r vif}
vif_scores <- fregre.vif(model, data = fdataobj)

df_vif <- data.frame(
  Component = paste0("PC", seq_along(vif_scores$vif)),
  VIF = vif_scores$vif
)

ggplot(df_vif, aes(x = .data$Component, y = .data$VIF)) +
  geom_col(fill = "#7B2D8E", alpha = 0.7) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "#D55E00") +
  annotate("text", x = length(vif_scores$vif), y = 5.3,
           label = "VIF = 5 threshold", color = "#D55E00", hjust = 1, size = 3) +
  labs(title = "Variance Inflation Factor",
       subtitle = paste("Mean VIF:", round(vif_scores$mean_vif, 2),
                        "— values near 1 confirm orthogonal FPC scores"),
       x = "Component", y = "VIF")
```

---

## Residual Diagnostics

Standard regression diagnostics check the model assumptions visually.
These plots use only the `fregre.lm` model's residuals and fitted values.

### Residuals vs Fitted

**In plain English:** If the model is well-specified, residuals should scatter
randomly around zero with constant spread. Patterns (curves, funnels) indicate
model misspecification: a curve means a nonlinear effect is missing; a funnel
means heteroscedasticity (variance changes with the fitted value). Point size
encodes leverage, so large distant points are the most concerning.

```{r residuals}
df_resid <- data.frame(
  Fitted = model$fitted.values,
  Residual = model$residuals,
  Leverage = influence$leverage
)

ggplot(df_resid, aes(x = .data$Fitted, y = .data$Residual)) +
  geom_point(aes(size = .data$Leverage), color = "#4A90D9", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00", linewidth = 0.8) +
  scale_size_continuous(range = c(1, 5)) +
  labs(title = "Residuals vs Fitted Values",
       subtitle = "Point size = leverage; orange line = loess smoother (should be flat at zero)",
       x = "Fitted Values", y = "Residuals")
```

### Normal Q-Q Plot

**In plain English:** If the errors are normally distributed, the points follow
the diagonal line. Deviations at the tails indicate heavy tails (outliers) or
skewness. Normality is assumed by prediction intervals (see the
[uncertainty quantification](uncertainty-quantification.html) article) but
not by conformal prediction.

```{r qq}
df_qq <- data.frame(Residual = model$residuals)

ggplot(df_qq, aes(sample = .data$Residual)) +
  stat_qq(color = "#4A90D9", size = 2, alpha = 0.7) +
  stat_qq_line(color = "#D55E00") +
  labs(title = "Normal Q-Q Plot of Residuals",
       subtitle = "Points on the line = Gaussian residuals; tails off = heavy tails / skewness",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

### Scale-Location Plot

**In plain English:** This plot checks for heteroscedasticity from a different
angle. The y-axis shows $\sqrt{|e_i|}$ against fitted values. If the spread
is constant, the smoother should be roughly flat.

```{r scale-location}
df_sl <- data.frame(
  Fitted = model$fitted.values,
  SqrtAbsResid = sqrt(abs(model$residuals))
)

ggplot(df_sl, aes(x = .data$Fitted, y = .data$SqrtAbsResid)) +
  geom_point(color = "#2E8B57", size = 2, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00", linewidth = 0.8) +
  labs(title = "Scale-Location Plot",
       subtitle = "Flat smoother = constant variance (homoscedasticity)",
       x = "Fitted Values", y = expression(sqrt("|Residual|")))
```

---

## Importance and Robustness

These methods assess which FPC scores matter and how stable the model is.

**Supported models: `fregre.lm` + `functional.logistic`** (except stability,
which is model-free).

```{=html}
<img src="../reference/figures/diag-robustness.svg" alt="Importance and robustness: permutation importance, conditional importance, regression depth, stability" style="width:100%; max-width:700px;">
```

### Permutation Importance

**In plain English:** For each FPC score, we randomly shuffle its values
(breaking the relationship with $y$) and measure how much the prediction
error increases. A large increase means the model relies heavily on that
score. This is a *model-agnostic* importance measure (works for any model
that predicts from FPC scores).

**Mathematically:** The importance of score $k$ is

$$\text{Imp}_k = \frac{1}{P}\sum_{p=1}^{P}\left[\text{MSE}(\hat{y};\,\tilde{\xi}_k^{(p)}) - \text{MSE}(\hat{y})\right],$$

where $\tilde{\xi}_k^{(p)}$ is the $p$-th random permutation of score $k$
and $P$ is the number of permutation replicates. The importance is always
non-negative for a useful predictor.

```{r importance}
imp <- fregre.importance(model, data = fdataobj, y = y, n.perm = 50, seed = 42)

df_imp <- data.frame(
  Component = paste0("PC", seq_along(imp$importance)),
  Importance = imp$importance
)

ggplot(df_imp, aes(x = reorder(.data$Component, -.data$Importance),
                   y = .data$Importance)) +
  geom_col(fill = "#4A90D9", alpha = 0.7) +
  labs(title = "Permutation Importance",
       subtitle = paste("Baseline MSE:", round(imp$baseline_metric, 4),
                        "— bars show MSE increase when shuffling each score"),
       x = "Component", y = "Importance (MSE increase)")
```

### Conditional Permutation Importance

**In plain English:** Standard permutation importance can overstate a score's
importance when predictors are correlated, because shuffling one score
implicitly breaks its correlation with other scores. Conditional permutation
fixes this by shuffling *within bins* of the other predictors, preserving the
correlation structure. The difference between unconditional and conditional
importance reveals how much of a score's apparent importance comes from
correlation rather than direct prediction.

**Mathematically:** Instead of permuting $\xi_k$ freely, we partition the other
scores $\boldsymbol{\xi}_{-k}$ into $B$ bins and permute $\xi_k$ only within
observations that share the same bin. This keeps the conditional distribution
$\xi_k \mid \boldsymbol{\xi}_{-k}$ approximately intact.

Since FPC scores are orthogonal by construction, conditional and unconditional
importance should agree closely here. When scalar covariates are included, they
may diverge.

```{r cond-importance}
cimp <- fregre.conditional.importance(model, data = fdataobj, y = y,
                                      n.bins = 5, n.perm = 50, seed = 42)

df_cimp <- data.frame(
  Component = paste0("PC", seq_along(cimp$importance)),
  Conditional = cimp$importance,
  Unconditional = cimp$unconditional_importance
)
df_cimp_long <- rbind(
  data.frame(Component = df_cimp$Component, Value = df_cimp$Conditional,
             Type = "Conditional"),
  data.frame(Component = df_cimp$Component, Value = df_cimp$Unconditional,
             Type = "Unconditional")
)

ggplot(df_cimp_long, aes(x = .data$Component, y = .data$Value,
                         fill = .data$Type)) +
  geom_col(position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("Conditional" = "#4A90D9",
                                "Unconditional" = "#D55E00")) +
  labs(title = "Conditional vs Unconditional Permutation Importance",
       subtitle = "Similar values confirm FPC scores are orthogonal (no spurious importance)",
       x = "Component", y = "Importance") +
  theme(legend.position = "bottom")
```

### Regression Depth

**In plain English:** Regression depth measures how "central" the estimated
coefficient vector is among bootstrap resamples. A high depth means the
estimate sits near the middle of the bootstrap cloud — it is robust. A low
depth means it is near the boundary, suggesting that a small data perturbation
could substantially change the coefficients. This is particularly useful for
detecting sensitivity to outliers.

**Mathematically:** Given $B$ bootstrap replicates
$\hat{\boldsymbol{b}}^{*(1)},\ldots,\hat{\boldsymbol{b}}^{*(B)}$, the
regression depth of $\hat{\boldsymbol{b}}$ is the minimum fraction of
bootstrap replicates that lie on one side of any hyperplane through
$\hat{\boldsymbol{b}}$. Depth ranges from 0 (boundary) to 0.5 (deepest
possible).

The per-component score depths reveal which specific coefficients are most
and least stable.

```{r reg-depth}
rdepth <- fregre.depth(model, data = fdataobj, y = y, n.boot = 50, seed = 42)

cat("Beta depth:", round(rdepth$beta_depth, 4), "\n")
cat("Mean score depth:", round(rdepth$mean_score_depth, 4), "\n")
cat("Bootstrap replicates:", rdepth$n_boot_success, "\n")
```

```{r plot-score-depths}
df_sdepth <- data.frame(
  Component = paste0("PC", seq_along(rdepth$score_depths)),
  Depth = rdepth$score_depths
)

ggplot(df_sdepth, aes(x = .data$Component, y = .data$Depth)) +
  geom_col(fill = "#2E8B57", alpha = 0.7) +
  labs(title = "Score-Level Regression Depth",
       subtitle = "Higher = more robust; low depth = sensitive to data perturbations",
       x = "Component", y = "Depth")
```

### Explanation Stability

**In plain English:** Stability analysis asks: "if I collected a slightly
different sample, how much would $\beta(t)$ change?" It bootstraps the entire
model-fitting process (FPCA + regression) and computes the coefficient of
variation (CV) of $\hat{\beta}(t)$ at each domain point. Low CV means the
coefficient is stable; high CV means it fluctuates across samples and should be
interpreted cautiously.

**Mathematically:** For $B$ bootstrap resamples, the pointwise CV is

$$\text{CV}(t) = \frac{\hat{\sigma}_{\text{boot}}(\hat{\beta}(t))}{|\overline{\hat{\beta}}(t)| + \epsilon},$$

where $\hat{\sigma}_{\text{boot}}$ is the bootstrap standard deviation and
$\epsilon$ is a small constant to avoid division by zero near $\beta(t) = 0$.

**Model-free:** This function takes `data`, `y`, `ncomp` rather than a fitted
model, because it refits the model from scratch on each bootstrap sample.

```{r stability}
stab <- fregre.stability(data = fdataobj, y = y, ncomp = 5,
                         n.boot = 50, seed = 42)

cat("Mean beta(t) CV:", round(mean(stab$beta_t_cv), 4), "\n")
cat("Coefficient std devs:", round(stab$coefficient_std, 4), "\n")
cat("Bootstrap successes:", stab$n_boot_success, "\n")
```

```{r plot-stability}
df_stab <- data.frame(t = t_grid, cv = stab$beta_t_cv)

ggplot(df_stab, aes(x = .data$t, y = .data$cv)) +
  geom_line(color = "#C0392B", linewidth = 1) +
  geom_ribbon(aes(ymin = 0, ymax = .data$cv), fill = "#C0392B", alpha = 0.1) +
  labs(title = "Coefficient of Variation for beta(t)",
       subtitle = "Lower = more stable; peaks indicate unreliable regions",
       x = "t", y = "CV")
```

---

## Summary

| Diagnostic | Question | Model | Key threshold |
|:-----------|:---------|:------|:--------------|
| Leverage | How unusual is this curve's shape? | lm | $h_{ii} > 2(K+1)/n$ |
| Cook's D | How much does removing this observation change the fit? | lm | $D_i > 4/n$ |
| DFBETAS | Which coefficient does this observation affect? | lm | $> 2/\sqrt{n}$ |
| DFFITS | How much does the fitted value change? | lm | $> 2\sqrt{(K+1)/n}$ |
| VIF | Are the FPC scores collinear? | lm, logistic | VIF > 5 |
| Permutation importance | Which scores drive predictions? | lm, logistic | Relative ranking |
| Conditional importance | Same, corrected for correlations | lm, logistic | Relative ranking |
| Regression depth | Is the coefficient estimate robust? | lm, logistic | Near 0.5 = robust |
| Stability CV | Does $\beta(t)$ change across samples? | model-free | Low CV = stable |

## References

- Belsley, D.A., Kuh, E. and Welsch, R.E. (1980). *Regression Diagnostics*.
  Wiley.

- Breiman, L. (2001). Random forests. *Machine Learning*, 45(1), 5--32.

- Cook, R.D. (1977). Detection of influential observations in linear
  regression. *Technometrics*, 19(1), 15--18.

- Kutner, M.H., Nachtsheim, C.J., Neter, J. and Li, W. (2004). *Applied
  Linear Statistical Models*. 5th ed. McGraw-Hill/Irwin.

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  2nd ed. Springer.

- Rousseeuw, P.J. and Hubert, M. (1999). Regression depth. *Journal of
  the American Statistical Association*, 94(446), 388--402.

- Strobl, C., Boulesteix, A.-L., Kneib, T., Augustin, T. and Zeileis, A.
  (2008). Conditional variable importance for random forests. *BMC
  Bioinformatics*, 9, 307.

## See Also

- `vignette("explainability")` --- global and local model explanations
- `vignette("uncertainty-quantification")` --- prediction intervals, bootstrap CI
- `vignette("articles/scalar-on-function")` --- scalar-on-function regression methods
