---
title: "Robust Functional Regression"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

Standard least-squares regression estimates the conditional mean of the
response by minimizing $\sum (y_i - \hat{y}_i)^2$. This works well when the
data are clean, but even a single outlier can drastically distort the fitted
model. In functional data settings the problem is amplified: contaminated
spectra, sensor glitches, or mislabelled samples can silently wreck an OLS
fit.

**fdars** provides two robust alternatives to `fregre.lm()`:
- **`fregre.l1()`** --- L1 / Least Absolute Deviation (median) regression
- **`fregre.huber()`** --- Huber M-estimation, a smooth blend of L2 and L1

Both use the same FPC projection strategy as `fregre.lm()` but replace the
OLS loss with a robust loss, solved via iteratively reweighted least squares
(IRLS). The result is an `fregre.robust` S3 object with the same interface:
`predict()`, `print()`, and access to fitted values, residuals, and $R^2$.

```{=html}
<img src="../reference/figures/robust-regression-overview.svg" alt="Robust Functional Regression: OLS vs L1 vs Huber" style="width:100%; max-width:700px;">
```

### Which Method Should I Use?

| Method | Function | Loss | Key parameter | Best when |
|--------|----------|------|---------------|-----------|
| **OLS** | `fregre.lm()` | $\sum (y_i - \hat{y}_i)^2$ | $K$ (# components) | Clean data; full diagnostics ecosystem |
| **L1 (LAD)** | `fregre.l1()` | $\sum \lvert y_i - \hat{y}_i \rvert$ | $K$ (# components) | Heavy contamination; outliers in $y$ |
| **Huber** | `fregre.huber()` | Huber loss | $K$, $k$ (tuning) | Moderate contamination; efficiency near OLS |

**Quick decision rule:**

1. Start with `fregre.lm()` --- it is the most efficient estimator when the
   data are clean and has full downstream support (explainability, diagnostics,
   conformal prediction).
2. If you suspect outliers in the response $y$, fit `fregre.huber()` as a
   robust check. With the default $k = 1.345$ it retains 95% asymptotic
   efficiency under Gaussian errors.
3. If contamination is severe (> 10--15% of observations), use `fregre.l1()`
   for maximum resistance.

## Simulating Clean and Contaminated Data

We simulate a near-infrared (NIR) spectroscopy scenario: 200 absorbance
spectra measured at 100 wavelengths, with a linear relationship between the
spectra and a scalar property (e.g., moisture content). We then contaminate
15% of the responses with large *one-sided* errors --- simulating systematic
measurement faults such as detector saturation or sample mislabelling.

```{r simulate-data}
set.seed(42)

# --- Dimensions ---
n <- 200          # number of spectra
m <- 100          # number of wavelength channels
t_grid <- seq(900, 1700, length.out = m)   # NIR wavelength range (nm)

# --- Generate smooth spectra ---
# Each spectrum is a sum of three smooth absorption bands plus noise.
# The amplitude of each band varies across samples.
X <- matrix(0, n, m)
amp1 <- rnorm(n, 1.0, 0.5)   # dominant band at 1200 nm
amp2 <- rnorm(n, 0.5, 0.15)  # secondary band at 1400 nm
amp3 <- rnorm(n, 0.3, 0.08)  # minor band at 1600 nm

for (i in 1:n) {
  X[i, ] <- amp1[i] * dnorm(t_grid, 1200, 80) * 200 +
            amp2[i] * dnorm(t_grid, 1400, 60) * 200 +
            amp3[i] * dnorm(t_grid, 1600, 50) * 200 +
            rnorm(m, sd = 0.01)
}

fd <- fdata(X, argvals = t_grid)

# --- Clean response ---
# Moisture content depends linearly on the 1200 nm and 1400 nm band amplitudes.
y_clean <- 2.0 * amp1 - 1.5 * amp2 + rnorm(n, sd = 0.3)

# --- Contaminated response (15% one-sided outliers) ---
# Simulates detector saturation: affected readings are biased *upward*.
y_contam <- y_clean
n_outlier <- round(0.15 * n)   # 30 outliers
outlier_idx <- sample(n, n_outlier)
y_contam[outlier_idx] <- y_contam[outlier_idx] + runif(n_outlier, 5, 15)
```

The contamination is one-sided (all positive), which is the worst case for OLS:
the bias systematically shifts the coefficient estimates. Clean errors have
$\sigma = 0.3$; outlier shifts range from 5 to 15 --- roughly 15--50 times
the noise scale.

```{r plot-spectra}
plot(fd, main = "NIR Absorbance Spectra",
     xlab = "Wavelength (nm)", ylab = "Absorbance")
```

```{r plot-responses}
df_y <- data.frame(
  idx = 1:n,
  clean = y_clean,
  contaminated = y_contam,
  is_outlier = seq_len(n) %in% outlier_idx
)
ggplot(df_y, aes(x = clean, y = contaminated, color = is_outlier)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("FALSE" = "#2E8B57", "TRUE" = "#D55E00"),
                     labels = c("Clean", "Outlier")) +
  labs(title = "Clean vs Contaminated Responses",
       x = "Clean y", y = "Contaminated y", color = "")
```

## OLS Regression --- Baseline

We first fit ordinary least squares via `fregre.lm()` on both the clean and
contaminated data.

```{r ols-fit}
# --- Clean data ---
fit_ols_clean <- fregre.lm(fd, y_clean, ncomp = 5)

# --- Contaminated data ---
fit_ols_contam <- fregre.lm(fd, y_contam, ncomp = 5)

cat("OLS R-squared (clean):", round(fit_ols_clean$r.squared, 4), "\n")
cat("OLS R-squared (contaminated):", round(fit_ols_contam$r.squared, 4), "\n")
```

On clean data, OLS performs well (R² > 0.9). On contaminated data, the 30
one-sided outliers systematically bias the coefficient estimates, inflating
the residuals for *all* observations --- even the clean ones. This is the
hallmark of OLS fragility: outliers have unbounded influence.

## L1 (LAD) Regression

`fregre.l1()` minimises the sum of absolute residuals rather than squared
residuals. Because the absolute value penalises large residuals linearly (not
quadratically), outliers exert much less influence on the fit.

```{r l1-fit}
# --- Clean data ---
fit_l1_clean <- fregre.l1(fd, y_clean, ncomp = 5)

# --- Contaminated data ---
fit_l1_contam <- fregre.l1(fd, y_contam, ncomp = 5)

cat("L1 R-squared (clean):", round(fit_l1_clean$r.squared, 4), "\n")
cat("L1 R-squared (contaminated):", round(fit_l1_contam$r.squared, 4), "\n")
```

### Examining the IRLS Weights

```{r l1-weights}
# Observations with small weights are effectively identified as outliers
w <- fit_l1_contam$weights
cat("Mean weight (clean obs):", round(mean(w[-outlier_idx]), 3), "\n")
cat("Mean weight (outlier obs):", round(mean(w[outlier_idx]), 3), "\n")
cat("Iterations:", fit_l1_contam$iterations, "\n")
cat("Converged:", fit_l1_contam$converged, "\n")
```

## Huber M-Estimation

Huber's M-estimator provides a smooth compromise between L2 and L1. The
Huber loss function is:

$$\rho_k(r) = \begin{cases} \frac{1}{2} r^2 & \text{if } |r| \leq k \\ k |r| - \frac{1}{2} k^2 & \text{if } |r| > k \end{cases}$$

where $k > 0$ is a tuning parameter. For small residuals ($|r| \leq k$),
the loss is quadratic (like OLS); for large residuals ($|r| > k$), the loss
is linear (like L1).

```{r huber-fit}
# --- Default k = 1.345 ---
fit_huber_contam <- fregre.huber(fd, y_contam, ncomp = 5)

cat("Huber R-squared (contaminated):", round(fit_huber_contam$r.squared, 4), "\n")
cat("Iterations:", fit_huber_contam$iterations, "\n")
cat("Converged:", fit_huber_contam$converged, "\n")
```

### The Tuning Parameter $k$

| $k$ | Behaviour | Efficiency (Gaussian) | Robustness |
|-----|-----------|----------------------|------------|
| 0.5 | Very aggressive | ~72% | High |
| 1.0 | Moderate | ~89% | Good |
| **1.345** | **Default** | **95%** | **Good** |
| 2.0 | Mild | ~98% | Moderate |
| $\infty$ | Identical to OLS | 100% | None |

```{r huber-weights}
# Compare weights across k values
w_default <- fit_huber_contam$weights

cat("Huber k=1.345:\n")
cat("  Mean weight (clean):", round(mean(w_default[-outlier_idx]), 3), "\n")
cat("  Mean weight (outlier):", round(mean(w_default[outlier_idx]), 3), "\n")
```

## Comparison: OLS vs L1 vs Huber

We compare all three methods on clean and contaminated data using a
train/test split.

```{r comparison}
# --- Train/test split ---
set.seed(123)
train_idx <- sample(n, 160)
test_idx <- setdiff(1:n, train_idx)

fd_train <- fd[train_idx, ]
fd_test <- fd[test_idx, ]

# ==========================================
# Clean data
# ==========================================
y_tr_clean <- y_clean[train_idx]
y_te_clean <- y_clean[test_idx]

fit_ols_c <- fregre.lm(fd_train, y_tr_clean, ncomp = 5)
fit_l1_c <- fregre.l1(fd_train, y_tr_clean, ncomp = 5)
fit_hub_c <- fregre.huber(fd_train, y_tr_clean, ncomp = 5)

pred_ols_c <- predict(fit_ols_c, fd_test)
pred_l1_c <- predict(fit_l1_c, fd_test)
pred_hub_c <- predict(fit_hub_c, fd_test)

# ==========================================
# Contaminated data
# ==========================================
y_tr_contam <- y_contam[train_idx]

fit_ols_d <- fregre.lm(fd_train, y_tr_contam, ncomp = 5)
fit_l1_d <- fregre.l1(fd_train, y_tr_contam, ncomp = 5)
fit_hub_d <- fregre.huber(fd_train, y_tr_contam, ncomp = 5)

# Predict on test set --- compare to *clean* truth
pred_ols_d <- predict(fit_ols_d, fd_test)
pred_l1_d <- predict(fit_l1_d, fd_test)
pred_hub_d <- predict(fit_hub_d, fd_test)
```

### Results Table

```{r results-table}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
r2 <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

results <- data.frame(
  Method = c("OLS (fregre.lm)", "L1 (fregre.l1)", "Huber (fregre.huber)"),
  RMSE_Clean = round(c(rmse(y_te_clean, pred_ols_c),
                        rmse(y_te_clean, pred_l1_c),
                        rmse(y_te_clean, pred_hub_c)), 4),
  R2_Clean = round(c(r2(y_te_clean, pred_ols_c),
                      r2(y_te_clean, pred_l1_c),
                      r2(y_te_clean, pred_hub_c)), 4),
  RMSE_Contam = round(c(rmse(y_te_clean, pred_ols_d),
                         rmse(y_te_clean, pred_l1_d),
                         rmse(y_te_clean, pred_hub_d)), 4),
  R2_Contam = round(c(r2(y_te_clean, pred_ols_d),
                       r2(y_te_clean, pred_l1_d),
                       r2(y_te_clean, pred_hub_d)), 4)
)

knitr::kable(results,
  col.names = c("Method", "RMSE", "R\u00b2", "RMSE", "R\u00b2"),
  caption = "Test-set performance on clean vs contaminated training data"
)
```

### Visualizing Predictions

```{r vis-predictions}
df_pred <- data.frame(
  Observed = rep(y_te_clean, 3),
  Predicted = c(pred_ols_d, pred_l1_d, pred_hub_d),
  Method = rep(c("OLS (fregre.lm)", "L1 (fregre.l1)", "Huber (fregre.huber)"),
               each = length(y_te_clean))
)

ggplot(df_pred, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7, color = "#4A90D9") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Method) +
  labs(title = "Observed vs Predicted (Contaminated Training Data)",
       x = "True Response", y = "Predicted")
```

## Explainability Support

Both `fregre.l1()` and `fregre.huber()` models support the full
explainability ecosystem: PDP, ALE, SHAP, permutation importance,
saliency maps, LIME, and Sobol indices --- the same tools available for
`fregre.lm()`:

```{r explainability}
# PDP for robust model
pdp <- fregre.pdp(fit_l1_contam, fd, component = 1)
cat("PDP grid values:", length(pdp$grid_values), "points\n")

# SHAP values
shap <- fregre.shap(fit_l1_contam, fd)
cat("SHAP values computed for", nrow(shap$shap_values), "observations\n")
```

## When to Use Robust Regression

**Use `fregre.l1()` when:**

- You suspect a substantial fraction (> 5%) of response values may be
  corrupted by gross errors.
- You want maximum resistance to outliers and are willing to accept a
  small loss of efficiency on clean data.

**Use `fregre.huber()` when:**

- You want a compromise: near-OLS efficiency on clean data with
  protection against moderate contamination.
- You are unsure whether outliers are present --- Huber is a safe default
  that costs very little when the data are clean.

**Stick with `fregre.lm()` when:**

- The data are clean and well-curated.
- You need the full downstream ecosystem: diagnostics (Cook's distance,
  leverage, DFBETAS), uncertainty quantification (conformal prediction),
  and cross-validation (`fregre.lm.cv()`).

## Mathematical Details

### FPC Projection

Both robust methods begin the same way as `fregre.lm()`: perform FPCA on the
functional predictor and project each curve onto $K$ principal components:

$$\xi_{ik} = \int (X_i(t) - \bar{X}(t))\phi_k(t)\,dt, \quad k = 1, \ldots, K$$

This reduces the infinite-dimensional functional regression to a
finite-dimensional problem:

$$y_i = \alpha + \sum_{k=1}^{K} \gamma_k \xi_{ik} + \epsilon_i$$

The only difference from `fregre.lm()` is the loss function used to
estimate $(\alpha, \gamma_1, \ldots, \gamma_K)$.

### IRLS Algorithm

Both L1 and Huber are solved by iteratively reweighted least squares. At
iteration $t$:

1. Compute residuals: $r_i^{(t)} = y_i - \alpha^{(t)} - \boldsymbol{\xi}_i^\top \boldsymbol{\gamma}^{(t)}$
2. Compute weights: $w_i^{(t)} = \psi(r_i^{(t)}) / r_i^{(t)}$
3. Solve weighted least squares: $(\alpha^{(t+1)}, \boldsymbol{\gamma}^{(t+1)}) = \arg\min \sum w_i^{(t)} (y_i - \alpha - \boldsymbol{\xi}_i^\top \boldsymbol{\gamma})^2$

The weight function $\psi(r)/r$ differs by method:

- **L1:** $w_i = 1/|r_i|$ (with a floor to avoid division by zero)
- **Huber:** $w_i = \min(1, k/|r_i|)$

## See Also

- `vignette("articles/scalar-on-function")` --- overview of all
  scalar-on-function methods
- `vignette("articles/outlier-detection")` --- detecting outliers in
  functional data
- `vignette("articles/regression-diagnostics")` --- influence and
  diagnostics for `fregre.lm()`
- `vignette("articles/explainability")` --- interpreting `fregre.lm()` models

## References

- Huber, P.J. (1964). Robust Estimation of a Location Parameter. *Annals of
  Mathematical Statistics*, 35(1), 73--101.
- Huber, P.J. and Ronchetti, E.M. (2009). *Robust Statistics*, 2nd ed.
  Wiley.
- Maronna, R.A., Martin, R.D., Yohai, V.J., and Salibian-Barrera, M.
  (2019). *Robust Statistics: Theory and Methods*, 2nd ed. Wiley.
- Cardot, H., Crambes, C., and Sarda, P. (2005). Quantile Regression when
  the Covariates are Functions. *Journal of Nonparametric Statistics*,
  17(7), 841--856.
