---
title: "Scalar-on-Shape Regression"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Standard scalar-on-function regression assumes curves are well-aligned before
analysis. When functional predictors contain **phase variation** --- differences
in the *timing* of features like peaks, valleys, or inflection points ---
standard methods confound shape with timing. The estimated coefficient function
$\hat\beta(t)$ becomes blurred, and predictive power drops.

`scalar.on.shape()` solves this by building a regression model that is
**invariant to phase variation**. It uses the Fisher-Rao elastic framework to
separate *shape* (amplitude after alignment) from *timing* (warping), then
regresses the scalar response on the shape of each curve.

```{=html}
<img src="../reference/figures/scalar-on-shape-overview.svg" alt="Scalar-on-Shape Regression Pipeline" style="width:100%; max-width:700px;">
```

### When to Use Scalar-on-Shape

| Scenario | Recommended method |
|----------|--------------------|
| Curves well-aligned, linear relationship | `fregre.lm()` |
| Curves misaligned, response depends on *amplitude* | `elastic.regression()` |
| Curves misaligned, response depends on *shape* (phase-invariant) | `scalar.on.shape()` |
| Unknown alignment quality | Compare all three |

**Key insight:** `elastic.regression()` aligns curves then regresses on
amplitude. `scalar.on.shape()` goes further --- it extracts a
*phase-invariant shape score* via the Fisher-Rao inner product, then maps
that score to the response through a flexible index function. This makes it
robust even when the predictive signal lives entirely in shape.

## The Scalar-on-Shape Model

### Mathematical Formulation

Given functional observations $X_1(t), \ldots, X_n(t)$ and scalar responses
$y_1, \ldots, y_n$, the model proceeds in four stages.

**1. SRSF transformation.** Each curve is mapped to its square-root slope
function (SRSF):

$$q_i(t) = \text{sign}(X_i'(t)) \sqrt{|X_i'(t)|}$$

The SRSF representation makes the $\mathbb{L}^2$ distance between
transformed curves equal to the Fisher-Rao distance between original curves.

**2. Elastic alignment.** Each SRSF $q_i$ is aligned to a shape index
function $\beta(t)$ via dynamic programming, producing optimal warping
functions $\gamma_i \in \Gamma$:

$$q_i^*(\cdot) = (q_i \circ \gamma_i) \sqrt{\dot\gamma_i}$$

**3. Shape scores.** Each aligned curve is projected onto the shape index
function:

$$s_i = \langle q_i^*, \beta \rangle = \int_0^1 q_i^*(t) \, \beta(t) \, dt$$

These scalar shape scores capture *how much* of the common shape pattern
each curve exhibits, invariant to timing.

**4. Index function.** The response is linked to the shape score through a
flexible function $h$:

$$\hat{y}_i = h(s_i)$$

The index function $h$ can be the identity (linear), a polynomial, or a
Nadaraya-Watson kernel smoother.

### Iterative Optimization

The shape index $\beta(t)$, warping functions $\gamma_i$, and link function
$h$ are estimated jointly by alternating:

1. Fix $\beta$ and $h$, update each $\gamma_i$ via DP alignment
2. Fix $\gamma_i$'s, update $\beta$ via penalized least squares on the
   Fourier basis
3. Fix $\beta$ and $\gamma_i$'s, update $h$ (polynomial fit or kernel smooth)

The algorithm converges when the relative change in SSE falls below `tol`.

## Simulating Data with Phase Variation

We simulate a biomechanics-inspired scenario: 100 ground reaction force
curves with both amplitude and phase variation. The scalar response (e.g.,
vertical jump height) depends on the *shape* of the force curve, not its
timing or scale.

```{r sim-data, eval = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)

n <- 100
m <- 80
t_grid <- seq(0, 1, length.out = m)

# Template: ground reaction force profile with two peaks
template <- function(s) {
  0.8 * exp(-40 * (s - 0.3)^2) + 1.2 * exp(-30 * (s - 0.7)^2)
}

X <- matrix(0, n, m)
y <- numeric(n)

for (i in 1:n) {
  # Shape variation: relative height of the two peaks
  shape_param <- rnorm(1, 0, 0.5)

  # Phase variation: random nonlinear warp
  a <- runif(1, 0.6, 1.8)
  b <- runif(1, 0.6, 1.8)
  gamma <- pbeta(t_grid, a, b)

  # Build the curve: shape variation + warp + noise
  base <- 0.8 * exp(-40 * (t_grid - 0.3)^2) * (1 + 0.3 * shape_param) +
          1.2 * exp(-30 * (t_grid - 0.7)^2) * (1 - 0.2 * shape_param)

  # Apply phase warp
  X[i, ] <- approx(t_grid, base, xout = gamma, rule = 2)$y +
            rnorm(m, sd = 0.03)

  # Response depends on shape, not timing
  y[i] <- 3 * shape_param + rnorm(1, sd = 0.5)
}

fd <- fdata(X, argvals = t_grid)
```

The response $y$ depends on `shape_param`, which controls the relative
height of the two peaks. The nonlinear warping (`pbeta`) shifts peak
*locations* randomly, creating phase variation that is unrelated to $y$.

```{r plot-sim, eval = FALSE}
plot(fd, main = "Simulated Ground Reaction Force Curves",
     xlab = "Normalized Time", ylab = "Force (BW)")
```

## Fitting the Model

### Identity Index (Linear Link)

The simplest model uses `index.method = "identity"`, which assumes a linear
relationship between the shape score and the response.

```{r fit-identity, eval = FALSE}
fit_id <- scalar.on.shape(fd, y, nbasis = 11, lambda = 1e-3,
                          index.method = "identity",
                          max.iter.outer = 15, tol = 1e-4)
print(fit_id)
```

### Polynomial Index (Nonlinear Link)

When the relationship between shape score and response may be nonlinear,
use a polynomial index with `index.method = "polynomial"`:

```{r fit-poly, eval = FALSE}
fit_poly <- scalar.on.shape(fd, y, nbasis = 11, lambda = 1e-3,
                            index.method = "polynomial",
                            index.param = 3,  # cubic polynomial
                            max.iter.outer = 15, tol = 1e-4)
print(fit_poly)
```

### Nadaraya-Watson Index (Fully Nonparametric Link)

For maximum flexibility, the Nadaraya-Watson kernel smoother makes no
parametric assumptions about $h$:

```{r fit-nw, eval = FALSE}
fit_nw <- scalar.on.shape(fd, y, nbasis = 11, lambda = 1e-3,
                          index.method = "nadaraya.watson",
                          index.param = 0.5,  # bandwidth
                          max.iter.outer = 15, tol = 1e-4)
print(fit_nw)
```

### Comparing Index Methods

```{r compare-index, eval = FALSE}
comp_index <- data.frame(
  Method = c("Identity (linear)", "Polynomial (cubic)", "Nadaraya-Watson"),
  R2 = round(c(fit_id$r.squared, fit_poly$r.squared, fit_nw$r.squared), 4),
  SSE = round(c(fit_id$sse, fit_poly$sse, fit_nw$sse), 2),
  Outer_Iter = c(fit_id$n.iter.outer, fit_poly$n.iter.outer,
                 fit_nw$n.iter.outer)
)
knitr::kable(comp_index, caption = "Index method comparison")
```

When the true relationship is linear (as in our simulation), the identity
index should perform well. Polynomial and NW indices add flexibility at the
cost of potential overfitting on small samples.

### Examining the Fitted Model

```{r examine-fit, eval = FALSE}
# Shape index function beta(t)
df_beta <- data.frame(t = t_grid, beta = fit_id$beta)
ggplot(df_beta, aes(x = t, y = beta)) +
  geom_line(color = "#4A90D9", linewidth = 1.2) +
  labs(title = "Estimated Shape Index Function",
       subtitle = "beta(t): the shape pattern that predicts y",
       x = "Normalized Time", y = expression(beta(t)))
```

The shape index $\hat\beta(t)$ reveals *which aspects of curve shape*
drive the response. Peaks in $\beta(t)$ correspond to regions where the
curve shape is most predictive.

```{r examine-scores, eval = FALSE}
# Shape scores vs observed response
df_scores <- data.frame(score = fit_id$shape.scores, y = y)
ggplot(df_scores, aes(x = score, y = y)) +
  geom_point(color = "#2E8B57", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#D55E00",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Shape Scores vs Response",
       subtitle = paste("R-squared =", round(fit_id$r.squared, 3)),
       x = "Shape Score", y = "Response (y)")
```

### Prediction

```{r predict-sos, eval = FALSE}
# Hold-out evaluation
set.seed(123)
train_idx <- sample(n, 70)
test_idx <- setdiff(1:n, train_idx)

fd_train <- fd[train_idx, ]
fd_test <- fd[test_idx, ]
y_train <- y[train_idx]
y_test <- y[test_idx]

fit_train <- scalar.on.shape(fd_train, y_train, nbasis = 11, lambda = 1e-3,
                             index.method = "identity",
                             max.iter.outer = 15, tol = 1e-4)

y_pred <- predict(fit_train, fd_test)

cat("Test RMSE:", round(sqrt(mean((y_test - y_pred)^2)), 4), "\n")
cat("Test R2:", round(1 - sum((y_test - y_pred)^2) /
                          sum((y_test - mean(y_test))^2), 4), "\n")
```

## Comparison with Other Methods

We compare three regression approaches on our phase-variable data to
demonstrate when scalar-on-shape regression is the right tool.

### Standard Regression (Ignoring Phase)

`fregre.lm()` treats the raw (misaligned) curves as input. Because FPCA
captures the dominant *phase* variation first, the shape signal that drives
$y$ is buried in higher-order components.

```{r fit-standard, eval = FALSE}
fit_std <- fregre.lm(fd_train, y_train, ncomp = 5)
pred_std <- predict(fit_std, fd_test)

cat("fregre.lm R-squared (train):", round(fit_std$r.squared, 4), "\n")
cat("fregre.lm RMSE (test):",
    round(sqrt(mean((y_test - pred_std)^2)), 4), "\n")
```

### Elastic Regression (Alignment-Aware)

`elastic.regression()` jointly aligns curves and fits a coefficient function.
It captures amplitude variation after alignment, but does not extract a
phase-invariant shape score.

```{r fit-elastic, eval = FALSE}
fit_elastic <- elastic.regression(fd_train, y_train, ncomp.beta = 5,
                                  lambda = 0.01, max.iter = 20, tol = 1e-3)
pred_elastic <- predict(fit_elastic, fd_test)

cat("elastic.regression R-squared (train):",
    round(fit_elastic$r.squared, 4), "\n")
cat("elastic.regression RMSE (test):",
    round(sqrt(mean((y_test - pred_elastic)^2)), 4), "\n")
```

### Scalar-on-Shape (Full Phase-Invariant)

`scalar.on.shape()` extracts a phase-invariant shape score via the
Fisher-Rao inner product, giving it the cleanest signal when $y$ depends
on shape.

```{r fit-sos-compare, eval = FALSE}
pred_sos <- predict(fit_train, fd_test)

cat("scalar.on.shape R-squared (train):",
    round(fit_train$r.squared, 4), "\n")
cat("scalar.on.shape RMSE (test):",
    round(sqrt(mean((y_test - pred_sos)^2)), 4), "\n")
```

### Summary Table

```{r summary-table, eval = FALSE}
rmse_std <- sqrt(mean((y_test - pred_std)^2))
rmse_elastic <- sqrt(mean((y_test - pred_elastic)^2))
rmse_sos <- sqrt(mean((y_test - pred_sos)^2))

r2_test <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}

comparison <- data.frame(
  Method = c("fregre.lm (standard)",
             "elastic.regression",
             "scalar.on.shape"),
  Handles_Phase = c("No", "Yes (alignment)", "Yes (shape score)"),
  Train_R2 = round(c(fit_std$r.squared,
                      fit_elastic$r.squared,
                      fit_train$r.squared), 4),
  Test_RMSE = round(c(rmse_std, rmse_elastic, rmse_sos), 4),
  Test_R2 = round(c(r2_test(y_test, pred_std),
                     r2_test(y_test, pred_elastic),
                     r2_test(y_test, pred_sos)), 4)
)
knitr::kable(comparison, caption = "Method comparison on phase-variable data")
```

### Visualizing Predictions

```{r vis-predictions, eval = FALSE}
df_pred <- data.frame(
  Observed = rep(y_test, 3),
  Predicted = c(pred_std, pred_elastic, pred_sos),
  Method = rep(c("fregre.lm", "elastic.regression", "scalar.on.shape"),
               each = length(y_test))
)

ggplot(df_pred, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7, color = "#4A90D9") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Method) +
  labs(title = "Observed vs Predicted by Method",
       x = "Observed", y = "Predicted")
```

## When to Use Which Method

| Criterion | `fregre.lm` | `elastic.regression` | `scalar.on.shape` |
|-----------|:-----------:|:--------------------:|:------------------:|
| **Phase variation** | Ignores it | Removes via alignment | Invariant via shape score |
| **Signal source** | Raw curve values | Amplitude after alignment | Shape (Fisher-Rao inner product) |
| **Link to response** | Linear (FPC scores) | Linear ($\beta(t)$ integral) | Identity / polynomial / NW |
| **Diagnostics** | Full (explainability, influence, etc.) | R-squared, residuals | R-squared, residuals |
| **Speed** | Fast | Moderate (iterative alignment) | Moderate (iterative alignment) |
| **Best for** | Well-aligned curves | Misaligned; amplitude signal | Misaligned; shape signal |

**Decision rule:**

1. If curves are well-aligned or phase variation is small, use `fregre.lm()`.
   It has the richest diagnostics and explainability support.
2. If curves are misaligned and the response depends on *amplitude*
   (overall scale or height differences), use `elastic.regression()`.
3. If curves are misaligned and the response depends on *shape*
   (the functional form after removing timing and scale), use
   `scalar.on.shape()`.
4. When unsure, fit all three and compare hold-out RMSE.

## Tuning Parameters

| Parameter | Default | Role | Guidance |
|-----------|---------|------|----------|
| `nbasis` | 11 | Fourier basis dimension for $\beta(t)$ | Increase for complex shapes; odd values recommended |
| `lambda` | 1e-3 | Roughness penalty on $\beta(t)$ | Increase to smooth $\beta$; decrease to capture detail |
| `index.method` | `"identity"` | Link function type | Start with identity; try polynomial if nonlinear |
| `index.param` | 2 | Polynomial degree or NW bandwidth | Higher = more flexible (risk of overfitting) |
| `g.degree` | 2 | Amplitude link polynomial degree | Usually left at default |
| `dp.lambda` | 0 | DP alignment penalty | Increase to discourage extreme warps |
| `max.iter.outer` | 10 | Outer loop iterations | 10--20 usually sufficient |
| `tol` | 1e-4 | Convergence threshold (relative SSE change) | Tighter = more iterations, better fit |

## References

- Srivastava, A., Klassen, E., Joshi, S.H. and Jermyn, I.H. (2011).
  Shape Analysis of Elastic Curves in Euclidean Spaces. *IEEE Transactions
  on Pattern Analysis and Machine Intelligence*, 33(7), 1415--1428.

- Tucker, J.D., Wu, W. and Srivastava, A. (2013). Generative Models for
  Functional Data Using Phase and Amplitude Separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.

- Xie, W., Kurtek, S., Bharath, K. and Sun, Y. (2017). A Geometric
  Approach to Visualization of Variability in Functional Data. *Journal
  of the American Statistical Association*, 112(519), 979--993.

## See Also

- `vignette("articles/elastic-regression")` --- elastic scalar-on-function
  regression and logistic classification
- `vignette("articles/scalar-on-function")` --- standard scalar-on-function
  regression methods
- `vignette("articles/elastic-alignment")` --- curve alignment without
  regression
- `vignette("elastic-fpca")` --- elastic FPCA for amplitude/phase separation
