---
title: "Scalar-on-Shape Regression"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

Standard scalar-on-function regression assumes curves are well-aligned before
analysis. When functional predictors contain **phase variation** --- differences
in the *timing* of features like peaks, valleys, or inflection points ---
standard methods confound shape with timing. The estimated coefficient function
$\hat\beta(t)$ becomes blurred, and predictive power drops.

`scalar.on.shape()` solves this by building a regression model that is
**invariant to phase variation**. It uses the Fisher-Rao elastic framework to
separate *shape* (amplitude after alignment) from *timing* (warping), then
regresses the scalar response on the shape of each curve.

```{=html}
<img src="../reference/figures/scalar-on-shape-overview.svg" alt="Scalar-on-Shape Regression Pipeline" style="width:100%; max-width:700px;">
```

### When to Use Scalar-on-Shape

| Scenario | Recommended method |
|----------|--------------------|
| Curves well-aligned, linear relationship | `fregre.lm()` |
| Curves misaligned, response depends on *amplitude* | `elastic.regression()` |
| Curves misaligned, response depends on *shape* (phase-invariant) | `scalar.on.shape()` |
| Unknown alignment quality | Compare all three |

## Simulating Data with Phase Variation

We simulate a biomechanics-inspired scenario: 200 force--time curves where
each curve is a single bump (e.g., a ground reaction force impulse during
a vertical jump). The bump's **location** varies across subjects (phase
variation from different jump timings), but the scalar response --- vertical
jump height --- depends on the bump's **width** (a shape feature that is
invariant to when the jump occurred).

This creates a challenging problem for standard FPCA-based regression: the
first several principal components capture the dominant *phase* variation
(where the bump is), not the *shape* variation (how wide it is). The shape
signal is buried in higher-order components that are hard to select in
practice.

```{r sim-data}
set.seed(42)

n <- 200
m <- 60
t_grid <- seq(0, 1, length.out = m)

X <- matrix(0, n, m)
y <- numeric(n)

for (i in 1:n) {
  # Shape: bump width (this predicts jump height)
  width_param <- rnorm(1, 0, 1)
  width <- 0.06 + 0.015 * width_param

  # Phase: bump center varies continuously (dominant variation)
  center <- runif(1, 0.25, 0.75)

  X[i, ] <- exp(-(t_grid - center)^2 / (2 * width^2)) +
            rnorm(m, sd = 0.015)

  # Response depends only on width (shape), not center (phase)
  y[i] <- 3 * width_param + rnorm(1, sd = 0.3)
}

fd <- fdata(X, argvals = t_grid)
```

```{r plot-sim}
plot(fd, main = "Simulated Ground Reaction Force Curves",
     xlab = "Normalized Time", ylab = "Force (BW)")
```

Note how the bumps are located at different positions along the time axis ---
this is phase variation. The response $y$ depends only on bump *width*, a
shape feature that standard FPCA mixes with phase.

## Fitting the Model

### Identity Index (Linear Link)

The simplest model uses `index.method = "identity"`, which assumes a linear
relationship between the shape score and the response.

```{r fit-identity}
fit_id <- scalar.on.shape(fd, y, nbasis = 11, lambda = 1e-3,
                          index.method = "identity",
                          max.iter.outer = 15, tol = 1e-4)
print(fit_id)
```

### Examining the Fitted Model

```{r examine-fit}
# Shape index function beta(t)
df_beta <- data.frame(t = t_grid, beta = fit_id$beta)
ggplot(df_beta, aes(x = t, y = beta)) +
  geom_line(color = "#4A90D9", linewidth = 1.2) +
  labs(title = "Estimated Shape Index Function",
       subtitle = "beta(t): the shape pattern that predicts y",
       x = "Normalized Time", y = expression(beta(t)))
```

The shape index $\hat\beta(t)$ reveals *which aspects of curve shape*
drive the response. Peaks in $\beta(t)$ correspond to regions where the
curve shape is most predictive.

```{r examine-scores}
# Shape scores vs observed response
df_scores <- data.frame(score = fit_id$shape.scores, y = y)
ggplot(df_scores, aes(x = score, y = y)) +
  geom_point(color = "#2E8B57", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#D55E00",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Shape Scores vs Response",
       subtitle = paste("R-squared =", round(fit_id$r.squared, 3)),
       x = "Shape Score", y = "Response (y)")
```

### Prediction

```{r predict-sos}
# Hold-out evaluation
set.seed(123)
train_idx <- sample(n, 140)
test_idx <- setdiff(1:n, train_idx)

fd_train <- fd[train_idx, ]
fd_test <- fd[test_idx, ]
y_train <- y[train_idx]
y_test <- y[test_idx]

fit_train <- scalar.on.shape(fd_train, y_train, nbasis = 11, lambda = 1e-3,
                             index.method = "identity",
                             max.iter.outer = 15, tol = 1e-4)

y_pred <- predict(fit_train, fd_test)

cat("Test RMSE:", round(sqrt(mean((y_test - y_pred)^2)), 4), "\n")
cat("Test R2:", round(1 - sum((y_test - y_pred)^2) /
                          sum((y_test - mean(y_test))^2), 4), "\n")
```

## Comparison with Other Methods

We compare three regression approaches on our phase-variable data. The key
question is: can each method extract the shape signal (bump width) when the
dominant variation is phase (bump location)?

### Standard Regression (Ignoring Phase)

`fregre.lm()` treats the raw (misaligned) curves as input. Because FPCA
captures the dominant *phase* variation first, the shape signal that drives
$y$ is buried in higher-order components. With a practical choice of 5
components, fregre.lm struggles to find the signal.

```{r fit-standard}
fit_std <- fregre.lm(fd_train, y_train, ncomp = 5)
pred_std <- predict(fit_std, fd_test)

cat("fregre.lm R-squared (train):", round(fit_std$r.squared, 4), "\n")
cat("fregre.lm RMSE (test):",
    round(sqrt(mean((y_test - pred_std)^2)), 4), "\n")
```

To understand *why* `fregre.lm()` fails, consider the FPCA decomposition:
the first few eigenfunctions capture *where* the bump is (phase), not *how
wide* it is (shape). The shape information only appears in higher-order
components that are typically discarded.

### Elastic Regression (Alignment-Aware)

`elastic.regression()` jointly aligns curves and fits a coefficient function.
It captures amplitude variation after alignment but is not specifically
designed to isolate shape features.

```{r fit-elastic}
fit_elastic <- elastic.regression(fd_train, y_train, ncomp.beta = 5,
                                  lambda = 0.01, max.iter = 20, tol = 1e-3)
pred_elastic <- predict(fit_elastic, fd_test)

cat("elastic.regression R-squared (train):",
    round(fit_elastic$r.squared, 4), "\n")
cat("elastic.regression RMSE (test):",
    round(sqrt(mean((y_test - pred_elastic)^2)), 4), "\n")
```

### Summary Table

```{r summary-table}
pred_sos <- predict(fit_train, fd_test)

rmse_std <- sqrt(mean((y_test - pred_std)^2))
rmse_elastic <- sqrt(mean((y_test - pred_elastic)^2))
rmse_sos <- sqrt(mean((y_test - pred_sos)^2))

r2_test <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}

comparison <- data.frame(
  Method = c("fregre.lm (ncomp=5)",
             "elastic.regression",
             "scalar.on.shape"),
  Handles_Phase = c("No", "Yes (alignment)", "Yes (shape score)"),
  Train_R2 = round(c(fit_std$r.squared,
                      fit_elastic$r.squared,
                      fit_train$r.squared), 4),
  Test_RMSE = round(c(rmse_std, rmse_elastic, rmse_sos), 4),
  Test_R2 = round(c(r2_test(y_test, pred_std),
                     r2_test(y_test, pred_elastic),
                     r2_test(y_test, pred_sos)), 4)
)
knitr::kable(comparison, caption = "Method comparison on phase-variable data")
```

`scalar.on.shape()` achieves the best predictive accuracy because it directly
targets the phase-invariant shape feature. Standard regression with 5
components captures almost none of the signal, while elastic regression
provides an intermediate result.

### Visualizing Predictions

```{r vis-predictions}
df_pred <- data.frame(
  Observed = rep(y_test, 3),
  Predicted = c(pred_std, pred_elastic, pred_sos),
  Method = factor(rep(c("fregre.lm (ncomp=5)", "elastic.regression",
                         "scalar.on.shape"),
                       each = length(y_test)),
                  levels = c("fregre.lm (ncomp=5)", "elastic.regression",
                             "scalar.on.shape"))
)

ggplot(df_pred, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7, color = "#4A90D9") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Method) +
  labs(title = "Observed vs Predicted by Method",
       x = "Observed", y = "Predicted")
```

### Why Standard Regression Fails

To see why more components don't fully solve the problem for `fregre.lm()`,
consider what happens as we increase `ncomp`:

```{r ncomp-sweep}
ncomp_vals <- c(3, 5, 8, 12)
r2_sweep <- sapply(ncomp_vals, function(k) {
  fit <- fregre.lm(fd_train, y_train, ncomp = k)
  pred <- predict(fit, fd_test)
  r2_test(y_test, pred)
})

df_sweep <- data.frame(ncomp = ncomp_vals, test_r2 = r2_sweep)
ggplot(df_sweep, aes(x = ncomp, y = test_r2)) +
  geom_line(color = "#4A90D9", linewidth = 1) +
  geom_point(color = "#4A90D9", size = 3) +
  geom_hline(yintercept = r2_test(y_test, pred_sos),
             linetype = "dashed", color = "#D55E00") +
  annotate("text", x = max(ncomp_vals), y = r2_test(y_test, pred_sos),
           label = "scalar.on.shape", vjust = -0.5, color = "#D55E00") +
  labs(title = "fregre.lm Performance vs Number of Components",
       x = "Number of FPC Components", y = expression("Test R"^2)) +
  ylim(-0.2, 1)
```

`fregre.lm()` eventually catches up with many components, but requires 8--12
to begin matching `scalar.on.shape()`. In practice, selecting the right
number of components is difficult, and using too many risks overfitting. The
phase-invariant approach avoids this model selection problem entirely.

## Tuning Parameters

| Parameter | Default | Role | Guidance |
|-----------|---------|------|----------|
| `nbasis` | 11 | Fourier basis dimension for $\beta(t)$ | Increase for complex shapes; odd values recommended |
| `lambda` | 1e-3 | Roughness penalty on $\beta(t)$ | Increase to smooth $\beta$; decrease to capture detail |
| `index.method` | `"identity"` | Link function type | Start with identity; try polynomial if nonlinear |
| `index.param` | 2 | Polynomial degree or NW bandwidth | Higher = more flexible (risk of overfitting) |
| `max.iter.outer` | 10 | Outer loop iterations | 10--20 usually sufficient |
| `tol` | 1e-4 | Convergence threshold (relative SSE change) | Tighter = more iterations, better fit |

## References

- Srivastava, A., Klassen, E., Joshi, S.H. and Jermyn, I.H. (2011).
  Shape Analysis of Elastic Curves in Euclidean Spaces. *IEEE Transactions
  on Pattern Analysis and Machine Intelligence*, 33(7), 1415--1428.

- Tucker, J.D., Wu, W. and Srivastava, A. (2013). Generative Models for
  Functional Data Using Phase and Amplitude Separation. *Computational
  Statistics & Data Analysis*, 61, 50--66.

## See Also

- `vignette("articles/elastic-regression")` --- elastic scalar-on-function
  regression and logistic classification
- `vignette("articles/scalar-on-function")` --- standard scalar-on-function
  regression methods
- `vignette("articles/elastic-alignment")` --- curve alignment without
  regression
