---
title: "Scalar-on-Function Regression"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Scalar-on-function regression predicts a **scalar response** $Y$ from a
**functional predictor** $X(t)$. This is the most common regression setting in
functional data analysis: you have curves and want to predict a number.

```{=html}
<img src="../reference/figures/scalar-on-function-diagram.svg" alt="Scalar-on-Function Regression: Four Estimation Paradigms" style="width:100%; max-width:700px;">
```

### Which Method Should I Use?

| Method | Function | Key parameter | Best when |
|--------|----------|---------------|-----------|
| **FPC Regression** | `fregre.lm()` | $K$ (# components) | Default choice; full diagnostics/explainability |
| **PC Regression** | `fregre.pc()` | $K$ (# components) | Pure R alternative to `fregre.lm()` |
| **Basis Expansion** | `fregre.basis()` | $\lambda$ (penalty) | Smooth $\beta(t)$, interpretable coefficient |
| **Nonparametric** | `fregre.np()` | $h$ (bandwidth) or $k$ | Nonlinear relationships |
| **Elastic** | `elastic.regression()` | $\lambda$, warps | Phase-variable curves |

**Quick decision rule:**

1. Start with `fregre.lm()` — it has the richest downstream support
   (explainability, diagnostics, uncertainty quantification).
2. Test linearity with `flm.test()`. If rejected, try `fregre.np()`.
3. If curves are misaligned, consider `elastic.regression()`
   (see `vignette("articles/elastic-regression")`).

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())

# Generate example data
set.seed(42)
n <- 100
m <- 50
t_grid <- seq(0, 1, length.out = m)

# Functional predictors
X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) * rnorm(1, 1, 0.3) +
            cos(4 * pi * t_grid) * rnorm(1, 0, 0.2) +
            rnorm(m, sd = 0.1)
}

fd <- fdata(X, argvals = t_grid)

# True coefficient function
beta_true <- sin(2 * pi * t_grid)

# Generate response: Y = integral(beta * X) + noise
y <- numeric(n)
for (i in 1:n) {
  y[i] <- sum(beta_true * X[i, ]) / m + rnorm(1, sd = 0.5)
}

plot(fd)
```

## The Functional Linear Model

The foundational model in scalar-on-function regression is:

$$Y_i = \alpha + \int_{\mathcal{T}} \beta(t) X_i(t) \, dt + \epsilon_i, \quad \epsilon_i \overset{\text{iid}}{\sim} (0, \sigma^2)$$

where:

- $Y_i$ is the scalar response for observation $i$
- $X_i(t)$ is the functional predictor observed over domain $\mathcal{T}$
- $\alpha$ is the intercept
- $\beta(t)$ is the **coefficient function** (unknown, to be estimated)
- $\epsilon_i$ are i.i.d. errors

The coefficient function $\beta(t)$ describes *how* the predictor curve at time
$t$ influences the response. A positive $\beta(t)$ means higher predictor
values at time $t$ increase $Y$.

### The Estimation Challenge

Unlike classical regression where we estimate a finite number of parameters,
here we must estimate an entire *function* $\beta(t)$. This is an ill-posed
inverse problem:

1. **Ill-posedness.** The integral operator is compact, so infinitely many
   solutions may exist and the least-squares estimate is wildly unstable.

2. **Curse of dimensionality.** Discretizing $X_i(t)$ at $m$ grid points
   gives $m$ parameters from $n$ observations. When $m \gg n$, the model is
   massively overparameterized.

**fdars** provides four approaches to regularize this problem, each described
in the sections below.

## FPC Regression (`fregre.lm`)

`fregre.lm()` estimates $\beta(t)$ by projecting onto **functional principal
components** (FPCs). This is the recommended default: it has the Rust backend,
supports standard errors and GCV, and is the only model with full
[explainability](explainability.html),
[diagnostics](regression-diagnostics.html), and
[uncertainty quantification](uncertainty-quantification.html) support.

### Mathematical Formulation

Let $\phi_1(t), \phi_2(t), \ldots$ be the eigenfunctions of the covariance
operator of $X$, and let $\xi_{ik} = \int (X_i(t) - \bar{X}(t))\phi_k(t)\,dt$
be the FPC scores. Expanding $\beta(t) = \sum_{k=1}^\infty \gamma_k \phi_k(t)$
and substituting gives:

$$Y_i \approx \alpha + \sum_{k=1}^K \gamma_k \xi_{ik} + \epsilon_i$$

This reduces the functional regression to an ordinary multiple regression on
$K$ FPC scores. The coefficient function is recovered as:

$$\hat\beta(t) = \sum_{k=1}^K \hat\gamma_k \phi_k(t)$$

The number of components $K$ controls the bias--variance trade-off: too few
components oversimplify $\beta(t)$; too many introduce noise.

### Standard Errors and GCV

With the FPC scores as predictors, standard OLS inference applies. The variance
of $\hat\gamma$ is $\sigma^2 (\Xi^T\Xi)^{-1}$ where $\Xi$ is the $n \times K$
score matrix. Generalized cross-validation provides a computationally efficient
alternative to leave-one-out CV:

$$\text{GCV} = \frac{1}{n}\sum_{i=1}^n \left(\frac{y_i - \hat{y}_i}{1 - h_{ii}}\right)^2$$

where $h_{ii}$ are the diagonal elements of the hat matrix
$H = \Xi(\Xi^T\Xi)^{-1}\Xi^T$.

### Basic Usage

```{r fregre-lm}
set.seed(42)
n_lm <- 80
m_lm <- 100
t_lm <- seq(0, 1, length.out = m_lm)

# Simulate functional predictor
X_lm <- matrix(0, n_lm, m_lm)
for (i in 1:n_lm) {
  X_lm[i, ] <- sin(2 * pi * t_lm) * runif(1, 0.5, 2) +
    cos(4 * pi * t_lm) * rnorm(1, sd = 0.3) +
    rnorm(m_lm, sd = 0.1)
}

beta_true_lm <- sin(2 * pi * t_lm)
y_lm <- X_lm %*% beta_true_lm / m_lm + rnorm(n_lm, sd = 0.3)

fd_lm <- fdata(X_lm, argvals = t_lm)
fit_lm <- fregre.lm(fd_lm, y_lm, ncomp = 3)
print(fit_lm)
```

### Cross-Validation for Component Selection

`fregre.lm.cv()` selects the optimal number of FPC components via k-fold
cross-validation.

```{r fregre-lm-cv}
cv_lm <- fregre.lm.cv(fd_lm, y_lm, k.range = 1:8, nfold = 10)
cat("Optimal ncomp:", cv_lm$optimal.k, "\n")
cat("CV errors:", round(cv_lm$cv.errors, 4), "\n")
```

## PC Regression (`fregre.pc`)

`fregre.pc()` uses the same FPC-based approach as `fregre.lm()` but is
implemented in pure R. It is mathematically equivalent — choose `fregre.lm()`
when you need the downstream explainability/diagnostics ecosystem; choose
`fregre.pc()` if you prefer a pure-R solution.

### Mathematical Formulation

Using FPCA, each curve is represented as:

$$X_i(t) = \bar{X}(t) + \sum_{k=1}^{\infty} \xi_{ik} \phi_k(t)$$

Truncating at $K$ components and substituting into the functional linear model:

$$Y_i = \alpha + \sum_{k=1}^{K} \gamma_k \xi_{ik} + \epsilon_i$$

The coefficient function is reconstructed as:

$$\hat{\beta}(t) = \sum_{k=1}^{K} \hat{\gamma}_k \phi_k(t)$$

### Basic Usage

```{r fregre-pc}
# Fit PC regression with 3 components
fit_pc <- fregre.pc(fd, y, ncomp = 3)
print(fit_pc)
```

### Examining the Fit

```{r pc-details}
# Fitted values
fitted_pc <- fit_pc$fitted.values

# Residuals
residuals_pc <- y - fitted_pc

# R-squared
r2_pc <- 1 - sum(residuals_pc^2) / sum((y - mean(y))^2)
cat("R-squared:", round(r2_pc, 3), "\n")
```

### Cross-Validation for Component Selection

```{r fregre-pc-cv}
# Find optimal number of components
cv_pc <- fregre.pc.cv(fd, y, ncomp.range = 1:10, seed = 42)

cat("Optimal number of components:", cv_pc$optimal.ncomp, "\n")
cat("CV error by component:\n")
print(round(cv_pc$cv.errors, 4))
```

### Prediction

```{r predict-pc}
# Split data
train_idx <- 1:80
test_idx <- 81:100

fd_train <- fd[train_idx, ]
fd_test <- fd[test_idx, ]
y_train <- y[train_idx]
y_test <- y[test_idx]

# Fit on training data
fit_train <- fregre.pc(fd_train, y_train, ncomp = 3)

# Predict on test data
y_pred <- predict(fit_train, fd_test)

# Evaluate
cat("Test RMSE:", round(pred.RMSE(y_test, y_pred), 3), "\n")
cat("Test R2:", round(pred.R2(y_test, y_pred), 3), "\n")
```

## Basis Expansion Regression (`fregre.basis`)

Basis expansion regression represents both the functional predictor $X(t)$ and
the coefficient function $\beta(t)$ using a finite set of basis functions,
providing a different regularization strategy.

### Mathematical Formulation

Let $\{B_j(t)\}_{j=1}^{J}$ be a set of basis functions (e.g., B-splines or
Fourier). We expand:

$$X_i(t) = \sum_{j=1}^{J} c_{ij} B_j(t) \quad \text{and} \quad \beta(t) = \sum_{j=1}^{J} b_j B_j(t)$$

Substituting into the functional linear model:

$$Y_i = \alpha + \mathbf{c}_i^\top \mathbf{W} \mathbf{b} + \epsilon_i$$

where $\mathbf{W}$ is the **inner product matrix** with entries
$W_{jk} = \int B_j(t) B_k(t) \, dt$.

### Ridge Regularization

To prevent overfitting, we add a roughness penalty:

$$\min_{\alpha, \mathbf{b}} \sum_{i=1}^{n} \left(Y_i - \alpha - \mathbf{c}_i^\top \mathbf{W} \mathbf{b}\right)^2 + \lambda \int \left[\beta''(t)\right]^2 dt$$

The penalty discourages rapid oscillations.

### Basis Choice

- **B-splines**: Flexible, local support, good for non-periodic data
- **Fourier**: Natural for periodic data, global support

### Basic Usage

```{r fregre-basis}
# Fit basis regression with 15 B-spline basis functions
fit_basis <- fregre.basis(fd, y, nbasis = 15, type = "bspline")
print(fit_basis)
```

### Regularization

```{r basis-lambda}
# Higher lambda = more regularization
fit_basis_reg <- fregre.basis(fd, y, nbasis = 15, type = "bspline", lambda = 1)
```

### Cross-Validation for Lambda

```{r fregre-basis-cv}
# Find optimal lambda
cv_basis <- fregre.basis.cv(fd, y, nbasis = 15, type = "bspline",
                            lambda.range = c(0.001, 0.01, 0.1, 1, 10))

cat("Optimal lambda:", cv_basis$optimal.lambda, "\n")
cat("CV error by lambda:\n")
print(round(cv_basis$cv.errors, 4))
```

### Fourier Basis

For periodic data, use Fourier basis:

```{r basis-fourier}
fit_fourier <- fregre.basis(fd, y, nbasis = 11, type = "fourier")
```

### Visualizing the Estimated Beta(t)

The coefficient function $\hat{\beta}(t)$ reveals *which time points* drive the
response. Positive regions mean higher predictor values there increase $Y$;
negative regions decrease it.

```{r beta-viz, fig.height = 4}
# Reconstruct beta_hat(t) from basis regression coefficients
beta_hat <- fit_basis$beta.est

# Compare estimated vs true beta(t)
df_beta <- data.frame(
  t = rep(t_grid, 2),
  beta = c(beta_hat, beta_true),
  Type = rep(c("Estimated", "True"), each = m)
)

ggplot(df_beta, aes(x = t, y = beta, color = Type, linetype = Type)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Estimated" = "#0072B2", "True" = "#D55E00")) +
  scale_linetype_manual(values = c("Estimated" = "solid", "True" = "dashed")) +
  labs(title = "Estimated vs True Coefficient Function",
       x = "t", y = expression(beta(t))) +
  theme(legend.position = "bottom")
```

## Nonparametric Regression (`fregre.np`)

Nonparametric functional regression makes no parametric assumptions about the
relationship between $X(t)$ and $Y$. Instead, it estimates
$\mathbb{E}[Y | X = x]$ directly using local averaging techniques.

### The General Framework

Given a new functional observation $X^*$, the predicted response is:

$$\hat{Y}^* = \hat{m}(X^*) = \sum_{i=1}^{n} w_i(X^*) Y_i$$

where $w_i(X^*)$ are weights that depend on the "distance" between $X^*$ and
the training curves $X_i$.

### Nadaraya-Watson Estimator

The **Nadaraya-Watson** (kernel regression) estimator uses:

$$\hat{m}(X^*) = \frac{\sum_{i=1}^{n} K\left(\frac{d(X^*, X_i)}{h}\right) Y_i}{\sum_{i=1}^{n} K\left(\frac{d(X^*, X_i)}{h}\right)}$$

where $K(\cdot)$ is a kernel function and $h > 0$ is the **bandwidth**.

| Kernel | Formula $K(u)$ |
|--------|----------------|
| Gaussian | $\frac{1}{\sqrt{2\pi}} e^{-u^2/2}$ |
| Epanechnikov | $\frac{3}{4}(1-u^2) \mathbf{1}_{|u| \leq 1}$ |
| Uniform | $\frac{1}{2} \mathbf{1}_{|u| \leq 1}$ |

### k-Nearest Neighbors

The **k-NN** estimator averages the responses of the $k$ closest curves:

$$\hat{m}(X^*) = \frac{1}{k} \sum_{i \in \mathcal{N}_k(X^*)} Y_i$$

Two variants are available:

- **Global k-NN** (`kNN.gCV`): single $k$ selected by leave-one-out CV
- **Local k-NN** (`kNN.lCV`): adaptive $k$ that may vary per prediction point

### Nadaraya-Watson Example

```{r fregre-np}
# Fit nonparametric regression with Nadaraya-Watson
fit_np <- fregre.np(fd, y, type.S = "S.NW")
print(fit_np)
```

### k-Nearest Neighbors

```{r knn}
# Global k-NN (single k for all observations)
fit_knn_global <- fregre.np(fd, y, type.S = "kNN.gCV")

# Local k-NN (adaptive k per observation)
fit_knn_local <- fregre.np(fd, y, type.S = "kNN.lCV")

cat("Global k-NN optimal k:", fit_knn_global$knn, "\n")
```

### Bandwidth Selection

```{r np-cv}
# Cross-validation for bandwidth
cv_np <- fregre.np.cv(fd, y, h.range = seq(0.1, 1, by = 0.1))

cat("Optimal bandwidth:", cv_np$optimal.h, "\n")
```

### Different Kernels

```{r kernels}
# Epanechnikov kernel
fit_epa <- fregre.np(fd, y, Ker = "epa")

# Available kernels: "norm", "epa", "tri", "quar", "cos", "unif"
```

### Different Metrics

```{r np-metrics}
# Use L1 metric instead of default L2
fit_np_l1 <- fregre.np(fd, y, metric = metric.lp, p = 1)

# Use semimetric based on PCA
fit_np_pca <- fregre.np(fd, y, metric = semimetric.pca, ncomp = 5)
```

### Multiple Functional Predictors

When the response depends on more than one functional predictor,
`fregre.np.multi()` extends nonparametric regression to handle a list of
functional predictors.

```{r np-multi}
# Simulate a second functional predictor
set.seed(123)
X2 <- matrix(0, n, m)
for (i in 1:n) {
  X2[i, ] <- cos(2 * pi * t_grid) * rnorm(1, 0, 0.5) +
    rnorm(m, sd = 0.1)
}
fd2 <- fdata(X2, argvals = t_grid)

# Response depends on both predictors
beta_true2 <- cos(4 * pi * t_grid)
y_multi <- numeric(n)
for (i in 1:n) {
  y_multi[i] <- sum(beta_true * X[i, ]) / m +
    0.5 * sum(beta_true2 * X2[i, ]) / m +
    rnorm(1, sd = 0.3)
}

# Fit with multiple functional predictors
fit_multi <- fregre.np.multi(list(fd, fd2), y_multi)
print(fit_multi)
```

### Mixed Functional and Scalar Predictors

Real datasets often include both functional and scalar covariates.
`fregre.np.mixed()` handles this by selecting separate bandwidths for the
functional and scalar parts.

```{r np-mixed}
# Add a scalar covariate
z <- rnorm(n)
y_mixed <- y + 0.8 * z  # scalar effect added

# Fit with functional + scalar predictors
fit_mixed <- fregre.np.mixed(fd, y_mixed, scalar.covariates = z)
print(fit_mixed)
```

## Testing the Linear Model Assumption

Before choosing between linear and nonparametric methods, `flm.test()` tests
whether the functional linear model is adequate. If the null hypothesis (linear
relationship) is not rejected, PC and basis regression are justified; otherwise
nonparametric methods may be preferable.

```{r flm-test}
# Test H0: the relationship between X(t) and Y is linear
flm_result <- flm.test(fd_train, y_train, B = 200)
print(flm_result)
```

A large p-value supports the linear model; a small p-value ($< 0.05$) suggests
the true relationship is nonlinear and parametric methods may be misspecified.

## Comparing Methods

```{r compare}
# Fit all methods on training data
fit1 <- fregre.pc(fd_train, y_train, ncomp = 3)
fit2 <- fregre.basis(fd_train, y_train, nbasis = 15)
fit3 <- fregre.np(fd_train, y_train, type.S = "S.NW")
fit4 <- fregre.np(fd_train, y_train, type.S = "kNN.gCV")
fit5 <- fregre.lm(fd_train, y_train, ncomp = 3)

# Predict on test data
pred1 <- predict(fit1, fd_test)
pred2 <- predict(fit2, fd_test)
pred3 <- predict(fit3, fd_test)
pred4 <- predict(fit4, fd_test)
pred5 <- predict(fit5, fd_test)

# Compare performance
results <- data.frame(
  Method = c("fregre.pc", "fregre.basis",
             "Nadaraya-Watson", "k-NN", "fregre.lm (Rust)"),
  RMSE = round(c(pred.RMSE(y_test, pred1),
                  pred.RMSE(y_test, pred2),
                  pred.RMSE(y_test, pred3),
                  pred.RMSE(y_test, pred4),
                  pred.RMSE(y_test, pred5)), 4),
  R2 = round(c(pred.R2(y_test, pred1),
                pred.R2(y_test, pred2),
                pred.R2(y_test, pred3),
                pred.R2(y_test, pred4),
                pred.R2(y_test, pred5)), 4),
  MAE = round(c(pred.MAE(y_test, pred1),
                 pred.MAE(y_test, pred2),
                 pred.MAE(y_test, pred3),
                 pred.MAE(y_test, pred4),
                 pred.MAE(y_test, pred5)), 4)
)
knitr::kable(results, caption = "Hold-out test set performance")
```

### Visualizing Predictions

```{r vis-predictions, fig.height = 5}
df_pred <- data.frame(
  Observed = rep(y_test, 5),
  Predicted = c(pred1, pred2, pred3, pred4, pred5),
  Method = rep(c("fregre.pc", "fregre.basis",
                 "Nadaraya-Watson", "k-NN", "fregre.lm"),
               each = length(y_test))
)

ggplot(df_pred, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  facet_wrap(~ Method, nrow = 2) +
  labs(title = "Observed vs Predicted by Method",
       x = "Observed", y = "Predicted")
```

### Comparison Table

| Method | Model | Key Parameter | Speed | Interpretability |
|--------|-------|---------------|-------|------------------|
| `fregre.lm` | $Y = \alpha + \sum_k \gamma_k \xi_{ik}$ | $K$ (# components) | Fast | High |
| `fregre.pc` | $Y = \alpha + \sum_k \gamma_k \xi_{ik}$ | $K$ (# components) | Fast | High |
| `fregre.basis` | $Y = \alpha + \int \beta(t) X(t) dt$ | $\lambda$ (penalty) | Fast | High |
| `fregre.np` (NW) | $Y = m(X)$ (nonparametric) | $h$ (bandwidth) | Moderate | Low |
| `fregre.np` (kNN) | $Y = \frac{1}{k}\sum_{j \in \mathcal{N}_k} Y_j$ | $k$ (neighbors) | Moderate | Low |

## Model Diagnostics

After fitting a functional linear model, standard regression diagnostics help
check whether the model assumptions hold.

### Residual Plots

```{r diagnostics, fig.height = 4}
df_diag <- data.frame(
  Fitted = fit_lm$fitted.values,
  Residuals = fit_lm$residuals
)

ggplot(df_diag, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Fitted vs Residuals (fregre.lm)",
       x = "Fitted values", y = "Residuals")
```

A random scatter around zero supports the linearity and constant-variance
assumptions. Fan-shaped or curved patterns suggest heteroscedasticity or
nonlinearity.

### QQ-Plot of Residuals

```{r qq-plot, fig.height = 4}
ggplot(df_diag, aes(sample = Residuals)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Normal QQ-Plot of Residuals",
       x = "Theoretical quantiles", y = "Sample quantiles")
```

### Visualizing Beta(t) with Confidence Bands

```{r beta-ci, fig.height = 4}
# fregre.lm stores beta(t) as an fdata object and SEs from the FPC regression
coefs <- fit_lm$coefficients[-1]  # exclude intercept
loadings <- fit_lm$.fpca_rotation  # m x K matrix of FPC loadings

beta_hat_lm <- as.numeric(loadings %*% coefs)

# Approximate pointwise SE via delta method using FPC score regression SEs
se_coefs <- fit_lm$std.errors[-1]  # exclude intercept SE
beta_se <- sqrt(as.numeric((loadings^2) %*% (se_coefs^2)))

df_beta_ci <- data.frame(
  t = t_lm,
  beta = beta_hat_lm,
  lower = beta_hat_lm - 1.96 * beta_se,
  upper = beta_hat_lm + 1.96 * beta_se,
  true_beta = beta_true_lm
)

ggplot(df_beta_ci, aes(x = t)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#0072B2", alpha = 0.2) +
  geom_line(aes(y = beta), color = "#0072B2", linewidth = 1) +
  geom_line(aes(y = true_beta), color = "#D55E00", linetype = "dashed",
            linewidth = 0.8) +
  labs(title = "Estimated Beta(t) with 95% Pointwise Confidence Band",
       x = "t", y = expression(hat(beta)(t))) +
  annotate("text", x = 0.8, y = max(beta_hat_lm) * 0.9,
           label = "True", color = "#D55E00") +
  annotate("text", x = 0.8, y = max(beta_hat_lm) * 0.7,
           label = "Estimated", color = "#0072B2")
```

Regions where the confidence band excludes zero indicate time points where the
predictor significantly influences the response.

For comprehensive diagnostics (influence, Cook's distance, DFBETAS, etc.), see
`vignette("articles/regression-diagnostics")`.

## Prediction Metrics

Model performance is evaluated using standard regression metrics:

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| MAE | $\frac{1}{n}\sum |y_i - \hat{y}_i|$ | Average absolute error |
| MSE | $\frac{1}{n}\sum (y_i - \hat{y}_i)^2$ | Average squared error |
| RMSE | $\sqrt{\text{MSE}}$ | Error in original units |
| $R^2$ | $1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$ | Proportion of variance explained |

```{r metrics}
cat("MAE:", pred.MAE(y_test, pred1), "\n")
cat("MSE:", pred.MSE(y_test, pred1), "\n")
cat("RMSE:", pred.RMSE(y_test, pred1), "\n")
cat("R2:", pred.R2(y_test, pred1), "\n")
```

All methods support **leave-one-out cross-validation** (LOOCV) for parameter
selection:

$$\text{CV} = \frac{1}{n} \sum_{i=1}^{n} (Y_i - \hat{Y}_{-i})^2$$

This is implemented efficiently using the "hat matrix trick" for linear methods.

## Practical Workflow

A recommended workflow for scalar-on-function regression:

1. **Start simple:** fit `fregre.lm.cv()` to find the optimal number of FPC
   components via cross-validation.
2. **Check linearity:** run `flm.test()` on the training data. A significant
   p-value suggests nonlinearity.
3. **If linear holds:** compare `fregre.lm()`, `fregre.pc()`, and
   `fregre.basis()` — all estimate the functional linear model with different
   regularization strategies.
4. **If linearity is rejected:** try `fregre.np()` with bandwidth
   cross-validation (`fregre.np.cv()`).
5. **Evaluate:** hold out a test set and compare methods with `pred.RMSE()`,
   `pred.R2()`, and `pred.MAE()`.
6. **Diagnose:** check residual plots and the estimated $\hat{\beta}(t)$ for
   interpretability.

```{r workflow}
set.seed(42)
tr <- sample(n, 70)
te <- setdiff(1:n, tr)

fd_tr <- fd[tr, ]
fd_te <- fd[te, ]
y_tr <- y[tr]
y_te <- y[te]

# Step 1: CV for component selection
cv <- fregre.lm.cv(fd_tr, y_tr, k.range = 1:6, nfold = 5)

# Step 2: Test linearity
linearity <- flm.test(fd_tr, y_tr, B = 200)
cat("FLM test p-value:", linearity$p.value, "\n")

# Step 3-4: Fit best linear and nonparametric
fit_best_lm <- fregre.lm(fd_tr, y_tr, ncomp = cv$optimal.k)
fit_best_np <- fregre.np(fd_tr, y_tr, type.S = "kNN.gCV")

# Step 5: Evaluate
pred_lm_wf <- predict(fit_best_lm, fd_te)
pred_np_wf <- predict(fit_best_np, fd_te)

cat("Linear RMSE:", round(pred.RMSE(y_te, pred_lm_wf), 4),
    "| R2:", round(pred.R2(y_te, pred_lm_wf), 4), "\n")
cat("Nonpar RMSE:", round(pred.RMSE(y_te, pred_np_wf), 4),
    "| R2:", round(pred.R2(y_te, pred_np_wf), 4), "\n")
```

## Method Selection Guide

**FPC Regression** (`fregre.lm` / `fregre.pc`):

- Best when the functional predictor has clear dominant modes of variation
- Computationally efficient for large datasets
- Interpretable: each PC represents a pattern in the data
- `fregre.lm()` is the only model with full explainability support

**Basis Expansion Regression** (`fregre.basis`):

- Best when you believe $\beta(t)$ is smooth
- Use B-splines for local features, Fourier for periodic patterns
- The penalty parameter $\lambda$ provides automatic regularization
- Good when you want to visualize and interpret $\hat{\beta}(t)$

**Nonparametric Regression** (`fregre.np`):

- Best when the relationship between $X$ and $Y$ may be nonlinear
- Makes minimal assumptions about the data-generating process
- Computationally more expensive (requires distance calculations)
- May require larger sample sizes for stable estimation

## See Also

- `vignette("articles/function-on-scalar")` — when the *response* is a curve
- `vignette("articles/elastic-regression")` — alignment-aware regression
- `vignette("articles/functional-classification")` — classifying curves
  (including `functional.logistic()`)
- `vignette("articles/explainability")` — interpreting `fregre.lm()` models
- `vignette("articles/regression-diagnostics")` — influence, Cook's distance, VIF
- `vignette("articles/uncertainty-quantification")` — prediction intervals
- `vignette("articles/cross-validation")` — cross-validation framework
- `vignette("fpca", package = "fdars")` — functional principal component analysis
- `vignette("basis-representation", package = "fdars")` — basis function representations

## References

**Foundational texts:**

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*, 2nd ed. Springer.
- Ferraty, F. and Vieu, P. (2006). *Nonparametric Functional Data Analysis: Theory and Practice*. Springer.
- Horváth, L. and Kokoszka, P. (2012). *Inference for Functional Data with Applications*. Springer.

**Key methodological papers:**

- Cardot, H., Ferraty, F., and Sarda, P. (1999). Functional Linear Model. *Statistics & Probability Letters*, 45(1), 11--22.
- Reiss, P.T. and Ogden, R.T. (2007). Functional Principal Component Regression and Functional Partial Least Squares. *Journal of the American Statistical Association*, 102(479), 984--996.
- Goldsmith, J., Bobb, J., Crainiceanu, C., Caffo, B., and Reich, D. (2011). Penalized Functional Regression. *Journal of Computational and Graphical Statistics*, 20(4), 830--851.
- Ferraty, F., Laksaci, A., and Vieu, P. (2006). Estimating Some Characteristics of the Conditional Distribution in Nonparametric Functional Models. *Statistical Inference for Stochastic Processes*, 9, 47--76.
