---
title: "Shape Analysis"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Two curves may look different pointwise yet represent the *same shape* ---
they differ only in how they are parameterized (traversal speed), translated
(vertical shift), or scaled (overall magnitude). **Shape analysis** provides
a rigorous framework for comparing curves after factoring out these nuisance
transformations.

The key idea is to work in a **quotient space**: the set of all curves
modulo a group of transformations. Two curves that differ only by a
reparameterization, translation, or scaling are identified as the same
point in the quotient space. The elastic (Fisher-Rao) framework makes
this quotient geometry computationally tractable by using the Square-Root
Slope Function (SRSF) representation.

**fdars** provides four core functions for shape analysis:

| Function | Purpose |
|----------|---------|
| `shape.representative()` | Canonical form of a single curve in the quotient space |
| `shape.distance()` | Elastic distance between two curves modulo nuisance transformations |
| `shape.mean()` | Karcher mean in the quotient space (average shape) |
| `shape.distance.matrix()` | Pairwise shape distance matrix for a set of curves |

```{=html}
<img src="../reference/figures/shape-analysis-overview.svg" alt="Shape Analysis Overview" style="width:100%; max-width:720px; display:block; margin:1em auto;" />
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Simulated Data

We simulate 30 curves that represent two groups of bump-shaped profiles,
inspired by letter-stroke outlines from different writers. Group 1 has
bumps centered near $t = 0.3$ and Group 2 near $t = 0.7$. Within each
group, the bump location varies slightly, mimicking natural variation in
handwriting.

```{r sim-data}
set.seed(42)
n <- 30
m <- 50
argvals <- seq(0, 1, length.out = m)

# Two groups of shapes - bumps at different locations
X <- matrix(0, n, m)
for (i in 1:15) {
  center <- 0.3 + rnorm(1, sd = 0.05)
  X[i, ] <- exp(-((argvals - center)^2) / 0.02) + rnorm(m, sd = 0.05)
}
for (i in 16:30) {
  center <- 0.7 + rnorm(1, sd = 0.05)
  X[i, ] <- exp(-((argvals - center)^2) / 0.02) + rnorm(m, sd = 0.05)
}
fd <- fdata(X, argvals = argvals)
group <- rep(c("Left bump", "Right bump"), each = 15)
```

```{r plot-data}
plot(fd, main = "Simulated Letter-Stroke Profiles (2 groups)",
     xlab = "Normalized Arc Length", ylab = "Curvature (a.u.)")
```

Within each group, the curves share a common shape (a single bump) but
differ in the exact position of the peak --- this is phase variation that
standard pointwise analysis would treat as genuine shape difference.

## Orbit Representatives

Every curve belongs to an **orbit** in the quotient space: the set of all
curves that can be obtained from it by applying the allowed transformations
(reparameterization, translation, or scaling). `shape.representative()`
computes a canonical member of this orbit --- the orbit representative ---
which provides a standard form for the curve's shape.

```{r orbit-rep}
# Compute orbit representative for a single curve
f_original <- X[1, ]
rep_result <- shape.representative(f_original, argvals = argvals,
                                    quotient = "reparameterization")
```

```{r plot-orbit}
df_orbit <- data.frame(
  argval = rep(argvals, 2),
  value = c(f_original, rep_result$representative),
  type = factor(rep(c("Original", "Representative"), each = m),
                levels = c("Original", "Representative"))
)

ggplot(df_orbit, aes(x = .data$argval, y = .data$value,
                     color = .data$type)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Original" = "#4A90D9",
                                "Representative" = "#D55E00")) +
  labs(title = "Orbit Representative (Reparameterization Quotient)",
       subtitle = "Canonical form after factoring out phase variation",
       x = "t", y = "f(t)", color = NULL) +
  theme(legend.position = "bottom")
```

The representative has the same shape as the original but is
reparameterized to a canonical speed. Two curves with the same shape
will map to the same representative (up to numerical precision),
regardless of how they were originally parameterized.

### Quotient Spaces

The `quotient` argument controls which transformations are factored out:

| Quotient | Factors out | Use case |
|----------|-------------|----------|
| `"reparameterization"` | Warping (timing differences) | Curves with different traversal speeds |
| `"translation"` | Vertical shifts | Curves at different baselines |
| `"scale"` | Reparameterization + scaling | Curves of different magnitudes |

## Shape Distance

`shape.distance()` computes the elastic distance between two curves in the
quotient space. It returns the distance along with the optimal warping that
aligns the second curve to the first.

```{r shape-dist}
# Compare a left-bump curve to another left-bump curve
d_same <- shape.distance(X[1, ], X[5, ], argvals = argvals,
                          quotient = "reparameterization")

# Compare a left-bump curve to a right-bump curve
d_diff <- shape.distance(X[1, ], X[20, ], argvals = argvals,
                          quotient = "reparameterization")

cat("Same group (curves 1 vs 5):", round(d_same$distance, 4), "\n")
cat("Different groups (curves 1 vs 20):", round(d_diff$distance, 4), "\n")
```

Curves from the same group have a small shape distance because their bumps
have the same form --- only the location differs, and that is factored out.
Curves from different groups have a larger distance because their shapes
genuinely differ.

```{r plot-alignment}
df_align <- data.frame(
  argval = rep(argvals, 3),
  value = c(X[1, ], X[20, ], d_diff$f2.aligned),
  curve = factor(rep(c("Curve 1 (left bump)", "Curve 20 (right bump)",
                        "Curve 20 (aligned)"), each = m),
                 levels = c("Curve 1 (left bump)", "Curve 20 (right bump)",
                            "Curve 20 (aligned)"))
)

ggplot(df_align, aes(x = .data$argval, y = .data$value,
                     color = .data$curve, linetype = .data$curve)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("#4A90D9", "#D55E00", "#2E8B57")) +
  scale_linetype_manual(values = c("solid", "solid", "dashed")) +
  labs(title = "Shape Distance: Alignment Result",
       subtitle = paste("Shape distance =", round(d_diff$distance, 4)),
       x = "t", y = "f(t)", color = NULL, linetype = NULL) +
  theme(legend.position = "bottom")
```

## Shape Mean

`shape.mean()` computes the Karcher (Frechet) mean in the quotient space.
This is the curve that minimizes the total squared shape distance to all
input curves. Unlike a pointwise mean, the shape mean is not blurred by
phase variation.

```{r shape-mean}
# Shape mean of left-bump group
fd_left <- fd[1:15, ]
sm_left <- shape.mean(fd_left, quotient = "reparameterization",
                       max.iter = 20, tol = 1e-4)
print(sm_left)
```

The built-in plot method shows aligned curves (grey) with the mean curve
(red):

```{r plot-mean}
plot(sm_left)
```

### Comparing Group Means

```{r group-means}
fd_right <- fd[16:30, ]
sm_right <- shape.mean(fd_right, quotient = "reparameterization",
                        max.iter = 20, tol = 1e-4)

df_means <- data.frame(
  argval = rep(argvals, 2),
  value = c(sm_left$mean, sm_right$mean),
  group = factor(rep(c("Left Bump Mean", "Right Bump Mean"), each = m))
)

ggplot(df_means, aes(x = .data$argval, y = .data$value,
                     color = .data$group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Left Bump Mean" = "#4A90D9",
                                "Right Bump Mean" = "#D55E00")) +
  labs(title = "Shape Means by Group",
       subtitle = "Karcher means in the reparameterization quotient space",
       x = "t", y = "f(t)", color = NULL) +
  theme(legend.position = "bottom")
```

Each shape mean captures the representative bump profile for its group
without the blurring that a pointwise average would produce.

## Shape Distance Matrix

`shape.distance.matrix()` computes all pairwise shape distances for a set
of curves. This is the foundation for downstream tasks like clustering,
multidimensional scaling, or nearest-neighbor classification.

```{r dist-matrix}
D <- shape.distance.matrix(fd, quotient = "reparameterization")
```

### Heatmap

```{r heatmap}
# Convert to long format for ggplot
df_heat <- expand.grid(Curve1 = 1:n, Curve2 = 1:n)
df_heat$distance <- as.vector(D)
df_heat$Curve1 <- factor(df_heat$Curve1)
df_heat$Curve2 <- factor(df_heat$Curve2)

ggplot(df_heat, aes(x = .data$Curve1, y = .data$Curve2,
                    fill = .data$distance)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "Shape\nDistance") +
  labs(title = "Pairwise Shape Distance Matrix",
       subtitle = "Block structure reveals two shape groups",
       x = "Curve Index", y = "Curve Index") +
  coord_equal() +
  theme(axis.text = element_text(size = 6))
```

The heatmap shows clear block-diagonal structure: curves within the same
group (1--15 and 16--30) have small pairwise distances, while curves from
different groups have large distances. This confirms that shape distance
correctly distinguishes the two bump patterns.

### Using with Standard Clustering

The shape distance matrix can be passed directly to standard R clustering
tools:

```{r hclust-shape}
# Hierarchical clustering on shape distances
hc <- hclust(as.dist(D), method = "complete")
plot(hc, main = "Dendrogram from Shape Distances",
     xlab = "Curve", ylab = "Shape Distance", cex = 0.7)

# Cut into 2 clusters
shape_clusters <- cutree(hc, k = 2)
table(Shape_Cluster = shape_clusters, True_Group = group)
```

## Best Practices

1. **Choose the right quotient space.** Use `"reparameterization"` when
   curves differ in speed/timing. Add `"translation"` or `"scale"` when
   baseline shifts or magnitude differences are also nuisance factors.
2. **Check convergence** of `shape.mean()`. If `converged = FALSE`,
   increase `max.iter` or relax `tol`.
3. **Regularization.** Set `lambda > 0` to prevent extreme warpings when
   curves are noisy or sparsely sampled.
4. **Preprocess consistently.** Smooth curves and evaluate on a common grid
   before shape analysis.
5. **Visualize warpings.** The `gamma` output from `shape.distance()` and
   `shape.representative()` reveals how much reparameterization was needed.

## See Also

- `vignette("articles/elastic-alignment")` --- elastic curve alignment
  and Karcher means
- `vignette("articles/elastic-clustering")` --- elastic k-means and
  hierarchical clustering
- `vignette("articles/scalar-on-shape")` --- scalar-on-shape regression
  (phase-invariant prediction)
- `vignette("articles/tsrvf")` --- the SRSF representation underlying
  elastic shape analysis

## References

- Srivastava, A. and Klassen, E. (2016). *Functional and Shape Data
  Analysis*. Springer.

- Srivastava, A., Klassen, E., Joshi, S.H. and Jermyn, I.H. (2011).
  Shape Analysis of Elastic Curves in Euclidean Spaces. *IEEE Transactions
  on Pattern Analysis and Machine Intelligence*, 33(7), 1415--1428.

- Kurtek, S., Srivastava, A., Klassen, E. and Ding, Z. (2012). Statistical
  Modeling of Curves Using Shapes and Related Features. *Journal of the
  American Statistical Association*, 107(499), 1152--1165.
