---
title: "Statistical Process Monitoring for Functional Data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

Statistical Process Monitoring (SPM) extends classical control charts to
functional data. When the "quality characteristic" is an entire curve
(e.g., a spectral profile, a temperature trajectory, or a process
signature), univariate control charts on summary statistics lose information.
Functional SPM monitors the full curve structure.

```{=html}
<img src="../reference/figures/spm-overview.svg" alt="SPM Pipeline: Phase I estimation and Phase II monitoring" style="width:100%; max-width:720px;">
```

The workflow has two phases:

1. **Phase I** — Estimate the in-control process using historical "good" data:
   perform FPCA, compute Hotelling $T^2$ and Squared Prediction Error (SPE)
   statistics, and set control limits.

2. **Phase II** — Monitor new observations by projecting onto the Phase I
   principal components and comparing statistics to the control limits.
   Alarms fire when a statistic exceeds its Upper Control Limit (UCL).

### When to Use Functional SPM

| Scenario | Method |
|----------|--------|
| Single functional quality variable | `spm.phase1()` + `spm.monitor()` |
| Multiple functional variables | `mf.spm.phase1()` + `mf.spm.monitor()` |
| Detecting gradual drifts | `spm.ewma()` |
| Process with known covariates | `frcc()` + `frcc.monitor()` |
| Diagnosing alarm root cause | `spm.contributions()` |

## Simulating Process Data

We simulate semiconductor wafer thickness profiles: each curve is a spectral
measurement across the wafer surface. Phase I uses 200 in-control profiles;
Phase II introduces a shift at observation 30.

```{r simulate}
set.seed(42)

m <- 50  # grid points
t_grid <- seq(0, 1, length.out = m)

# In-control process: smooth curves with random variation
n_ic <- 200  # Phase I observations
X_ic <- matrix(0, n_ic, m)
for (i in seq_len(n_ic)) {
  scores <- rnorm(3, sd = c(2, 1, 0.5))
  X_ic[i, ] <- 5 + scores[1] * sin(2 * pi * t_grid) +
                    scores[2] * cos(4 * pi * t_grid) +
                    scores[3] * sin(6 * pi * t_grid) +
                    rnorm(m, sd = 0.3)
}

# Phase II: 50 observations, shift at observation 30
n_new <- 50
X_new <- matrix(0, n_new, m)
for (i in seq_len(n_new)) {
  scores <- rnorm(3, sd = c(2, 1, 0.5))
  shift <- if (i >= 30) 1.5 * sin(2 * pi * t_grid) else 0
  X_new[i, ] <- 5 + scores[1] * sin(2 * pi * t_grid) +
                     scores[2] * cos(4 * pi * t_grid) +
                     scores[3] * sin(6 * pi * t_grid) +
                     shift + rnorm(m, sd = 0.3)
}

fd_ic <- fdata(X_ic, argvals = t_grid)
fd_new <- fdata(X_new, argvals = t_grid)
```

```{r plot-data}
plot(fd_ic, main = "Phase I: In-Control Profiles",
     xlab = "Measurement Position", ylab = "Thickness")
```

## Phase I: Estimating Control Limits

`spm.phase1()` performs FPCA on the in-control data, computes $T^2$ and SPE
statistics for each training observation, and estimates control limits at
significance level $\alpha$.

```{r phase1}
chart <- spm.phase1(fd_ic, ncomp = 3, alpha = 0.05, seed = 42)
print(chart)
```

The returned `spm.chart` object contains:

- **T2 control limit** — based on the $F$-distribution with `ncomp` and
  `n - ncomp` degrees of freedom
- **SPE control limit** — estimated from the empirical distribution of
  Phase I SPE values
- **Phase I statistics** — $T^2$ and SPE for each training observation
  (useful for checking that Phase I data is truly in-control)

```{r phase1-plot}
plot(chart)
```

## Phase II: Online Monitoring

Project new observations onto the Phase I principal components and compare
to control limits:

```{r phase2}
monitor <- spm.monitor(chart, fd_new)
print(monitor)

# Which observations triggered alarms?
alarm_idx <- which(monitor$t2.alarm | monitor$spe.alarm)
cat("Alarms at observations:", alarm_idx, "\n")
```

```{r phase2-plot}
plot(monitor)
```

The monitoring result contains:

| Field | Description |
|-------|-------------|
| `t2` | Hotelling $T^2$ values for each new observation |
| `spe` | SPE values for each new observation |
| `t2.alarm` | Logical: did $T^2$ exceed the UCL? |
| `spe.alarm` | Logical: did SPE exceed the UCL? |
| `scores` | PC scores for the new observations |

## EWMA Monitoring

The Exponentially Weighted Moving Average (EWMA) chart smooths the monitoring
statistics over time, making it more sensitive to small, persistent shifts
than the standard Shewhart-type chart:

```{r ewma}
ewma_result <- spm.ewma(chart, fd_new, lambda = 0.2, alpha = 0.05)

cat("EWMA T2 alarms:", sum(ewma_result$t2.alarm), "\n")
cat("EWMA SPE alarms:", sum(ewma_result$spe.alarm), "\n")
```

The smoothing parameter $\lambda \in (0, 1]$ controls the memory:
- $\lambda = 1$ gives the standard Shewhart chart (no smoothing)
- $\lambda = 0.2$ (default) gives moderate smoothing
- Smaller $\lambda$ detects smaller shifts but responds more slowly

## Multivariate FPCA

When multiple functional variables are measured simultaneously (e.g.,
temperature, pressure, and flow rate profiles), Multivariate FPCA
extracts joint modes of variation across all variables:

```{r mfpca}
# Simulate two correlated functional variables
fd_var1 <- fdata(X_ic, argvals = t_grid)
fd_var2 <- fdata(X_ic + matrix(rnorm(n_ic * m, sd = 0.5), n_ic, m),
                 argvals = t_grid)

mfpca_result <- mfpca(list(fd_var1, fd_var2), ncomp = 3)
cat("Eigenvalues:", round(mfpca_result$eigenvalues, 3), "\n")
```

The result contains joint scores and per-variable eigenfunctions, which
can be used for multivariate SPM via `mf.spm.phase1()`.

## Contribution Diagnostics

When an alarm fires, the next question is: **what went wrong?** Contribution
analysis decomposes the aggregate $T^2$ and SPE statistics into per-variable
(or per-grid-region) contributions, pinpointing which part of the curve
triggered the alarm.

### T-squared Contributions

$T^2$ contributions break down the Hotelling statistic into additive terms,
one per principal component. A large contribution from PC $k$ means that the
curve's score on the $k$-th eigenfunction is unusually far from the in-control
mean. This tells you *which mode of variation* shifted:

```{r contributions-t2}
# T2 contributions: one value per PC per observation
t2_contrib <- spm.contributions(monitor$scores,
                                chart$eigenvalues,
                                grid.sizes = c(m),
                                type = "t2")

# Identify the first alarming observation
alarm_idx <- which(monitor$t2.alarm | monitor$spe.alarm)
first_alarm <- alarm_idx[1]
cat("First alarm at observation:", first_alarm, "\n")
cat("T2 contributions (per PC):", round(t2_contrib[first_alarm, ], 3), "\n")
```

```{r plot-t2-contrib}
# Visualize T2 contributions for all monitoring observations
n_mon <- nrow(t2_contrib)
n_pc <- ncol(t2_contrib)
df_t2 <- data.frame(
  obs = rep(seq_len(n_mon), n_pc),
  contribution = as.vector(t2_contrib),
  pc = factor(rep(paste0("PC", seq_len(n_pc)), each = n_mon))
)

ggplot(df_t2, aes(x = obs, y = contribution, fill = pc)) +
  geom_col(position = "stack", width = 1) +
  geom_vline(xintercept = 29.5, linetype = "dashed", color = "red", linewidth = 0.5) +
  annotate("text", x = 29.5, y = max(rowSums(t2_contrib)) * 0.9,
           label = "Shift onset", hjust = 1.1, color = "red", size = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = expression("T"^2 * " Contributions by Principal Component"),
       x = "Monitoring Observation", y = expression("Contribution to T"^2),
       fill = "Component") +
  theme(legend.position = "bottom")
```

Since we injected a shift in the first eigenfunction direction ($\sin(2\pi t)$),
PC 1 should dominate the $T^2$ contributions after the shift onset at
observation 30.

### Identifying the Dominant Component

We can extract which principal component drives the alarm for each
observation. This is especially useful for directing root-cause investigation:

```{r dominant-pc}
# Which PC has the largest T2 contribution per observation?
dominant_pc <- apply(t2_contrib, 1, which.max)

df_dom <- data.frame(
  obs = seq_len(nrow(t2_contrib)),
  dominant = factor(paste0("PC", dominant_pc)),
  alarm = monitor$t2.alarm | monitor$spe.alarm
)

ggplot(df_dom, aes(x = obs, fill = dominant)) +
  geom_bar(width = 1) +
  geom_vline(xintercept = 29.5, linetype = "dashed", color = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Dominant PC per Observation",
       subtitle = "Which component drives the T\u00b2 statistic",
       x = "Monitoring Observation", y = "Count",
       fill = "Dominant PC") +
  theme(legend.position = "bottom")
```

After the shift onset (observation 30), PC 1 dominates because we injected
a shift aligned with the first eigenfunction ($\sin 2\pi t$).

### Interpreting Contributions in Practice

| Diagnostic | What it tells you | Typical action |
|-----------|-------------------|----------------|
| **T2 dominated by PC 1** | Shift in the dominant mode of variation | Check for process mean shift |
| **T2 dominated by higher PCs** | Shift in a minor mode | Check for subtle process changes |
| **SPE spike in a specific region** | New variation not in the model | Inspect that region for sensor fault |
| **SPE uniformly elevated** | Global model inadequacy | Consider adding more PCs |

In a multi-variable setting (using `mf.spm.phase1()`), the contributions are
further broken down by variable, allowing you to identify which sensor or
measurement channel is responsible for the alarm.

## Elastic SPM

When functional data exhibit **phase variation** (timing differences between
curves), standard SPM charts based on Euclidean FPCA can produce false alarms.
Elastic SPM separates amplitude and phase components, monitoring each with its
own control chart.

```{r elastic-spm, eval = FALSE}
# Phase I: build elastic chart
chart_elastic <- spm.elastic.phase1(fd_train, ncomp = 3, alpha = 0.05,
                                     monitor.phase = TRUE, warp.ncomp = 3)

# Phase II: monitor new data
mon_elastic <- spm.elastic.monitor(chart_elastic, fd_test)
```

The elastic chart produces dual T²/SPE statistics for both amplitude and phase.
Amplitude alarms indicate changes in curve shape; phase alarms indicate changes
in timing or dynamics.

For a comprehensive tutorial on elastic SPM and advanced monitoring methods, see
`vignette("articles/advanced-spm")`.

## Partial Domain Monitoring

In many applications, we need to monitor curves **before they are fully
observed** — for example, monitoring a batch process while it is still running.
Partial domain monitoring completes the unobserved portion of the curve and
projects it through the existing control chart.

```{r partial-monitor, eval = FALSE}
# Monitor a curve observed over only the first 70% of the domain
partial_values <- new_curve[1:35]  # first 35 of 50 grid points
result <- spm.monitor.partial(chart, partial.data = partial_values,
                               n.observed = 35,
                               completion = "conditional")
result$t2        # T² statistic
result$t2.alarm  # whether alarm is triggered
```

Three completion methods are available: `"conditional"` (BLUP, recommended),
`"projection"` (scaled inner products), and `"zero_pad"` (mean padding).

For profile monitoring with covariates and batch processing of partial curves,
see `vignette("articles/profile-partial-monitoring")`.

## Comparison of Methods

| Method | Detects | Best for |
|--------|---------|----------|
| **T²** | Mean shifts in the PC score space | Location changes in dominant modes |
| **SPE** | Changes not captured by the PCs | New types of variation, sensor drift |
| **EWMA** | Small persistent shifts | Gradual process degradation |
| **FRCC** | Shifts in the relationship between curves and covariates | Regression-based monitoring |
| **CUSUM** | Small sustained shifts | Drift detection with tunable sensitivity |
| **MEWMA** | Multivariate small shifts | Persistent shifts across multiple components |
| **AMEWMA** | Shifts of unknown size | Adaptive smoothing for mixed shift sizes |
| **Elastic** | Phase/amplitude changes | Phase-variable processes |
| **Partial** | Early detection mid-curve | In-progress batch monitoring |

## See Also

- `vignette("articles/fpca")` — Functional PCA fundamentals
- `vignette("articles/streaming-depth")` — Online depth-based monitoring
- `vignette("articles/outlier-detection")` — Depth-based outlier detection

## References

- Colosimo, B.M. and Pacella, M. (2010). A comparison study of control charts
  for statistical monitoring of functional data. *International Journal of
  Production Research*, 48(6), 1575--1601.
- Paynabar, K., Jin, J., and Pacella, M. (2013). Monitoring and diagnosis of
  multichannel nonlinear profiles using uncorrelated multilinear principal
  component analysis. *IIE Transactions*, 45(11), 1235--1247.
- Grasso, M., Colosimo, B.M., and Pacella, M. (2014). Profile monitoring via
  sensor fusion: the use of PCA methods for multi-channel data.
  *International Journal of Production Research*, 52(20), 6110--6135.
