---
title: "Statistical Process Monitoring for Functional Data"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
```

## Introduction

Statistical Process Monitoring (SPM) extends classical control charts to
functional data. When the "quality characteristic" is an entire curve
(e.g., a spectral profile, a temperature trajectory, or a process
signature), univariate control charts on summary statistics lose information.
Functional SPM monitors the full curve structure.

```{=html}
<img src="../reference/figures/spm-overview.svg" alt="SPM Pipeline: Phase I estimation and Phase II monitoring" style="width:100%; max-width:720px;">
```

The workflow has two phases:

1. **Phase I** — Estimate the in-control process using historical "good" data:
   perform FPCA, compute Hotelling $T^2$ and Squared Prediction Error (SPE)
   statistics, and set control limits.

2. **Phase II** — Monitor new observations by projecting onto the Phase I
   principal components and comparing statistics to the control limits.
   Alarms fire when a statistic exceeds its Upper Control Limit (UCL).

### When to Use Functional SPM

| Scenario | Method |
|----------|--------|
| Single functional quality variable | `spm.phase1()` + `spm.monitor()` |
| Multiple functional variables | `mf.spm.phase1()` + `mf.spm.monitor()` |
| Detecting gradual drifts | `spm.ewma()` |
| Process with known covariates | `frcc()` + `frcc.monitor()` |
| Diagnosing alarm root cause | `spm.contributions()` |

## Simulating Process Data

We simulate semiconductor wafer thickness profiles: each curve is a spectral
measurement across the wafer surface. Phase I uses 200 in-control profiles;
Phase II introduces a shift at observation 30.

```{r simulate}
set.seed(42)

m <- 50  # grid points
t_grid <- seq(0, 1, length.out = m)

# In-control process: smooth curves with random variation
n_ic <- 200  # Phase I observations
X_ic <- matrix(0, n_ic, m)
for (i in seq_len(n_ic)) {
  scores <- rnorm(3, sd = c(2, 1, 0.5))
  X_ic[i, ] <- 5 + scores[1] * sin(2 * pi * t_grid) +
                    scores[2] * cos(4 * pi * t_grid) +
                    scores[3] * sin(6 * pi * t_grid) +
                    rnorm(m, sd = 0.3)
}

# Phase II: 50 observations, shift at observation 30
n_new <- 50
X_new <- matrix(0, n_new, m)
for (i in seq_len(n_new)) {
  scores <- rnorm(3, sd = c(2, 1, 0.5))
  shift <- if (i >= 30) 1.5 * sin(2 * pi * t_grid) else 0
  X_new[i, ] <- 5 + scores[1] * sin(2 * pi * t_grid) +
                     scores[2] * cos(4 * pi * t_grid) +
                     scores[3] * sin(6 * pi * t_grid) +
                     shift + rnorm(m, sd = 0.3)
}

fd_ic <- fdata(X_ic, argvals = t_grid)
fd_new <- fdata(X_new, argvals = t_grid)
```

```{r plot-data}
plot(fd_ic, main = "Phase I: In-Control Profiles",
     xlab = "Measurement Position", ylab = "Thickness")
```

## Phase I: Estimating Control Limits

`spm.phase1()` performs FPCA on the in-control data, computes $T^2$ and SPE
statistics for each training observation, and estimates control limits at
significance level $\alpha$.

```{r phase1}
chart <- spm.phase1(fd_ic, ncomp = 3, alpha = 0.05, seed = 42)
print(chart)
```

The returned `spm.chart` object contains:

- **T2 control limit** — based on the $F$-distribution with `ncomp` and
  `n - ncomp` degrees of freedom
- **SPE control limit** — estimated from the empirical distribution of
  Phase I SPE values
- **Phase I statistics** — $T^2$ and SPE for each training observation
  (useful for checking that Phase I data is truly in-control)

```{r phase1-plot}
plot(chart)
```

## Phase II: Online Monitoring

Project new observations onto the Phase I principal components and compare
to control limits:

```{r phase2}
monitor <- spm.monitor(chart, fd_new)
print(monitor)

# Which observations triggered alarms?
alarm_idx <- which(monitor$t2.alarm | monitor$spe.alarm)
cat("Alarms at observations:", alarm_idx, "\n")
```

```{r phase2-plot}
plot(monitor)
```

The monitoring result contains:

| Field | Description |
|-------|-------------|
| `t2` | Hotelling $T^2$ values for each new observation |
| `spe` | SPE values for each new observation |
| `t2.alarm` | Logical: did $T^2$ exceed the UCL? |
| `spe.alarm` | Logical: did SPE exceed the UCL? |
| `scores` | PC scores for the new observations |

## EWMA Monitoring

The Exponentially Weighted Moving Average (EWMA) chart smooths the monitoring
statistics over time, making it more sensitive to small, persistent shifts
than the standard Shewhart-type chart:

```{r ewma}
ewma_result <- spm.ewma(chart, fd_new, lambda = 0.2, alpha = 0.05)

cat("EWMA T2 alarms:", sum(ewma_result$t2.alarm), "\n")
cat("EWMA SPE alarms:", sum(ewma_result$spe.alarm), "\n")
```

The smoothing parameter $\lambda \in (0, 1]$ controls the memory:
- $\lambda = 1$ gives the standard Shewhart chart (no smoothing)
- $\lambda = 0.2$ (default) gives moderate smoothing
- Smaller $\lambda$ detects smaller shifts but responds more slowly

## Multivariate FPCA

When multiple functional variables are measured simultaneously (e.g.,
temperature, pressure, and flow rate profiles), Multivariate FPCA
extracts joint modes of variation across all variables:

```{r mfpca}
# Simulate two correlated functional variables
fd_var1 <- fdata(X_ic, argvals = t_grid)
fd_var2 <- fdata(X_ic + matrix(rnorm(n_ic * m, sd = 0.5), n_ic, m),
                 argvals = t_grid)

mfpca_result <- mfpca(list(fd_var1, fd_var2), ncomp = 3)
cat("Eigenvalues:", round(mfpca_result$eigenvalues, 3), "\n")
```

The result contains joint scores and per-variable eigenfunctions, which
can be used for multivariate SPM via `mf.spm.phase1()`.

## Contribution Diagnostics

When an alarm fires, contribution analysis identifies *which* region of
the curve or *which* variable caused the alarm:

```{r contributions}
# T2 contributions show which PC scores are responsible
t2_contrib <- spm.contributions(monitor$scores,
                                chart$eigenvalues,
                                grid.sizes = c(m),
                                type = "t2")

# Each row = observation, each column = variable/region contribution
cat("Contribution of PC1 to observation 35:", round(t2_contrib[35, 1], 3), "\n")
```

## Comparison of Methods

| Method | Detects | Best for |
|--------|---------|----------|
| **T2** | Mean shifts in the PC score space | Location changes in dominant modes |
| **SPE** | Changes not captured by the PCs | New types of variation, sensor drift |
| **EWMA** | Small persistent shifts | Gradual process degradation |
| **FRCC** | Shifts in the relationship between curves and covariates | Regression-based monitoring |

## See Also

- `vignette("articles/fpca")` — Functional PCA fundamentals
- `vignette("articles/streaming-depth")` — Online depth-based monitoring
- `vignette("articles/outlier-detection")` — Depth-based outlier detection

## References

- Colosimo, B.M. and Pacella, M. (2010). A comparison study of control charts
  for statistical monitoring of functional data. *International Journal of
  Production Research*, 48(6), 1575--1601.
- Paynabar, K., Jin, J., and Pacella, M. (2013). Monitoring and diagnosis of
  multichannel nonlinear profiles using uncorrelated multilinear principal
  component analysis. *IIE Transactions*, 45(11), 1235--1247.
- Grasso, M., Colosimo, B.M., and Pacella, M. (2014). Profile monitoring via
  sensor fusion: the use of PCA methods for multi-channel data.
  *International Journal of Production Research*, 52(20), 6110--6135.
