---
title: "Streaming Depth Computation"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Standard depth computations require $O(N^2 T)$ time for $N$ curves on $T$ grid points, which becomes expensive for large reference samples. **Streaming depth** decouples reference-set construction from query evaluation: the reference data is pre-sorted at each time point, enabling $O(T \log N)$ depth per query curve.

This is particularly useful for:

- **Online monitoring**: Evaluating new curves as they arrive against a fixed reference
- **Large reference samples**: When $N$ is large, the $O(\log N)$ lookup is much faster than $O(N)$

## Self-Depth (Batch)

The simplest use case computes the depth of each curve relative to the full dataset:

```{r batch}
library(fdars)
set.seed(42)

# Generate functional data
n <- 50
m <- 100
argvals <- seq(0, 1, length.out = m)
data <- matrix(0, n, m)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.3) + rnorm(m, 0, 0.05)
}

# Add an outlier
data[1, ] <- 5 * cos(4 * pi * argvals)
fd <- fdata(data, argvals = argvals)

# Compute streaming FM depth
d_fm <- streaming.depth(fd, method = "FM")
head(round(d_fm, 4))
```

The outlier (curve 1) has the lowest depth:

```{r outlier}
cat("Outlier depth:", round(d_fm[1], 4), "\n")
cat("Median depth:", round(median(d_fm[-1]), 4), "\n")
```

## Monitoring New Curves

A key application is evaluating new observations against a fixed reference sample:

```{r monitoring}
# Split into reference and new data
ref <- fd[1:40]
new_curves <- fd[41:50]

# Compute depth of new curves against reference
d_new <- streaming.depth(new_curves, fdataori = ref, method = "FM")
round(d_new, 4)
```

Curves with unusually low depth may be considered anomalous.

## Comparing FM, MBD, and BD

Three streaming depth methods are available:

- **FM** (Fraiman-Muniz): Integrates pointwise depth ranks; robust general-purpose
- **MBD** (Modified Band Depth): Based on fraction of time a curve is within bands; always in [0, 1]
- **BD** (Band Depth): All-or-nothing band containment; more conservative

```{r comparison}
d_fm <- streaming.depth(fd, method = "FM")
d_mbd <- streaming.depth(fd, method = "MBD")
d_bd <- streaming.depth(fd, method = "BD")

# Compare rankings
df <- data.frame(
  curve = 1:n,
  FM = d_fm,
  MBD = d_mbd,
  BD = d_bd
)

# Correlations between methods
round(cor(df[, -1]), 3)
```

The methods generally agree on ranking, though MBD and FM tend to be more concordant:

```{r rank-plot}
par(mfrow = c(1, 2))
plot(rank(d_fm), rank(d_mbd), pch = 19,
     xlab = "FM rank", ylab = "MBD rank", main = "FM vs MBD")
abline(0, 1, col = "red")

plot(rank(d_fm), rank(d_bd), pch = 19,
     xlab = "FM rank", ylab = "BD rank", main = "FM vs BD")
abline(0, 1, col = "red")
```

## Integration with the Depth Dispatcher

Streaming depth is available through the `depth()` dispatcher:

```{r dispatcher}
d <- depth(fd, method = "streaming")
all.equal(d, streaming.depth(fd, method = "FM"))
```

The `depth.streaming()` alias also works:

```{r alias}
d2 <- depth.streaming(fd)
all.equal(d, d2)
```

## References

- Fraiman, R. and Muniz, G. (2001). Trimmed means for functional data. *Test*, 10(2):419--440.
- Lopez-Pintado, S. and Romo, J. (2009). On the concept of depth for functional data. *Journal of the American Statistical Association*, 104(486):718--734.
