---
title: "Streaming Depth Computation"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Standard depth computations require $O(N^2 T)$ time for $N$ curves on $T$ grid points, which becomes expensive for large reference samples. **Streaming depth** decouples reference-set construction from query evaluation: the reference data is pre-sorted at each time point, enabling $O(T \log N)$ depth per query curve.

This is particularly useful for:

- **Online monitoring**: Evaluating new curves as they arrive against a fixed reference
- **Large reference samples**: When $N$ is large, the $O(\log N)$ lookup is much faster than $O(N)$

## Self-Depth (Batch)

The simplest use case computes the depth of each curve relative to the full dataset:

```{r batch}
library(fdars)
set.seed(42)

# Generate functional data
n <- 50
m <- 100
argvals <- seq(0, 1, length.out = m)
data <- matrix(0, n, m)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.3) + rnorm(m, 0, 0.05)
}

# Add an outlier
data[1, ] <- 5 * cos(4 * pi * argvals)
fd <- fdata(data, argvals = argvals)

# Compute streaming FM depth
d_fm <- streaming.depth(fd, method = "FM")
head(round(d_fm, 4))
```

The outlier (curve 1) has the lowest depth:

```{r outlier}
cat("Outlier depth:", round(d_fm[1], 4), "\n")
cat("Median depth:", round(median(d_fm[-1]), 4), "\n")
```

## Monitoring New Curves

A key application is evaluating new observations against a fixed reference sample:

```{r monitoring}
# Split into reference and new data
ref <- fd[1:40]
new_curves <- fd[41:50]

# Compute depth of new curves against reference
d_new <- streaming.depth(new_curves, fdataori = ref, method = "FM")
round(d_new, 4)
```

Curves with unusually low depth may be considered anomalous.

## Comparing FM, MBD, and BD

Three streaming depth methods are available:

- **FM** (Fraiman-Muniz): Integrates pointwise depth ranks; robust general-purpose
- **MBD** (Modified Band Depth): Based on fraction of time a curve is within bands; always in [0, 1]
- **BD** (Band Depth): All-or-nothing band containment; more conservative

```{r comparison}
d_fm <- streaming.depth(fd, method = "FM")
d_mbd <- streaming.depth(fd, method = "MBD")
d_bd <- streaming.depth(fd, method = "BD")

# Compare rankings
df <- data.frame(
  curve = 1:n,
  FM = d_fm,
  MBD = d_mbd,
  BD = d_bd
)

# Correlations between methods
round(cor(df[, -1]), 3)
```

The methods generally agree on ranking, though MBD and FM tend to be more concordant:

```{r rank-plot}
par(mfrow = c(1, 2))
plot(rank(d_fm), rank(d_mbd), pch = 19,
     xlab = "FM rank", ylab = "MBD rank", main = "FM vs MBD")
abline(0, 1, col = "red")

plot(rank(d_fm), rank(d_bd), pch = 19,
     xlab = "FM rank", ylab = "BD rank", main = "FM vs BD")
abline(0, 1, col = "red")
```

## Outlier Detection

Depth naturally provides an outlier score: curves with low depth are far from
the centre of the distribution. A simple threshold on the depth values identifies
magnitude outliers (shifted too high or too low) and shape outliers (unusual
patterns).

### Threshold-Based Detection

```{r outlier-threshold}
# Depth-based threshold: flag curves below the 5th percentile
threshold <- quantile(d_fm, 0.05)
outlier_idx <- which(d_fm < threshold)
cat("Outlier indices:", outlier_idx, "\n")
cat("Outlier depths:", round(d_fm[outlier_idx], 4), "\n")
```

### Combining with the Functional Boxplot

The functional boxplot uses depth to define a central region and fences.
Streaming depth can serve as the depth engine:

```{r outlier-boxplot, fig.height = 5}
bp <- boxplot(fd, depth.func = depth.streaming)
cat("Functional boxplot outliers:", bp$outliers, "\n")
```

### Magnitude vs Shape Outliers

Different outlier types have different depth signatures. **Magnitude outliers**
(vertically shifted) are detected by all depth methods. **Shape outliers**
(unusual pattern but similar range) are harder — MBD is particularly
sensitive to them:

```{r outlier-types, fig.height = 5}
set.seed(42)
n <- 40
data_out <- matrix(0, n, m)
for (i in 1:n) {
  data_out[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.2) + rnorm(m, 0, 0.05)
}

# Add a magnitude outlier (shifted up)
data_out[1, ] <- sin(2 * pi * argvals) + 3

# Add a shape outlier (different frequency)
data_out[2, ] <- sin(6 * pi * argvals)

fd_out <- fdata(data_out, argvals = argvals)

d_fm_out <- streaming.depth(fd_out, method = "FM")
d_mbd_out <- streaming.depth(fd_out, method = "MBD")

cat("Magnitude outlier (curve 1):\n")
cat("  FM depth: ", round(d_fm_out[1], 4),
    "  MBD depth:", round(d_mbd_out[1], 4), "\n")
cat("Shape outlier (curve 2):\n")
cat("  FM depth: ", round(d_fm_out[2], 4),
    "  MBD depth:", round(d_mbd_out[2], 4), "\n")
cat("Typical curve (curve 3):\n")
cat("  FM depth: ", round(d_fm_out[3], 4),
    "  MBD depth:", round(d_mbd_out[3], 4), "\n")
```

```{r outlier-types-plot, fig.height = 5}
cols <- rep("grey70", n)
cols[1] <- "red"
cols[2] <- "blue"
matplot(argvals, t(fd_out$data), type = "l", lty = 1, col = cols,
        xlab = "t", ylab = "value",
        main = "Magnitude outlier (red) vs Shape outlier (blue)")
legend("topright", c("Magnitude", "Shape", "Normal"),
       col = c("red", "blue", "grey70"), lty = 1, cex = 0.8)
```

## Depth-Based Trimming

Streaming depth can drive robust summary statistics. The **trimmed mean**
drops the least deep curves before averaging, providing a robust functional
mean that is less sensitive to outliers than `colMeans`:

```{r trimming}
# Use streaming depth for trimmed mean
tm_10 <- trimmed(fd_out, trim = 0.10, method = "FM")
tm_25 <- trimmed(fd_out, trim = 0.25, method = "FM")
raw_mean <- colMeans(fd_out$data)

cat("Max |raw mean - trimmed 10%|:", round(max(abs(raw_mean - tm_10$data)), 4), "\n")
cat("Max |raw mean - trimmed 25%|:", round(max(abs(raw_mean - tm_25$data)), 4), "\n")
```

```{r trimming-plot, fig.height = 5}
plot(argvals, raw_mean, type = "l", col = "black", lwd = 2,
     ylim = range(c(raw_mean, tm_25$data)),
     xlab = "t", ylab = "value",
     main = "Effect of Depth-Based Trimming on the Mean")
lines(argvals, as.numeric(tm_10$data), col = "steelblue", lwd = 2)
lines(argvals, as.numeric(tm_25$data), col = "firebrick", lwd = 2)
lines(argvals, sin(2 * pi * argvals), col = "green3", lwd = 1, lty = 2)
legend("topright",
       c("Raw mean", "Trimmed 10%", "Trimmed 25%", "True signal"),
       col = c("black", "steelblue", "firebrick", "green3"),
       lwd = c(2, 2, 2, 1), lty = c(1, 1, 1, 2), cex = 0.8)
```

With outliers present, trimming at 25% recovers a mean much closer to the
true generating function.

## Process Monitoring

A practical use case for streaming depth is **statistical process control**:
a reference sample defines "normal" behaviour, and incoming curves are flagged
when their depth falls below a control limit.

```{r process-monitoring}
set.seed(42)

# Phase 1: Establish reference from in-control process
n_ref <- 100
data_ref <- matrix(0, n_ref, m)
for (i in 1:n_ref) {
  data_ref[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.2) + rnorm(m, 0, 0.05)
}
fd_ref <- fdata(data_ref, argvals = argvals)

# Compute self-depth to establish control limit
d_ref <- streaming.depth(fd_ref, method = "FM")
control_limit <- quantile(d_ref, 0.01)  # 1st percentile
cat("Control limit (1st percentile):", round(control_limit, 4), "\n")
```

```{r process-monitoring-sim}
# Phase 2: Monitor 50 new curves (some from a shifted process)
n_new <- 50
data_new <- matrix(0, n_new, m)
shift_point <- 35  # process shifts at curve 35
for (i in 1:n_new) {
  if (i < shift_point) {
    # In-control
    data_new[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.2) + rnorm(m, 0, 0.05)
  } else {
    # Out-of-control: mean shift
    data_new[i, ] <- sin(2 * pi * argvals) + 0.8 + rnorm(1, 0, 0.2) + rnorm(m, 0, 0.05)
  }
}
fd_new <- fdata(data_new, argvals = argvals)

# Evaluate each new curve against reference
d_monitor <- streaming.depth(fd_new, fdataori = fd_ref, method = "FM")
alarms <- which(d_monitor < control_limit)
```

```{r process-monitoring-plot, fig.height = 5}
plot(1:n_new, d_monitor, type = "b", pch = 19, cex = 0.7,
     col = ifelse(d_monitor < control_limit, "red", "black"),
     xlab = "Curve index", ylab = "Depth",
     main = "Streaming Depth Control Chart")
abline(h = control_limit, col = "red", lty = 2, lwd = 1.5)
abline(v = shift_point - 0.5, col = "grey50", lty = 3)
text(shift_point + 5, max(d_monitor) * 0.95, "Process shift", col = "grey50", cex = 0.8)
legend("bottomleft", c("In control", "Alarm", "Control limit"),
       col = c("black", "red", "red"), pch = c(19, 19, NA),
       lty = c(NA, NA, 2), cex = 0.8)
cat("Alarms at curves:", alarms, "\n")
cat("First alarm:", min(alarms), "(shift at", shift_point, ")\n")
```

The depth chart detects the process shift shortly after it occurs. Unlike
univariate control charts, this approach monitors the entire curve shape
simultaneously.

## Depth-Weighted Density Estimation

Depth values can weight curves for density estimation or visualization,
giving more influence to central curves and down-weighting outliers:

```{r depth-weighted, fig.height = 5}
# Use depth as weights for a weighted mean
d_weights <- d_fm / sum(d_fm)
weighted_mean <- colSums(fd$data * d_weights)
simple_mean <- colMeans(fd$data)

plot(argvals, simple_mean, type = "l", col = "black", lwd = 2,
     ylim = range(c(simple_mean, weighted_mean)),
     xlab = "t", ylab = "value",
     main = "Simple Mean vs Depth-Weighted Mean")
lines(argvals, weighted_mean, col = "steelblue", lwd = 2)
legend("topright", c("Simple mean", "Depth-weighted mean"),
       col = c("black", "steelblue"), lwd = 2, cex = 0.8)
```

The depth-weighted mean down-weights the outlier (curve 1), producing a more
representative central tendency.

## Performance: Streaming vs Standard Depth

Streaming depth is designed for the scenario where a reference sample is
fixed and many query curves must be evaluated. The pre-sorting step is
$O(N T \log N)$, but each subsequent query is only $O(T \log N)$ instead of
$O(N T)$:

```{r performance}
set.seed(42)
sizes <- c(50, 100, 500, 1000)
m_perf <- 100
argvals_perf <- seq(0, 1, length.out = m_perf)

cat("Streaming depth timing (self-depth, FM method):\n")
for (n_size in sizes) {
  data_perf <- matrix(rnorm(n_size * m_perf), n_size, m_perf)
  fd_perf <- fdata(data_perf, argvals = argvals_perf)
  t_elapsed <- system.time(streaming.depth(fd_perf, method = "FM"))["elapsed"]
  cat(sprintf("  N = %4d: %.4f sec\n", n_size, t_elapsed))
}
```

For large reference samples with many incoming queries, the streaming approach
provides substantial speedup over recomputing standard depth from scratch.

## Integration with the Depth Dispatcher

Streaming depth is available through the `depth()` dispatcher:

```{r dispatcher}
d <- depth(fd, method = "streaming")
all.equal(d, streaming.depth(fd, method = "FM"))
```

The `depth.streaming()` alias also works:

```{r alias}
d2 <- depth.streaming(fd)
all.equal(d, d2)
```

## See Also

- `vignette("depth-functions", package = "fdars")` — functional depth measures and ranking
- `vignette("outlier-detection", package = "fdars")` — functional outlier detection methods

## References

- Fraiman, R. and Muniz, G. (2001). Trimmed means for functional data. *Test*, 10(2):419--440.
- Lopez-Pintado, S. and Romo, J. (2009). On the concept of depth for functional data. *Journal of the American Statistical Association*, 104(486):718--734.
