---
title: "Functional Tolerance Bands"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

A **tolerance band** for functional data is a region expected to contain a given fraction of individual curves in the population -- the functional analogue of classical tolerance intervals. Unlike confidence bands (which target the mean), tolerance bands characterize the spread of individual curves.

**fdars** provides five tolerance band methods:

| Method | Key Properties |
|--------|---------------|
| FPCA | Bootstrap on PC scores; pointwise or simultaneous |
| Conformal | Distribution-free; uses calibration/training split |
| SCB (Degras) | Simultaneous confidence band for the mean |
| Exponential family | For non-Gaussian data (Binomial, Poisson) |
| Elastic | Alignment-based; removes phase variability first |

## Generate Sample Data

```{r setup}
library(fdars)
set.seed(42)

argvals <- seq(0, 1, length.out = 50)
n <- 60
data <- matrix(0, n, 50)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.3) +
               rnorm(50, 0, 0.1)
}
fd <- fdata(data, argvals = argvals)
plot(fd)
```

## FPCA Bootstrap Band

The FPCA method reconstructs curves from their principal component scores and uses bootstrap resampling to estimate quantiles. Two types are available:

- **Pointwise**: Independent interval at each evaluation point (narrower)
- **Simultaneous**: Single scaling factor across all points (wider, controls family-wise error)

```{r fpca}
band_pw <- tolerance.band(fd, method = "fpca", coverage = 0.95,
                          band.type = "pointwise", nb = 200, seed = 42)
print(band_pw)
plot(band_pw)
```

```{r fpca-sim}
band_sim <- tolerance.band(fd, method = "fpca", coverage = 0.95,
                           band.type = "simultaneous", nb = 200, seed = 42)
plot(band_sim)
```

## Conformal Prediction Band

The conformal method is **distribution-free**: it makes no parametric assumptions about the data-generating process. It splits data into a training set and calibration set, computing non-conformity scores on the calibration set to determine band width.

Two score types are available:
- **supnorm**: Maximum deviation (constant band width)
- **l2**: Integrated squared deviation (variable band width)

```{r conformal}
band_conf <- tolerance.band(fd, method = "conformal", coverage = 0.95,
                            score.type = "supnorm", seed = 42)
plot(band_conf)
```

## SCB Degras (Mean Confidence Band)

The Degras method constructs a **simultaneous confidence band for the mean function** using a multiplier bootstrap. This is technically a confidence band rather than a tolerance band, but is useful for assessing whether the mean is well-estimated.

```{r scb}
band_scb <- tolerance.band(fd, method = "scb", coverage = 0.95,
                           nb = 200, seed = 42)
plot(band_scb)
```

## Exponential Family Band

For data from exponential family distributions (e.g., count data), the exponential family method applies the appropriate link function transformation before computing FPCA bands:

```{r exponential}
# Gaussian data (identity link)
band_exp <- tolerance.band(fd, method = "exponential", family = "gaussian",
                           nb = 200, seed = 42)
plot(band_exp)
```

## Elastic (Alignment-Based) Band

The elastic method first aligns curves using the Karcher mean in the elastic metric, removing phase variability. It then computes an FPCA tolerance band on the aligned data. This produces tighter bands when curves exhibit timing differences:

```{r elastic}
# Generate data with phase variability
set.seed(42)
data_phase <- matrix(0, n, 50)
for (i in 1:n) {
  shift <- runif(1, -0.05, 0.05)
  data_phase[i, ] <- sin(2 * pi * (argvals - shift)) + rnorm(1, 0, 0.2) +
                     rnorm(50, 0, 0.05)
}
fd_phase <- fdata(data_phase, argvals = argvals)

band_elastic <- tolerance.band(fd_phase, method = "elastic", coverage = 0.95,
                               nb = 200, max.iter = 10, seed = 42)
print(band_elastic)
plot(band_elastic)
```

Compare the elastic band to FPCA on the same phase-shifted data:

```{r elastic-compare, fig.height = 4}
band_fpca_phase <- tolerance.band(fd_phase, method = "fpca", coverage = 0.95,
                                  nb = 200, seed = 42)
cat("FPCA mean half-width:   ", round(mean(band_fpca_phase$half_width), 4), "\n")
cat("Elastic mean half-width:", round(mean(band_elastic$half_width), 4), "\n")
```

The elastic band is narrower because alignment concentrates variance into amplitude rather than spreading it across both amplitude and phase.

## Method Comparison

```{r compare, fig.height = 8}
methods <- c("fpca", "conformal", "scb")
bands <- lapply(methods, function(m) {
  tolerance.band(fd, method = m, coverage = 0.95, nb = 100, seed = 42)
})

par(mfrow = c(3, 1), mar = c(3, 4, 2, 1))
for (i in seq_along(methods)) {
  b <- bands[[i]]
  plot(argvals, b$center, type = "l", col = "blue", lwd = 2,
       ylim = range(c(b$lower, b$upper)),
       main = paste0(methods[i], " (mean width: ",
                     round(mean(b$upper - b$lower), 3), ")"),
       xlab = "", ylab = "value")
  polygon(c(argvals, rev(argvals)),
          c(b$lower, rev(b$upper)),
          col = rgb(0.27, 0.51, 0.71, 0.3), border = NA)
}
```

## References

- Rathnayake, L.N. and Cuevas, A. (2016). Tolerance bands for functional data. *Technometrics*, 58(3):326--334.
- Lei, J. and Wasserman, L. (2014). Distribution-free prediction bands for non-parametric regression. *Journal of the Royal Statistical Society: Series B*, 76(1):71--96.
- Degras, D. (2011). Simultaneous confidence bands for nonparametric regression with functional data. *Statistica Sinica*, 21(4):1735--1765.
