---
title: "Functional Tolerance Bands"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

A **tolerance band** for functional data is a region expected to contain a given fraction of individual curves in the population -- the functional analogue of classical tolerance intervals. Unlike confidence bands (which target the mean), tolerance bands characterize the spread of individual curves.

**fdars** provides five tolerance band methods:

| Method | Key Properties |
|--------|---------------|
| FPCA | Bootstrap on PC scores; pointwise or simultaneous |
| Conformal | Distribution-free; uses calibration/training split |
| SCB (Degras) | Simultaneous confidence band for the mean |
| Exponential family | For non-Gaussian data (Binomial, Poisson) |
| Elastic | Alignment-based; removes phase variability first |

## Generate Sample Data

```{r setup}
library(fdars)
set.seed(42)

argvals <- seq(0, 1, length.out = 50)
n <- 60
data <- matrix(0, n, 50)
for (i in 1:n) {
  data[i, ] <- sin(2 * pi * argvals) + rnorm(1, 0, 0.3) +
               rnorm(50, 0, 0.1)
}
fd <- fdata(data, argvals = argvals)
plot(fd)
```

## FPCA Bootstrap Band

The FPCA method reconstructs curves from their principal component scores and uses bootstrap resampling to estimate quantiles. Two types are available:

- **Pointwise**: Independent interval at each evaluation point (narrower)
- **Simultaneous**: Single scaling factor across all points (wider, controls family-wise error)

```{r fpca}
band_pw <- tolerance.band(fd, method = "fpca", coverage = 0.95,
                          band.type = "pointwise", nb = 200, seed = 42)
print(band_pw)
plot(band_pw)
```

```{r fpca-sim}
band_sim <- tolerance.band(fd, method = "fpca", coverage = 0.95,
                           band.type = "simultaneous", nb = 200, seed = 42)
plot(band_sim)
```

## Conformal Prediction Band

The conformal method is **distribution-free**: it makes no parametric assumptions about the data-generating process. It splits data into a training set and calibration set, computing non-conformity scores on the calibration set to determine band width.

Two score types are available:
- **supnorm**: Maximum deviation (constant band width)
- **l2**: Integrated squared deviation (variable band width)

```{r conformal}
band_conf <- tolerance.band(fd, method = "conformal", coverage = 0.95,
                            score.type = "supnorm", seed = 42)
plot(band_conf)
```

## SCB Degras (Mean Confidence Band)

The Degras method constructs a **simultaneous confidence band for the mean function** using a multiplier bootstrap. This is technically a confidence band rather than a tolerance band, but is useful for assessing whether the mean is well-estimated.

```{r scb}
band_scb <- tolerance.band(fd, method = "scb", coverage = 0.95,
                           nb = 200, seed = 42)
plot(band_scb)
```

## Exponential Family Band

For data from exponential family distributions (e.g., count data), the exponential family method applies the appropriate link function transformation before computing FPCA bands:

```{r exponential}
# Gaussian data (identity link)
band_exp <- tolerance.band(fd, method = "exponential", family = "gaussian",
                           nb = 200, seed = 42)
plot(band_exp)
```

## Method Comparison

```{r compare, fig.height = 8}
methods <- c("fpca", "conformal", "scb")
bands <- lapply(methods, function(m) {
  tolerance.band(fd, method = m, coverage = 0.95, nb = 100, seed = 42)
})

par(mfrow = c(3, 1), mar = c(3, 4, 2, 1))
for (i in seq_along(methods)) {
  b <- bands[[i]]
  plot(argvals, b$center, type = "l", col = "blue", lwd = 2,
       ylim = range(c(b$lower, b$upper)),
       main = paste0(methods[i], " (mean width: ",
                     round(mean(b$upper - b$lower), 3), ")"),
       xlab = "", ylab = "value")
  polygon(c(argvals, rev(argvals)),
          c(b$lower, rev(b$upper)),
          col = rgb(0.27, 0.51, 0.71, 0.3), border = NA)
}
```
