---
title: "TSRVF: Linearized Elastic Analysis"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

After elastic alignment, the aligned curves live on a **nonlinear manifold** --
the quotient space $\mathcal{F} / \Gamma$ of functions modulo reparameterization.
Standard linear operations (PCA, regression, hypothesis testing) are not directly
valid on this curved space.

The **Transported Square-Root Velocity Function (TSRVF)** solves this by mapping
aligned curves to a **tangent space** at the Karcher mean, where the geometry is
locally Euclidean. In this tangent space, standard multivariate methods apply
directly, giving a rigorous framework for **elastic functional data analysis**.

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
library(patchwork)
theme_set(theme_minimal())
```

```{=html}
<img src="../reference/figures/tsrvf-diagram.svg" alt="TSRVF: From Curved Manifold to Flat Tangent Space" style="width:100%; max-width:700px;">
```

## How It Works (Intuition)

After elastic alignment, curve shapes live on a curved surface (a manifold) --
think of points on the surface of a globe. Standard statistical methods like
PCA assume data lives in a flat space (like a map), so applying them directly
to aligned curves introduces distortion.

The TSRVF solves this by **projecting curves onto a flat tangent plane** at the
mean shape -- like laying a flat map tangent to the globe at one point. Near
the point of contact, the map is a good approximation of the globe's surface.

Each curve becomes a **tangent vector** -- a direction and magnitude showing how
that curve's shape differs from the mean. These tangent vectors live in ordinary
Euclidean space, so you can:

- Run **PCA** to find the main modes of shape variation
- Use **regression** with tangent vectors as predictors
- Apply any **clustering** or **classification** method that works with vectors

The projection is invertible: you can go from tangent vectors back to curves
(by "projecting back onto the globe"), so you can reconstruct curves from
their PC scores or model predictions.

## Mathematical Framework

### The Shape Space

Let $\mathcal{F}$ denote the space of absolutely continuous functions
$f : [0, 1] \to \mathbb{R}$, and let $\Gamma$ be the warping group of
orientation-preserving diffeomorphisms of $[0, 1]$. The **shape space** is the
quotient:

$$\mathcal{S} = \mathcal{F} / \Gamma$$

where two functions $f_1, f_2$ are equivalent if $f_1 = f_2 \circ \gamma$ for
some $\gamma \in \Gamma$. Points in $\mathcal{S}$ represent curve *shapes*,
stripped of parameterization effects.

Via the SRSF transform $q(t) = \text{sign}(f'(t)) \sqrt{|f'(t)|}$, the shape
space is isometric to the quotient of the unit Hilbert sphere
$\mathbb{S}^\infty \subset L^2$ by $\Gamma$. This is a nonlinear manifold with
nontrivial curvature.

### Exponential and Logarithmic Maps

At a point $[\mu]$ on the shape space (represented by the Karcher mean), the
**tangent space** $T_{[\mu]} \mathcal{S}$ is a flat (Euclidean) vector space.
The **exponential map** $\exp_{[\mu]}$ projects from the tangent space onto the
manifold, and the **logarithmic map** $\log_{[\mu]}$ does the reverse:

$$v_i = \log_{[\mu]}([f_i]) \in T_{[\mu]} \mathcal{S}$$

These tangent vectors $v_i$ are the TSRVF representation. They capture how each
curve's *shape* differs from the mean, in a coordinate system where Euclidean
distances approximate geodesic distances on the manifold.

### The TSRVF Transform

Given aligned SRSFs $\tilde{q}_i = (q_i \circ \gamma_i^*) \sqrt{\dot\gamma_i^*}$
and the mean SRSF $\bar{q}$ (the Karcher mean in SRSF space), the TSRVF of the
$i$-th curve is computed via the **inverse exponential map** on the sphere:

$$v_i = \frac{\theta_i}{\sin(\theta_i)} \left( \tilde{q}_i - \cos(\theta_i) \, \bar{q} \right)$$

where $\theta_i = \cos^{-1}(\langle \tilde{q}_i, \bar{q} \rangle_{L^2})$ is the
geodesic distance between $\tilde{q}_i$ and $\bar{q}$ on the sphere.

Key properties:

1. **$v_i \in T_{\bar{q}} \mathbb{S}^\infty$**: the tangent vectors are
   orthogonal to $\bar{q}$, i.e., $\langle v_i, \bar{q} \rangle = 0$
2. **Euclidean distances approximate geodesic distances**: $\|v_i - v_j\|_{L^2}
   \approx d_e(f_i, f_j)$ for curves near the mean
3. **Linear operations are valid**: addition, scalar multiplication, inner
   products in the tangent space have geometric meaning

### Inverse Transform

To reconstruct a curve from its tangent vector $v$, we apply the **exponential
map**:

$$\tilde{q} = \cos(\|v\|) \, \bar{q} + \sin(\|v\|) \, \frac{v}{\|v\|}$$

This gives the aligned SRSF, from which the aligned curve is recovered via
SRSF inversion.

## Quick Start

```{r quick-start}
set.seed(42)
argvals <- seq(0, 1, length.out = 100)

# Simulate curves with amplitude and phase variation
n <- 20
data <- matrix(0, n, 100)
for (i in 1:n) {
  amp <- rnorm(1, 1, 0.2)
  shift <- runif(1, -0.08, 0.08)
  data[i, ] <- amp * sin(2 * pi * (argvals - shift)) +
               0.3 * rnorm(1) * cos(4 * pi * argvals)
}
fd <- fdata(data, argvals = argvals)

# Compute TSRVF
tv <- tsrvf.transform(fd, max.iter = 15, tol = 1e-4)
print(tv)
```

```{r quick-start-plot}
p1 <- plot(fd) + labs(title = "Original Curves")
p2 <- plot(tv, type = "tangent") + labs(title = "TSRVF Tangent Vectors")
p3 <- plot(tv, type = "mean")
p1 + p2 + p3
```

The tangent vectors represent each curve's shape deviation from the mean in a
Euclidean space -- ready for PCA, regression, or any linear method.

## From a Pre-Computed Alignment

If you have already computed a Karcher mean (which is the expensive step), you
can obtain the TSRVF without re-running the alignment:

```{r from-alignment}
km <- karcher.mean(fd, max.iter = 15)
tv_from <- tsrvf.from.alignment(km)

# Same tangent space, no extra alignment cost
cat("Dimensions:", nrow(tv_from$tangent_vectors$data), "x",
    ncol(tv_from$tangent_vectors$data), "\n")
```

## Inverse Transform

The TSRVF is invertible -- you can reconstruct (aligned) curves from their
tangent vectors:

```{r inverse}
recon <- tsrvf.inverse(tv)

# Compare original aligned curves to reconstruction
p1 <- plot(km, type = "aligned") + labs(title = "Aligned Curves")
p2 <- plot(recon) + labs(title = "Reconstructed from TSRVF")
p1 + p2
```

## TSRVF-Based PCA

The main application of TSRVF is enabling standard PCA in the tangent space.
This gives **elastic FPCA** -- principal components that respect the geometry of
the shape space.

```{r tsrvf-pca}
# Standard PCA on tangent vectors
tv_data <- tv$tangent_vectors$data
pca_result <- prcomp(tv_data, center = TRUE, scale. = FALSE)

# Variance explained
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cat("Variance explained by first 3 PCs:",
    round(cumsum(var_explained)[1:3] * 100, 1), "%\n")
```

```{r tsrvf-pca-scores}
scores <- pca_result$x[, 1:2]
df_scores <- data.frame(PC1 = scores[, 1], PC2 = scores[, 2])

ggplot(df_scores, aes(x = PC1, y = PC2)) +
  geom_point(size = 2) +
  labs(title = "TSRVF PCA Scores", x = "PC1", y = "PC2") +
  coord_equal()
```

### Comparison: Raw FPCA vs TSRVF FPCA

Standard FPCA on the raw (unaligned) data confounds amplitude and phase
variation. TSRVF FPCA operates on aligned shapes, so the principal components
capture only amplitude variation:

```{r comparison}
# Raw FPCA
raw_pca <- prcomp(fd$data, center = TRUE, scale. = FALSE)
raw_var <- raw_pca$sdev^2 / sum(raw_pca$sdev^2)

# TSRVF FPCA
tsrvf_var <- var_explained

df_var <- data.frame(
  PC = rep(1:5, 2),
  Variance = c(cumsum(raw_var[1:5]), cumsum(tsrvf_var[1:5])) * 100,
  Method = rep(c("Raw FPCA", "TSRVF FPCA"), each = 5)
)

ggplot(df_var, aes(x = PC, y = Variance, color = Method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative % Variance") +
  scale_color_manual(values = c("Raw FPCA" = "grey50", "TSRVF FPCA" = "steelblue"))
```

TSRVF FPCA typically needs fewer components because it has already removed
phase variation -- the remaining amplitude variation is lower-dimensional.

## When to Use TSRVF

| Scenario | Recommended approach |
|----------|---------------------|
| Exploratory alignment & visualization | `karcher.mean()` / `elastic.align()` |
| PCA on aligned data | `tsrvf.transform()` + `prcomp()` |
| Regression with aligned predictors | `tsrvf.transform()` + standard regression |
| Clustering aligned curves | `tsrvf.transform()` + any Euclidean clustering |
| Classification | TSRVF scores as features |

The key insight: **TSRVF converts a nonlinear shape analysis problem into a
standard multivariate statistics problem**, at the cost of a one-time alignment
step.

## See Also

- `vignette("elastic-alignment", package = "fdars")` -- the underlying alignment
  framework
- `karcher.mean()` -- compute the Karcher mean (prerequisite for TSRVF)
- `alignment.quality()` -- diagnostics for the alignment step
