---
title: "Uncertainty Quantification"
description: >
  Prediction intervals, LOO cross-validation, conformal prediction, and
  bootstrap confidence bands for functional regression models.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty Quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

Point predictions alone are insufficient for critical applications. **fdars**
provides four complementary approaches to uncertainty quantification for
functional regression models:

```{=html}
<img src="../reference/figures/uncertainty-quantification-diagram.svg" alt="Uncertainty Quantification: four methods overview" style="width:100%; max-width:700px;">
```

| Method | Question | Assumptions | Current Models | Function |
|:-------|:---------|:------------|:------:|:---------|
| Prediction Intervals | How uncertain is $\hat{y}$? | Gaussian errors | **lm only** | `fregre.prediction.interval()` |
| Conformal Prediction | How uncertain is $\hat{y}$? | None (distribution-free) | **lm** (expanding) | `fregre.conformal()` |
| LOO Cross-Validation | How well does the model generalise? | None | **lm only** | `fregre.loo()` |
| Bootstrap CI for $\beta(t)$ | Where is the coefficient significant? | None (bootstrap) | lm + logistic | `fregre.bootstrap.ci()` |

### Supported Models

Most methods currently require a `fregre.lm` model (fitted by `fregre.lm()`).
`fregre.bootstrap.ci()` also supports `functional.logistic` models.

**Prediction intervals and LOO-CV** require the hat matrix $\boldsymbol{H}$,
residual standard error $\hat{\sigma}$, and the FPC score matrix
$\boldsymbol{\Xi}$ — internal structures only `fregre.lm` provides.
Nonparametric models like `fregre.np()` do not produce a hat matrix or
parametric variance estimate, so the formulas for prediction intervals and
LOO-CV (which exploit $e_i^{(-i)} = e_i / (1 - h_{ii})$) do not apply.

**Conformal prediction** is different. The algorithm is inherently
model-agnostic — it only needs a `fit`/`predict` function and a set of
calibration residuals. The coverage guarantee
$P(y_{\text{new}} \in C_\alpha) \geq 1 - \alpha$ holds for *any* model, with
no distributional assumptions. The current R wrapper `fregre.conformal()` is
coupled to `fregre.lm` internals (it passes FPC scores to the Rust backend),
but the underlying Rust crate (`fdars-core`) is being extended with
model-specific conformal functions for nonparametric regression
(`fregre.np`), elastic regression, logistic regression, classification, and
generic models via closure-based interfaces. These will be exposed to R as
they become available.

## Setup

```{r setup, include = FALSE}
library(fdars)
library(ggplot2)
theme_set(theme_minimal())
set.seed(42)
```

```{r create-data}
n <- 80
m <- 100
t_grid <- seq(0, 1, length.out = m)

beta_true <- 2 * dnorm(t_grid, 0.3, 0.08) - 1.5 * dnorm(t_grid, 0.7, 0.08)
beta_true <- beta_true / max(abs(beta_true))

X <- matrix(0, n, m)
for (i in 1:n) {
  X[i, ] <- sin(2 * pi * t_grid) + 0.5 * rnorm(1) * cos(4 * pi * t_grid) +
            rnorm(m, sd = 0.1)
}

dt <- t_grid[2] - t_grid[1]
y <- numeric(n)
for (i in 1:n) {
  y[i] <- sum(beta_true * X[i, ]) * dt + rnorm(1, sd = 0.3)
}

fdataobj <- fdata(X, argvals = t_grid)
model <- fregre.lm(fdataobj, y, ncomp = 5)
```

---

## Prediction-Level Uncertainty

```{=html}
<img src="../reference/figures/uq-prediction.svg" alt="Prediction-Level Uncertainty: parametric vs conformal" style="width:100%; max-width:700px;">
```

Both methods answer the same question — *how uncertain is $\hat{y}$ for a new
observation?* — but differ in assumptions and interval shape.

### Parametric Prediction Intervals

*Requires `fregre.lm`.*

**In plain English.** A prediction interval is like an error bar around each
prediction. It says: "the true value will fall in this range with 95%
confidence." Observations that look unusual compared to the training data get
wider intervals — the model is less sure about them.

**Mathematically.** For a new observation with FPC scores
$\boldsymbol{\xi}_{\text{new}}$, the prediction interval accounts for both
estimation uncertainty and irreducible noise:

$$\hat{y}_{\text{new}} \pm t_{n-K-1,\,1-\alpha/2}\;\hat{\sigma}\sqrt{1 + \boldsymbol{\xi}_{\text{new}}^\top (\boldsymbol{\Xi}^\top\boldsymbol{\Xi})^{-1} \boldsymbol{\xi}_{\text{new}}}.$$

The term $\boldsymbol{\xi}_{\text{new}}^\top (\boldsymbol{\Xi}^\top\boldsymbol{\Xi})^{-1} \boldsymbol{\xi}_{\text{new}}$
is the leverage — how far the new point is from the training centre in FPC
score space. High leverage observations receive wider intervals.

```{r prediction-intervals}
pred_int <- fregre.prediction.interval(
  model,
  train.data = fdataobj,
  new.data = fdataobj[1:20],
  confidence = 0.95
)

df_pi <- data.frame(
  Observation = 1:20,
  Observed = y[1:20],
  Fitted = pred_int$predictions,
  Lower = pred_int$lower,
  Upper = pred_int$upper
)

ggplot(df_pi, aes(x = .data$Observation)) +
  geom_ribbon(aes(ymin = .data$Lower, ymax = .data$Upper),
              fill = "#4A90D9", alpha = 0.2) +
  geom_point(aes(y = .data$Observed), color = "#D55E00", size = 2) +
  geom_line(aes(y = .data$Fitted), color = "#4A90D9", linewidth = 1) +
  labs(title = "95% Prediction Intervals",
       subtitle = "Blue band = prediction interval, red points = observed values",
       x = "Observation", y = "y")
```

#### Coverage and Width

```{r coverage}
covered <- df_pi$Observed >= df_pi$Lower & df_pi$Observed <= df_pi$Upper
cat("Coverage:", round(mean(covered) * 100, 1), "%  (expected: 95%)\n")
```

Wider intervals indicate observations far from the training data centre:

```{r interval-width}
df_pi$Width <- df_pi$Upper - df_pi$Lower

ggplot(df_pi, aes(x = .data$Observation, y = .data$Width)) +
  geom_col(fill = "#7B2D8E", alpha = 0.7) +
  labs(title = "Prediction Interval Widths",
       subtitle = "Wider = more uncertain (high leverage)",
       x = "Observation", y = "Interval Width")
```

### Conformal Prediction

*Currently `fregre.lm` only. The Rust backend is being extended to support
`fregre.np`, elastic regression, logistic regression, classification, and
generic models — see [Supported Models](#supported-models) above.*

**In plain English.** Conformal prediction makes no assumptions about the data
distribution. Instead of relying on Gaussian theory, it sets aside a portion
of the training data as a "calibration set", measures how wrong the model is
on those held-out points, and uses the worst-case error as the interval width.
The result: guaranteed coverage (e.g., 90% of future observations will fall
in the interval) no matter what the data looks like.

**Why is it model-agnostic in theory?** The algorithm only needs two things:
(1) a `fit` function that trains a model on data, and (2) a `predict` function
that produces predictions for new data. It computes residuals on the
calibration set, takes the $(1-\alpha)$-quantile, and adds/subtracts that from
predictions. No hat matrix, no variance formula, no distributional assumptions.

**Mathematically.** (Vovk et al., 2005) The split conformal approach provides
finite-sample coverage:

$$P(y_{\text{new}} \in C_\alpha) \geq 1 - \alpha,$$

for any distribution of $(X, y)$. Given calibration residuals
$|e_1|, \ldots, |e_m|$ and quantile
$q = \lceil (1 - \alpha)(m + 1) \rceil / m$-th order statistic, the interval is

$$C_\alpha = \left[\hat{y}_{\text{new}} - q,\;\hat{y}_{\text{new}} + q\right].$$

All intervals have the same width $2q$ — unlike parametric intervals, which
vary with leverage.

```{r conformal}
# Train/test split
train_fd <- fdataobj[1:60]
train_y <- y[1:60]
test_fd <- fdataobj[61:80]
test_y <- y[61:80]

model_train <- fregre.lm(train_fd, train_y, ncomp = 5)

# Conformal prediction intervals (90% coverage)
conf <- fregre.conformal(
  model_train,
  train.data = train_fd,
  train.y = train_y,
  test.data = test_fd,
  cal.fraction = 0.25,
  alpha = 0.10,
  seed = 42
)

cat("Residual quantile:", round(conf$residual_quantile, 4), "\n")
```

```{r conformal-plot}
df_conf <- data.frame(
  Observation = 1:20,
  Observed = test_y,
  Predicted = conf$predictions,
  Lower = conf$lower,
  Upper = conf$upper
)

ggplot(df_conf, aes(x = .data$Observation)) +
  geom_ribbon(aes(ymin = .data$Lower, ymax = .data$Upper),
              fill = "#2E8B57", alpha = 0.2) +
  geom_point(aes(y = .data$Observed), color = "#D55E00", size = 2) +
  geom_line(aes(y = .data$Predicted), color = "#2E8B57", linewidth = 1) +
  labs(title = "90% Conformal Prediction Intervals",
       subtitle = "Green band = conformal interval (distribution-free)",
       x = "Test Observation", y = "y")
```

```{r conformal-coverage}
covered_conf <- df_conf$Observed >= df_conf$Lower &
                df_conf$Observed <= df_conf$Upper
cat("Empirical coverage:", round(mean(covered_conf) * 100, 1),
    "%  (target: 90%)\n")
```

### Comparison: Parametric vs Conformal

| | Parametric | Conformal |
|:---|:-----------|:----------|
| **Assumptions** | Gaussian errors | None |
| **Width** | Variable (leverage-dependent) | Constant |
| **Coverage guarantee** | Exact if assumptions hold | Finite-sample, always |
| **Data cost** | Uses all training data | Sacrifices a calibration fraction |
| **Model support** | `fregre.lm` only | Any model (in theory) |

```{r compare-intervals}
pred_param <- fregre.prediction.interval(
  model_train, train.data = train_fd, new.data = test_fd, confidence = 0.90
)

df_compare <- data.frame(
  Method = rep(c("Parametric", "Conformal"), each = 20),
  Width = c(pred_param$upper - pred_param$lower,
            conf$upper - conf$lower)
)

ggplot(df_compare, aes(x = .data$Method, y = .data$Width, fill = .data$Method)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Parametric" = "#4A90D9",
                               "Conformal" = "#2E8B57")) +
  labs(title = "Interval Width: Parametric vs Conformal",
       subtitle = "Both at 90% nominal level",
       y = "Interval Width") +
  theme(legend.position = "none")
```

---

## Model Assessment: LOO Cross-Validation

*Requires `fregre.lm`.*

**In plain English.** Leave-one-out cross-validation asks: "If I re-fit the
model without observation $i$, how well does it predict that observation?"
Doing this for every observation gives an honest estimate of how the model
will perform on new data. A large gap between in-sample $R^2$ and LOO $R^2$
signals overfitting — the model has memorised the training data.

The clever part: for linear models, you don't actually need to re-fit $n$
times. The hat matrix gives you the answer in one step.

**Mathematically.** The LOO residual for observation $i$ is efficiently
computed via the hat matrix:

$$e_{i}^{(-i)} = \frac{e_i}{1 - h_{ii}},$$

where $h_{ii}$ is the $i$-th diagonal of $\boldsymbol{H} = \boldsymbol{\Xi}(\boldsymbol{\Xi}^\top\boldsymbol{\Xi})^{-1}\boldsymbol{\Xi}^\top$.
High-leverage points ($h_{ii}$ close to 1) have inflated LOO residuals —
removing them changes the fit substantially.

The PRESS statistic is

$$\text{PRESS} = \sum_{i=1}^{n} \left(e_i^{(-i)}\right)^2,$$

and the LOO $R^2 = 1 - \text{PRESS} / \text{TSS}$ provides an honest estimate
of out-of-sample performance.

```{r loo}
loo <- fregre.loo(model, data = fdataobj, y = y)

cat("PRESS:", round(loo$press, 4), "\n")
cat("LOO R\u00b2:", round(loo$loo_r_squared, 4), "\n")
cat("In-sample R\u00b2:", round(model$r.squared, 4), "\n")
```

A large gap between in-sample $R^2$ and LOO $R^2$ indicates overfitting.

```{r loo-residuals}
df_loo <- data.frame(
  Observation = seq_along(loo$loo_residuals),
  LOO_Residual = loo$loo_residuals,
  In_Sample_Residual = model$residuals
)

ggplot(df_loo, aes(x = .data$In_Sample_Residual, y = .data$LOO_Residual)) +
  geom_point(color = "#7B2D8E", size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "gray50") +
  labs(title = "In-Sample vs LOO-CV Residuals",
       subtitle = "Points above the diagonal indicate optimistic in-sample fit",
       x = "In-sample Residual", y = "LOO Residual")
```

```{r loo-leverage}
df_leverage <- data.frame(
  Observation = seq_along(loo$loo_residuals),
  LOO_Residual = abs(loo$loo_residuals),
  In_Sample_Residual = abs(model$residuals)
)
df_leverage$Inflation <- df_leverage$LOO_Residual / df_leverage$In_Sample_Residual

ggplot(df_leverage, aes(x = .data$Observation, y = .data$Inflation)) +
  geom_col(fill = "#7B2D8E", alpha = 0.6) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  labs(title = "LOO Residual Inflation Factor",
       subtitle = "|LOO residual| / |in-sample residual| — high values = high leverage",
       x = "Observation", y = "Inflation (1 / (1 − h_ii))")
```

---

## Coefficient Uncertainty: Bootstrap CI for β(t)

*Supports both `fregre.lm` and `functional.logistic`.*

```{=html}
<img src="../reference/figures/uq-bootstrap.svg" alt="Bootstrap CI: pointwise vs simultaneous bands" style="width:100%; max-width:700px;">
```

**In plain English.** Previous sections quantified uncertainty in *predictions*
($\hat{y}$). This section asks a different question: where on the functional
domain does the predictor actually matter? The functional coefficient
$\beta(t)$ tells you how each part of the curve contributes to the response.
Bootstrap confidence bands show which parts of $\beta(t)$ are reliably
different from zero — those are the regions where the predictor has a
statistically significant effect.

The bootstrap works by resampling the data many times, fitting the model each
time, and observing how $\hat{\beta}(t)$ varies. If it varies a lot at some
$t$, we're uncertain about that part of the coefficient.

**Mathematically.** Given $B$ bootstrap replicates
$\hat{\beta}^{*1}(t), \ldots, \hat{\beta}^{*B}(t)$, we construct two types
of confidence bands:

### Pointwise Bands

At each grid point $t$ independently:

$$CI_{\text{pw}}(t) = \left[\hat{\beta}_{(\alpha/2)}^*(t),\;\hat{\beta}_{(1-\alpha/2)}^*(t)\right],$$

where $\hat{\beta}_{(q)}^*(t)$ is the $q$-quantile of the bootstrap
distribution at $t$. This controls coverage at each $t$ separately but not
simultaneously across all $t$.

```{r bootstrap-ci}
ci <- fregre.bootstrap.ci(model, n.boot = 200, alpha = 0.05, seed = 42)
print(ci)
```

```{r plot-bootstrap-ci}
plot(ci)
```

### Simultaneous Bands

**In plain English.** Pointwise bands control the error at each grid point
separately. If you have 100 grid points and test each at 5%, you'd expect about
5 false positives. Simultaneous bands are wider — they guarantee that the
*entire curve* $\beta(t)$ lies within the band with $(1-\alpha)$ probability.
Regions where the simultaneous band excludes zero are statistically significant
at the chosen level, corrected for the multiple testing across all $t$.

**Mathematically.** The simultaneous band uses the sup-norm critical value
$c_\alpha$ estimated from the bootstrap distribution:

$$CI_{\text{sim}}(t) = \hat{\beta}(t) \pm c_\alpha \cdot \widehat{SE}_{\text{boot}}(t),$$

where $c_\alpha$ is the $(1-\alpha)$-quantile of
$\sup_t |\hat{\beta}^*(t) - \hat{\beta}(t)| / \widehat{SE}_{\text{boot}}(t)$
over the bootstrap replicates.

```{r plot-bootstrap-ci-sim}
plot(ci, simultaneous = TRUE)
```

The simultaneous band is wider because it guarantees
$P(\beta(t) \in CI_{\text{sim}}(t) \;\forall\, t) \geq 1 - \alpha$, rather
than controlling coverage at each $t$ separately.

---

## Conformal Prediction: Beyond Linear Models

The `fregre.conformal()` function currently requires `fregre.lm`, but conformal
prediction is **theoretically model-agnostic**. The algorithm only needs a
fit/predict function and a calibration set — no distributional assumptions, no
hat matrix, no parametric variance estimate.

The Rust backend (`fdars-core`) is being extended with a comprehensive
conformal prediction framework supporting many model types:

| Conformal variant | Model type | Key idea |
|:------------------|:-----------|:---------|
| Split conformal | `fregre.lm` | Calibrate on held-out residuals (current) |
| Split conformal | `fregre.np` | Same algorithm, nonparametric model |
| Split conformal | Elastic regression | Conformal intervals in SRVF space |
| Split conformal | Elastic PCR | Phase-aware conformal intervals |
| Split conformal | Logistic / classification | Conformal prediction sets (labels, not intervals) |
| CV+ conformal | Any regression | Cross-validation instead of a single split |
| Jackknife+ | Any regression | LOO-based, uses all data for training |
| Generic closure | Any `fit`/`predict` | User provides fit and predict closures |

**Why does this matter?**

- **Nonparametric models** (`fregre.np`) have no parametric prediction intervals
  — conformal is the *only* rigorous uncertainty quantification available.
- **Elastic models** operate in SRVF space — standard intervals in function
  space are meaningless; conformal intervals respect the registration geometry.
- **Classification models** can produce prediction *sets* (multiple plausible
  labels with coverage guarantees) instead of just point predictions.

These functions will be exposed to R as they become available in the vendored
crate. The interface will follow the same `fregre.conformal()` pattern, with
automatic dispatch based on model class.

---

## Summary

| Aspect | Prediction Intervals | Conformal | LOO-CV | Bootstrap CI |
|:-------|:---------------------|:----------|:-------|:-------------|
| Target | $\hat{y}$ | $\hat{y}$ | Model fit | $\beta(t)$ |
| Models | lm only | lm (expanding to all) | lm only | lm + logistic |
| Assumptions | Gaussian | None | None | None |
| Width | Variable (leverage) | Constant | — | Variable (SE) |
| Guarantee | Exact if Gaussian | Finite-sample | — | Asymptotic |
| Key output | Interval bounds | Interval bounds | PRESS, LOO $R^2$ | Pointwise + simultaneous bands |

## References

- Allen, D.M. (1974). The relationship between variable selection and data
  augmentation and a method for prediction. *Technometrics*, 16(1), 125--127.

- Ramsay, J.O. and Silverman, B.W. (2005). *Functional Data Analysis*.
  2nd ed. Springer.

- Vovk, V., Gammerman, A. and Shafer, G. (2005). *Algorithmic Learning in
  a Random World*. Springer.

## See Also

- `vignette("explainability")` --- global and local model explanations
- `vignette("regression-diagnostics")` --- influence analysis and VIF
- `vignette("cross-validation")` --- K-fold cross-validation framework
